### 2014-04-05 2014-04-07-adv-bd-mtg/beta.tex
switch to latex, already being switched to keynote

\documentclass[aspectratio=169,14pt]{beamer}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{fontspec}

\usetheme{Singapore}
\usecolortheme[RGB={41,92,35}]{structure}
\setbeamercolor{section in head/foot}{bg=white}
\setbeamercovered{again covered={\opaqueness<1->{50}}}

\defaultfontfeatures{Mapping=tex-text}
%\setsansfont{Sofia Pro}
\setsansfont{Open Sans}

\title[Advisory Board Meeting]{Overview of beta site}
\subtitle{Advisory Board Meeting}
\institute{Databrary}
\date{April 7, 2014}

\pgfdeclareimage[height=1cm]{databrary}{../databrary}
\logo{\pgfuseimage{databrary}}

\begin{document}

\begin{frame}
	\titlepage
\end{frame}

\section{Progress}

\begin{frame}{6 months ago}{Website}
	\begin{itemize}
		\item	Lots of videos
		\begin{itemize}
			\item Piled up into studies
			\item Minimal organization
		\end{itemize}
		\item	``Static'' views
		\begin{itemize}
			\item Single presentation option
			\item Limited flexibility
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Updates}{Website}
	\begin{itemize}
		\item	Revised technical requirements
		\begin{itemize}
			\item Flexible discovery options
			\begin{itemize}
				\item searching: by keyword, text, descriptions, names, etc. (text box)
				\item filtering: limiting results by age, gender, numeric values (side bar)
			\end{itemize}
			\item Efficient, intuitive editing, uploading
			\item High-level visualizations, summaries
			\item API, scripting interface
		\end{itemize}
		\item Built new completely dynamic JS interface, using API
	\end{itemize}
\end{frame}

\begin{frame}{Updates}{Previously collected data}
	\begin{itemize}
		\item Learned a lot from manually curating and organizing existing and new datasets
		\item Broadened types of possible, standard metadata
		\item Developed better ways to group and relate data, studies, datasets
	\end{itemize}
\end{frame}

\begin{frame}{Updates}{Data model}
	\begin{itemize}
		\item There are always exceptions, irregularities in real data
		\begin{itemize}
			\item Missing/multiple videos
			\item Doesn't fit cleanly in defined groups
			\item Excluded and later re-purposed data
			\item Always something \emph{non-rectangular}
		\end{itemize}
		\item Different types of data (classroom, public, longitudinal) have different needs
		\item Need responsive, iterative, ``data-driven'' approach
		\item Better understanding of constraints, required flexibility
	\end{itemize}
\end{frame}

\section{Data structures}

\begin{frame}{Session}{Basic organizational unit}
	\begin{columns}
	\column{2.5in}
	\includegraphics[width=\textwidth]{session.pdf}

	\column{\textwidth-2.25in}
	\begin{itemize}
		\item Covered by a \emph{release level}
		\item Data collected at the same time
		\item Contains timeline of collected raw data files
		\item Annotated with metadata
		\item Analysis files layered on
	\end{itemize}

	\end{columns}
\end{frame}

\begin{frame}{Datasets and Studies}{Data provenance}
	\begin{columns}
	\column{\textwidth-2.95in}
	\begin{itemize}
	\item Datasets: raw data collected directly during a session
	\item Studies: analyses, generated data, published papers
	\end{itemize}

	\column{2.5in}
	\includegraphics[width=2.5in]{studies.png}
	\end{columns}

	\begin{itemize}
	\item Studies collect included sessions from datasets and ``layer'' files over them
	\item Allows re-purposing, re-collecting data
	\end{itemize}
\end{frame}

\section{Walkthrough}

\begin{frame}{Walkthrough}{Live site}
\end{frame}

\section{Next steps}
\begin{frame}<1>[label=browser]{Next steps}{Enhancing the browser}
	\begin{itemize}
		\item<1> Improve searching, filtering, sorting within and across datasets
		\item<1> Have many user stores and use cases to drive this
		\item<2> Mainly focused around browsing, searching, using
		\item<2> Also want to build views that make sense for entering data, and can be reused for editing
	\end{itemize}
\end{frame}

\begin{frame}{User stories}{Education and Teaching}
	\begin{itemize}
		\item Video clips for teaching
		\item Illustrate an idea
		\item Show the range of behaviors and exceptions
		\item Show an excerpt in a talk
	\end{itemize}
\end{frame}
\begin{frame}{User stories}{Pre-research}
	\begin{itemize}
		\item Browse the work in my field
		\item Decide whether a study is worth doing
		\item Preliminary data for grant proposal
		\item Ideas and inspiration
		\item Replicate, expand on, or review previous work based on the procedure or coding manual
	\end{itemize}
\end{frame}
\begin{frame}{User stories}{Research}
	\begin{itemize}
		\item Repurpose videos for new uses
		\item Replicate existing work by recoding videos
		\item Grow sample size
		\item Include participants from other contexts and populations
		\item Conduct integrative analyses
		\item Complete a grant progress report
	\end{itemize}
\end{frame}
\againframe<2>{browser}

\begin{frame}{Databrary 1.0}{Public release}
	\begin{itemize}
		\item Better video, timeline management
		\item Improved browsing within a dataset/study
		\item Uploading as you go
		\item Entering, editing data
		\item Organizing, grouping data
	\end{itemize}
\end{frame}

\end{document}

### 2015-01-14 jcdl-15/paper.tex
Reorg and cleanup jcdl directory

% This is "sig-alternate.tex" V2.0 May 2012
% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.5 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.5) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V2.0 - May 2012

\documentclass{sig-alternate}

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{JCDL}{'15 Knoxville, Tennesse USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Databrary: A digital data library for sharing research video}

\numberofauthors{6}

\author{
\alignauthor
Rick O. Gilmore, Ph.D.\\
	\affaddress{The Pennsylvania State University}\\
	\affaddress{Department of Psychology}\\
	\affadddress{University Park, PA 16802}\\
	\email{rogilmore@psu.edu}

\alignauthor
Lisa A. Steiger\\
	\affaddress{Databrary Project}\\
	\affaddress{New York University}
	\affaddress{196 Mercer Street, Room 807}\\
	\affaddress{New York, NY 10012}\\
	\email{lisa@databrary.org}

\alignauthor
Dylan A. Simon\\
	\affaddress{Databrary Project}\\
	\affaddress{New York University}
	\affaddress{196 Mercer Street, Room 807}\\
	\affaddress{New York, NY 10012}\\
	\email{dylan@databrary.org}

\and

\alignauthor
Drew Gordon\\
	\affaddress{Databrary Project}\\
	\affaddress{New York University}
	\affaddress{196 Mercer Street, Room 807}\\
	\affaddress{New York, NY 10012}\\
	\email{drew@databrary.org}

\alignauthor
David S. Millman\\
	\affaddress{New York University}
	\affaddress{New York, NY 10012}\\
	\email{dsm@nyu.edu}

\alignauthor
Karen E. Adolph, Ph.D.\\
	\affaddress{New York University}
	\affaddress{Department of Psychology}
	\affaddress{New York, NY 10012}\\
	\email{karen.adolph@nyu.edu}

}

\maketitle

\begin{abstract}

Video captures the complexity, richness, and diversity of behavior
unlike any other tool. As a result, large numbers of researchers who
study human or animal behavior employ video. Video documents itself to a
large degree, and therefore has significant potential for data re-use.
Still, researchers rarely share video. Video often contains information
about personal identities, so considerations about research ethics pose
challenges to sharing. The relatively large size of video files and
diversity of formats pose technical challenges. In this paper, we will
describe how the web-based Databrary data library has overcome the most
significant barriers to sharing video within the developmental and
learning sciences community, including solutions to maintaining
participant privacy, data tagging, and data management.
\end{abstract}

**Categories and Subject Descriptors**

\category{H.2.8}{Database Management}{Database Applications}[image
databases, scientific databases]

\terms{Management, Documentation, Design, Security, Legal Aspects}

\keywords{Data sharing, open science, video, psychology, developmental science}

\section{Introduction}

Permission to make digital or hard copies of all or part of this work
for personal or classroom use is granted without fee provided that
copies are not made or distributed for profit or commercial advantage
and that copies bear this notice and the full citation on the first
page. To copy otherwise, or republish, to post on servers or to
redistribute to lists, requires prior specific permission and/or a fee.

Film and video recording have been a mainstay of research in the
developmental sciences for nearly 100 years [CITE]. A typical lab
acquires, on average, 12 hours of video per week depicting behavior in
natural or laboratory environments [CITE]. These recordings constitute a
primary source of raw data. Prior to descriptive or statistical
analyses, researchers score or “code” recordings, mostly manually, using
commercial software or makeshift tools. However, even the most detailed
and laborious video coding does not exhaust the possibilities for
analyses by other researchers with different questions. Thus, sharing
video promises a wide range use cases, perhaps more than sharing
flat-file data or static images. Despite the wealth of untapped
information available in research videos and the labor expended in
collecting and coding them, developmental researchers rarely share data.
In the prevailing culture, investigators conduct research privately
within their own laboratories. Data are destroyed or molder away on a
shelf after the research is published. The Databrary project aims to
change the status quo, making sharing standard practice.

Databrary enables researchers to share video and associated flat-file
data for research and educational purposes. For example, a researcher
interested in language development might reuse a set of video recordings
created originally to study the development of walking. Researchers can
conduct integrative analyses by combining data across disparate studies,
increase sample sizes, verify the feasibility of proposed projects,
compare methods, learn about procedures, and independently verify
results. Similarly, investigators can reuse shared videos for
educational purposes such as to find illustrative examples for teaching,
use exemplars in presentations, show students how to conduct procedures,
or provide supplemental materials for reviewers.

Data sharing and open science practices are emerging or accepted norms
in the biological, physical and earth sciences with demonstrated
benefits for scientific transparency and accelerated discovery [1-4].
Open research practices are gaining traction in the behavioral and
social sciences [5-10]. However, open data sharing among developmental
researchers is still the exception [11] not common practice. In seeking
to expand sharing of video-based data, Databrary faces challenges
concerning the interplay among institutional research policies, ethical
issues related to obtaining participants’ consent for sharing, and
technical authorization systems. This paper describes the current state
of the Databrary project including the project organization, how content
is acquired, stored, preserved, and accessed, policy considerations, and
how the library has affected and hopes to affect the behavioral science
research community.

\section{Project Organization}

Databrary is a joint project of New York University (NYU) and The
Pennsylvania State University (Penn State). The Databrary project began
with a workshop on open video data sharing funded by the National
Science Foundation (NSF) under Grant No. BCS-1139702 at which
developmental scientists, computer scientists, library scientists, and
federal agency program officers discussed the promises and challenges of
sharing video data. NSF and the National Institute of Child Health and
Human Development (NICHD) have provided funding under grant No.
BCS-1238599 and Cooperative Agreement U01-HD-076595, respectively. The
interdisciplinary team bring expertise in developmental and neural
science, information technology, library science, software development,
data curation, and community engagement. In addition, a board of expert
advisors composed of developmental scientists, library scientists, and
leaders of data repositories in the behavioral and social sciences
provide guidance.

The project is organized around five core activities.

\subsection{Curation}

To seed the Databrary, staff identified and selected video datasets for
manual curation. The datasets were selected based on availability,
presumed interest to developmental science community, suitability for
the organization and file structure of the repository, diversity of
study type, and on the availability of appropriate sharing permissions
(see 2.2 below). Manual curation of archival datasets continues under
the direction of a staff member with expertise in library and
information sciences.

Nevertheless, we have found manual data curation time consuming and
difficult, especially for researchers. The more difficult and
time-consuming data sharing becomes, the less likely it will be that
researchers participate despite best intentions. So, we have created
user-friendly spreadsheet-like interfaces that enable researchers to
self-curate datasets as they are collected. At the end of a data
collection session, researchers may upload videos and other data files
collected along with metadata about participant demographic
characteristics or experimental conditions. Databrary transcodes videos
to a standard format in the background, and the system alerts the user
when the video can be previewed. Researchers may also upload study-level
materials -- code books, flat data files used for statistical analyses,
images of experimental equipment, links to related external information
(e.g., GitHub repositories), etc.

This scheme, which we call "upload-as-you-go", makes the data curation
required for sharing part of the normal scientific workflow, not a
burdensome after thought. No data are shared until the researcher grants
access, but unlike the *ad hoc* organization in a typical investigator's
laboratory, on Databrary, the data are organized in a standardized way
that could be readily and easily shared with others when the
investigator is ready to do so.

Sharing video requires special considerations because the data cannot be
de-identified without reducing its value. No previously established
research data repository has specialized in video data sharing.
Therefore, Databrary had to develop a new policy and permission
framework for sharing identifiable and sensitive data while
simultaneously preserving the rights of research participants.

The framework has two components, both of which build on
well-established principles. One extends the idea of informed consent to
participate in research by requiring that participants give explicit
permission for their identifiable or sensitive data to be shared with
other researchers. Databrary makes available on its website template
language researchers may use for seeking permission to share, along with
video demonstrations and additional guidance. So, researchers who wish
to share video must either use the template or demonstrate that
participants have granted an equivalent level of permission. The second
component involves limiting access to individuals who have prior formal
authorization. Researchers must either secure authorization to access
Databrary from their institution, typically a university where they
conduct independent research, or must have another authorized researcher
grant them access. Institution-level authorization requires a formal
agreement with Databrary. Fundamentally, the Databrary framework ensures
that participants control whether other researchers can view their video
(or other identifiable data), and only researchers who pledge to uphold
ethical principles, including maintaining participant confidentiality,
may access Databrary's sensitive materials.

Databrary developed these policies and agreements in close consultation
with Institutional Review Boards (IRBs), sponsored research privacy
administrators, and legal staff at NYU and Penn State, with other
experts in research data ethics, and with Federal officials. See
http://databrary.org/access/guide.html for more details.

The centerpiece of the Databrary project is the repository itself. Our
team of programmers has developed digital library software, including
asset management, workflow, access control, discovery services, and a
user interface. This work is closely coordinated with the central NYU IT
infrastructure providers for services such as high-capacity storage,
high-security storage, and high-performance computing (for video
processing). The long-term goal is to build the cyber-infrastructure for
seamless search, streaming, uploading, and downloading and to provide an
open-source framework for other entities to build on the Databrary
software or extend it to other content domains.

\subsection{Annotation and Metadata}

Video becomes most useful for conducting research when it is transformed
into quantitative and qualitative data in a flat-file format. Thus, the
Databrary team is enhancing a general-purpose video-coding software
application specialized for exploring, annotating, tagging, scoring, and
visualizing the video files in preparation for qualitative and
quantitative analyses. Toward that end, we have released Datavyu
(datayvu.org), a free, open source, multi-platform desktop application.
Datavyu is a fork of the OpenSHAPA tool used by some developmental
researchers. The long-term goal is to build a web-based version of
Datavyu that will provide both data analysis and data management
services required for preparing and organizing video material prior to
its storage in Databrary.

\subsection{Community Practices}

The current zeitgeist in the developmental science community is one of
privacy and data hoarding. A primary activity of the Databrary project
is to introduce the practice of open data sharing to the developmental
science community. We are doing so through announcements to
developmental research societies, colloquia, and conference
presentations. The team participates in workshops to discuss sharing
issues and hosts events to teach potential users about sharing and
analysis tools. We post links on websites frequented by developmental
researchers and participate in user forums. In addition, we have asked
leaders in developmental science to set an example for other researchers
by publicly committing to open video data sharing. More than a hundred
have responded to the call
(http://databrary.org/community/contributors.html). Our long-term goal
is a self-sustaining community of researchers committed to open
science.

\section{Building the Library}

Existing data management practices pose challenges for video sharing.
Despite similar research methods, study designs vary widely. No two
developmental science labs manage data in the same way. Some studies are
longitudinal, where researchers observe the same participants at
multiple sessions. Some are cross-sectional, where researchers observe
each participant at only one session. The timing of observations may be
determined by participants’ age (e.g., 4-month-olds, 8-month-olds,
12-month-olds) or abilities (e.g., preverbal, one-word utterances,
sentences), experimentally determined variables (e.g., pre- and
post-intervention), or other factors. Some labs organize longitudinal
data by grouping files first by participant and then by session date.
Other labs organize longitudinal data by session and then by
participant. Still others organize longitudinal data based on task (book
reading, block building, free play). Some researchers institute a
central data management system to be followed by all the lab members,
providing easier access and greater transparency for the entire lab but
not necessarily providing the structure for similar benefits to
researchers unfamiliar with the lab’s practices. Other researchers allow
their students to keep separate records, making it difficult to share
data even within the lab. Some researchers keep videos, metadata, and
analyses together, and some do not. Idiosyncratic terms, record-keeping,
and data management systems are the norm. Databrary must enable
researchers to discover and understand each other’s materials,
regardless of the original investigator’s data management system.

Databrary takes several approaches to address this challenge: A flexible
data model, informal metadata contributions by users via tagging and
commenting, mediated curation, and incentives to deposit materials more
consistently. Figure 1 illustrates the principal data model. It
accommodates different hierarchies and organizations, such as those
described in the examples above.

The full expressive flexibility of the data model is not evident to
Databrary users. Instead, users see different aspects of the data model
at different times, depending on the context. Researchers organize their
materials by acquisition date and time into structures called sessions.
A session corresponds to a unique recording episode and contains one or
more recordings or related flat files (assets). Researchers can group
sessions in different ways, using whatever groupings are appropriate for
the particular dataset. These groupings may also identify particular
time segments of a recording to distinguish tasks, events, or
participants that comprise the session. In the data model, the aggregate
of these flexible groupings is called a volume. In practice, researchers
combine sessions or segments of sessions within and across datasets to
form the raw material that are subsequently described in published
articles and presentations. Thus, Databrary contributors can combine
sessions or segments within and across datasets with ancillary
materials, such as coding manuals, Datavyu spreadsheets, statistical
analyses, questionnaires, IRB documents, computer code, sample displays,
and links to published journal articles in an aggregate structure
investigators call a study. Like datasets, studies are represented as
volumes in the data model.

The discovery and browsing interfaces offer content items at the volume
(study or dataset) level to users searching the library. The
contributor-assigned groupings (records), which may include relevant
metadata about participants (measures, such as participant demographic
information, domain-specific survey information, location data,
condition variables, task descriptions, or other properties), allow a
second level of filtering and organization within and between relevant
identified studies or datasets. This enables users to quickly identify
data that may be relevant to their own interests or research.

Authorized Databrary Investigators can also add keyword tags at the
study/dataset, session, or segment level, and can endorse or deprecate
keywords that have been suggested by other authorized researchers. These
abilities are critical because different researchers may have very
different reasons for citing a study or using data. Future
implementations may extend the ability of authorized researchers to
annotate studies with other kinds of metadata.

The Databrary team recognizes that other data repositories enforce
strict metadata ontologies and that doing so may have benefits in
sub-domains of research when there is community consensus [12]. We will
support standard data coding ontologies among researchers, but only
enforce standardization for a small set of standard tags such as study
date, participant birth date, and sex. We will also encourage
contributors to report race, ethnicity, primary language, language of
dataset, and location of session. The Databrary team chose not to
require strict metadata ontologies for several reasons. We hope to
reduce the pre-deposit curational demands on contributors, encourage the
repurposing of shared data, and foster the rapid adoption of data
sharing. Developing and achieving agreement on metadata ontologies can
take significant amounts of time. Video data are so rich and complex
that in many domains, researchers have not settled on standard
definitions for particular behaviors and may have little current need
for standard tasks, procedures, or terminology. However, we will also
encourage users to re-use tags and terminology by suggesting common or
similar terms, without confining users to these suggestions. User
communities within Databrary may eventually converge on common
conceptual and metadata ontologies based on the most common (and
commonly endorsed) keyword tags, but standardized ontologies are not
necessary for browsing and searching in most of the use cases we
envision.

Our experiences curating and ingesting archival datasets have
highlighted the considerable value of contributors entering raw video
data into Databrary as soon as recordings are acquired. Immediate
uploading reduces the workload on investigators, minimizes the risk of
data loss and corruption, and accelerates the speed with which materials
become openly available. To encourage immediate uploading, Databrary
provides a complete set of controls so that researchers can restrict
access to their data to only their own labs or to other users of their
choosing. Datasets can be shared at a later point when data collection
and ancillary materials are complete, whenever the contributor is
comfortable sharing, or when journals or funders require it. Databrary
has published a Data Sharing Manifesto [13] that explains to researchers
the Databrary philosophy. Standards about when data should be shared are
evolving. Our philosophy is consistent with concepts and practices in
other domains where data sharing is the norm (e.g., www.iedata.org).
Planned enhancements to the Datavyu tool will allow contributors to
organize video files and metadata as part of the analysis process. This
will facilitate the contribution of packaged datasets to Databrary.

As part of the curation process, Databrary stores at least two versions
of each item of Databrary video content: a copy for access, and the
received original file if it was digital, or a 10-bit YUV digital
preservation copy if the original version was not digital. Currently,
the access version format is H.264 (HiP) with AAC audio in an MPEG-4
container, although we expect the appropriate video formats to change
over time, as has been the case with many digital video formats in
recent years.

For preservation, the original file (if digital) or the preservation
copy will be stored in a long-term preservation repository managed
jointly by the NYU Libraries and the central Information Technology
Services unit. This repository ensures that each content item has a METS
[14] structural metadata file that associates the digital asset with its
metadata. It stores files in two mirrored and geographically distributed
locations, and a third copy on offsite tape; it performs regular fixity
checks; and it provides a format migration capacity, in the event that a
stored format becomes at risk of obsolescence.

\section{Policies Framework}

Sharing video recordings poses a unique challenge to existing data
sharing policies because videos contain personally identifying
information—specifically participants’ faces and voices and often the
insides of their homes and classrooms. Sharing personally identifiable
information puts research participants and others depicted in recordings
at increased risk for loss of privacy. At the same time, blurring or
altering original recordings to hide identities undermines or eliminates
their value to other researchers. Often, participants’ faces and voices
produce the behaviors of interest. So, Databrary has elected to maximize
the potential for data re-use by keeping recordings in their original
unaltered form. Instead of removing participants’ identities, Databrary
restricts access to identifiable or sensitive data to authorized
researchers. Further, Databrary provides access only when the people
depicted have given permission for their information to be shared with
other researchers.

\subsection{Restricting Access}

Databrary provides access to shared data only to authorized researchers
who have agreed to uphold common practices concerning the responsible
and ethical use of identifiable and sensitive data. To become an
authorized investigator, applicants must register on the site and
electronically sign the Authorized Investigator Agreement, which must
also be co-signed by the applicant’s institution. Full privileges will
be granted only to researchers with principal investigator (PI) status
at their institutions. Other researchers may be granted privileges if
they are affiliated with a PI who agrees to sponsor and supervise their
application. Initially there will be a manual process to identify the
institutional representative—typically the authorizing official of the
university—who can co-sign the Investigator Agreement. However, as the
user groups at each university expand, Databrary may implement
administrative accounts at each institution. This will enable the
authorizing official to independently manage the authorizations of
individual researchers at her institution.

\subsection{Seeking permission to share identifiable data}

Data from a particular session may be stored in Databrary for the
contributing researcher’s use whether the records are shared with other
scientists or not. When a researcher chooses to share, Databrary makes
records openly available to the community of authorized researchers only
if the people depicted in the recordings have given permission to
release the data for sharing. Thus, Databrary requires that people
depicted in recordings grant permission before their information can be
shared. Databrary’s policies extend currently accepted principles of
informed consent to the situation where participants are granted
authority to consent to (or refuse) the release of their identifiable
data.

We developed these ideas in close collaboration with the NYU and PSU IRB
staff. To formalize the process of acquiring permission, we developed a
Participant Release Form Template, based on photo or video release
language many researchers use currently. The template release form has
standard language that Databrary recommends investigators should use
with study participants. This language makes it easy for participants to
understand what is involved in sharing their video data, with whom it
will be shared, and the potential risks associated with releasing their
video and other identifiable data to other researchers. Use of the
template also allows for the standardization of language associated with
the release of identifiable or sensitive participant data.

\subsection{Technical assistance with IRBs}

Some IRBs may deem an investigator’s existing, approved video or photo
release form equivalent to the Databrary release. This enables a
researcher to share with Databrary recordings they have already
collected. However, most researchers will need to modify their research
protocols, by adding the Databrary sharing permission procedures, prior
to collecting new shareable video data. Databrary staff are available to
advise potential data contributors about how to amend existing research
protocols so that the information acquired is Databrary-compliant.
Protocol amendments involve seeking approval for use of the Databrary
template release form and modifying the time period over which collected
data will be made available. Specifically, researchers must remove any
clauses in research consent documents that require data destruction
after some fixed period of time since Databrary intends to store shared
data indefinitely.

\section{Impact}

The Databrary project has already begun to have an impact. As of
mid-January 2015, more than 57 scientists from 35 institutions in North
America, the UK, and Europe have received authorization for full access
to and data sharing with Databrary. More than 50 additional researchers
are in the process of securing institutional approval. In consultation
with Databrary staff, many of these researchers have secured or are in
the process of securing permission from their research ethics boards to
share archival or new data. Thus, the Databrary project has laid the
groundwork for change in scientific culture around data sharing within
the community of developmental science researchers who are most familiar
with it. In a similar vein, Databrary is attempting to lead by example
where open science practices are concerned. Databrary is an open source
project. The entire code base is available on Github
(github.com/databrary) as are all policy documents. The project team
consults regularly with other leaders in the open science, data sharing,
and data repository communities, and we share best practices among them.

Databrary has also begun to have an impact in the policy arena. The
notion that identifiable research data may be shared – under the right
circumstances – is not new to Databrary [15,16]. But, Databrary has
created a set of policies and template documents that will help IRBs to
come to see that seeking permission to share data is merely an extension
of the principle of informed consent. Furthermore, Databrary’s
Authorized Investigator Agreement combines provisions for data
contribution with those guiding data use, part of an overall effort to
reduce barriers to sharing and re-use. Our colleagues in the data
sharing community tell us that this combination represents an innovation
in itself.

The Databrary team has come to understand that laboratory data
management practices pose significant challenges to widespread data
sharing. Simply put, many researchers deploy workflows that would
require significant modification in order for video files and associated
metadata to be readily shared. In response, Databrary has expanded its
data curation expertise and capacity. However, hand-curating significant
volumes of research data in this way is not sustainable in the long
term. Building tools that can enable self-curation will be essential to
the future success and sustainability of Databrary. As a result, the
team has begun to compile best practices for data management and
curation that will be folded into future data coding (Datavyu) and data
library (Databrary) features. We note that the process of understanding
the diversity of data management techniques in developmental science
poses unique opportunities for librarians and curators to learn how
researchers organize and manage their data for daily use. It also
affords researchers opportunities to learn from their library science
colleagues about the value of adopting best practices in data
organization and management.

\section{Conclusions}

The Databrary project aims to increase scientific transparency and
accelerate discovery in developmental science by creating a
user-friendly and powerful infrastructure for researchers to share video
and related data. Clearly, sharing video data poses technical and policy
challenges, but it presents significant opportunities for accelerating
discovery if these challenges can be met successfully. The Databrary
project has already made significant strides in identifying and
overcoming many of the obstacles, and the tools and infrastructure that
we develop promise to enhance data sharing and management practices in
the entire behavioral science community.

\section{Acknowledgments}

This work was supported by the National Science Foundation (BCS-1238599)
and the National Institute of Child Health and Human Development
(U01-HD-076595-01). The authors gratefully acknowledge the NYU
Libraries, Human Connectome Project, the Personal Genome Project, the
Inter-university Consortium for Political and Social Research, the
Center for Open Science, Dataverse, Data Dryad, and staff from the
National Database for Autism Research for their valuable advice and
consultation.

REFERENCES
==========

Curry A (2011) Rescue of Old Data Offers Lesson for Particle Physicists.
Science 331:694-695. doi:10.1126/science.331.6018.694

Integrated Earth Data Applications. (2014). http://www.iedadata.org.
Accessed 10 February 2014

Kaye J, Heeney C, Hawkins N, de Vries J, Boddington P (2009) Data
sharing in genomics [mdash] re-shaping scientific practice. Nat Rev
Genet 10 (5):331-335. doi:10.1038/nrg2573

Overpeck JT, Meehl GA, Bony S, Easterling DR (2011) Climate Data
Challenges in the 21st Century. Science 331 (6018):700-702.
doi:10.1126/science.1197869

Lunshof J, Church G, Prainsack B (2014) Raw Personal Data: Providing
Access. Science 343 (6169):373-374. doi:10.1126/science.1249382

Adolph K, Gilmore RO, Freeman C, Sanderson P, Millman D (2012) Toward
Open Behavioral Science. Psychological Inquiry: An International Journal
for the Advancement of Psychological Theory 23 (3):244-247.
doi:10.1080/1047840X.2012.705133

King G (2011) Ensuring the Data-Rich Future of the Social Sciences.
Science 331 (6018):719-721. doi:10.1126/science.1197872.

Association for Psychological Science (2012) Special Section on
Replicability in Psychological Science: A Crisis of Confidence.
Perspectives on Psychological Science 7 (6)

Nosek BA, Bar-Anan Y (2012) Scientific Utopia: I. Opening Scientific
Communication. Psychological Inquiry 23 (3):217-243.
doi:10.1080/1047840X.2012.692215

Nosek BA, Spies JR, Motyl M (2012) Scientific Utopia: II. Restructuring
Incentives and Practices to Promote Truth Over Publishability.
Perspectives on Psychological Science 7 (6):615-631.
doi:10.1177/1745691612459058

MacWhinney B (2001) From CHILDES to TalkBank. In: M. Almgren AB, M.
Ezeizaberrena, I. Idiazabal & B. MacWhinney (ed) Research on Child
Language Acquisition. Cascadilla, Massachusetts, pp 17-34

Cognitive Atlas. (2014). http://www.cognitiveatlas.org. Accessed 10
February 2014

Databrary (2013) Databrary Data Sharing Manifesto.
http://databrary.org/user-guide/policies/data-sharing-manifesto.html.
Accessed 10 February 2014

METS (2014) XSLT-based Metadata Encoding and Transcription Standards
Viewers. http://dlib.nyu.edu/metstools/metsviewer/. Accessed 10 February
2014

HCP (2014) Human Connectome Project.
http://www.humanconnectomeproject.org. Accessed 10 February 2014

PGP (2014) Personal Genome Project. http://www.personalgenomes.org.
Accessed 10 February 2014

\end{document}

### 2015-01-14 jcdl-15/paper.tex
Cleanup some cruft and fix syntax errors
Use references and add basic makefile

@@
-% This is "sig-alternate.tex" V2.0 May 2012
-% This file should be compiled with V2.5 of "sig-alternate.cls" May 2012
-%
-% This example file demonstrates the use of the 'sig-alternate.cls'
-% V2.5 LaTeX2e document class file. It is for those submitting
-% articles to ACM Conference Proceedings WHO DO NOT WISH TO
-% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
-% The 'sig-alternate.cls' file will produce a similar-looking,
-% albeit, 'tighter' paper resulting in, invariably, fewer pages.
-%
-% ----------------------------------------------------------------------------------------------------------------
-% This .tex file (and associated .cls V2.5) produces:
-%       1) The Permission Statement
-%       2) The Conference (location) Info information
-%       3) The Copyright Line with ACM data
-%       4) NO page numbers
-%
-% as against the acm_proc_article-sp.cls file which
-% DOES NOT produce 1) thru' 3) above.
-%
-% Using 'sig-alternate.cls' you have control, however, from within
-% the source .tex file, over both the CopyrightYear
-% (defaulted to 200X) and the ACM Copyright Data
-% (defaulted to X-XXXXX-XX-X/XX/XX).
-% e.g.
-% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
-% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
-%
-% ---------------------------------------------------------------------------------------------------------------
-% This .tex source is an example which *does* use
-% the .bib file (from which the .bbl file % is produced).
-% REMEMBER HOWEVER: After having produced the .bbl file,
-% and prior to final submission, you *NEED* to 'insert'
-% your .bbl file into your source .tex file so as to provide
-% ONE 'self-contained' source file.
-%
-% ================= IF YOU HAVE QUESTIONS =======================
-% Questions regarding the SIGS styles, SIGS policies and
-% procedures, Conferences etc. should be sent to
-% Adrienne Griscti (griscti@acm.org)
-%
-% Technical questions _only_ to
-% Gerald Murray (murray@hq.acm.org)
-% ===============================================================
-%
-% For tracking purposes - this is V2.0 - May 2012
-
 \documentclass{sig-alternate}
@@
 \begin{document}
-%
-% --- Author Metadata here ---
 \conferenceinfo{JCDL}{'15 Knoxville, Tennesse USA}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
-% --- End of Author Metadata ---
 
@@
 
-\numberofauthors{6}
+\numberofauthors{3}
 
@@
 \alignauthor
-Rick O. Gilmore, Ph.D.\\
-	\affaddress{The Pennsylvania State University}\\
-	\affaddress{Department of Psychology}\\
-	\affadddress{University Park, PA 16802}\\
-	\email{rogilmore@psu.edu}
-
-\alignauthor
-Lisa A. Steiger\\
-	\affaddress{Databrary Project}\\
-	\affaddress{New York University}
-	\affaddress{196 Mercer Street, Room 807}\\
-	\affaddress{New York, NY 10012}\\
-	\email{lisa@databrary.org}
-
-\alignauthor
 Dylan A. Simon\\
-	\affaddress{Databrary Project}\\
-	\affaddress{New York University}
-	\affaddress{196 Mercer Street, Room 807}\\
-	\affaddress{New York, NY 10012}\\
+	\affaddr{Databrary Project}\\
+	\affaddr{New York University}\\
+	\affaddr{196 Mercer Street, Room 807}\\
+	\affaddr{New York, NY 10012}\\
 	\email{dylan@databrary.org}
-
-\and
-
 \alignauthor
 Drew Gordon\\
-	\affaddress{Databrary Project}\\
-	\affaddress{New York University}
-	\affaddress{196 Mercer Street, Room 807}\\
-	\affaddress{New York, NY 10012}\\
+	\affaddr{Databrary Project}\\
+	\affaddr{New York University}
+	\affaddr{196 Mercer Street, Room 807}\\
+	\affaddr{New York, NY 10012}\\
 	\email{drew@databrary.org}
-
 \alignauthor
-David S. Millman\\
-	\affaddress{New York University}
-	\affaddress{New York, NY 10012}\\
-	\email{dsm@nyu.edu}
-
-\alignauthor
-Karen E. Adolph, Ph.D.\\
-	\affaddress{New York University}
-	\affaddress{Department of Psychology}
-	\affaddress{New York, NY 10012}\\
-	\email{karen.adolph@nyu.edu}
-
+Rick O. Gilmore, Ph.D.\\
+	\affaddr{The Pennsylvania State University}\\
+	\affaddr{Department of Psychology}\\
+	\affaddr{University Park, PA 16802}\\
+	\email{rogilmore@psu.edu}
 }
@@
 \end{abstract}
-
-**Categories and Subject Descriptors**
 
@@
 
-REFERENCES
-==========
-
-Curry A (2011) Rescue of Old Data Offers Lesson for Particle Physicists.
-Science 331:694-695. doi:10.1126/science.331.6018.694
-
-Integrated Earth Data Applications. (2014). http://www.iedadata.org.
-Accessed 10 February 2014
-
-Kaye J, Heeney C, Hawkins N, de Vries J, Boddington P (2009) Data
-sharing in genomics [mdash] re-shaping scientific practice. Nat Rev
-Genet 10 (5):331-335. doi:10.1038/nrg2573
-
-Overpeck JT, Meehl GA, Bony S, Easterling DR (2011) Climate Data
-Challenges in the 21st Century. Science 331 (6018):700-702.
-doi:10.1126/science.1197869
-
-Lunshof J, Church G, Prainsack B (2014) Raw Personal Data: Providing
-Access. Science 343 (6169):373-374. doi:10.1126/science.1249382
-
-Adolph K, Gilmore RO, Freeman C, Sanderson P, Millman D (2012) Toward
-Open Behavioral Science. Psychological Inquiry: An International Journal
-for the Advancement of Psychological Theory 23 (3):244-247.
-doi:10.1080/1047840X.2012.705133
-
-King G (2011) Ensuring the Data-Rich Future of the Social Sciences.
-Science 331 (6018):719-721. doi:10.1126/science.1197872.
-
-Association for Psychological Science (2012) Special Section on
-Replicability in Psychological Science: A Crisis of Confidence.
-Perspectives on Psychological Science 7 (6)
-
-Nosek BA, Bar-Anan Y (2012) Scientific Utopia: I. Opening Scientific
-Communication. Psychological Inquiry 23 (3):217-243.
-doi:10.1080/1047840X.2012.692215
-
-Nosek BA, Spies JR, Motyl M (2012) Scientific Utopia: II. Restructuring
-Incentives and Practices to Promote Truth Over Publishability.
-Perspectives on Psychological Science 7 (6):615-631.
-doi:10.1177/1745691612459058
-
-MacWhinney B (2001) From CHILDES to TalkBank. In: M. Almgren AB, M.
-Ezeizaberrena, I. Idiazabal & B. MacWhinney (ed) Research on Child
-Language Acquisition. Cascadilla, Massachusetts, pp 17-34
-
-Cognitive Atlas. (2014). http://www.cognitiveatlas.org. Accessed 10
-February 2014
-
-Databrary (2013) Databrary Data Sharing Manifesto.
-http://databrary.org/user-guide/policies/data-sharing-manifesto.html.
-Accessed 10 February 2014
-
-METS (2014) XSLT-based Metadata Encoding and Transcription Standards
-Viewers. http://dlib.nyu.edu/metstools/metsviewer/. Accessed 10 February
-2014
-
-HCP (2014) Human Connectome Project.
-http://www.humanconnectomeproject.org. Accessed 10 February 2014
-
-PGP (2014) Personal Genome Project. http://www.personalgenomes.org.
-Accessed 10 February 2014
+\bibliographystyle{abbrv}
+\bibliography{references}
 

### 2015-01-14 jcdl-15/paper.tex
Fix inline citations

@@
 in the biological, physical and earth sciences with demonstrated
-benefits for scientific transparency and accelerated discovery [1-4].
+benefits for scientific transparency and accelerated discovery \cite{Curry2011,iedadata,Kaye2009,Overpeck2011}.
 Open research practices are gaining traction in the behavioral and
-social sciences [5-10]. However, open data sharing among developmental
-researchers is still the exception [11] not common practice. In seeking
+social sciences \cite{Lunshof2014,Adolph2012,King2011,APS2012,Nosek2012i,Nosek2012ii}. However, open data sharing among developmental
+researchers is still the exception \cite{MacWhinney2001} not common practice. In seeking
 to expand sharing of video-based data, Databrary faces challenges
@@
 strict metadata ontologies and that doing so may have benefits in
-sub-domains of research when there is community consensus [12]. We will
+sub-domains of research when there is community consensus \cite{cogatlas}. We will
 support standard data coding ontologies among researchers, but only
@@
 comfortable sharing, or when journals or funders require it. Databrary
-has published a Data Sharing Manifesto [13] that explains to researchers
+has published a Data Sharing Manifesto \cite{manifesto} that explains to researchers
 the Databrary philosophy. Standards about when data should be shared are
@@
 Services unit. This repository ensures that each content item has a METS
-[14] structural metadata file that associates the digital asset with its
+\cite{METS} structural metadata file that associates the digital asset with its
 metadata. It stores files in two mirrored and geographically distributed
@@
 notion that identifiable research data may be shared – under the right
-circumstances – is not new to Databrary [15,16]. But, Databrary has
+circumstances – is not new to Databrary \cite{HCP,PGP}. But, Databrary has
 created a set of policies and template documents that will help IRBs to

### 2015-01-15 jcdl-15/paper.tex
basic outline

@@
 
-Permission to make digital or hard copies of all or part of this work
-for personal or classroom use is granted without fee provided that
-copies are not made or distributed for profit or commercial advantage
-and that copies bear this notice and the full citation on the first
-page. To copy otherwise, or republish, to post on servers or to
-redistribute to lists, requires prior specific permission and/or a fee.
+\subsection{Video-based human research data}
+
+\subsection{Discovering and re-purposing data}
+
+\section{The ``Session''}
+
+\section{Organizing and documenting collections}
+
+\section{Collecting and annotating temporal data}
+
+\section{Open Problems}
+
+\section{Introduction}
 

### 2015-01-15 jcdl-15/paper.tex
Some intro sentences

@@
 \section{Introduction}
+
+Video plays a key role in many areas of human research, particularly in developmental psychology, education, ...
+Unfortunately, in standard practice, due largely to the perceived burdens of ethics and privacy restrictions, the use of these data is isolated to the researchers who collected them.
+The goal of the Databrary project establish policies and tools to allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interest.
+Accordingly, Databrary, with developmental psychology as a use case, has created a web-based library and organizational structures to accommodate existing video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
 

### 2015-01-15 jcdl-15/paper.tex
more sentences, questionable

@@
 Accordingly, Databrary, with developmental psychology as a use case, has created a web-based library and organizational structures to accommodate existing video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+
+% this is questionable...
+Most existing research data repositories provide either unstructured storage of arbitrary files, or have rigorously structured schema that require contributors to properly extract and format contributions.
+While the former is well suited for archival of entire datasets and general-purpose hosting of materials, it does not facilitate discoverability, understanding, or re-use across various datasets.
+The latter, on the other hand, generate large-scale, homogonous data sources for meta- and re-analysis, but limited to specific domains and research questions, losing the full richness and innovation of the source materials and field.
+% we're a hybrid?
 

### 2015-01-16 jcdl-15/paper.tex
checkpoint: first pass on intro, sessions
decided to reorder session/record

@@
 The latter, on the other hand, generate large-scale, homogonous data sources for meta- and re-analysis, but limited to specific domains and research questions, losing the full richness and innovation of the source materials and field.
+We have attempted here to adopt a middle ground: the bar for contributing a collection, but when data are provided we encourage a good deal of common structure, either by capturing data as they are collected, or by curation after the fact.
+It is this structure that is the focus of this paper.
 % we're a hybrid?
 
-\subsection{Video-based human research data}
+We have intentionally chosen to limit the bulk of the data hosted in the library to human-consumable, self-documenting, timeseries data: video and audio.
+These data can and largely are consumed and analyzed, at least initially, without sophisticated analysis tools: once digitally decoded, humans can glean the bulk of the richness of such data directly.
+Similarly, these data can often be used on their own or with minimal explanation, as the nature of the activity and participants are readily apparent.
+This naturally aids in the problems of documentation and re-use that plague most flat-file or dense datasets.
+On the other hand, since automated analysis of videos is an active and immature research area, the problems of search, discoverability, and meta-analysis remain.
+We attempt to address these issues by augmenting the primary data with additional, structured annotations and metadata.
 
-\subsection{Discovering and re-purposing data}
+\section{Data Schema}
 
-\section{The ``Session''}
+As with most user-driven databases, we separate individual data contributions or packages into individual collections variously referred to as datasets, studies, or volumes.
+Volumes can be associated with a minimal amount of metadata (title, description, permissions, etc.) as well as additional files (documents and other materials).
+These volumes are to some extent embeddable and referenceable, both among themselves and to and from external entities by DOI.
+The remainder of this work will address the available structures of data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
 
-\section{Organizing and documenting collections}
+% should we actually use the term "record"?  Alternatives: "grouping", "label", "annotation", ...
 
-\section{Collecting and annotating temporal data}
+There are two key, innovative structural elements that comprise a volume: the session and the record.
+Sessions are the primary container for data files, basically an elaboration on traditional file folders with additional temporal metadata.
+Records are used to represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as providing an organizational structure among sessions.
+The combination of these two elements allows data and metadata to be flexibly related and annotated.
 
-\section{Open Problems}
+\subsection{The ``Session''}
+
+Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
+Thus, the video itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
+However, in many cases a single session can comprise multiple videos: from serial data streams due to recording equipment being reset or producing multiple files, and/or in parallel from multiple data capture sources that overlap in time.
+These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, all of which have specific and critical temporal relationships with the primary data.
+
+We capture these relationships within an object called a session, which can comprises not only arbitrary files but also their absolute or relative temporal positioning.
+The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or a date of collection and relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units).
+Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
+Although some advanced recording equipment provides automatic timestamping or synchronization than can be used, most researchers collecting multiple parallel data streams often use ``sync points'' such as an electronic trigger signal, flash, or tone to align recordings.
+
+Once the timeframe of a session is established, this can allow temporally-positioned annotations to be attached to the session.
+% this is sort of an aside:
+In fact, this practice is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio and annotate events, utterances, actions, or phases at any temporal scale, from individual samples/frames to the entire session.
+Most of these tools currently focus on a single recording file, however, and despite the plethora of tools and features, there is little interoperability.
+
+\subsection{The ``Record''}
 

### 2015-01-16 jcdl-15/paper.tex
checkpoint: records and a bit more

@@
 
-We capture these relationships within an object called a session, which can comprises not only arbitrary files but also their absolute or relative temporal positioning.
+% should we be using more language like, "we observed", "in practice"?
+When dealing with multiple primary recordings, researchers often reduce these raw files into a single representation.
+For example, they often concatenate serial recordings, or make a composite video by joining or overlaying frames.
+Secondary data is also often overlaid or combined in this way, such as by adding an additional waveform or frequency visualization below a video, or by overlaying crosshairs representing location or gaze data.
+Doing so allows for more convenient human analysis of the data in standard tools that only support a single data stream, but also often loses some of the raw data, by obscuring, masking, or most often downsampling spatially or temporally.
+
+Instead, we capture these relationships within an object called a session, which can comprise not only arbitrary files but also their absolute or relative temporal positioning.
 The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or a date of collection and relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units).
@@
 
+We currently allow keyword-based tagging and comment-based discussions on any temporal segment of a session.
+These annotations could be applied manually by users or by automated video analysis systems.
+We expect that these annotations will provide an index target for additional discovery capabilities for users looking for particular features within data.
+
 \subsection{The ``Record''}
+
+An object called a record serves a number of purposes around session and metadata organization.
+A record itself represents any number of variables of measurements---what would traditionally be a row of a database or flat file.
+For example, a record may represent an individual participant within the dataset, and contain their subject identifier, gender, birthdate, and any other demographic, survey, or measurement data about that participant.
+A record could also represent a particular experimental task, condition, location, or outcome, and contain variables with corresponding parameters or descriptions.
+Records are thus labeled and divided by the type of entity they represent.
+These variables themselves a simply textual, numeric, date, or potentially any other scalar data as in any traditional tabular data interface, with the records of a given type constituting the rows.
+
+However, records can also be used to organize sessions.
+We observed that different researchers organize their data files in different ways according to the type of research.
+Many consider participants to be the primary item of analysis, and so create a directory for each participant.
+Others organize their data based on date, experimental condition, phase of analysis, or age groups.
+Sometimes even single video files are split up into segments according to phase or activity, and researchers consider this the primary property organizational principal.
+
+Rather than trying to handle all of these strategies separately, we instead leverage these records.
+That is, records can be attached to sessions, or, equivalently, sessions can be grouped into records, in a many-to-many relationship.
+For example, in a traditional psychology study, each participant may appear in exactly one session, so the collected information about that participant, added to the record, is attached one-to-one with their session.
+A dataset may be composed of a few different sub-sets based on experimental procedures or location, and here each session would be placed into exactly one of these few records.
+In some research, multiple participants may also appear in multiple sessions, such as recordings of the same class of students over a span of time.
+
+Since researchers can create whichever record types apply to their data, they can then choose to group their data by whichever record type they wish.
+This also allows users to either import or export data in what they consider the native directory structure by indicating the appropriate nesting of record types.
+By providing some standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants of a particular age range.
+
+Finally, records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
+That is, if any record only applies to part of a session, for example a single activity, or a participant being present for only part of a recording, it can be applied within the timeline of a session to only the appropriate part.
+This allows the final type of organizational structure, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
+It also can be used to represent annotations at a finer temporal scale, for example individual events interactions that are normally handled by off-line annotation softaware.
+However, the appropriate interfaces and tools to input or import these annotations are still being developed.
+
+\section{Standardization vs. Flexibility}
+
+As the extant research data libraries demonstrate, there is always the tension between standardization and flexibility.
+Since we're primarily dealing with video and audio data, we have, without much loss of flexibility, standardized on a particular encoding format (codecs and parameters) that allows for convenient access and HTML5-based streaming, while still allowing arbitrary resolutions and sample/frame-rates.
+We also allow for ``dumps'' using certain portable data formats to include any data that cannot be encompased by our standard formats, reasoning that these data may be relevant for re-use, but not broadly adopted by the community.
+
+Here we have also introduced a number of metadata concepts where this tension becomes most apparent, namely the available record types, variables (and values within categorical data types), and tag keywords.
+We have adopted an approach of encouraging adherence to existing standardized options but also allowing new types to be created as the data require it.
+We encourage standardization both on-line as data are collected and entered via our user interface using auto-completion, suggestions, and drop-down options, and off-line via manual review and curaion of contributed data.
+
+Achieving the appropriate balance will no doubt be challanging, but by limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
 

### 2015-01-16 jcdl-15/paper.tex
a few tiny edits
didn't get very far past first paragraph

@@
 
-Video plays a key role in many areas of human research, particularly in developmental psychology, education, ...
-Unfortunately, in standard practice, due largely to the perceived burdens of ethics and privacy restrictions, the use of these data is isolated to the researchers who collected them.
-The goal of the Databrary project establish policies and tools to allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interest.
+Video plays a key role in many areas of human research, particularly in developmental psychology, education, and is even a primary data source ...
+Unfortunately, in standard practice, due largely to the perceived requirements about privacy restrictions, the use of these video data is isolated to the researchers who collected them.
+The goal of the Databrary project is to establish policies and tools to allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interest.
 Accordingly, Databrary, with developmental psychology as a use case, has created a web-based library and organizational structures to accommodate existing video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.

### 2015-01-16 jcdl-15/paper.tex
sending up some notes and edits
probably have more, but overall I think now is a good time, since we have some structure and details, to reflect on the overall gist of what we are trying to communicate, that is to keep site of the larger picture while discussing the nitty gritty.

@@
 \begin{abstract}
-
+%%dg being that the conference them is big data, should we be trying to be more explicit about that?
+% big data is generally referring to "more rows" rather than larger files. I think the closest we get is the size of video files
+% which is already noted below. Just wondering if it makes sense to be more explicit or not.
 Video captures the complexity, richness, and diversity of behavior
@@
 participant privacy, data tagging, and data management.
+%%dg also preservation and discoverability - though maybe that's included in data management
 \end{abstract}
@@
 Video plays a key role in many areas of human research, particularly in developmental psychology, education, and is even a primary data source ...
-Unfortunately, in standard practice, due largely to the perceived requirements about privacy restrictions, the use of these video data is isolated to the researchers who collected them.
+%%dg unclear what is meant by primary data source in this context
+Unfortunately, in standard practice, the use of these video data is isolated to the researchers who collected them because of the perceived requirements about privacy restrictions.
 The goal of the Databrary project is to establish policies and tools to allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interest.
-Accordingly, Databrary, with developmental psychology as a use case, has created a web-based library and organizational structures to accommodate existing video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+Accordingly, Databrary, with a focus on developmental psychology research, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
 
@@
 Most existing research data repositories provide either unstructured storage of arbitrary files, or have rigorously structured schema that require contributors to properly extract and format contributions.
+%%dg should pick examples to illustrate both sides of the spectrum (ICPSR - ???)
 While the former is well suited for archival of entire datasets and general-purpose hosting of materials, it does not facilitate discoverability, understanding, or re-use across various datasets.
+%%dg i'd say that the general purpose ones are fine on discoverability, but they might no facilitate re-use or reflect very much of the original 
+%context of the data
 The latter, on the other hand, generate large-scale, homogonous data sources for meta- and re-analysis, but limited to specific domains and research questions, losing the full richness and innovation of the source materials and field.
-We have attempted here to adopt a middle ground: the bar for contributing a collection, but when data are provided we encourage a good deal of common structure, either by capturing data as they are collected, or by curation after the fact.
-It is this structure that is the focus of this paper.
+Databrary attempts to strike a balance by creating a common structure that is general enough to provide the flexibility for incorporating heterogenous datasets while at the same time facilitating a depth of access to configure data more to the original context in which it was collected and analyzed.
+While the architecture of the backend allows us to curate and include diverse datasets, another important component is a well-define user experience that allows researchers to define how their data is kept in Databrary by uploading it as they collect it.
+The following provides further detail on the design and decision-making of this structure.
 % we're a hybrid?
+%%dg more like a system informed by the pitfalls of committing solely to either extreme.
 
+%%dg perhaps better to frame below as the affordances and challenges of working with video data. 
+%That is..."A key component to understanding the design decisions made in creating Databrary is the unique affordances and challenges of video and audio data"
+%Affordances
 We have intentionally chosen to limit the bulk of the data hosted in the library to human-consumable, self-documenting, timeseries data: video and audio.
@@
 This naturally aids in the problems of documentation and re-use that plague most flat-file or dense datasets.
-On the other hand, since automated analysis of videos is an active and immature research area, the problems of search, discoverability, and meta-analysis remain.
-We attempt to address these issues by augmenting the primary data with additional, structured annotations and metadata.
+%Challenges
+On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain.
+We attempt to address these issues by augmenting the primary video data with additional, structured annotations and metadata.
+%%dg probably need to make this section more cohesive, but the main gist is 1) We focus on video data, 2)video data has these inherent challenges.
+3) At the same time there are already challenges involved in creating a research data repository, independant of the format of the data travels on.
+4) This paper dicusses the challenges we have resolved and the challenges we have coming up.
 
 \section{Data Schema}
+
+%%dg would a figure illustration be helpful here? do we already have one?
 
@@
 
-Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
+(Behvioral|Developmental) Science Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
 Thus, the video itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
@@
 % this is sort of an aside:
+%%dg maybe the value here is that we've designed things with this specific type of research in mind.
 In fact, this practice is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio and annotate events, utterances, actions, or phases at any temporal scale, from individual samples/frames to the entire session.
@@
 
-We currently allow keyword-based tagging and comment-based discussions on any temporal segment of a session.
+In addition to providing an organizational structure that allows users to visualize data in ways familiar to their research, we also currently allow keyword-based tagging and comment-based discussions on any temporal segment of a session.
 These annotations could be applied manually by users or by automated video analysis systems.
 We expect that these annotations will provide an index target for additional discovery capabilities for users looking for particular features within data.
+
+%%dg, i guess it's crucial to make sure the audience understands that one of the specific challenges for us is that we're not just storing comprehensive single video files, but multiple files that capture the same phenomenon and how do we keep all those together in a meaningful way...to differentiate us from being "just another you tube" ..
+which i guess is what is already being said above.
+Though might need to do a little more hand holding for the reader as to why this is an issue in general and why it's important to even those who might be faced with different types of video data to collect (?
+maybe I'm rambling now ?).
 

### 2015-01-18 jcdl-15/paper.tex
more notes
More updates, respond to some comments, culling
No new sections

@@
 % which is already noted below. Just wondering if it makes sense to be more explicit or not.
-Video captures the complexity, richness, and diversity of behavior
-unlike any other tool. As a result, large numbers of researchers who
-study human or animal behavior employ video. Video documents itself to a
-large degree, and therefore has significant potential for data re-use.
-Still, researchers rarely share video. Video often contains information
-about personal identities, so considerations about research ethics pose
-challenges to sharing. The relatively large size of video files and
-diversity of formats pose technical challenges. In this paper, we will
-describe how the web-based Databrary data library has overcome the most
-significant barriers to sharing video within the developmental and
-learning sciences community, including solutions to maintaining
-participant privacy, data tagging, and data management.
 %%dg also preservation and discoverability - though maybe that's included in data management
@@
 
-Video plays a key role in many areas of human research, particularly in developmental psychology, education, and is even a primary data source ...
-%%dg unclear what is meant by primary data source in this context
-Unfortunately, in standard practice, the use of these video data is isolated to the researchers who collected them because of the perceived requirements about privacy restrictions.
+%%dg maybe break this part up into subsections: Content/Domain; Users; Design contraints; Approach (or something like that)
+
+Video and audio recordings play a key role in many areas of human research, particularly in psychology, linguistics, education, and ethnography.
+It can document experimental procedures or behaviors, convey ideas and examples in teaching and scientific communication, and often act primary data source for later analysis.
+Unfortunately, in standard practice, the use of these video data is isolated to the researchers who collected them, due largely to the perceived requirements around privacy restrictions.
 The goal of the Databrary project is to establish policies and tools to allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interest.
-Accordingly, Databrary, with a focus on developmental psychology research, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+Accordingly, Databrary, with an initial focus on developmental psychology research, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
 
@@
 %%dg should pick examples to illustrate both sides of the spectrum (ICPSR - ???)
-While the former is well suited for archival of entire datasets and general-purpose hosting of materials, it does not facilitate discoverability, understanding, or re-use across various datasets.
-%%dg i'd say that the general purpose ones are fine on discoverability, but they might no facilitate re-use or reflect very much of the original 
-%context of the data
-The latter, on the other hand, generate large-scale, homogonous data sources for meta- and re-analysis, but limited to specific domains and research questions, losing the full richness and innovation of the source materials and field.
-Databrary attempts to strike a balance by creating a common structure that is general enough to provide the flexibility for incorporating heterogenous datasets while at the same time facilitating a depth of access to configure data more to the original context in which it was collected and analyzed.
-While the architecture of the backend allows us to curate and include diverse datasets, another important component is a well-define user experience that allows researchers to define how their data is kept in Databrary by uploading it as they collect it.
-The following provides further detail on the design and decision-making of this structure.
-% we're a hybrid?
-%%dg more like a system informed by the pitfalls of committing solely to either extreme.
+While the former is well suited for archival of entire datasets and general-purpose hosting of resources, it does not facilitate detailed search, understanding, or re-use across various datasets.
+The latter, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
+We have attempted to strike a balance between these extremes by focusing on video and audio recordings as the unifying factor and allowing additional resources to be built and discovered around this core.
 
@@
 %Affordances
-We have intentionally chosen to limit the bulk of the data hosted in the library to human-consumable, self-documenting, timeseries data: video and audio.
+We have intentionally chosen to limit the bulk of the data hosted in the library to human-consumable, self-documenting, timeseries data: video and audio recordings.
 These data can and largely are consumed and analyzed, at least initially, without sophisticated analysis tools: once digitally decoded, humans can glean the bulk of the richness of such data directly.
-Similarly, these data can often be used on their own or with minimal explanation, as the nature of the activity and participants are readily apparent.
+Similarly, these data can often be used on their own or with minimal explanation, as the nature of the subjects and activity are readily apparent.
 This naturally aids in the problems of documentation and re-use that plague most flat-file or dense datasets.
@@
 On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain.
-We attempt to address these issues by augmenting the primary video data with additional, structured annotations and metadata.
+We attempt to address these issues by augmenting the primary recording data with additional structured and unstructured features.
 %%dg probably need to make this section more cohesive, but the main gist is 1) We focus on video data, 2)video data has these inherent challenges.
@@
 
-\section{Data Schema}
+Databrary can incorporate heterogeneous datasets within our target domain by allowing secondary, general-purpose flat-file data, materials, and references to be associated with the primary data.
+This facilitates a depth of access to configure data more to the original context in which it was collected and analyzed.
+% that use of "configure" doesn't make sense to me -dylan
+At the same time, it provides a common, standard structure for recordings and their associated metadata, including context, contents, and annotations.
+We achieve this conformance of datasets in the library through a combination of in-depth investigative curation and ingest processes for already-collected data, and a well-defined user experience that provides a platform for researchers to contribute and organize all of their research data as they are collected.
+The following sections provide specific detail on the design and decision-making behind this unifying structure.
+%%dg more like a system informed by the pitfalls of committing solely to either extreme.
+
+\section{Data Architecture}
 
@@
 Sessions are the primary container for data files, basically an elaboration on traditional file folders with additional temporal metadata.
-Records are used to represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as providing an organizational structure among sessions.
+Records are used to represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as to provide an organizational structure among sessions.
 The combination of these two elements allows data and metadata to be flexibly related and annotated.
@@
 
-(Behvioral|Developmental) Science Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
-Thus, the video itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
-However, in many cases a single session can comprise multiple videos: from serial data streams due to recording equipment being reset or producing multiple files, and/or in parallel from multiple data capture sources that overlap in time.
-These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, all of which have specific and critical temporal relationships with the primary data.
+Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
+Thus, the recording itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
+However, in many cases a single session can comprise multiple videos: from serial data streams due to recording equipment being producing multiple files, and/or in parallel from multiple data capture sources operating at the same time.
+These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
 
@@
 When dealing with multiple primary recordings, researchers often reduce these raw files into a single representation.
-For example, they often concatenate serial recordings, or make a composite video by joining or overlaying frames.
+For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
 Secondary data is also often overlaid or combined in this way, such as by adding an additional waveform or frequency visualization below a video, or by overlaying crosshairs representing location or gaze data.
-Doing so allows for more convenient human analysis of the data in standard tools that only support a single data stream, but also often loses some of the raw data, by obscuring, masking, or most often downsampling spatially or temporally.
+Doing so allows for more convenient human analysis of the data in standard tools that only support a single data stream, but also loses some of the raw data, by obscuring, masking, or downsampling spatially or temporally.
 
 Instead, we capture these relationships within an object called a session, which can comprise not only arbitrary files but also their absolute or relative temporal positioning.
-The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or a date of collection and relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units).
+The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
-Although some advanced recording equipment provides automatic timestamping or synchronization than can be used, most researchers collecting multiple parallel data streams often use ``sync points'' such as an electronic trigger signal, flash, or tone to align recordings.
+Although some advanced recording equipment provides automatic timestamping or synchronization than can be used, most researchers collecting multiple parallel data streams use ``sync points'' such as an electronic trigger signal, flash, or tone to align recordings.
+
+\subsubsection{Simple annotations}
 
@@
 %%dg maybe the value here is that we've designed things with this specific type of research in mind.
-In fact, this practice is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio and annotate events, utterances, actions, or phases at any temporal scale, from individual samples/frames to the entire session.
+In fact, this practice is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio and annotate events, utterances, actions, or scenes at any temporal scale, from individual samples/frames to the entire session.
 Most of these tools currently focus on a single recording file, however, and despite the plethora of tools and features, there is little interoperability.
@@
 An object called a record serves a number of purposes around session and metadata organization.
-A record itself represents any number of variables of measurements---what would traditionally be a row of a database or flat file.
-For example, a record may represent an individual participant within the dataset, and contain their subject identifier, gender, birthdate, and any other demographic, survey, or measurement data about that participant.
+A record itself represents any number of variables or measurements---what would traditionally be a row of a database or flat file.
+For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birthdate, and any other demographic, survey, or measurement data about that participant.
 A record could also represent a particular experimental task, condition, location, or outcome, and contain variables with corresponding parameters or descriptions.
 Records are thus labeled and divided by the type of entity they represent.
-These variables themselves a simply textual, numeric, date, or potentially any other scalar data as in any traditional tabular data interface, with the records of a given type constituting the rows.
+The variables themselves are textual, numeric, date, or potentially any other scalar data as in any traditional tabular data interface, with the records of a given type constituting the tabular rows.
 
-However, records can also be used to organize sessions.
+Records can also be used to organize sessions.
 We observed that different researchers organize their data files in different ways according to the type of research.
@@
 Others organize their data based on date, experimental condition, phase of analysis, or age groups.
-Sometimes even single video files are split up into segments according to phase or activity, and researchers consider this the primary property organizational principal.
+Sometimes even single video files are split up into segments according to scene or activity, and researchers consider this property the primary organizational principal.
 
-Rather than trying to handle all of these strategies separately, we instead leverage these records.
+Rather than trying to handle all of these strategies separately, we instead leverage records.
 That is, records can be attached to sessions, or, equivalently, sessions can be grouped into records, in a many-to-many relationship.
 For example, in a traditional psychology study, each participant may appear in exactly one session, so the collected information about that participant, added to the record, is attached one-to-one with their session.
-A dataset may be composed of a few different sub-sets based on experimental procedures or location, and here each session would be placed into exactly one of these few records.
+A dataset may also be composed of a few different sub-sets based on experimental procedures or location, and here each session would be placed into exactly one of these few records.
 In some research, multiple participants may also appear in multiple sessions, such as recordings of the same class of students over a span of time.
@@
 
+%\subsection{Technical infrastructure}
+
+%%dg other notes: should we give a bit more discussion of the infrastruction - Postgresql, ffmpeg, using HPC for transcoding, storage, frontend technologies etc. 
+% i'm not sure. do other papers have this sort of "methods" section? -dylan
+
 \section{Standardization vs. Flexibility}
@@
 Since we're primarily dealing with video and audio data, we have, without much loss of flexibility, standardized on a particular encoding format (codecs and parameters) that allows for convenient access and HTML5-based streaming, while still allowing arbitrary resolutions and sample/frame-rates.
-We also allow for ``dumps'' using certain portable data formats to include any data that cannot be encompased by our standard formats, reasoning that these data may be relevant for re-use, but not broadly adopted by the community.
+We also allow for ``dumps'' using certain portable data formats to include any data that cannot be encompassed by our standard formats, reasoning that these data may be relevant for re-use, but not broadly adopted by the community.
 
@@
 
-Achieving the appropriate balance will no doubt be challanging, but by limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
+Achieving the appropriate balance will no doubt be challenging, but by limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
 
-\section{Introduction}
+%%dg not sure if this drives us off course, but wondering how much we should discuss library specific challenges in the "stuff to come section", both pertaining to video data and just in general, largely stuff that we will still have to deal with, such as: 1) creation and organization of metadata is still largely manual and time consuming - both in combing videos for stuff to clip out and also trying to extract more metadata from the contributed data itself as well as third party sources (e.g.
+keywords from journals).
+2) interpretation of data lends to this, we can't always rely on contributors to be available to tell us what we need to know about their data 3) metadata interoperability: DOIs; Export in friendly formats (Dublin Core, METS), OAI-PMH compliance.
 
-Film and video recording have been a mainstay of research in the
-developmental sciences for nearly 100 years [CITE]. A typical lab
-acquires, on average, 12 hours of video per week depicting behavior in
-natural or laboratory environments [CITE]. These recordings constitute a
-primary source of raw data. Prior to descriptive or statistical
-analyses, researchers score or “code” recordings, mostly manually, using
-commercial software or makeshift tools. However, even the most detailed
-and laborious video coding does not exhaust the possibilities for
-analyses by other researchers with different questions. Thus, sharing
-video promises a wide range use cases, perhaps more than sharing
-flat-file data or static images. Despite the wealth of untapped
-information available in research videos and the labor expended in
-collecting and coding them, developmental researchers rarely share data.
-In the prevailing culture, investigators conduct research privately
-within their own laboratories. Data are destroyed or molder away on a
-shelf after the research is published. The Databrary project aims to
-change the status quo, making sharing standard practice.
-
-Databrary enables researchers to share video and associated flat-file
-data for research and educational purposes. For example, a researcher
-interested in language development might reuse a set of video recordings
-created originally to study the development of walking. Researchers can
-conduct integrative analyses by combining data across disparate studies,
-increase sample sizes, verify the feasibility of proposed projects,
-compare methods, learn about procedures, and independently verify
-results. Similarly, investigators can reuse shared videos for
-educational purposes such as to find illustrative examples for teaching,
-use exemplars in presentations, show students how to conduct procedures,
-or provide supplemental materials for reviewers.
-
-Data sharing and open science practices are emerging or accepted norms
-in the biological, physical and earth sciences with demonstrated
-benefits for scientific transparency and accelerated discovery \cite{Curry2011,iedadata,Kaye2009,Overpeck2011}.
-Open research practices are gaining traction in the behavioral and
-social sciences \cite{Lunshof2014,Adolph2012,King2011,APS2012,Nosek2012i,Nosek2012ii}. However, open data sharing among developmental
-researchers is still the exception \cite{MacWhinney2001} not common practice. In seeking
-to expand sharing of video-based data, Databrary faces challenges
-concerning the interplay among institutional research policies, ethical
-issues related to obtaining participants’ consent for sharing, and
-technical authorization systems. This paper describes the current state
-of the Databrary project including the project organization, how content
-is acquired, stored, preserved, and accessed, policy considerations, and
-how the library has affected and hopes to affect the behavioral science
-research community.
-
-\section{Project Organization}
-
-Databrary is a joint project of New York University (NYU) and The
-Pennsylvania State University (Penn State). The Databrary project began
-with a workshop on open video data sharing funded by the National
-Science Foundation (NSF) under Grant No. BCS-1139702 at which
-developmental scientists, computer scientists, library scientists, and
-federal agency program officers discussed the promises and challenges of
-sharing video data. NSF and the National Institute of Child Health and
-Human Development (NICHD) have provided funding under grant No.
-BCS-1238599 and Cooperative Agreement U01-HD-076595, respectively. The
-interdisciplinary team bring expertise in developmental and neural
-science, information technology, library science, software development,
-data curation, and community engagement. In addition, a board of expert
-advisors composed of developmental scientists, library scientists, and
-leaders of data repositories in the behavioral and social sciences
-provide guidance.
-
-The project is organized around five core activities.
-
-\subsection{Curation}
-
-To seed the Databrary, staff identified and selected video datasets for
-manual curation. The datasets were selected based on availability,
-presumed interest to developmental science community, suitability for
-the organization and file structure of the repository, diversity of
-study type, and on the availability of appropriate sharing permissions
-(see 2.2 below). Manual curation of archival datasets continues under
-the direction of a staff member with expertise in library and
-information sciences.
-
-Nevertheless, we have found manual data curation time consuming and
-difficult, especially for researchers. The more difficult and
-time-consuming data sharing becomes, the less likely it will be that
-researchers participate despite best intentions. So, we have created
-user-friendly spreadsheet-like interfaces that enable researchers to
-self-curate datasets as they are collected. At the end of a data
-collection session, researchers may upload videos and other data files
-collected along with metadata about participant demographic
-characteristics or experimental conditions. Databrary transcodes videos
-to a standard format in the background, and the system alerts the user
-when the video can be previewed. Researchers may also upload study-level
-materials -- code books, flat data files used for statistical analyses,
-images of experimental equipment, links to related external information
-(e.g., GitHub repositories), etc.
-
-This scheme, which we call "upload-as-you-go", makes the data curation
-required for sharing part of the normal scientific workflow, not a
-burdensome after thought. No data are shared until the researcher grants
-access, but unlike the *ad hoc* organization in a typical investigator's
-laboratory, on Databrary, the data are organized in a standardized way
-that could be readily and easily shared with others when the
-investigator is ready to do so.
-
-Sharing video requires special considerations because the data cannot be
-de-identified without reducing its value. No previously established
-research data repository has specialized in video data sharing.
-Therefore, Databrary had to develop a new policy and permission
-framework for sharing identifiable and sensitive data while
-simultaneously preserving the rights of research participants.
-
-The framework has two components, both of which build on
-well-established principles. One extends the idea of informed consent to
-participate in research by requiring that participants give explicit
-permission for their identifiable or sensitive data to be shared with
-other researchers. Databrary makes available on its website template
-language researchers may use for seeking permission to share, along with
-video demonstrations and additional guidance. So, researchers who wish
-to share video must either use the template or demonstrate that
-participants have granted an equivalent level of permission. The second
-component involves limiting access to individuals who have prior formal
-authorization. Researchers must either secure authorization to access
-Databrary from their institution, typically a university where they
-conduct independent research, or must have another authorized researcher
-grant them access. Institution-level authorization requires a formal
-agreement with Databrary. Fundamentally, the Databrary framework ensures
-that participants control whether other researchers can view their video
-(or other identifiable data), and only researchers who pledge to uphold
-ethical principles, including maintaining participant confidentiality,
-may access Databrary's sensitive materials.
-
-Databrary developed these policies and agreements in close consultation
-with Institutional Review Boards (IRBs), sponsored research privacy
-administrators, and legal staff at NYU and Penn State, with other
-experts in research data ethics, and with Federal officials. See
-http://databrary.org/access/guide.html for more details.
-
-The centerpiece of the Databrary project is the repository itself. Our
-team of programmers has developed digital library software, including
-asset management, workflow, access control, discovery services, and a
-user interface. This work is closely coordinated with the central NYU IT
-infrastructure providers for services such as high-capacity storage,
-high-security storage, and high-performance computing (for video
-processing). The long-term goal is to build the cyber-infrastructure for
-seamless search, streaming, uploading, and downloading and to provide an
-open-source framework for other entities to build on the Databrary
-software or extend it to other content domains.
-
-\subsection{Annotation and Metadata}
-
-Video becomes most useful for conducting research when it is transformed
-into quantitative and qualitative data in a flat-file format. Thus, the
-Databrary team is enhancing a general-purpose video-coding software
-application specialized for exploring, annotating, tagging, scoring, and
-visualizing the video files in preparation for qualitative and
-quantitative analyses. Toward that end, we have released Datavyu
-(datayvu.org), a free, open source, multi-platform desktop application.
-Datavyu is a fork of the OpenSHAPA tool used by some developmental
-researchers. The long-term goal is to build a web-based version of
-Datavyu that will provide both data analysis and data management
-services required for preparing and organizing video material prior to
-its storage in Databrary.
-
-\subsection{Community Practices}
-
-The current zeitgeist in the developmental science community is one of
-privacy and data hoarding. A primary activity of the Databrary project
-is to introduce the practice of open data sharing to the developmental
-science community. We are doing so through announcements to
-developmental research societies, colloquia, and conference
-presentations. The team participates in workshops to discuss sharing
-issues and hosts events to teach potential users about sharing and
-analysis tools. We post links on websites frequented by developmental
-researchers and participate in user forums. In addition, we have asked
-leaders in developmental science to set an example for other researchers
-by publicly committing to open video data sharing. More than a hundred
-have responded to the call
-(http://databrary.org/community/contributors.html). Our long-term goal
-is a self-sustaining community of researchers committed to open
-science.
-
-\section{Building the Library}
-
-Existing data management practices pose challenges for video sharing.
-Despite similar research methods, study designs vary widely. No two
-developmental science labs manage data in the same way. Some studies are
-longitudinal, where researchers observe the same participants at
-multiple sessions. Some are cross-sectional, where researchers observe
-each participant at only one session. The timing of observations may be
-determined by participants’ age (e.g., 4-month-olds, 8-month-olds,
-12-month-olds) or abilities (e.g., preverbal, one-word utterances,
-sentences), experimentally determined variables (e.g., pre- and
-post-intervention), or other factors. Some labs organize longitudinal
-data by grouping files first by participant and then by session date.
-Other labs organize longitudinal data by session and then by
-participant. Still others organize longitudinal data based on task (book
-reading, block building, free play). Some researchers institute a
-central data management system to be followed by all the lab members,
-providing easier access and greater transparency for the entire lab but
-not necessarily providing the structure for similar benefits to
-researchers unfamiliar with the lab’s practices. Other researchers allow
-their students to keep separate records, making it difficult to share
-data even within the lab. Some researchers keep videos, metadata, and
-analyses together, and some do not. Idiosyncratic terms, record-keeping,
-and data management systems are the norm. Databrary must enable
-researchers to discover and understand each other’s materials,
-regardless of the original investigator’s data management system.
-
-Databrary takes several approaches to address this challenge: A flexible
-data model, informal metadata contributions by users via tagging and
-commenting, mediated curation, and incentives to deposit materials more
-consistently. Figure 1 illustrates the principal data model. It
-accommodates different hierarchies and organizations, such as those
-described in the examples above.
-
-The full expressive flexibility of the data model is not evident to
-Databrary users. Instead, users see different aspects of the data model
-at different times, depending on the context. Researchers organize their
-materials by acquisition date and time into structures called sessions.
-A session corresponds to a unique recording episode and contains one or
-more recordings or related flat files (assets). Researchers can group
-sessions in different ways, using whatever groupings are appropriate for
-the particular dataset. These groupings may also identify particular
-time segments of a recording to distinguish tasks, events, or
-participants that comprise the session. In the data model, the aggregate
-of these flexible groupings is called a volume. In practice, researchers
-combine sessions or segments of sessions within and across datasets to
-form the raw material that are subsequently described in published
-articles and presentations. Thus, Databrary contributors can combine
-sessions or segments within and across datasets with ancillary
-materials, such as coding manuals, Datavyu spreadsheets, statistical
-analyses, questionnaires, IRB documents, computer code, sample displays,
-and links to published journal articles in an aggregate structure
-investigators call a study. Like datasets, studies are represented as
-volumes in the data model.
-
-The discovery and browsing interfaces offer content items at the volume
-(study or dataset) level to users searching the library. The
-contributor-assigned groupings (records), which may include relevant
-metadata about participants (measures, such as participant demographic
-information, domain-specific survey information, location data,
-condition variables, task descriptions, or other properties), allow a
-second level of filtering and organization within and between relevant
-identified studies or datasets. This enables users to quickly identify
-data that may be relevant to their own interests or research.
-
-Authorized Databrary Investigators can also add keyword tags at the
-study/dataset, session, or segment level, and can endorse or deprecate
-keywords that have been suggested by other authorized researchers. These
-abilities are critical because different researchers may have very
-different reasons for citing a study or using data. Future
-implementations may extend the ability of authorized researchers to
-annotate studies with other kinds of metadata.
-
-The Databrary team recognizes that other data repositories enforce
-strict metadata ontologies and that doing so may have benefits in
-sub-domains of research when there is community consensus \cite{cogatlas}. We will
-support standard data coding ontologies among researchers, but only
-enforce standardization for a small set of standard tags such as study
-date, participant birth date, and sex. We will also encourage
-contributors to report race, ethnicity, primary language, language of
-dataset, and location of session. The Databrary team chose not to
-require strict metadata ontologies for several reasons. We hope to
-reduce the pre-deposit curational demands on contributors, encourage the
-repurposing of shared data, and foster the rapid adoption of data
-sharing. Developing and achieving agreement on metadata ontologies can
-take significant amounts of time. Video data are so rich and complex
-that in many domains, researchers have not settled on standard
-definitions for particular behaviors and may have little current need
-for standard tasks, procedures, or terminology. However, we will also
-encourage users to re-use tags and terminology by suggesting common or
-similar terms, without confining users to these suggestions. User
-communities within Databrary may eventually converge on common
-conceptual and metadata ontologies based on the most common (and
-commonly endorsed) keyword tags, but standardized ontologies are not
-necessary for browsing and searching in most of the use cases we
-envision.
-
-Our experiences curating and ingesting archival datasets have
-highlighted the considerable value of contributors entering raw video
-data into Databrary as soon as recordings are acquired. Immediate
-uploading reduces the workload on investigators, minimizes the risk of
-data loss and corruption, and accelerates the speed with which materials
-become openly available. To encourage immediate uploading, Databrary
-provides a complete set of controls so that researchers can restrict
-access to their data to only their own labs or to other users of their
-choosing. Datasets can be shared at a later point when data collection
-and ancillary materials are complete, whenever the contributor is
-comfortable sharing, or when journals or funders require it. Databrary
-has published a Data Sharing Manifesto \cite{manifesto} that explains to researchers
-the Databrary philosophy. Standards about when data should be shared are
-evolving. Our philosophy is consistent with concepts and practices in
-other domains where data sharing is the norm (e.g., www.iedata.org).
-Planned enhancements to the Datavyu tool will allow contributors to
-organize video files and metadata as part of the analysis process. This
-will facilitate the contribution of packaged datasets to Databrary.
-
-As part of the curation process, Databrary stores at least two versions
-of each item of Databrary video content: a copy for access, and the
-received original file if it was digital, or a 10-bit YUV digital
-preservation copy if the original version was not digital. Currently,
-the access version format is H.264 (HiP) with AAC audio in an MPEG-4
-container, although we expect the appropriate video formats to change
-over time, as has been the case with many digital video formats in
-recent years.
-
-For preservation, the original file (if digital) or the preservation
-copy will be stored in a long-term preservation repository managed
-jointly by the NYU Libraries and the central Information Technology
-Services unit. This repository ensures that each content item has a METS
-\cite{METS} structural metadata file that associates the digital asset with its
-metadata. It stores files in two mirrored and geographically distributed
-locations, and a third copy on offsite tape; it performs regular fixity
-checks; and it provides a format migration capacity, in the event that a
-stored format becomes at risk of obsolescence.
-
-\section{Policies Framework}
-
-Sharing video recordings poses a unique challenge to existing data
-sharing policies because videos contain personally identifying
-information—specifically participants’ faces and voices and often the
-insides of their homes and classrooms. Sharing personally identifiable
-information puts research participants and others depicted in recordings
-at increased risk for loss of privacy. At the same time, blurring or
-altering original recordings to hide identities undermines or eliminates
-their value to other researchers. Often, participants’ faces and voices
-produce the behaviors of interest. So, Databrary has elected to maximize
-the potential for data re-use by keeping recordings in their original
-unaltered form. Instead of removing participants’ identities, Databrary
-restricts access to identifiable or sensitive data to authorized
-researchers. Further, Databrary provides access only when the people
-depicted have given permission for their information to be shared with
-other researchers.
-
-\subsection{Restricting Access}
-
-Databrary provides access to shared data only to authorized researchers
-who have agreed to uphold common practices concerning the responsible
-and ethical use of identifiable and sensitive data. To become an
-authorized investigator, applicants must register on the site and
-electronically sign the Authorized Investigator Agreement, which must
-also be co-signed by the applicant’s institution. Full privileges will
-be granted only to researchers with principal investigator (PI) status
-at their institutions. Other researchers may be granted privileges if
-they are affiliated with a PI who agrees to sponsor and supervise their
-application. Initially there will be a manual process to identify the
-institutional representative—typically the authorizing official of the
-university—who can co-sign the Investigator Agreement. However, as the
-user groups at each university expand, Databrary may implement
-administrative accounts at each institution. This will enable the
-authorizing official to independently manage the authorizations of
-individual researchers at her institution.
-
-\subsection{Seeking permission to share identifiable data}
-
-Data from a particular session may be stored in Databrary for the
-contributing researcher’s use whether the records are shared with other
-scientists or not. When a researcher chooses to share, Databrary makes
-records openly available to the community of authorized researchers only
-if the people depicted in the recordings have given permission to
-release the data for sharing. Thus, Databrary requires that people
-depicted in recordings grant permission before their information can be
-shared. Databrary’s policies extend currently accepted principles of
-informed consent to the situation where participants are granted
-authority to consent to (or refuse) the release of their identifiable
-data.
-
-We developed these ideas in close collaboration with the NYU and PSU IRB
-staff. To formalize the process of acquiring permission, we developed a
-Participant Release Form Template, based on photo or video release
-language many researchers use currently. The template release form has
-standard language that Databrary recommends investigators should use
-with study participants. This language makes it easy for participants to
-understand what is involved in sharing their video data, with whom it
-will be shared, and the potential risks associated with releasing their
-video and other identifiable data to other researchers. Use of the
-template also allows for the standardization of language associated with
-the release of identifiable or sensitive participant data.
-
-\subsection{Technical assistance with IRBs}
-
-Some IRBs may deem an investigator’s existing, approved video or photo
-release form equivalent to the Databrary release. This enables a
-researcher to share with Databrary recordings they have already
-collected. However, most researchers will need to modify their research
-protocols, by adding the Databrary sharing permission procedures, prior
-to collecting new shareable video data. Databrary staff are available to
-advise potential data contributors about how to amend existing research
-protocols so that the information acquired is Databrary-compliant.
-Protocol amendments involve seeking approval for use of the Databrary
-template release form and modifying the time period over which collected
-data will be made available. Specifically, researchers must remove any
-clauses in research consent documents that require data destruction
-after some fixed period of time since Databrary intends to store shared
-data indefinitely.
-
-\section{Impact}
-
-The Databrary project has already begun to have an impact. As of
-mid-January 2015, more than 57 scientists from 35 institutions in North
-America, the UK, and Europe have received authorization for full access
-to and data sharing with Databrary. More than 50 additional researchers
-are in the process of securing institutional approval. In consultation
-with Databrary staff, many of these researchers have secured or are in
-the process of securing permission from their research ethics boards to
-share archival or new data. Thus, the Databrary project has laid the
-groundwork for change in scientific culture around data sharing within
-the community of developmental science researchers who are most familiar
-with it. In a similar vein, Databrary is attempting to lead by example
-where open science practices are concerned. Databrary is an open source
-project. The entire code base is available on Github
-(github.com/databrary) as are all policy documents. The project team
-consults regularly with other leaders in the open science, data sharing,
-and data repository communities, and we share best practices among them.
-
-Databrary has also begun to have an impact in the policy arena. The
-notion that identifiable research data may be shared – under the right
-circumstances – is not new to Databrary \cite{HCP,PGP}. But, Databrary has
-created a set of policies and template documents that will help IRBs to
-come to see that seeking permission to share data is merely an extension
-of the principle of informed consent. Furthermore, Databrary’s
-Authorized Investigator Agreement combines provisions for data
-contribution with those guiding data use, part of an overall effort to
-reduce barriers to sharing and re-use. Our colleagues in the data
-sharing community tell us that this combination represents an innovation
-in itself.
-
-The Databrary team has come to understand that laboratory data
-management practices pose significant challenges to widespread data
-sharing. Simply put, many researchers deploy workflows that would
-require significant modification in order for video files and associated
-metadata to be readily shared. In response, Databrary has expanded its
-data curation expertise and capacity. However, hand-curating significant
-volumes of research data in this way is not sustainable in the long
-term. Building tools that can enable self-curation will be essential to
-the future success and sustainability of Databrary. As a result, the
-team has begun to compile best practices for data management and
-curation that will be folded into future data coding (Datavyu) and data
-library (Databrary) features. We note that the process of understanding
-the diversity of data management techniques in developmental science
-poses unique opportunities for librarians and curators to learn how
-researchers organize and manage their data for daily use. It also
-affords researchers opportunities to learn from their library science
-colleagues about the value of adopting best practices in data
-organization and management.
-
-\section{Conclusions}
-
-The Databrary project aims to increase scientific transparency and
-accelerate discovery in developmental science by creating a
-user-friendly and powerful infrastructure for researchers to share video
-and related data. Clearly, sharing video data poses technical and policy
-challenges, but it presents significant opportunities for accelerating
-discovery if these challenges can be met successfully. The Databrary
-project has already made significant strides in identifying and
-overcoming many of the obstacles, and the tools and infrastructure that
-we develop promise to enhance data sharing and management practices in
-the entire behavioral science community.
+\section{Conclusion}
 

### 2015-01-18 jcdl-15/paper.tex
some more notes, clean out stuff that has been addressed, some edits to paper and references

@@
 Databrary can incorporate heterogeneous datasets within our target domain by allowing secondary, general-purpose flat-file data, materials, and references to be associated with the primary data.
-This facilitates a depth of access to configure data more to the original context in which it was collected and analyzed.
-% that use of "configure" doesn't make sense to me -dylan
+This facilitates storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
 At the same time, it provides a common, standard structure for recordings and their associated metadata, including context, contents, and annotations.
@@
 %%dg would a figure illustration be helpful here? do we already have one?
+%%dg in general, how much room do we have for screenshots, if they help explain?
 
@@
 % should we actually use the term "record"?  Alternatives: "grouping", "label", "annotation", ...
+% as long as we're clear in defining it. the record section appears to be doing this.
 
@@
 
+%dg "The session was designed both to organize videos in a way that is familiar to researchers, as well as to allow them to 'go into' a session and add or edit data." - stuff like this to remind the reader why we're discussing these components in specific (rather than some other aspect of the system).
 Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
@@
 % should we be using more language like, "we observed", "in practice"?
-When dealing with multiple primary recordings, researchers often reduce these raw files into a single representation.
+%%dg - if it helps to illustrate particular points better, sure. 
+When working with multiple primary recordings, researchers often reduce these raw files into a single representation.
 For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
@@
 
-%%dg, i guess it's crucial to make sure the audience understands that one of the specific challenges for us is that we're not just storing comprehensive single video files, but multiple files that capture the same phenomenon and how do we keep all those together in a meaningful way...to differentiate us from being "just another you tube" ..
-which i guess is what is already being said above.
-Though might need to do a little more hand holding for the reader as to why this is an issue in general and why it's important to even those who might be faced with different types of video data to collect (?
-maybe I'm rambling now ?).
+%%dg, i guess it's crucial to make the audience aware that one of the specific challenges for us is that we're not just storing comprehensive single video files, but multiple files that capture the same phenomenon and how do we keep all those together in a meaningful way...to differentiate us from being "just another you tube".
 
@@
 
+%%dg maybe add some langauge to the beginning of this section that is explicit that the "record" is designed with the flexibility we talk about above in mind.
 An object called a record serves a number of purposes around session and metadata organization. 
@@
 % i'm not sure. do other papers have this sort of "methods" section? -dylan
+%%dg - other papers in this general area (see: Peer_2012, Griffin_etal_2014, and Witt_2012 from references) do comment on the technology used (if there actually is a product and not just a proposed framework) to varying degrees.
+I don't think we have to spend more than a paragraph on it, but it would be good to mention the stack in terms of taking, augmenting, and storing video files and metadata overall.
 
@@
 Since we're primarily dealing with video and audio data, we have, without much loss of flexibility, standardized on a particular encoding format (codecs and parameters) that allows for convenient access and HTML5-based streaming, while still allowing arbitrary resolutions and sample/frame-rates.
-We also allow for ``dumps'' using certain portable data formats to include any data that cannot be encompassed by our standard formats, reasoning that these data may be relevant for re-use, but not broadly adopted by the community.
+We also allow for ``dumps'', which allow users - using certain portable data formats (e.g.
+...) - to include any data that cannot be encompassed by our standard formats.
+In this way, we can incorporate data which may be relevant for re-use, but not broadly adopted by the community, without making the system too rigid.
 
@@
 We have adopted an approach of encouraging adherence to existing standardized options but also allowing new types to be created as the data require it.
-We encourage standardization both on-line as data are collected and entered via our user interface using auto-completion, suggestions, and drop-down options, and off-line via manual review and curaion of contributed data.
+We encourage standardization both on-line as data are collected and entered via our user interface using auto-completion, suggestions, and drop-down options, and off-line via manual review and curation of contributed data.
 

### 2015-01-18 jcdl-15/paper.tex
Add technical infrastructure paragraph
Needs citations, though that goes without saying since everything does

@@
 This allows the final type of organizational structure, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
-It also can be used to represent annotations at a finer temporal scale, for example individual events interactions that are normally handled by off-line annotation softaware.
+It also can be used to represent annotations at a finer temporal scale, for example individual events interactions that are normally handled by off-line annotation software.
 However, the appropriate interfaces and tools to input or import these annotations are still being developed.
 
-%\subsection{Technical infrastructure}
+\subsection{Technical infrastructure}
 
@@
 
+Databrary is a new, cohesive web application, built on the Play Framework to enable a responsive user interface, a complete API, and high-performance streaming.
+Recordings and other files are placed in content-based filesystem storage, and all structured data are stored in a PostgreSQL database, leveraging its geometric indexing capabilities for temporal data.
+All uploaded recordings are automatically transcoded to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
+This transcoding utilizes the high performance computing cluster at the hosting campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
+The user interface is built primarily on the Angular web framework, and all data access is performed through an open JSON API.
+Although hosted data are protected, all source code is released under a GPLv3 license on github.
+
 \section{Standardization vs. Flexibility}
@@
 As the extant research data libraries demonstrate, there is always the tension between standardization and flexibility.
-Since we're primarily dealing with video and audio data, we have, without much loss of flexibility, standardized on a particular encoding format (codecs and parameters) that allows for convenient access and HTML5-based streaming, while still allowing arbitrary resolutions and sample/frame-rates.
-We also allow for ``dumps'', which allow users - using certain portable data formats (e.g.
-...) - to include any data that cannot be encompassed by our standard formats.
+Since we're primarily dealing with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
+We also allow for ``dumps'', which allow users---using certain portable data formats (e.g.
+...)---to include any data that cannot be encompassed by our standard formats.
 In this way, we can incorporate data which may be relevant for re-use, but not broadly adopted by the community, without making the system too rigid.
@@
 
-\section{Conclusion}
+\section{Conclusions}
 
-\section{Acknowledgments}
+\section*{Acknowledgments}
 

### 2015-01-19 jcdl-15/paper.tex
add reference, edits

@@
 Video and audio recordings play a key role in many areas of human research, particularly in psychology, linguistics, education, and ethnography.
-It can document experimental procedures or behaviors, convey ideas and examples in teaching and scientific communication, and often act primary data source for later analysis.
-Unfortunately, in standard practice, the use of these video data is isolated to the researchers who collected them, due largely to the perceived requirements around privacy restrictions.
+It can document experimental procedures or behaviors, convey ideas and examples in teaching and scientific communication, and is a rich primary data source for later analysis.
+Unfortunately, in standard practice, the use of these video data is isolated to the researchers who collected them, due largely to the perceived requirements around privacy restrictions, not to mention.
 The goal of the Databrary project is to establish policies and tools to allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interest.
@@
 % this is questionable...
-Most existing research data repositories provide either unstructured storage of arbitrary files, or have rigorously structured schema that require contributors to properly extract and format contributions.
-%%dg should pick examples to illustrate both sides of the spectrum (ICPSR - ???)
-While the former is well suited for archival of entire datasets and general-purpose hosting of resources, it does not facilitate detailed search, understanding, or re-use across various datasets.
-The latter, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
+%%dg probably want to try a different approach to highlighting this dichotomy
+We have found in designing an appropriate data schema that there are two competing polarities that need to be balanced out.
+On the one hand, a system might tend towards more generality, providing unstructured storage of arbitrary files.
+On the other, it can present a rigorously structured schema that requires contributors to properly extract and format contributions in a highly specified manner.
+While the former is well suited for archival of entire datasets and general-purpose hosting of resources, it does not facilitate detailed search, understanding, or re-use across various datasets with a great deal of curation \cite{Peer_2012}.
+The latter, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but are limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
 We have attempted to strike a balance between these extremes by focusing on video and audio recordings as the unifying factor and allowing additional resources to be built and discovered around this core.
@@
 %Challenges
-On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain.
+On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
 We attempt to address these issues by augmenting the primary recording data with additional structured and unstructured features.
@@
 The following sections provide specific detail on the design and decision-making behind this unifying structure.
-%%dg more like a system informed by the pitfalls of committing solely to either extreme.
 

### 2015-01-19 jcdl-15/paper.tex
tweaked 1st part of intro

@@
 \alignauthor
-Rick O. Gilmore, Ph.D.\\
+Rick O. Gilmore\\
 	\affaddr{The Pennsylvania State University}\\
@@
 
-Video and audio recordings play a key role in many areas of human research, particularly in psychology, linguistics, education, and ethnography.
-It can document experimental procedures or behaviors, convey ideas and examples in teaching and scientific communication, and is a rich primary data source for later analysis.
-Unfortunately, in standard practice, the use of these video data is isolated to the researchers who collected them, due largely to the perceived requirements around privacy restrictions, not to mention.
-The goal of the Databrary project is to establish policies and tools to allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interest.
-Accordingly, Databrary, with an initial focus on developmental psychology research, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+Video and audio recordings play a key role in many areas of research, especially in psychology, linguistics, education, ethnography, and ethology.
+Video documents experimental procedures or behaviors, conveys ideas and provides examples for teaching and scientific communication, and serves as a rich primary data source for later analysis.
+Unfortunately, in standard practice, the use of video data is restricted to the researchers who collect them, due largely to the perceived requirements around privacy restrictions and the relatively large file sizes that make email an impractical means of distribution.
+The goal of the Databrary project is to establish tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interests.
+The project also aims to create policies that enable the personally idenfifying information contained in videos to be shared among researchers while protecting the rights of research participants.
+Accordingly, Databrary, with an initial focus on research in developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+In this paper, we discuss design and technical challenges the Databrary team has overcome in launching the system, and we suggest ways that our experiences building Databrary may have relevance to other teams charged with building large scale data repositories that are specialized to serve the needs of specific research communities.
+
+\subsection{Video is Special}
+%%dg perhaps better to frame below as the affordances and challenges of working with video data. 
+%That is..."A key component to understanding the design decisions made in creating Databrary is the unique affordances and challenges of video and audio data"
+%Affordances
+We have intentionally chosen to focus our efforts on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
+Video is an inexpensive and easy-to-use medium for capturing the richness, diversity, and complexity of behavior.
+As a consequence, film and video recording have been a mainstay of research in the developmental sciences for nearly 100 years. 
+A typical developmental lab acquires, on average, 12 hours of video per week depicting behavior in natural or laboratory environments.
+Video data can and largely are consumed and analyzed, at least initially, without sophisticated analysis tools: Once digitally decoded, trained human observers can glean directly much of the data's richness.
+Similarly, video data collected in one context can often be used by others for different purposes with minimal explanation, as the nature of the subjects and activity are readily apparent.
+This naturally reduces the problems of documentation and re-use that plague most flat-file or dense datasets.
+%Challenges
+On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
+We attempt to address these issues by augmenting the primary recording data with additional structured and unstructured features.
+%%dg probably need to make this section more cohesive, but the main gist is 1) We focus on video data, 2)video data has these inherent challenges.
+3) At the same time there are already challenges involved in creating a research data repository, independant of the format of the data travels on.
+4) This paper dicusses the challenges we have resolved and the challenges we have coming up.
 
@@
 We have attempted to strike a balance between these extremes by focusing on video and audio recordings as the unifying factor and allowing additional resources to be built and discovered around this core.
-
-%%dg perhaps better to frame below as the affordances and challenges of working with video data. 
-%That is..."A key component to understanding the design decisions made in creating Databrary is the unique affordances and challenges of video and audio data"
-%Affordances
-We have intentionally chosen to limit the bulk of the data hosted in the library to human-consumable, self-documenting, timeseries data: video and audio recordings.
-These data can and largely are consumed and analyzed, at least initially, without sophisticated analysis tools: once digitally decoded, humans can glean the bulk of the richness of such data directly.
-Similarly, these data can often be used on their own or with minimal explanation, as the nature of the subjects and activity are readily apparent.
-This naturally aids in the problems of documentation and re-use that plague most flat-file or dense datasets.
-%Challenges
-On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
-We attempt to address these issues by augmenting the primary recording data with additional structured and unstructured features.
-%%dg probably need to make this section more cohesive, but the main gist is 1) We focus on video data, 2)video data has these inherent challenges.
-3) At the same time there are already challenges involved in creating a research data repository, independant of the format of the data travels on.
-4) This paper dicusses the challenges we have resolved and the challenges we have coming up.
 

### 2015-01-19 jcdl-15/paper.tex
Fix addresses, cedar reference, more on uayg

--- 2015-01-19 jcdl-15/paper.tex
@@
 	\affaddr{New York University}\\
-	\affaddr{196 Mercer Street, Room 807}\\
-	\affaddr{New York, NY 10012}\\
+	\affaddr{New York, NY, USA}\\
 	\email{dylan@databrary.org}
@@
 	\affaddr{Databrary Project}\\
-	\affaddr{New York University}
-	\affaddr{196 Mercer Street, Room 807}\\
-	\affaddr{New York, NY 10012}\\
+	\affaddr{New York University}\\
+	\affaddr{New York, NY, USA}\\
 	\email{drew@databrary.org}
@@
 	\affaddr{Department of Psychology}\\
-	\affaddr{University Park, PA 16802}\\
+	\affaddr{University Park, PA, USA}\\
 	\email{rogilmore@psu.edu}
@@
 For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
-Secondary data is also often overlaid or combined in this way, such as by adding an additional waveform or frequency visualization below a video, or by overlaying crosshairs representing location or gaze data.
+Secondary data is also often overlaid or combined in this way, such as by adding an additional waveform or frequency visualization below a video, or by overlaying cross-hairs representing location or gaze data.
 Doing so allows for more convenient human analysis of the data in standard tools that only support a single data stream, but also loses some of the raw data, by obscuring, masking, or downsampling spatially or temporally.
@@
 
-\subsubsection{Simple annotations}
+\subsubsection*{Simple annotations}
 
@@
 A record itself represents any number of variables or measurements---what would traditionally be a row of a database or flat file.
-For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birthdate, and any other demographic, survey, or measurement data about that participant.
+For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birth date, and any other demographic, survey, or measurement data about that participant.
 A record could also represent a particular experimental task, condition, location, or outcome, and contain variables with corresponding parameters or descriptions.
@@
 By providing some standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants of a particular age range.
+
+\subsubsection*{Structured annotations}
 
@@
 Since we're primarily dealing with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
-We also allow for ``dumps'', which allow users---using certain portable data formats (e.g.
-...)---to include any data that cannot be encompassed by our standard formats.
-In this way, we can incorporate data which may be relevant for re-use, but not broadly adopted by the community, without making the system too rigid.
+We also allow for ``dumps'', which allow users---using certain portable formats (flat-files, documents, images, data)---to include any data that cannot be encompassed by our standard formats.
+In this way, we can incorporate data which may be relevant for re-use but not broadly adopted by the community, without making the system too rigid.
 
@@
 We have adopted an approach of encouraging adherence to existing standardized options but also allowing new types to be created as the data require it.
-We encourage standardization both on-line as data are collected and entered via our user interface using auto-completion, suggestions, and drop-down options, and off-line via manual review and curation of contributed data.
+Initially we have achieved this through manual review and off-line curation of contributions, dynamically extending the system as new types of data were discovered.
+For many previously collected datasets, this has proven the most effective method to both learn from data and avoid burdening contributors.
+However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a separate, user-driven approach.
 
-Achieving the appropriate balance will no doubt be challenging, but by limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
+We found that while some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools, often using a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and available filesystem storage (often just external hard-drives).
+This suggested the opportunity to supply a cohesive data management platform where all collected raw data, including recordings, contextual information, and measurements, could be entered, stored, and exported for later off-line analysis.
+Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset through liberal use of auto-completion, suggestions, and drop-down options.
+With the considerable benefit of reducing storage requirements within researcher's labs, this has proven an attractive prospect for researchers.
+
+Achieving the appropriate balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
 

### 2015-01-19 jcdl-15/paper.tex
Another full pass

@@
 Video and audio recordings play a key role in many areas of research, especially in psychology, linguistics, education, ethnography, and ethology.
-Video documents experimental procedures or behaviors, conveys ideas and provides examples for teaching and scientific communication, and serves as a rich primary data source for later analysis.
-Unfortunately, in standard practice, the use of video data is restricted to the researchers who collect them, due largely to the perceived requirements around privacy restrictions and the relatively large file sizes that make email an impractical means of distribution.
-The goal of the Databrary project is to establish tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose existing videos for their own research interests.
-The project also aims to create policies that enable the personally idenfifying information contained in videos to be shared among researchers while protecting the rights of research participants.
+Video documents experimental procedures or behaviors, conveys ideas, provides examples for teaching and scientific communication, and serves as a rich primary data source for later analysis.
+Unfortunately, in standard practice, the use of video data is restricted to the researchers who collect them, due largely to the perceived requirements around privacy restrictions.
+The goal of the Databrary project is to create tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose them for their own research interests.
 Accordingly, Databrary, with an initial focus on research in developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
-In this paper, we discuss design and technical challenges the Databrary team has overcome in launching the system, and we suggest ways that our experiences building Databrary may have relevance to other teams charged with building large scale data repositories that are specialized to serve the needs of specific research communities.
+In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
 
-\subsection{Video is Special}
+%\subsection{Affordances and Challenges of Video}
+
 %%dg perhaps better to frame below as the affordances and challenges of working with video data. 
@@
 We have intentionally chosen to focus our efforts on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
-Video is an inexpensive and easy-to-use medium for capturing the richness, diversity, and complexity of behavior.
-As a consequence, film and video recording have been a mainstay of research in the developmental sciences for nearly 100 years. 
-A typical developmental lab acquires, on average, 12 hours of video per week depicting behavior in natural or laboratory environments.
-Video data can and largely are consumed and analyzed, at least initially, without sophisticated analysis tools: Once digitally decoded, trained human observers can glean directly much of the data's richness.
-Similarly, video data collected in one context can often be used by others for different purposes with minimal explanation, as the nature of the subjects and activity are readily apparent.
+These types of data have unique affordances but also present additional challenges.
+On one hand, recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can glean directly much of the data's richness.
+In particular, video data collected in one context can often be used by others for different purposes with minimal explanation, as the nature of the subjects and activity are readily apparent.
 This naturally reduces the problems of documentation and re-use that plague most flat-file or dense datasets.
@@
 
-% this is questionable...
-%%dg probably want to try a different approach to highlighting this dichotomy
-We have found in designing an appropriate data schema that there are two competing polarities that need to be balanced out.
+%\section{Normalization vs. Flexibility}
+
+We have found in researching existing research data repositories that there are two competing approaches that need to be balanced out.
 On the one hand, a system might tend towards more generality, providing unstructured storage of arbitrary files.
 On the other, it can present a rigorously structured schema that requires contributors to properly extract and format contributions in a highly specified manner.
-While the former is well suited for archival of entire datasets and general-purpose hosting of resources, it does not facilitate detailed search, understanding, or re-use across various datasets with a great deal of curation \cite{Peer_2012}.
+While the former is well suited for archival of entire datasets and general-purpose hosting of resources, it does not facilitate detailed search, understanding, or re-use across various datasets without a great deal of curation \cite{Peer_2012}.
 The latter, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but are limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
@@
 We achieve this conformance of datasets in the library through a combination of in-depth investigative curation and ingest processes for already-collected data, and a well-defined user experience that provides a platform for researchers to contribute and organize all of their research data as they are collected.
-The following sections provide specific detail on the design and decision-making behind this unifying structure.
+The following section provides specific detail on the design and decision-making behind this unifying structure.
 
@@
 
-As with most user-driven databases, we separate individual data contributions or packages into individual collections variously referred to as datasets, studies, or volumes.
+As with most user-driven databases, we separate individual data contributions into collections or packages variously referred to as datasets, studies, or volumes.
 Volumes can be associated with a minimal amount of metadata (title, description, permissions, etc.) as well as additional files (documents and other materials).
-These volumes are to some extent embeddable and referenceable, both among themselves and to and from external entities by DOI.
-The remainder of this work will address the available structures of data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
+These volumes are to some extent embeddable and referenceable, both among themselves and to and from external entities by DOI or URI.
+The remainder of this work will address the available structures for data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
 
@@
 
-%dg "The session was designed both to organize videos in a way that is familiar to researchers, as well as to allow them to 'go into' a session and add or edit data." - stuff like this to remind the reader why we're discussing these components in specific (rather than some other aspect of the system).
+%dg "" - stuff like this to remind the reader why we're discussing these components in specific (rather than some other aspect of the system).
 Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
 Thus, the recording itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
-However, in many cases a single session can comprise multiple videos: from serial data streams due to recording equipment being producing multiple files, and/or in parallel from multiple data capture sources operating at the same time.
+However, in many cases a single session can comprise multiple videos: from serial data streams due to recording equipment producing multiple files, and/or in parallel from multiple data capture sources operating at the same time.
 These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
 
-% should we be using more language like, "we observed", "in practice"?
-%%dg - if it helps to illustrate particular points better, sure. 
 When working with multiple primary recordings, researchers often reduce these raw files into a single representation.
@@
 Secondary data is also often overlaid or combined in this way, such as by adding an additional waveform or frequency visualization below a video, or by overlaying cross-hairs representing location or gaze data.
-Doing so allows for more convenient human analysis of the data in standard tools that only support a single data stream, but also loses some of the raw data, by obscuring, masking, or downsampling spatially or temporally.
+Doing so allows for more convenient human analysis of the data in standard playback tools that only support a single data stream, but also loses some of the raw data, by obscuring, masking, or downsampling spatially or temporally.
+The session was thus designed to organize data in a way that is familiar to researchers without requiring these modifications.
 
-Instead, we capture these relationships within an object called a session, which can comprise not only arbitrary files but also their absolute or relative temporal positioning.
+A session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
 The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
@@
 Although some advanced recording equipment provides automatic timestamping or synchronization than can be used, most researchers collecting multiple parallel data streams use ``sync points'' such as an electronic trigger signal, flash, or tone to align recordings.
+These sync points can then be used to manually place recordings on the timeline appropriately.
 
@@
 %%dg maybe the value here is that we've designed things with this specific type of research in mind.
-In fact, this practice is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio and annotate events, utterances, actions, or scenes at any temporal scale, from individual samples/frames to the entire session.
+In fact, this practice is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio to annotate events, utterances, actions, or scenes at any temporal scale, from individual samples/frames to the entire session.
 Most of these tools currently focus on a single recording file, however, and despite the plethora of tools and features, there is little interoperability.
@@
 A record itself represents any number of variables or measurements---what would traditionally be a row of a database or flat file.
-For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birth date, and any other demographic, survey, or measurement data about that participant.
+For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birth date, and any other demographic, survey, or measurement data about that individual.
 A record could also represent a particular experimental task, condition, location, or outcome, and contain variables with corresponding parameters or descriptions.
 Records are thus labeled and divided by the type of entity they represent.
-The variables themselves are textual, numeric, date, or potentially any other scalar data as in any traditional tabular data interface, with the records of a given type constituting the tabular rows.
+The variables themselves are textual, numeric, date, or potentially any other scalar data as in any traditional tabular data interface, with the records of a given type then constituting the tabular rows.
 
-Records can also be used to organize sessions.
+Records are also used to organize sessions.
 We observed that different researchers organize their data files in different ways according to the type of research.
@@
 
-\section{Standardization vs. Flexibility}
+\section{Normalization vs. Flexibility}
 
-As the extant research data libraries demonstrate, there is always the tension between standardization and flexibility.
+As the extant research data libraries demonstrate, there is always the tension between normalization and flexibility.
 Since we're primarily dealing with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
@@
 
-This work was supported by the National Science Foundation (BCS-1238599)
-and the National Institute of Child Health and Human Development
-(U01-HD-076595-01). The authors gratefully acknowledge the NYU
-Libraries, Human Connectome Project, the Personal Genome Project, the
-Inter-university Consortium for Political and Social Research, the
-Center for Open Science, Dataverse, Data Dryad, and staff from the
-National Database for Autism Research for their valuable advice and
-consultation.
+This work was supported by the National Science Foundation (BCS-1238599) and the National Institute of Child Health and Human Development (U01-HD-076595-01).
+The authors gratefully acknowledge the NYU Libraries for their valuable advice and consultation.
 

### 2015-01-19 jcdl-15/paper.tex
Update keywords, minor addition

@@
 
-\category{H.2.8}{Database Management}{Database Applications}[image
-databases, scientific databases]
+\category{H.2.8}{Database Management}{Database Applications}[scientific databases, image databases]
+\category{H.3.5}{Information Storage and Retrieval}{Online Information Systems}[data sharing, web-based services]
 
-\terms{Management, Documentation, Design, Security, Legal Aspects}
+\terms{Design, Standardization}
 
@@
 Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset through liberal use of auto-completion, suggestions, and drop-down options.
-With the considerable benefit of reducing storage requirements within researcher's labs, this has proven an attractive prospect for researchers.
+With the considerable benefit of reducing storage requirements within researcher's labs, and often adding convenience over existing practices, this has proven an attractive prospect for researchers.
 

### 2015-01-20 jcdl-15/paper.tex
meager updates, adding a reference in paper
maybe premature, but starts to pin things down
comments on 2015-01-20 draft
smoothing some transitions, adding some language, clearing out notes that have been addressed

@@
 %%dg also preservation and discoverability - though maybe that's included in data management
+%%rg video is big/dense in time and space.
+But, much of the size is masked without intensive analysis by humans or machines.
+Compare it with brain imaging data.
+A typical brain imaging study might include 3-5 runs of 200 brain volumes with 30 128x128 pixel x 8 bit greyscale slices per volume, with each volume collected about every 2-3 sec (.33-.5 Hz).
+Video is much "bigger" because it samples space and time densely.
+
 \end{abstract}
@@
 %That is..."A key component to understanding the design decisions made in creating Databrary is the unique affordances and challenges of video and audio data"
+%rog The text you removed explained how and why video is actually "data." Maybe that's not important for this audience.
+What are the questions we are trying to answer with this paper?
+I'd argue for the following: 1) Why video?
+2) What challenges does video pose?
+3) How does Databrary solve these challenges?
+4) What challenges remain unmet?
+5) What lessons does Databrary offer other data repositories?
 %Affordances
 We have intentionally chosen to focus our efforts on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
-These types of data have unique affordances but also present additional challenges.
+%rog I don't know that many people understand what the term affordances means.
+These types of data have unique affordances but also present additional challenges \cite{Lanagan_Smeaton_2012}.
 On one hand, recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can glean directly much of the data's richness.
@@
 The session was thus designed to organize data in a way that is familiar to researchers without requiring these modifications.
+By also supporting the upload of multiple video files for a session, Databrary provides a workflow for their research in progress and not merely as a final resting place.
+%dg-I am not married to the above sentence.
 
@@
 
-Once the timeframe of a session is established, this can allow temporally-positioned annotations to be attached to the session.
-% this is sort of an aside:
-%%dg maybe the value here is that we've designed things with this specific type of research in mind.
-In fact, this practice is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio to annotate events, utterances, actions, or scenes at any temporal scale, from individual samples/frames to the entire session.
+Once the timeframe of a session is established, this allows temporally-positioned annotations to be attached to the session.
+Currently these annotations take the form of keyword-based tagging and comment-based discussions on any temporal segment of a session supplied manually by users.
+In future releases we expect to support annotations created by automated video analyis systems.
+
+This practice of annotation is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio to annotate events, utterances, actions, or scenes at any temporal scale, from individual samples/frames to the entire session.
 Most of these tools currently focus on a single recording file, however, and despite the plethora of tools and features, there is little interoperability.
+By supporting this annotation of positioned segments in a session, researchers can not only organize their data, and see it in ways familiar to their research, but also create additional metadata, and thus value of the dataset for other users, on top of it.
 
-In addition to providing an organizational structure that allows users to visualize data in ways familiar to their research, we also currently allow keyword-based tagging and comment-based discussions on any temporal segment of a session.
-These annotations could be applied manually by users or by automated video analysis systems.
-We expect that these annotations will provide an index target for additional discovery capabilities for users looking for particular features within data.
-
-%%dg, i guess it's crucial to make the audience aware that one of the specific challenges for us is that we're not just storing comprehensive single video files, but multiple files that capture the same phenomenon and how do we keep all those together in a meaningful way...to differentiate us from being "just another you tube".
+In addition to enriching datasets for the creators of the data, these annotations will also provide an index target for additional discovery capabilities for users looking for particular features within data.
 

### 2015-01-20 jcdl-15/paper.tex
removing notes
wording

@@
 Video is much "bigger" because it samples space and time densely.
-
 \end{abstract}
@@
 
-%%dg maybe break this part up into subsections: Content/Domain; Users; Design contraints; Approach (or something like that)
-
 Video and audio recordings play a key role in many areas of research, especially in psychology, linguistics, education, ethnography, and ethology.
@@
 
-%\subsection{Affordances and Challenges of Video}
+%\subsection{Benefits and Challenges of Video}
 
-%%dg perhaps better to frame below as the affordances and challenges of working with video data. 
-%That is..."A key component to understanding the design decisions made in creating Databrary is the unique affordances and challenges of video and audio data"
 %rog The text you removed explained how and why video is actually "data." Maybe that's not important for this audience.
@@
 5) What lessons does Databrary offer other data repositories?
-%Affordances
+%Benefitss
 We have intentionally chosen to focus our efforts on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
-%rog I don't know that many people understand what the term affordances means.
-These types of data have unique affordances but also present additional challenges \cite{Lanagan_Smeaton_2012}.
+Though while these types of data are an accessible and potent source for research analysis, they also present additional challenges \cite{Lanagan_Smeaton_2012}.
 On one hand, recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can glean directly much of the data's richness.
@@
 %%dg would a figure illustration be helpful here? do we already have one?
-%%dg in general, how much room do we have for screenshots, if they help explain?
 
@@
 The remainder of this work will address the available structures for data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
-
-% should we actually use the term "record"?  Alternatives: "grouping", "label", "annotation", ...
-% as long as we're clear in defining it. the record section appears to be doing this.
 
@@
 
-%dg "" - stuff like this to remind the reader why we're discussing these components in specific (rather than some other aspect of the system).
 Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
@@
 A session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
-The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
+The metaphor we use here is a timeline <good place for a screenshot> , wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
@@
 
-%%dg maybe add some langauge to the beginning of this section that is explicit that the "record" is designed with the flexibility we talk about above in mind.
 An object called a record serves a number of purposes around session and metadata organization. 
 A record itself represents any number of variables or measurements---what would traditionally be a row of a database or flat file.
+Defining this component broadly supports the flexibility of the system, while giving this data a discrete place in the data schema.
 For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birth date, and any other demographic, survey, or measurement data about that individual.
@@
 \subsection{Technical infrastructure}
-
-%%dg other notes: should we give a bit more discussion of the infrastruction - Postgresql, ffmpeg, using HPC for transcoding, storage, frontend technologies etc. 
-% i'm not sure. do other papers have this sort of "methods" section? -dylan
-%%dg - other papers in this general area (see: Peer_2012, Griffin_etal_2014, and Witt_2012 from references) do comment on the technology used (if there actually is a product and not just a proposed framework) to varying degrees.
-I don't think we have to spend more than a paragraph on it, but it would be good to mention the stack in terms of taking, augmenting, and storing video files and metadata overall.
 

### 2015-01-21 jcdl-15/paper.tex
clean out more notes
reference change, add things to come section
add reference

@@
 In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
-
-%\subsection{Benefits and Challenges of Video}
-
 %rog The text you removed explained how and why video is actually "data." Maybe that's not important for this audience.
@@
 5) What lessons does Databrary offer other data repositories?
-%Benefitss
 We have intentionally chosen to focus our efforts on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
@@
 This naturally reduces the problems of documentation and re-use that plague most flat-file or dense datasets.
-%Challenges
 On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
 We attempt to address these issues by augmenting the primary recording data with additional structured and unstructured features.
-%%dg probably need to make this section more cohesive, but the main gist is 1) We focus on video data, 2)video data has these inherent challenges.
-3) At the same time there are already challenges involved in creating a research data repository, independant of the format of the data travels on.
-4) This paper dicusses the challenges we have resolved and the challenges we have coming up.
 
@@
 
-Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection.
+Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection \cite{Bakeman_2012}.
 Thus, the recording itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
@@
 
-%%dg not sure if this drives us off course, but wondering how much we should discuss library specific challenges in the "stuff to come section", both pertaining to video data and just in general, largely stuff that we will still have to deal with, such as: 1) creation and organization of metadata is still largely manual and time consuming - both in combing videos for stuff to clip out and also trying to extract more metadata from the contributed data itself as well as third party sources (e.g.
-keywords from journals).
-2) interpretation of data lends to this, we can't always rely on contributors to be available to tell us what we need to know about their data 3) metadata interoperability: DOIs; Export in friendly formats (Dublin Core, METS), OAI-PMH compliance.
+\section{Ongoing Challenges and Next Steps}
+
+ 1) Improving the API for interoperability with other systems
+ 2) Metadata interoperability: DOIs; Export in friendly formats (Dublin Core, METS), OAI-PMH compliance.
+ 3) ...
 

### 2015-01-21 jcdl-15/paper.tex
some text and moving notes around

@@
 
-Video and audio recordings play a key role in many areas of research, especially in psychology, linguistics, education, ethnography, and ethology.
-Video documents experimental procedures or behaviors, conveys ideas, provides examples for teaching and scientific communication, and serves as a rich primary data source for later analysis.
-Unfortunately, in standard practice, the use of video data is restricted to the researchers who collect them, due largely to the perceived requirements around privacy restrictions.
-The goal of the Databrary project is to create tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose them for their own research interests.
-Accordingly, Databrary, with an initial focus on research in developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
-In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
 %rog The text you removed explained how and why video is actually "data." Maybe that's not important for this audience.
@@
 5) What lessons does Databrary offer other data repositories?
+
+Video and audio recordings play a key role in many areas of research, especially in psychology, linguistics, education, ethnography, and ethology.
+Video documents experimental procedures or behaviors, conveys ideas, provides examples for teaching and scientific communication, and serves as a rich primary data source for later analysis.
+Unfortunately, in standard practice, the use of video data is restricted to the researchers who collect them, due largely to the perceived requirements around privacy restrictions.
+The goal of the Databrary project is to create tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose them for their own research interests.
+Accordingly, Databrary, with an initial focus on research in developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
+The larger discussion involving research data repositories and the role of libraries in automating and facilitating access to research data well after its original use is very much active and ongoing \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
+While Databrary is a model for tackling some of the challenges there, the scope of this paper focuses more on how we handle research data of a specific format and from a specific discipline, with implications for the future of data cataloging and reuse in other contexts.
+
 We have intentionally chosen to focus our efforts on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.

### 2015-01-22 jcdl-15/paper.tex
add a possible section
though maybe it's not overloading
Reflow only

@@
 The goal of the Databrary project is to create tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose them for their own research interests.
-Accordingly, Databrary, with an initial focus on research in developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+Accordingly, Databrary, with an initial focus on developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
 In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
@@
 
+\section{Implications for Digital Libraries}
+%%dg this might or might not be valuable but I am wondering if this is a section that might help draw all the pieces together.
+Or maybe this just boils down to a sentence or two in the next steps or conclusions
+
 \section{Ongoing Challenges and Next Steps}

### 2015-01-22 jcdl-15/paper.tex
Un-reorder and edit intro

--- 2015-01-21 jcdl-15/paper.tex
@@
 The goal of the Databrary project is to create tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose them for their own research interests.
-Accordingly, Databrary, with an initial focus on research in developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+Accordingly, Databrary, with an initial target audience of developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
 In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
-%rog The text you removed explained how and why video is actually "data." Maybe that's not important for this audience.
-What are the questions we are trying to answer with this paper?
-I'd argue for the following: 1) Why video?
-2) What challenges does video pose?
-3) How does Databrary solve these challenges?
-4) What challenges remain unmet?
-5) What lessons does Databrary offer other data repositories?
-We have intentionally chosen to focus our efforts on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
-Though while these types of data are an accessible and potent source for research analysis, they also present additional challenges \cite{Lanagan_Smeaton_2012}.
-On one hand, recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can glean directly much of the data's richness.
+
+Existing research data repositories present two competing approaches that need to be balanced out.
+A system might tend towards more generality, providing unstructured storage of arbitrary files.
+Alternatively, it can present a rigorously structured schema that requires contributors to properly extract and format contributions in a highly specific manner.
+While general-purpose repositories are well suited for archival of entire datasets and flexible hosting of resources, they do not facilitate detailed search, understanding, or re-use across various datasets without a great deal of curation \cite{Peer_2012}.
+Targeted repositories, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but are limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
+We have attempted to strike a balance between these extremes by focusing on video and audio recordings as the unifying factor and allowing additional structured and unstructured resources to be built and discovered around this core.
+
+We have intentionally chosen to focus Databrary on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
+While these types of data are an accessible source for research analysis, they also present additional challenges \cite{Lanagan_Smeaton_2012}.
+Recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can glean directly much of the data's richness.
 In particular, video data collected in one context can often be used by others for different purposes with minimal explanation, as the nature of the subjects and activity are readily apparent.
 This naturally reduces the problems of documentation and re-use that plague most flat-file or dense datasets.
-On the other hand, since automated analysis of videos is an active yet immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
-We attempt to address these issues by augmenting the primary recording data with additional structured and unstructured features.
+However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
 
 %\section{Normalization vs. Flexibility}
-
-We have found in researching existing research data repositories that there are two competing approaches that need to be balanced out.
-On the one hand, a system might tend towards more generality, providing unstructured storage of arbitrary files.
-On the other, it can present a rigorously structured schema that requires contributors to properly extract and format contributions in a highly specified manner.
-While the former is well suited for archival of entire datasets and general-purpose hosting of resources, it does not facilitate detailed search, understanding, or re-use across various datasets without a great deal of curation \cite{Peer_2012}.
-The latter, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but are limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
-We have attempted to strike a balance between these extremes by focusing on video and audio recordings as the unifying factor and allowing additional resources to be built and discovered around this core.
 
@@
 
+\section{Implications for Digital Libraries}
+%%dg this might or might not be valuable but I am wondering if this is a section that might help draw all the pieces together.
+Or maybe this just boils down to a sentence or two in the next steps or conclusions
+
+There is active discussion around the role of repositories and libraries in automating and facilitating access to research data \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
+While Databrary has some implications for the future of data cataloging and reuse in these contexts, our focus is on handling active research data of specific formats from a specific discipline.
+
 \section{Ongoing Challenges and Next Steps}

### 2015-01-23 jcdl-15/paper.tex
Another full pass, some reorg, cleanup some refs
Still need an end, probably more on datavyu2

@@
 
+\subsection{Video-driven research}
+
 We have intentionally chosen to focus Databrary on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
@@
 
-%\section{Normalization vs. Flexibility}
+Research using recordings of these types typically proceeds by annotating or scoring raw data.
+In order to extract value from recordings for further analysis, researchers add temporally-positioned annotations, either by direct observation or with the aid of simple visualizations or analysis tools.
+Myriad specialized desktop software tools are employed to review video and audio to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
+Most of these tools currently focus on a single recording file, and despite the plethora of tools and features, there is little interoperability.
+Moreover, there is almost no standardization of these annotations themselves, even across studies performed by the same researcher.
+
+\subsection{Approach}
 
@@
 
-There are two key, innovative structural elements that comprise a volume: the session and the record.
-Sessions are the primary container for data files, basically an elaboration on traditional file folders with additional temporal metadata.
+There are two key, innovative structural elements that comprise a volume: the \emph{session} and the \emph{record}.
+Sessions are the primary container for data files, an elaboration on traditional file folders with additional temporal metadata.
 Records are used to represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as to provide an organizational structure among sessions.
@@
 
-\subsection{The ``Session''}
+\subsection{Sessions}
 
@@
 
-When working with multiple primary recordings, researchers often reduce these raw files into a single representation.
+When working with multiple primary recordings, researchers often combine these raw files into a single representation.
 For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
@@
 The session was thus designed to organize data in a way that is familiar to researchers without requiring these modifications.
-By also supporting the upload of multiple video files for a session, Databrary provides a workflow for their research in progress and not merely as a final resting place.
-%dg-I am not married to the above sentence.
 
 A session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
-The metaphor we use here is a timeline <good place for a screenshot> , wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
+The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
+% good place for a screenshot?
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
@@
 Once the timeframe of a session is established, this allows temporally-positioned annotations to be attached to the session.
-Currently these annotations take the form of keyword-based tagging and comment-based discussions on any temporal segment of a session supplied manually by users.
-In future releases we expect to support annotations created by automated video analyis systems.
+In addition to the structured and organizational annotations discussed in the next section, we currently support simple keyword-based tagging and comment-based discussions on any temporal segment of a session.
+Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
+In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data.
 
-This practice of annotation is very common in research when analyzing recordings, and specialized desktop software tools are employed to review video and audio to annotate events, utterances, actions, or scenes at any temporal scale, from individual samples/frames to the entire session.
-Most of these tools currently focus on a single recording file, however, and despite the plethora of tools and features, there is little interoperability.
-By supporting this annotation of positioned segments in a session, researchers can not only organize their data, and see it in ways familiar to their research, but also create additional metadata, and thus value of the dataset for other users, on top of it.
-
-In addition to enriching datasets for the creators of the data, these annotations will also provide an index target for additional discovery capabilities for users looking for particular features within data.
-
-\subsection{The ``Record''}
+\subsection{Records}
 
 An object called a record serves a number of purposes around session and metadata organization.
-A record itself represents any number of variables or measurements---what would traditionally be a row of a database or flat file.
-Defining this component broadly supports the flexibility of the system, while giving this data a discrete place in the data schema.
+A record itself represents any number of variables or measurements: what would traditionally be a row of a database or flat file.
+Records can thus represent most any item of analysis or organization typically represented in this form.
 For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birth date, and any other demographic, survey, or measurement data about that individual.
-A record could also represent a particular experimental task, condition, location, or outcome, and contain variables with corresponding parameters or descriptions.
+A record could also represent a particular experimental trial, task, condition, location, or outcome, and contain variables with corresponding measurements, parameters, descriptions.
 Records are thus labeled and divided by the type of entity they represent.
@@
 We observed that different researchers organize their data files in different ways according to the type of research.
-Many consider participants to be the primary item of analysis, and so create a directory for each participant.
+Many consider participants to be the primary item of analysis, and so create a filesystem directory for each participant.
 Others organize their data based on date, experimental condition, phase of analysis, or age groups.
@@
 Since researchers can create whichever record types apply to their data, they can then choose to group their data by whichever record type they wish.
-This also allows users to either import or export data in what they consider the native directory structure by indicating the appropriate nesting of record types.
-By providing some standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants of a particular age range.
+This also allows users to either import or export data in what they consider the native directory structure by indicating the appropriate nesting of record types, along with the record variables themselves in traditional CSV formats.
+By providing some standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants within a particular age range.
 
@@
 Finally, records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
-That is, if any record only applies to part of a session, for example a single activity, or a participant being present for only part of a recording, it can be applied within the timeline of a session to only the appropriate part.
+That is, if any record only applies to part of a session, such as a single scene or section---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only the appropriate part.
 This allows the final type of organizational structure, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
-It also can be used to represent annotations at a finer temporal scale, for example individual events interactions that are normally handled by off-line annotation software.
+It also can be used to represent annotations at a finer temporal scale, for example individual trials or events that are normally handled by off-line annotation software.
 However, the appropriate interfaces and tools to input or import these annotations are still being developed.
@@
 
-Databrary is a new, cohesive web application, built on the Play Framework to enable a responsive user interface, a complete API, and high-performance streaming.
+Databrary is a new, cohesive web application, built in Scala on the Play Framework\footnote{http://playframework.com/} to enable a responsive user interface, a complete API, and high-performance streaming.
 Recordings and other files are placed in content-based filesystem storage, and all structured data are stored in a PostgreSQL database, leveraging its geometric indexing capabilities for temporal data.
@@
 This transcoding utilizes the high performance computing cluster at the hosting campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
-The user interface is built primarily on the Angular web framework, and all data access is performed through an open JSON API.
-Although hosted data are protected, all source code is released under a GPLv3 license on github.
+The user interface is built primarily on the Angular web framework\footnote{http://angularjs.org/}, and all data access is performed through an open JSON API.
+Although hosted data are protected, all source code is released under a GPLv3 license on github\footnote{http://github.com/databrary}.
 
-\section{Normalization vs. Flexibility}
+\section{Repository}
+
+The Databrary website was opened for general use in October 2014 and accepting contributions since earlier that year.
+It currently hosts 5700 video files totaling 1600 hours of recordings along with 2200 additional files.
+These files comprise 2400 sessions and are covered by 1300 records (including 1100 individual participants).
+Data originates from 35 individual contributors across 25 different universities.
+
+\subsection{Normalization vs. Flexibility}
 
 As the extant research data libraries demonstrate, there is always the tension between normalization and flexibility.
-Since we're primarily dealing with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
-We also allow for ``dumps'', which allow users---using certain portable formats (flat-files, documents, images, data)---to include any data that cannot be encompassed by our standard formats.
+Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
+We also allow allow users, using certain portable formats (flat-files, documents, images), to include any data that cannot be encompassed by our standard formats.
 In this way, we can incorporate data which may be relevant for re-use but not broadly adopted by the community, without making the system too rigid.
@@
 For many previously collected datasets, this has proven the most effective method to both learn from data and avoid burdening contributors.
-However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a separate, user-driven approach.
+However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a distinct, user-driven approach.
 
@@
 
-Achieving the appropriate balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
+Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
 
-\section{Implications for Digital Libraries}
+\subsection{Implications for Digital Libraries}
 %%dg this might or might not be valuable but I am wondering if this is a section that might help draw all the pieces together.
@@
 
-\section{Ongoing Challenges and Next Steps}
+\section{Conclusions}
+
+\subsection{Ongoing Challenges and Next Steps}
 
@@
 
-\section{Conclusions}
+coding:
+We are also working to support import and export of these annotations so researchers can continue to use their existing tools.
+By supporting temporally-positioned session annotations directly within our system, researchers will be able to work with their recordings directly in raw form without having to transfer or transform it.
 

### 2015-01-23 jcdl-15/paper.tex
Add discovery section

@@
 
-The Databrary website was opened for general use in October 2014 and accepting contributions since earlier that year.
+The Databrary website was opened for general use in October 2014 and has accepted contributions since earlier that year.
 It currently hosts 5700 video files totaling 1600 hours of recordings along with 2200 additional files.
@@
 
-\subsection{Normalization vs. Flexibility}
+\subsection{Normalization}
 
@@
 
+\subsection{Discovery}
+
+While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely unsolved.
+This problem is compounded for us because we expect researchers not only to reuse and reanalyze data within their own research area, but potentially to find new uses data originally collected for entirely different purposes.
+Recordings with potential value in one research area may be primarily labeled with contextual information from a different research area.
+Ultimately, making this determination requires direct, human observation of the recordings, and so a large part of solving the discovery problem will be simply to present researchers with short, representative samples from a variety of datasets.
+
+Another important source for search targets comes from annotations supplied both by the original researchers during analysis as well as by the community.
+In the former case, normalization remains a concern, as most annotation schemes researchers employ involve opaque, symbolic or numeric codes with little semantic content.
+We are currently working to support import and export workflows for existing annotations tools that will be able to add meaningful definitions to these codes while researchers continue their current practices.
+Beyond that, we also plan to build a new, more efficient and structured annotation web interface enabling more of the analysis workflow directly within our system
+In this way, Databrary will allow researchers to work with their recordings directly in raw form without having to transfer or transform them, while also capturing more meaningful and standardized contextual information from this process.
+
+\section{Conclusions}
+
 \subsection{Implications for Digital Libraries}
@@
 While Databrary has some implications for the future of data cataloging and reuse in these contexts, our focus is on handling active research data of specific formats from a specific discipline.
-
-\section{Conclusions}
 
@@
 coding:
-We are also working to support import and export of these annotations so researchers can continue to use their existing tools.
-By supporting temporally-positioned session annotations directly within our system, researchers will be able to work with their recordings directly in raw form without having to transfer or transform it.
 

### 2015-01-23 jcdl-15/paper.tex
add notes, some edits, move some stuff around

@@
 A system might tend towards more generality, providing unstructured storage of arbitrary files.
-Alternatively, it can present a rigorously structured schema that requires contributors to properly extract and format contributions in a highly specific manner.
-While general-purpose repositories are well suited for archival of entire datasets and flexible hosting of resources, they do not facilitate detailed search, understanding, or re-use across various datasets without a great deal of curation \cite{Peer_2012}.
+Alternatively, it can present a rigorously structured schema that requires contributors to format contributions in a highly specific manner.
+While general-purpose repositories are well-suited for archival of entire datasets and flexible hosting of resources, they do not facilitate detailed search, understanding, or re-use across various datasets without a great deal of curation \cite{Peer_2012}.
 Targeted repositories, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but are limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
@@
 While these types of data are an accessible source for research analysis, they also present additional challenges \cite{Lanagan_Smeaton_2012}.
-Recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can glean directly much of the data's richness.
+Recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can directly glean much of the data's richness.
 In particular, video data collected in one context can often be used by others for different purposes with minimal explanation, as the nature of the subjects and activity are readily apparent.
-This naturally reduces the problems of documentation and re-use that plague most flat-file or dense datasets.
+This naturally reduces the problems of documentation and re-use that plague most flat-file, largely text-based datasets.
 However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
@@
 Most of these tools currently focus on a single recording file, and despite the plethora of tools and features, there is little interoperability.
-Moreover, there is almost no standardization of these annotations themselves, even across studies performed by the same researcher.
+Moreover, there is almost no standardization of these annotations themselves, even across studies performed by the same researcher \cite{?}.
 
 \subsection{Approach}
-
-Databrary can incorporate heterogeneous datasets within our target domain by allowing secondary, general-purpose flat-file data, materials, and references to be associated with the primary data.
+%%dg tried to add to this section, but it feels like it needs a bit more
+Databrary's approach to data is to provide as much context around the video data as possible and provide users a variety of ways to bring their data in.
+To keep datasets intact, Databrary can incorporate heterogeneous datasets within our target domain by allowing secondary, general-purpose flat-file data, materials, and references to be associated with the primary, video data.
 This facilitates storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
 At the same time, it provides a common, standard structure for recordings and their associated metadata, including context, contents, and annotations.
-We achieve this conformance of datasets in the library through a combination of in-depth investigative curation and ingest processes for already-collected data, and a well-defined user experience that provides a platform for researchers to contribute and organize all of their research data as they are collected.
+We achieve this conformance of datasets in the library through a combination of in-depth investigative curation and automated, server-side ingest processes for already-collected data.
+For datasets that are in the process of being collecting, a well-defined user experience provides a platform for researchers to contribute and organize all of their research data as they proceed with their study.
 The following section provides specific detail on the design and decision-making behind this unifying structure.
@@
 
-As with most user-driven databases, we separate individual data contributions into collections or packages variously referred to as datasets, studies, or volumes.
+At the top level of organization, we separate individual data contributions into collections or packages variously referred to as datasets, studies, or volumes.
 Volumes can be associated with a minimal amount of metadata (title, description, permissions, etc.) as well as additional files (documents and other materials).
 These volumes are to some extent embeddable and referenceable, both among themselves and to and from external entities by DOI or URI.
+%%dg can we expand on the above sentence, what does this mean and what are the implications?
 The remainder of this work will address the available structures for data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
@@
 
-Researchers record activities with individuals either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection \cite{Bakeman_2012}.
+The session in Databrary draws from the accepted and documented practice of researchers recording activities with individuals in sessions either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection \cite{Bakeman_2012}.
 Thus, the recording itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
@@
 Doing so allows for more convenient human analysis of the data in standard playback tools that only support a single data stream, but also loses some of the raw data, by obscuring, masking, or downsampling spatially or temporally.
-The session was thus designed to organize data in a way that is familiar to researchers without requiring these modifications.
+Though by allowing the option to add multiple videos that are related, without compiling them, the session was thus designed to organize data in a way that is familiar to researchers without requiring these modifications as well as to allow the research keep their data in the rawest form, so as not to lose analysis options in uploading their data.
+This additionally allows for greater options in reuse by other researchers.
 
-A session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
+Because Databrary allows for multiple videos in a single session that may occur at different times, data cab be represented temporally.
+That is, a session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
 The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
@@
 \subsubsection*{Simple annotations}
-
+%%dg this feels like it might be padded a little bit.
 Once the timeframe of a session is established, this allows temporally-positioned annotations to be attached to the session.
@@
 Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
-In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data.
+In addition to enriching datasets and enabling dialog between researchers through the additional layer of information on top of the original data, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
 
@@
 \subsubsection*{Structured annotations}
-
+%%dg - might confuse the audience a bit in the difference betweetn simple and structured annotations
 Finally, records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
@@
 
-While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely unsolved.
+While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely to be implemented.
 This problem is compounded for us because we expect researchers not only to reuse and reanalyze data within their own research area, but potentially to find new uses data originally collected for entirely different purposes.
@@
 
+%%dg is any of the following redundant with the annotation discussion above?
 Another important source for search targets comes from annotations supplied both by the original researchers during analysis as well as by the community.
@@
 \section{Conclusions}
+%%dg - do we not want to have this as the last section?
 
 \subsection{Implications for Digital Libraries}
-%%dg this might or might not be valuable but I am wondering if this is a section that might help draw all the pieces together.
-Or maybe this just boils down to a sentence or two in the next steps or conclusions
 
 There is active discussion around the role of repositories and libraries in automating and facilitating access to research data \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
-While Databrary has some implications for the future of data cataloging and reuse in these contexts, our focus is on handling active research data of specific formats from a specific discipline.
+In the technical structure and approach laid out above, Databrary does present a functioning model for a system that can wrangle and facilitate research data management in a way that could make the job of libraries aspiring to store and provide access to research data more feasible.
+By building in increased interoperability of datastreams for the contents in Databrary using library-based metadata schemas (Dublin Core, METS) and OAI-PMH compliance, the data researchers add to systems like Databrary can then automatically be added into federated library searches with other domain specific data repositories.
+Additional interoperability like a more refined API and assigning Digital Obejct Identifiers (DOIs) to Databrary volumes, we can enhance the ways library and other information systems can tap into the datasets compiled in Databrary.
 
 \subsection{Ongoing Challenges and Next Steps}
-
- 1) Improving the API for interoperability with other systems
- 2) Metadata interoperability: DOIs; Export in friendly formats (Dublin Core, METS), OAI-PMH compliance.
- 3) ...
 

### 2015-01-24 jcdl-15/paper.tex
Another pass, mainly trimming

@@
 In particular, video data collected in one context can often be used by others for different purposes with minimal explanation, as the nature of the subjects and activity are readily apparent.
-This naturally reduces the problems of documentation and re-use that plague most flat-file, largely text-based datasets.
+This naturally reduces the problems of documentation and re-use that plague most flat-file datasets. % wanted to include raw, opaque data here, too
 However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
@@
 \subsection{Approach}
+
 %%dg tried to add to this section, but it feels like it needs a bit more
-Databrary's approach to data is to provide as much context around the video data as possible and provide users a variety of ways to bring their data in.
-To keep datasets intact, Databrary can incorporate heterogeneous datasets within our target domain by allowing secondary, general-purpose flat-file data, materials, and references to be associated with the primary, video data.
+Databrary's approach to data is to provide as much context around the primary recordings as possible while still providing contributors some flexibility.
+To keep datasets intact, Databrary can incorporate heterogeneous datasets within our target domain by allowing secondary, general-purpose flat-file data, materials, and references to be associated with the primary data.
 This facilitates storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
 At the same time, it provides a common, standard structure for recordings and their associated metadata, including context, contents, and annotations.
-We achieve this conformance of datasets in the library through a combination of in-depth investigative curation and automated, server-side ingest processes for already-collected data.
-For datasets that are in the process of being collecting, a well-defined user experience provides a platform for researchers to contribute and organize all of their research data as they proceed with their study.
+We achieve this conformance for previously collected datasets through a combination of in-depth investigative curation and automated, server-side ingest processes.
+For datasets that are in the process of being collected, a well-defined user experience provides a platform for researchers to contribute and organize all of their research data as they proceed with their study.
 The following section provides specific detail on the design and decision-making behind this unifying structure.
@@
 
-The session in Databrary draws from the accepted and documented practice of researchers recording activities with individuals in sessions either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection \cite{Bakeman_2012}.
+The session in Databrary draws from accepted and documented practices around how researchers record activities with individuals.
+Recordings are made either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection \cite{Bakeman_2012}.
 Thus, the recording itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
@@
 Doing so allows for more convenient human analysis of the data in standard playback tools that only support a single data stream, but also loses some of the raw data, by obscuring, masking, or downsampling spatially or temporally.
-Though by allowing the option to add multiple videos that are related, without compiling them, the session was thus designed to organize data in a way that is familiar to researchers without requiring these modifications as well as to allow the research keep their data in the rawest form, so as not to lose analysis options in uploading their data.
+The session was thus designed to organize original recordings in a way that is familiar to researchers, keeping their data in the rawest and most flexible form without requiring these modifications.
 This additionally allows for greater options in reuse by other researchers.
 
-Because Databrary allows for multiple videos in a single session that may occur at different times, data cab be represented temporally.
-That is, a session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
+A session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
 The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
@@
 \subsubsection*{Simple annotations}
+
 %%dg this feels like it might be padded a little bit.
@@
 Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
-In addition to enriching datasets and enabling dialog between researchers through the additional layer of information on top of the original data, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
+In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
 
@@
 
-While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely to be implemented.
+While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely unsolved.
 This problem is compounded for us because we expect researchers not only to reuse and reanalyze data within their own research area, but potentially to find new uses data originally collected for entirely different purposes.
@@
 
-There is active discussion around the role of repositories and libraries in automating and facilitating access to research data \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
-In the technical structure and approach laid out above, Databrary does present a functioning model for a system that can wrangle and facilitate research data management in a way that could make the job of libraries aspiring to store and provide access to research data more feasible.
-By building in increased interoperability of datastreams for the contents in Databrary using library-based metadata schemas (Dublin Core, METS) and OAI-PMH compliance, the data researchers add to systems like Databrary can then automatically be added into federated library searches with other domain specific data repositories.
-Additional interoperability like a more refined API and assigning Digital Obejct Identifiers (DOIs) to Databrary volumes, we can enhance the ways library and other information systems can tap into the datasets compiled in Databrary.
+There is active discussion around the role of repositories and libraries in archiving and providing access to research data \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
+In the approach laid out above, Databrary presents a functioning system for research data management that may bring data closer established library standards that achieve these goals.
+Ultimately, we hope to provide increased interoperability with library-based metadata schemas (Dublin Core, METS) and reach OAI-PMH compliance, so that data researchers add to Databrary can be automatically incorporated into federated library searches with other domain specific data repositories.
+By providing a refined API and assigning Digital Obejct Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
 
 \subsection{Ongoing Challenges and Next Steps}
-
-coding:
 

### 2015-01-25 jcdl-15/paper.tex
Complete conclusion, but don't like the end

@@
 \section{Conclusions}
-%%dg - do we not want to have this as the last section?
 
-\subsection{Implications for Digital Libraries}
+Starting from the perspective of a general-purpose repository, we have introduced a number of significant restrictions to create a specialized, domain repository for specific types of data.
+By limiting both file and metadata organization to a flat (but overlapping) set of sessions and records, rather than supporting the full hierarchical richness of other systems, we can present and capture session-based research data with a cohesive and intuitive user interface.
+By focusing on video and audio recordings we can normalize presentation and discovery of data across the site, and also introduce temporal structure to data and annotations driven by the nature of these recordings.
+Although we allow a few types of additional data files, we encourage users with other types of rich data, such as physiological recordings, to store and link to those resources externally, ideally in similarly focused and better-suited repositories.
+Similarly, by targeting use cases around finding new uses for existing research data, rather than replication or meta-analysis, we can limit the scope of accepted data to the original recordings and early analysis phases that generate the most value, rather than trying to capture the entire provenance.
 
-There is active discussion around the role of repositories and libraries in archiving and providing access to research data \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
-In the approach laid out above, Databrary presents a functioning system for research data management that may bring data closer established library standards that achieve these goals.
-Ultimately, we hope to provide increased interoperability with library-based metadata schemas (Dublin Core, METS) and reach OAI-PMH compliance, so that data researchers add to Databrary can be automatically incorporated into federated library searches with other domain specific data repositories.
-By providing a refined API and assigning Digital Obejct Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
+Having made significant progress building ingest and data entry services that have let us grow the library, we now must turn our efforts towards discovery and reuse, finding ways to better describe and elaborate on collected data.
+To do this we look both to the original data contributors for ways that their existing analysis workflows can generate usable content descriptions, and to the community of researchers to label and comment on datasets as they explore.
+Thus we seek to incorporate these existing annotation processes into Databrary, both to capture and define these annotations as generalizable search terms, and to make these workflows more convenient by obviating existing file transfer and transformation burdens.
+Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.
 
-\subsection{Ongoing Challenges and Next Steps}
+There is also active discussion around the role of repositories and libraries in archiving and providing access to research data \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
+Although this is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer established library standards that achieve these goals.
+We may be well positioned to provide interoperability with library-based metadata schemas (Dublin Core, METS) and reach OAI-PMH compliance, so that data researchers add to Databrary can be automatically incorporated into federated library searches with other domain specific data repositories.
+By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
+
+% this is kinda weird, but maybe suitable for a last paragraph?
+Our single-minded focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers gives us a novel perspective on building domain-specific research repositories.
+The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
+By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the library science community to start reaping their benefits.
 

### 2015-01-25 jcdl-15/paper.tex
fill out the rest, good enough for a first draft
some jcdl edits

@@
 
-%%dg tried to add to this section, but it feels like it needs a bit more
 Databrary's approach to data is to provide as much context around the primary recordings as possible while still providing contributors some flexibility.
@@
 Although this is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer established library standards that achieve these goals.
-We may be well positioned to provide interoperability with library-based metadata schemas (Dublin Core, METS) and reach OAI-PMH compliance, so that data researchers add to Databrary can be automatically incorporated into federated library searches with other domain specific data repositories.
+We may be well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS schemas) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
 By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
@@
 The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
-By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the library science community to start reaping their benefits.
+By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.
 

### 2015-01-26 jcdl-15/paper.tex
add reference to conclusion

@@
 
-There is also active discussion around the role of repositories and libraries in archiving and providing access to research data \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014}.
-Although this is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer established library standards that achieve these goals.
+There is also active discussion around the role of repositories and libraries in archiving and providing access to research data, where the roles and responsibilities between libraries and research departments is not entirely established \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014, Pinfield_etal_2014}.
+Although resolving this complicated issue is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer established library standards that achieve these goals.
 We may be well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS schemas) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.

### 2015-01-26 jcdl-15/paper.tex
combine annotations section, leaving a little in a messy state

@@
 
-\subsubsection*{Simple annotations}
-
-%%dg this feels like it might be padded a little bit.
-Once the timeframe of a session is established, this allows temporally-positioned annotations to be attached to the session.
-In addition to the structured and organizational annotations discussed in the next section, we currently support simple keyword-based tagging and comment-based discussions on any temporal segment of a session.
-Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
-In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
-
 \subsection{Records}
@@
 
-\subsubsection*{Structured annotations}
-%%dg - might confuse the audience a bit in the difference betweetn simple and structured annotations
-Finally, records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
-That is, if any record only applies to part of a session, such as a single scene or section---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only the appropriate part.
+\subsubsection*{Annotations}
+
+%%dg - note on this section: 1) we can add tags to sessions and portions of sections 2) in adding tags to portions of sessions, we should be able to import data from desktop coding software as tags on portions of sessions.
+
+%%simple
+The establishing a timeline for each session allows temporally-positioned, simple annotations to be attached to that session.
+In addition to the structured and organizational annotations discussed in the next section, we currently support simple keyword-based tagging and comment-based discussions on any temporal segment of a session.
+Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
+In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
+
+%%structured
+Yet to be implemented is the ability to add structured annotations to sections of a session. Records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
+If any record only applies to part of a session, such as a single scene or section---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only that appropriate part.
 This allows the final type of organizational structure, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
+%might need to refit this idea in
 It also can be used to represent annotations at a finer temporal scale, for example individual trials or events that are normally handled by off-line annotation software.
@@
 
-%%dg is any of the following redundant with the annotation discussion above?
-Another important source for search targets comes from annotations supplied both by the original researchers during analysis as well as by the community.
+Another important source for search targets would come from annotations supplied both by the original researchers during analysis as well as by the community.
 In the former case, normalization remains a concern, as most annotation schemes researchers employ involve opaque, symbolic or numeric codes with little semantic content.

### 2015-01-26 jcdl-15/paper.tex
Reorg new annotation section

@@
 
-\subsubsection*{Annotations}
+\subsection{Annotations}
 
-%%dg - note on this section: 1) we can add tags to sessions and portions of sections 2) in adding tags to portions of sessions, we should be able to import data from desktop coding software as tags on portions of sessions.
+Establishing a timeline for each session allows temporally-positioned annotations to be attached to the session.
+Currently we support two types of annotations: structured, organizational annotations based on records and added by the data contributors; and simple annotations contributed by the community.
+Eventually we hope to extend both of these to support the full range of annotations performed during analysis, both directly on the web and imported from existing software.
 
-%%simple
-The establishing a timeline for each session allows temporally-positioned, simple annotations to be attached to that session.
-In addition to the structured and organizational annotations discussed in the next section, we currently support simple keyword-based tagging and comment-based discussions on any temporal segment of a session.
+First, records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
+If any record relates to a part, scene, or section of a session---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only that appropriate part.
+This allows an additional type of organizational structure researchers employ, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
+This can also theoretically be used to represent annotations at a finer temporal scale, for example individual trials or events that are normally handled by off-line annotation software.
+However, the appropriate interfaces and tools to input or import these annotations are still being developed.
+
+Second, we support simple keyword-based tagging and comment-based discussions on any temporal segment of a session.
 Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
 In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
-
-%%structured
-Yet to be implemented is the ability to add structured annotations to sections of a session. Records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
-If any record only applies to part of a session, such as a single scene or section---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only that appropriate part.
-This allows the final type of organizational structure, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
-%might need to refit this idea in
-It also can be used to represent annotations at a finer temporal scale, for example individual trials or events that are normally handled by off-line annotation software.
-However, the appropriate interfaces and tools to input or import these annotations are still being developed.
 

### 2015-01-27 jcdl-15/paper.tex
add references and some precarious text

@@
 In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
+Where research of community tagging provides evidence for the case that not many users actually do participate in tagging, we intend to leverage automated tagging through already existing metadata and analysis to seed a base of tags for reuse by the community \cite{Yang_Lu_Giles_2011, Farooq_etal_2007}.
+%%dg - might move the last sentence around, but provides a gist.
 
@@
 We are currently working to support import and export workflows for existing annotations tools that will be able to add meaningful definitions to these codes while researchers continue their current practices.
-Beyond that, we also plan to build a new, more efficient and structured annotation web interface enabling more of the analysis workflow directly within our system
+Beyond that, we also plan to build a new, more efficient and structured annotation web interface enabling more of the analysis workflow directly within our system.
 In this way, Databrary will allow researchers to work with their recordings directly in raw form without having to transfer or transform them, while also capturing more meaningful and standardized contextual information from this process.

### 2015-01-28 jcdl-15/paper.tex
updates only to first line

@@
 
-Video and audio recordings play a key role in many areas of research, especially in psychology, linguistics, education, ethnography, and ethology.
-Video documents experimental procedures or behaviors, conveys ideas, provides examples for teaching and scientific communication, and serves as a rich primary data source for later analysis.
+Video and audio recordings play a key role in many areas of research and often serves as a primary data source in domains such as psychology, linguistics, education, anthropology, and numerous other fields.
 Unfortunately, in standard practice, the use of video data is restricted to the researchers who collect them, due largely to the perceived requirements around privacy restrictions.
@@
 Our single-minded focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers gives us a novel perspective on building domain-specific research repositories.
-The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
+The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discocvery that many more traditional repositories avoid.
 By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.

### 2015-01-28 jcdl-15/paper.tex
Fix typo

@@
 Our single-minded focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers gives us a novel perspective on building domain-specific research repositories.
-The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discocvery that many more traditional repositories avoid.
+The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
 By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.

### 2015-01-29 jcdl-15/paper.tex
comments thru section 1

@@
 Rick O. Gilmore\\
-	\affaddr{The Pennsylvania State University}\\
+	\affaddr{Penn State}\\
 	\affaddr{Department of Psychology}\\
@@
 
-Video and audio recordings play a key role in many areas of research and often serves as a primary data source in domains such as psychology, linguistics, education, anthropology, and numerous other fields.
-Unfortunately, in standard practice, the use of video data is restricted to the researchers who collect them, due largely to the perceived requirements around privacy restrictions.
-The goal of the Databrary project is to create tools that allow video data to be documented, organized, and shared so that other researchers can discover and re-purpose them for their own research interests.
-Accordingly, Databrary, with an initial target audience of developmental psychology and related fields, has created a web-based library to accommodate the storage and organization of video data, capture contextual information necessary for re-use, and standardize across datasets without placing undue burden on data contributors.
+Video and audio recordings serve as a primary data source in psychology, linguistics, education, anthropology, and numerous other fields, especially the study of human and animal development.
+Unfortunately, researchers who collect video rarely share recordings with others, due largely to perceived requirements associated with privacy restrictions.
+The goal of the Databrary project is to increase the repurposing and reuse of recordings, especially video, by creating tools that allow video to be documented, organized, and shared among researchers.
+We have created a web-based data library to accommodate the storage and organization of video data, capture contextual information necessary for reuse, and standardize across datasets without placing undue burden on data contributors.
 In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
 
-Existing research data repositories present two competing approaches that need to be balanced out.
-A system might tend towards more generality, providing unstructured storage of arbitrary files.
-Alternatively, it can present a rigorously structured schema that requires contributors to format contributions in a highly specific manner.
-While general-purpose repositories are well-suited for archival of entire datasets and flexible hosting of resources, they do not facilitate detailed search, understanding, or re-use across various datasets without a great deal of curation \cite{Peer_2012}.
-Targeted repositories, on the other hand, generate large-scale, homogeneous data sources for meta- and re-analysis, but are limited to specific domains and research questions, losing the full richness, context, and innovation of the source materials and field.
-We have attempted to strike a balance between these extremes by focusing on video and audio recordings as the unifying factor and allowing additional structured and unstructured resources to be built and discovered around this core.
+\subsection{Balancing Generality and Specificity}
+
+Research data repositories require a balance between generality and specificity.
+Archives that permit the unstructured storage of arbitrary files lie at one extreme;
+those that present a rigorously structured schema and required contribution formats lie at the other.
+The flexibility of general-purpose repositories make them especially well-suited to the archival of entire datasets, but these systems do not facilitate search, understanding, or re-use without extensive, usually labor-intensive, curation \cite{Peer_2012}.
+Targeted repositories, on the other hand, generate large-scale, homogeneous data sources that facilitate meta- and re-analysis, but serve only specific domains. 
+We have attempted to strike a balance between these extremes by making video and audio recordings the unifying focus of Databrary.
+This allows additional structured and unstructured resources to be built around and discovered through this functional core.
 
@@
 
-We have intentionally chosen to focus Databrary on hosting human-consumable, self-documenting, timeseries data: video and audio recordings.
-While these types of data are an accessible source for research analysis, they also present additional challenges \cite{Lanagan_Smeaton_2012}.
-Recordings can be and largely are consumed and analyzed, at least initially, without sophisticated tools: once digitally decoded, trained human observers can directly glean much of the data's richness.
-In particular, video data collected in one context can often be used by others for different purposes with minimal explanation, as the nature of the subjects and activity are readily apparent.
-This naturally reduces the problems of documentation and re-use that plague most flat-file datasets. % wanted to include raw, opaque data here, too
+Video and audio recordings constitute a form of human-consumable, self-documenting, timeseries data.
+Recordings can be and largely are consumed and analyzed, at least initially, without the use of sophisticated tools: once digitally decoded, trained human observers can directly glean much of the data's richness.
+Video data collected in one context can often be used by others for different purposes with minimal explanation.
+This naturally reduces the problems of documentation and re-use that plague many flat-file datasets. % wanted to include raw, opaque data here, too
 However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
+In short, recordings pose challenges that older, less dynamic media do not \cite{Lanagan_Smeaton_2012}.
 
-Research using recordings of these types typically proceeds by annotating or scoring raw data.
-In order to extract value from recordings for further analysis, researchers add temporally-positioned annotations, either by direct observation or with the aid of simple visualizations or analysis tools.
-Myriad specialized desktop software tools are employed to review video and audio to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
-Most of these tools currently focus on a single recording file, and despite the plethora of tools and features, there is little interoperability.
-Moreover, there is almost no standardization of these annotations themselves, even across studies performed by the same researcher \cite{?}.
+Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
+Specialized %commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (Transana, Elan, CLAN, and Datavyu) 
+desktop software tools enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
+Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
+Moreover, with rare exceptions \cite{MacWhinney2001} there are few standards for annotations themselves, even across studies performed by the same researcher \cite{?}. %Adolph, personal communication?
 
@@
 
-Databrary's approach to data is to provide as much context around the primary recordings as possible while still providing contributors some flexibility.
-To keep datasets intact, Databrary can incorporate heterogeneous datasets within our target domain by allowing secondary, general-purpose flat-file data, materials, and references to be associated with the primary data.
-This facilitates storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
-At the same time, it provides a common, standard structure for recordings and their associated metadata, including context, contents, and annotations.
+Our approach has been to provide as much context as possible around the primary recordings  while still providing contributors flexibility that they have come to expect from their own, possibly \textit{ad hoc}, archival systems.
+To keep datasets intact, Databrary incorporates allows heterogeneous flat-file data, materials, and references to be associated with the primary video or audio recordings.
+We provide a standard structure for recordings and their associated metadata, including context, contents, and annotations.
+This facilitates the storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
 We achieve this conformance for previously collected datasets through a combination of in-depth investigative curation and automated, server-side ingest processes.
-For datasets that are in the process of being collected, a well-defined user experience provides a platform for researchers to contribute and organize all of their research data as they proceed with their study.
-The following section provides specific detail on the design and decision-making behind this unifying structure.
+For datasets that are in the process of being collected, a well-defined user experience provides a platform for researchers to actively curate their own research data as they acquire it.
+The following section provides specific details about the design and decision-making behind this unifying structure.
 

### 2015-01-29 jcdl-15/paper.tex
through records...

@@
 \begin{document}
-\conferenceinfo{JCDL}{'15 Knoxville, Tennesse USA}
+\conferenceinfo{JCDL}{'15 Knoxville, Tennessee USA}
 
@@
 \begin{abstract}
+
+Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences. 
+Video is dense in both time and space.
+It uniquely captures the complexity of behavior, and when mined by human coders or computer algorithms, video provides big streams of data that can yield novel scientific insights \cite{raudies_visual_2014}.
+Thus, video presents substantial opportunities for data reuse if challenges relating to tagging/annotation, search and discovery, and curation can be overcome.
+The Databrary data library represents one approach to meeting these challenges within a particular scholarly domain, the study of human and animal development.
+
 %%dg being that the conference them is big data, should we be trying to be more explicit about that?
@@
 
-Our approach has been to provide as much context as possible around the primary recordings  while still providing contributors flexibility that they have come to expect from their own, possibly \textit{ad hoc}, archival systems.
-To keep datasets intact, Databrary incorporates allows heterogeneous flat-file data, materials, and references to be associated with the primary video or audio recordings.
+Our approach has been to provide consistent structure related to the storage and presentation of primary recordings while providing contributors the flexibility that they have come to expect from their own, possibly \emph{ad hoc}, archival systems.
+To keep datasets intact, Databrary associates heterogeneous flat-file data, materials, and references with the primary video or audio recordings.
 We provide a standard structure for recordings and their associated metadata, including context, contents, and annotations.
@@
 
-At the top level of organization, we separate individual data contributions into collections or packages variously referred to as datasets, studies, or volumes.
+At the top level of organization, we separate individual data contributions into collections or packages called \emph{volumes}.
+Most researchers in the target domain organize their research activity around the collection and analysis of specific datasets; scholarly products like papers or presentations draw upon one or more datasets.
+The volume enables Databrary to reflect the many-to-many structural relationship between datasets and studies in a flexible and powerful way.
 Volumes can be associated with a minimal amount of metadata (title, description, permissions, etc.) as well as additional files (documents and other materials).
@@
 %%dg can we expand on the above sentence, what does this mean and what are the implications?
-The remainder of this work will address the available structures for data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
+%The remainder of this work will address the available structures for data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
 
-There are two key, innovative structural elements that comprise a volume: the \emph{session} and the \emph{record}.
-Sessions are the primary container for data files, an elaboration on traditional file folders with additional temporal metadata.
-Records are used to represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as to provide an organizational structure among sessions.
+Two structural elements comprise a volume: the \emph{session} and the \emph{record}.
+Sessions provide the primary container for data files, an elaboration on traditional file folders with additional temporal metadata. 
+Records represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as provide an organizational structure among sessions.
 The combination of these two elements allows data and metadata to be flexibly related and annotated.
@@
 
-The session in Databrary draws from accepted and documented practices around how researchers record activities with individuals.
-Recordings are made either for the purpose of a later, closer review or analysis, or simply to document and verify measurements and observations made during the collection \cite{Bakeman_2012}.
-Thus, the recording itself is often the core item of research, representing a single data collection, experiment, trial, visit, or case depending on the nature of the research.
-However, in many cases a single session can comprise multiple videos: from serial data streams due to recording equipment producing multiple files, and/or in parallel from multiple data capture sources operating at the same time.
+The session in Databrary draws upon documented practices in the ways researchers use video. 
+The session was designed to organize recordings and related data streams in a way familiar to researchers, keeping data in its rawest and most flexible form without requiring  modifications that reduce the precision or utility of the data to others.
+
+Recordings are often the core component of data collection.
+Recordings provide raw materials for subsequent detailed analysis or documentation about what measurements were made during the collection \cite{Bakeman_2012}.
+In some cases a single session can comprise multiple videos, either serially in time representing different segments of activity or in parallel from multiple cameras recording the same events.
 These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
@@
 For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
-Secondary data is also often overlaid or combined in this way, such as by adding an additional waveform or frequency visualization below a video, or by overlaying cross-hairs representing location or gaze data.
-Doing so allows for more convenient human analysis of the data in standard playback tools that only support a single data stream, but also loses some of the raw data, by obscuring, masking, or downsampling spatially or temporally.
-The session was thus designed to organize original recordings in a way that is familiar to researchers, keeping their data in the rawest and most flexible form without requiring these modifications.
-This additionally allows for greater options in reuse by other researchers.
+Researchers may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing location or gaze data.
+Doing so allows for more convenient human analysis of the data in standard playback tools that support single data streams, but this spatial or temporal downsampling can obscure or mask some data patterns.
 
-A session object captures these relationships to represent a single data acquisition, comprising not only arbitrary files but also their absolute or relative temporal positioning.
-The metaphor we use here is a timeline, wherein each session defines a time range, with absolute timestamps representing the exact time of data collection, or relative positioning defined by a 0 starting point of the collection (or potentially any other meaningful wall-time units) with an optional date.
+A session container captures the relationships among data elements from a single data acquisition, capturing heterogeneous file types \emph{and} their absolute or relative temporal positioning.
+We use a timeline metaphor: Each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
 % good place for a screenshot?
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
-Although some advanced recording equipment provides automatic timestamping or synchronization than can be used, most researchers collecting multiple parallel data streams use ``sync points'' such as an electronic trigger signal, flash, or tone to align recordings.
-These sync points can then be used to manually place recordings on the timeline appropriately.
+Although some advanced recording equipment provides automatic timestamping or synchronization features, most researchers who collect parallel data streams use ``sync points'' such as an electronic trigger signal, flash, or tone to align recordings.
+These sync points can then be used to manually place recordings on the timeline.
+In this way, the session container represents the structure of a data collection session but retains significant flexibility.
 
@@
 
-An object called a record serves a number of purposes around session and metadata organization.
-A record itself represents any number of variables or measurements: what would traditionally be a row of a database or flat file.
-Records can thus represent most any item of analysis or organization typically represented in this form.
-For example, a record may represent an individual participant within the dataset and contain their subject identifier, gender, birth date, and any other demographic, survey, or measurement data about that individual.
-A record could also represent a particular experimental trial, task, condition, location, or outcome, and contain variables with corresponding measurements, parameters, descriptions.
-Records are thus labeled and divided by the type of entity they represent.
-The variables themselves are textual, numeric, date, or potentially any other scalar data as in any traditional tabular data interface, with the records of a given type then constituting the tabular rows.
+Record objects represent any number of variables, measurements, or items of analysis within a session: What would traditionally be a row of a database or flat file.
+A record may represent an individual participant and contain a subject identifier, gender, birth date, and demographic, survey, or measurement data about that person.
+It could also represent a particular experimental trial, task, condition, location, or outcome, and contain variables with corresponding measurements, parameters, or descriptions.
+Accordingly, records are labeled and organized by the type of entity they represent.
+The variables themselves are textual, numeric, date, or any scalar data form.
 
-Records are also used to organize sessions.
-We observed that different researchers organize their data files in different ways according to the type of research.
+Records help to organize sessions.
+We have observed that different researchers organize data files in different ways.
 Many consider participants to be the primary item of analysis, and so create a filesystem directory for each participant.
-Others organize their data based on date, experimental condition, phase of analysis, or age groups.
-Sometimes even single video files are split up into segments according to scene or activity, and researchers consider this property the primary organizational principal.
+Others organize their data based on date, experimental condition, phase of analysis, or age group.
+Sometimes even single video files are split up into segments according to scene or activity, and researchers consider this property the primary organizational principle.
 
-Rather than trying to handle all of these strategies separately, we instead leverage records.
-That is, records can be attached to sessions, or, equivalently, sessions can be grouped into records, in a many-to-many relationship.
-For example, in a traditional psychology study, each participant may appear in exactly one session, so the collected information about that participant, added to the record, is attached one-to-one with their session.
-A dataset may also be composed of a few different sub-sets based on experimental procedures or location, and here each session would be placed into exactly one of these few records.
-In some research, multiple participants may also appear in multiple sessions, such as recordings of the same class of students over a span of time.
+Rather than trying to handle all of these strategies separately, Databrary leverages the power of records.
+Records can be attached to sessions, or, sessions can be grouped into records, in a many-to-many relationship.
+For example, in one type of lab-based psychology study, a participant may appear in exactly one session, so the collected information about that participant, added to a record, is attached in a one-to-one relationship with a specific session.
+In other research, multiple participants may appear in multiple sessions, such as recordings of the same class of students over a span of time.
+A dataset may consist of distinct sub-sets of participants based on differences in experimental procedures or locations; here each session would be placed into a record representing the procedure or location.
 
-Since researchers can create whichever record types apply to their data, they can then choose to group their data by whichever record type they wish.
-This also allows users to either import or export data in what they consider the native directory structure by indicating the appropriate nesting of record types, along with the record variables themselves in traditional CSV formats.
-By providing some standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants within a particular age range.
+Since researchers can create whichever record types apply to their data, they can also choose to group their data as they wish.
+This allows users to import data in what they consider the native directory structure by indicating the appropriate nesting of record types along with record variables themselves in traditional CSV formats.
+The same flexbility applies to data export.
+By providing standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants within a particular age range.
 

### 2015-01-29 jcdl-15/paper.tex
clear out some notes, edit some text

@@
 
-%%dg being that the conference them is big data, should we be trying to be more explicit about that?
-% big data is generally referring to "more rows" rather than larger files. I think the closest we get is the size of video files
-% which is already noted below. Just wondering if it makes sense to be more explicit or not.
-%%dg also preservation and discoverability - though maybe that's included in data management
 %%rg video is big/dense in time and space.
@@
 Volumes can be associated with a minimal amount of metadata (title, description, permissions, etc.) as well as additional files (documents and other materials).
-These volumes are to some extent embeddable and referenceable, both among themselves and to and from external entities by DOI or URI.
-%%dg can we expand on the above sentence, what does this mean and what are the implications?
-%The remainder of this work will address the available structures for data within these volumes, highlighting how these structures allow commonalities between volumes to be represented.
+These volumes are citeable in that they are persistently identified (with a URI) and can cite external resources using a Digital Object Identifier (DOI) and an Open Researcher and Contributor ID (ORCID).
 
@@
 We use a timeline metaphor: Each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
-% good place for a screenshot?
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
@@
 Where research of community tagging provides evidence for the case that not many users actually do participate in tagging, we intend to leverage automated tagging through already existing metadata and analysis to seed a base of tags for reuse by the community \cite{Yang_Lu_Giles_2011, Farooq_etal_2007}.
-%%dg - might move the last sentence around, but provides a gist.
 

### 2015-01-29 jcdl-15/paper.tex
removing more notes

@@
 Video data collected in one context can often be used by others for different purposes with minimal explanation.
-This naturally reduces the problems of documentation and re-use that plague many flat-file datasets. % wanted to include raw, opaque data here, too
+This naturally reduces the problems of documentation and re-use that plague many flat-file datasets or repositories for data that require propietary software to view.
 However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
@@
 Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
-Specialized %commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (Transana, Elan, CLAN, and Datavyu) 
-desktop software tools enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
+Specialized desktop software tools --- both commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (e.g.
+Transana, Elan, CLAN, and Datavyu) --- enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
 Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
@@
 \section{Data Architecture}
-
-%%dg would a figure illustration be helpful here? do we already have one?
 
@@
 
-% this is kinda weird, but maybe suitable for a last paragraph?
 Our single-minded focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers gives us a novel perspective on building domain-specific research repositories.

### 2015-01-29 jcdl-15/paper.tex
typo

@@
 Video data collected in one context can often be used by others for different purposes with minimal explanation.
-This naturally reduces the problems of documentation and re-use that plague many flat-file datasets or repositories for data that require propietary software to view.
+This naturally reduces the problems of documentation and re-use that plague many flat-file datasets or repositories for data that require proprietary software to view.
 However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.

### 2015-01-29 jcdl-15/paper.tex
Update abstract; last sentence needs some tweaking

@@
 \begin{abstract}
-
-Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences. 
-Video is dense in both time and space.
-It uniquely captures the complexity of behavior, and when mined by human coders or computer algorithms, video provides big streams of data that can yield novel scientific insights \cite{raudies_visual_2014}.
-Thus, video presents substantial opportunities for data reuse if challenges relating to tagging/annotation, search and discovery, and curation can be overcome.
-The Databrary data library represents one approach to meeting these challenges within a particular scholarly domain, the study of human and animal development.
-
-%%rg video is big/dense in time and space.
-But, much of the size is masked without intensive analysis by humans or machines.
-Compare it with brain imaging data.
-A typical brain imaging study might include 3-5 runs of 200 brain volumes with 30 128x128 pixel x 8 bit greyscale slices per volume, with each volume collected about every 2-3 sec (.33-.5 Hz).
-Video is much "bigger" because it samples space and time densely.
+Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences, and present unique opportunities for reuse and reanalysis for novel scientific purposes.
+However, there are numerous technical and policy challenges preventing existing recordings from being accessed and reused by other researchers.
+By investigating how researchers organize, analyze, and mine their own recording data, we have implemented a number of structural generalizations to capture these data and their context allowing standardization, sharing, and understanding.
+There are remaining challenges around discovery and annotation, and we discuss some options and initial approaches.
+While many of Databrary's strengths have derived from its focus on a particular scholarly domain, many of these issues have broader relevance to the digital library community.
 \end{abstract}

### 2015-01-29 jcdl-15/paper.tex
more tweaks, but not yet to annotations

@@
+% rog We need to cut a half page or more...
+
 \documentclass{sig-alternate}
@@
 \begin{abstract}
-Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences, and present unique opportunities for reuse and reanalysis for novel scientific purposes.
+Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences.
+Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes.
 However, there are numerous technical and policy challenges preventing existing recordings from being accessed and reused by other researchers.
-By investigating how researchers organize, analyze, and mine their own recording data, we have implemented a number of structural generalizations to capture these data and their context allowing standardization, sharing, and understanding.
+By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share these data in a standardized way.
 There are remaining challenges around discovery and annotation, and we discuss some options and initial approaches.
@@
 Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
-Moreover, with rare exceptions \cite{MacWhinney2001} there are few standards for annotations themselves, even across studies performed by the same researcher \cite{?}. %Adolph, personal communication?
+Moreover, with rare exceptions \cite{MacWhinney2001} there are few standards for annotations themselves, even across studies performed by the same researcher. %Adolph, personal communication?
 
@@
 
-The session in Databrary draws upon documented practices in the ways researchers use video. 
-The session was designed to organize recordings and related data streams in a way familiar to researchers, keeping data in its rawest and most flexible form without requiring  modifications that reduce the precision or utility of the data to others.
+The session draws upon documented practices in the ways researchers use video. 
+It was designed to organize recordings and related data streams in a way familiar to researchers, keeping data in its rawest and most flexible form without requiring  modifications that reduce the precision or utility of the data to others.
 
@@
 For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
-Researchers may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing location or gaze data.
+Some may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing location or gaze data.
 Doing so allows for more convenient human analysis of the data in standard playback tools that support single data streams, but this spatial or temporal downsampling can obscure or mask some data patterns.
@@
 A session container captures the relationships among data elements from a single data acquisition, capturing heterogeneous file types \emph{and} their absolute or relative temporal positioning.
-We use a timeline metaphor: Each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
+We use a timeline metaphor for storage and visualization: Each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
@@
 Records help to organize sessions.
-We have observed that different researchers organize data files in different ways.
-Many consider participants to be the primary item of analysis, and so create a filesystem directory for each participant.
+We have observed that researchers organize data files in substantially different ways.
+Many consider participants to be the primary focus of analysis, and so create a filesystem directory for each participant.
 Others organize their data based on date, experimental condition, phase of analysis, or age group.
@@
 
+% rog this section needs work.
+
 Establishing a timeline for each session allows temporally-positioned annotations to be attached to the session.
-Currently we support two types of annotations: structured, organizational annotations based on records and added by the data contributors; and simple annotations contributed by the community.
+Currently we support two types of annotations: structured annotations based on records and added by the data contributors; and simple annotations contributed by the community.
 Eventually we hope to extend both of these to support the full range of annotations performed during analysis, both directly on the web and imported from existing software.
@@
 All uploaded recordings are automatically transcoded to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
-This transcoding utilizes the high performance computing cluster at the hosting campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
+This transcoding utilizes the high performance computing cluster at the host campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
 The user interface is built primarily on the Angular web framework\footnote{http://angularjs.org/}, and all data access is performed through an open JSON API.
@@
 
-The Databrary website was opened for general use in October 2014 and has accepted contributions since earlier that year.
-It currently hosts 5700 video files totaling 1600 hours of recordings along with 2200 additional files.
-These files comprise 2400 sessions and are covered by 1300 records (including 1100 individual participants).
+The Databrary website accepted contributions beginning in early 2014 and opened for general use in October 2014.
+It currently hosts 5,700 video files totaling 1,600 hours of recordings along with 2,200 additional files.
+These files comprise 2,400 sessions and are covered by 1,300 records (including 1100 individual participants).
 Data originates from 35 individual contributors across 25 different universities.
@@
 
-As the extant research data libraries demonstrate, there is always the tension between normalization and flexibility.
 Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
@@
 Here we have also introduced a number of metadata concepts where this tension becomes most apparent, namely the available record types, variables (and values within categorical data types), and tag keywords.
-We have adopted an approach of encouraging adherence to existing standardized options but also allowing new types to be created as the data require it.
+We have adopted an approach that encourages adherence to existing standardized options but  allows new types to be created as the data require it.
 Initially we have achieved this through manual review and off-line curation of contributions, dynamically extending the system as new types of data were discovered.
-For many previously collected datasets, this has proven the most effective method to both learn from data and avoid burdening contributors.
+For many previously collected datasets, this has proven the most effective method to learn from data and avoid burdening contributors.
 However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a distinct, user-driven approach.
 
-We found that while some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools, often using a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and available filesystem storage (often just external hard-drives).
+We found that while some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools, often using a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and available filesystem storage.
 This suggested the opportunity to supply a cohesive data management platform where all collected raw data, including recordings, contextual information, and measurements, could be entered, stored, and exported for later off-line analysis.
@@
 
-Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve a good amount of standardization across the entire data library without overly burdening contributors.
+Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
 
@@
 
-Starting from the perspective of a general-purpose repository, we have introduced a number of significant restrictions to create a specialized, domain repository for specific types of data.
-By limiting both file and metadata organization to a flat (but overlapping) set of sessions and records, rather than supporting the full hierarchical richness of other systems, we can present and capture session-based research data with a cohesive and intuitive user interface.
+Starting from the perspective of a general-purpose repository, we have introduced a number of  restrictions to create one specialized for specific types of data.
 By focusing on video and audio recordings we can normalize presentation and discovery of data across the site, and also introduce temporal structure to data and annotations driven by the nature of these recordings.
-Although we allow a few types of additional data files, we encourage users with other types of rich data, such as physiological recordings, to store and link to those resources externally, ideally in similarly focused and better-suited repositories.
-Similarly, by targeting use cases around finding new uses for existing research data, rather than replication or meta-analysis, we can limit the scope of accepted data to the original recordings and early analysis phases that generate the most value, rather than trying to capture the entire provenance.
+By limiting both file and metadata organization to a flat (but overlapping) set of sessions and records, we capture session-based research data within a cohesive and intuitive user interface.
+Although we allow additional data files, we encourage users with other types of rich data, such as physiological recordings, to store and link to those resources externally.
+Similarly, by targeting reuse of existing research data, rather than replication or meta-analysis, we limit the scope of accepted data to the original recordings and early analysis phases that generate the most value.
 
@@
 
-This work was supported by the National Science Foundation (BCS-1238599) and the National Institute of Child Health and Human Development (U01-HD-076595-01).
+This work was supported by the NSF (BCS-1238599) and the NICHD (U01-HD-076595-01).
 The authors gratefully acknowledge the NYU Libraries for their valuable advice and consultation.

### 2015-01-29 jcdl-15/paper.tex
Minor editing intro, reflow

@@
 
-Video and audio recordings serve as a primary data source in psychology, linguistics, education, anthropology, and numerous other fields, especially the study of human and animal development.
-Unfortunately, researchers who collect video rarely share recordings with others, due largely to perceived requirements associated with privacy restrictions.
-The goal of the Databrary project is to increase the repurposing and reuse of recordings, especially video, by creating tools that allow video to be documented, organized, and shared among researchers.
+Video and audio recordings serve as a primary data source in psychology, linguistics, education, anthropology, and numerous other fields, and especially in the study of human development.
+Unfortunately, researchers who collect recordings rarely share with others, due largely to perceived requirements associated with privacy restrictions.
+The goal of the Databrary project is build tools and policies that simplify repurposing and reuse of recordings.
 We have created a web-based data library to accommodate the storage and organization of video data, capture contextual information necessary for reuse, and standardize across datasets without placing undue burden on data contributors.
@@
 Research data repositories require a balance between generality and specificity.
-Archives that permit the unstructured storage of arbitrary files lie at one extreme;
-those that present a rigorously structured schema and required contribution formats lie at the other.
+Archives that permit the unstructured storage of arbitrary files lie at one extreme; those that present a rigorously structured schema and required contribution formats lie at the other.
 The flexibility of general-purpose repositories make them especially well-suited to the archival of entire datasets, but these systems do not facilitate search, understanding, or re-use without extensive, usually labor-intensive, curation \cite{Peer_2012}.
-Targeted repositories, on the other hand, generate large-scale, homogeneous data sources that facilitate meta- and re-analysis, but serve only specific domains. 
+Targeted repositories, on the other hand, generate large-scale, homogeneous data sources that facilitate meta- and re-analysis, but are limited to specific research questions, losing the full richness, context, and innovation of the source materials and field.
 We have attempted to strike a balance between these extremes by making video and audio recordings the unifying focus of Databrary.
@@
 Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
-Specialized desktop software tools --- both commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (e.g.
-Transana, Elan, CLAN, and Datavyu) --- enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
+Specialized desktop software tools---both commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (e.g., Transana, Elan, CLAN, and Datavyu)---enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
 Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.

### 2015-01-30 jcdl-15/paper.tex
updates to records and structure for intro
notes need to be fleshed out into full paragraphs

@@
 There are remaining challenges around discovery and annotation, and we discuss some options and initial approaches.
-While many of Databrary's strengths have derived from its focus on a particular scholarly domain, many of these issues have broader relevance to the digital library community.
+While many of Databrary's strengths have derived from its focus on particular scholarly domains, many of these issues have broader relevance to the digital library community.
 \end{abstract}
@@
 \section{Introduction}
+%video data are special because they are reusable
+Video and audio recordings constitute a form of human-consumable, self-documenting, timeseries data.
+Recordings can be and largely are consumed and analyzed, at least initially, without the use of sophisticated tools: once digitally decoded, trained human observers can directly glean much of the data's richness.
+Video data collected in one context can often be used by others for different purposes with minimal explanation.
+This naturally reduces the problems of documentation and re-use that plague many flat-file datasets or repositories for data that require proprietary software to view.
+However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
 
+%video data are not reused, databrary aims to provide solutions that allow for reuse.
 Video and audio recordings serve as a primary data source in psychology, linguistics, education, anthropology, and numerous other fields, and especially in the study of human development.
 Unfortunately, researchers who collect recordings rarely share with others, due largely to perceived requirements associated with privacy restrictions.
-The goal of the Databrary project is build tools and policies that simplify repurposing and reuse of recordings.
-We have created a web-based data library to accommodate the storage and organization of video data, capture contextual information necessary for reuse, and standardize across datasets without placing undue burden on data contributors.
+%say this better: This inhibits reuse. 
+To address this problem, the Databrary project aims to build tools and policies that allow repurposing and reuse of recordings.
+We have created a web-based data library to accommodate the storage and organization of video data, capture contextual information necessary for reuse, and create standardization across datasets that enable discovery without placing undue burden on data contributors.
 In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
@@
 Archives that permit the unstructured storage of arbitrary files lie at one extreme; those that present a rigorously structured schema and required contribution formats lie at the other.
-The flexibility of general-purpose repositories make them especially well-suited to the archival of entire datasets, but these systems do not facilitate search, understanding, or re-use without extensive, usually labor-intensive, curation \cite{Peer_2012}.
+The flexibility of general-purpose repositories make them especially well-suited to the archival of entire datasets, but these systems do not facilitate search, understanding, or re-use without extensive, usually labor-intensive, curation to summarize or extract \cite{Peer_2012}.
 Targeted repositories, on the other hand, generate large-scale, homogeneous data sources that facilitate meta- and re-analysis, but are limited to specific research questions, losing the full richness, context, and innovation of the source materials and field.
+
+\subsection{Approach}
+%How databrary uniquely can deal with this problem because of its focus on video: by focusing on specific domain and focusing primarily on video, and inferring structure from actual data, databary has figured out how to standardize enough of the contextual metadata to facilitate understanding/enable re-use.
+We provide a standard structure for recordings and their associated metadata, including context, contents, and annotations.
+This facilitates the storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
+%our novel contribution to this is the identification of generalizable set of principles around which to structure this contextual metadata, which enables re-use of recordings at an unprecendented scale.
+%annotating data is unique advantage of video, hope to leverage add'l data generated during analysis (i.e. annotations) to solve remaining problem is search.
+
 We have attempted to strike a balance between these extremes by making video and audio recordings the unifying focus of Databrary.
 This allows additional structured and unstructured resources to be built around and discovered through this functional core.
-
-\subsection{Video-driven research}
-
-Video and audio recordings constitute a form of human-consumable, self-documenting, timeseries data.
-Recordings can be and largely are consumed and analyzed, at least initially, without the use of sophisticated tools: once digitally decoded, trained human observers can directly glean much of the data's richness.
-Video data collected in one context can often be used by others for different purposes with minimal explanation.
-This naturally reduces the problems of documentation and re-use that plague many flat-file datasets or repositories for data that require proprietary software to view.
-However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013}.
-In short, recordings pose challenges that older, less dynamic media do not \cite{Lanagan_Smeaton_2012}.
-
-Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
-Specialized desktop software tools---both commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (e.g., Transana, Elan, CLAN, and Datavyu)---enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
-Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
-Moreover, with rare exceptions \cite{MacWhinney2001} there are few standards for annotations themselves, even across studies performed by the same researcher. %Adolph, personal communication?
-
-\subsection{Approach}
-
 Our approach has been to provide consistent structure related to the storage and presentation of primary recordings while providing contributors the flexibility that they have come to expect from their own, possibly \emph{ad hoc}, archival systems.
 To keep datasets intact, Databrary associates heterogeneous flat-file data, materials, and references with the primary video or audio recordings.
-We provide a standard structure for recordings and their associated metadata, including context, contents, and annotations.
-This facilitates the storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
+
 We achieve this conformance for previously collected datasets through a combination of in-depth investigative curation and automated, server-side ingest processes.
@@
 At the top level of organization, we separate individual data contributions into collections or packages called \emph{volumes}.
-Most researchers in the target domain organize their research activity around the collection and analysis of specific datasets; scholarly products like papers or presentations draw upon one or more datasets.
+Most researchers in the target domains organize their research activity around the collection and analysis of specific datasets; scholarly products like papers or presentations draw upon one or more datasets.
 The volume enables Databrary to reflect the many-to-many structural relationship between datasets and studies in a flexible and powerful way.
@@
 \subsection{Records}
+A record can represent many things that are typically conceived of as a row of a database.
+Each may represent a single data sample, such as a participant, dyad, classroom, a level of analysis, such as an experimental trial, task, condition, or location, or some other factor. 
+A record is thus defined by one of the above types along with any number of scalar values, which can be textual, numeric, or date.
+In the case of a record representing an individual participant, these values might include subject identifier, gender, birth date, demographic information, measurement data, or other collected variables.
+For other record types, these values may include relevant independent variables, parameters, or descriptions. 
+We can thus display the records of a given type as a table, with the values of each record constituting a row.
 
-Record objects represent any number of variables, measurements, or items of analysis within a session: What would traditionally be a row of a database or flat file.
-A record may represent an individual participant and contain a subject identifier, gender, birth date, and demographic, survey, or measurement data about that person.
-It could also represent a particular experimental trial, task, condition, location, or outcome, and contain variables with corresponding measurements, parameters, or descriptions.
-Accordingly, records are labeled and organized by the type of entity they represent.
-The variables themselves are textual, numeric, date, or any scalar data form.
-
-Records help to organize sessions.
-We have observed that researchers organize data files in substantially different ways.
-Many consider participants to be the primary focus of analysis, and so create a filesystem directory for each participant.
-Others organize their data based on date, experimental condition, phase of analysis, or age group.
+In addition to containing values, records are also used to organize sessions.
+We have found that different researchers organize data files in substantially different ways.
+Researchers differentially consider their primary organizational unit, often depending on the primary focus of analysis for their specific study or sub-domain.
+They may organize their data by participants, session date, experimental condition, phase of analysis, or age group.
 Sometimes even single video files are split up into segments according to scene or activity, and researchers consider this property the primary organizational principle.
@@
 Records can be attached to sessions, or, sessions can be grouped into records, in a many-to-many relationship.
-For example, in one type of lab-based psychology study, a participant may appear in exactly one session, so the collected information about that participant, added to a record, is attached in a one-to-one relationship with a specific session.
+For example, a study participant may appear in exactly one session, so the collected information about that participant, added to a record, is attached in a one-to-one relationship with a specific session.
 In other research, multiple participants may appear in multiple sessions, such as recordings of the same class of students over a span of time.
-A dataset may consist of distinct sub-sets of participants based on differences in experimental procedures or locations; here each session would be placed into a record representing the procedure or location.
+% for brevity, can we either exclude or condense this sentence?: A dataset may consist of distinct sub-sets of participants based on differences in experimental procedures or locations; here each session would be placed into a record representing the procedure or location.
 
@@
 % rog this section needs work.
+
+Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
+Specialized desktop software tools---both commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (e.g., Transana, Elan, CLAN, and Datavyu)---enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
+Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
+Moreover, with rare exceptions \cite{MacWhinney2001} there are few standards for annotations themselves, even across studies performed by the same researcher. 
+%moved ^ from intro
 
@@
 
-The Databrary website accepted contributions beginning in early 2014 and opened for general use in October 2014.
+Databrary accepted contributions beginning in early 2014 and opened for general use in October 2014.
 It currently hosts 5,700 video files totaling 1,600 hours of recordings along with 2,200 additional files.
@@
 
-Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
-We also allow allow users, using certain portable formats (flat-files, documents, images), to include any data that cannot be encompassed by our standard formats.
+%Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
+We allow users, using certain portable formats (flat-files, documents, images), to include any data that cannot be encompassed by our standard formats.
 In this way, we can incorporate data which may be relevant for re-use but not broadly adopted by the community, without making the system too rigid.
@@
 Initially we have achieved this through manual review and off-line curation of contributions, dynamically extending the system as new types of data were discovered.
-For many previously collected datasets, this has proven the most effective method to learn from data and avoid burdening contributors.
+For many previously collected datasets, this has proven the most effective method to infer from data and avoid burdening contributors.
 However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a distinct, user-driven approach.
@@
 
-Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
+Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to
+
+particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
 

### 2015-01-30 jcdl-15/paper.tex
reviewed paper, changed some over specific language and edits
* need to focus annotation section more
* can probably collapse Normalization and Discovery sections down

@@
 \section{Introduction}
-%video data are special because they are reusable
+
 Video and audio recordings constitute a form of human-consumable, self-documenting, timeseries data.
@@
 
-%video data are not reused, databrary aims to provide solutions that allow for reuse.
 Video and audio recordings serve as a primary data source in psychology, linguistics, education, anthropology, and numerous other fields, and especially in the study of human development.
 Unfortunately, researchers who collect recordings rarely share with others, due largely to perceived requirements associated with privacy restrictions.
-%say this better: This inhibits reuse. 
-To address this problem, the Databrary project aims to build tools and policies that allow repurposing and reuse of recordings.
+Despite the inherent reusability of video data, difficulties in sharing them limit this scenario from happening more often.
+
+To address this problem, the Databrary project aims to build tools and policies that facilitate repurposing and reuse of recordings.
 We have created a web-based data library to accommodate the storage and organization of video data, capture contextual information necessary for reuse, and create standardization across datasets that enable discovery without placing undue burden on data contributors.
@@
 Research data repositories require a balance between generality and specificity.
-Archives that permit the unstructured storage of arbitrary files lie at one extreme; those that present a rigorously structured schema and required contribution formats lie at the other.
+General purpose repositories that permit the unstructured storage of arbitrary files lie at one extreme \cite{dataverse, dryad, purr}; those that present a rigorously structured schema and required contribution formats lie at the other \cite{HCP, PGP, talkbank}.
 The flexibility of general-purpose repositories make them especially well-suited to the archival of entire datasets, but these systems do not facilitate search, understanding, or re-use without extensive, usually labor-intensive, curation to summarize or extract \cite{Peer_2012}.
@@
 This allows additional structured and unstructured resources to be built around and discovered through this functional core.
-Our approach has been to provide consistent structure related to the storage and presentation of primary recordings while providing contributors the flexibility that they have come to expect from their own, possibly \emph{ad hoc}, archival systems.
+Our approach has been to provide a basic scaffold related to the storage and presentation of primary recordings while providing contributors the flexibility that they have come to expect from their own, possibly \emph{ad hoc}, archival systems.
 To keep datasets intact, Databrary associates heterogeneous flat-file data, materials, and references with the primary video or audio recordings.
@@
 Recordings are often the core component of data collection.
-Recordings provide raw materials for subsequent detailed analysis or documentation about what measurements were made during the collection \cite{Bakeman_2012}.
+They provide raw materials for subsequent detailed analysis or documentation about what measurements were made during the collection \cite{Bakeman_2012}.
 In some cases a single session can comprise multiple videos, either serially in time representing different segments of activity or in parallel from multiple cameras recording the same events.
@@
 For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
-Some may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing location or gaze data.
+Some may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing location or gaze data (that is, where a subject is looking during data collection).
 Doing so allows for more convenient human analysis of the data in standard playback tools that support single data streams, but this spatial or temporal downsampling can obscure or mask some data patterns.
@@
 A record can represent many things that are typically conceived of as a row of a database.
-Each may represent a single data sample, such as a participant, dyad, classroom, a level of analysis, such as an experimental trial, task, condition, or location, or some other factor. 
+Each may represent a single data sample, such as a participant, participant pairs interacting, classroom, a level of analysis, such as an experimental trial, task, condition, or location, or some other factor.
 A record is thus defined by one of the above types along with any number of scalar values, which can be textual, numeric, or date.
@@
 Where research of community tagging provides evidence for the case that not many users actually do participate in tagging, we intend to leverage automated tagging through already existing metadata and analysis to seed a base of tags for reuse by the community \cite{Yang_Lu_Giles_2011, Farooq_etal_2007}.
+%%dg - reflagging this last sentence, it's valuable, but seems to do a major belly flop right here.
 
@@
 
+%%dg - voting to put this and the repository section up higher in the paper, Approach, maybe?
 Databrary is a new, cohesive web application, built in Scala on the Play Framework\footnote{http://playframework.com/} to enable a responsive user interface, a complete API, and high-performance streaming.
@@
 \subsection{Normalization}
+
+%%dg - Reading this again, this section seems a little redudant with what is discussed above with making records and sessions flexible?
+Might be able to sum up this section and the discovery section in a single 'Lessons learned and Upcoming Challenges' section
 

### 2015-01-30 jcdl-15/paper.tex
mostly work on intro and other small changes.
annotations section of intro may be very out of place

@@
 Video and audio recordings constitute a form of human-consumable, self-documenting, timeseries data. 
-Recordings can be and largely are consumed and analyzed, at least initially, without the use of sophisticated tools: once digitally decoded, trained human observers can directly glean much of the data's richness.
-Video data collected in one context can often be used by others for different purposes with minimal explanation.
-This naturally reduces the problems of documentation and re-use that plague many flat-file datasets or repositories for data that require proprietary software to view.
-However, since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
-
-Video and audio recordings serve as a primary data source in psychology, linguistics, education, anthropology, and numerous other fields, and especially in the study of human development.
-Unfortunately, researchers who collect recordings rarely share with others, due largely to perceived requirements associated with privacy restrictions.
-Despite the inherent reusability of video data, difficulties in sharing them limit this scenario from happening more often.
-
+Moreover, recordings are a primary source of raw data for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
+Recordings collected in one context can often be used by others for different purposes with minimal explanation. 
+This, therefore, naturally reduces the problems of documentation and provenance that plague many flat-file datasets and data repositories, especially those containing data that requires proprietary software to access.
+Yet researchers rarely share recordings with others, due largely to entrenched practices and perceived requirements associated with privacy restrictions.
+%something here or somewhere else about aversion (fear) to changing practices...
+Despite the inherent reusability of recordings, this lack of sharing inhibits the reuse of raw data and impedes scientific discovery.
 To address this problem, the Databrary project aims to build tools and policies that facilitate repurposing and reuse of recordings.
 We have created a web-based data library to accommodate the storage and organization of video data, capture contextual information necessary for reuse, and create standardization across datasets that enable discovery without placing undue burden on data contributors.
-In this paper, we discuss the design and technical challenges of the system and describe some findings and structural solutions to these challenges.
 
+Recordings can be and largely are consumed and analyzed, at least initially, without the use of sophisticated tools.
+Once recordings are digitally decoded, trained human observers can directly glean much of the data's richness.
+Since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
+%^either need to say more here about annotations or do it in approach. 
 \subsection{Balancing Generality and Specificity}
@@
 Records represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as provide an organizational structure among sessions.
-The combination of these two elements allows data and metadata to be flexibly related and annotated.
+The combination of these two elements allows data and metadata to be flexibly related and annotated. %another word besides annotated here?
 
@@
 The session draws upon documented practices in the ways researchers use video.
-It was designed to organize recordings and related data streams in a way familiar to researchers, keeping data in its rawest and most flexible form without requiring  modifications that reduce the precision or utility of the data to others.
+It was designed to organize recordings and related data streams in a way familiar to researchers, keeping data in its most raw and flexible form without requiring  modifications that reduce the precision or utility of the data to others.
 
@@
 
+%for brevity, maybe this paragraph can go. it's adding nuance but does it advance our point about contextual metadata, structure and reuse? yes, because reuse is harder with only composite. 
 When working with multiple primary recordings, researchers often combine these raw files into a single representation.
@@
 A session container captures the relationships among data elements from a single data acquisition, capturing heterogeneous file types \emph{and} their absolute or relative temporal positioning.
-We use a timeline metaphor for storage and visualization: Each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
+The metaphor we use here is a timeline: each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
-Although some advanced recording equipment provides automatic timestamping or synchronization features, most researchers who collect parallel data streams use ``sync points'' such as an electronic trigger signal, flash, or tone to align recordings.
+Although some advanced recording equipment provides automatic timestamping or synchronization features, most researchers who collect parallel data streams use ``sync points'' such as an electronic trigger signal, light flash, or tone to align recordings.
 These sync points can then be used to manually place recordings on the timeline.

### 2015-01-30 jcdl-15/paper.tex
Introduction pass

@@
-% rog We need to cut a half page or more...
-
 \documentclass{sig-alternate}
@@
 
-Video and audio recordings constitute a form of human-consumable, self-documenting, timeseries data. 
-Moreover, recordings are a primary source of raw data for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
-Recordings collected in one context can often be used by others for different purposes with minimal explanation. 
-This, therefore, naturally reduces the problems of documentation and provenance that plague many flat-file datasets and data repositories, especially those containing data that requires proprietary software to access.
-Yet researchers rarely share recordings with others, due largely to entrenched practices and perceived requirements associated with privacy restrictions.
+Video and audio recordings serve as a primary raw data source for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
+They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools as human observers can directly glean much of the data's richness.
+% access in the sense of "what format is it in":
+This allows recordings collected in one context to be used by others for different purposes with minimal explanation, naturally reducing the problems of documentation, provenance, and access that plague many research repositories.
+On the other hand, automated analysis of videos is an active but still immature research area, so the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
+
+Unfortunately, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
 %something here or somewhere else about aversion (fear) to changing practices...
-Despite the inherent reusability of recordings, this lack of sharing inhibits the reuse of raw data and impedes scientific discovery.
-To address this problem, the Databrary project aims to build tools and policies that facilitate repurposing and reuse of recordings.
-We have created a web-based data library to accommodate the storage and organization of video data, capture contextual information necessary for reuse, and create standardization across datasets that enable discovery without placing undue burden on data contributors.
-
-Recordings can be and largely are consumed and analyzed, at least initially, without the use of sophisticated tools.
-Once recordings are digitally decoded, trained human observers can directly glean much of the data's richness.
-Since automated analysis of videos is an active yet still immature research area, the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
-%^either need to say more here about annotations or do it in approach. 
-\subsection{Balancing Generality and Specificity}
+Despite their inherent reusability, this lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
+To address this problem, the Databrary project aims to build tools and policies that facilitate repurposing and reuse.
+We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways, in order to enable understanding and discovery without placing undue burden on data contributors.
 
 Research data repositories require a balance between generality and specificity.
-General purpose repositories that permit the unstructured storage of arbitrary files lie at one extreme \cite{dataverse, dryad, purr}; those that present a rigorously structured schema and required contribution formats lie at the other \cite{HCP, PGP, talkbank}.
-The flexibility of general-purpose repositories make them especially well-suited to the archival of entire datasets, but these systems do not facilitate search, understanding, or re-use without extensive, usually labor-intensive, curation to summarize or extract \cite{Peer_2012}.
+General purpose repositories permit unstructured storage of arbitrary files \cite{dataverse, dryad, purr}, while those with rigorously structured schema require contributions to be to carefully formatted \cite{HCP, PGP, talkbank}.
+The flexibility of general-purpose repositories make them especially well-suited to capture and archive entire datasets, but these systems do not facilitate search, understanding, or reuse without extensive, usually labor-intensive, curation to summarize or extract metadata \cite{Peer_2012}.
 Targeted repositories, on the other hand, generate large-scale, homogeneous data sources that facilitate meta- and re-analysis, but are limited to specific research questions, losing the full richness, context, and innovation of the source materials and field.
 
-\subsection{Approach}
-%How databrary uniquely can deal with this problem because of its focus on video: by focusing on specific domain and focusing primarily on video, and inferring structure from actual data, databary has figured out how to standardize enough of the contextual metadata to facilitate understanding/enable re-use.
-We provide a standard structure for recordings and their associated metadata, including context, contents, and annotations.
-This facilitates the storage and retrieval of data more closely mapped to the original context in which it was collected and analyzed.
-%our novel contribution to this is the identification of generalizable set of principles around which to structure this contextual metadata, which enables re-use of recordings at an unprecendented scale.
-%annotating data is unique advantage of video, hope to leverage add'l data generated during analysis (i.e. annotations) to solve remaining problem is search.
-
-We have attempted to strike a balance between these extremes by making video and audio recordings the unifying focus of Databrary.
-This allows additional structured and unstructured resources to be built around and discovered through this functional core.
-Our approach has been to provide a basic scaffold related to the storage and presentation of primary recordings while providing contributors the flexibility that they have come to expect from their own, possibly \emph{ad hoc}, archival systems.
-To keep datasets intact, Databrary associates heterogeneous flat-file data, materials, and references with the primary video or audio recordings.
-
-We achieve this conformance for previously collected datasets through a combination of in-depth investigative curation and automated, server-side ingest processes.
-For datasets that are in the process of being collected, a well-defined user experience provides a platform for researchers to actively curate their own research data as they acquire it. 
-The following section provides specific details about the design and decision-making behind this unifying structure.
+Our approach is to focus on video and audio recordings in particular research domains, and build appropriate structured and unstructured resources around this functional core.
+By investigating of how researchers studying human development and education organize their own data, we have identified a unifying set of principles around which to structure contextual metadata.
+This structure facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
+We achieve this conformance through in-depth, investigative curation of previously collected datasets, along with familiar user interfaces allowing researchers to structure their own research data as they acquire it.
+In the future, we hope to capture additional annotation data researchers already produce while analyzing recordings to further improve our discovery capabilities.
 
@@
 At the top level of organization, we separate individual data contributions into collections or packages called \emph{volumes}.
-Most researchers in the target domains organize their research activity around the collection and analysis of specific datasets; scholarly products like papers or presentations draw upon one or more datasets.
-The volume enables Databrary to reflect the many-to-many structural relationship between datasets and studies in a flexible and powerful way.
-Volumes can be associated with a minimal amount of metadata (title, description, permissions, etc.) as well as additional files (documents and other materials).
-These volumes are citeable in that they are persistently identified (with a URI) and can cite external resources using a Digital Object Identifier (DOI) and an Open Researcher and Contributor ID (ORCID).
+Volumes allow a minimal amount of metadata (title, description, permissions, etc.) as well as additional supporting files (documents and other materials).
+They are persistently identified and can cite external resources by URI or Digital Object Identifier (DOI).
 
-Two structural elements comprise a volume: the \emph{session} and the \emph{record}.
+Most importantly, a volume involves two key structural elements: the \emph{session} and the \emph{record}.
 Sessions provide the primary container for data files, an elaboration on traditional file folders with additional temporal metadata.
-Records represent all types of scalar variable-based data, traditionally contained in flat files or databases, as well as provide an organizational structure among sessions.
-The combination of these two elements allows data and metadata to be flexibly related and annotated. %another word besides annotated here?
+Records represent all types of scalar variable-based data, traditionally contained in flat files or databases, and also provide an organizational structure among sessions.
+The combination of these two elements allows data and metadata to be flexibly structured, related, and annotated.
 
@@
 
-The session draws upon documented practices in the ways researchers use video.
-It was designed to organize recordings and related data streams in a way familiar to researchers, keeping data in its most raw and flexible form without requiring  modifications that reduce the precision or utility of the data to others.
-
-Recordings are often the core component of data collection.
-They provide raw materials for subsequent detailed analysis or documentation about what measurements were made during the collection \cite{Bakeman_2012}.
+Recordings are often the core component of a data collection, and so sessions were designed to structure them in a way familiar to researchers.
+Sessions provide raw materials for subsequent detailed analysis or documentation about what measurements were made during the collection \cite{Bakeman_2012}.
 In some cases a single session can comprise multiple videos, either serially in time representing different segments of activity or in parallel from multiple cameras recording the same events.
@@
 A record can represent many things that are typically conceived of as a row of a database.
-Each may represent a single data sample, such as a participant, participant pairs interacting, classroom, a level of analysis, such as an experimental trial, task, condition, or location, or some other factor.
+Each may represent a single data sample, such as a participant, group, or classroom, a level of analysis, such as an experimental trial, task, condition, or location, or some other factor. 
 A record is thus defined by one of the above types along with any number of scalar values, which can be textual, numeric, or date.
@@
 We allow users, using certain portable formats (flat-files, documents, images), to include any data that cannot be encompassed by our standard formats.
-In this way, we can incorporate data which may be relevant for re-use but not broadly adopted by the community, without making the system too rigid.
+In this way, we can incorporate data which may be relevant for reuse but not broadly adopted by the community, without making the system too rigid.
 

### 2015-01-30 jcdl-15/paper.tex
A bit farther

@@
 
-Recordings are often the core component of a data collection, and so sessions were designed to structure them in a way familiar to researchers.
-Sessions provide raw materials for subsequent detailed analysis or documentation about what measurements were made during the collection \cite{Bakeman_2012}.
-In some cases a single session can comprise multiple videos, either serially in time representing different segments of activity or in parallel from multiple cameras recording the same events.
+Sessions provide a container for raw materials and measurements that were made during a data collection \cite{Bakeman_2012}.
+As recordings are often the core component of a data collection, a session in many cases is defined by a single recording.
+In some cases a single session can comprise multiple videos, either serially in time representing different segments or in parallel from multiple cameras recording the same events.
 These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
 
-%for brevity, maybe this paragraph can go. it's adding nuance but does it advance our point about contextual metadata, structure and reuse? yes, because reuse is harder with only composite. 
-When working with multiple primary recordings, researchers often combine these raw files into a single representation.
+When working with multiple files, researchers often combine them into a single representation.
 For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
-Some may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing location or gaze data (that is, where a subject is looking during data collection).
-Doing so allows for more convenient human analysis of the data in standard playback tools that support single data streams, but this spatial or temporal downsampling can obscure or mask some data patterns.
+Some may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing object or eye movement.
+Doing so allows for more convenient human analysis of the data in standard playback tools that support single data streams, but often involves losing data either through masking or downsampling.
 
-A session container captures the relationships among data elements from a single data acquisition, capturing heterogeneous file types \emph{and} their absolute or relative temporal positioning.
+A session captures these relationships among data elements from a single data acquisition, representing files along with their absolute or relative temporal positioning.
 The metaphor we use here is a timeline: each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
@@
 These sync points can then be used to manually place recordings on the timeline.
-In this way, the session container represents the structure of a data collection session but retains significant flexibility.
+In this way, the session represents recordings and related data streams in a way familiar to researchers, while keeping data in its most raw and flexible form.
 
 \subsection{Records}
+
 A record can represent many things that are typically conceived of as a row of a database.
@@
 In the case of a record representing an individual participant, these values might include subject identifier, gender, birth date, demographic information, measurement data, or other collected variables.
-For other record types, these values may include relevant independent variables, parameters, or descriptions. 
+For other record types, these values may include relevant independent variables, parameters, identifiers, or descriptions. 
 We can thus display the records of a given type as a table, with the values of each record constituting a row.
@@
 We have found that different researchers organize data files in substantially different ways.
-Researchers differentially consider their primary organizational unit, often depending on the primary focus of analysis for their specific study or sub-domain.
-They may organize their data by participants, session date, experimental condition, phase of analysis, or age group.
+Researchers employ different organizational structures, often depending on the primary focus of analysis for their specific study or sub-domain.
+Common directory structures include participants, session date, experimental condition, phase of analysis, and age group.
 Sometimes even single video files are split up into segments according to scene or activity, and researchers consider this property the primary organizational principle.
 
-Rather than trying to handle all of these strategies separately, Databrary leverages the power of records.
-Records can be attached to sessions, or, sessions can be grouped into records, in a many-to-many relationship.
+Rather than trying to handle all of these strategies separately, Databrary leverages records to provide arbitrary groupings.
+Records can be attached to sessions, or, equivalently, sessions can be grouped into records, in a many-to-many relationship.
 For example, a study participant may appear in exactly one session, so the collected information about that participant, added to a record, is attached in a one-to-one relationship with a specific session.
-In other research, multiple participants may appear in multiple sessions, such as recordings of the same class of students over a span of time.
-% for brevity, can we either exclude or condense this sentence?: A dataset may consist of distinct sub-sets of participants based on differences in experimental procedures or locations; here each session would be placed into a record representing the procedure or location.
+In other research, multiple participants may appear in multiple sessions, such as recordings of the same class of students over a span of time, and so the record may be attached to multiple sessions.
+This similarly extends to any other groupings of sessions based on experimental procedures or other factors by attaching the appropriate record of the corresponding type.
 
 Since researchers can create whichever record types apply to their data, they can also choose to group their data as they wish.
-This allows users to import data in what they consider the native directory structure by indicating the appropriate nesting of record types along with record variables themselves in traditional CSV formats.
-The same flexbility applies to data export.
+This allows users to import, browse, or export data in what they consider the native directory structure by indicating the appropriate nesting of record types
 By providing standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants within a particular age range.
@@
 
-% rog this section needs work.
+Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
+Specialized desktop software tools enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
+Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
+Moreover, with rare exceptions in specific research domains \cite{MacWhinney2001}, there are few standards for annotations themselves, even across studies performed by the same researcher.
 
-Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
-Specialized desktop software tools---both commercial (e.g., Noldus Observer, Mangold Interact, and Studiocode) and academic (e.g., Transana, Elan, CLAN, and Datavyu)---enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
-Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
-Moreover, with rare exceptions \cite{MacWhinney2001} there are few standards for annotations themselves, even across studies performed by the same researcher. 
-%moved ^ from intro
-
-Establishing a timeline for each session allows temporally-positioned annotations to be attached to the session.
+Establishing a timeline for each session also allows these sorts of temporally-positioned annotations to be attached.
 Currently we support two types of annotations: structured annotations based on records and added by the data contributors; and simple annotations contributed by the community.
@@
 This allows an additional type of organizational structure researchers employ, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
-This can also theoretically be used to represent annotations at a finer temporal scale, for example individual trials or events that are normally handled by off-line annotation software.
+This can also theoretically be used to represent annotations at a finer temporal scale, for example individual trials or events.
 However, the appropriate interfaces and tools to input or import these annotations are still being developed.
 
-Second, we support simple keyword-based tagging and comment-based discussions on any temporal segment of a session.
+Second, we allow users to contribute simple keyword-based tagging and comment-based discussions on any temporal segment of a session as they watch the recordings.
 Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
 In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
-Where research of community tagging provides evidence for the case that not many users actually do participate in tagging, we intend to leverage automated tagging through already existing metadata and analysis to seed a base of tags for reuse by the community \cite{Yang_Lu_Giles_2011, Farooq_etal_2007}.
-%%dg - reflagging this last sentence, it's valuable, but seems to do a major belly flop right here.
+There may also be useful automated approaches for seeding tags to be refined by the community based on existing metadata and analysis \cite{Yang_Lu_Giles_2011,Farooq_etal_2007,Giles2013}.
 
@@
 
-%Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
+Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
 We allow users, using certain portable formats (flat-files, documents, images), to include any data that cannot be encompassed by our standard formats.
@@
 
-Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to
-
-particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
+Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
 

### 2015-01-30 jcdl-15/paper.tex
A bit more work on normalization, discovery

@@
 
-\numberofauthors{3}
+\numberofauthors{4}
 
@@
 \alignauthor
+Lisa Steiger\\
+	\affaddr{Databrary Project}\\
+	\affaddr{New York University}\\
+	\affaddr{New York, NY, USA}\\
+	\email{lisa@databrary.org}
+\and
 Rick O. Gilmore\\
@@
 It currently hosts 5,700 video files totaling 1,600 hours of recordings along with 2,200 additional files.
-These files comprise 2,400 sessions and are covered by 1,300 records (including 1100 individual participants).
+These files comprise 2,400 sessions and are covered by 1,300 records (including 1,100 individual participants).
 Data originates from 35 individual contributors across 25 different universities.
@@
 
-%%dg - Reading this again, this section seems a little redudant with what is discussed above with making records and sessions flexible?
-Might be able to sum up this section and the discovery section in a single 'Lessons learned and Upcoming Challenges' section
+Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
+We additionally allow users, using certain portable formats (flat files, documents, images), to include data is not encompassed by our standard formats, but which may still be relevant for reuse.
+Beyond that, we have introduced a number of metadata concepts where this tension becomes most apparent, namely the available record types, variables, and tag keywords.
+We have adopted an approach that encourages adherence to existing standardized options but still allows new types to be created as the data require it.
+Initially we have achieved this through manual review and off-line curation of contributions, dynamically extending the system as new types of data were discovered without burdening or constraining researchers.
 
-Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
-We allow users, using certain portable formats (flat-files, documents, images), to include any data that cannot be encompassed by our standard formats.
-In this way, we can incorporate data which may be relevant for reuse but not broadly adopted by the community, without making the system too rigid.
-
-Here we have also introduced a number of metadata concepts where this tension becomes most apparent, namely the available record types, variables (and values within categorical data types), and tag keywords.
-We have adopted an approach that encourages adherence to existing standardized options but  allows new types to be created as the data require it.
-Initially we have achieved this through manual review and off-line curation of contributions, dynamically extending the system as new types of data were discovered.
-For many previously collected datasets, this has proven the most effective method to infer from data and avoid burdening contributors.
 However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a distinct, user-driven approach.
-
-We found that while some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools, often using a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and available filesystem storage.
+We found that while some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools, often using a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and local filesystem storage.
 This suggested the opportunity to supply a cohesive data management platform where all collected raw data, including recordings, contextual information, and measurements, could be entered, stored, and exported for later off-line analysis.
-Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset through liberal use of auto-completion, suggestions, and drop-down options.
+Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of auto-completion, suggestions, and drop-down options.
 With the considerable benefit of reducing storage requirements within researcher's labs, and often adding convenience over existing practices, this has proven an attractive prospect for researchers.
-
-Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
 
@@
 While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely unsolved.
-This problem is compounded for us because we expect researchers not only to reuse and reanalyze data within their own research area, but potentially to find new uses data originally collected for entirely different purposes.
-Recordings with potential value in one research area may be primarily labeled with contextual information from a different research area.
-Ultimately, making this determination requires direct, human observation of the recordings, and so a large part of solving the discovery problem will be simply to present researchers with short, representative samples from a variety of datasets.
+This problem is compounded for us because we expect that data collected in one research area, to be useful for entirely different purposes.
+Thus, data arrives labeled with contextual information that may not fully describe the range of features present.
+Ultimately, making determination of usefulness requires direct, human observation of the recordings, and so a large part the discovery solution will be simply to present researchers with short, representative samples from a variety of datasets.
 
@@
 Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.
+%Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
 

### 2015-01-30 jcdl-15/paper.tex
Minor edit conclusion

@@
 
-Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample/frame-rates.
+Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample\-/frame-rates.
 We additionally allow users, using certain portable formats (flat files, documents, images), to include data is not encompassed by our standard formats, but which may still be relevant for reuse.
@@
 
-Starting from the perspective of a general-purpose repository, we have introduced a number of  restrictions to create one specialized for specific types of data.
-By focusing on video and audio recordings we can normalize presentation and discovery of data across the site, and also introduce temporal structure to data and annotations driven by the nature of these recordings.
+We have built a research data repository focused on video and audio recordings that allows for normalized presentation and discovery of data across the site and introduces temporal structure to data and annotations driven by the nature of these recordings.
 By limiting both file and metadata organization to a flat (but overlapping) set of sessions and records, we capture session-based research data within a cohesive and intuitive user interface.
@@
 Similarly, by targeting reuse of existing research data, rather than replication or meta-analysis, we limit the scope of accepted data to the original recordings and early analysis phases that generate the most value.
+Focus on a particular research domain allows us to achieve a sustainable balance between standardization and flexibility without burdening contributors.
 
@@
 Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.
-%Achieving a sustainable balance will no doubt be challenging, but by using the above approaches and limiting our focus to particular research domains we have been able to achieve substantial standardization across the entire data library without overly burdening contributors.
 
-There is also active discussion around the role of repositories and libraries in archiving and providing access to research data, where the roles and responsibilities between libraries and research departments is not entirely established \cite{Castelli_etal_2013, Nielson_Hjørland_2014, Macmillan_2014, Pinfield_etal_2014}.
+There is also active discussion around the role of repositories and libraries in archiving and providing access to research data, where the roles and responsibilities between libraries and research departments is not entirely established \cite{Castelli_etal_2013,Nielson_Hjørland_2014,Macmillan_2014,Pinfield_etal_2014}.
 Although resolving this complicated issue is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer established library standards that achieve these goals.

### 2015-01-30 jcdl-15/paper.tex
Move repository infrastructure overview sections

@@
 
+\subsection*{Repository Overview}
+
+Databrary began accepting contributions in early 2014 and opened for general use in October 2014.
+It currently hosts 5,700 video files totaling 1,600 hours of recordings along with 2,200 additional files.
+These files comprise 2,400 sessions and are covered by 1,300 metadata records (including 1,100 individual participants).
+Data originates from 35 individual contributors across 25 different universities.
+
+Databrary is a new, cohesive web application, built in Scala on the Play Framework\footnote{http://playframework.com/} to enable a responsive user interface, a complete API, and high-performance streaming.
+Recordings and other files are placed in content-based filesystem storage, and all structured data are stored in a PostgreSQL database, leveraging its geometric indexing capabilities for temporal data.
+All uploaded recordings are automatically transcoded to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
+This transcoding utilizes the high performance computing cluster at the host campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
+The user interface is built primarily on the Angular web framework\footnote{http://angularjs.org/}, and all data access is performed through an open JSON API.
+Although hosted data are protected, all source code is released under a GPLv3 license on github\footnote{http://github.com/databrary}.
+
 \section{Data Architecture}
@@
 
-\subsection{Technical infrastructure}
-
-%%dg - voting to put this and the repository section up higher in the paper, Approach, maybe?
-Databrary is a new, cohesive web application, built in Scala on the Play Framework\footnote{http://playframework.com/} to enable a responsive user interface, a complete API, and high-performance streaming.
-Recordings and other files are placed in content-based filesystem storage, and all structured data are stored in a PostgreSQL database, leveraging its geometric indexing capabilities for temporal data.
-All uploaded recordings are automatically transcoded to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
-This transcoding utilizes the high performance computing cluster at the host campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
-The user interface is built primarily on the Angular web framework\footnote{http://angularjs.org/}, and all data access is performed through an open JSON API.
-Although hosted data are protected, all source code is released under a GPLv3 license on github\footnote{http://github.com/databrary}.
-
-\section{Repository}
-
-Databrary accepted contributions beginning in early 2014 and opened for general use in October 2014.
-It currently hosts 5,700 video files totaling 1,600 hours of recordings along with 2,200 additional files.
-These files comprise 2,400 sessions and are covered by 1,300 records (including 1,100 individual participants).
-Data originates from 35 individual contributors across 25 different universities.
+\section{Interface and Interoperability}
 

### 2015-01-30 jcdl-15/paper.tex
Move DLS paragraph into discovery
some clean up on intro

@@
 Despite their inherent reusability, this lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
-To address this problem, the Databrary project aims to build tools and policies that facilitate repurposing and reuse.
+To address this problem, the Databrary project aims to build tools and policies that facilitate repurposing and reuse of recordings.
 We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways, in order to enable understanding and discovery without placing undue burden on data contributors.
@@
 Our approach is to focus on video and audio recordings in particular research domains, and build appropriate structured and unstructured resources around this functional core.
-By investigating of how researchers studying human development and education organize their own data, we have identified a unifying set of principles around which to structure contextual metadata.
+By investigating how researchers studying human development and education organize their own data, we have identified a unifying set of principles around which to structure contextual metadata.
 This structure facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
@@
 
+There is also active discussion around the role of repositories and libraries in archiving and providing access to research data, where the roles and responsibilities between libraries and research departments is not entirely established \cite{Castelli_etal_2013,Nielson_Hjørland_2014,Macmillan_2014,Pinfield_etal_2014}.
+Although resolving this complicated issue is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer established library standards that achieve these goals.
+We may be well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS schemas) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
+By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
+
 \section{Conclusions}
@@
 Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.
-
-There is also active discussion around the role of repositories and libraries in archiving and providing access to research data, where the roles and responsibilities between libraries and research departments is not entirely established \cite{Castelli_etal_2013,Nielson_Hjørland_2014,Macmillan_2014,Pinfield_etal_2014}.
-Although resolving this complicated issue is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer established library standards that achieve these goals.
-We may be well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS schemas) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
-By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
 

### 2015-01-30 jcdl-15/paper.tex
library rewording

--- 2015-01-30 jcdl-15/paper.tex
@@
 
+There is also active discussion around the role libraries in providing repositories and services for research data \cite{Castelli_etal_2013,Nielson_Hjørland_2014,Macmillan_2014,Pinfield_etal_2014}.
+Although resolving the complicated issues that arise within the library domain is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer to established library standards that achieve these goals, further enhancing the discoverability of these data.
+Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS schemas) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
+By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
+
 \section{Conclusions}

### 2015-01-30 jcdl-15/paper.tex
minor edits

@@
 Since researchers can create whichever record types apply to their data, they can also choose to group their data as they wish.
-This allows users to import, browse, or export data in what they consider the native directory structure by indicating the appropriate nesting of record types
+This allows users to import, browse, or export data in what they consider the native directory structure by indicating the appropriate nesting of record types.
 By providing standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants within a particular age range.
@@
 
-Establishing a timeline for each session also allows these sorts of temporally-positioned annotations to be attached.
+Establishing a timeline for each session allows these sorts of temporally-positioned annotations to be attached.
 Currently we support two types of annotations: structured annotations based on records and added by the data contributors; and simple annotations contributed by the community.

### 2015-01-30 jcdl-15/paper.tex
change url refs to footnotes

@@
 Research data repositories require a balance between generality and specificity.
-General purpose repositories permit unstructured storage of arbitrary files \cite{dataverse, dryad, purr}, while those with rigorously structured schema require contributions to be to carefully formatted \cite{HCP, PGP, talkbank}.
+General purpose repositories permit unstructured storage of arbitrary files (e.g.
+Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, and PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be to carefully formatted (e.g.
+Human Connectome Project\footnote{http://www.humanconnectomeproject.org}, Personal Genome Project\footnote{http://www.personalgenomes.org}, and TalkBank\footnote{http://talkbank.org/}).
 The flexibility of general-purpose repositories make them especially well-suited to capture and archive entire datasets, but these systems do not facilitate search, understanding, or reuse without extensive, usually labor-intensive, curation to summarize or extract metadata \cite{Peer_2012}.

### 2015-01-30 jcdl-15/paper.tex
Combine some sentences; Drew -> Andrew

@@
 \alignauthor
-Drew Gordon\\
+Andrew Gordon\\
 	\affaddr{Databrary Project}\\
@@
 In addition to containing values, records are also used to organize sessions.
-We have found that different researchers organize data files in substantially different ways.
-Researchers employ different organizational structures, often depending on the primary focus of analysis for their specific study or sub-domain.
+We have found that different researchers employ substantially different organizational structures for their files depending on their needs.
 Common directory structures include participants, session date, experimental condition, phase of analysis, and age group.

### 2015-01-30 jcdl-15/paper.tex
changing some wording
structure

@@
 Our approach is to focus on video and audio recordings in particular research domains, and build appropriate structured and unstructured resources around this functional core.
-By investigating how researchers studying human development and education organize their own data, we have identified a unifying set of principles around which to structure contextual metadata.
-This structure facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
-We achieve this conformance through in-depth, investigative curation of previously collected datasets, along with familiar user interfaces allowing researchers to structure their own research data as they acquire it.
+By investigating how researchers studying human development and education organize their own data, we have identified a unifying set of principles for organizing contextual metadata.
+This facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
+We achieve this conformance through in-depth, investigative curation of previously collected datasets, and user interfaces allowing researchers to upload their own research data as they acquire it.
 In the future, we hope to capture additional annotation data researchers already produce while analyzing recordings to further improve our discovery capabilities.
@@
 A record can represent many things that are typically conceived of as a row of a database.
-Each may represent a single data sample, such as a participant, group, or classroom, a level of analysis, such as an experimental trial, task, condition, or location, or some other factor. 
+Each may represent a single data sample: such as a participant; group; or classroom, or a level of analysis: such as an experimental trial; task; condition; or location, or some other factor. 
 A record is thus defined by one of the above types along with any number of scalar values, which can be textual, numeric, or date.
@@
 In addition to containing values, records are also used to organize sessions.
-We have found that different researchers employ substantially different organizational structures for their files depending on their needs.
+We have found that different researchers employ varied arrangements for their files depending on their needs.
 Common directory structures include participants, session date, experimental condition, phase of analysis, and age group.

### 2015-01-30 jcdl-15/paper.tex
Tweak footnotes and title

@@
 
-\title{Databrary: A digital data library for sharing research video}
+\title{Databrary: Enabling sharing and reuse of research video data}
 
@@
 Research data repositories require a balance between generality and specificity.
-General purpose repositories permit unstructured storage of arbitrary files (e.g.
-Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, and PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be to carefully formatted (e.g.
-Human Connectome Project\footnote{http://www.humanconnectomeproject.org}, Personal Genome Project\footnote{http://www.personalgenomes.org}, and TalkBank\footnote{http://talkbank.org/}).
+General purpose repositories permit unstructured storage of arbitrary files (e.g., Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be to carefully formatted (e.g., Human Connectome Project\footnote{http://www.humanconnectomeproject.org/}, Personal Genome Project\footnote{http://www.personalgenomes.org/}, TalkBank\footnote{http://talkbank.org/}).
 The flexibility of general-purpose repositories make them especially well-suited to capture and archive entire datasets, but these systems do not facilitate search, understanding, or reuse without extensive, usually labor-intensive, curation to summarize or extract metadata \cite{Peer_2012}.

### 2015-01-30 jcdl-15/paper.tex
just removed stuff that seemed to be tangential or another issue unto itself

--- 2015-01-30 jcdl-15/paper.tex
@@
 
-Since we primarily deal with video and audio data, standardizing on an encoding format allows for convenient, consistent access without much loss of flexibility, while still allowing arbitrary resolutions, encoding qualities, and sample\-/frame-rates.
-We additionally allow users, using certain portable formats (flat files, documents, images), to include data is not encompassed by our standard formats, but which may still be relevant for reuse.
-Beyond that, we have introduced a number of metadata concepts where this tension becomes most apparent, namely the available record types, variables, and tag keywords.
 We have adopted an approach that encourages adherence to existing standardized options but still allows new types to be created as the data require it.
 Initially we have achieved this through manual review and off-line curation of contributions, dynamically extending the system as new types of data were discovered without burdening or constraining researchers.
+However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a distinct, user-driven approach.
 
-However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a distinct, user-driven approach.
 We found that while some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools, often using a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and local filesystem storage.
@@
 Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of auto-completion, suggestions, and drop-down options.
-With the considerable benefit of reducing storage requirements within researcher's labs, and often adding convenience over existing practices, this has proven an attractive prospect for researchers.
+In turn, this adds convenience for researchers over existing practices.
 
@@
 Another important source for search targets would come from annotations supplied both by the original researchers during analysis as well as by the community.
-In the former case, normalization remains a concern, as most annotation schemes researchers employ involve opaque, symbolic or numeric codes with little semantic content.
 We are currently working to support import and export workflows for existing annotations tools that will be able to add meaningful definitions to these codes while researchers continue their current practices.
@@
 Thus we seek to incorporate these existing annotation processes into Databrary, both to capture and define these annotations as generalizable search terms, and to make these workflows more convenient by obviating existing file transfer and transformation burdens.
+%%below sentence too long? remove or shorten?
 Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.
 
-Our single-minded focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers gives us a novel perspective on building domain-specific research repositories.
-The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
+The success we've had in engaging, involving, and extracting data from researchers is due largely to our focus on a specific community of researchers, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
 By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.

### 2015-01-30 jcdl-15/paper.tex
Manually merge in some changes from "paper-rog.tex"

--- 2015-01-30 jcdl-15/paper.tex
@@
 Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes.
-However, there are numerous technical and policy challenges preventing existing recordings from being accessed and reused by other researchers.
-By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share these data in a standardized way.
+However, technical and policy challenges prevent existing recordings from being accessed and reused by other researchers.
+By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
 There are remaining challenges around discovery and annotation, and we discuss some options and initial approaches.
-While many of Databrary's strengths have derived from its focus on particular scholarly domains, many of these issues have broader relevance to the digital library community.
+While Databrary's strengths derive from its focus on particular scholarly domains, many of these issues have broader relevance to the digital library community.
 \end{abstract}
@@
 
-Video and audio recordings serve as a primary raw data source for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
+Video and audio recordings serve as a primary data source for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
 They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools as human observers can directly glean much of the data's richness.
@@
 At the top level of organization, we separate individual data contributions into collections or packages called \emph{volumes}.
-Volumes allow a minimal amount of metadata (title, description, permissions, etc.) as well as additional supporting files (documents and other materials).
+Volumes support a minimal amount of metadata (title, description, permissions, etc.) along with additional supporting files (documents and other materials).
 They are persistently identified and can cite external resources by URI or Digital Object Identifier (DOI).
@@
 As recordings are often the core component of a data collection, a session in many cases is defined by a single recording.
-In some cases a single session can comprise multiple videos, either serially in time representing different segments or in parallel from multiple cameras recording the same events.
+In some cases a single session comprises multiple videos, either serially in time representing different segments or in parallel from multiple cameras recording the same events.
 These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
@@
 
-A session captures these relationships among data elements from a single data acquisition, representing files along with their absolute or relative temporal positioning.
-The metaphor we use here is a timeline: each session defines a time range, with absolute timestamps representing the exact time of data collection or relative positioning with an optional date.
+A session captures relationships among data elements from a single data acquisition, representing files along with their absolute or relative temporal positioning.
+The metaphor we use here is a timeline: each session defines a time range, with absolute timestamps representing the exact point of data collection or relative positioning with an optional date.
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
@@
 
-A record can represent many things that are typically conceived of as a row of a database.
-Each may represent a single data sample: such as a participant; group; or classroom, or a level of analysis: such as an experimental trial; task; condition; or location, or some other factor. 
+A record can represent many things typically conceived of as a row of a database.
+Each may represent: a single data sample, such as a participant, group, or classroom; a level of analysis, such as an experimental trial, task, condition, or location; or some other factor. 
 A record is thus defined by one of the above types along with any number of scalar values, which can be textual, numeric, or date.
-In the case of a record representing an individual participant, these values might include subject identifier, gender, birth date, demographic information, measurement data, or other collected variables.
-For other record types, these values may include relevant independent variables, parameters, identifiers, or descriptions. 
+In the case of a record representing an individual participant, these values might include subject identifier, gender, birth date, demographic information, measurements, or other collected variables.
+For other record types, these values may include independent variables, parameters, identifiers, or descriptions. 
 We can thus display the records of a given type as a table, with the values of each record constituting a row.
@@
 In addition to containing values, records are also used to organize sessions.
-We have found that different researchers employ varied arrangements for their files depending on their needs.
-Common directory structures include participants, session date, experimental condition, phase of analysis, and age group.
-Sometimes even single video files are split up into segments according to scene or activity, and researchers consider this property the primary organizational principle.
+Different researchers employ substantially different directory structures for their files depending on their needs.
+Common groupings include participants, session date, experimental condition, phase of analysis, and age group.
+Sometimes single video files are split into segments according to scene or activity, and researchers consider this property the primary organizational principle.
 
@@
 For example, a study participant may appear in exactly one session, so the collected information about that participant, added to a record, is attached in a one-to-one relationship with a specific session.
-In other research, multiple participants may appear in multiple sessions, such as recordings of the same class of students over a span of time, and so the record may be attached to multiple sessions.
+In other research, multiple participants appear in multiple sessions, such as recordings of the same class of students over a span of time, and so the record may be attached to multiple sessions.
 This similarly extends to any other groupings of sessions based on experimental procedures or other factors by attaching the appropriate record of the corresponding type.
@@
 Since researchers can create whichever record types apply to their data, they can also choose to group their data as they wish.
-This allows users to import, browse, or export data in what they consider the native directory structure by indicating the appropriate nesting of record types.
-By providing standardization of record types, we can also perform broad, meta-searches across the library, for example for all videos containing participants within a particular age range.
+This allows users to import, browse, or export data in what they consider the most intuitive directory structure by indicating the appropriate nesting of record types.
+By standardizing record types, we can also perform targeted searches across the library, for example for all videos containing participants within a particular age range.
 
@@
 Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
-Specialized desktop software tools enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
-Most of these tools currently focus on a single recording, and despite the plethora of tools and features, there is little interoperability.
+Various specialized software tools enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
+Most tools focus on a single recording, and there is little interoperability among them.
 Moreover, with rare exceptions in specific research domains \cite{MacWhinney2001}, there are few standards for annotations themselves, even across studies performed by the same researcher.
@@
 Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
-In addition to enriching datasets and enabling dialog between researchers, we expect these annotations to provide an index target for additional discovery capabilities for users looking for particular features within data \cite{Lanagan_Smeaton_2012}.
-There may also be useful automated approaches for seeding tags to be refined by the community based on existing metadata and analysis \cite{Yang_Lu_Giles_2011,Farooq_etal_2007,Giles2013}.
+In addition to enriching datasets and enabling dialog between researchers, we expect annotations to provide an index target for additional discovery capabilities for finding for particular features within data \cite{Lanagan_Smeaton_2012}.
+We will also evaluate automated approaches for seeding tags to be refined by the community based on existing metadata and analysis \cite{Yang_Lu_Giles_2011,Farooq_etal_2007,Giles2013}.
 
@@
 
-Our single-minded focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers gives us a novel perspective on building domain-specific research repositories.
+Databrary's focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers provides a novel perspective on building domain-specific research repositories.
 The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.

### 2015-01-30 jcdl-15/paper.tex

@@
 Thus we seek to incorporate these existing annotation processes into Databrary, both to capture and define these annotations as generalizable search terms, and to make these workflows more convenient by obviating existing file transfer and transformation burdens.
+%%below sentence too long? remove or shorten?
 Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.
 
-Databrary's focus on making collected, identifiable recordings of human research available and reusable to a whole community of researchers provides a novel perspective on building domain-specific research repositories.
-The success we've had in engaging, involving, and extracting data from researchers is due largely to this focus, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
+The success we've had in engaging, involving, and extracting data from researchers is due largely to our focus on a specific community of researchers, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
 By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.

### 2015-01-30 jcdl-15/paper.tex
Minor editing, remove word from title

@@
 
-\title{Databrary: Enabling sharing and reuse of research video data}
+\title{Databrary: Enabling sharing and reuse of research video}
 
@@
 Research data repositories require a balance between generality and specificity.
-General purpose repositories permit unstructured storage of arbitrary files (e.g., Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be to carefully formatted (e.g., Human Connectome Project\footnote{http://www.humanconnectomeproject.org/}, Personal Genome Project\footnote{http://www.personalgenomes.org/}, TalkBank\footnote{http://talkbank.org/}).
+General purpose repositories permit unstructured storage of arbitrary files (e.g., Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be to carefully formatted (e.g., Human Connectome Project\footnote{http://www.humanconnectomeproject.org/}, Per\-son\-al Ge\-nome Pro\-ject\footnote{http://www.personalgenomes.org/}, TalkBank\footnote{http://talkbank.org/}).
 The flexibility of general-purpose repositories make them especially well-suited to capture and archive entire datasets, but these systems do not facilitate search, understanding, or reuse without extensive, usually labor-intensive, curation to summarize or extract metadata \cite{Peer_2012}.
@@
 Another important source for search targets would come from annotations supplied both by the original researchers during analysis as well as by the community.
-We are currently working to support import and export workflows for existing annotations tools that will be able to add meaningful definitions to these codes while researchers continue their current practices.
+We are currently working to support import and export workflows for existing annotations tools that will be able to add meaningful definitions to these currently opaque codes while researchers continue their current practices.
 Beyond that, we also plan to build a new, more efficient and structured annotation web interface enabling more of the analysis workflow directly within our system.

### 2015-01-30 jcdl-15/paper.tex
more tiny language tweaks

@@
 Our approach is to focus on video and audio recordings in particular research domains, and build appropriate structured and unstructured resources around this functional core.
-By investigating how researchers studying human development and education organize their own data, we have identified a unifying set of principles for organizing contextual metadata.
+We designed the system by investigating how researchers studying human development and education organize their own data, deriving from those observations a unifying set of principles for organizing contextual metadata.
 This facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.

### 2015-01-30 jcdl-15/paper.tex
remove reference

@@
 In addition to enriching datasets and enabling dialog between researchers, we expect annotations to provide an index target for additional discovery capabilities for finding for particular features within data \cite{Lanagan_Smeaton_2012}.
-We will also evaluate automated approaches for seeding tags to be refined by the community based on existing metadata and analysis \cite{Yang_Lu_Giles_2011,Farooq_etal_2007,Giles2013}.
+We will also evaluate automated approaches for seeding tags to be refined by the community based on existing metadata and analysis \cite{Yang_Lu_Giles_2011,Giles2013}.
 

### 2015-01-30 jcdl-15/paper.tex
Lose open source line, merge into first

@@
 
-Databrary is a new, cohesive web application, built in Scala on the Play Framework\footnote{http://playframework.com/} to enable a responsive user interface, a complete API, and high-performance streaming.
+Databrary is a new, open-source web application\footnote{http://github.com/databrary/databrary}, built in Scala on the Play Framework\footnote{http://playframework.com/} to support a responsive user interface, a complete API, and high-performance streaming.
 Recordings and other files are placed in content-based filesystem storage, and all structured data are stored in a PostgreSQL database, leveraging its geometric indexing capabilities for temporal data.
@@
 The user interface is built primarily on the Angular web framework\footnote{http://angularjs.org/}, and all data access is performed through an open JSON API.
-Although hosted data are protected, all source code is released under a GPLv3 license on github\footnote{http://github.com/databrary}.
 

### 2015-01-30 jcdl-15/paper.tex
some minor phrasing edits

@@
 In some cases a single session comprises multiple videos, either serially in time representing different segments or in parallel from multiple cameras recording the same events.
-These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
+These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys.
+All of which have specific and critical temporal relationships with the primary data.
 
@@
 There is also active discussion around the role libraries in providing repositories and services for research data \cite{Castelli_etal_2013,Nielson_Hjørland_2014,Macmillan_2014,Pinfield_etal_2014}.
-Although resolving the complicated issues that arise within the library domain is not directly within our purview, we have laid out a functioning system for research data management that may bring data closer to established library standards that achieve these goals, further enhancing the discoverability of these data.
+Although resolving the complicated issues that arise within the library domain is not directly within our purview, we have laid out a functioning system for research data management that may bring established library standards to bear on research data, further enhancing their discoverability.
 Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS schemas) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
@@
 Thus we seek to incorporate these existing annotation processes into Databrary, both to capture and define these annotations as generalizable search terms, and to make these workflows more convenient by obviating existing file transfer and transformation burdens.
-%%below sentence too long? remove or shorten?
 Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.

### 2015-01-31 jcdl-15/paper.tex
Fix fragment

@@
 In some cases a single session comprises multiple videos, either serially in time representing different segments or in parallel from multiple cameras recording the same events.
-These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys.
-All of which have specific and critical temporal relationships with the primary data.
+These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
 

### 2015-01-31 jcdl-15/paper.tex
a word

@@
 Although resolving the complicated issues that arise within the library domain is not directly within our purview, we have laid out a functioning system for research data management that may bring established library standards to bear on research data, further enhancing their discoverability.
-Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS schemas) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
+Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
 By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.

### 2015-01-31 jcdl-15/paper.tex
Final minor edits through intro

@@
 Video and audio recordings serve as a primary data source for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
-They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools as human observers can directly glean much of the data's richness.
+They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools, as human observers can directly glean much of the data's richness.
 % access in the sense of "what format is it in":
-This allows recordings collected in one context to be used by others for different purposes with minimal explanation, naturally reducing the problems of documentation, provenance, and access that plague many research repositories.
-On the other hand, automated analysis of videos is an active but still immature research area, so the problems of search, discoverability, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
+This allows recordings collected in one context to be used by others for different purposes with minimal explanation, naturally reducing the problems of documentation, provenance, formatting, and access that plague many research repositories.
+On the other hand, automated analysis of videos is an active but still immature research area, so the problems of search, discovery, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
 
@@
 Despite their inherent reusability, this lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
-To address this problem, the Databrary project aims to build tools and policies that facilitate repurposing and reuse of recordings.
+To address this problem, the Databrary project\footnote{https://databrary.org/} aims to build tools and policies that facilitate repurposing and reuse of recordings.
 We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways, in order to enable understanding and discovery without placing undue burden on data contributors.
@@
 Research data repositories require a balance between generality and specificity.
-General purpose repositories permit unstructured storage of arbitrary files (e.g., Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be to carefully formatted (e.g., Human Connectome Project\footnote{http://www.humanconnectomeproject.org/}, Per\-son\-al Ge\-nome Pro\-ject\footnote{http://www.personalgenomes.org/}, TalkBank\footnote{http://talkbank.org/}).
-The flexibility of general-purpose repositories make them especially well-suited to capture and archive entire datasets, but these systems do not facilitate search, understanding, or reuse without extensive, usually labor-intensive, curation to summarize or extract metadata \cite{Peer_2012}.
+General purpose repositories permit unstructured storage of arbitrary files (e.g., Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be carefully formatted (e.g., Human Connectome Project\footnote{http://www.humanconnectomeproject.org/}, Per\-son\-al Ge\-nome Pro\-ject\footnote{http://www.personalgenomes.org/}, TalkBank\footnote{http://talkbank.org/}).
+The flexibility of general-purpose repositories make them especially well-suited to capture and archive entire datasets, but these systems do not facilitate search, understanding, or reuse without extensive, usually labor-intensive curation to summarize or extract metadata \cite{Peer_2012}.
 Targeted repositories, on the other hand, generate large-scale, homogeneous data sources that facilitate meta- and re-analysis, but are limited to specific research questions, losing the full richness, context, and innovation of the source materials and field.

### 2015-02-01 jcdl-15/paper.tex
Full final editing pass

@@
 Despite their inherent reusability, this lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
-To address this problem, the Databrary project\footnote{https://databrary.org/} aims to build tools and policies that facilitate repurposing and reuse of recordings.
-We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways, in order to enable understanding and discovery without placing undue burden on data contributors.
+To address this problem, the Databrary project\footnote{https://databrary.org/} aims to build tools and policies that facilitate repurposing and reuse of recordings without placing undue burden on data contributors.
+We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways, allowing for understanding and discovery.
 
@@
 This facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
-We achieve this conformance through in-depth, investigative curation of previously collected datasets, and user interfaces allowing researchers to upload their own research data as they acquire it.
-In the future, we hope to capture additional annotation data researchers already produce while analyzing recordings to further improve our discovery capabilities.
+We achieve this conformance through in-depth, investigative curation of previously collected datasets, along with user interfaces allowing researchers to upload their own research data as they acquire it.
+In the future, we hope to capture the additional annotation data researchers already produce while analyzing recordings to further improve our discovery capabilities.
 
@@
 It currently hosts 5,700 video files totaling 1,600 hours of recordings along with 2,200 additional files.
-These files comprise 2,400 sessions and are covered by 1,300 metadata records (including 1,100 individual participants).
+These files make up 2,400 sessions and are covered by 1,300 metadata records (including 1,100 individual participants).
 Data originates from 35 individual contributors across 25 different universities.
@@
 As recordings are often the core component of a data collection, a session in many cases is defined by a single recording.
-In some cases a single session comprises multiple videos, either serially in time representing different segments or in parallel from multiple cameras recording the same events.
+In some cases a single session comprises multiple videos, either serially in time representing different segments, or in parallel from multiple cameras recording the same events.
 These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
@@
 
-A session captures relationships among data elements from a single data acquisition, representing files along with their absolute or relative temporal positioning.
-The metaphor we use here is a timeline: each session defines a time range, with absolute timestamps representing the exact point of data collection or relative positioning with an optional date.
+A session captures relationships among data elements from a single data acquisition, representing files along with their temporal positioning and optional date of collection.
+The metaphor we use here is a timeline: each session defines a time range, with absolute timestamps representing the exact point of data collection, or relative positioning from an arbitrary 0-point.
 Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
@@
 
-In addition to containing values, records are also used to organize sessions.
+In addition to containing values, records are used to organize sessions.
 Different researchers employ substantially different directory structures for their files depending on their needs.
-Common groupings include participants, session date, experimental condition, phase of analysis, and age group.
+Common approaches involve creating one directory per participant, session date, experimental condition, phase of analysis, and/or age group.
 Sometimes single video files are split into segments according to scene or activity, and researchers consider this property the primary organizational principle.
@@
 Most tools focus on a single recording, and there is little interoperability among them.
-Moreover, with rare exceptions in specific research domains \cite{MacWhinney2001}, there are few standards for annotations themselves, even across studies performed by the same researcher.
+Moreover, with rare exceptions in specific research domains \cite{MacWhinney2001}, there are few standards or definitions for the annotations themselves, even across studies performed by the same researcher.
 
@@
 
-First, records can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
+The first type is created using records, which can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
 If any record relates to a part, scene, or section of a session---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only that appropriate part.
@@
 
-Second, we allow users to contribute simple keyword-based tagging and comment-based discussions on any temporal segment of a session as they watch the recordings.
-Such annotations can also be added through our API by automated video analysis according to features of interest identified in the data.
+For the second, we allow users to contribute keyword-based tags and comment-based discussions for any temporal segment of a session as they watch the recordings.
+Such annotations could also be added by automated video analysis according to features of interest identified in the data, using our API.
 In addition to enriching datasets and enabling dialog between researchers, we expect annotations to provide an index target for additional discovery capabilities for finding for particular features within data \cite{Lanagan_Smeaton_2012}.
@@
 
-We found that while some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools, often using a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and local filesystem storage.
-This suggested the opportunity to supply a cohesive data management platform where all collected raw data, including recordings, contextual information, and measurements, could be entered, stored, and exported for later off-line analysis.
-Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of auto-completion, suggestions, and drop-down options.
-In turn, this adds convenience for researchers over existing practices.
+While some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools.
+Most use a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and local filesystem storage.
+This suggested the opportunity to supply a cohesive data management platform where all collected raw data---including recordings, contextual information, and measurements---could be entered, stored, and exported for later off-line analysis.
+Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
+In turn, this adds convenience and functionality for researchers over existing practices while alleviating storage burdens.
 
@@
 While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely unsolved.
-This problem is compounded for us because we expect that data collected in one research area, to be useful for entirely different purposes.
-Thus, data arrives labeled with contextual information that may not fully describe the range of features present.
-Ultimately, making determination of usefulness requires direct, human observation of the recordings, and so a large part the discovery solution will be simply to present researchers with short, representative samples from a variety of datasets.
+This problem is compounded for us because we expect data collected in one research area to be useful for entirely different purposes.
+That is, data arrives labeled with contextual information that may not fully describe the range of features present.
+Ultimately, determining usefulness requires direct, human observation of the recordings, and so a large part the discovery solution will be simply to present researchers with short, representative samples from a variety of datasets.
 
-Another important source for search targets would come from annotations supplied both by the original researchers during analysis as well as by the community.
-We are currently working to support import and export workflows for existing annotations tools that will be able to add meaningful definitions to these currently opaque codes while researchers continue their current practices.
-Beyond that, we also plan to build a new, more efficient and structured annotation web interface enabling more of the analysis workflow directly within our system.
+Another important source for search targets comes from annotations supplied by the original researchers during analysis, as well as by the community.
+We are currently working to support import and export workflows for existing annotation tools that add meaningful definitions to these often opaque codes, while allowing researchers to continue their current practices.
+Beyond that, we also plan to build a new, more efficient and structured annotation web interface supporting more of the analysis workflow directly within our system.
 In this way, Databrary will allow researchers to work with their recordings directly in raw form without having to transfer or transform them, while also capturing more meaningful and standardized contextual information from this process.
@@
 Although resolving the complicated issues that arise within the library domain is not directly within our purview, we have laid out a functioning system for research data management that may bring established library standards to bear on research data, further enhancing their discoverability.
-Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS) and reach OAI-PMH compliance, so as to automatically incorporate the data that researchers add to Databrary into federated library searches with other domain specific data repositories.
+Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS) and reach OAI-PMH compliance.
+This will automatically incorporate data that researchers add to Databrary into federated library searches with other domain specific data repositories.
 By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
@@
 
-We have built a research data repository focused on video and audio recordings that allows for normalized presentation and discovery of data across the site and introduces temporal structure to data and annotations driven by the nature of these recordings.
+We have built a research data repository focused on video and audio recordings, allowing for normalized presentation and discovery of data across the site, and introducing temporal structure to metadata driven by the nature of these recordings.
 By limiting both file and metadata organization to a flat (but overlapping) set of sessions and records, we capture session-based research data within a cohesive and intuitive user interface.
@@
 Having made significant progress building ingest and data entry services that have let us grow the library, we now must turn our efforts towards discovery and reuse, finding ways to better describe and elaborate on collected data.
-To do this we look both to the original data contributors for ways that their existing analysis workflows can generate usable content descriptions, and to the community of researchers to label and comment on datasets as they explore.
-Thus we seek to incorporate these existing annotation processes into Databrary, both to capture and define these annotations as generalizable search terms, and to make these workflows more convenient by obviating existing file transfer and transformation burdens.
-Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users and existing data can continuously increase in value for researchers.
+To do this we look both to the original data contributors for ways that their existing analysis workflows can generate usable content descriptions, and to the community of researchers to label and comment on datasets as they explore and reuse.
+Thus we seek to incorporate these annotation processes into Databrary, both to capture and define these annotations as generalizable search terms, and to make these workflows more convenient by obviating existing file transfer and transformation burdens.
+Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users, and existing data can continuously increase in value for researchers.
 
-The success we've had in engaging, involving, and extracting data from researchers is due largely to our focus on a specific community of researchers, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
+The success we've had in engaging, involving, and extracting data from researchers is due largely to our focus on a specific community, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
 By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.

### 2015-02-01 jcdl-15/paper.tex
Edit distance 1

@@
 Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS) and reach OAI-PMH compliance.
-This will automatically incorporate data that researchers add to Databrary into federated library searches with other domain specific data repositories.
+This will automatically incorporate data that researchers add to Databrary into federated library searches with other domain-specific data repositories.
 By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.

### 2015-02-01 jcdl-15/paper.tex
One word

@@
 The first type is created using records, which can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
-If any record relates to a part, scene, or section of a session---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only that appropriate part.
+If any record relates to a part, scene, or section of a session---for example a participant being present for only part of a recording---it can be applied within the timeline of that session to only that appropriate part.
 This allows an additional type of organizational structure researchers employ, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.

### 2015-04-02 jcdl-15/paper_demo.tex
start a new copy for the jcdl demo
probably need to get rid of all of it and start from scratch, but could salvage a little for background.

--- 2015-02-01 jcdl-15/paper.tex
@@
 Rick O. Gilmore\\
+	\affaddr{Department of Psychology}\\
 	\affaddr{Penn State}\\
-	\affaddr{Department of Psychology}\\
 	\affaddr{University Park, PA, USA}\\
@@
 Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes.
-However, technical and policy challenges prevent existing recordings from being accessed and reused by other researchers.
+Databrary provides a web-based service for sharing and reusing the video data created by researchers in the developmental and learning sciences.
 By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
-There are remaining challenges around discovery and annotation, and we discuss some options and initial approaches.
-While Databrary's strengths derive from its focus on particular scholarly domains, many of these issues have broader relevance to the digital library community.
+This demo will take a tour through the Databrary service, hightlighting how it controls user access to restricted data based on human research subjects, how it facilitates browsing and discoverability of datasets, and how it provides a service for researchers to upload and manage their data as they collect it.
 \end{abstract}
@@
 
-\section{Introduction}
+\section{Introduction/Background}
 
@@
 
-\section{Data Architecture}
+\section{???}
 
-At the top level of organization, we separate individual data contributions into collections or packages called \emph{volumes}.
-Volumes support a minimal amount of metadata (title, description, permissions, etc.) along with additional supporting files (documents and other materials).
-They are persistently identified and can cite external resources by URI or Digital Object Identifier (DOI).
-
-Most importantly, a volume involves two key structural elements: the \emph{session} and the \emph{record}.
-Sessions provide the primary container for data files, an elaboration on traditional file folders with additional temporal metadata.
-Records represent all types of scalar variable-based data, traditionally contained in flat files or databases, and also provide an organizational structure among sessions.
-The combination of these two elements allows data and metadata to be flexibly structured, related, and annotated.
-
-\subsection{Sessions}
-
-Sessions provide a container for raw materials and measurements that were made during a data collection \cite{Bakeman_2012}.
-As recordings are often the core component of a data collection, a session in many cases is defined by a single recording.
-In some cases a single session comprises multiple videos, either serially in time representing different segments, or in parallel from multiple cameras recording the same events.
-These primary recordings may also be augmented with other forms of timeseries data such as motion capture, eye tracking, or physiological recordings, as well as with additional point-in-time measurements, observations, or surveys, all of which have specific and critical temporal relationships with the primary data.
-
-When working with multiple files, researchers often combine them into a single representation.
-For example, they may concatenate serial recordings or make a composite video by joining or overlaying frames.
-Some may overlay or combine secondary data, by adding waveform or frequency visualizations below a video, or by overlaying cross-hairs representing object or eye movement.
-Doing so allows for more convenient human analysis of the data in standard playback tools that support single data streams, but often involves losing data either through masking or downsampling.
-
-A session captures relationships among data elements from a single data acquisition, representing files along with their temporal positioning and optional date of collection.
-The metaphor we use here is a timeline: each session defines a time range, with absolute timestamps representing the exact point of data collection, or relative positioning from an arbitrary 0-point.
-Files placed within a session can then be (optionally) positioned, much as in video or audio editing software.
-Although some advanced recording equipment provides automatic timestamping or synchronization features, most researchers who collect parallel data streams use ``sync points'' such as an electronic trigger signal, light flash, or tone to align recordings.
-These sync points can then be used to manually place recordings on the timeline.
-In this way, the session represents recordings and related data streams in a way familiar to researchers, while keeping data in its most raw and flexible form.
-
-\subsection{Records}
-
-A record can represent many things typically conceived of as a row of a database.
-Each may represent: a single data sample, such as a participant, group, or classroom; a level of analysis, such as an experimental trial, task, condition, or location; or some other factor. 
-A record is thus defined by one of the above types along with any number of scalar values, which can be textual, numeric, or date.
-In the case of a record representing an individual participant, these values might include subject identifier, gender, birth date, demographic information, measurements, or other collected variables.
-For other record types, these values may include independent variables, parameters, identifiers, or descriptions. 
-We can thus display the records of a given type as a table, with the values of each record constituting a row.
-
-In addition to containing values, records are used to organize sessions.
-Different researchers employ substantially different directory structures for their files depending on their needs.
-Common approaches involve creating one directory per participant, session date, experimental condition, phase of analysis, and/or age group.
-Sometimes single video files are split into segments according to scene or activity, and researchers consider this property the primary organizational principle.
-
-Rather than trying to handle all of these strategies separately, Databrary leverages records to provide arbitrary groupings.
-Records can be attached to sessions, or, equivalently, sessions can be grouped into records, in a many-to-many relationship.
-For example, a study participant may appear in exactly one session, so the collected information about that participant, added to a record, is attached in a one-to-one relationship with a specific session.
-In other research, multiple participants appear in multiple sessions, such as recordings of the same class of students over a span of time, and so the record may be attached to multiple sessions.
-This similarly extends to any other groupings of sessions based on experimental procedures or other factors by attaching the appropriate record of the corresponding type.
-
-Since researchers can create whichever record types apply to their data, they can also choose to group their data as they wish.
-This allows users to import, browse, or export data in what they consider the most intuitive directory structure by indicating the appropriate nesting of record types.
-By standardizing record types, we can also perform targeted searches across the library, for example for all videos containing participants within a particular age range.
-
-\subsection{Annotations}
-
-Research using recordings typically proceeds with the addition of temporally-positioned annotations, either by direct observation or with the aid of simple visualization or analysis tools.
-Various specialized software tools enable researchers to annotate events, measurements, utterances, or behaviors at any temporal scale, from individual samples/frames to the entire session.
-Most tools focus on a single recording, and there is little interoperability among them.
-Moreover, with rare exceptions in specific research domains \cite{MacWhinney2001}, there are few standards or definitions for the annotations themselves, even across studies performed by the same researcher.
-
-Establishing a timeline for each session allows these sorts of temporally-positioned annotations to be attached.
-Currently we support two types of annotations: structured annotations based on records and added by the data contributors; and simple annotations contributed by the community.
-Eventually we hope to extend both of these to support the full range of annotations performed during analysis, both directly on the web and imported from existing software.
-
-The first type is created using records, which can be attached not only to entire sessions, but also to arbitrary temporal sub-sets of a session.
-If any record relates to a part, scene, or section of a session---for example a participant being present for only part of a recording---it can be applied within the timeline of a session to only that appropriate part.
-This allows an additional type of organizational structure researchers employ, wherein records define virtual sub-sessions that can be grouped and interacted with as units, without needing to cut up source videos.
-This can also theoretically be used to represent annotations at a finer temporal scale, for example individual trials or events.
-However, the appropriate interfaces and tools to input or import these annotations are still being developed.
-
-For the second, we allow users to contribute keyword-based tags and comment-based discussions for any temporal segment of a session as they watch the recordings.
-Such annotations could also be added by automated video analysis according to features of interest identified in the data, using our API.
-In addition to enriching datasets and enabling dialog between researchers, we expect annotations to provide an index target for additional discovery capabilities for finding for particular features within data \cite{Lanagan_Smeaton_2012}.
-We will also evaluate automated approaches for seeding tags to be refined by the community based on existing metadata and analysis \cite{Yang_Lu_Giles_2011,Giles2013}.
-
-\section{Interface and Interoperability}
+\section{Interface}
 
@@
 In this way, Databrary will allow researchers to work with their recordings directly in raw form without having to transfer or transform them, while also capturing more meaningful and standardized contextual information from this process.
-
-There is also active discussion around the role libraries in providing repositories and services for research data \cite{Castelli_etal_2013,Nielson_Hjørland_2014,Macmillan_2014,Pinfield_etal_2014}.
-Although resolving the complicated issues that arise within the library domain is not directly within our purview, we have laid out a functioning system for research data management that may bring established library standards to bear on research data, further enhancing their discoverability.
-Going forward, we are well positioned to provide interoperability with library-based metadata schemas (such as export of data packages cross-walked to Dublin Core or METS) and reach OAI-PMH compliance.
-This will automatically incorporate data that researchers add to Databrary into federated library searches with other domain-specific data repositories.
-By providing a refined API and assigning Digital Object Identifiers (DOIs) to volumes, we can also allow libraries and other information systems to tap into Databrary datasets.
 

### 2015-04-02 jcdl-15/paper_demo.tex
take out more and move things around

@@
 They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools, as human observers can directly glean much of the data's richness.
-% access in the sense of "what format is it in":
 This allows recordings collected in one context to be used by others for different purposes with minimal explanation, naturally reducing the problems of documentation, provenance, formatting, and access that plague many research repositories.
-On the other hand, automated analysis of videos is an active but still immature research area, so the problems of search, discovery, and meta-analysis over these media remain \cite{Albertson_2013, Lanagan_Smeaton_2012}.
 
@@
 We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways, allowing for understanding and discovery.
-
-Research data repositories require a balance between generality and specificity.
-General purpose repositories permit unstructured storage of arbitrary files (e.g., Dataverse\footnote{http://dataverse.org/}, Dryad\footnote{http://datadryad.org/}, PURR\footnote{https://purr.purdue.edu/}), while those with rigorously structured schema require contributions to be carefully formatted (e.g., Human Connectome Project\footnote{http://www.humanconnectomeproject.org/}, Per\-son\-al Ge\-nome Pro\-ject\footnote{http://www.personalgenomes.org/}, TalkBank\footnote{http://talkbank.org/}).
-The flexibility of general-purpose repositories make them especially well-suited to capture and archive entire datasets, but these systems do not facilitate search, understanding, or reuse without extensive, usually labor-intensive curation to summarize or extract metadata \cite{Peer_2012}.
-Targeted repositories, on the other hand, generate large-scale, homogeneous data sources that facilitate meta- and re-analysis, but are limited to specific research questions, losing the full richness, context, and innovation of the source materials and field.
 
@@
 
-\section{???}
+\section{Designated Community}
+
+Focus on a particular research domain allows us to achieve a sustainable balance between standardization and flexibility without burdening contributors.
+The success we've had in engaging, involving, and extracting data from researchers is due largely to our focus on a specific community, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
+By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.
 
@@
 
-\section{Conclusions}
-
-We have built a research data repository focused on video and audio recordings, allowing for normalized presentation and discovery of data across the site, and introducing temporal structure to metadata driven by the nature of these recordings.
-By limiting both file and metadata organization to a flat (but overlapping) set of sessions and records, we capture session-based research data within a cohesive and intuitive user interface.
-Although we allow additional data files, we encourage users with other types of rich data, such as physiological recordings, to store and link to those resources externally.
-Similarly, by targeting reuse of existing research data, rather than replication or meta-analysis, we limit the scope of accepted data to the original recordings and early analysis phases that generate the most value.
-Focus on a particular research domain allows us to achieve a sustainable balance between standardization and flexibility without burdening contributors.
-
-Having made significant progress building ingest and data entry services that have let us grow the library, we now must turn our efforts towards discovery and reuse, finding ways to better describe and elaborate on collected data.
-To do this we look both to the original data contributors for ways that their existing analysis workflows can generate usable content descriptions, and to the community of researchers to label and comment on datasets as they explore and reuse.
-Thus we seek to incorporate these annotation processes into Databrary, both to capture and define these annotations as generalizable search terms, and to make these workflows more convenient by obviating existing file transfer and transformation burdens.
-Ultimately we hope to allow and capture instances of data reuse in this same way, so that content descriptions are aggregated from multiple annotation passes from different users, and existing data can continuously increase in value for researchers.
-
-The success we've had in engaging, involving, and extracting data from researchers is due largely to our focus on a specific community, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
-By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.
-
 \section*{Acknowledgments}

### 2015-04-03 jcdl-15/paper_demo.tex
getting close to a proposed structure and pulling in some stock language for shaping

@@
 We achieve this conformance through in-depth, investigative curation of previously collected datasets, along with user interfaces allowing researchers to upload their own research data as they acquire it.
-In the future, we hope to capture the additional annotation data researchers already produce while analyzing recordings to further improve our discovery capabilities.
 
 \subsection*{Repository Overview}
-
+%might not want to keep this section, 2-pages is pretty short.
 Databrary began accepting contributions in early 2014 and opened for general use in October 2014.
@@
 
-\section{Interface}
-
-\subsection{Normalization}
-
-We have adopted an approach that encourages adherence to existing standardized options but still allows new types to be created as the data require it.
-Initially we have achieved this through manual review and off-line curation of contributions, dynamically extending the system as new types of data were discovered without burdening or constraining researchers.
-However, since we expect the bulk of the data in the library to come from newly collected sources (due largely to privacy constraints on existing datasets), we have additionally taken a distinct, user-driven approach.
-
 While some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools.
@@
 
-\subsection{Discovery}
+\section{Uploading and Managing Data}
+
+The team incorporated insights drawn from observations of data management practices in a sample of laboratories. 
+From them we created a set of data management features that empower researchers to actively curate their own projects – to upload data with accompanying metadata – as each study unfolds. 
+We have designed and implemented a spreadsheet interface (see Figure 1) for entering, editing, and viewing session-level metadata (e.g.
+participants, conditions of study, tasks in the experiment, session access levels, study groups etc.).
+Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality transparent to users.
+We have also implemented a timeline for uploading, viewing, and tagging video assets related to sessions.
+The timeline view is designed to look and operate like video editing software commonly used in many research labs (see Figure 2).
+It allows users to upload video files, position them to reflect the temporal order of each component of data collection, and annotate video sections with user-generated tags. 
+These tags become additional metadata indices for search and discovery. 
+Databrary’s staff continue to refine the active curation features on the basis of user feedback.
+
+\section{Discovery and Scholarly Communication}
 
@@
 Ultimately, determining usefulness requires direct, human observation of the recordings, and so a large part the discovery solution will be simply to present researchers with short, representative samples from a variety of datasets.
-
-Another important source for search targets comes from annotations supplied by the original researchers during analysis, as well as by the community.
-We are currently working to support import and export workflows for existing annotation tools that add meaningful definitions to these often opaque codes, while allowing researchers to continue their current practices.
-Beyond that, we also plan to build a new, more efficient and structured annotation web interface supporting more of the analysis workflow directly within our system.
-In this way, Databrary will allow researchers to work with their recordings directly in raw form without having to transfer or transform them, while also capturing more meaningful and standardized contextual information from this process.
 

### 2015-04-10 jcdl-15/paper_demo.tex
add numbers to sections as mentioned in review

@@
 
-\section{Introduction/Background}
+\section{1. Introduction/Background}
 
@@
 Unfortunately, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
-%something here or somewhere else about aversion (fear) to changing practices...
 Despite their inherent reusability, this lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
@@
 This facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
-We achieve this conformance through in-depth, investigative curation of previously collected datasets, along with user interfaces allowing researchers to upload their own research data as they acquire it.
 
-\subsection*{Repository Overview}
-%might not want to keep this section, 2-pages is pretty short.
+\section{2. Designated Community}
+
+Focus on a particular research domain allows us to achieve a sustainable balance between standardization and flexibility without burdening contributors.
+While some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools.
+Most use a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and local filesystem storage.
+This suggested the opportunity to supply a cohesive data management platform where all collected raw data---including recordings, contextual information, and measurements---could be entered, stored, and exported for later off-line analysis.
+Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
+In turn, this adds convenience and functionality for researchers over existing practices while alleviating storage burdens.
+
+\section{3. Access Policies}
+
+A primary design principle for the development of the Databrary repository is site accessibility. 
+Because video data involving human subjects contains many points of identifiability, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
+User registration also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
+Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of reseach data that has a high disclosure risk.
+Finally, contributors are offerred session release levels and file classifications that allow them the capacity to fine-tune who has access to the data they contribute.
+
+\section{4. Repository Overview}
+%might need to rethink this section, 2-pages is pretty short.
 Databrary began accepting contributions in early 2014 and opened for general use in October 2014.
@@
 
-\section{Designated Community}
+\section{5. Uploading and Managing Data}
 
-Focus on a particular research domain allows us to achieve a sustainable balance between standardization and flexibility without burdening contributors.
-The success we've had in engaging, involving, and extracting data from researchers is due largely to our focus on a specific community, but it has come at the cost of limitations in scope and difficulties of interoperability and discovery that many more traditional repositories avoid.
-By slowly growing our scope and relaxing some of these limitations, we hope to leverage more of the established knowledge of the digital library community in order to make this repository more globally applicable to the storage and accessibility of research data.
-
-While some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools.
-Most use a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and local filesystem storage.
-This suggested the opportunity to supply a cohesive data management platform where all collected raw data---including recordings, contextual information, and measurements---could be entered, stored, and exported for later off-line analysis.
-Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
-In turn, this adds convenience and functionality for researchers over existing practices while alleviating storage burdens.
-
-\section{Uploading and Managing Data}
-
-The team incorporated insights drawn from observations of data management practices in a sample of laboratories. 
+Databrary was designed from insights drawn from observations of data management practices in a sample of laboratories. 
 From them we created a set of data management features that empower researchers to actively curate their own projects – to upload data with accompanying metadata – as each study unfolds. 
@@
 
-\section{Discovery and Scholarly Communication}
+\section{6. Discovery and Scholarly Communication}
 
-While increased normalization facilitates filtering and targeted searches for data, the general problems of browsing and discovery remain largely unsolved.
-This problem is compounded for us because we expect data collected in one research area to be useful for entirely different purposes.
-That is, data arrives labeled with contextual information that may not fully describe the range of features present.
-Ultimately, determining usefulness requires direct, human observation of the recordings, and so a large part the discovery solution will be simply to present researchers with short, representative samples from a variety of datasets.
+As a platform for also sharing and reusing video-based research data, discoverability and features for facilitating the referencing and display of datasets in scholarly communication are an important aspect of the repository.
+Currently basic text search over the entire repository is supported and soon we will begin to implement more refined search that allows faceting of search results based on dataset tags and keywords and session metadata such as participant age group, task, study condition, and location.
+Additionally, we allow for the creation of tags that annotate videos and video segments, which will also serve as an index point for this more refined search functionality.  
+Each dataset comes with a formatted citation, a persistent URI, and soon will also mint and provide a DOI for referencing in scholarly publications and presentations.
+One of the more unique features of Databrary is the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences [figure?]
 
-\section*{Acknowledgments}
+\section*{7. Acknowledgments}
 
@@
 \bibliographystyle{abbrv}
-\bibliography{references}
+\bibliography{8. references}
 

### 2015-04-13 jcdl-15/paper_demo.tex
apparently they are auto-inserted
let's just see how many commits we can push on this
make changes - first draft

@@
 By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
-This demo will take a tour through the Databrary service, hightlighting how it controls user access to restricted data based on human research subjects, how it facilitates browsing and discoverability of datasets, and how it provides a service for researchers to upload and manage their data as they collect it.
+This demo will provide a tour through the Databrary service, hightlighting how it promotes storage, sharing, and data management for research data, controls user access to restricted data based on human research subjects, and how it facilitates browsing and discoverability of datasets.
 \end{abstract}
@@
 
-\section{1. Introduction/Background}
+\section{Introduction/Background}
 
@@
 
-Unfortunately, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
+Unfortunately, researchers rarely share recordings with others \cite{MacWhinney2001}, due largely to entrenched practices and perceived privacy requirements.
 Despite their inherent reusability, this lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
@@
 
-\section{2. Designated Community}
+\section{Designated Community}
 
@@
 
-\section{3. Access Policies}
+\section{Access Policies}
 
-A primary design principle for the development of the Databrary repository is site accessibility. 
+A primary design principle for the development of the Databrary repository is restricting site accessibility to ensure contributors that their data will retain the privacy contraints defined in their studies' release consent language.
 Because video data involving human subjects contains many points of identifiability, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
-User registration also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
-Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of reseach data that has a high disclosure risk.
-Finally, contributors are offerred session release levels and file classifications that allow them the capacity to fine-tune who has access to the data they contribute.
+User registration for gaining access to Databrary also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
+Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of reseach data that has a high disclosure risk\footnote{https://databrary.org/policies/agreement.pdf}.
+Finally, contributors are offerred session release levels and file classifications that allow them the capacity to fine-tune who has access to their data.
 
-\section{4. Repository Overview}
+\section{Repository Overview}
 %might need to rethink this section, 2-pages is pretty short.
@@
 
-Databrary is a new, open-source web application\footnote{http://github.com/databrary/databrary}, built in Scala on the Play Framework\footnote{http://playframework.com/} to support a responsive user interface, a complete API, and high-performance streaming.
+Databrary is an open-source web application\footnote{http://github.com/databrary/databrary}, built in Scala on the Play Framework\footnote{http://playframework.com/} to support a responsive user interface, a complete API, and high-performance streaming.
 Recordings and other files are placed in content-based filesystem storage, and all structured data are stored in a PostgreSQL database, leveraging its geometric indexing capabilities for temporal data.
-All uploaded recordings are automatically transcoded to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
+Databrary stores at least two versions of each item of Databrary video content: a copy for access, and the received original file.
+The access copy is generated by automatically transcoded the uploaded file to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
 This transcoding utilizes the high performance computing cluster at the host campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
@@
 
-\section{5. Uploading and Managing Data}
+\section{Uploading and Managing Data}
 
@@
 We have designed and implemented a spreadsheet interface (see Figure 1) for entering, editing, and viewing session-level metadata (e.g.
-participants, conditions of study, tasks in the experiment, session access levels, study groups etc.).
+participants, conditions of study, tasks in the experiment, session access levels, study groups and condition etc.).
 Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality transparent to users. 
@@
 These tags become additional metadata indices for search and discovery.
-Databrary’s staff continue to refine the active curation features on the basis of user feedback.
 
-\section{6. Discovery and Scholarly Communication}
+\section{Discovery and Scholarly Communication}
 
@@
 
-\section*{7. Acknowledgments}
+\section*{Acknowledgments}
 
@@
 \bibliographystyle{abbrv}
-\bibliography{8. references}
+\bibliography{references}
 

### 2015-04-13 jcdl-15/paper_demo.tex
Minor superficial fixes, no reading yet

@@
 By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
-This demo will provide a tour through the Databrary service, hightlighting how it promotes storage, sharing, and data management for research data, controls user access to restricted data based on human research subjects, and how it facilitates browsing and discoverability of datasets.
+This demo will provide a tour through the Databrary service, highlighting how it promotes storage, sharing, and data management for research data, controls user access to restricted data based on human research subjects, and how it facilitates browsing and discoverability of datasets.
 \end{abstract}
@@
 
-A primary design principle for the development of the Databrary repository is restricting site accessibility to ensure contributors that their data will retain the privacy contraints defined in their studies' release consent language.
+A primary design principle for the development of the Databrary repository is restricting site accessibility to ensure contributors that their data will retain the privacy constraints defined in their studies' release consent language.
 Because video data involving human subjects contains many points of identifiability, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
 User registration for gaining access to Databrary also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
-Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of reseach data that has a high disclosure risk\footnote{https://databrary.org/policies/agreement.pdf}.
-Finally, contributors are offerred session release levels and file classifications that allow them the capacity to fine-tune who has access to their data.
+Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of research data that has a high disclosure risk\footnote{https://databrary.org/policies/agreement.pdf}.
+Finally, contributors are offered session release levels and file classifications that allow them the capacity to fine-tune who has access to their data.
 
@@
 Databrary was designed from insights drawn from observations of data management practices in a sample of laboratories. 
-From them we created a set of data management features that empower researchers to actively curate their own projects – to upload data with accompanying metadata – as each study unfolds. 
-We have designed and implemented a spreadsheet interface (see Figure 1) for entering, editing, and viewing session-level metadata (e.g.
-participants, conditions of study, tasks in the experiment, session access levels, study groups and condition etc.).
+From them we created a set of data management features that empower researchers to actively curate their own projects---to upload data with accompanying metadata---as each study unfolds. 
+We have designed and implemented a spreadsheet interface (see Figure 1) for entering, editing, and viewing session-level metadata (e.g., participants, conditions of study, tasks in the experiment, session access levels, study groups and conditions, etc.).
 Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality transparent to users. 

### 2015-04-13 jcdl-15/paper_demo.tex
More minor edits from read-through; nothing structural

@@
 By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
-This demo will provide a tour through the Databrary service, highlighting how it promotes storage, sharing, and data management for research data, controls user access to restricted data based on human research subjects, and how it facilitates browsing and discoverability of datasets.
+This demo will provide a tour through the Databrary service, highlighting how it promotes storage, sharing, and data management for research data, controls user access to restricted data based on human research subjects, and facilitates browsing and discoverability of datasets.
 \end{abstract}
@@
 
-Unfortunately, researchers rarely share recordings with others \cite{MacWhinney2001}, due largely to entrenched practices and perceived privacy requirements.
-Despite their inherent reusability, this lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
+Unfortunately, despite their inherent reusability, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
+This lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
 To address this problem, the Databrary project\footnote{https://databrary.org/} aims to build tools and policies that facilitate repurposing and reuse of recordings without placing undue burden on data contributors.
-We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways, allowing for understanding and discovery.
+We have created a web-based data library to store and organize recordings and to capture contextual information necessary for reuse in standardized ways.
 
@@
 We designed the system by investigating how researchers studying human development and education organize their own data, deriving from those observations a unifying set of principles for organizing contextual metadata.
-This facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
+Thus it facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
 
@@
 Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
-In turn, this adds convenience and functionality for researchers over existing practices while alleviating storage burdens.
+In turn, this adds convenience and functionality for researchers over existing practices and alleviates storage burdens.
 
@@
 
-A primary design principle for the development of the Databrary repository is restricting site accessibility to ensure contributors that their data will retain the privacy constraints defined in their studies' release consent language.
+A primary design principle for the Databrary repository is restricting data access in order to ensure that contributors' data will retain the privacy constraints defined by their studies' release language.
 Because video data involving human subjects contains many points of identifiability, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
@@
 
-\section{Repository Overview}
+\section{Technical details}
+
 %might need to rethink this section, 2-pages is pretty short.
@@
 Databrary stores at least two versions of each item of Databrary video content: a copy for access, and the received original file.
-The access copy is generated by automatically transcoded the uploaded file to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
+The access copy is generated by automatically transcoding the uploaded file to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
 This transcoding utilizes the high performance computing cluster at the host campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
@@
 Databrary was designed from insights drawn from observations of data management practices in a sample of laboratories. 
-From them we created a set of data management features that empower researchers to actively curate their own projects---to upload data with accompanying metadata---as each study unfolds. 
-We have designed and implemented a spreadsheet interface (see Figure 1) for entering, editing, and viewing session-level metadata (e.g., participants, conditions of study, tasks in the experiment, session access levels, study groups and conditions, etc.).
-Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality transparent to users. 
+From them we created a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds. 
+We have designed and implemented a spreadsheet interface for entering, editing, and viewing session-level metadata (e.g., participants, conditions of study, tasks in the experiment, session access levels, study groups and conditions, etc.).
+Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality intuitive to users. 
 We have also implemented a timeline for uploading, viewing, and tagging video assets related to sessions.
-The timeline view is designed to look and operate like video editing software commonly used in many research labs (see Figure 2).
+The timeline view is designed to look and operate like video editing software commonly used in many research labs.
 It allows users to upload video files, position them to reflect the temporal order of each component of data collection, and annotate video sections with user-generated tags. 
@@
 
-As a platform for also sharing and reusing video-based research data, discoverability and features for facilitating the referencing and display of datasets in scholarly communication are an important aspect of the repository.
-Currently basic text search over the entire repository is supported and soon we will begin to implement more refined search that allows faceting of search results based on dataset tags and keywords and session metadata such as participant age group, task, study condition, and location.
+As a platform for sharing and reusing video-based research data, features for finding, displaying, and referencing datasets from scholarly communication are an important aspect of the repository. 
+Currently basic text search over the entire repository is supported and soon we will begin to implement more refined search that allows faceting of search results based on tags, keywords, and session metadata such as participant age group, task, study condition, and location.
 Additionally, we allow for the creation of tags that annotate videos and video segments, which will also serve as an index point for this more refined search functionality.  
 Each dataset comes with a formatted citation, a persistent URI, and soon will also mint and provide a DOI for referencing in scholarly publications and presentations.
-One of the more unique features of Databrary is the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences [figure?]
+One of the more unique features of Databrary is the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences.
 

### 2015-04-14 jcdl-15/paper_demo.tex
some clean up, some additions.

@@
 Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes.
-Databrary provides a web-based service for sharing and reusing the video data created by researchers in the developmental and learning sciences.
+Though they also present a challenge in that they are often more identifiable than data provided in other formats.
+Databrary is a web-based service for sharing and reusing the video data created by researchers in the developmental and learning sciences.
 By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
@@
 They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools, as human observers can directly glean much of the data's richness.
-This allows recordings collected in one context to be used by others for different purposes with minimal explanation, naturally reducing the problems of documentation, provenance, formatting, and access that plague many research repositories.
+This allows recordings collected in one context to be used by others for different purposes with minimal explanation.
 
-Unfortunately, despite their inherent reusability, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
+While video data has many benefits, it also presents a challenge in being highly identifiable.
+Video often depicts participants’ faces, and their names are often spoken aloud. 
+Because of this, sharing video data presents challenges related to the privacy of human subjects. 
+In contrast to other forms of data which are non-identifiable or can be easily de-identified, video data cannot be easily de-identified without significantly diminishing its value and richness. 
+Therefore, any solution for video sharing must provide appropriate policy and access frameworks to ensure appropriate protection of research participants.
+
+Despite their inherent reusability, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
 This lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
@@
 
-Our approach is to focus on video and audio recordings in particular research domains, and build appropriate structured and unstructured resources around this functional core.
+Our approach is to focus on video and audio recordings in a particular research domain, and build appropriate structured and unstructured resources around this functional core.
 We designed the system by investigating how researchers studying human development and education organize their own data, deriving from those observations a unifying set of principles for organizing contextual metadata.
@@
 
-\section{Designated Community}
-
-Focus on a particular research domain allows us to achieve a sustainable balance between standardization and flexibility without burdening contributors.
-While some researchers in our target domains have standardized data management practices, a large majority have no special data management or collection tools.
-Most use a combination of hand-written paper, simple spreadsheets, existing video annotation tools, statistical analysis software, and local filesystem storage.
-This suggested the opportunity to supply a cohesive data management platform where all collected raw data---including recordings, contextual information, and measurements---could be entered, stored, and exported for later off-line analysis.
-Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
-In turn, this adds convenience and functionality for researchers over existing practices and alleviates storage burdens.
-
-\section{Access Policies}
-
-A primary design principle for the Databrary repository is restricting data access in order to ensure that contributors' data will retain the privacy constraints defined by their studies' release language.
-Because video data involving human subjects contains many points of identifiability, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
-User registration for gaining access to Databrary also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
-Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of research data that has a high disclosure risk\footnote{https://databrary.org/policies/agreement.pdf}.
-Finally, contributors are offered session release levels and file classifications that allow them the capacity to fine-tune who has access to their data.
-
 \section{Technical details}
 
-%might need to rethink this section, 2-pages is pretty short.
+%we should probably update these stats.
 Databrary began accepting contributions in early 2014 and opened for general use in October 2014.
@@
 
-Databrary is an open-source web application\footnote{http://github.com/databrary/databrary}, built in Scala on the Play Framework\footnote{http://playframework.com/} to support a responsive user interface, a complete API, and high-performance streaming.
-Recordings and other files are placed in content-based filesystem storage, and all structured data are stored in a PostgreSQL database, leveraging its geometric indexing capabilities for temporal data.
+Databrary is an open-source web application\footnote{http://github.com/databrary/databrary}, featuring a responsive user interface, a RESTful API, and high-performance video streaming.
 Databrary stores at least two versions of each item of Databrary video content: a copy for access, and the received original file.
 The access copy is generated by automatically transcoding the uploaded file to a standard format to enable cross-platform HTML5-based streaming and downloading for off-line access, currently H.264/AAC in an MPEG-4 container for video.
-This transcoding utilizes the high performance computing cluster at the host campus of New York University, using ffmpeg's libav bindings for both this and direct access to video frames and clips.
-The user interface is built primarily on the Angular web framework\footnote{http://angularjs.org/}, and all data access is performed through an open JSON API.
 
-\section{Uploading and Managing Data}
+\section{Data Management for a Designated Community}
 
-Databrary was designed from insights drawn from observations of data management practices in a sample of laboratories. 
-From them we created a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds. 
+Focusing on a particular research domain allows us to develop a tool that is informed by how the intended community of users currently works, while enhancing that labor and adhering to the concerns and priorities for data that often has a high disclosure risk.
+This has resulted in the creation of a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds.
 We have designed and implemented a spreadsheet interface for entering, editing, and viewing session-level metadata (e.g., participants, conditions of study, tasks in the experiment, session access levels, study groups and conditions, etc.).
@@
 
+Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
+In turn, this adds convenience and functionality for researchers over existing practices and alleviates storage burdens.
+
+Because video data involving human subjects contains many points of identifiability, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
+User registration for gaining access to Databrary also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
+Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of research data involving human subjects\footnote{https://databrary.org/policies/agreement.pdf}.
+Finally, contributors are offered release levels for files and groups of files that allow them to fine-tune who has access to their data.
+
 \section{Discovery and Scholarly Communication}
 
-As a platform for sharing and reusing video-based research data, features for finding, displaying, and referencing datasets from scholarly communication are an important aspect of the repository. 
+Though Databrary also strongly encourages sharing and reuse as a complement to data storage and management.
+As such, we have put a priority on creating an interface that allows other researchers to find, filter, and download video files they might be able to use in their own research or classroom teaching. 
 Currently basic text search over the entire repository is supported and soon we will begin to implement more refined search that allows faceting of search results based on tags, keywords, and session metadata such as participant age group, task, study condition, and location.
@@
 Each dataset comes with a formatted citation, a persistent URI, and soon will also mint and provide a DOI for referencing in scholarly publications and presentations.
-One of the more unique features of Databrary is the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences.
+Based on input from site users we have also built in the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences.
+As we continue to hear feedback from our intended user base, we not only see Databrary as a tool that facilitates the management of research data, but also as an opportunity to enhance the ability for academics to communicate about their research in all of its richness.
 

### 2015-04-14 jcdl-15/paper_demo.tex
some meager changes

@@
 
-Though Databrary also strongly encourages sharing and reuse as a complement to data storage and management.
-As such, we have put a priority on creating an interface that allows other researchers to find, filter, and download video files they might be able to use in their own research or classroom teaching. 
+Though central to Databrary's mission is not merely the storage and management of research data, but making that data discoverable by other researchers and facilitating the reuse of that data in future studies and in scholarly communication as a citeable resource as well as an aid to educational presentation.
 Currently basic text search over the entire repository is supported and soon we will begin to implement more refined search that allows faceting of search results based on tags, keywords, and session metadata such as participant age group, task, study condition, and location.
-Additionally, we allow for the creation of tags that annotate videos and video segments, which will also serve as an index point for this more refined search functionality.  
-Each dataset comes with a formatted citation, a persistent URI, and soon will also mint and provide a DOI for referencing in scholarly publications and presentations.
+Additionally, Databrary allows for the annotatation of videos and video segments with tags, which will also serve as an index point for this more refined search functionality.
+
+To aid in using datasets as a bibliographic source, each dataset comes with a formatted citation, a persistent URI, and soon will also be assigned a Digital Object Identifier (DOI) for referencing in scholarly publications and presentations.
 Based on input from site users we have also built in the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences.
-As we continue to hear feedback from our intended user base, we not only see Databrary as a tool that facilitates the management of research data, but also as an opportunity to enhance the ability for academics to communicate about their research in all of its richness.
+As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics communicate about their research in all of its richness.
 

### 2015-04-15 jcdl-15/paper_demo.tex
update stats and kind of sappy ending that needs reworking

@@
 
-%we should probably update these stats.
 Databrary began accepting contributions in early 2014 and opened for general use in October 2014.
-It currently hosts 5,700 video files totaling 1,600 hours of recordings along with 2,200 additional files.
-These files make up 2,400 sessions and are covered by 1,300 metadata records (including 1,100 individual participants).
+As of April 2015, it hosts 7,340 video files totaling 2,409 hours of recordings along with 2,647 additional files.
+These files make up 3,108 study sessions and are covered by 2,000 metadata records (including 1,804 individual participants).
 Data originates from 35 individual contributors across 25 different universities.
@@
 
-Focusing on a particular research domain allows us to develop a tool that is informed by how the intended community of users currently works, while enhancing that labor and adhering to the concerns and priorities for data that often has a high disclosure risk.
+Focusing on a particular research domain allows us to develop a tool that is informed by how the intended community of users currently works, while enhancing that labor as well as accomodating the the privacy concerns for data that often has a high disclosure risk.
 This has resulted in the creation of a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds.
-We have designed and implemented a spreadsheet interface for entering, editing, and viewing session-level metadata (e.g., participants, conditions of study, tasks in the experiment, session access levels, study groups and conditions, etc.).
+To this end, we have designed and implemented a spreadsheet interface for entering, editing, and viewing session-level metadata (e.g., participants, conditions of study, tasks in the experiment, session access levels, study groups and conditions, etc.).
 Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality intuitive to users. 
@@
 
-Providing a targeted user interface to meet the existing and evolving needs of our contributed data allows us to control the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
+Providing a targeted user interface to meet the existing and evolving needs of our user base allows us to control and standardize the means of data entry and thus achieve a greater amount of normalization in the data from the outset, through liberal use of visualization, auto-completion, and suggestions.
 In turn, this adds convenience and functionality for researchers over existing practices and alleviates storage burdens.
 
-Because video data involving human subjects contains many points of identifiability, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
+In addition, because video data involving human subjects is more highly indentifiable, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
 User registration for gaining access to Databrary also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
@@
 Based on input from site users we have also built in the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences.
+
 As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics communicate about their research in all of its richness.
+We already addressed the basic upload, data management, and storage needs of our target community, and have sufficiently answered the question of how to protect sensitive video data while providing wider opportunities to share that data.
+Going forward, we hope to focus more on developing the tools for searching, discovering, and reusing that data, allowing researchers to work with their data in ways they had not previously imagined.
 

### 2015-04-15 jcdl-15/paper_demo.tex
some changes

@@
 
-Focusing on a particular research domain allows us to develop a tool that is informed by how the intended community of users currently works, while enhancing that labor as well as accomodating the the privacy concerns for data that often has a high disclosure risk.
+Focusing on a particular research domain allows us to develop a tool that is informed by how the intended community of users currently works, while enhancing that labor as well as accomodating the specific privacy concerns for data that often has a high disclosure risk.
 This has resulted in the creation of a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds.
@@
 
-\section{Discovery and Scholarly Communication}
+\section{Discovery, Reuse, and Next Steps}
 
@@
 
-As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics communicate about their research in all of its richness.
+As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics communicate about and augment their research.
 We already addressed the basic upload, data management, and storage needs of our target community, and have sufficiently answered the question of how to protect sensitive video data while providing wider opportunities to share that data.

### 2015-04-16 jcdl-15/demo.tex
Rename paper_demo to demo
Pass minor editing

--- 2015-04-15 jcdl-15/paper_demo.tex
@@
 Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences.
-Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes.
-Though they also present a challenge in that they are often more identifiable than data provided in other formats.
+Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes, but also present challenges around respecting the privacy of individuals depicted.
 Databrary is a web-based service for sharing and reusing the video data created by researchers in the developmental and learning sciences.
 By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
-This demo will provide a tour through the Databrary service, highlighting how it promotes storage, sharing, and data management for research data, controls user access to restricted data based on human research subjects, and facilitates browsing and discoverability of datasets.
+This demo will provide a tour through the Databrary service, highlighting how it promotes storage, sharing, and management for research data, controls user access to restricted human subject data, and facilitates browsing and discoverability of datasets.
 \end{abstract}
@@
 
-\section{Introduction/Background}
+\section{Introduction}
 
 Video and audio recordings serve as a primary data source for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
-They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools, as human observers can directly glean much of the data's richness.
+They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools as human observers can directly glean much of the data's richness.
 This allows recordings collected in one context to be used by others for different purposes with minimal explanation.
 
-While video data has many benefits, it also presents a challenge in being highly identifiable.
-Video often depicts participants’ faces, and their names are often spoken aloud. 
+While video data have many benefits, they also contain highly identifiable aspects of depicted human participants: recordings contain participants' faces and voices, and their names are often spoken aloud.
 Because of this, sharing video data presents challenges related to the privacy of human subjects.
-In contrast to other forms of data which are non-identifiable or can be easily de-identified, video data cannot be easily de-identified without significantly diminishing its value and richness. 
-Therefore, any solution for video sharing must provide appropriate policy and access frameworks to ensure appropriate protection of research participants.
+In contrast to other forms of data that are non-identifiable or can be easily de-identified, video data cannot be easily de-identified without significantly diminishing its value and richness. 
+Therefore, any solution for video sharing must provide a policy and access framework to ensure appropriate protection of research participants.
 
 Despite their inherent reusability, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
-This lack of sharing inhibits the reuse of recordings and impedes scientific discovery.
+This lack of sharing prevents extracting the full value from recordings and impedes scientific discovery.
 To address this problem, the Databrary project\footnote{https://databrary.org/} aims to build tools and policies that facilitate repurposing and reuse of recordings without placing undue burden on data contributors.
@@
 
-Our approach is to focus on video and audio recordings in a particular research domain, and build appropriate structured and unstructured resources around this functional core.
+Our approach is to focus on video and audio recordings in a particular research domain, and build appropriate structured and unstructured resources around this core.
 We designed the system by investigating how researchers studying human development and education organize their own data, deriving from those observations a unifying set of principles for organizing contextual metadata.
-Thus it facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
+Thus, it facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
 
-\section{Technical details}
+\section{Repository Details}
 
@@
 These files make up 3,108 study sessions and are covered by 2,000 metadata records (including 1,804 individual participants).
-Data originates from 35 individual contributors across 25 different universities.
+Data originate from 35 individual contributors across 25 different universities.
 
-Databrary is an open-source web application\footnote{http://github.com/databrary/databrary}, featuring a responsive user interface, a RESTful API, and high-performance video streaming.
+Databrary is an open-source web application\footnote{http://github.com/databrary/databrary}, featuring a responsive user interface, a JSON API, and high-performance video streaming.
 Databrary stores at least two versions of each item of Databrary video content: a copy for access, and the received original file.
@@
 
-\section{Data Management for a Designated Community}
+\section{Data Management}
 
-Focusing on a particular research domain allows us to develop a tool that is informed by how the intended community of users currently works, while enhancing that labor as well as accomodating the specific privacy concerns for data that often has a high disclosure risk.
-This has resulted in the creation of a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds.
-To this end, we have designed and implemented a spreadsheet interface for entering, editing, and viewing session-level metadata (e.g., participants, conditions of study, tasks in the experiment, session access levels, study groups and conditions, etc.).
+Focusing on a particular research domain allows us to develop tools informed by how the intended community of users currently works, while accommodating the specific privacy concerns for data that often has a high disclosure risk.
+We have created of a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds. 
+These features include a spreadsheet interface for entering, editing, and viewing session-level metadata: participant demographics, conditions of study, tasks in the experiment, permission levels, study groups and conditions, etc.
 Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality intuitive to users. 
@@
 It allows users to upload video files, position them to reflect the temporal order of each component of data collection, and annotate video sections with user-generated tags.
-These tags become additional metadata indices for search and discovery.
 
@@
 
-In addition, because video data involving human subjects is more highly indentifiable, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the original participant consents under which the data were collected.
-User registration for gaining access to Databrary also involves an authorization process by which site users are verified to be faculty, student, or researcher affiliated with an educational institution.
-Users are also required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of research data involving human subjects\footnote{https://databrary.org/policies/agreement.pdf}.
-Finally, contributors are offered release levels for files and groups of files that allow them to fine-tune who has access to their data.
+In addition, because video data involving human subjects is identifiable, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the participant consents under which the data were collected.
+User registration for gaining access to Databrary involves an authorization process by which site users are verified to be faculty, students, or researchers affiliated with an educational institution.
+Users are required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of research data involving human subjects\footnote{http://databrary.org/access/policies/agreement.html}.
+Finally, contributors are offered permission levels for files and groups of files that allow them to fine-tune who has access to their data.
 
-\section{Discovery, Reuse, and Next Steps}
+\section{Discovery, Reuse, and Beyond}
 
-Though central to Databrary's mission is not merely the storage and management of research data, but making that data discoverable by other researchers and facilitating the reuse of that data in future studies and in scholarly communication as a citeable resource as well as an aid to educational presentation.
-Currently basic text search over the entire repository is supported and soon we will begin to implement more refined search that allows faceting of search results based on tags, keywords, and session metadata such as participant age group, task, study condition, and location.
-Additionally, Databrary allows for the annotatation of videos and video segments with tags, which will also serve as an index point for this more refined search functionality.
+Databrary's mission is not merely the storage and management of research data, but also to make these data discoverable by other researchers and to facilitate their reuse.
+Currently we support text searching over the entire repository, and we plan to continue implement more refined searches that allows faceting of search results based on tags, keywords, and session metadata such as participant age, task, study condition, and context.
+Additionally, Databrary allows for the annotation of videos and video segments with tags, which can serve as an index point for more refined search functionality.
 
+We aim to allow data to be reused in future studies, cited in scholarly communication, as well as used an aid in scientific and educational presentations.
 To aid in using datasets as a bibliographic source, each dataset comes with a formatted citation, a persistent URI, and soon will also be assigned a Digital Object Identifier (DOI) for referencing in scholarly publications and presentations.
@@
 As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics communicate about and augment their research.
-We already addressed the basic upload, data management, and storage needs of our target community, and have sufficiently answered the question of how to protect sensitive video data while providing wider opportunities to share that data.
-Going forward, we hope to focus more on developing the tools for searching, discovering, and reusing that data, allowing researchers to work with their data in ways they had not previously imagined.
+We have already addressed the basic upload, data management, and storage needs of our target community, and have sufficiently answered the question of how to protect sensitive video data while providing wider opportunities to share those data.
+Going forward, we hope to focus more on developing the tools for searching, discovering, and reusing those data, allowing researchers to work with their data in ways they had not previously imagined.
 

### 2015-04-16 jcdl-15/demo.tex
minor edits, think it's ready after a few last passes

@@
 Databrary began accepting contributions in early 2014 and opened for general use in October 2014.
-As of April 2015, it hosts 7,340 video files totaling 2,409 hours of recordings along with 2,647 additional files.
-These files make up 3,108 study sessions and are covered by 2,000 metadata records (including 1,804 individual participants).
+As of April 2015, it hosts 7,300 video files totaling 2,400 hours of recordings along with 2,600 additional files.
+These files make up 3,100 study sessions and are covered by 2,000 metadata records (including 1,800 individual participants).
 Data originate from 35 individual contributors across 25 different universities.
@@
 Focusing on a particular research domain allows us to develop tools informed by how the intended community of users currently works, while accommodating the specific privacy concerns for data that often has a high disclosure risk.
-We have created of a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds. 
+We have created a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds. 
 These features include a spreadsheet interface for entering, editing, and viewing session-level metadata: participant demographics, conditions of study, tasks in the experiment, permission levels, study groups and conditions, etc.
@@
 
-In addition, because video data involving human subjects is identifiable, a core challenge to storing and sharing this type of research data is creating the access checks necessary to honor the participant consents under which the data were collected.
+In addition, because video data involving human subjects is identifiable, we have built the access checks necessary to honor the participant consents under which the data were collected. 
 User registration for gaining access to Databrary involves an authorization process by which site users are verified to be faculty, students, or researchers affiliated with an educational institution.
@@
 
-\section{Discovery, Reuse, and Beyond}
+\section{Discovery and Reuse}
 
 Databrary's mission is not merely the storage and management of research data, but also to make these data discoverable by other researchers and to facilitate their reuse.
-Currently we support text searching over the entire repository, and we plan to continue implement more refined searches that allows faceting of search results based on tags, keywords, and session metadata such as participant age, task, study condition, and context.
+Currently we support text searching over the entire repository, and we plan to continue to implement more refined searches that allows faceting of search results based on session metadata such as participant details, tasks involved in the experiment, study conditions, and context.
 Additionally, Databrary allows for the annotation of videos and video segments with tags, which can serve as an index point for more refined search functionality.
 
-We aim to allow data to be reused in future studies, cited in scholarly communication, as well as used an aid in scientific and educational presentations.
-To aid in using datasets as a bibliographic source, each dataset comes with a formatted citation, a persistent URI, and soon will also be assigned a Digital Object Identifier (DOI) for referencing in scholarly publications and presentations.
+We aim to allow data to be reused in future studies, cited in scholarly communication, as well as used to aid in scientific and educational presentations.
+To promote using datasets as a bibliographic source, each dataset comes with a formatted citation, a persistent URI, and soon will also be assigned a Digital Object Identifier (DOI) for referencing in scholarly publications and presentations.
 Based on input from site users we have also built in the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences.
 
-As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics communicate about and augment their research.
+As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics to communicate about and augment their research.
 We have already addressed the basic upload, data management, and storage needs of our target community, and have sufficiently answered the question of how to protect sensitive video data while providing wider opportunities to share those data.

### 2015-04-16 jcdl-15/demo.tex
Push introduction title onto next column; remove references
adding some text

@@
 
+\break
 \section{Introduction}
@@
 We have created a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds. 
-These features include a spreadsheet interface for entering, editing, and viewing session-level metadata: participant demographics, conditions of study, tasks in the experiment, permission levels, study groups and conditions, etc.
-Most researchers use desktop spreadsheets for precisely this purpose in their own labs, making the interface and functionality intuitive to users. 
+These features include a spreadsheet interface for entering, editing, and viewing the metadata for all sessions related to a study.
+Session-level metadata might participant details, conditions of study, tasks in the experiment, permission levels, study groups and conditions, and serves the purpose of providing ways to browse and search for specific video data in Databrary.
+We have found that most researchers use desktop spreadsheets for managind data in their own labs, making the interface and functionality intuitive to users. 
 We have also implemented a timeline for uploading, viewing, and tagging video assets related to sessions.
 The timeline view is designed to look and operate like video editing software commonly used in many research labs.
-It allows users to upload video files, position them to reflect the temporal order of each component of data collection, and annotate video sections with user-generated tags.
+A common practice for researchers in this domain who make use of video data is to have multiple, parallel camera views, or a sequence of recordings documenting the same study session.
+The timeline view then allows users to upload video files, position them to reflect the temporal order of each video collected, and annotate video sections with user-generated tags.
 
@@
 To promote using datasets as a bibliographic source, each dataset comes with a formatted citation, a persistent URI, and soon will also be assigned a Digital Object Identifier (DOI) for referencing in scholarly publications and presentations.
-Based on input from site users we have also built in the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset or be used in the classroom as an education tool or at professional conferences.
+Based on input from site users we have also built in the ability to excerpt or highlight clips from videos in such a way that they might serve as an exemplar clip for the dataset---providing users unfamiliar with a dataset a quick summary of what is contained therein---or be used in the classroom as an education tool or at professional conferences.
 
@@
 
-\bibliographystyle{abbrv}
-\bibliography{references}
-
 \end{document}

### 2015-04-16 jcdl-15/demo.tex
Minor edits, fix typos

@@
 These features include a spreadsheet interface for entering, editing, and viewing the metadata for all sessions related to a study.
-Session-level metadata might participant details, conditions of study, tasks in the experiment, permission levels, study groups and conditions, and serves the purpose of providing ways to browse and search for specific video data in Databrary.
-We have found that most researchers use desktop spreadsheets for managind data in their own labs, making the interface and functionality intuitive to users. 
+Session-level metadata might include participant details, conditions of study, tasks in the experiment, permission levels, study groups, or conditions, and serves the purpose of providing ways to browse and search for specific video data in Databrary.
+We have found that most researchers use desktop spreadsheets for managing data in their own labs, making the interface and functionality intuitive to users.
+
 We have also implemented a timeline for uploading, viewing, and tagging video assets related to sessions.

### 2015-04-16 jcdl-15/demo.tex
some words

@@
 
-Focusing on a particular research domain allows us to develop tools informed by how the intended community of users currently works, while accommodating the specific privacy concerns for data that often has a high disclosure risk.
+Focusing on a particular research domain allows us to develop tools informed by how the intended community of users currently works, while accommodating the specific privacy concerns for data that often have a high disclosure risk.
 We have created a set of data management features that empower researchers to actively manage their own projects---to upload data with accompanying metadata---as each study unfolds. 
@@
 Users are required to review and sign an access agreement stating that they will treat the data they are able to access through Databrary in a manner that complies with the ethical handling of research data involving human subjects\footnote{http://databrary.org/access/policies/agreement.html}.
-Finally, contributors are offered permission levels for files and groups of files that allow them to fine-tune who has access to their data.
+Finally, contributors are offered the ability to set permission levels for files and groups of files that allow them to fine-tune who has access to their data.
 

### 2015-04-21 vss-15/poster.tex
vss-15 poster draft

\documentclass[final]{beamer}
\usetheme{databrary}
\usepackage[orientation=landscape,size=a0,scale=1.4,debug]{beamerposter}
\usepackage[absolute,overlay]{textpos}
\usepackage{graphics}

\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}

\title{Sharing displays and data from vision science research with Databrary}
\author{R.O. Gilmore, K.E. Adolph, D.S. Millman, L. Steiger, and D.A. Simon}
\footer{NSF BCS-1238599, NICHD U01-HD-076595, \& SRCD}
\date{}

\begin{document}
\begin{frame}{}

\begin{columns}[t]
	\column{.3\textwidth}
		\begin{block}{Motivation}
			The open sharing of data and workflows from scientific research has gained increasing attention in many fields and has become a priority for research funders. 
			Databrary (databrary.org) is a digital data library specialized for storing and sharing video and research data and metadata. Vision scientists often present temporally varying visual displays to human or non-human animal observers and may collect gaze position, kinematic data, or other physiological recordings simultaneously. 
			Some vision scientists collect identifiable video recordings of participants for use in off-line tagging or coding. 
			Databrary may be especially well suited for these use cases because of the library's focus on uploading, organizing, storing, and sharing research data and metadata collected in the course of behavioral resesearch. 
			Databrary has also developed policies that enable identifiable information to be shared with the permission of research participants. 
		\end{block}
		\begin{block}{Privacy \& Security}
			Videos may contain identifiable information such as faces, voices, and images of places. 
			Databrary ensures that identifiable information can be shared in two ways:
			\begin{itemize}
			\item Restricting Access:
				\begin{indent}
					Only those with independent researcher status at an institution may apply for authorization. 
					Applicants promise to uphold ethical principles and sign an access agreement. The institution co-signs the agreement.
				\end{indent}
			\item Securing Permission to Share:
				\begin{indent}
					Identifiable data may only be shared if participants have given permisssion.
					Databrary has developed a release template whose language may be adapted for specific research protocols (https://databrary.org/access/policies.html).
				\end{indent}
			\end{itemize}
		\end{block}
		\begin{block}{What Can Be Shared}
			Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
			Video and audio data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4. Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
		\end{block}
		\begin{block}{Technology Stack}
			Databary is built on a PostreSQL database using the Scala Play framework and JavaScript.
			Data are preserved indefinitely in a secure data storage facility at NYU managed by the university's central information technologies organization. Central IT staff handle storage, network, and backup systems. NYU does routine tape backups that are stored off site. There is no cost to use the system; an institutional subscription model is under development.
		\end{block}
	\column{.3\textwidth}
		\begin{block}{Databrary.org}
			\begin{center}
				\includegraphics[width=\textwidth]{img/databrary-splash.png}
			\end{center}
			Databrary's splash page shows a timeline of recent activity, a set of clickable tags that link to studies with those search terms, a featured dataset, and links that enable users to search for data or manage their own data.
		\end{block}
		\begin{block}{Each Study/Dataset Has Its Own Page}
			\begin{center}
				\includegraphics[width=\textwidth]{img/volume-31.png}
			\end{center}
		\end{block}
	\column{.3\textwidth}
		\begin{block}{Data Management for Ongoing Projects}
			\begin{center}
				\includegraphics[width=\textwidth]{img/spreadsheet.png}
			\end{center}
			Users may upload video and related data \emph{as they collect it}, recording information about individual data collection sessions such as participant characteristics, sharing permission levels, and study conditions. 
		\end{block}
		\begin{block}{Organize, View. Tag Data in Timeline}
			\begin{center}
				\includegraphics[width=\textwidth]{img/timeline.png}
			\end{center}
			Within a session, data are represented on a timeline that reflects when events occurred. This is particularlly useful for studies with multiple camera views (e.g., eye-tracking) or with combined behavioral and image/display streams.
		\end{block}
\end{columns}

\end{frame}
\end{document}

### 2015-04-22 jcdl-15/demo.tex
add permissions to front page

@@
-\documentclass{sig-alternate}
+\documentclass{sig-alternate-2013}
+
+\newfont{\mycrnotice}{ptmr8t at 7pt}
+\newfont{\myconfname}{ptmri8t at 7pt}
+\let\crnotice\mycrnotice%
+\let\confname\myconfname%
+
+\permission{Permission to make digital or hard copies of part or all of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+Copyrights for third-party components of this work must be honored.
+For all other uses, contact the Owner/Author.
+Copyright is held by the owner/author(s).}
+\conferenceinfo{JCDL'15,}{June 21--25, 2015, Knoxville, Tennessee, USA.}
+\copyrightetc{ACM \the\acmcopyr}
+\crdata{978-1-4503-3594-2/15/06. \\
+http://dx.doi.org/10.1145/2756406.2756951}
+
+\clubpenalty=10000
+\widowpenalty = 10000
 

### 2015-04-23 jcdl-15/demo.tex
final pass from me - looks good!

@@
 Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences.
-Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes, but also present challenges around respecting the privacy of individuals depicted.
+Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes, but also present challenges due around respecting the privacy of individuals depicted.
 Databrary is a web-based service for sharing and reusing the video data created by researchers in the developmental and learning sciences.
 By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
-This demo will provide a tour through the Databrary service, highlighting how it promotes storage, sharing, and management for research data, controls user access to restricted human subject data, and facilitates browsing and discoverability of datasets.
+This demo will provide a tour through the Databrary service, highlighting how it promotes storage, management, sharing, and reuse of research data, controls access privileges to restricted human subject data, and facilitates browsing and discoverability of datasets.
 \end{abstract}
@@
 Video and audio recordings serve as a primary data source for scientific research in psychology, linguistics, education, anthropology, and many other disciplines.
-They are, in contrast to most timeseries data, human-consumable and largely self-documenting, and are often analyzed without the use of sophisticated tools as human observers can directly glean much of the data's richness.
+They are, in contrast to other timeseries data, human-consumable and largely self-documenting. 
 This allows recordings collected in one context to be used by others for different purposes with minimal explanation.
+Moreover, video and audio recordings are often analyzed without the use of sophisticated tools as human observers can directly glean much of the data's richness.
 
@@
 Despite their inherent reusability, researchers rarely share recordings with others, due largely to entrenched practices and perceived privacy requirements.
-This lack of sharing prevents extracting the full value from recordings and impedes scientific discovery.
+This lack of sharing prevents others from extracting the full value out of recordings and impedes scientific discovery.
 To address this problem, the Databrary project\footnote{https://databrary.org/} aims to build tools and policies that facilitate repurposing and reuse of recordings without placing undue burden on data contributors.
@@
 We designed the system by investigating how researchers studying human development and education organize their own data, deriving from those observations a unifying set of principles for organizing contextual metadata.
-Thus, it facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings at an unprecedented scale.
+We aimed to design a system that facilitates the capture, standardization, discovery, and understanding of data to enable reuse of recordings for novel purposes at an unprecedented scale.
 
@@
 These files make up 3,100 study sessions and are covered by 2,000 metadata records (including 1,800 individual participants).
-Data originate from 35 individual contributors across 25 different universities.
+Data originate from 35 individual contributors across 25 different universities. 50 universities have granted access to the repository to over 95 researchers.
 
@@
 
-As we continue to gather feedback from our intended user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics to communicate about and augment their research.
+As we continue to gather feedback from our user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics to communicate about and enhance their own research.
 We have already addressed the basic upload, data management, and storage needs of our target community, and have sufficiently answered the question of how to protect sensitive video data while providing wider opportunities to share those data.

### 2015-04-23 jcdl-15/demo.tex
few changes for submission

@@
 Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences.
-Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes, but also present challenges due around respecting the privacy of individuals depicted.
+Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes, but also present challenges related to respecting the privacy of individuals depicted.
 Databrary is a web-based service for sharing and reusing the video data created by researchers in the developmental and learning sciences.
@@
 These files make up 3,100 study sessions and are covered by 2,000 metadata records (including 1,800 individual participants).
-Data originate from 35 individual contributors across 25 different universities. 50 universities have granted access to the repository to over 95 researchers.
+Data originate from 35 individual contributors across 25 different universities. Fifty universities have granted access to the repository to over 95 researchers.
 
@@
 These features include a spreadsheet interface for entering, editing, and viewing the metadata for all sessions related to a study.
-Session-level metadata might include participant details, conditions of study, tasks in the experiment, permission levels, study groups, or conditions, and serves the purpose of providing ways to browse and search for specific video data in Databrary.
+Session-level metadata might include participant details, tasks in the experiment, permission levels, study groups, or conditions, and serves the purpose of providing ways to browse and search for specific video data in Databrary.
 We have found that most researchers use desktop spreadsheets for managing data in their own labs, making the interface and functionality intuitive to users.
@@
 
-As we continue to gather feedback from our user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics to communicate about and enhance their own research.
+As we continue to gather feedback from our user base, we not only see Databrary as a tool that facilitates the standardized management of research data, but also as an opportunity to enhance the ability for academics to communicate about and bolster their own research.
 We have already addressed the basic upload, data management, and storage needs of our target community, and have sufficiently answered the question of how to protect sensitive video data while providing wider opportunities to share those data.

### 2015-04-26 vss-15/poster_landscape.tex
change to baposter

\documentclass[landscape,final,a0paper,fontscale=0.285]{baposter}

\usepackage{calc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{relsize}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{bm}
\usepackage{url}

\usepackage{graphicx}
\usepackage{multicol}

%\usepackage{times}
%\usepackage{helvet}
%\usepackage{bookman}
\usepackage{palatino}

\newcommand{\captionfont}{\footnotesize}

\graphicspath{{images/}{../images/}}
\usetikzlibrary{calc}

\newcommand{\SET}[1]  {\ensuremath{\mathcal{#1}}}
\newcommand{\MAT}[1]  {\ensuremath{\boldsymbol{#1}}}
\newcommand{\VEC}[1]  {\ensuremath{\boldsymbol{#1}}}
\newcommand{\Video}{\SET{V}}
\newcommand{\video}{\VEC{f}}
\newcommand{\track}{x}
\newcommand{\Track}{\SET T}
\newcommand{\LMs}{\SET L}
\newcommand{\lm}{l}
\newcommand{\PosE}{\SET P}
\newcommand{\posE}{\VEC p}
\newcommand{\negE}{\VEC n}
\newcommand{\NegE}{\SET N}
\newcommand{\Occluded}{\SET O}
\newcommand{\occluded}{o}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Multicol Settings
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setlength{\columnsep}{1.5em}
\setlength{\columnseprule}{0mm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Begin of Document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Here starts the poster
%%%---------------------------------------------------------------------------
%%% Format it to your taste with the options
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Define some colors

%\definecolor{lightblue}{cmyk}{0.83,0.24,0,0.12}
\definecolor{lightblue}{rgb}{0.145,0.6666,1}

%%
\begin{poster}%
  % Poster Options
  {
  % Show grid to help with alignment
  grid=false,
  % Column spacing
  colspacing=1em,
  columns=3,
  % Color style
  bgColorOne=white,
  bgColorTwo=white,
  borderColor=lightblue,
  headerColorOne=black,
  headerColorTwo=lightblue,
  headerFontColor=white,
  boxColorOne=white,
  boxColorTwo=lightblue,
  % Format of textbox
  textborder=roundedleft,
  % Format of text header
  eyecatcher=true,
  headerborder=closed,
  headerheight=0.1\textheight,
%  textfont=\sc, An example of changing the text font
  headershape=roundedright,
  headershade=shadelr,
  headerfont=\Large\bf\textsc, %Sans Serif
  textfont={\setlength{\parindent}{1.5em}},
  boxshade=plain,
%  background=shade-tb,
  background=plain,
  linewidth=2pt
  }
  % Eye Catcher
  {
    \includegraphics[height=5em]{img/datavyu-leaf-large.png}
  } 
  %Title
  {\bf{Sharing displays and data from vision science research with Databrary} }
  % Authors
  {R.O. Gilmore, K.E. Adolph, D.S. Millman, L. Steiger, and D.A. Simon \\
   \small{Vision Sciences Society 2015 -- Poster 26.4031 -- \emph{rick@databrary.org}}}
  % University logo
  {
    \includegraphics[height=5em]{img/databrary-leaf-large.png}
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Now define the boxes that make up the poster
%%%---------------------------------------------------------------------------
%%% Each box has a name and can be placed absolutely or relatively.
%%% The only inconvenience is that you can only specify a relative position 
%%% towards an already declared box. So if you have a box attached to the 
%%% bottom, one to the top and a third one which should be in between, you 
%%% have to specify the top and bottom boxes before you specify the middle 
%%% box.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Problem}{name=problem,column=0,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      The open sharing of data and workflows from scientific research has gained increasing attention in many fields and has become a priority for research funders. 
      Databrary (databrary.org) is a digital data library specialized for storing and sharing video and research data and metadata. Vision scientists often present temporally varying visual displays to human or non-human animal observers and may collect gaze position, kinematic data, or other physiological recordings simultaneously. 
      Some vision scientists collect identifiable video recordings of participants for use in off-line tagging or coding. 
      Databrary may be especially well suited for these use cases because of the library's focus on uploading, organizing, storing, and sharing research data and metadata collected in the course of behavioral resesearch.
      Databrary has developed policies that enable identifiable information to be shared with the permission of research participants.
 }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Privacy \& Security}{name=privacy,column=0,row=1, below=problem}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      Videos may contain identifiable information such as faces, voices, and images of places. 
      Databrary ensures that identifiable information can be shared in two ways:
      \begin{itemize}
      \item{\bf{Restricting Access}}

      Only those with independent researcher status at an institution may apply for authorization. 
      Applicants promise to uphold ethical principles and sign an access agreement co-signed by the institution.

      \item{\bf{Securing Permission to Share}}

      Identifiable data may only be shared if participants have given permisssion.
      The Databrary release template may be adapted for specific research protocols.

      \end{itemize}
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{What Can Be Shared}{name=shared,column=1,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
    Video and audio data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4. Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Technology Stack}{name=stack,column=0,below=privacy}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    Databary is built on a PostreSQL database using the Scala Play framework and JavaScript.
    Data are preserved indefinitely in a secure data storage facility at NYU managed by central IT. 
    There is no cost to use the system; an institutional subscription model is under development.
    All code is hosted on GitHub.
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Databrary.org}{name=splash,column=1,span=1,row=0}{
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{center}
    \includegraphics[width=\textwidth]{img/databrary-splash.png}
  \end{center}
  % Databrary's splash page shows a timeline of recent activity, a set of clickable tags that link to studies with those search terms, a featured dataset, and links that enable users to search for data or manage their own data.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Volume/Dataset Page}{name=volume,column=2,span=1,row=0,bottomaligned=splash}{
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{center}
  \includegraphics[width=\textwidth]{img/volume-31.png}
  \end{center}
  Each dataset or study has a page with summary, owner, funding, citation information, materials, and links to publications.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Acknowledgments}{name=thanks,column=0,span=2,above=bottom}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  Databrary is based on work supported by the National Science Foundation under Grant No. BCS-1238599 and the Eunice Kennedy Shriver National Institute of Child Health and Human Development under Cooperative Agreement U01-HD-076595. 
  Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author(s) and do not necessarily reflect the views of the National Science Foundation or the Eunice Kennedy Shriver National Institute of Child Health and Human Development.
  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Session-level Data Management}{name=data,column=1,span=1,below=splash,above=thanks}{
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{center}
    \includegraphics[width=.9\textwidth]{img/spreadsheet.png}
  \end{center}
  % Users may upload video and related data \emph{as they collect it}, recording information about individual data collection sessions such as participant characteristics, sharing permission levels, and study conditions.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Timeline}{name=data,column=2,span=1,row=1,below=volume,aligned=data}{
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{center}
  \includegraphics[width=\textwidth]{img/timeline.png}
  \end{center}
  % Within a session, data are represented on a timeline that reflects when events occurred. This is particularlly useful for studies with multiple camera views (e.g., eye-tracking) or with combined behavioral and image/display streams.
}

\end{poster}

\end{document}

### 2015-04-30 2015-06-01-nsf-ed-workshop/overleaf-apa6-template.tex
add nsf ed workshop

%https://www.overleaf.com/2636873dcbhqd#/6964401/
\documentclass[a4paper,man,natbib]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}

\title{Your APA6-Style Manuscript}
\shorttitle{Your APA6-Style Manuscript}
\author{You}
\affiliation{Somewhere}

\abstract{Your abstract here.}

\begin{document}
\maketitle

\section{Introduction}

Your introduction goes here! Some examples of commonly used commands and features are listed below, to help you get started.

If you have a question, please use the support box in the bottom right of the screen to get in touch.

\section{Some \LaTeX{} Examples}
\label{sec:examples}

\subsection{Sections}

Use section and subsection commands to organize your document. \LaTeX{} handles all the formatting and numbering automatically. Use ref and label commands for cross-references.

\subsection{Comments}

You can add inline TODO comments with the todonotes package, like this:
\todo[inline, color=green!40]{This is an inline comment.}

\subsection{References}

LaTeX automatically generates a bibliography in the APA style from your .bib file. The citep command generates a formatted citation in parentheses \citep{Lamport1986}. The cite command generates one without parentheses. LaTeX was first discovered by \cite{Lamport1986}.

\subsection{Tables and Figures}

Use the table and tabular commands for basic tables --- see Table~\ref{tab:widgets}, for example. You can upload a figure (JPEG, PNG or PDF) using the files menu. To include it in your document, use the includegraphics command as in the code for Figure~\ref{fig:frog} below.

% Commands to include a figure:
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{frog.jpg}
\caption{\label{fig:frog}This is a figure caption.}
\end{figure}

\begin{table}
\centering
\begin{tabular}{l|r}
Item & Quantity \\\hline
Widgets & 42 \\
Gadgets & 13
\end{tabular}
\caption{\label{tab:widgets}An example table.}
\end{table}

\subsection{Mathematics}

\LaTeX{} is great at typesetting mathematics. Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
      = \frac{1}{n}\sum_{i}^{n} X_i$$
denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

\subsection{Lists}

You can make lists with automatic numbering \dots

\begin{enumerate}
\item Like this,
\item and like this.
\end{enumerate}
\dots or bullet points \dots
\begin{itemize}
\item Like this,
\item and like this.
\end{itemize}

We hope you find write\LaTeX\ useful, and please let us know if you have any feedback using the help menu above.

\bibliography{example}

\end{document}

%
% Please see the package documentation for more information
% on the APA6 document class:
%
% http://www.ctan.org/pkg/apa6
%

### 2015-04-30 2015-06-01-nsf-ed-workshop/paper.tex
add nsf ed workshop

--- 2015-04-30 2015-06-01-nsf-ed-workshop/overleaf-apa6-template.tex
@@
-%https://www.overleaf.com/2636873dcbhqd#/6964401/
 \documentclass[a4paper,man,natbib]{apa6}
@@
 
-\title{Your APA6-Style Manuscript}
-\shorttitle{Your APA6-Style Manuscript}
-\author{You}
-\affiliation{Somewhere}
+\title{The Role of Video Data Sharing in Advancing Education Policy and Practice}
+\shorttitle{Video Data Sharing in Education}
+\author{Rick O. Gilmore}
+\affiliation{The Databrary (databrary.org) Project \\ \& The Pennsylvania State University}
 
@@
 
-\section{Some \LaTeX{} Examples}
+\section{Promise of Video Data Sharing}
 \label{sec:examples}
@@
 
-LaTeX automatically generates a bibliography in the APA style from your .bib file.
-The citep command generates a formatted citation in parentheses \citep{Lamport1986}.
-The cite command generates one without parentheses.
-LaTeX was first discovered by \cite{Lamport1986}.
-
 \subsection{Tables and Figures}
-
-Use the table and tabular commands for basic tables --- see Table~\ref{tab:widgets}, for example.
-You can upload a figure (JPEG, PNG or PDF) using the files menu.
-To include it in your document, use the includegraphics command as in the code for Figure~\ref{fig:frog} below.
-
-% Commands to include a figure:
-\begin{figure}
-\centering
-\includegraphics[width=0.5\textwidth]{frog.jpg}
-\caption{\label{fig:frog}This is a figure caption.}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{l|r}
-Item & Quantity \\\hline
-Widgets & 42 \\
-Gadgets & 13
-\end{tabular}
-\caption{\label{tab:widgets}An example table.}
-\end{table}
-
-\subsection{Mathematics}
-
-\LaTeX{} is great at typesetting mathematics.
-Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-      = \frac{1}{n}\sum_{i}^{n} X_i$$
-denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
 
@@
 \end{document}
-
-%
-% Please see the package documentation for more information
-% on the APA6 document class:
-%
-% http://www.ctan.org/pkg/apa6
-%

### 2015-04-30 2015-06-01-nsf-ed-workshop/paper.tex
full outline

@@
-\documentclass[a4paper,man,natbib]{apa6}
+\documentclass[letterpaper,man,natbib]{apa6}
 
@@
 
-Your introduction goes here! Some examples of commonly used commands and features are listed below, to help you get started.
+Video is a uniquely rich medium for capturing in real time the interactions that go on in classroom, laboratory, and informal learning settings.
 
-If you have a question, please use the support box in the bottom right of the screen to get in touch.
+\section{The Promise of Video Data Sharing}
 
-\section{Promise of Video Data Sharing}
-\label{sec:examples}
+\subsection{Video is Uniquely Rich}
 
-\subsection{Sections}
+\subsection{Video Uses in Education}
 
-Use section and subsection commands to organize your document. \LaTeX{} handles all the formatting and numbering automatically. Use ref and label commands for cross-references.
+\section{The Challenges of Video Data Sharing}
 
-\subsection{Comments}
+\subsection{Technical}
 
-You can add inline TODO comments with the todonotes package, like this:
-\todo[inline, color=green!40]{This is an inline comment.}
+\subsubsection{Online Video Sharing}
 
-\subsection{References}
-
-\subsection{Tables and Figures}
-
-\subsection{Lists}
-
-You can make lists with automatic numbering \dots
-
-\begin{enumerate}
-\item Like this,
-\item and like this.
-\end{enumerate}
-\dots or bullet points \dots
 \begin{itemize}
-\item Like this,
-\item and like this.
+\item Size of files
+\item Storage/streaming
+\item Formats/transcoding
+\item Preservation
 \end{itemize}
 
-We hope you find write\LaTeX\ useful, and please let us know if you have any feedback using the help menu above.
+\todo[inline, color=green!40]{Grab text from other papers?}
+
+\subsubsection{Tools for Coding Video}
+
+Human coders evaluate, apply text-based labels to videos using a wide-range of commercial and academic coding tools: Transana, StudioCode, V-code, MaxQDA, Noldus Observer, Mangold Interact, Datavyu.
+No standard data format for coding tools.
+Some offer web/cloud based storage (e.g., Transana, V-code), but otherwise coded data may not be easily shared within tools, much less between them.
+With few exceptions, machine learning has not been applied to video data coding.
+
+\subsection{Ethics and Privacy}
+
+\subsection{Data Management}
+
+\subsection{Changing Community Practices}
+
+\section{Databrary.org}
+
+\subsection{System Design}
+
+\subsection{Policies for Safe, Secure, \& Sharing}
+
+\subsection{Managing Data for Sharing}
+
+\section{Conclusion}
 

### 2015-05-03 vss-15/poster.tex
small edits

@@
 					Only those with independent researcher status at an institution may apply for authorization. 
-					Applicants promise to uphold ethical principles and sign an access agreement. The institution co-signs the agreement.
+					Applicants promise to uphold ethical principles and sign an access agreement. The institution co-signs the agreement to grant authorization.
 				\end{indent}
@@
 				\begin{indent}
-					Identifiable data may only be shared if participants have given permisssion.
+					Identifiable data may only be shared if participants have given informed consent.
 					Databrary has developed a release template whose language may be adapted for specific research protocols (https://databrary.org/access/policies.html).
@@
 			Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
-			Video and audio data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4.
+			Video data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4.
 Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
@@
 NYU does routine tape backups that are stored off site.
-There is no cost to use the system; an institutional subscription model is under development.
+There is no cost to use the system.
 		\end{block}
@@
 			\end{center}
-			Databrary's splash page shows a timeline of recent activity, a set of clickable tags that link to studies with those search terms, a featured dataset, and links that enable users to search for data or manage their own data.
+			Databrary's splash page shows a feed of recent activity, a set of clickable tags that link to studies with those search terms, a featured dataset, and links that enable users to search for data or manage their own data.
 		\end{block}
@@
 			Users may upload video and related data \emph{as they collect it}, recording information about individual data collection sessions such as participant characteristics, sharing permission levels, and study conditions.
+Once uploaded, the project remains private only to named collaborators until the data owner chooses to share it with Databrary.
 		\end{block}

### 2015-05-03 vss-15/poster_landscape.tex
replace timeline figure

@@
   % Authors
-  {R.O. Gilmore, K.E. Adolph, D.S. Millman, L. Steiger, and D.A. Simon \\
-   \small{Vision Sciences Society 2015 -- Poster 26.4031 -- \emph{rick@databrary.org}}}
+  {Rick O. Gilmore (\emph{rick@databrary.org}), Karen E. Adolph, David S. Millman, Lisa Steiger, and Dylan A. Simon \\
+  VSS 2015 -- Poster 26.4031}
   % University logo
@@
   \begin{center}
-  \includegraphics[width=\textwidth]{img/volume-31.png}
+  \includegraphics[width=\textwidth]{img/volume-31-new.jpg}
   \end{center}
-  Each dataset or study has a page with summary, owner, funding, citation information, materials, and links to publications.
 }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{Timeline}{name=data,column=2,span=1,row=1,below=volume,aligned=data}{
+\headerbox{Timeline}{name=data,column=2,span=1,row=1,below=volume, above=bottom, aligned=data}{
   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
   \begin{center}
-  \includegraphics[width=\textwidth]{img/timeline.png}
+  \includegraphics[width=.95\textwidth]{img/timeline.png}
   \end{center}

### 2015-05-06 2015-06-01-nsf-ed-workshop/paper.tex
copied raw from many sources

@@
 \author{Rick O. Gilmore}
-\affiliation{The Databrary (databrary.org) Project \\ \& The Pennsylvania State University}
+\author{Karen E. Adolph}
+\author{David S. Millman}
+\affiliation{The Databrary (databrary.org) Project}
 
-\abstract{Your abstract here.}
+\abstract{
+Video captures the complexity, richness, and diversity of behavior unlike any other measure. 
+As a result, large numbers of people who study teaching and learning or seek to improve it employ video. 
+Video documents itself to a large degree, and therefore has significant potential for data re-use.
+This potential remains largely unrealized because videos are rarely shared.
+Video often contains information about personal identities, so considerations about research ethics pose
+challenges to sharing. 
+The relatively large size of video files and diversity of formats and software tools for coding pose technical challenges. 
+In this thought paper, we will describe how the web-based Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
+The Databrary model and system provides insights about how video and other identifiable data collected in the context of education research might be shared.
+Greater data sharing will advance education policy and improve practice.
+}
 
@@
 \section{Introduction}
+Scientific discovery and the application of discovery to human needs such as improved educational policy and practice depends on the free flow of information. 
+According to the NIH Data Sharing Policy (NIH 1), open data sharing is essential for translating research findings into improvements in human health. 
+Open data sharing promotes greater transparency and peer oversight, supports inquiry into data collection methods and measurement, stimulates new work, encourages diversity of analysis and opinion, allows for testing of new or alternative hypotheses, facilitates education of new researchers, permits integrative analyses by creating new datasets from multiple sources, allows future research to build on earlier efforts, and enables exploration of topics not envisioned by the original investigators.
+In short, open data sharing leads to new uses and scientific insights and thus increases the impact of public investments in research.
+Open data sharing is a scientific imperative and common practice in many areas of biomedical, physical, and earth sciences where it has accelerated the pace of discovery (NIH 2–5).
+Unfortunately, research on behavioral development remains shrouded in a culture of isolation.
+Rather than providing direct access to raw data, researchers share interpretations of distilled data through publications and presentations5.
 
-Video is a uniquely rich medium for capturing in real time the interactions that go on in classroom, laboratory, and informal learning settings.
+Reliance on video impedes data sharing in developmental science. 
+Most developmental researchers rely on video because it is cheap, easy to use, and faithfully renders the richness and complexity of behavior6. 
+However, researchers rarely share raw video data—even within labs and among collaborators. 
+After publication, the rich, largely under-used videos molder quietly on a shelf.
+This situation is ironic because video data are conducive to sharing; the behaviors recorded are directly available for peer inspection and the complexity of video data makes it amenable to mining for new information.
+Certainly, sharing digital video poses technical problems and challenges to protection of participant privacy.
+But, failure to overcome the obstacles to sharing has impeded transparency and peer oversight, limited the generalizability of findings, slowed progress, and diminished the impact of Federal funding.
+The Databrary project aims to increase scientific transparency and speed insights into behavioral development by building the community practices and infrastructure for open video data sharing.
+The contribution of a particular dataset will no longer depend on the private activities of one researcher, but will instead benefit from the imagination of many researchers with different viewpoints.
+
+Video is a uniquely rich medium for capturing in real time the interactions that occur in classroom, laboratory, and informal learning settings. 
+Most researchers in the developmental and learning sciences collect video as raw data, but ethical and technical barriers to sharing video have made open data sharing uncommon. 
+While personally identifying information can be removed from text-based data, videos may contain faces, voices, names spoken aloud, and sometimes views of the homes of research participants. 
+These cannot be removed without reducing the information content.
+Further, the collection of video or other identifiable or sensitive information requires approval by a research ethics board and informed consent from the participants.
+The consent process formalizes a promise by the research team that a participant's identity will not be disclosed.
+Thus, researchers risk violating participants’ privacy if digital images are viewed or released to others without authorization.
+
+In this paper, we describe the promise of video data sharing, the challenges it poses, and the solutions that the Databrary team has created to address them.
+In developing technology and policies that enable the safe and secure sharing of video and other identifiable research data, Databrary illustrates how researchers interested in advancing educational policy and practice through widespread video data sharing might achieve their ambitious aims.
 
@@
 \section{The Challenges of Video Data Sharing}
+
+Sharing video poses challenges. 
+These include technical infrastructure, research ethics, privacy, and security, and data management. 
+Further, making open data sharing in education research an established norm requires changing understanding about how identifiable data might be shared and building a community of researchers and educators committed to making video data sharing work.
 
@@
 
-\todo[inline, color=green!40]{Grab text from other papers?}
-
 \subsubsection{Tools for Coding Video}
-
-Human coders evaluate, apply text-based labels to videos using a wide-range of commercial and academic coding tools: Transana, StudioCode, V-code, MaxQDA, Noldus Observer, Mangold Interact, Datavyu.
-No standard data format for coding tools.
+After videos are recorded, human coders evaluate videos and apply text-based labels or numeric codes to videos.
+Coders use a wide-range of commercial and academic software tools: Transana, StudioCode, V-code, MaxQDA, Noldus Observer, Mangold Interact, and Datavyu, among others.
+Some use spreadsheets or pencil and paper.
+The tools have no standard format for data storage or export.
 Some offer web/cloud based storage (e.g., Transana, V-code), but otherwise coded data may not be easily shared within tools, much less between them.
 With few exceptions, machine learning has not been applied to video data coding.
+\subsection{Ethics and Privacy}
+Sharing video recordings poses a unique challenge to existing data sharing policies because videos contain personally identifying information—specifically participants’ faces and voices and often the
+insides of their homes and classrooms. 
+Sharing personally identifiable information puts research participants and others depicted in recordings at increased risk for loss of privacy. 
+At the same time, blurring or altering original recordings to hide identities undermines or eliminates
+their value to other researchers. 
+Often, participants’ faces and voices produce the behaviors of interest.
+A challenge for future education research is to find ways to preserve the value of video for reuse by others while protecting the privacy of research participants.
+Databrary has developed an access model described below that provides an example of how this might be accomplished.
+\subsection{Data Management}
+A primary impediment to sharing is researchers’ lack of time to find, label, clean, organize, and copy their files into formats that can be used and understood by others15. 
+Even scientists committed to data sharing find it difficult to do. Data management systems that make sharing convenient, reliable—and optimally automatic29—can help.
 
-\subsection{Ethics and Privacy}
+Existing data management practices pose challenges for video sharing.
+Despite similar research methods, study designs vary widely. 
+No two developmental science labs manage data in the same way. 
+Some studies are longitudinal, where researchers observe the same participants at
+multiple sessions. 
+Some are cross-sectional, where researchers observe
+each participant at only one session. 
+The timing of observations may be
+determined by participants’ age (e.g., 4-month-olds, 8-month-olds,
+12-month-olds) or abilities (e.g., preverbal, one-word utterances,
+sentences), experimentally determined variables (e.g., pre- and
+post-intervention), or other factors. 
+Some labs organize longitudinal data by grouping files first by participant and then by session date.
+Other labs organize longitudinal data by session and then by
+participant. 
+Still others organize longitudinal data based on task (book reading, block building, free play).
+Some researchers institute a central data management system to be followed by all the lab members, providing easier access and greater transparency for the entire lab but not necessarily providing the structure for similar benefits to
+researchers unfamiliar with the lab’s practices. 
+Other researchers allow their students to keep separate records, making it difficult to share
+data even within the lab. 
+Some researchers keep videos, metadata, and analyses together, and some do not. Idiosyncratic terms, record-keeping, and data management systems are the norm. 
+Databrary must enable researchers to discover and understand each other’s materials, regardless of the original investigator’s data management system.
+As in other fields22, developmental science is inundated by an explosion of digital data, most of which is inaccessible to other researchers.
+The data deluge stems from the advent of cheap, high-resolution digital cameras.
+The average developmental lab collects 12 hours of video/week23.
+Raw data without information about workflow and data provenance is essentially “orphaned” and cannot be understood by subsequent researchers who weren’t part of the original team24.
+Moreover, video files must be electronically linked with participant permissions and relevant metadata.
+An effective data management system for video does not currently exist.
+We will build one to allow researchers within labs to benefit from each other’s work, facilitate interdisciplinary collaborations, and provide a solid foundation for open sharing25
+\subsection{Changing Community Practices}
+Data sharing is an established best practice in some areas of educational research (AERA ethics guide), but it is not necessarily widespread.
+\section{Databrary.org}
+Databrary (databrary.org) is a digital data library specialized for storing and sharing video and research data and metadata.
+Databrary is a joint project of New York University (NYU) and The Pennsylvania State University (Penn State). 
+The Databrary project began with a workshop on open video data sharing funded by the National
+Science Foundation (NSF) under Grant No.
+BCS-1139702 at which developmental scientists, computer scientists, library scientists, and federal agency program officers discussed the promises and challenges of
+sharing video data. 
+NSF and the National Institute of Child Health and Human Development (NICHD) have provided funding under grant No. BCS-1238599 and Cooperative Agreement U01-HD-076595, respectively. 
+The interdisciplinary team bring expertise in developmental and neural
+science, information technology, library science, software development,
+data curation, and community engagement. 
+In addition, a board of expert
+advisors composed of developmental scientists, library scientists, and
+leaders of data repositories in the behavioral and social sciences
+provide guidance.
+No large-scale repository for sharing video data currently exists. So, we will create a web-based Databrary repository. 
+\subsection{System Design}
+Databary is built on a PostreSQL database using the Scala Play framework and JavaScript.
+Data are preserved indefinitely in a secure data storage facility at NYU managed by central IT.
+There is no cost to use the system; an institutional subscription model is under development.
+All code is hosted on GitHub.
+Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
+Video and audio data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4.
+Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
 
-\subsection{Data Management}
+Common video and coding file formats are required for users to build on earlier coding efforts, perform integrative analyses across shared files, and search effectively in a shared repository.
+We will develop tools for converting files from other coding tools into a common format for the Databrary.
 
-\subsection{Changing Community Practices}
+Behavior is rich and complex.
+Research interests and methods are diverse.
+Naturally, video coding is idiosyncratic.
+The only universal tags in developmental research are children’s age and sex.
+Thus, OpenSHAPA is powerful and flexible.
+It is designed to help researchers understand behavior—or any other time series—using codes of their choosing.
+As the PI's experience demonstrates, OpenSHAPA will strengthen work within labs.
+It will also provide the common coding format for files in the Databrary.
+\subsection{Policies for Safe, Secure, \& Sharing}
+Shared videos are maximally informative when participants’ faces are visible and their vocalizations are audible. Thus, anonymity cannot be ensured as it is for flat file or imaging data. 
+To address issues of confidentiality, access, and IRB approval for sharing video data, we will develop template data sharing permission forms and contributor and user agreements with input from the community, from human compliance officers, and from leaders of previous data sharing efforts.
+Databrary has elected to maximize the potential for data re-use by keeping recordings in their original
+unaltered form. 
+Instead of removing participants’ identities, Databrary restricts access to identifiable or sensitive data to authorized researchers. 
+Further, Databrary provides access only when the peopledepicted have given permission for their information to be shared with other researchers.
+\subsubsection{Restricting Access}
+Databrary provides access to shared data only to authorized researchers who have agreed to uphold common practices concerning the responsible and ethical use of identifiable and sensitive data. 
+To become an authorized investigator, applicants must register on the site and electronically sign an Access Agreement, which must also be co-signed by the applicant’s institution. 
+Full privileges will be granted only to researchers with principal investigator (PI) status
+at their institutions. 
+Other researchers may be granted privileges if they are affiliated with a PI who agrees to sponsor and supervise their application. 
+Initially there will be a manual process to identify the institutional representative—typically the authorizing official of the university—who can co-sign the Investigator Agreement. 
+However, as the user groups at each university expand, Databrary may implement
+administrative accounts at each institution. 
+This will enable the authorizing official to independently manage the authorizations of
+individual researchers at her institution.
 
-\section{Databrary.org}
+Data from a particular session may be stored in Databrary for the contributing researcher’s use whether the records are shared with other scientists or not. 
+When a researcher chooses to share, Databrary makes records openly available to the community of authorized researchers only if the people depicted in the recordings have given permission to release the data for sharing.
+Thus, Databrary requires that people depicted in recordings grant permission before their information can be
+shared.
+Databrary’s policies extend currently accepted principles of informed consent to the situation where participants are granted authority to consent to (or refuse) the release of their identifiable data.
 
-\subsection{System Design}
+We developed these ideas in close collaboration with the NYU and PSU IRB staff. 
+To formalize the process of acquiring permission, we developed a Participant Release Form Template, based on photo or video release language many researchers use currently. 
+The template release form has standard language that Databrary recommends investigators should use
+with study participants. 
+This language makes it easy for participants to understand what is involved in sharing their video data, with whom it will be shared, and the potential risks associated with releasing their video and other identifiable data to other researchers.
+Use of the template also allows for the standardization of language associated with the release of identifiable or sensitive participant data.
 
-\subsection{Policies for Safe, Secure, \& Sharing}
+Some IRBs may deem an investigator’s existing, approved video or photo release form equivalent to the Databrary release. 
+This enables a researcher to share with Databrary recordings they have already
+collected. 
+However, most researchers will need to modify their research protocols, by adding the Databrary sharing permission procedures, prior to collecting new shareable video data. 
+Databrary staff are available to advise potential data contributors about how to amend existing research
+protocols so that the information acquired is Databrary-compliant.
+Protocol amendments involve seeking approval for use of the Databrary
+template release form and modifying the time period over which collected
+data will be made available. 
+Specifically, researchers must remove any clauses in research consent documents that require data destruction
+after some fixed period of time since Databrary intends to store shared
+data indefinitely.
+\subsection{Managing Data for Sharing}
+Raw video is most informative when it is linked with coding spreadsheets, codebooks, protocols, statistical analyses, and manuscripts. 
+A data management system is required to link related files, document workflow and data provenance, and tag files with appropriate permission levels for sharing. 
+Consistent data management practices will strengthen work within labs and among collaborators, enable convenient and reliable file uploading, and enhance the value of shared data in the Databrary.
 
-\subsection{Managing Data for Sharing}
+The 2011 NSF workshop8 and PI Adolph's MacSHAPA/OpenSHAPA workshops over the last decade have revealed widely varying laboratory data management practices.
+Some minimal standardization across labs is required to make data sharing feasible.
+In the biomedical sciences, use of laboratory information systems (LIMS) is increasingly common and shows promise for lowering barriers to sharing in other fields15. 
+Consistent with our overall approach, we will research, adapt, and adopt best laboratory management practices and open-source tools from other fields34, and we will prioritize implementation based on the preferences and priorities of the developmental science community.
+
+The full expressive flexibility of the data model is not evident to Databrary users. 
+Instead, users see different aspects of the data model at different times, depending on the context. Researchers organize their materials by acquisition date and time into structures called sessions.
+A session corresponds to a unique recording episode and contains one or more recordings or related flat files (assets).
+Researchers can group sessions in different ways, using whatever groupings are appropriate for
+the particular dataset. 
+These groupings may also identify particular time segments of a recording to distinguish tasks, events, or
+participants that comprise the session. 
+In the data model, the aggregate of these flexible groupings is called a volume. 
+In practice, researchers combine sessions or segments of sessions within and across datasets to
+form the raw material that are subsequently described in published
+articles and presentations. 
+Thus, Databrary contributors can combine sessions or segments within and across datasets with ancillary
+materials, such as coding manuals, Datavyu spreadsheets, statistical analyses, questionnaires, IRB documents, computer code, sample displays, and links to published journal articles in an aggregate structure
+investigators call a study. 
+Like datasets, studies are represented as volumes in the data model.
+
+The discovery and browsing interfaces offer content items at the volume (study or dataset) level to users searching the library.
+The contributor-assigned groupings (records), which may include relevant
+metadata about participants (measures, such as participant demographic information, domain-specific survey information, location data, condition variables, task descriptions, or other properties), allow a
+second level of filtering and organization within and between relevant
+identified studies or datasets.
+This enables users to quickly identify data that may be relevant to their own interests or research.
+
+Authorized Databrary Investigators can also add keyword tags at the study/dataset, session, or segment level, and can endorse or deprecate keywords that have been suggested by other authorized researchers.
+These
+abilities are critical because different researchers may have very different reasons for citing a study or using data. Future implementations may extend the ability of authorized researchers to
+annotate studies with other kinds of metadata.
+
+The Databrary team recognizes that other data repositories enforce strict metadata ontologies and that doing so may have benefits in sub-domains of research when there is community consensus [12]. 
+We will support standard data coding ontologies among researchers, but only enforce standardization for a small set of standard tags such as study date, participant birth date, and sex.
+We will also encourage
+contributors to report race, ethnicity, primary language, language of
+dataset, and location of session. 
+The Databrary team chose not to require strict metadata ontologies for several reasons. We hope to
+reduce the pre-deposit curational demands on contributors, encourage the repurposing of shared data, and foster the rapid adoption of data sharing. 
+Developing and achieving agreement on metadata ontologies can take significant amounts of time. 
+Video data are so rich and complex that in many domains, researchers have not settled on standard definitions for particular behaviors and may have little current need for standard tasks, procedures, or terminology.
+However, we will also encourage users to re-use tags and terminology by suggesting common or similar terms, without confining users to these suggestions. 
+User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not
+necessary for browsing and searching in most of the use cases we
+envision.
+
+Our experiences curating and ingesting archival datasets have highlighted the considerable value of contributors entering raw video data into Databrary as soon as recordings are acquired. 
+Immediate uploading reduces the workload on investigators, minimizes the risk of
+data loss and corruption, and accelerates the speed with which materials
+become openly available. 
+To encourage immediate uploading, Databrary provides a complete set of controls so that researchers can restrict
+access to their data to only their own labs or to other users of their
+choosing. 
+Datasets can be shared at a later point when data collection and ancillary materials are complete, whenever the contributor is comfortable sharing, or when journals or funders require it. 
+Databrary has published a Data Sharing Manifesto [13] that explains to researchers the Databrary philosophy. Standards about when data should be shared are evolving. 
+Our philosophy is consistent with concepts and practices in other domains where data sharing is the norm (e.g., www.iedata.org).
+Planned enhancements to the Datavyu tool will allow contributors to organize video files and metadata as part of the analysis process. 
+This will facilitate the contribution of packaged datasets to Databrary.
+
+Based on discussions at the NSF workshop, we expect to see most contributions after data collection, coding, and analyses are completed.
+Researchers have widely differing practices for designing codebooks and spreadsheets, managing data collections, coding, analyses, collecting and managing participant data (demographic, anthropometric, etc.), and documenting workflow and data provenance.
+Our goal for Years 2-3 is to incorporate this more detailed type of information into the data management system.
+Doing so will strengthen sharing within labs, expand the usefulness of Databrary, and enable more powerful searches.
+The data management system will be extended to organize and package other files associated with contributors (position, expertise), studies (codebooks, protocols, analysis spreadsheets graphs, manuscripts, presentations, grant proposals, etc.), and participants (coded spreadsheets, test results, demographics, etc.).
+The same basic ideas will apply— standard file naming conventions and data tagged by contributor, study, and participant level.
+OpenSHAPA and other applications will communicate with the data management tool.
+The data management tool will be used to stitch together all the sharable files that comprise a study.
+From data sharing to best lab practices.
+The developmental science community is hungry for new practices that will enhance research, as demonstrated by the number of committed contributors.
+LIMS in other fields embody additional functions.
+We envision that the data and tool sharing enabled by Databrary will spark interest in other lab management enhancements.
+For example, the data management system could facilitate participant recruitment and scheduling, task/project management and scheduling, workflow processing, and quality assurance (e.g., reliability testing).
+We plan to seek community input—via online comments, conference workshops on best lab practices, and other avenues—about which features should be the highest priorities for data management tool developers.
+Extending the data management tool to incorporate the highest priority features will be a focus of the front-end programmer in Years 4-5.
+
+\subsection{Building A Community That Embracing Sharing}
+Data sharing works only when the scientific community embraces it. 
+To make video-based data sharing succeed, we will build a diverse contributor/user community committed to open data sharing. 
+Based on continual community input, we will expand the OpenSHAPA video coding tool, build data management tools, and develop tools for uploading, browsing, and searching data in the Databrary repository.
+These tools will enable sharing within labs, among collaborators, and among users in the larger community.
+To avoid pitfalls of previous data sharing efforts, leaders in data sharing, developmental science, and computer science will guide our work through an advisory board.
+To ensure continued community engagement, Databrary personnel will provide technical support to individual labs and provide information to the community about new data contributions, citations of shared data, and use cases.
 
 \section{Conclusion}
+As Gesell once noted, cameras can record child behavior in ways that make it "...as tangible as tissue."38 The Databrary project will strengthen traditional ways of conducting video-based research by facilitating better coding, data mining, and data management practices within labs while building the culture and infrastructure to enable a new tradition of open data sharing in developmental science.
+The Databrary project is potentially transformative.
+We will increase transparency by sharing the traditionally private activity of data collection, coding, and analysis.
+We will speed progress by enabling later researchers to benefit from previous work.
+Ultimately, we will better understand human development and how to transform knowledge into improvements in health.
+
+\section{Acknowledgments}
+Databrary is based on work supported by the National Science Foundation under Grant No.
+BCS-1238599, the Eunice Kennedy Shriver National Institute of Child Health and Human Development under Cooperative Agreement U01-HD-076595, and the Society for Research in Child Development.
+Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author(s) and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.
 

### 2015-05-06 vss-15/poster_landscape.tex
vss-15 final
Fix spelling errors

@@
   % Authors
-  {Rick O. Gilmore (\emph{rick@databrary.org}), Karen E. Adolph, David S. Millman, Lisa Steiger, and Dylan A. Simon \\
+  {Rick O. Gilmore \emph{(rick@databrary.org)}, Karen E. Adolph, David S. Millman, Lisa Steiger, and Dylan A. Simon \\
   VSS 2015 -- Poster 26.4031}
@@
       Some vision scientists collect identifiable video recordings of participants for use in off-line tagging or coding. 
-      Databrary may be especially well suited for these use cases because of the library's focus on uploading, organizing, storing, and sharing research data and metadata collected in the course of behavioral resesearch.
+      Databrary may be especially well suited for these use cases because of the library's focus on uploading, organizing, storing, and sharing research data and metadata collected in the course of behavioral research.
       Databrary has developed policies that enable identifiable information to be shared with the permission of research participants.
@@
 
-      Identifiable data may only be shared if participants have given permisssion.
+      Identifiable data may only be shared if participants have given permission.
       The Databrary release template may be adapted for specific research protocols.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    Databary is built on a PostreSQL database using the Scala Play framework and JavaScript.
+    Databrary is built on a PostreSQL database using the Scala Play framework and JavaScript.
     Data are preserved indefinitely in a secure data storage facility at NYU managed by central IT. 
@@
   % Within a session, data are represented on a timeline that reflects when events occurred.
-This is particularlly useful for studies with multiple camera views (e.g., eye-tracking) or with combined behavioral and image/display streams.
+This is particularly useful for studies with multiple camera views (e.g., eye-tracking) or with combined behavioral and image/display streams.
 }

### 2015-05-06 vss-15/poster_landscape.tex
Minor edits for accuracy

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
-    Video and audio data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4.
+    Databrary can house video, audio, PDF, spreadsheet, image, and text-based files along with associated metadata and annotations.
+    Video files are transcoded into standard, HTML5-compatible formats, currently H.264+AAC in MP4.
 Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
@@
     Data are preserved indefinitely in a secure data storage facility at NYU managed by central IT. 
-    There is no cost to use the system; an institutional subscription model is under development.
+    There is no cost to use the system; a sustainability model is under development.
     All code is hosted on GitHub.

### 2015-05-07 vss-15/poster_landscape.tex
modify policies language for accuracy

@@
 
-      Only those with independent researcher status at an institution may apply for authorization. 
-      Applicants promise to uphold ethical principles and sign an access agreement co-signed by the institution.
+      Only those with independent researcher status at an institution may be granted authorized investigator status.
+      Applicants promise to uphold ethical principles and sign an access agreement, which must be co-signed by their institution.
 
@@
       Identifiable data may only be shared if participants have given permission.
+      Databrary has developed a release template for obtaining informed consent for sharing identifiable data.
       The Databrary release template may be adapted for specific research protocols.

### 2015-05-07 vss-15/poster_landscape.tex
Change styling and minor edits to fit in space

@@
       Databrary ensures that identifiable information can be shared in two ways:
-      \begin{itemize}
-      \item{\bf{Restricting Access}}
-
+      \begin{description}
+      \item[Restricting Access:]
       Only those with independent researcher status at an institution may be granted authorized investigator status.
@@
 
-      \item{\bf{Securing Permission to Share}}
-
+      \item[Securing Permission to Share:]
       Identifiable data may only be shared if participants have given permission.
@@
 
-      \end{itemize}
+      \end{description}
   }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% N.B.: this section is currently obscured by splash
   \headerbox{What Can Be Shared}{name=shared,column=1,row=0}{
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    Databrary is built on a PostreSQL database using the Scala Play framework and JavaScript.
-    Data are preserved indefinitely in a secure data storage facility at NYU managed by central IT. 
+    Databrary is built on PostreSQL using the Scala Play framework and AngularJS.
+    Data are preserved indefinitely in a secure data storage facility at NYU. 
     There is no cost to use the system; a sustainability model is under development.
-    All code is hosted on GitHub.
+    All code is GPL on GitHub.
   }

### 2015-05-08 2015-06-01-nsf-ed-workshop/paper.tex
draft of intro

@@
 As a result, large numbers of people who study teaching and learning or seek to improve it employ video. 
-Video documents itself to a large degree, and therefore has significant potential for data re-use.
+Video documents itself to a large degree, and therefore has significant potential for reuse by others.
 This potential remains largely unrealized because videos are rarely shared.
@@
 The relatively large size of video files and diversity of formats and software tools for coding pose technical challenges. 
-In this thought paper, we will describe how the web-based Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
-The Databrary model and system provides insights about how video and other identifiable data collected in the context of education research might be shared.
-Greater data sharing will advance education policy and improve practice.
+In this thought paper, we will describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
+Databrary suggests ways that video and other identifiable data collected in the context of education research might be shared.
+Widespread open sharing of high value, high impact data will advance education policy and improve practice.
 }
@@
 \section{Introduction}
-Scientific discovery and the application of discovery to human needs such as improved educational policy and practice depends on the free flow of information. 
-According to the NIH Data Sharing Policy (NIH 1), open data sharing is essential for translating research findings into improvements in human health. 
-Open data sharing promotes greater transparency and peer oversight, supports inquiry into data collection methods and measurement, stimulates new work, encourages diversity of analysis and opinion, allows for testing of new or alternative hypotheses, facilitates education of new researchers, permits integrative analyses by creating new datasets from multiple sources, allows future research to build on earlier efforts, and enables exploration of topics not envisioned by the original investigators.
-In short, open data sharing leads to new uses and scientific insights and thus increases the impact of public investments in research.
-Open data sharing is a scientific imperative and common practice in many areas of biomedical, physical, and earth sciences where it has accelerated the pace of discovery (NIH 2–5).
-Unfortunately, research on behavioral development remains shrouded in a culture of isolation.
-Rather than providing direct access to raw data, researchers share interpretations of distilled data through publications and presentations5.
 
-Reliance on video impedes data sharing in developmental science. 
-Most developmental researchers rely on video because it is cheap, easy to use, and faithfully renders the richness and complexity of behavior6. 
-However, researchers rarely share raw video data—even within labs and among collaborators. 
-After publication, the rich, largely under-used videos molder quietly on a shelf.
-This situation is ironic because video data are conducive to sharing; the behaviors recorded are directly available for peer inspection and the complexity of video data makes it amenable to mining for new information.
-Certainly, sharing digital video poses technical problems and challenges to protection of participant privacy.
-But, failure to overcome the obstacles to sharing has impeded transparency and peer oversight, limited the generalizability of findings, slowed progress, and diminished the impact of Federal funding.
-The Databrary project aims to increase scientific transparency and speed insights into behavioral development by building the community practices and infrastructure for open video data sharing.
-The contribution of a particular dataset will no longer depend on the private activities of one researcher, but will instead benefit from the imagination of many researchers with different viewpoints.
+The application of insights from research to essential human needs such as improved educational policy and practice depends on the free flow of information. 
+Open data sharing promotes greater transparency and peer oversight, supports inquiry into data collection methods and measurement, stimulates new work, encourages diversity of analysis and opinion, allows for testing of new or alternative hypotheses, facilitates the education of new researchers, permits integrative analyses by creating new datasets from multiple sources, allows future research to build on earlier efforts, and enables exploration of topics not envisioned by the original investigators.
+Open data sharing increases the impact of public investments in research and leads to better and more effective public policy.
+Open data sharing is a scientific imperative and common practice in many areas of biomedical, physical, and earth sciences where it has accelerated the pace of discovery. 
+Unfortunately, research on human development and learning remains shrouded in a culture of isolation. 
+Rather than providing direct access to raw data, researchers in the developmental and learning sciences share interpretations of distilled data almost exclusively through publications and presentations.
+The path from raw data to finding to conclusion cannot be traced or validated by others, nor can new questions be posed by others building on the same, often expensive-to-collect, raw data.
+To realize significant advances in our understanding of teaching and learning and for those insights to have meaningful impact on educational practice and policy, open data sharing must become an established norm, not the rare exception.
+Moreover, we must find ways to share the most valuable raw data, the data with the greatest potential for reuse by others, and the data with the greatest potential for yielding meaningful insights into the learning process.
 
-Video is a uniquely rich medium for capturing in real time the interactions that occur in classroom, laboratory, and informal learning settings. 
-Most researchers in the developmental and learning sciences collect video as raw data, but ethical and technical barriers to sharing video have made open data sharing uncommon. 
-While personally identifying information can be removed from text-based data, videos may contain faces, voices, names spoken aloud, and sometimes views of the homes of research participants. 
-These cannot be removed without reducing the information content.
-Further, the collection of video or other identifiable or sensitive information requires approval by a research ethics board and informed consent from the participants.
-The consent process formalizes a promise by the research team that a participant's identity will not be disclosed.
-Thus, researchers risk violating participants’ privacy if digital images are viewed or released to others without authorization.
+Video is a uniquely rich medium for capturing in real time the interactions that occur in classrooms, muesuems, laboratory, home, and other informal learning settings.
+Many researchers and practitioners in the developmental, learning, and education sciences collect video as raw data.
+Video recordings capture what teachers say and do and what students say and do in ways that offer the potential to discover when, why, and how learning actually works.
+Still, the open sharing of video data is rare, even within labs or among collaborators.
+Video data should be conducive to sharing.
+Video captures behavior that others can readily inspect.
+It captures multiple dimensions of complex real-world learning settings in ways that others can easily build upon.
+But, technical, ethical, and cultural barriers have made open sharing of video data rare.
+Videos come in varied formats and generage large files; these pose conversion and storage challenges.
+Personally identifying information can be removed from text-based data.
+Videos may contain faces, voices, names spoken aloud, views or voices of non-participants, views of classroom interiors and sometimes views of the homes of research participants. 
+These elements cannot be removed without reducing the information content and reuse value to others. 
+The collection of video or other identifiable or sensitive information requires approval by a research ethics board.
+It also requires informed consent or assent from the participants, and possibly parental, teacher, and school district approval.
+Researchers risk violating participants’ privacy if digital images are viewed or released to others without authorization.
+For these and other reasons, no culture of widespread open sharing of video data has emerged in the education, learning, and developmental sciences.
 
-In this paper, we describe the promise of video data sharing, the challenges it poses, and the solutions that the Databrary team has created to address them.
-In developing technology and policies that enable the safe and secure sharing of video and other identifiable research data, Databrary illustrates how researchers interested in advancing educational policy and practice through widespread video data sharing might achieve their ambitious aims.
+The Databrary project has built a digital data library specialized for the open sharing of video data from research in these areas of inquiry.
+In this paper, we will describe how Databrary has overcome the most significant barriers to sharing video, including solutions to maintaining participant privacy, data tagging, and data management.
+In developing technology and policies that enable the safe and secure sharing of video and other identifiable research data, Databrary suggests ways that video and other identifiable data collected in the context of research on teaching and learning might be shared.
+Fostering the open sharing and reuse of high value, high impact data like video promises to advance education policy and improve practice.
 
 \section{The Promise of Video Data Sharing}
+
+More than 1 billion users worldwide uploade more than 300 hours of video to YouTube every minute (https://www.youtube.com/yt/press/statistics.html).
+Google, YouTube's owner, has built an immense data infrastructure to upload, store, convert, tag, and stream video.
+The current scale of video collection in developmental and education contexts is undoubtedly smaller, but it is not small.
+The Gates Foundation-funded Measures of Effective Teaching (MET) Project, hosted at ICPSR, contains X video segments from X classroom settings, representing tens of terabytes of data.
+In just over a year of operation, the Databrary digital library has collected more than 7,000 individual videos, representing 2,400 hours of recording, featuring more than 1,800 infant, child, and adult participants.
+Video data is big, and the interest in recording and sharing video for research, education, and policy purposes has grown.
 
@@
 
+We record teachers in training and teachers in-service for evaluation.
+We record classrooms to capture what teachers do and how students respond.
+We record visitors and docents in museum settings.
 \section{The Challenges of Video Data Sharing}
@@
 
-\subsubsection{Online Video Sharing}
+Video poses two types of technical challenges related to the goal of increasing widespread sharing.
+One concerns the requirements for storing, sharing, streaming, and preserving large numbers of videos.
+The other concerns tools for extracting meaningful information from the episodes and behavior videos capture.
 
-\begin{itemize}
-\item Size of files
-\item Storage/streaming
-\item Formats/transcoding
-\item Preservation
-\end{itemize}
+\subsubsection{Video Data is Big}
+
+Video often accompanies other data streams (physiological recordings, brain imaging, EMG, motion tracking, eye tracking, verbal transcripts). 
+Multiple data streams—including two or more camera views—require tools for synchronization and integration and benefit from multivariate time series analyses. 
+Many labs lack the resources to adopt tools for detailed video coding and complex analyses. 
+Furthermore, widespread use of video creates a data explosion: The typical developmental lab collects 8-12 hours of video/week 18 in widely varied formats. 
+Thus, sharing digital video requires substantial storage capacity, powerful search and streaming tools, and significant computational resources for transcoding videos into common, preservable formats.
 
 \subsubsection{Tools for Coding Video}
-After videos are recorded, human coders evaluate videos and apply text-based labels or numeric codes to videos.
+After videos are recorded, human coders evaluate videos and apply text-based annotations or numeric codes to videos.
 Coders use a wide-range of commercial and academic software tools: Transana, StudioCode, V-code, MaxQDA, Noldus Observer, Mangold Interact, and Datavyu, among others.

### 2015-05-08 2015-06-01-nsf-ed-workshop/paper.tex
fix authors
1st pass thru Technical Challenges

@@
 \shorttitle{Video Data Sharing in Education}
-\author{Rick O. Gilmore}
-\author{Karen E. Adolph}
-\author{David S. Millman}
+\author{Rick O. Gilmore, Karen E. Adolph, David S. Millman}
 \affiliation{The Databrary (databrary.org) Project}
@@
 Many researchers and practitioners in the developmental, learning, and education sciences collect video as raw data.
-Video recordings capture what teachers say and do and what students say and do in ways that offer the potential to discover when, why, and how learning actually works.
+Video recordings capture what teachers, docents, parents, tutors and peers say and do and what learners say and do in response.
+The recordings capture these dynamics in ways that offer the potential to discover who did what, and when, why, and how it influenced learning.
 Still, the open sharing of video data is rare, even within labs or among collaborators.
@@
 
-\section{The Promise of Video Data Sharing}
+\section{The Promise of Video}
 
@@
 Google, YouTube's owner, has built an immense data infrastructure to upload, store, convert, tag, and stream video. 
-The current scale of video collection in developmental and education contexts is undoubtedly smaller, but it is not small.
-The Gates Foundation-funded Measures of Effective Teaching (MET) Project, hosted at ICPSR, contains X video segments from X classroom settings, representing tens of terabytes of data.
-In just over a year of operation, the Databrary digital library has collected more than 7,000 individual videos, representing 2,400 hours of recording, featuring more than 1,800 infant, child, and adult participants.
-Video data is big, and the interest in recording and sharing video for research, education, and policy purposes has grown.
+The current scale of video collection in developmental and education contexts is undoubtedly smaller and more focused in purpose, but it is large and growing. 
+The Gates Foundation-funded Measures of Effective Teaching (MET; http://www.metproject.org/) Project generated more than 1,000 full-length video segments from K-12 classrooms, covering core subjects such as mathematics, language arts from multiple camera angles.
+The data constituting tens of terabytes of storage are hosted at the University of Michigan (http://soe.mivideo.it.umich.edu/) and securely streamed to registered viewers across the country.
+Similarly, in just over a year of operation, the Databrary (http://databrary.org) digital library has collected more than 7,000 individual videos, representing 2,400 hours of recording, and featuring more than 1,800 infant, child, and adult participants.
+Databrary has more than 100 authorized researchers representing more than 50 institutions across the globe who access the videos.
+Video data is big, and the interest in recording and sharing video for research, education, and policy purposes continues to grow.
 
-\subsection{Video is Uniquely Rich}
+Video captures the richness and complexity of human behavior and interactions between people and their environment like no other measure.
+Video captures what people say and what they do.
+It captures when and where they look or gesture, when, where and how they move, and how their looking, gesturing, and moving corresponds to what others are doing.
+These characteristics mean that video has multiple use cases in the study of learning and development.
+In laboratory contexts, researchers use videos to study how infants, children, and adults behave spontaneously or when presented with specific researcher-specified tasks.
+In educational contexts, educators, teacher trainers and evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance.
+Researchers may record videos of learners in classrooms or museums to understand what teachers and docents do and how audiences respond.
+In lab, classroom, or museum settings, video-based eye tracking systems provide information about where people look and for how long.
+While not strictly video recordings, image streams from computer-based tasks or learning-rich video games may be combined with information about user keyboard or controller inputs to study real-time perfomance.
+When combined with other quantitative and qualitative measures about learning outcomes, video can provide information about what kinds of teaching techniques boost learning for what kinds of students.
+Video may even be used to study engagement by students enrolled in online/web-based courses or to monitor compliance with academic integrity rules.
+Because video closely mimics the multisensory experiences of live human observers, recordings collected by one person for a particular purpose may be readily understood by another person and reused for another purpose.
+In other words, video requires less metadata and less manual curation than other data types, but offers greater potential for reuse.
+Moreover, the success, growth, and popularity of YouTube and other video-based social media (Vimeo, Periscope, Meerkat, SnapChat) demonstrates that video garners high levels of user interest and attention for interpersonal communication and entertainment uses.
+The success of these services also suggests that web-based video storage and streaming systems are now sufficiently well developed to satisfy large-scale demand.
+The question for researchers and policymakers is how to seize the moment in order to realize video's substantial potential to improve teaching and learning.
 
-\subsection{Video Uses in Education}
+\section{The Challenge of Video}
 
-We record teachers in training and teachers in-service for evaluation.
-We record classrooms to capture what teachers do and how students respond.
-We record visitors and docents in museum settings.
-\section{The Challenges of Video Data Sharing}
-
-Sharing video poses challenges. 
+Sharing research video poses challenges.
 These include technical infrastructure, research ethics, privacy, and security, and data management. 
@@
 
-\subsection{Technical}
+\subsection{Technical Challenges}
 
@@
 
-\subsubsection{Video Data is Big}
+Videos are large.
+One hour of HD video (1080p, 1440x1080 @ 60 Hz) can consume more than 10 GB.
+The size of the file varies by format and resolution, and recordings vary widely, making it essential for video storage systems to convert files to a standard format.
+Many settings, like classrooms, require multiple camera views to capture the desired behaviors.
+Some questions involve the collection of multiple, temporally dense data streams (physiological recordings, motion tracking, eye tracking, verbal transcripts, keyboard/controller responses).
+These require tools for synchronization across camera views and data streams.
+The use of video creates a data explosion: A typical lab studying infant or child development collects 8-12 hours of video/week [cite18] in widely varied formats. 
+Thus, sharing digital video requires substantial storage capacity, significant computational resources for transcoding videos into common, preservable formats, and powerful search and streaming tools,.
 
-Video often accompanies other data streams (physiological recordings, brain imaging, EMG, motion tracking, eye tracking, verbal transcripts). 
-Multiple data streams—including two or more camera views—require tools for synchronization and integration and benefit from multivariate time series analyses. 
-Many labs lack the resources to adopt tools for detailed video coding and complex analyses. 
-Furthermore, widespread use of video creates a data explosion: The typical developmental lab collects 8-12 hours of video/week 18 in widely varied formats. 
-Thus, sharing digital video requires substantial storage capacity, powerful search and streaming tools, and significant computational resources for transcoding videos into common, preservable formats.
+After videos are recorded, human coders evaluate videos and apply text-based annotations or numeric codes to videos.
+Coders use a wide-range of commercial (Noldus Observer, Mangold Interact, MaxQDA, V-Note), and academic (Transana, Elan, Datavyu) software tools, among others.
+Some researchers use spreadsheets or pencil and paper.
+Some tools (Datavyu, Mangold, V-Note) offer text-based export to interoperable file formats, but the tools have no standard format for data storage or export.
+Some tools offer web/cloud based storage (e.g., Datavyu via Databrary, Transana, V-Note), but otherwise coded data are easily shared within tools, much less between them.
 
-\subsubsection{Tools for Coding Video}
-After videos are recorded, human coders evaluate videos and apply text-based annotations or numeric codes to videos.
-Coders use a wide-range of commercial and academic software tools: Transana, StudioCode, V-code, MaxQDA, Noldus Observer, Mangold Interact, and Datavyu, among others.
-Some use spreadsheets or pencil and paper.
-The tools have no standard format for data storage or export.
-Some offer web/cloud based storage (e.g., Transana, V-code), but otherwise coded data may not be easily shared within tools, much less between them.
-With few exceptions, machine learning has not been applied to video data coding.
+With few exceptions (Chen Yu; Gilmore \& Raudies, Rehg?, others), machine learning has not been applied to video data coding in the developmental and learning sciences.
+But, there is explosive growth in the use of machine learning and computer vision tools to tag images and videos in commercial settings.
+This may create opportunities for future projects that supply machine-generated tags to videos collected in educational and developmental research.
+
 \subsection{Ethics and Privacy}
+
 Sharing video recordings poses a unique challenge to existing data sharing policies because videos contain personally identifying information—specifically participants’ faces and voices and often the
@@
 We will build one to allow researchers within labs to benefit from each other’s work, facilitate interdisciplinary collaborations, and provide a solid foundation for open sharing25
+
 \subsection{Changing Community Practices}
+
 Data sharing is an established best practice in some areas of educational research (AERA ethics guide), but it is not necessarily widespread.
+
 \section{Databrary.org}
@@
 No large-scale repository for sharing video data currently exists. So, we will create a web-based Databrary repository.
+
 \subsection{System Design}
+
 Databary is built on a PostreSQL database using the Scala Play framework and JavaScript.

### 2015-05-08 2015-06-01-nsf-ed-workshop/paper.tex
add some refs

@@
-\documentclass[letterpaper,man,natbib]{apa6}
+\documentclass[letterpaper,man,apacite]{apa6}
 
@@
 \usepackage{graphicx}
-\usepackage[colorinlistoftodos]{todonotes}
 
@@
 \shorttitle{Video Data Sharing in Education}
-\author{Rick O. Gilmore, Karen E. Adolph, David S. Millman}
-\affiliation{The Databrary (databrary.org) Project}
+
+\twoauthors{Rick O. Gilmore}{Karen E. Adolph, David S. Millman}
+\twoaffiliations{The Pennsylvania State University}{New York University}
 
@@
 The relatively large size of video files and diversity of formats and software tools for coding pose technical challenges. 
-In this thought paper, we will describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
+In this paper, we will describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
 Databrary suggests ways that video and other identifiable data collected in the context of education research might be shared.
@@
 
-More than 1 billion users worldwide uploade more than 300 hours of video to YouTube every minute (https://www.youtube.com/yt/press/statistics.html). 
+More than 1 billion users worldwide uploade more than 300 hours of video to YouTube every minute \cite{YouTube2015}. 
 Google, YouTube's owner, has built an immense data infrastructure to upload, store, convert, tag, and stream video. 
@@
 These characteristics mean that video has multiple use cases in the study of learning and development.
-In laboratory contexts, researchers use videos to study how infants, children, and adults behave spontaneously or when presented with specific researcher-specified tasks.
-In educational contexts, educators, teacher trainers and evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance.
-Researchers may record videos of learners in classrooms or museums to understand what teachers and docents do and how audiences respond.
-In lab, classroom, or museum settings, video-based eye tracking systems provide information about where people look and for how long.
+In laboratory contexts, researchers use videos to study how \cite{Karasik2014}.
+In educational contexts, educators, teacher trainers and evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance \cite{Blomberg2014, Masats20111151, Baecher2013189}.
+.
+Researchers may record videos of learners in classrooms \cite{Alibali2012} or museums to understand what teachers do and how audiences respond or how using video affects what museum-goers learn \cite{Bakken2015}.
+In lab \cite{Kretch2014}, classroom \cite{Mason2015, Prieto:2014:STC:2669485.2669543}, or museum settings video-based eye tracking systems provide information about where people look and for how long.
 While not strictly video recordings, image streams from computer-based tasks or learning-rich video games may be combined with information about user keyboard or controller inputs to study real-time perfomance.
@@
 The use of video creates a data explosion: A typical lab studying infant or child development collects 8-12 hours of video/week [cite18] in widely varied formats. 
-Thus, sharing digital video requires substantial storage capacity, significant computational resources for transcoding videos into common, preservable formats, and powerful search and streaming tools,.
+Thus, sharing digital video requires substantial storage capacity, significant computational resources for transcoding videos into common, preservable formats, and powerful search and streaming tools.
 
-After videos are recorded, human coders evaluate videos and apply text-based annotations or numeric codes to videos.
-Coders use a wide-range of commercial (Noldus Observer, Mangold Interact, MaxQDA, V-Note), and academic (Transana, Elan, Datavyu) software tools, among others.
-Some researchers use spreadsheets or pencil and paper.
-Some tools (Datavyu, Mangold, V-Note) offer text-based export to interoperable file formats, but the tools have no standard format for data storage or export.
-Some tools offer web/cloud based storage (e.g., Datavyu via Databrary, Transana, V-Note), but otherwise coded data are easily shared within tools, much less between them.
+After videos are recorded, human raters evaluate videos and apply text-based annotations or numeric codes to videos.
+Raters use a wide-range of commercial (Noldus Observer, Mangold Interact, MaxQDA, V-Note), and academic (Transana, Elan, Datavyu) software tools, spreadsheets or even pencil and paper.
+Some tools (Datavyu, Mangold, V-Note) offer text-based export to interoperable file formats like comma-separated value (CSV), but the tools have no standard format for data storage or export.
+Some tools offer web/cloud based storage (e.g., Datavyu via Databrary, Transana, V-Note), but otherwise coded data are not easily shared within tools, much less between them.
 
@@
 Databrary has developed an access model described below that provides an example of how this might be accomplished.
+
 \subsection{Data Management}
@@
 data indefinitely.
+
 \subsection{Managing Data for Sharing}
@@
 \subsection{Building A Community That Embracing Sharing}
+
 Data sharing works only when the scientific community embraces it. 
@@
 \section{Conclusion}
+
 As Gesell once noted, cameras can record child behavior in ways that make it "...as tangible as tissue."38 The Databrary project will strengthen traditional ways of conducting video-based research by facilitating better coding, data mining, and data management practices within labs while building the culture and infrastructure to enable a new tradition of open data sharing in developmental science.
@@
 \section{Acknowledgments}
+
 Databrary is based on work supported by the National Science Foundation under Grant No.
@@
 
-\bibliography{example}
+\bibliography{paper}
 

### 2015-05-08 2015-06-01-nsf-ed-workshop/paper.tex
thru system design

@@
 Open data sharing is a scientific imperative and common practice in many areas of biomedical, physical, and earth sciences where it has accelerated the pace of discovery. 
-Unfortunately, research on human development and learning remains shrouded in a culture of isolation. 
+Unfortunately, despite the efforts of professional organizations to make data sharing a norm \cite[e.g.]{AERA-Code-2011} most research on human development and learning remains shrouded in a culture of isolation.
 Rather than providing direct access to raw data, researchers in the developmental and learning sciences share interpretations of distilled data almost exclusively through publications and presentations.
@@
 The collection of video or other identifiable or sensitive information requires approval by a research ethics board.
-It also requires informed consent or assent from the participants, and possibly parental, teacher, and school district approval.
+It also requires informed consent from participants, and possibly parental, teacher, and school district approval.
 Researchers risk violating participants’ privacy if digital images are viewed or released to others without authorization.
@@
 These require tools for synchronization across camera views and data streams.
-The use of video creates a data explosion: A typical lab studying infant or child development collects 8-12 hours of video/week [cite18] in widely varied formats. 
+The use of video creates a data explosion: A typical lab studying infant or child development collects 8-12 hours of video/week in widely varied formats \cite{Survey2012}. 
 Thus, sharing digital video requires substantial storage capacity, significant computational resources for transcoding videos into common, preservable formats, and powerful search and streaming tools.
@@
 
-With few exceptions (Chen Yu; Gilmore \& Raudies, Rehg?, others), machine learning has not been applied to video data coding in the developmental and learning sciences.
-But, there is explosive growth in the use of machine learning and computer vision tools to tag images and videos in commercial settings.
-This may create opportunities for future projects that supply machine-generated tags to videos collected in educational and developmental research.
+There has been explosive growth in the use of machine learning and computer vision tools to tag images and videos in commercial settings %\cite{Google2014}. 
+But, with few exceptions %\cite{ChenYu2013, Raudies2014, Fathi2012}
+, these techniques have not been applied to video data coding in the developmental and learning sciences. 
+The application of machine-generated tags to research videos presents an as-yet-untapped opportunity to apply big data tools to educational and developmental research.
 
@@
 
-Sharing video recordings poses a unique challenge to existing data sharing policies because videos contain personally identifying information—specifically participants’ faces and voices and often the
-insides of their homes and classrooms. 
-Sharing personally identifiable information puts research participants and others depicted in recordings at increased risk for loss of privacy. 
-At the same time, blurring or altering original recordings to hide identities undermines or eliminates
-their value to other researchers. 
-Often, participants’ faces and voices produce the behaviors of interest.
-A challenge for future education research is to find ways to preserve the value of video for reuse by others while protecting the privacy of research participants.
-Databrary has developed an access model described below that provides an example of how this might be accomplished.
+Sharing video recordings poses a challenge to existing data sharing policies.
+De-identified data %\cite{HHS} 
+may often be openly shared, but since the faces and and voices contained in video are considered identifiable information, de-identifying video requires their removal.
+This severely diminishes the reuse value of video and poses additional costs and burdens on researchers.
+So, sharing video data openly requires the development of new policies that protect the privacy of research participants while preserving the integrity of raw video for reuse by others.
+Databrary has developed an access model that demonstrates how this might be accomplished by building on the well-established principle of informed consent and restricting access to authorized individuals who meet specific criteria.
+We discuss the access model in detail below.
 
 \subsection{Data Management}
-A primary impediment to sharing is researchers’ lack of time to find, label, clean, organize, and copy their files into formats that can be used and understood by others15. 
-Even scientists committed to data sharing find it difficult to do. Data management systems that make sharing convenient, reliable—and optimally automatic29—can help.
+
+As in other fields \cite{Overpeck2011}, developmental and education science is inundated by an explosion of digital data, most of which is inaccessible to other researchers.
+A primary impediment to sharing is researchers’ lack of time to find, label, clean, organize, and copy their files into formats that can be used and understood by others \cite{Ascoli2006}. 
+Even scientists committed to data sharing can find it difficult to do. 
+Data management systems that make sharing convenient, reliable—and optimally automatic—can help.
 
@@
 Despite similar research methods, study designs vary widely. 
-No two developmental science labs manage data in the same way. 
-Some studies are longitudinal, where researchers observe the same participants at
-multiple sessions. 
-Some are cross-sectional, where researchers observe
-each participant at only one session. 
-The timing of observations may be
-determined by participants’ age (e.g., 4-month-olds, 8-month-olds,
-12-month-olds) or abilities (e.g., preverbal, one-word utterances,
-sentences), experimentally determined variables (e.g., pre- and
-post-intervention), or other factors. 
-Some labs organize longitudinal data by grouping files first by participant and then by session date.
-Other labs organize longitudinal data by session and then by
-participant. 
-Still others organize longitudinal data based on task (book reading, block building, free play).
-Some researchers institute a central data management system to be followed by all the lab members, providing easier access and greater transparency for the entire lab but not necessarily providing the structure for similar benefits to
-researchers unfamiliar with the lab’s practices. 
+No two labs manage data in the same way. 
+Some studies are longitudinal, where researchers observe the same participants at multiple sessions. 
+Some are cross-sectional, where researchers observe each participant at only one session. 
+The timing of observations may by determined by participants’ age (e.g., kindergarteners) or abilities (e.g., readers), experimentally determined variables (e.g., pre- and post-intervention), or other factors.
 Other researchers allow their students to keep separate records, making it difficult to share
 data even within the lab. 
-Some researchers keep videos, metadata, and analyses together, and some do not. Idiosyncratic terms, record-keeping, and data management systems are the norm. 
-Databrary must enable researchers to discover and understand each other’s materials, regardless of the original investigator’s data management system.
-As in other fields22, developmental science is inundated by an explosion of digital data, most of which is inaccessible to other researchers.
-The data deluge stems from the advent of cheap, high-resolution digital cameras.
-The average developmental lab collects 12 hours of video/week23.
-Raw data without information about workflow and data provenance is essentially “orphaned” and cannot be understood by subsequent researchers who weren’t part of the original team24.
-Moreover, video files must be electronically linked with participant permissions and relevant metadata.
-An effective data management system for video does not currently exist.
-We will build one to allow researchers within labs to benefit from each other’s work, facilitate interdisciplinary collaborations, and provide a solid foundation for open sharing25
+Some researchers keep videos, metadata, and analyses together, and some do not. 
+Idiosyncratic terms, record-keeping, and data management systems are the norm. 
+Raw data without information about workflow and data provenance is essentially “orphaned” and cannot be understood by subsequent researchers who weren’t part of the original team \cite{curry2011}. 
+Moreover, even though video requires minimal metadata to be suitable for reuse, video files should be electronically linked to whatever relevant metadata do exist, and there must be information linked to the videos that determines what level of access is permitted.
+We describe below Databrary's data management system that enables users to manage the uploading of research videos and related metadata as these are collected.
 
@@
 
-Data sharing is an established best practice in some areas of educational research (AERA ethics guide), but it is not necessarily widespread.
+Data sharing is an established best practice in some areas of educational research \cite{AERA-Code-2011}, but it is not necessarily widespread.
+
+[SAY MORE OR CUT]
 
 \section{Databrary.org}
-Databrary (databrary.org) is a digital data library specialized for storing and sharing video and research data and metadata.
-Databrary is a joint project of New York University (NYU) and The Pennsylvania State University (Penn State). 
-The Databrary project began with a workshop on open video data sharing funded by the National
-Science Foundation (NSF) under Grant No.
-BCS-1139702 at which developmental scientists, computer scientists, library scientists, and federal agency program officers discussed the promises and challenges of
-sharing video data. 
-NSF and the National Institute of Child Health and Human Development (NICHD) have provided funding under grant No. BCS-1238599 and Cooperative Agreement U01-HD-076595, respectively. 
-The interdisciplinary team bring expertise in developmental and neural
-science, information technology, library science, software development,
-data curation, and community engagement. 
-In addition, a board of expert
-advisors composed of developmental scientists, library scientists, and
-leaders of data repositories in the behavioral and social sciences
-provide guidance.
-No large-scale repository for sharing video data currently exists. So, we will create a web-based Databrary repository.
+
+The Databrary project (databrary.org) arose to meet the challenges of sharing research video and deliver on the promise of open data sharing in educational and developmental science.
+With funding from NSF (BCS-1238599) and the National Institute of Child Health and Human Development (NICHD U01-HD-076595) the project has focused on building a data library specialized for video, creating data management tools, creating new policies to enable video sharing, and creating a community of researchers who embrace video sharing.
+A fifth aim focuses on the development of a free, open-source video annotation tool, Datavyu (http://datavyu.org).
+Our interdisciplinary team brings expertise in developmental and neural science, information technology, library science, software development, data curation, and community engagement. 
+The project receives guidance from a board of expert advisors composed of developmental scientists, library scientists, and leaders of data repositories in the behavioral and social sciences provide guidance.
+The project received funding in 2012-2013, began a private beta testing phase in the spring of 2014 and opened for full use in October 2015.
 
@@
 Data are preserved indefinitely in a secure data storage facility at NYU managed by central IT.
-There is no cost to use the system; an institutional subscription model is under development.
-All code is hosted on GitHub.
+There is no cost to use the system, but an institutional subscription model is under development.
+All code is hosted on GitHub (http://github.com/databrary).
 Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
 Video and audio data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4.
+This ensures that video and audio data can be streamed and downloaded by any operating system that supports a modern browser. 
 Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
+Since common video coding/tagging file formats are required for users to build on earlier coding efforts, Databrary plans to develop tools for converting files from other coding tools into a common format.
 
-Common video and coding file formats are required for users to build on earlier coding efforts, perform integrative analyses across shared files, and search effectively in a shared repository.
-We will develop tools for converting files from other coding tools into a common format for the Databrary.
+\subsection{Policies for Safe, Secure, \& Sharing}
 
-Behavior is rich and complex.
-Research interests and methods are diverse.
-Naturally, video coding is idiosyncratic.
-The only universal tags in developmental research are children’s age and sex.
-Thus, OpenSHAPA is powerful and flexible.
-It is designed to help researchers understand behavior—or any other time series—using codes of their choosing.
-As the PI's experience demonstrates, OpenSHAPA will strengthen work within labs.
-It will also provide the common coding format for files in the Databrary.
-\subsection{Policies for Safe, Secure, \& Sharing}
 Shared videos are maximally informative when participants’ faces are visible and their vocalizations are audible. Thus, anonymity cannot be ensured as it is for flat file or imaging data. 

### 2015-05-08 2015-06-01-nsf-ed-workshop/paper.tex
thru system design

@@
 \twoauthors{Rick O. Gilmore}{Karen E. Adolph, David S. Millman}
-\twoaffiliations{The Pennsylvania State University}{New York University}
+\twoaffiliations{The Pennsylvania State University, The Databrary Project}{New York University, The Databrary Project}
 
@@
 Video captures the complexity, richness, and diversity of behavior unlike any other measure. 
-As a result, large numbers of people who study teaching and learning or seek to improve it employ video. 
+As a result, large numbers of people who study teaching and learning employ video. 
 Video documents itself to a large degree, and therefore has significant potential for reuse by others.
 This potential remains largely unrealized because videos are rarely shared.
-Video often contains information about personal identities, so considerations about research ethics pose
-challenges to sharing. 
-The relatively large size of video files and diversity of formats and software tools for coding pose technical challenges. 
+Video often contains information about personal identities, so considerations about privacy and research ethics pose challenges to sharing. 
+The relatively large size of video files, diversity of formats, and incompatible software tools for coding pose technical challenges. 
 In this paper, we will describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
@@
 The application of insights from research to essential human needs such as improved educational policy and practice depends on the free flow of information. 
-Open data sharing promotes greater transparency and peer oversight, supports inquiry into data collection methods and measurement, stimulates new work, encourages diversity of analysis and opinion, allows for testing of new or alternative hypotheses, facilitates the education of new researchers, permits integrative analyses by creating new datasets from multiple sources, allows future research to build on earlier efforts, and enables exploration of topics not envisioned by the original investigators.
-Open data sharing increases the impact of public investments in research and leads to better and more effective public policy.
+Open data sharing embodies many virtues: It bolsters transparency and peer oversight, encourages diversity of analysis and opinion, accelerates the education of new researchers, stimulates the exploration of topics not envisioned by the original investigators.
+Open data sharing increases the impact of public investments in research, accelerates the pace of discovery, and leads to better and more effective public policy.
 Open data sharing is a scientific imperative and common practice in many areas of biomedical, physical, and earth sciences where it has accelerated the pace of discovery.
-Unfortunately, despite the efforts of professional organizations to make data sharing a norm \cite[e.g.]{AERA-Code-2011} most research on human development and learning remains shrouded in a culture of isolation.
+
+Unfortunately, despite the efforts of professional organizations to make data sharing a norm \cite{AERA-Code-2011} most research on human development and learning remains shrouded in a culture of isolation.
 Rather than providing direct access to raw data, researchers in the developmental and learning sciences share interpretations of distilled data almost exclusively through publications and presentations.
@@
 Many researchers and practitioners in the developmental, learning, and education sciences collect video as raw data.
-Video recordings capture what teachers, docents, parents, tutors and peers say and do and what learners say and do in response.
-The recordings capture these dynamics in ways that offer the potential to discover who did what, and when, why, and how it influenced learning.
+Video recordings capture what teachers, parents, tutors and peers say and do and what learners say and do in response.
+The recordings capture these dynamics in ways that offer the potential to discover who did what, and where, when, why, and how it influenced learning.
 Still, the open sharing of video data is rare, even within labs or among collaborators.
@@
 But, technical, ethical, and cultural barriers have made open sharing of video data rare.
-Videos come in varied formats and generage large files; these pose conversion and storage challenges.
+Videos come in varied formats and generate large files; these pose conversion and storage challenges.
 Personally identifying information can be removed from text-based data.
@@
 These characteristics mean that video has multiple use cases in the study of learning and development.
-In laboratory contexts, researchers use videos to study how \cite{Karasik2014}.
+In laboratory contexts, researchers use videos to study how \cite{Karasik2014} infants, children, and adults behave in response to natural or experimenter tasks.
 In educational contexts, educators, teacher trainers and evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance \cite{Blomberg2014, Masats20111151, Baecher2013189}.
-.
-Researchers may record videos of learners in classrooms \cite{Alibali2012} or museums to understand what teachers do and how audiences respond or how using video affects what museum-goers learn \cite{Bakken2015}.
-In lab \cite{Kretch2014}, classroom \cite{Mason2015, Prieto:2014:STC:2669485.2669543}, or museum settings video-based eye tracking systems provide information about where people look and for how long.
+Researchers record videos of learners in classrooms \cite{Alibali2012} to understand what teachers do and how audiences respond.
+Some explore how \emph{using} video affects what museum-goers learn \cite{Bakken2015}.
+In lab \cite{Kretch2014} and classroom \cite{Mason2015, Prieto:2014:STC:2669485.2669543} settings video-based eye tracking systems provide information about where people look and for how long.
 While not strictly video recordings, image streams from computer-based tasks or learning-rich video games may be combined with information about user keyboard or controller inputs to study real-time perfomance.
@@
 Video may even be used to study engagement by students enrolled in online/web-based courses or to monitor compliance with academic integrity rules.
+
 Because video closely mimics the multisensory experiences of live human observers, recordings collected by one person for a particular purpose may be readily understood by another person and reused for another purpose.
 In other words, video requires less metadata and less manual curation than other data types, but offers greater potential for reuse.
-Moreover, the success, growth, and popularity of YouTube and other video-based social media (Vimeo, Periscope, Meerkat, SnapChat) demonstrates that video garners high levels of user interest and attention for interpersonal communication and entertainment uses.
+Moreover, the success, growth, and popularity of YouTube and other video-based social media (e.g., Vimeo, Periscope, Meerkat, SnapChat) demonstrates that video garners high levels of user interest and attention for interpersonal communication and entertainment uses.
 The success of these services also suggests that web-based video storage and streaming systems are now sufficiently well developed to satisfy large-scale demand.
-The question for researchers and policymakers is how to seize the moment in order to realize video's substantial potential to improve teaching and learning.
+The question for researchers and policymakers is how to capitalize on this momentum in order to realize video's substantial potential to improve teaching and learning.
 
@@
 These include technical infrastructure, research ethics, privacy, and security, and data management. 
-Further, making open data sharing in education research an established norm requires changing understanding about how identifiable data might be shared and building a community of researchers and educators committed to making video data sharing work.
+Further, making open data sharing in education research an established norm requires changing minds. 
+Some researchers and research ethics officials believe that identifiable research data may not be shared at all, when in fact, open sharing of this sort of data can be achieved with proper informed consent.
 
 \subsection{Technical Challenges}
-
-Video poses two types of technical challenges related to the goal of increasing widespread sharing. 
-One concerns the requirements for storing, sharing, streaming, and preserving large numbers of videos. 
-The other concerns tools for extracting meaningful information from the episodes and behavior videos capture.
 
@@
 Many settings, like classrooms, require multiple camera views to capture the desired behaviors.
-Some questions involve the collection of multiple, temporally dense data streams (physiological recordings, motion tracking, eye tracking, verbal transcripts, keyboard/controller responses).
+Some questions involve the collection of multiple, temporally dense data streams (physiological recordings, motion tracking, eye tracking, verbal transcripts, keyboard/controller responses) alongside video.
 These require tools for synchronization across camera views and data streams.
@@
 Raters use a wide-range of commercial (Noldus Observer, Mangold Interact, MaxQDA, V-Note), and academic (Transana, Elan, Datavyu) software tools, spreadsheets or even pencil and paper.
-Some tools (Datavyu, Mangold, V-Note) offer text-based export to interoperable file formats like comma-separated value (CSV), but the tools have no standard format for data storage or export.
-Some tools offer web/cloud based storage (e.g., Datavyu via Databrary, Transana, V-Note), but otherwise coded data are not easily shared within tools, much less between them.
+Some tools (Datavyu, Mangold, V-Note) offer text-based export to interoperable formats like comma-separated value (CSV) files, but the tools lack standard formats for data storage or export.
+Some tools offer web/cloud-based storage (e.g., Datavyu via Databrary, Transana, V-Note), but otherwise coded data are not easily shared between users employing the same tools, much less between them.
 
-There has been explosive growth in the use of machine learning and computer vision tools to tag images and videos in commercial settings %\cite{Google2014}. 
-But, with few exceptions %\cite{ChenYu2013, Raudies2014, Fathi2012}
-, these techniques have not been applied to video data coding in the developmental and learning sciences. 
+Human-centered video annotation is likely to remain the primary mode of data coding and reduction for some time given the unique abilities of trained observers to capture important behavioral dimensions.
+However, there has been explosive growth in the use of machine learning and computer vision tools to tag images and videos \cite{Google2014}
+. 
+With few exceptions \cite{ChenYu2013,Raudies2014,Fathi2012}, 
+these techniques have not been applied to video data coding in the developmental and learning sciences. 
 The application of machine-generated tags to research videos presents an as-yet-untapped opportunity to apply big data tools to educational and developmental research.
@@
 Sharing video recordings poses a challenge to existing data sharing policies.
-De-identified data %\cite{HHS} 
+De-identified data \cite{HHS} 
 may often be openly shared, but since the faces and and voices contained in video are considered identifiable information, de-identifying video requires their removal.
@@
 So, sharing video data openly requires the development of new policies that protect the privacy of research participants while preserving the integrity of raw video for reuse by others.
-Databrary has developed an access model that demonstrates how this might be accomplished by building on the well-established principle of informed consent and restricting access to authorized individuals who meet specific criteria.
+Databrary has developed an access model that achieves this by building on the well-established principle of informed consent and restricting access to authorized individuals who meet specific criteria.
 We discuss the access model in detail below.
@@
 Even scientists committed to data sharing can find it difficult to do. 
-Data management systems that make sharing convenient, reliable—and optimally automatic—can help.
+Data management systems that make sharing convenient, reliable—and optimally automatic—could help.
 
-Existing data management practices pose challenges for video sharing.
+But, even with the availability and widespread adoption of suitable data management tools, existing data management practices pose challenges for video sharing.
 Despite similar research methods, study designs vary widely. 
@@
 The timing of observations may by determined by participants’ age (e.g., kindergarteners) or abilities (e.g., readers), experimentally determined variables (e.g., pre- and post-intervention), or other factors.
-Other researchers allow their students to keep separate records, making it difficult to share
-data even within the lab. 
+Some researchers maintain central, organized file systems, while others allow trainees to keep separate records, making it difficult to share data even within the lab. 
 Some researchers keep videos, metadata, and analyses together, and some do not. 
@@
 
-Data sharing is an established best practice in some areas of educational research \cite{AERA-Code-2011}, but it is not necessarily widespread.
-
-[SAY MORE OR CUT]
+Finall, we note that while data sharing is an established best practice in some areas of educational research \cite{AERA-Code-2011}, it is not widespread.
+The reasons are many \cite{Ascoli2006b,Ferguson2014}.
+Researchers cite intellectual property and privacy issues, the lack of data sharing requirements from funding agencies, and fears about the misuse, misinterpretation, or harm that might come from sharing.
+Researchers also note that data shaving diverts energy and resources from other scholarly activities that are more heavily and frequently rewarded \cite{Ascoli2006b,Ferguson2014}.
+Put another way, data sharing involves mostly sticks and few carrots.
+In order to make data sharing become a scientific norm, these factors must be addressed.
 
@@
 This ensures that video and audio data can be streamed and downloaded by any operating system that supports a modern browser.
+Preservation copies of original video and audio files are also stored.
 Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
@@
 
-\subsection{Policies for Safe, Secure, \& Sharing}
+\subsection{Policies for Safe \& Secure Video Sharing}
 
@@
 Further, Databrary provides access only when the peopledepicted have given permission for their information to be shared with other researchers.
+
 \subsubsection{Restricting Access}

### 2015-05-09 2015-06-01-nsf-ed-workshop/paper.tex
full draft

@@
 Even scientists committed to data sharing can find it difficult to do. 
-Data management systems that make sharing convenient, reliable—and optimally automatic—could help.
+Data management systems that make sharing convenient, reliable—and optimally automatic—would help.
 
@@
 The reasons are many \cite{Ascoli2006b,Ferguson2014}.
-Researchers cite intellectual property and privacy issues, the lack of data sharing requirements from funding agencies, and fears about the misuse, misinterpretation, or harm that might come from sharing.
+Researchers cite intellectual property and privacy issues, the lack of data sharing requirements from funding agencies, and fears about the misuse, misinterpretation, or professional harm that might come from sharing.
 Researchers also note that data shaving diverts energy and resources from other scholarly activities that are more heavily and frequently rewarded \cite{Ascoli2006b,Ferguson2014}.
@@
 
+No data sharing system would be complete without an information technology infrastructure that enables large numbers of video and related files to be uploaded, organized, stored, streamed, and tagged.
+Considerations about what file formats to support are also paramount.
 Databary is built on a PostreSQL database using the Scala Play framework and JavaScript.
@@
 
-\subsection{Policies for Safe \& Secure Video Sharing}
-
-Shared videos are maximally informative when participants’ faces are visible and their vocalizations are audible. Thus, anonymity cannot be ensured as it is for flat file or imaging data. 
-To address issues of confidentiality, access, and IRB approval for sharing video data, we will develop template data sharing permission forms and contributor and user agreements with input from the community, from human compliance officers, and from leaders of previous data sharing efforts.
-Databrary has elected to maximize the potential for data re-use by keeping recordings in their original
-unaltered form. 
-Instead of removing participants’ identities, Databrary restricts access to identifiable or sensitive data to authorized researchers. 
-Further, Databrary provides access only when the peopledepicted have given permission for their information to be shared with other researchers.
-
-\subsubsection{Restricting Access}
-Databrary provides access to shared data only to authorized researchers who have agreed to uphold common practices concerning the responsible and ethical use of identifiable and sensitive data. 
-To become an authorized investigator, applicants must register on the site and electronically sign an Access Agreement, which must also be co-signed by the applicant’s institution. 
-Full privileges will be granted only to researchers with principal investigator (PI) status
-at their institutions. 
-Other researchers may be granted privileges if they are affiliated with a PI who agrees to sponsor and supervise their application. 
-Initially there will be a manual process to identify the institutional representative—typically the authorizing official of the university—who can co-sign the Investigator Agreement. 
-However, as the user groups at each university expand, Databrary may implement
-administrative accounts at each institution. 
-This will enable the authorizing official to independently manage the authorizations of
-individual researchers at her institution.
-
-Data from a particular session may be stored in Databrary for the contributing researcher’s use whether the records are shared with other scientists or not. 
-When a researcher chooses to share, Databrary makes records openly available to the community of authorized researchers only if the people depicted in the recordings have given permission to release the data for sharing.
-Thus, Databrary requires that people depicted in recordings grant permission before their information can be
-shared.
-Databrary’s policies extend currently accepted principles of informed consent to the situation where participants are granted authority to consent to (or refuse) the release of their identifiable data.
-
-We developed these ideas in close collaboration with the NYU and PSU IRB staff. 
-To formalize the process of acquiring permission, we developed a Participant Release Form Template, based on photo or video release language many researchers use currently. 
-The template release form has standard language that Databrary recommends investigators should use
-with study participants. 
-This language makes it easy for participants to understand what is involved in sharing their video data, with whom it will be shared, and the potential risks associated with releasing their video and other identifiable data to other researchers.
-Use of the template also allows for the standardization of language associated with the release of identifiable or sensitive participant data.
-
-Some IRBs may deem an investigator’s existing, approved video or photo release form equivalent to the Databrary release. 
-This enables a researcher to share with Databrary recordings they have already
-collected. 
-However, most researchers will need to modify their research protocols, by adding the Databrary sharing permission procedures, prior to collecting new shareable video data. 
-Databrary staff are available to advise potential data contributors about how to amend existing research
-protocols so that the information acquired is Databrary-compliant.
-Protocol amendments involve seeking approval for use of the Databrary
-template release form and modifying the time period over which collected
-data will be made available. 
-Specifically, researchers must remove any clauses in research consent documents that require data destruction
-after some fixed period of time since Databrary intends to store shared
-data indefinitely.
-
-\subsection{Managing Data for Sharing}
-Raw video is most informative when it is linked with coding spreadsheets, codebooks, protocols, statistical analyses, and manuscripts. 
-A data management system is required to link related files, document workflow and data provenance, and tag files with appropriate permission levels for sharing. 
-Consistent data management practices will strengthen work within labs and among collaborators, enable convenient and reliable file uploading, and enhance the value of shared data in the Databrary.
-
-The 2011 NSF workshop8 and PI Adolph's MacSHAPA/OpenSHAPA workshops over the last decade have revealed widely varying laboratory data management practices.
-Some minimal standardization across labs is required to make data sharing feasible.
-In the biomedical sciences, use of laboratory information systems (LIMS) is increasingly common and shows promise for lowering barriers to sharing in other fields15. 
-Consistent with our overall approach, we will research, adapt, and adopt best laboratory management practices and open-source tools from other fields34, and we will prioritize implementation based on the preferences and priorities of the developmental science community.
-
-The full expressive flexibility of the data model is not evident to Databrary users. 
-Instead, users see different aspects of the data model at different times, depending on the context. Researchers organize their materials by acquisition date and time into structures called sessions.
+The system's data model embodies flexibility. 
+Researchers organize their materials by acquisition date and time into structures called sessions.
 A session corresponds to a unique recording episode and contains one or more recordings or related flat files (assets). 
@@
 participants that comprise the session. 
-In the data model, the aggregate of these flexible groupings is called a volume. 
-In practice, researchers combine sessions or segments of sessions within and across datasets to
-form the raw material that are subsequently described in published
-articles and presentations. 
+The aggregate of these flexible groupings is called a \emph{volume}. 
+In practice, researchers combine sessions or segments of sessions within and across datasets to form the raw material that are subsequently described in published articles and presentations. 
 Thus, Databrary contributors can combine sessions or segments within and across datasets with ancillary
-materials, such as coding manuals, Datavyu spreadsheets, statistical analyses, questionnaires, IRB documents, computer code, sample displays, and links to published journal articles in an aggregate structure
-investigators call a study. 
+materials, such as coding manuals, coding spreadsheets, statistical analyses, questionnaires, IRB documents, computer code, sample displays, and links to published journal articles in an aggregate structure investigators call a \emph{study}.
 Like datasets, studies are represented as volumes in the data model.
@@
 The discovery and browsing interfaces offer content items at the volume (study or dataset) level to users searching the library. 
-The contributor-assigned groupings (records), which may include relevant
-metadata about participants (measures, such as participant demographic information, domain-specific survey information, location data, condition variables, task descriptions, or other properties), allow a
-second level of filtering and organization within and between relevant
-identified studies or datasets.
+The contributor-assigned groupings may include metadata about participants, tasks or measures, and locations.
 This enables users to quickly identify data that may be relevant to their own interests or research.
+Users can also add keyword tags at the study/dataset, session, or segment level, and can endorse or deprecate keywords that have been suggested by others. 
+Different users may have very different reasons for citing a study or using data, so these features are critical for fostering resuse. 
+Future implementations may extend the ability of authorized researchers to annotate studies with other kinds of metadata.
 
-Authorized Databrary Investigators can also add keyword tags at the study/dataset, session, or segment level, and can endorse or deprecate keywords that have been suggested by other authorized researchers.
-These
-abilities are critical because different researchers may have very different reasons for citing a study or using data. Future implementations may extend the ability of authorized researchers to
-annotate studies with other kinds of metadata.
+The Databrary team recognizes that other data repositories enforce strict metadata ontologies and that these Doing may have benefits in sub-domains of research when there is community consensus on core concepts.
+The Databrary team has chosen not to require strict metadata ontologies for several reasons. 
+We hope to reduce the pre-deposit curational demands on contributors and foster the rapid adoption of data sharing.Developing and achieving agreement on metadata ontologies can take significant amounts of time.
+Video data are so rich and complex that in many domains, researchers have not settled on standard definitions for particular behaviors and may have little current need for standard tasks, procedures, or terminology.
+However, Databrary empowers users to re-use tags and terminology by suggesting common or similar terms, without confining users to these suggestions. 
+User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not necessary for browsing and searching in most of the use cases we envision.
+The Databrary team believes that the effort spent creating this powerful and flexible system will enable it to be extended to a diverse communities engaged in educational research.
 
-The Databrary team recognizes that other data repositories enforce strict metadata ontologies and that doing so may have benefits in sub-domains of research when there is community consensus [12]. 
-We will support standard data coding ontologies among researchers, but only enforce standardization for a small set of standard tags such as study date, participant birth date, and sex.
-We will also encourage
-contributors to report race, ethnicity, primary language, language of
-dataset, and location of session. 
-The Databrary team chose not to require strict metadata ontologies for several reasons. We hope to
-reduce the pre-deposit curational demands on contributors, encourage the repurposing of shared data, and foster the rapid adoption of data sharing. 
-Developing and achieving agreement on metadata ontologies can take significant amounts of time. 
-Video data are so rich and complex that in many domains, researchers have not settled on standard definitions for particular behaviors and may have little current need for standard tasks, procedures, or terminology.
-However, we will also encourage users to re-use tags and terminology by suggesting common or similar terms, without confining users to these suggestions. 
-User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not
-necessary for browsing and searching in most of the use cases we
-envision.
+\subsection{Policies for Safe \& Secure Video Sharing}
 
+Equally important for the success of a video data sharing enterprise are policies for openly identifiable data in ways that securely preserve participant privacy.
+As discussed previously, videos cannot be easily or cheaply de-identified without undermining their reuse value.
+Databrary has chosen to maximize the potential for data re-use by keeping recordings in their original
+unaltered form.
+To make unaltered raw videos available to others for reuse, Databrary has developed a two-pronged access model that i) restricts access to authorized individuals, and ii) enables access to identifiable data only with the permission of the people depicted.
+
+To gain access to Databrary a person must register on the site and electronically sign an access agreement, which must be co-signed by the applicant's institution. 
+Applicants must agree uphold common practices concerning the responsible and ethical use of identifiable and sensitive data. 
+Full privileges will be granted only to researchers with independent researcher status at their institutions. 
+Others may be granted privileges if they are affiliated with researcher who agrees to sponsor and supervise their application.
+Ethics board or IRB approval is not required to gain access to Databrary since many use cases do not involve research.
+Approval is required for uses that involve research.
+Once authorized, a user may have full access to the site's shared data, and may browse, tag, download for later viewing, and conduct non- or pre-research activities.
+
+Unique among data repositories, the Databrary access agreement authorizes data use and contribution.
+However, users may only store materials on Databrary for which they have approval from an ethics board or IRB and for identifiable data like video, permission from the people depicted. 
+Databrary has extended the commonly understood principle of informed consent to participate in research to encompass permission to share data with other researchers.
+In other words, identifiable data may be shared if the participants agree.
+
+Data from a particular session may be stored in Databrary for the contributing researcher’s use whether the records are shared with other scientists or not. 
+When a researcher chooses to share, Databrary makes records openly available to the community of authorized researchers only if the people depicted in the recordings have given permission to release the data for sharing.
+To formalize the process of acquiring permission, Databrary has developed a Participant Release Form Template, based on photo or video release language many researchers use currently. 
+The template release form has standard language that Databrary recommends investigators should use with study participants. 
+This language makes it easy for participants to understand what is involved in sharing their video data, with whom it will be shared, and the potential risks associated with releasing their video and other identifiable data to other researchers.
+Use of the template also allows for the standardization of language associated with the release of identifiable or sensitive participant data.
+
+Some IRBs may deem an investigator’s existing, approved video or photo release form equivalent to the Databrary release. 
+This enables a researcher to share with Databrary recordings they have already collected. 
+However, most researchers will need to modify their research protocols, by adding the Databrary sharing permission procedures, prior to collecting new shareable video data. 
+Databrary staff are available to advise potential data contributors about how to amend existing research
+protocols so that the information acquired is Databrary-compliant.
+Protocol amendments involve seeking approval for use of the Databrary template release form and modifying the time period over which collected data will be made available. 
+Specifically, researchers must remove any clauses in research consent documents that require data destruction after some fixed period of time since Databrary intends to store shared data indefinitely.
+
+We developed these ideas in close collaboration with the NYU and PSU research ethics and administration official, and we took inspiration from the Human Connectome Project and Open Genome Project that have similarities.
+As of May 2015, Databrary has secured agreements with 55 institutions on behalf of 105 resesearchers representing a diverse array of entities in North and South America, Europe, and Australia.
+We believe that this model of sharing identifiable video data can be extended to other data types and other contexts.
+In the realm of education research, we recognize that the access model will have to adapt to meet the needs of users who are not affiliated with an institution as currently conceived and to address the role of other parties involved in the educational research enterprise, namely teachers, schools, and school districts.
+Nevertheless, we believe that the core ideas provide a strong foundation for enabling the widespread sharing of video and other identifiable data to the educational research community.
+
+\subsection{Managing Data for Sharing}
+
+As we discussed earlier, even when researchers are willing to share data and have the required approvals to do so, varying data management practices make curating data for sharing after a research project has finished a difficult, often unrewarded chore.
+In order to make the raw data informative to others, it must be linked with coding spreadsheets, codebooks, protocols, statistical analyses, and manuscripts.
+But, the organization scheme that works for one research team does not necessarily work for general audiences unfamiliar with the nuances of a research project.
 Our experiences curating and ingesting archival datasets have highlighted the considerable value of contributors entering raw video data into Databrary as soon as recordings are acquired. 
-Immediate uploading reduces the workload on investigators, minimizes the risk of
-data loss and corruption, and accelerates the speed with which materials
-become openly available. 
+Immediate uploading reduces the workload on investigators, minimizes the risk of data loss and corruption, and accelerates the speed with which materials become openly available. 
+Accordingly, Databrary has developed a data management system that empowers researchers to curate their own data \emph{as it is collected}.
+
+The system employs familiar, easy-to-use spreadsheet and timeline-based interfaces that allow users to upload videos, add metadata about tasks, settings, and participants, link related files, document workflow and data provenance, and tag files with appropriate permission levels for sharing.
 To encourage immediate uploading, Databrary provides a complete set of controls so that researchers can restrict
-access to their data to only their own labs or to other users of their
-choosing. 
+access to their data to only their own labs or to other users of their choosing. 
 Datasets can be shared at a later point when data collection and ancillary materials are complete, whenever the contributor is comfortable sharing, or when journals or funders require it.
-Databrary has published a Data Sharing Manifesto [13] that explains to researchers the Databrary philosophy. Standards about when data should be shared are evolving. 
-Our philosophy is consistent with concepts and practices in other domains where data sharing is the norm (e.g., www.iedata.org).
-Planned enhancements to the Datavyu tool will allow contributors to organize video files and metadata as part of the analysis process. 
-This will facilitate the contribution of packaged datasets to Databrary.
 
-Based on discussions at the NSF workshop, we expect to see most contributions after data collection, coding, and analyses are completed.
-Researchers have widely differing practices for designing codebooks and spreadsheets, managing data collections, coding, analyses, collecting and managing participant data (demographic, anthropometric, etc.), and documenting workflow and data provenance.
-Our goal for Years 2-3 is to incorporate this more detailed type of information into the data management system.
-Doing so will strengthen sharing within labs, expand the usefulness of Databrary, and enable more powerful searches.
-The data management system will be extended to organize and package other files associated with contributors (position, expertise), studies (codebooks, protocols, analysis spreadsheets graphs, manuscripts, presentations, grant proposals, etc.), and participants (coded spreadsheets, test results, demographics, etc.).
-The same basic ideas will apply— standard file naming conventions and data tagged by contributor, study, and participant level.
-OpenSHAPA and other applications will communicate with the data management tool.
-The data management tool will be used to stitch together all the sharable files that comprise a study.
-From data sharing to best lab practices.
-The developmental science community is hungry for new practices that will enhance research, as demonstrated by the number of committed contributors.
-LIMS in other fields embody additional functions.
-We envision that the data and tool sharing enabled by Databrary will spark interest in other lab management enhancements.
-For example, the data management system could facilitate participant recruitment and scheduling, task/project management and scheduling, workflow processing, and quality assurance (e.g., reliability testing).
-We plan to seek community input—via online comments, conference workshops on best lab practices, and other avenues—about which features should be the highest priorities for data management tool developers.
-Extending the data management tool to incorporate the highest priority features will be a focus of the front-end programmer in Years 4-5.
+It is too early to evaluate the impact of Databrary's data management tools on researcher's willingness to share data.
+But, the developmental science community is hungry for new practices that will enhance research productivity, as demonstrated by the rapid growth in the number of authorized investigators.
+We are confident that consistent data management practices will strengthen work within labs and among collaborators, enable convenient and reliable file uploading, and enhance the value of shared data in the Databrary.
 
-\subsection{Building A Community That Embracing Sharing}
+\subsection{Building A Community That Embraces Sharing}
 
 Data sharing works only when the scientific community embraces it.
-To make video-based data sharing succeed, we will build a diverse contributor/user community committed to open data sharing. 
-Based on continual community input, we will expand the OpenSHAPA video coding tool, build data management tools, and develop tools for uploading, browsing, and searching data in the Databrary repository.
-These tools will enable sharing within labs, among collaborators, and among users in the larger community.
-To avoid pitfalls of previous data sharing efforts, leaders in data sharing, developmental science, and computer science will guide our work through an advisory board.
-To ensure continued community engagement, Databrary personnel will provide technical support to individual labs and provide information to the community about new data contributions, citations of shared data, and use cases.
+From the beginning, Databrary has sought to cultivate a community of researchers who support data sharing and commit to enacting that support in their own work flows.
+Those efforts involve many interacting components.
+They include active engagement with professional associations, conference-based exhibits and training workshops, communications with research ethics and administration staff, talks and presentations to diverse audiences, and one-on-one consultations with individual researchers and research teams.
+These activities are time and labor-intensive, but we believe that they are critical if we are to change community attitudes toward data sharing in the educational and learning sciences.
+Looking ahead, it will be critical to engage professional organizations that have already made fostering video data sharing a priority in the effort to forge community consensus about the importance, feasibility, and potential of open video data sharing.
 
@@
 
-As Gesell once noted, cameras can record child behavior in ways that make it "...as tangible as tissue."38 The Databrary project will strengthen traditional ways of conducting video-based research by facilitating better coding, data mining, and data management practices within labs while building the culture and infrastructure to enable a new tradition of open data sharing in developmental science.
-The Databrary project is potentially transformative.
-We will increase transparency by sharing the traditionally private activity of data collection, coding, and analysis.
-We will speed progress by enabling later researchers to benefit from previous work.
-Ultimately, we will better understand human development and how to transform knowledge into improvements in health.
+As Gesell once noted, cameras can record behavior in ways that make it "...as tangible as tissue." \cite{Scott2011}.
+
+Video has considerable untapped potential for making tangible the anatomy of successful teaching and learning.
+Realizing this potential requires a multifaceted approach that reduces barriers to sharing video and fosters new community values that embrace it.
+The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread video data sharing within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.,
+Databrary suggests ways that video and other identifiable data collected in the context of education research might also be shared.
+Widespread open sharing of high value, high impact data promises to advance education policy and improve practice.
+The Databrary team welcomes collaborations with other like-minded parties who share our vision \cite{Adolph2012} for a future where open data sharing is the norm in the behavioral and social sciences.
 

### 2015-05-09 2015-06-01-nsf-ed-workshop/paper.tex
chg acknowledgments to author note and move to front

@@
 }
+
+\authornote{Rick O.
+Gilmore is in the Department of Psychology, The Pennsylvania State University, University Park, PA 16802, rogilmore@psu.edu.
+Databrary is based on work supported by the National Science Foundation under Grant No.
+BCS-1238599, the Eunice Kennedy Shriver National Institute of Child Health and Human Development under Cooperative Agreement U01-HD-076595, and the Society for Research in Child Development.
+Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author(s) and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.}
 
@@
 User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not necessary for browsing and searching in most of the use cases we envision.
+Future challenges include building the capacity to searching for tagged segments inside of videos.
+Some of within-video tagging functionality exists in the current software, with more extensive search and preview capabilities on the near horizon.
+
 The Databrary team believes that the effort spent creating this powerful and flexible system will enable it to be extended to a diverse communities engaged in educational research.
@@
 
-\section{Acknowledgments}
-
-Databrary is based on work supported by the National Science Foundation under Grant No.
-BCS-1238599, the Eunice Kennedy Shriver National Institute of Child Health and Human Development under Cooperative Agreement U01-HD-076595, and the Society for Research in Child Development.
-Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author(s) and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.
-
 \bibliography{paper}

### 2015-05-09 2015-06-01-nsf-ed-workshop/paper.tex
add Nosek2012, edit future of technology and conclusion

@@
 
-Unfortunately, despite the efforts of professional organizations to make data sharing a norm \cite{AERA-Code-2011} most research on human development and learning remains shrouded in a culture of isolation.
+Unfortunately, despite the efforts of professional organizations to make data sharing a norm \cite{AERA-Code-2011} most research on human development and learning \cite{Adolph2012} and in the social sciences in general \cite{Nosek2012} remains shrouded in a culture of isolation.
 Rather than providing direct access to raw data, researchers in the developmental and learning sciences share interpretations of distilled data almost exclusively through publications and presentations.
@@
 User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not necessary for browsing and searching in most of the use cases we envision.
+
 Future challenges include building the capacity to searching for tagged segments inside of videos.
 Some of within-video tagging functionality exists in the current software, with more extensive search and preview capabilities on the near horizon.
-
-The Databrary team believes that the effort spent creating this powerful and flexible system will enable it to be extended to a diverse communities engaged in educational research.
+A related challenge involves developing mechanisms that enable files from video coding tools to be imported into Databrary so that user supplied codes may be visualized independent of the desktop coding tool deployed in a particular project.
+We envision a parallel set of export functions that permit full interoperability among coding tools.
+Since some tools use proprietary data formats, the priority will be to create interoperability with tools using open formats.
+The Databrary team believes that the effort spent creating this powerful and flexible system will enable it to be extended to a diverse set of communities engaged in educational research who collect and seek to share video.
 
@@
 Equally important for the success of a video data sharing enterprise are policies for openly identifiable data in ways that securely preserve participant privacy.
-As discussed previously, videos cannot be easily or cheaply de-identified without undermining their reuse value.
-Databrary has chosen to maximize the potential for data re-use by keeping recordings in their original
+As discussed previously, videos cannot be easily or inexpensively deidentified without undermining their reuse value.
+Databrary has chosen to maximize the potential for data reuse by keeping recordings in their original
 unaltered form.
@@
 Realizing this potential requires a multifaceted approach that reduces barriers to sharing video and fosters new community values that embrace it.
-The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread video data sharing within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.,
+The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread video data sharing within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
 Databrary suggests ways that video and other identifiable data collected in the context of education research might also be shared.
-Widespread open sharing of high value, high impact data promises to advance education policy and improve practice.
-The Databrary team welcomes collaborations with other like-minded parties who share our vision \cite{Adolph2012} for a future where open data sharing is the norm in the behavioral and social sciences.
+Technologies and policies for providing secure access to videos for broader use cases will have to be developed, and tools that allow desktop coding software files to be seamlessly imported to and exported from Databrary will have to be perfected.
+Communities of scholars dedicated to sharing education-related video will have to be cultivated.
+But, we believe that the widespread sharing of high value, high impact data of the sort that video can provide promises to advance education policy and improve practice.
+The Databrary team welcomes collaborations with other like-minded parties who share our vision \cite{Adolph2012} for a future where open data sharing is the norm.
 

### 2015-05-09 2015-06-01-nsf-ed-workshop/paper.tex
refs to intro; tweaks

@@
 Databrary suggests ways that video and other identifiable data collected in the context of education research might be shared.
-Widespread open sharing of high value, high impact data will advance education policy and improve practice.
+Widespread open sharing of high value, high impact data of the sort provided by video will advance education policy and improve practice.
 }
@@
 
-The application of insights from research to essential human needs such as improved educational policy and practice depends on the free flow of information. 
-Open data sharing embodies many virtues: It bolsters transparency and peer oversight, encourages diversity of analysis and opinion, accelerates the education of new researchers, stimulates the exploration of topics not envisioned by the original investigators.
+The application of insights from scientific research to essential human needs such as improved educational policy and practice depends on the free flow of information. 
+Open data sharing embodies many virtues: It bolsters transparency and peer oversight, encourages diversity of analysis and opinion, accelerates the education of new researchers, and stimulates the exploration of topics not envisioned by the original investigators.
 Open data sharing increases the impact of public investments in research, accelerates the pace of discovery, and leads to better and more effective public policy.
-Open data sharing is a scientific imperative and common practice in many areas of biomedical, physical, and earth sciences where it has accelerated the pace of discovery.
+Open data sharing is a scientific imperative \cite{NSF2011,NIMH2015} and common practice in many areas of biomedical \cite{Kaye2009}, physical \cite{Young2011}, and earth sciences \cite{Kleiner2011} where it has accelerated the pace of discovery.
 
-Unfortunately, despite the efforts of professional organizations to make data sharing a norm \cite{AERA-Code-2011} most research on human development and learning \cite{Adolph2012} and in the social sciences in general \cite{Nosek2012} remains shrouded in a culture of isolation.
+Unfortunately, despite the efforts of funding agencies and professional organizations \cite{AERA-Code-2011} to make data sharing a norm, most research on human development and learning \cite{Adolph2012} and in the social sciences in general \cite{Nosek2012} remains shrouded in a culture of isolation.
 Rather than providing direct access to raw data, researchers in the developmental and learning sciences share interpretations of distilled data almost exclusively through publications and presentations.
@@
 . 
-With few exceptions \cite{ChenYu2013,Raudies2014,Fathi2012}, 
+With few exceptions \cite{Amso2014,ChenYu2013,Raudies2014,Fathi2012}, 
 these techniques have not been applied to video data coding in the developmental and learning sciences. 
@@
 
-Finall, we note that while data sharing is an established best practice in some areas of educational research \cite{AERA-Code-2011}, it is not widespread.
+Finally, we note that while data sharing is an established best practice in some areas of educational research \cite{AERA-Code-2011}, it is not widespread.
 The reasons are many \cite{Ascoli2006b,Ferguson2014}.
@@
 The system's data model embodies flexibility. 
-Researchers organize their materials by acquisition date and time into structures called sessions.
+Researchers organize their materials by acquisition date and time into structures called \emph{sessions}.
 A session corresponds to a unique recording episode and contains one or more recordings or related flat files (assets). 

### 2015-05-10 2015-06-01-nsf-ed-workshop/paper.tex
more blue sky thoughts in conclusion

@@
 
+Imagine a time in the near future when researchers interested in studying classroom teaching and learning can mine an integrated, synchronized, interoperable, and open data set.
+That data set includes multiple measurements -- video from multiple cameras, eye tracking, motion, and physiological measurements, and both historical and real-time student performance data.
+Imagine that this classroom-level data can be linked with grade, school, neighborhood, community, region, and state-level data about education practice, curriculum, and policy.
+Then, imagine training a cadre of experts with skills in the data science of learning and education who are made keenly sensitive to privacy, confidentiality and ethical issues involved in research involving sensitive, identifiable information.
+We then empower these learning scientists to extract from the data meaningful insights about how educational practice and policy might be improved.
+In short, imagine a science of teaching and learning that can be personally tailored to individuals in ways analogous to the impact of big data on medicine \cite{BBC2014}.
+The barriers to realizing this vision are similar to those that confront the vision of personalized medicine -- the development of technologies that enable data to be collected, synchronized, tagged, curated, stored, shared, linked, and aggregated, policies and practices that ensure security and individual privacy, and the cultivation of professional expertise needed to turn raw data into actionable insights.
+
 As Gesell once noted, cameras can record behavior in ways that make it "...as tangible as tissue." \cite{Scott2011}.
-
-Video has considerable untapped potential for making tangible the anatomy of successful teaching and learning.
-Realizing this potential requires a multifaceted approach that reduces barriers to sharing video and fosters new community values that embrace it.
-The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread video data sharing within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
+The Databrary team contends that video has a central role to play in efforts to make tangible the anatomy of successful teaching and learning.
+In fact, we argue that video could be the core around which other measures of teaching and learning cluster.
+This requires reducing barriers to sharing video and related data and one that fosters new community values around ethical data sharing.
+The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread sharing within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
 Databrary suggests ways that video and other identifiable data collected in the context of education research might also be shared.
-Technologies and policies for providing secure access to videos for broader use cases will have to be developed, and tools that allow desktop coding software files to be seamlessly imported to and exported from Databrary will have to be perfected.
-Communities of scholars dedicated to sharing education-related video will have to be cultivated.
+Technologies and policies for providing secure access to videos for broader use cases will have to be developed,  tools that allow desktop coding software files to be seamlessly converted to and from one another will have to be perfected, and ways of synchronizing and linking disparate data streams will have to be created.
+Equally important, communities of scholars dedicated to collecting, sharing, and mining education-related video data will have to be cultivated.
 But, we believe that the widespread sharing of high value, high impact data of the sort that video can provide promises to advance education policy and improve practice.
-The Databrary team welcomes collaborations with other like-minded parties who share our vision \cite{Adolph2012} for a future where open data sharing is the norm.
+The Databrary team welcomes collaborations with other like-minded parties who share our vision \cite{Adolph2012} for a future where open video data sharing is the norm and a personalized science of teaching and learning is the goal.
 

### 2015-05-10 2015-06-01-nsf-ed-workshop/paper.tex
full readthru & edit

@@
 Video captures the complexity, richness, and diversity of behavior unlike any other measure. 
-As a result, large numbers of people who study teaching and learning employ video. 
+As a result, large numbers of people who study teaching and learning or train and evaluate teachers employ video. 
 Video documents itself to a large degree, and therefore has significant potential for reuse by others.
 This potential remains largely unrealized because videos are rarely shared.
-Video often contains information about personal identities, so considerations about privacy and research ethics pose challenges to sharing. 
-The relatively large size of video files, diversity of formats, and incompatible software tools for coding pose technical challenges. 
-In this paper, we will describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
-Databrary suggests ways that video and other identifiable data collected in the context of education research might be shared.
-Widespread open sharing of high value, high impact data of the sort provided by video will advance education policy and improve practice.
+Video contains information about personal identities.
+This poses challenges to sharing. 
+The large size of video files, diversity of formats, and incompatible software tools pose technical challenges. 
+In this paper, we will describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community.
+Databrary has developed solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and their metadata.
+The Databrary experience suggests ways that video and other identifiable data collected in the context of education research might be shared.
+We imagine a science of teaching and learning, with video as its core, that allows educational experiences to be tailored to students in ways that big data promises to personalize medicine.
+The creation of repositories that enable the open sharing of richly informative, high value, high impact data about teaching and learning will help realize this ambitious vision.
 }
@@
 
-The application of insights from scientific research to essential human needs such as improved educational policy and practice depends on the free flow of information. 
+The application of insights from scientific research to essential human needs on the free flow of information and the open sharing of data. 
 Open data sharing embodies many virtues: It bolsters transparency and peer oversight, encourages diversity of analysis and opinion, accelerates the education of new researchers, and stimulates the exploration of topics not envisioned by the original investigators.
@@
 
-Unfortunately, despite the efforts of funding agencies and professional organizations \cite{AERA-Code-2011} to make data sharing a norm, most research on human development and learning \cite{Adolph2012} and in the social sciences in general \cite{Nosek2012} remains shrouded in a culture of isolation.
-Rather than providing direct access to raw data, researchers in the developmental and learning sciences share interpretations of distilled data almost exclusively through publications and presentations.
-The path from raw data to finding to conclusion cannot be traced or validated by others, nor can new questions be posed by others building on the same, often expensive-to-collect, raw data.
-To realize significant advances in our understanding of teaching and learning and for those insights to have meaningful impact on educational practice and policy, open data sharing must become an established norm, not the rare exception.
-Moreover, we must find ways to share the most valuable raw data, the data with the greatest potential for reuse by others, and the data with the greatest potential for yielding meaningful insights into the learning process.
+Unfortunately, despite the efforts to make data sharing a norm in the social and educational sciences \cite{AERA-Code-2011, Nosek2012}, most research on human development and learning \cite{Adolph2012} remains shrouded in a culture of isolation.
+Researchers share interpretations of distilled, not raw data, almost exclusively through publications and presentations.
+The path from raw data to finding to conclusion cannot be traced or validated by others, nor can new questions be posed by others building on the same, often expensive-to-collect, raw material.
+To realize significant advances in our understanding of teaching and learning and for those insights to influence educational practice and policy, open data sharing must become an established norm, not the rare exception.
+We must find ways to share raw data with the greatest potential for reuse by others for yielding meaningful insights into the learning process.
 
@@
 Video recordings capture what teachers, parents, tutors and peers say and do and what learners say and do in response.
-The recordings capture these dynamics in ways that offer the potential to discover who did what, and where, when, why, and how it influenced learning.
-Still, the open sharing of video data is rare, even within labs or among collaborators.
-Video data should be conducive to sharing.
-Video captures behavior that others can readily inspect.
-It captures multiple dimensions of complex real-world learning settings in ways that others can easily build upon.
+The recordings capture these dynamics in ways that shed light on who did what, and where, when, why, and how it influenced learning.
+% Still, the open sharing of video data is rare, even within labs or among collaborators.
+% Video data should be conducive to sharing.
+% Video captures behavior that others can readily inspect.
+% It captures multiple dimensions of complex real-world learning settings in ways that others can easily build upon.
 But, technical, ethical, and cultural barriers have made open sharing of video data rare.
 Videos come in varied formats and generate large files; these pose conversion and storage challenges.
-Personally identifying information can be removed from text-based data.
-Videos may contain faces, voices, names spoken aloud, views or voices of non-participants, views of classroom interiors and sometimes views of the homes of research participants. 
-These elements cannot be removed without reducing the information content and reuse value to others. 
-The collection of video or other identifiable or sensitive information requires approval by a research ethics board.
-It also requires informed consent from participants, and possibly parental, teacher, and school district approval.
-Researchers risk violating participants’ privacy if digital images are viewed or released to others without authorization.
+Personally identifying information can be removed from text-based data, but videos contain personally identifying information that cannot be removed without reducing the potential for reuse value to others.
+Identifiable information about research participants may not be shared without without consent and ethics board or IRB approval.
 For these and other reasons, no culture of widespread open sharing of video data has emerged in the education, learning, and developmental sciences.
 
-The Databrary project has built a digital data library specialized for the open sharing of video data from research in these areas of inquiry.
-In this paper, we will describe how Databrary has overcome the most significant barriers to sharing video, including solutions to maintaining participant privacy, data tagging, and data management.
-In developing technology and policies that enable the safe and secure sharing of video and other identifiable research data, Databrary suggests ways that video and other identifiable data collected in the context of research on teaching and learning might be shared.
-Fostering the open sharing and reuse of high value, high impact data like video promises to advance education policy and improve practice.
+The Databrary project has built a digital data library (http://databrary.org) specialized for the open sharing of video data.
+In this paper, we will describe how Databrary has overcome the most significant barriers to sharing video, including solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and their metadata
+In developing technology and policies that enable the safe and secure sharing of video data, Databrary suggests ways that comparable data collected in the context of research on teaching and learning might be shared.
+Fostering the open sharing and reuse of high value, high impact data promises to advance education policy and improve practice.
 
@@
 It captures when and where they look or gesture, when, where and how they move, and how their looking, gesturing, and moving corresponds to what others are doing.
-These characteristics mean that video has multiple use cases in the study of learning and development.
-In laboratory contexts, researchers use videos to study how \cite{Karasik2014} infants, children, and adults behave in response to natural or experimenter tasks.
-In educational contexts, educators, teacher trainers and evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance \cite{Blomberg2014, Masats20111151, Baecher2013189}.
+These characteristics mean that video has multiple use cases in the study of learning and development and a long history \cite{Goldman2007,Scott2011}.
+In laboratory contexts \cite{Karasik2014}, researchers use videos to study how infants, children, and adults behave in response to natural or experimenter-imposed tasks.
+Educators, teacher trainers and evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance \cite{Blomberg2014, Masats20111151, Baecher2013189}.
 Researchers record videos of learners in classrooms \cite{Alibali2012} to understand what teachers do and how audiences respond.
@@
 In other words, video requires less metadata and less manual curation than other data types, but offers greater potential for reuse.
-Moreover, the success, growth, and popularity of YouTube and other video-based social media (e.g., Vimeo, Periscope, Meerkat, SnapChat) demonstrates that video garners high levels of user interest and attention for interpersonal communication and entertainment uses.
+Moreover, the success, growth, and popularity of YouTube and other video-based social media (e.g., Vimeo, Periscope, Meerkat, Snapchat) demonstrates that video garners high levels of user interest and attention for interpersonal communication and entertainment uses.
 The success of these services also suggests that web-based video storage and streaming systems are now sufficiently well developed to satisfy large-scale demand.
@@
 Human-centered video annotation is likely to remain the primary mode of data coding and reduction for some time given the unique abilities of trained observers to capture important behavioral dimensions.
-However, there has been explosive growth in the use of machine learning and computer vision tools to tag images and videos \cite{Google2014}
-. 
+However, there has been explosive growth in the use of machine learning and computer vision tools to tag images and videos \cite{Google2014}. 
 With few exceptions \cite{Amso2014,ChenYu2013,Raudies2014,Fathi2012}, 
@@
 Sharing video recordings poses a challenge to existing data sharing policies.
-De-identified data \cite{HHS} 
-may often be openly shared, but since the faces and and voices contained in video are considered identifiable information, de-identifying video requires their removal.
+De-identified data \cite{HHS} may often be openly shared, but since the faces and and voices contained in video are considered identifiable information, de-identifying video requires their removal.
 This severely diminishes the reuse value of video and poses additional costs and burdens on researchers.
 So, sharing video data openly requires the development of new policies that protect the privacy of research participants while preserving the integrity of raw video for reuse by others.
-Databrary has developed an access model that achieves this by building on the well-established principle of informed consent and restricting access to authorized individuals who meet specific criteria.
-We discuss the access model in detail below.
+Sharing identifiable data from educational research faces similar challenges \cite{Asilomar2014}. 
+Databrary has developed an access model that builds on the well-established principle of informed consent and restricts access to authorized individuals who meet specific criteria.
+We discuss the access model in detail later.
 
@@
 Even scientists committed to data sharing can find it difficult to do. 
-Data management systems that make sharing convenient, reliable—and optimally automatic—would help.
+Data management systems that make sharing convenient, reliable—and optimally automatic—are needed.
 
 But, even with the availability and widespread adoption of suitable data management tools, existing data management practices pose challenges for video sharing.
-Despite similar research methods, study designs vary widely. 
-No two labs manage data in the same way. 
-Some studies are longitudinal, where researchers observe the same participants at multiple sessions. 
-Some are cross-sectional, where researchers observe each participant at only one session. 
-The timing of observations may by determined by participants’ age (e.g., kindergarteners) or abilities (e.g., readers), experimentally determined variables (e.g., pre- and post-intervention), or other factors.
-Some researchers maintain central, organized file systems, while others allow trainees to keep separate records, making it difficult to share data even within the lab. 
-Some researchers keep videos, metadata, and analyses together, and some do not. 
+Despite similar research methods, study designs vary widely, and no two labs manage data in the same way. 
+% Some studies are longitudinal, where researchers observe the same participants at multiple sessions. 
+% Some are cross-sectional, where researchers observe each participant at only one session. 
+% The timing of observations may by determined by participants’ age (e.g., kindergarteners) or abilities (e.g., readers), experimentally determined variables (e.g., pre- and post-intervention), or other factors.
+% Some researchers maintain central, organized file systems, while others allow trainees to keep separate records, making it difficult to share data even within the lab. 
+% Some researchers keep videos, metadata, and analyses together, and some do not. 
 Idiosyncratic terms, record-keeping, and data management systems are the norm. 
 Raw data without information about workflow and data provenance is essentially “orphaned” and cannot be understood by subsequent researchers who weren’t part of the original team \cite{curry2011}. 
-Moreover, even though video requires minimal metadata to be suitable for reuse, video files should be electronically linked to whatever relevant metadata do exist, and there must be information linked to the videos that determines what level of access is permitted.
-We describe below Databrary's data management system that enables users to manage the uploading of research videos and related metadata as these are collected.
+Moreover, even though video requires minimal metadata, video files must be electronically linked to whatever relevant metadata do exist, and there must be information linked to videos that specifies what level of access is permitted.
+We describe Databrary's data management system in a later section.
 
 \subsection{Changing Community Practices}
-
-Finally, we note that while data sharing is an established best practice in some areas of educational research \cite{AERA-Code-2011}, it is not widespread.
+Data sharing is an established best practice in some areas of educational and developmental research \cite{AERA-Code-2011}, but it is not widespread.
 The reasons are many \cite{Ascoli2006b,Ferguson2014}.
@@
 Put another way, data sharing involves mostly sticks and few carrots.
-In order to make data sharing become a scientific norm, these factors must be addressed.
+In order to make data sharing become a scientific norm, these barriers must be overcome.
 
 \section{Databrary.org}
-
 The Databrary project (databrary.org) arose to meet the challenges of sharing research video and deliver on the promise of open data sharing in educational and developmental science.
@@
 A fifth aim focuses on the development of a free, open-source video annotation tool, Datavyu (http://datavyu.org).
-Our interdisciplinary team brings expertise in developmental and neural science, information technology, library science, software development, data curation, and community engagement. 
-The project receives guidance from a board of expert advisors composed of developmental scientists, library scientists, and leaders of data repositories in the behavioral and social sciences provide guidance.
-The project received funding in 2012-2013, began a private beta testing phase in the spring of 2014 and opened for full use in October 2015.
+The project received funding in 2012-2013, began a private beta testing phase in the spring of 2014 and opened for public use in October 2015.
 
@@
 Considerations about what file formats to support are also paramount.
-Databary is built on a PostreSQL database using the Scala Play framework and JavaScript.
+Databary is a free, open-source (http://github.com/databrary) web application built on a PostreSQL database using the Scala Play framework and JavaScript.
 Data are preserved indefinitely in a secure data storage facility at NYU managed by central IT.
 There is no cost to use the system, but an institutional subscription model is under development.
-All code is hosted on GitHub (http://github.com/databrary).
 Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
@@
 
-The Databrary team recognizes that other data repositories enforce strict metadata ontologies and that these Doing may have benefits in sub-domains of research when there is community consensus on core concepts.
-The Databrary team has chosen not to require strict metadata ontologies for several reasons. 
-We hope to reduce the pre-deposit curational demands on contributors and foster the rapid adoption of data sharing.Developing and achieving agreement on metadata ontologies can take significant amounts of time.
+Other data repositories enforce strict metadata ontologies.
+Doing so may have benefits in sub-domains of research when there is community consensus on core concepts. 
+Databrary has chosen not to require strict metadata ontologies. 
+We hope to reduce the demands on contributors and foster the rapid adoption of data sharing.
+Achieving agreement on metadata ontologies can take significant amounts of time. 
 Video data are so rich and complex that in many domains, researchers have not settled on standard definitions for particular behaviors and may have little current need for standard tasks, procedures, or terminology.
-However, Databrary empowers users to re-use tags and terminology by suggesting common or similar terms, without confining users to these suggestions. 
+Instead, Databrary empowers users to re-use tags and terminology by suggesting common or similar terms, without confining users to these suggestions. 
 User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not necessary for browsing and searching in most of the use cases we envision.
@@
 Since some tools use proprietary data formats, the priority will be to create interoperability with tools using open formats.
-The Databrary team believes that the effort spent creating this powerful and flexible system will enable it to be extended to a diverse set of communities engaged in educational research who collect and seek to share video.
+Databrary also recognizes the need to develop open data standards and application program interfaces (APIs) that enable different types of data elements to be synchronized and linked.
+Nonetheless, we believe that Databrary is already sufficiently powerful and flexible to be extended to a diverse set of communities engaged in educational research who collect and seek to share video.
 
@@
 
-Equally important for the success of a video data sharing enterprise are policies for openly identifiable data in ways that securely preserve participant privacy.
-As discussed previously, videos cannot be easily or inexpensively deidentified without undermining their reuse value.
+Equally important for the success of a video data sharing repository are policies for openly sharing identifiable data in ways that securely preserve participant privacy.
+As discussed previously, videos cannot be deidentified without undermining their reuse value.
 Databrary has chosen to maximize the potential for data reuse by keeping recordings in their original
 unaltered form.
-To make unaltered raw videos available to others for reuse, Databrary has developed a two-pronged access model that i) restricts access to authorized individuals, and ii) enables access to identifiable data only with the permission of the people depicted.
+To make unaltered raw videos available to others for reuse, Databrary has developed a two-pronged access model that i) restricts access to authorized individuals, and ii) enables access to identifiable data only with the explicit permission.
 
-To gain access to Databrary a person must register on the site and electronically sign an access agreement, which must be co-signed by the applicant's institution. 
+To gain access to Databrary a person must register on the site and electronically sign an access agreement which must be co-signed by the applicant's institution. 
 Applicants must agree uphold common practices concerning the responsible and ethical use of identifiable and sensitive data. 
-Full privileges will be granted only to researchers with independent researcher status at their institutions. 
-Others may be granted privileges if they are affiliated with researcher who agrees to sponsor and supervise their application.
+Full privileges are be granted only to those with independent researcher status at their institutions. 
+Others may be granted privileges if they are affiliated with a researcher who agrees to sponsor their application and supervise their use.
 Ethics board or IRB approval is not required to gain access to Databrary since many use cases do not involve research.
@@
 To formalize the process of acquiring permission, Databrary has developed a Participant Release Form Template, based on photo or video release language many researchers use currently. 
-The template release form has standard language that Databrary recommends investigators should use with study participants. 
+The template release form has standard language that Databrary recommends investigators use with study participants. 
 This language makes it easy for participants to understand what is involved in sharing their video data, with whom it will be shared, and the potential risks associated with releasing their video and other identifiable data to other researchers.
@@
 
-Some IRBs may deem an investigator’s existing, approved video or photo release form equivalent to the Databrary release. 
-This enables a researcher to share with Databrary recordings they have already collected. 
-However, most researchers will need to modify their research protocols, by adding the Databrary sharing permission procedures, prior to collecting new shareable video data. 
-Databrary staff are available to advise potential data contributors about how to amend existing research
-protocols so that the information acquired is Databrary-compliant.
-Protocol amendments involve seeking approval for use of the Databrary template release form and modifying the time period over which collected data will be made available. 
+Databrary staff advise potential data contributors about how to amend existing research protocols so that the information acquired is Databrary-compliant.
+Protocol amendments involve seeking approval for use of the Databrary template release language and modifying the time period over which collected data will be made available. 
 Specifically, researchers must remove any clauses in research consent documents that require data destruction after some fixed period of time since Databrary intends to store shared data indefinitely.
 
-We developed these ideas in close collaboration with the NYU and PSU research ethics and administration official, and we took inspiration from the Human Connectome Project and Open Genome Project that have similarities.
-As of May 2015, Databrary has secured agreements with 55 institutions on behalf of 105 resesearchers representing a diverse array of entities in North and South America, Europe, and Australia.
-We believe that this model of sharing identifiable video data can be extended to other data types and other contexts.
-In the realm of education research, we recognize that the access model will have to adapt to meet the needs of users who are not affiliated with an institution as currently conceived and to address the role of other parties involved in the educational research enterprise, namely teachers, schools, and school districts.
-Nevertheless, we believe that the core ideas provide a strong foundation for enabling the widespread sharing of video and other identifiable data to the educational research community.
+Databrary developed these ideas in close collaboration with the NYU and PSU research ethics and administration officials.
+We also took inspiration from the Human Connectome Project and Open Genome Project that face similar challenges.
+As of May 2015, Databrary had secured agreements with 55 institutions on behalf of 105 resesearchers representing a diverse array of entities in North and South America, Europe, and Australia.
+We believe that Databrary's model of sharing identifiable video data can be extended to other data types and other contexts.
+In the realm of education research, we recognize that the access model will have to be adapted to meet the needs of users who are not affiliated with an institution as currently conceived.
+The role of other parties involved in the educational research enterprise, namely teachers, schools, and school districts will also have to be sorted out.
+Nevertheless, we believe that the core ideas provide a strong foundation for enabling the widespread sharing of video and other identifiable data among members of the educational research community.
 
@@
 
-As we discussed earlier, even when researchers are willing to share data and have the required approvals to do so, varying data management practices make curating data for sharing after a research project has finished a difficult, often unrewarded chore.
+As discussed earlier, even when researchers are willing to share data and have the required approvals to do so, varying data management practices make curating data for sharing after a research project has finished a difficult, often unrewarded chore.
 In order to make the raw data informative to others, it must be linked with coding spreadsheets, codebooks, protocols, statistical analyses, and manuscripts.
 But, the organization scheme that works for one research team does not necessarily work for general audiences unfamiliar with the nuances of a research project.
-Our experiences curating and ingesting archival datasets have highlighted the considerable value of contributors entering raw video data into Databrary as soon as recordings are acquired. 
+Databrary's experiences curating and ingesting archival datasets have highlighted the considerable value of contributors entering raw video data into Databrary as soon as recordings are acquired. 
 Immediate uploading reduces the workload on investigators, minimizes the risk of data loss and corruption, and accelerates the speed with which materials become openly available. 
-Accordingly, Databrary has developed a data management system that empowers researchers to curate their own data \emph{as it is collected}.
+Accordingly, Databrary has developed a data management system that empowers researchers to \emph{curate their own data as it is collected}.
 
@@
 To encourage immediate uploading, Databrary provides a complete set of controls so that researchers can restrict
-access to their data to only their own labs or to other users of their choosing. 
-Datasets can be shared at a later point when data collection and ancillary materials are complete, whenever the contributor is comfortable sharing, or when journals or funders require it.
+access to their own labs or to other users of their choosing. 
+Datasets can be openly shared with the broader research community at a later point when data collection and ancillary materials are complete, whenever the contributor is comfortable sharing, or when journals or funders require it.
 
 It is too early to evaluate the impact of Databrary's data management tools on researcher's willingness to share data.
-But, the developmental science community is hungry for new practices that will enhance research productivity, as demonstrated by the rapid growth in the number of authorized investigators.
-We are confident that consistent data management practices will strengthen work within labs and among collaborators, enable convenient and reliable file uploading, and enhance the value of shared data in the Databrary.
+But, the developmental science community is hungry for new practices that will enhance research productivity, as demonstrated by the rapid growth in the number of authorized Databrary investigators.
+We are confident that consistent, but flexible data management practices will strengthen work within labs and among collaborators, enable convenient and reliable file uploading, and enhance the value of shared data in Databrary.
 
@@
 They include active engagement with professional associations, conference-based exhibits and training workshops, communications with research ethics and administration staff, talks and presentations to diverse audiences, and one-on-one consultations with individual researchers and research teams.
-These activities are time and labor-intensive, but we believe that they are critical if we are to change community attitudes toward data sharing in the educational and learning sciences.
-Looking ahead, it will be critical to engage professional organizations that have already made fostering video data sharing a priority in the effort to forge community consensus about the importance, feasibility, and potential of open video data sharing.
+These activities are time and labor-intensive, but we believe that they are critical to change community attitudes toward data sharing in the educational and learning sciences.
+Looking ahead, it will be critical to engage funders, journals, and professional organizations in the effort to forge community consensus about the importance, feasibility, and potential of open video data sharing.
 
@@
 Imagine that this classroom-level data can be linked with grade, school, neighborhood, community, region, and state-level data about education practice, curriculum, and policy.
-Then, imagine training a cadre of experts with skills in the data science of learning and education who are made keenly sensitive to privacy, confidentiality and ethical issues involved in research involving sensitive, identifiable information.
+Then, imagine training a cadre of experts with skills in the data science of learning and education who are sensitive to privacy, confidentiality and ethical issues involved in research involving identifiable information.
 We then empower these learning scientists to extract from the data meaningful insights about how educational practice and policy might be improved.
@@
 The Databrary team contends that video has a central role to play in efforts to make tangible the anatomy of successful teaching and learning.
-In fact, we argue that video could be the core around which other measures of teaching and learning cluster.
-This requires reducing barriers to sharing video and related data and one that fosters new community values around ethical data sharing.
-The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread sharing within the developmental sciences community, including solutions to maintaining participant privacy, data tagging, and data management.
+In fact, we argue that video can be the core around which other measures of teaching and learning cluster.
+This requires reducing barriers to sharing video and related data and one that fosters new community values around the importance of data sharing.
+The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread sharing within the developmental sciences community. 
 Databrary suggests ways that video and other identifiable data collected in the context of education research might also be shared.
@@
 Equally important, communities of scholars dedicated to collecting, sharing, and mining education-related video data will have to be cultivated.
-But, we believe that the widespread sharing of high value, high impact data of the sort that video can provide promises to advance education policy and improve practice.
-The Databrary team welcomes collaborations with other like-minded parties who share our vision \cite{Adolph2012} for a future where open video data sharing is the norm and a personalized science of teaching and learning is the goal.
+But, we believe that the widespread sharing of high value, high impact data of the sort that video can provide promises to achieve this ambitious vision to advance education policy and improve practice.
+The Databrary team welcomes collaborations with other like-minded parties who share our vision \cite{Adolph2012} for a future where open video data sharing is the norm, a personalized science of teaching and learning is the goal, and what optimizes student learning is as tangible as tissue.
 

### 2015-05-14 2015-06-01-nsf-ed-workshop/paper.tex
updates

@@
 As a result, large numbers of people who study teaching and learning or train and evaluate teachers employ video. 
-Video documents itself to a large degree, and therefore has significant potential for reuse by others.
-This potential remains largely unrealized because videos are rarely shared.
+Video documents itself to a large degree.
+This presents significant potential for reuse by others.
+The potential remains largely unrealized because videos are rarely shared.
 Video contains information about personal identities.
@@
 The large size of video files, diversity of formats, and incompatible software tools pose technical challenges. 
-In this paper, we will describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community.
-Databrary has developed solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and their metadata.
+We describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community.
+Databrary has developed solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and associated metadata.
 The Databrary experience suggests ways that video and other identifiable data collected in the context of education research might be shared.
-We imagine a science of teaching and learning, with video as its core, that allows educational experiences to be tailored to students in ways that big data promises to personalize medicine.
-The creation of repositories that enable the open sharing of richly informative, high value, high impact data about teaching and learning will help realize this ambitious vision.
+We envision a data intensive science of teaching and learning, with video as its core, that allows educational experiences to be tailored to students in ways that big data promises to personalize medicine.
+The creation and support of repositories that enable the open sharing of dense, richly informative, high value, and high impact data about teaching and learning will help realize this ambitious vision.
 }
@@
 
-The application of insights from scientific research to essential human needs on the free flow of information and the open sharing of data. 
+The application of insights from scientific research to essential human needs requires the free flow of information and the open sharing of data. 
 Open data sharing embodies many virtues: It bolsters transparency and peer oversight, encourages diversity of analysis and opinion, accelerates the education of new researchers, and stimulates the exploration of topics not envisioned by the original investigators.
@@
 The path from raw data to finding to conclusion cannot be traced or validated by others, nor can new questions be posed by others building on the same, often expensive-to-collect, raw material.
-To realize significant advances in our understanding of teaching and learning and for those insights to influence educational practice and policy, open data sharing must become an established norm, not the rare exception.
-We must find ways to share raw data with the greatest potential for reuse by others for yielding meaningful insights into the learning process.
+To realize significant advances in our understanding of teaching and learning, open data sharing must become an established norm, not the rare exception.
+We must find ways to share raw data with the greatest potential for reuse by others in order to yield maximally meaningful insights into the learning process.
 
-Video is a uniquely rich medium for capturing in real time the interactions that occur in classrooms, muesuems, laboratory, home, and other informal learning settings.
+Video is a uniquely rich medium for capturing in real time the interactions that occur in classrooms, museums, laboratory, home, and other informal learning settings.
 Many researchers and practitioners in the developmental, learning, and education sciences collect video as raw data.
@@
 The recordings capture these dynamics in ways that shed light on who did what, and where, when, why, and how it influenced learning.
-% Still, the open sharing of video data is rare, even within labs or among collaborators.
-% Video data should be conducive to sharing.
-% Video captures behavior that others can readily inspect.
-% It captures multiple dimensions of complex real-world learning settings in ways that others can easily build upon.
-But, technical, ethical, and cultural barriers have made open sharing of video data rare.
+Unfortunately, technical, ethical, and cultural barriers have made the open sharing of video data rare.
 Videos come in varied formats and generate large files; these pose conversion and storage challenges.
-Personally identifying information can be removed from text-based data, but videos contain personally identifying information that cannot be removed without reducing the potential for reuse value to others.
-Identifiable information about research participants may not be shared without without consent and ethics board or IRB approval.
-For these and other reasons, no culture of widespread open sharing of video data has emerged in the education, learning, and developmental sciences.
+Videos contain personally identifying information that cannot be removed without reducing the potential for reuse value to others.
+For these reasons, no culture of widespread open sharing of video data has emerged in the education, learning, and developmental sciences.
 
 The Databrary project has built a digital data library (http://databrary.org) specialized for the open sharing of video data.
-In this paper, we will describe how Databrary has overcome the most significant barriers to sharing video, including solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and their metadata
+Here we describe how Databrary has overcome the most significant barriers to sharing video, including solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and their metadata.
 In developing technology and policies that enable the safe and secure sharing of video data, Databrary suggests ways that comparable data collected in the context of research on teaching and learning might be shared.
-Fostering the open sharing and reuse of high value, high impact data promises to advance education policy and improve practice.
+Fostering the open sharing and reuse of dense, richly informative, high value, and high impact data about teaching and learning promises to advance education policy and improve practice.
 
@@
 
-More than 1 billion users worldwide uploade more than 300 hours of video to YouTube every minute \cite{YouTube2015}. 
-Google, YouTube's owner, has built an immense data infrastructure to upload, store, convert, tag, and stream video. 
-The current scale of video collection in developmental and education contexts is undoubtedly smaller and more focused in purpose, but it is large and growing. 
+More than 1 billion users worldwide upload more than 300 hours of video to YouTube every minute \cite{YouTube2015}.The current scale of video collection in developmental and education contexts is undoubtedly smaller and more focused in purpose, but it is large and growing.
 The Gates Foundation-funded Measures of Effective Teaching (MET; http://www.metproject.org/) Project generated more than 1,000 full-length video segments from K-12 classrooms, covering core subjects such as mathematics, language arts from multiple camera angles.
-The data constituting tens of terabytes of storage are hosted at the University of Michigan (http://soe.mivideo.it.umich.edu/) and securely streamed to registered viewers across the country.
-Similarly, in just over a year of operation, the Databrary (http://databrary.org) digital library has collected more than 7,000 individual videos, representing 2,400 hours of recording, and featuring more than 1,800 infant, child, and adult participants.
+The data constituting tens of terabytes of storage are hosted at the University of Michigan (http://soe.mivideo.it.umich.edu/) and securely streamed to registered viewers.
+Similarly, in just over a year of operation, the Databrary (http://databrary.org) digital library has collected more than 7,000 individual videos, representing 2,400 hours of recording, featuring more than 1,800 infant, child, and adult participants.
 Databrary has more than 100 authorized researchers representing more than 50 institutions across the globe who access the videos.
@@
 It captures when and where they look or gesture, when, where and how they move, and how their looking, gesturing, and moving corresponds to what others are doing.
-These characteristics mean that video has multiple use cases in the study of learning and development and a long history \cite{Goldman2007,Scott2011}.
-In laboratory contexts \cite{Karasik2014}, researchers use videos to study how infants, children, and adults behave in response to natural or experimenter-imposed tasks.
-Educators, teacher trainers and evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance \cite{Blomberg2014, Masats20111151, Baecher2013189}.
-Researchers record videos of learners in classrooms \cite{Alibali2012} to understand what teachers do and how audiences respond.
-Some explore how \emph{using} video affects what museum-goers learn \cite{Bakken2015}.
+These characteristics explain video's long history of use \cite{Goldman2007,Scott2011} in the study of learning and development.
+Video is used in home and laboratory contexts \cite{Karasik2014}, to study how infants, children, and adults behave respond to natural or experimenter-imposed tasks.
+Educators, teacher trainers and teacher evaluators use video to record what other teachers do to help improve teaching practices or evaluate performance \cite{Blomberg2014, Masats20111151, Baecher2013189}.
+Researchers record videos of learners in classrooms \cite{Alibali2012} to understand what teachers do and how students respond or how \emph{using} video affects what museum-goers learn \cite{Bakken2015}.
 In lab \cite{Kretch2014} and classroom \cite{Mason2015, Prieto:2014:STC:2669485.2669543} settings video-based eye tracking systems provide information about where people look and for how long.
-While not strictly video recordings, image streams from computer-based tasks or learning-rich video games may be combined with information about user keyboard or controller inputs to study real-time perfomance.
-When combined with other quantitative and qualitative measures about learning outcomes, video can provide information about what kinds of teaching techniques boost learning for what kinds of students.
-Video may even be used to study engagement by students enrolled in online/web-based courses or to monitor compliance with academic integrity rules.
+While not strictly video recordings, image streams from computer-based tasks or learning-rich video games may be combined with information about user keyboard or controller inputs to study real-time performance.
+When combined with other quantitative and qualitative measures about learning outcomes, video can provide information about what kinds of teaching techniques boost learning and for what kinds of students.
 
 Because video closely mimics the multisensory experiences of live human observers, recordings collected by one person for a particular purpose may be readily understood by another person and reused for another purpose.
-In other words, video requires less metadata and less manual curation than other data types, but offers greater potential for reuse.
 Moreover, the success, growth, and popularity of YouTube and other video-based social media (e.g., Vimeo, Periscope, Meerkat, Snapchat) demonstrates that video garners high levels of user interest and attention for interpersonal communication and entertainment uses.
@@
 These include technical infrastructure, research ethics, privacy, and security, and data management. 
-Further, making open data sharing in education research an established norm requires changing minds. 
-Some researchers and research ethics officials believe that identifiable research data may not be shared at all, when in fact, open sharing of this sort of data can be achieved with proper informed consent.
+Further, making open data sharing in education research an established norm requires changing minds about the ethics and feasibility of sharing video data.
 
@@
 The size of the file varies by format and resolution, and recordings vary widely, making it essential for video storage systems to convert files to a standard format.
-Many settings, like classrooms, require multiple camera views to capture the desired behaviors.
-Some questions involve the collection of multiple, temporally dense data streams (physiological recordings, motion tracking, eye tracking, verbal transcripts, keyboard/controller responses) alongside video.
-These require tools for synchronization across camera views and data streams.
+Many settings, like classrooms, require multiple camera views to capture desired behaviors.
+Some questions involve the collection of multiple, temporally dense data streams (physiological recordings, motion tracking, eye tracking, verbal transcripts, keyboard/controller responses) time-locked to video.
 The use of video creates a data explosion: A typical lab studying infant or child development collects 8-12 hours of video/week in widely varied formats \cite{Survey2012}. 
@@
 Raters use a wide-range of commercial (Noldus Observer, Mangold Interact, MaxQDA, V-Note), and academic (Transana, Elan, Datavyu) software tools, spreadsheets or even pencil and paper.
-Some tools (Datavyu, Mangold, V-Note) offer text-based export to interoperable formats like comma-separated value (CSV) files, but the tools lack standard formats for data storage or export.
-Some tools offer web/cloud-based storage (e.g., Datavyu via Databrary, Transana, V-Note), but otherwise coded data are not easily shared between users employing the same tools, much less between them.
+Some tools (Datavyu, Mangold, V-Note) offer text-based export to interoperable formats, but the field lacks consensus about standard storage or export formats.
+Some tools offer web/cloud-based storage (e.g., Datavyu via Databrary, Transana, V-Note), but most coded data are not easily shared between users employing the same tools, much less between them.
 
 Human-centered video annotation is likely to remain the primary mode of data coding and reduction for some time given the unique abilities of trained observers to capture important behavioral dimensions.
-However, there has been explosive growth in the use of machine learning and computer vision tools to tag images and videos \cite{Google2014}. 
-With few exceptions \cite{Amso2014,ChenYu2013,Raudies2014,Fathi2012}, 
-these techniques have not been applied to video data coding in the developmental and learning sciences. 
-The application of machine-generated tags to research videos presents an as-yet-untapped opportunity to apply big data tools to educational and developmental research.
+However, there has been explosive growth in the use of machine learning and computer vision tools to tag images and videos \cite{Google2014} in some domains. 
+These techniques are beginning \cite{Amso2014,ChenYu2013,Raudies2014,Fathi2012} to be applied to video data coding in the developmental and learning sciences. 
+The application of machine-generated tags to research videos may present an as-yet-untapped opportunity to apply data intensive tools to educational and developmental research.
 
@@
 So, sharing video data openly requires the development of new policies that protect the privacy of research participants while preserving the integrity of raw video for reuse by others.
-Sharing identifiable data from educational research faces similar challenges \cite{Asilomar2014}. 
-Databrary has developed an access model that builds on the well-established principle of informed consent and restricts access to authorized individuals who meet specific criteria.
-We discuss the access model in detail later.
+Sharing identifiable data from other types of data intensive educational research faces similar challenges \cite{Asilomar2014}. 
+Databrary has developed an access model discussed later that builds on the well-established principle of informed consent and restricts access to authorized individuals who meet specific criteria.
 
@@
 
-As in other fields \cite{Overpeck2011}, developmental and education science is inundated by an explosion of digital data, most of which is inaccessible to other researchers.
-A primary impediment to sharing is researchers’ lack of time to find, label, clean, organize, and copy their files into formats that can be used and understood by others \cite{Ascoli2006}. 
-Even scientists committed to data sharing can find it difficult to do. 
-Data management systems that make sharing convenient, reliable—and optimally automatic—are needed.
+Like other fields \cite{Overpeck2011} education research is inundated by an explosion of digital data, most of which is inaccessible to other researchers.
+Researchers lack time to find, label, clean, organize, and copy their files into formats that can be used and understood by others \cite{Ascoli2006}. 
+Even scientists committed to data sharing can find it difficult. 
+Data management systems that make reduce the burdens of data sharing are desperately needed.
 
@@
 Despite similar research methods, study designs vary widely, and no two labs manage data in the same way. 
-% Some studies are longitudinal, where researchers observe the same participants at multiple sessions. 
-% Some are cross-sectional, where researchers observe each participant at only one session. 
-% The timing of observations may by determined by participants’ age (e.g., kindergarteners) or abilities (e.g., readers), experimentally determined variables (e.g., pre- and post-intervention), or other factors.
-% Some researchers maintain central, organized file systems, while others allow trainees to keep separate records, making it difficult to share data even within the lab. 
-% Some researchers keep videos, metadata, and analyses together, and some do not. 
 Idiosyncratic terms, record-keeping, and data management systems are the norm. 
-Raw data without information about workflow and data provenance is essentially “orphaned” and cannot be understood by subsequent researchers who weren’t part of the original team \cite{curry2011}. 
-Moreover, even though video requires minimal metadata, video files must be electronically linked to whatever relevant metadata do exist, and there must be information linked to videos that specifies what level of access is permitted.
-We describe Databrary's data management system in a later section.
+To interpret raw data requires information about workflows and data provenance \cite{curry2011}. 
+Moreover, even though video requires minimal metadata, video files must be electronically linked to what relevant metadata exist including information about what level of access is permitted. 
+Databrary has developed a data management system described in a later section.
 
@@
 Researchers cite intellectual property and privacy issues, the lack of data sharing requirements from funding agencies, and fears about the misuse, misinterpretation, or professional harm that might come from sharing.
-Researchers also note that data shaving diverts energy and resources from other scholarly activities that are more heavily and frequently rewarded \cite{Ascoli2006b,Ferguson2014}.
-Put another way, data sharing involves mostly sticks and few carrots.
-In order to make data sharing become a scientific norm, these barriers must be overcome.
+Data sharing diverts energy and resources from scholarly activities that are more heavily and frequently rewarded \cite{Ascoli2006b,Ferguson2014}.
+In order to make data sharing a scientific norm, these barriers must be overcome.
 
 \section{Databrary.org}
-The Databrary project (databrary.org) arose to meet the challenges of sharing research video and deliver on the promise of open data sharing in educational and developmental science.
-With funding from NSF (BCS-1238599) and the National Institute of Child Health and Human Development (NICHD U01-HD-076595) the project has focused on building a data library specialized for video, creating data management tools, creating new policies to enable video sharing, and creating a community of researchers who embrace video sharing.
-A fifth aim focuses on the development of a free, open-source video annotation tool, Datavyu (http://datavyu.org).
+The Databrary project (databrary.org) arose to meet the challenges of sharing research video and to deliver on the promise of open data sharing in educational and developmental science.
+With funding from NSF (BCS-1238599) and the National Institute of Child Health and Human Development (NICHD U01-HD-076595) Databrary has focused on building a data library specialized for video, creating data management tools, crafting new policies that enable video sharing, and fostering a community of researchers who embrace video sharing.
+A fifth aim has focused on the development of a free, open-source video annotation tool, Datavyu (http://datavyu.org).
 The project received funding in 2012-2013, began a private beta testing phase in the spring of 2014 and opened for public use in October 2015.
@@
 
-No data sharing system would be complete without an information technology infrastructure that enables large numbers of video and related files to be uploaded, organized, stored, streamed, and tagged.
-Considerations about what file formats to support are also paramount.
+No data sharing system would be complete without an information technology infrastructure that enables large numbers of video and related files to be uploaded, converted, organized, stored, streamed, and tagged.
 Databary is a free, open-source (http://github.com/databrary) web application built on a PostreSQL database using the Scala Play framework and JavaScript.
@@
 There is no cost to use the system, but an institutional subscription model is under development.
-Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab that facilitate data analysis.
+Databrary can house video, audio, PDF, spreadsheet, image, and text-based files (along with associated metadata), as well as executable scripts in Ruby, R and Matlab.
 Video and audio data are transcoded into standard and HTML5-compatible formats, currently H.264+AAC as MP4.
@@
 A session corresponds to a unique recording episode and contains one or more recordings or related flat files (assets). 
-Researchers can group sessions in different ways, using whatever groupings are appropriate for
-the particular dataset. 
-These groupings may also identify particular time segments of a recording to distinguish tasks, events, or
+Researchers may identify particular time segments of a recording to distinguish tasks, events, or
 participants that comprise the session. 
@@
 
-The discovery and browsing interfaces offer content items at the volume (study or dataset) level to users searching the library. 
+The discovery and browsing interfaces display content items at the volume (study or dataset) level to users searching the library. 
 The contributor-assigned groupings may include metadata about participants, tasks or measures, and locations.
@@
 Other data repositories enforce strict metadata ontologies.
-Doing so may have benefits in sub-domains of research when there is community consensus on core concepts. 
-Databrary has chosen not to require strict metadata ontologies. 
-We hope to reduce the demands on contributors and foster the rapid adoption of data sharing.
+Doing so may have benefits in sub-domains of research when there is community consensus on core concepts, but Databrary has chosen not to require strict metadata ontologies. 
 Achieving agreement on metadata ontologies can take significant amounts of time. 
+We aim to reduce the demands on contributors and foster the rapid adoption of data sharing.
 Video data are so rich and complex that in many domains, researchers have not settled on standard definitions for particular behaviors and may have little current need for standard tasks, procedures, or terminology.
 Instead, Databrary empowers users to re-use tags and terminology by suggesting common or similar terms, without confining users to these suggestions. 
-User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not necessary for browsing and searching in most of the use cases we envision.
+User communities within Databrary may eventually converge on common conceptual and metadata ontologies based on the most common (and commonly endorsed) keyword tags, but standardized ontologies are not necessary for most of the use cases we envision.
 
@@
 Ethics board or IRB approval is not required to gain access to Databrary since many use cases do not involve research.
-Approval is required for uses that involve research.
+Approval is of course required for uses that involve research.
 Once authorized, a user may have full access to the site's shared data, and may browse, tag, download for later viewing, and conduct non- or pre-research activities.
@@
 Databrary has extended the commonly understood principle of informed consent to participate in research to encompass permission to share data with other researchers.
-In other words, identifiable data may be shared if the participants agree.
 
-Data from a particular session may be stored in Databrary for the contributing researcher’s use whether the records are shared with other scientists or not. 
+Data from a particular session may be stored in Databrary for the contributing researcher’s use whether the records are shared with others or not. 
 When a researcher chooses to share, Databrary makes records openly available to the community of authorized researchers only if the people depicted in the recordings have given permission to release the data for sharing.
-To formalize the process of acquiring permission, Databrary has developed a Participant Release Form Template, based on photo or video release language many researchers use currently. 
-The template release form has standard language that Databrary recommends investigators use with study participants. 
-This language makes it easy for participants to understand what is involved in sharing their video data, with whom it will be shared, and the potential risks associated with releasing their video and other identifiable data to other researchers.
+To formalize the process of acquiring permission, Databrary has developed a Participant Release Template (https://databrary.org/access/policies/release-template.html), based on photo or video release language many researchers use currently.
+The template release has standard language that Databrary recommends investigators use with study participants. 
+This language helps participants understand what is involved in sharing their video data, with whom it will be shared, and the potential risks associated with releasing their video and other identifiable data to other researchers.
 Use of the template also allows for the standardization of language associated with the release of identifiable or sensitive participant data.
@@
 Databrary staff advise potential data contributors about how to amend existing research protocols so that the information acquired is Databrary-compliant.
-Protocol amendments involve seeking approval for use of the Databrary template release language and modifying the time period over which collected data will be made available. 
-Specifically, researchers must remove any clauses in research consent documents that require data destruction after some fixed period of time since Databrary intends to store shared data indefinitely.
+Protocol amendments involve seeking approval for use of the Databrary template release language and modifying the time period over which collected data will be made available to clarify that Databrary intends to store shared data indefinitely.
 
 Databrary developed these ideas in close collaboration with the NYU and PSU research ethics and administration officials.
-We also took inspiration from the Human Connectome Project and Open Genome Project that face similar challenges.
-As of May 2015, Databrary had secured agreements with 55 institutions on behalf of 105 resesearchers representing a diverse array of entities in North and South America, Europe, and Australia.
+We also took inspiration from the Human Connectome Project and Open Genome Project that face similar challenges in sharing identifiable or potentially identifiable data.
+As of May 2015, Databrary had secured agreements with 62 institutions on behalf of 112 resesearchers representing a diverse array of entities in North and South America, Europe, and Australia.
 We believe that Databrary's model of sharing identifiable video data can be extended to other data types and other contexts.
@@
 It is too early to evaluate the impact of Databrary's data management tools on researcher's willingness to share data.
-But, the developmental science community is hungry for new practices that will enhance research productivity, as demonstrated by the rapid growth in the number of authorized Databrary investigators.
+But, the developmental and learning science community is hungry for new practices that will enhance research productivity, as demonstrated by the rapid growth in the number of authorized Databrary investigators.
 We are confident that consistent, but flexible data management practices will strengthen work within labs and among collaborators, enable convenient and reliable file uploading, and enhance the value of shared data in Databrary.
@@
 
-Imagine a time in the near future when researchers interested in studying classroom teaching and learning can mine an integrated, synchronized, interoperable, and open data set.
-That data set includes multiple measurements -- video from multiple cameras, eye tracking, motion, and physiological measurements, and both historical and real-time student performance data.
+Imagine a time in the near future when researchers interested in studying classroom teaching and learning can mine an integrated, synchronized, interoperable, and open dataset.
+That data include multiple measurements -- video from multiple cameras, eye tracking, motion, and physiological measurements, and both historical and real-time student performance data.
 Imagine that this classroom-level data can be linked with grade, school, neighborhood, community, region, and state-level data about education practice, curriculum, and policy.
@@
 In fact, we argue that video can be the core around which other measures of teaching and learning cluster.
-This requires reducing barriers to sharing video and related data and one that fosters new community values around the importance of data sharing.
+This requires reducing barriers to sharing video and related data and fostering new community values around the  data sharing that make it indispensible.
 The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread sharing within the developmental sciences community. 

### 2015-05-17 2015-06-01-nsf-ed-workshop/paper_new.tex
new version with kea, ls

\documentclass[letterpaper,man,apacite]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
% \usepackage{hyperref}
% \usepackage{breakurl}

\title{Open Video Data Sharing Can Transform Education Research}
\shorttitle{Video Data Sharing}

\twoauthors{Rick O. Gilmore}{Karen E. Adolph, David S. Millman}
\twoaffiliations{The Pennsylvania State University, The Databrary Project}{New York University, The Databrary Project}

\abstract{
Video captures the complexity, richness, and diversity of behavior unlike any other measure. 
As a result, large numbers of people who study teaching and learning employ video. 
Video documents itself to a large degree.
This presents significant potential for reuse by others.
The potential remains largely unrealized because videos are rarely shared.
Video contains information about personal identities.
This poses challenges to sharing. 
The large size of video files, diversity of formats, and incompatible software tools pose technical challenges. 
We describe how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community.
Databrary has developed solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and associated metadata.
The Databrary experience suggests ways that video and other identifiable data collected in the context of education research might be shared.
We envision a data intensive science of teaching and learning, with video as its core, that allows educational experiences to be tailored to students in ways that big data promises to personalize medicine.
The creation and support of repositories that enable the open sharing of dense, richly informative, high value, and high impact data about teaching and learning will help realize this ambitious vision.
}

\authornote{Rick O. Gilmore is in the Department of Psychology, The Pennsylvania State University, University Park, PA 16802, rogilmore@psu.edu. Karen E. Adolph is in the Department of Psychology at New York University. David S. Millman is with the NYU Library. Databrary is based on work supported by the National Science Foundation under Grant No. BCS-1238599, the Eunice Kennedy Shriver National Institute of Child Health and Human Development under Cooperative Agreement U01-HD-076595, and the Society for Research in Child Development.
Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author(s) and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.}

\begin{document}
\maketitle

\section{Introduction}
Open data sharing can help to translate insights from scientific research into applications serving essential human needs. 
Open data sharing bolsters transparency and peer oversight, encourages diversity of analysis and opinion, accelerates the education of new researchers, and stimulates the exploration of new topics not envisioned by the original investigators. 
Data sharing and reuse increases the impact of public investments in research and leads to more effective public policy. 
Although many researchers in the developmental, learning, and education sciences collect video as raw research data, most research on human learning and development remains shrouded in a culture of isolation \cite{Adolph2012}. 
Researchers share interpretations of distilled, not raw data, almost exclusively through publications and presentations. 
The path from raw video to research findings to conclusions cannot be traced or validated by others.
Other researchers cannot pose new questions that build on the same raw materials. 
This paper describes how the Databrary data library has overcome the most significant barriers to sharing video within the developmental sciences community.
It highlights how open video data sharing might improve scientific practice and advance research on learning and development.

\section{The Promise and Challenge of Video}
Video is a uniquely rich, inexpensive, and adaptable medium for capturing the complex dynamics of behavior. 
Researchers use video in home and laboratory contexts to study how infants, children, and adults behave in natural or experimenter-imposed tasks \cite{Karasik2014}. 
Researchers record videos of students in classrooms \cite{Alibali2012} to understand what teachers do and how students respond. 
Because video closely mimics the multisensory experiences of live human observers, recordings collected by one person for a particular purpose may be readily understood by another person and reused for a different purpose. 
Moreover, the success of YouTube and other video-based social media demonstrates that web-based video storage and streaming systems are now sufficiently well developed to satisfy large-scale demand. 
The question for researchers and policymakers is how to capitalize on video's potential to improve teaching and learning.

The answer requires overcoming significant technical, ethical, practical, and cultural challenges to sharing research video. 
\emph{File sizes and diverse formats present special challenges} for sharing. 
Video files are large (one hour of HD video can consume 10+ GB of storage) and come in varied formats (from cell phones to high-speed video). 
Many studies require multiple camera views to capture desired behaviors. 
Research video creates a data explosion: A typical lab studying infant or child development collects 8-12 hours of video/week \cite{Survey2012}. 
Thus, sharing videos requires substantial storage capacity and significant computational resources for transcoding videos into common, preservable formats.

\emph{Technical challenges involved in searching the contents of videos} present barriers to sharing. 
Videos contain rich and diverse information that requires significant effort by human observers to extract. 
Researchers make use of videos by watching them and, using paper and pencil or more automated computerized coding software, translating observations into ideas and numbers. 
In many cases, researchers assign codes to particular portions of videos.
These codes make the contents of videos searchable by others, in principle. 
However, researchers focus on different questions from varied theoretical perspectives and lack consensus on conceptual ontologies.
So, in practice, most coded data are not easily shared. 
Although human-centered video coding capitalizes on the unique abilities of trained observers to capture important dimensions of behavior, machine learning and computer vision tools may provide new avenues for tagging the contents of videos for educational and developmental research \cite{Amso2014,ChenYu2013,Fathi2012,Google2014,Raudies2014}.

Open video sharing must overcome \emph{ethical challenges} linked to sharing personally identifiable data. 
Although policies exist for sharing de-identified data, video contains easily identifiable data: faces, voices, names, interiors of homes and classrooms, and so on. 
Removing identifiable information from video severely diminishes its reuse value and poses additional burdens on researchers. 
So, open video sharing requires new policies that protect the privacy of research participants while preserving the integrity of raw video for reuse by others.

Open video sharing faces practical \emph{challenges of data management}. 
Developmental and education research is inundated by an explosion of data, most of which is inaccessible to other researchers.
Researchers lack time to find, label, clean, organize, and copy their files into formats that can be used and understood by others \cite{Ascoli2006}. 
Study designs vary widely, and no two labs manage data in the same way. 
Idiosyncratic terms, record-keeping, and data management practices are the norm. 
Few researchers document workflows or data provenance. 
Although video requires minimal metadata to be useful, video files must be electronically linked to what relevant metadata exist including information whether participants have given permission to share.

Perhaps the most important \emph{challenge is cultural}--community practices must change. 
Most researchers in the education, learning, and developmental sciences do not reuse their own videos or videos collected by other researchers; they neither recognize nor endorse the value of open sharing. 
Contributing data is anathema and justifications against sharing are many. 
Researchers cite intellectual property and privacy issues, the lack of data sharing requirements from funding agencies, and fears about the misuse, misinterpretation, or professional harm that might come from sharing \cite{Ascoli2006b,Ferguson2014}. 
Data sharing diverts energy and resources from scholarly activities that are more heavily and frequently rewarded. 
These barriers must be overcome to make data sharing a scientific norm.

\section{Databrary.org}
The Databrary project has built a digital data library (http://databrary.org) specialized for open sharing of research videos. 
Databrary has overcome the most significant barriers to sharing video, including solutions to maintaining participant privacy, storing, streaming, and sharing video, and for managing video datasets and associated metadata. 
Databrary's technology and policies lay the groundwork for securely sharing research videos on teaching and learning. 
In only a year of operation, Databrary has collected more than 7,000 individual videos, representing 2,400 hours of recording, featuring more than 1,800 infant, child, and adult participants. 
Databrary has more than 100 authorized researchers representing more than 60 institutions across the globe. 
Video data is big data, and the interest in recording and sharing video for research, education, and policy purposes continues to grow.

The Databrary project (databrary.org) arose to meet the challenges of sharing research video and to deliver on the promise of open data sharing in educational and developmental science. 
With funding from NSF (BCS-1238599) and NIH (NICHD U01-HD-076595), Databrary has focused on building a data library specialized for video, creating data management tools, crafting new policies that enable video sharing, and fostering a community of researchers who embrace video sharing. 
Databrary also developed a free, open-source video annotation tool, Datavyu (http://datavyu.org). 
The project received funding in 2012-2013, began a private beta testing phase in the spring of 2014 and opened for public use in October 2014.

\subsection{System Design}
The Databrary system enables large numbers of video and related files to be uploaded, converted, organized, stored, streamed, and tagged. 
Databrary is a free, open-source (http://github.com/databrary) web application whose data are preserved indefinitely in a secure storage facility at NYU. 
Databrary can house video and audio files, along with associated materials, coding spreadsheets, and metadata. 
Video and audio data are transcoded into standard and HTML5-compatible formats. 
This ensures that video data can be streamed and downloaded by any operating system that supports a modern browser. 
Copies of original video files are also stored. 
Databrary stores other data in their original formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).

The system’s data model embodies flexibility. 
Researchers organize their materials by acquisition date and time into structures called \emph{sessions}. 
A session corresponds to a unique recording episode featuring specific participants.
It contains one or more videos and other file types and may be linked to user-defined metadata about the participants, tasks or measures, and locations.  
A group of sessions is called a \emph{volume}.
Databrary contributors may combine sessions or segments with coding manuals, coding spreadsheets, statistical analyses, questionnaires, IRB documents, computer code, sample displays, and links to published journal articles.

Databrary does not enforce strict ontologies for tagging volumes, sessions, or the contents of videos. 
Video data are so rich and complex that in many domains, researchers have not settled on standard definitions for particular behaviors and may have little current need for standardized tasks, procedures, or terminology. 
Indeed, standardized ontologies are not necessary for many use cases. 
Databrary empowers users to add keyword tags and to select terms that have been suggested by others without being confined to the suggestions. 
Moreover, Databrary encourages user communities within Databrary to converge on common conceptual and metadata ontologies based on the most common keyword tags, and to construct and enforce common procedures and tasks wherever this makes sense.

Future challenges include enhancing the capacity to search for tagged segments inside of videos. 
Some search functionality exists in the current software, with more extensive capabilities on the near horizon. 
A related challenge involves importing files from desktop video coding tools.
This will allow for the visualization of user-supplied codes independent of the desktop software deployed in a particular project. 
We envision a parallel set of export functions that permit full interoperability among coding tools.
The priority will be to create interoperability with tools using open, not proprietary file formats. 
Databrary also recognizes the need to develop open standards and interfaces that enable Databrary to link to and synchronize with outside sources that specialize in other data types.

\subsection{Policies for Safe and Secure Video Sharing}
Policies for openly sharing identifiable data in ways that securely preserve participant privacy are essential for sharing research video. 
Databrary does not attempt to de-identify videos. 
Instead, we maximize the potential for video reuse by keeping recordings in their original unaltered form. 
To make unaltered raw videos available to others for reuse, Databrary has developed a two-pronged access model that (a) restricts access to authorized researchers, and (b) enables access to identifiable data only with the explicit permission of participants.

To gain access to Databrary a person must register on the site.
Applicants agree to uphold Databrary's ethical principles and to follow accepted practices concerning the responsible use of sensitive data.
Each applicant's institution must co-sign an access agreement. 
Full privileges are granted only to those applicants with independent researcher status at their institutions. 
Others may be granted privileges if they are affiliated with a researcher who agrees to sponsor their application and supervise their use. 
Ethics board or IRB approval is not required to gain access to Databrary because many use cases do not involve research, but IRB approval is required for research uses. 
Once authorized, a user has full access to the site’s shared data, and may browse, tag, download for later viewing, and conduct non- or pre-research activities.

Unique among data repositories, the Databrary access agreement authorizes both data use and contribution. 
However, users agree to store on Databrary only materials for which they have ethics board or IRB approval. 
Data may be stored on Databrary for the contributing researcher’s use regardless of whether the records are shared with others or not. 
When a researcher chooses to share, Databrary makes the data openly available to the community of authorized researchers.

In addition to restricting access to authorized researchers, Databrary has extended the principle of informed consent to participate in research to encompass permission to share data with other researchers.
To formalize the process of acquiring permission, Databrary has developed a Participant Release Template \cite{Databrary2015} with standard language we recommended for use with study participants. 
This language helps participants to understand what is involved in sharing video data, with whom the data will be shared, and the potential risks of releasing video and other identifiable data to other researchers.

\subsection{Managing Data for Sharing}
When researchers \emph{do} share, standard practice involves organizing data after a project has finished, perhaps when a paper goes to press. 
This “preparing for sharing” after the fact presents a difficult and unrewarding chore for investigators. 
It makes curating and ingesting datasets challenging for repositories, as well.
Databrary has chosen a different route to curation.

We have developed a data management system that empowers researchers to upload and organize data as it is collected.
Immediate uploading reduces the workload on investigators, minimizes the risk of data loss and corruption, and accelerates the speed with which materials become openly available. 
The system employs familiar, easy-to-use spreadsheet and timeline-based interfaces that allow users to upload videos, add metadata about tasks, settings, and participants, link related files, and assign appropriate permission levels for sharing. 
To encourage immediate uploading, Databrary provides a complete set of controls so that researchers can restrict access to their own labs or to other users of their choosing. 
Datasets can be openly shared with the broader research community at a later point when data collection and ancillary materials are complete, whenever the contributor is comfortable sharing, or when journals or funders require it.

\subsection{Building a Community}
Data sharing works only when the scientific community embraces it. 
From the beginning, Databrary has sought to cultivate a community of researchers who support data sharing and commit to enacting that support in their own work flows. 
Our community building efforts involve many interacting components. 
They include active engagement with professional associations, conference-based exhibits and training workshops, communications with research ethics and administration staff, talks and presentations to diverse audiences, and one-on-one consultations with individual researchers and research teams. 
These activities are time and labor-intensive, but we believe that they are critical to changing community attitudes toward data sharing in the educational and learning sciences. 
Looking ahead, it will be critical to engage funders, journals, and professional organizations in the effort to forge community consensus about the importance, feasibility, and potential of open video data sharing.

\section{Conclusion}
Imagine a time in the near future when researchers interested in studying classroom teaching and learning can mine an integrated, synchronized, interoperable, open and widely shared dataset. 
The components include video from multiple cameras, eye tracking, motion, and physiological measurements, and information from both historical and real-time student performance measures. 
Imagine that this classroom-level data can be linked with grade, school, neighborhood, community, region, and state-level data about education practice, curriculum, and policy. 
Then, imagine training a cadre of experts with skills in the data science of learning and education who are sensitive to privacy, confidentiality and ethical issues involved in research involving identifiable information. 
We empower these learning scientists to extract from the data meaningful insights about how educational practice and policy might be improved. 
In short, imagine a science of teaching and learning that can be personally tailored to individuals in ways analogous to the impact of big data on medicine. 
The barriers to realizing this vision are similar to those that confront the vision of personalized medicine – the development of technologies that enable data to be collected, synchronized, tagged, curated, stored, shared, linked, and aggregated; policies and practices that ensure security and individual privacy; and the cultivation of professional expertise needed to turn raw data into actionable insights.

As Gesell once noted, cameras can record behavior in ways that make it “...as tangible as tissue” \cite{Scott2011}.
The Databrary team contends that video has a central role to play in efforts to make tangible the anatomy of successful teaching and learning. 
In fact, we argue that video can be the core around which other measures of teaching and learning cluster. 
This requires reducing barriers to sharing video and fostering new community values around data sharing that make it indispensible. 
The Databrary project has built technology and policies that overcome many of the most significant barriers to widespread sharing within the developmental sciences community. 
Databrary suggests ways that video and other identifiable data collected in the context of education research might also be shared. Technologies and policies for providing secure access to videos for broader use cases will have to be developed, tools that allow desktop coding software files to be seamlessly converted to and from one another will have to be perfected, and ways of synchronizing and linking disparate data streams will have to be created. 
Equally important, communities of scholars dedicated to collecting, sharing, and mining education-related video data will have to be cultivated. 
But, we believe that the widespread sharing of high value, high impact data of the sort that video can provide promises to achieve this ambitious vision to advance education policy and improve practice. 
Databrary is working toward a future where open video data sharing is the norm, a personalized science of teaching and learning is the goal, and what optimizes student learning is as tangible as tissue.

\bibliography{paper}

\end{document}

### 2015-06-18 jcdl-15/poster/poster_landscape.tex
add poster for jcdl, modified from vss

--- 2015-05-07 vss-15/poster_landscape.tex
@@
 %\definecolor{lightblue}{cmyk}{0.83,0.24,0,0.12}
-\definecolor{lightblue}{rgb}{0.145,0.6666,1}
+%\definecolor{lightblue}{rgb}{0.145,0.6666,1}
+\definecolor{dbrary}{RGB}{134, 190, 179}
 
@@
   bgColorTwo=white,
-  borderColor=lightblue,
-  headerColorOne=black,
-  headerColorTwo=lightblue,
+  borderColor=dbrary,
+  headerColorOne=dbrary,
+  headerColorTwo=dbrary,
   headerFontColor=white,
   boxColorOne=white,
-  boxColorTwo=lightblue,
+  boxColorTwo=dbrary,
   % Format of textbox
@@
   %Title
-  {\bf{Sharing displays and data from vision science research with Databrary} }
+  {\bf{Databrary: Enabling Sharing and Reuse of Research Video} }
   % Authors
-  {Rick O. Gilmore \emph{(rick@databrary.org)}, Karen E. Adolph, David S. Millman, Lisa Steiger, and Dylan A. Simon \\
-  VSS 2015 -- Poster 26.4031}
+  { Andrew S. Gordon \emph{(drew@databrary.org)}, Rick O. Gilmore, Karen E. Adolph, David S. Millman, Lisa Steiger, and Dylan A. Simon \\
+  JCDL 2015}
   % University logo
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Problem}{name=problem,column=0,row=0}{
+  \headerbox{Abstract}{name=abstract,column=0,row=0}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      The open sharing of data and workflows from scientific research has gained increasing attention in many fields and has become a priority for research funders. 
-      Databrary (databrary.org) is a digital data library specialized for storing and sharing video and research data and metadata.
-Vision scientists often present temporally varying visual displays to human or non-human animal observers and may collect gaze position, kinematic data, or other physiological recordings simultaneously.
-      Some vision scientists collect identifiable video recordings of participants for use in off-line tagging or coding. 
-      Databrary may be especially well suited for these use cases because of the library's focus on uploading, organizing, storing, and sharing research data and metadata collected in the course of behavioral research.
-      Databrary has developed policies that enable identifiable information to be shared with the permission of research participants.
+      Video and audio recordings serve as a primary data source in many fields, especially in the social and behavioral sciences.
+      Recordings present unique opportunities for reuse and reanalysis for novel scientific purposes, but also present challenges related to respecting the privacy of individuals depicted.
+      Databrary is a web-based service for sharing and reusing the video data created by researchers in the developmental and learning sciences.
+      By investigating how researchers organize, analyze, and mine their own recordings, we have implemented a system that empowers researchers to capture, store, and share recordings in a standardized way.
+      The Databrary service promotes storage, management, sharing, and reuse of research data, controls access privileges to restricted human subject data, and facilitates browsing and discoverability of datasets.
  }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Privacy \& Security}{name=privacy,column=0,row=1, below=problem}{
+  \headerbox{Access Restrictions}{name=access,column=0,row=1, below=abstract}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      Videos may contain identifiable information such as faces, voices, and images of places. 
-      Databrary ensures that identifiable information can be shared in two ways:
-      \begin{description}
-      \item[Restricting Access:]
-      Only those with independent researcher status at an institution may be granted authorized investigator status.
-      Applicants promise to uphold ethical principles and sign an access agreement, which must be co-signed by their institution.
-
-      \item[Securing Permission to Share:]
-      Identifiable data may only be shared if participants have given permission.
-      Databrary has developed a release template for obtaining informed consent for sharing identifiable data.
-      The Databrary release template may be adapted for specific research protocols.
-
-      \end{description}
+      Databrary ensures that identifiable information in videos can shared under four different release levels:
+        \begin{center}
+        \includegraphics[width=\textwidth]{img/release_levels.png}
+        \end{center}
   }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% N.B.: this section is currently obscured by splash
-  \headerbox{What Can Be Shared}{name=shared,column=1,row=0}{
+  \headerbox{What Can Be Shared}{name=shared,column=0,below=access}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     Video files are transcoded into standard, HTML5-compatible formats, currently H.264+AAC in MP4.
-Databrary stores other data in native formats (e.g., .doc, .docx, .xls, .xlsx, .txt, .csv, .pdf, .jpg, .png).
-  }
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Technology Stack}{name=stack,column=0,below=privacy}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    Databrary is built on PostreSQL using the Scala Play framework and AngularJS.
-    Data are preserved indefinitely in a secure data storage facility at NYU. 
-    There is no cost to use the system; a sustainability model is under development.
-    All code is GPL on GitHub.
   }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{Volume/Dataset Page}{name=volume,column=2,span=1,row=0,bottomaligned=splash}{
+  \headerbox{Technology Stack}{name=stack,column=2,span=1,row=0,bottomaligned=splash}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+    Databrary is built on PostreSQL using the Scala Play framework and AngularJS.
+    Data are preserved indefinitely in a secure data storage facility at NYU. 
+    There is no cost to use the system; a sustainability model is under development.
+    All code is GPL on GitHub.
     \begin{center}
-  \includegraphics[width=\textwidth]{img/volume-31-new.jpg}
+    \includegraphics[width=.9\textwidth]{img/architecture.png}
     \end{center}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{Timeline}{name=data,column=2,span=1,row=1,below=volume, above=bottom, aligned=data}{
+\headerbox{Timeline}{name=data,column=2,span=1,row=1,below=stack, above=bottom, aligned=data}{
   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-06-18 jcdl-15/poster/poster_landscape.tex
fix typo
alphabetical order for names
a word
Fix postresql typo in jcdl poster

@@
   % Authors
-  { Andrew S. Gordon \emph{(drew@databrary.org)}, Rick O. Gilmore, Karen E. Adolph, David S. Millman, Lisa Steiger, and Dylan A. Simon \\
+  { Andrew S. Gordon \emph{(drew@databrary.org)}, Karen E. Adolph, Rick O. Gilmore, David S. Millman, Lisa Steiger, and Dylan A. Simon \\
   JCDL 2015}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-      Databrary ensures that identifiable information in videos can shared under four different release levels:
+      Databrary ensures that data with identifiable information can be shared under four different release levels:
         \begin{center}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    Databrary is built on PostreSQL using the Scala Play framework and AngularJS.
+    Databrary is built on PostgreSQL using the Scala Play framework and AngularJS.
     Data are preserved indefinitely in a secure data storage facility at NYU. 

### 2015-07-07 wiley-big-data-devel/big-data.tex
new intro sections.

\documentclass[letterpaper,man,apacite]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}

\title{From big data to deep insight in developmental science}
\shorttitle{Big data in development}

\oneauthor{Rick O. Gilmore}
\twoaffiliations{The Pennsylvania State University, The Databrary Project}

\abstract{}

\authornote{Rick O. Gilmore is in the Department of Psychology, The Pennsylvania State University, University Park, PA 16802, rogilmore@psu.edu.
Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.}

\begin{document}
\maketitle

\section{Introduction}

A brash new kid has moved in down the street, and she has stolen some limelight from the old guard.
She's a data scientist, and she's teaching the old dogs of statistics some new tricks about how to detect, describe, and visualize patterns in complex sets of data.
Like successful newcomers in many fields, data scientists build on some well-established ideas and tools, but combine and apply them in new ways to new problems \cite{Press2013b}.
And, like many upstarts in recent memory, data scientists have ridden the wave created by the massive growth in the availability of information over the internet.
In this paper, I will focus on how the increased availability of and interest in analyzing large, complex data sets has influenced the study of human development.
I will begin by discussing the recent history of ``big data" and how, in principle, it has implications for developmental science.
Then, I will survey a set of case studies that illustrate how large, complex data sets have been and are being used currently.
To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future.

\subsection{``Big Data" is big}

When the government notices a cultural phenomenon, it is often well-established. 
In March 2012, the U.S. Obama Administration announced an initiative to spend \$200 M dollars on a ``big data'' research and development (R\&D) initiative \cite{Gianchandani2012, Obama2012} with the expressed intent of applying data science techniques to large collections of digital data in order ``to solve some of the Nation's most pressing challenges..." \cite{Obama2012}.
The press release that accompanied the announcement \cite{Obama2012} highlighted some of the big data initiatives already or soon-to-be underway at the federal level, and these include the U.S. National Science Foundation (NSF), U.S. National Institutes of Health (NIH), the Departments of Energy and Defense and the U.S. Geological Survey (USGS).
The European Union (EU) has also taken public positions \cite{EU2013} about the importance of data to fostering social and economic well-being and has made it a priority to grant more open public access to data sets collected or managed by government entities \cite{EU2015}.
These potential benefits for open access to public data extend to governments in the developing world \cite{Malik2015}, and to state and local governments in the U.S.
For example, Indiana has embarked on an effort to reduce infant mortality using big data \cite{Ravindranath2014}.
Notwithstanding the promises of big data to improving public well-being, governments recognize that privacy concerns loom large \cite{Obama2014}, a topic we will revisit in later sections.
For researchers, the promise of government interest in big data extends beyond potentially new or expanded funding sources, such as the NIH's Big Data to Knowledge Initiative \cite{BD2K2015} or the now closed National Children's Study \cite{NCS2015}.
It means potentially new or more widely available or readily accessible data sources accompanied by more stringent requirements for open data sharing \cite{NIMH2015, NSF2011}.
We will revisit these topics again, as well.

So, big data looms large in the minds of public agencies who fund most research on human development.
In recognizing this, we note that the recent focus on big data has a history longer than the 21st century \cite{Press2013a}. 
Indeed, developmental scientists have long recognized the importance of collecting and sharing data sets that are big along at least one dimension -- the number of individuals, number of time points sampled, or the number of measures collected.
In turn, researchers in developmental science have continued to pioneer advanced analytical techniques that unpack and explain developmental phenomena across multiple levels of analysis and across wide-ranging timescales \cite{CDS2014, QuantDev}.

 \subsection{What is big data?}

    % + Three V's [^1],[^2]
    %    + Volume
    %    + Velocity
    %    + Variety
    % + Add
    % 	+ variability
    % 	+ complexity
    % + Data types
    % 	+ text, numeric
    % 	+ image
    % 	+ audio, video streams
   	%  + Issues 
    % 	- Data that cannot be processed by traditional databases or systems
    % 	- Structured or unstructured data
   	%  	- creation, capture, storage, sharing, analysis, transfer, visualization, privacy
   	%  	- 2.5 exabytes (10^8 bytes)/day [^3]
   	%  	- descriptive statistics; inferential; non-linear models; time series.
   	%  - In developmental science, this may mean
    % 	+ Numbers of participants
    % 	+ Numbers of time points
    % 	+ Numbers of measures

\subsection{Big data about what?}

    % + Biology
    %     * Genes
    %     * Brain activity, structure
    %     * Growth and health
    % + Behavior
    %     * Perception
    %     * Action
    %     * Cognition
    %     * Language
    %     * Emotion / Personality / Temperament
    %     * Friendships, families, & relationships
    %     * Attitudes 
    % + Educational and Employment aptitude/outcomes
    % + Social/economic/cultural context

\subsection{Why big data in developmental science?}
    % + What develops?
    % + Why does it develop that way?
    %     * Changes in the environment?
    %     * Changes in the person?
    % + More precise estimates of small effects
    %     * Trends within person
    %     * across time...
    %     * Example recent replication of GWAS with 100K, <http://pss.sagepub.com/content/early/2014/10/06/0956797614545132.full
    % + Multi-level influences on development
    %     * Influences at different spatial, temporal scales
    %     * Nesting
    % + Improve outcomes for children
    %     * In general
    %     * Disease/disorder focus

\section{Case Studies}

\subsection{Biology and Health}

        % * Genes
        % 	- EGADs/NEAD (Jenae Neiderhiser)
        %   - Other twin/genetically informed studies 
        % * Brain activity/structure
        %     - Developmental Connectome Project, [website](http://www.developingconnectome.org/)
        %     - David Edwards, Oxford, Imperial College, Kings College, [Centre for the Developing Brain](http://www.kcl.ac.uk/medicine/research/divisions/imaging/centres/cdb/index.aspx)
        %     - John Richards' normative structural MRI series.
        % * Growth and health
        %     - NICHD Study of Early Child Care and Youth Development (SECCYD)
        %     - [National Children's Study](http://www.niehs.nih.gov/research/programs/children-study/), [^11]
        %         - In need of redesign? [^12], according to 2014 report from NAS [^13], 40 academic research centers have been dropped from the program [^14]
        %         - Similar to Japan Environment and Children's Study [^15]
        %         - WHO coordination efforts for Birth Cohort Studies [^16]
        %         - British birth cohort studies [^17], site specificially describing UK longitudinal research datasets [^18]
        %     - World Health Organization (WHO) Child Growth Standards, <http://www.who.int/childgrowth/en/>
        %     - WHO Multicentre Growth Reference Study (MGRS), <http://www.who.int/childgrowth/en/>
        %         - "The MGRS collected primary growth data and related information from approximately 8500 children from widely different ethnic backgrounds and cultural settings (Brazil, Ghana, India, Norway, Oman and the USA)."
        %     - Adolescent Health, <http://www.cpc.unc.edu/projects/addhealth>
        %     - Sex, drugs, alcohol and risky behavior
        %         + Rob Turrisi, Eva Lewkowicz, Chuck Geier
        %         + PSU Prevention Center
        %     - Nutrition

\subsection{Behavior}

   % + Mass media interest in longitudinal perspective
   %      * Up Series, <http://en.wikipedia.org/wiki/Up_Series>
   %      * Boyhood
   %       * Perception
   %          - Teller Acuity Card norms
   %              + <http://www.ncbi.nlm.nih.gov/pubmed/7890497>
   %              + <http://www.ncbi.nlm.nih.gov/pubmed/7890496>
   %          - Auditory testing norms
   %      * Action
   %          - Gesell?
   %          - Bayley Scales of Infant and Toddler Development
   %              + Nancy Bayley bio, <http://www.pearsonclinical.com/education/authors/bayley-nancy.html>
   %      * Cognition
   %          - Wechsler
   %              + <http://www.iupui.edu/~flip/wechsler.html>
   %              + Wechsler Adult Intelligence Scale (WAIS)
   %              + Wechsler Intelligence Scale for Children (WISC-III), 6-16 year-olds
   %              + Wechsler Preschool and Primary Scale of Intelligence (WPPSI-R), 4-6 1/2 years.
   %      * Language
   %          - Human Speechome, <http://www.media.mit.edu/cogmac/projects/hsp.html>
   %          - [CHILDES/TalkBank](http://talkbank.org/)
   %          - Smith, Aslin, Yu, Frank word learning, grammar
   %          - MacArthur, CDI and Mike Frank's project (wordbank.stanford.edu)
   %          - Cathie Tamis-Lemonda
   %          - Gedeon Deak
   %          - LENA, <http://www.lenafoundation.org/>
   %      * Emotion / Personality / Temperament
   %          - Ecological Momentary Assessment
   %              + Matthias Mehl
   %              + Diary studies
   %          - Infant Behavior Questionnaire, CBC

\subsection{Education and Employment}

        %    - Bureau of Labor Statistics (BLS), youth employment, <http://www.bls.gov/news.release/youth.nr0.htm>
        % * Education
        %     - MOOCs
        %         + Stanford, Coursera; Harvard, MIT, EdX.
        %     - SAT, ACT scores, <http://www.erikthered.com/tutor/sat-act-history.html>
        %     - Aptitude, <http://timss.bc.edu/>
        %     - OECD PISA, <http://pisa2000.acer.edu.au/index.php>
        %     - Measures of Effective Teaching (MET) Study, <http://www.metproject.org/>
        %         - "The MET project was a research partnership between 3,000 teacher volunteers and dozens of independent research teams. The project's goal was to build and test measures of effective teaching to find out how evaluation methods could best be used to tell teachers more about the skills that make them most effective and to help districts identify and develop great teaching. Launched in 2009, the study has identified multiple measures and tools that – taken together – can provide an accurate and reliable picture of teaching effectiveness."
        %         - "The project was funded by the Bill & Melinda Gates Foundation as part of ongoing efforts to give teachers the tools they need to be successful and to improve student achievement in public schools across the United States. 2. WHO WAS INVOLVED IN THE MET PROJECT?Truly understanding the skills and techniques that are hallmarks of great teaching requires collaboration between educators and researchers. The MET project was unprecedented: it was a partnership among thousands of teacher volunteers and administrators and union leaders from school districts across the country and dozens of independent researchers and education organizations. Teachers – more than 3,000 of them – were at the heart of the study. The teacher volunteers were recruited from Charlotte-Mecklenburg Schools, N.C.; Dallas Independent School District, Texas; Denver Public Schools, Colo.; Hillsborough County Public Schools, Fla.; Memphis City Schools, Tenn.; New York City Department of Education, N.Y.; and Pittsburgh Public Schools, Pa. Pittsburgh served as the project's pilot district, but no data from this district was analyzed. Lead researchers involved in the project were affiliated with Dartmouth College, Harvard University, the University of Michigan, the University of Virginia and Stanford University. Participating non-profits and education companies included RAND Corporation, Educational Testing Service, Teachscape, The Danielson Group, The New Teacher Center, National Math & Science Initiative and Westat."
          %      * Attitudes
   %          - Judith Torney-Purta, <http://www.nie.edu.sg/research-publications/cieclopediaorg/cieclopediaorg-a-to-z-listing/Judith-Torney-Purta>

\subsection{Social, Economic, and Cultural Contexts of Development}

       % * [Panel study on income dynamics](http://psidonline.isr.umich.edu/)
       %  * [NIH study of early child care and youth development (SECCYD)](https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx)
       %  * [Family Life Project](http://flp.fpg.unc.edu/)
       %  + Normative vs. perturbed development
       %      * Bucharest Early Intervention Project (BEIP), <http://www.childrenshospital.org/beip>
       %      * NPR story, <http://www.npr.org/blogs/health/2014/02/20/280237833/orphans-lonely-beginnings-reveal-how-parents-shape-a-childs-brain>
       %      * Ros Picard, ASD

\section{The Future of Big Data}

\subsection{Technical}

\subsubsection{Collecting data with New technologies, data streams}
        % * Social media
        %     * Twitter, Facebook, Snapchat, etc.
        % + Other web-based data collection tools
        %     * Lookit
        %     * Tablets
        %     * Portable devices

\subsubsection{Coding and analyzing big data}
        % * Audio
        % * Video
        % * Surveys
        % * Physiological data streams
        % * Geo-location

\subsection{Storing and sharing big data}
        % * Where to store
        % * What/when to share
        % * How to share
        % * NDAR
        % * Dataverse
        % * ICPSR
        % * Databases of datasets
        % * International Cross Time Cross System Dataset (XTXS), <http://www.intledstatsdatabase.org/>
        % * APA's list, <http://www.apa.org/research/responsible/data-links.aspx>
        %         * Domain-specific
        %     - NDAR, etc.
        %     - MacArthur Network
        % * Domain-general
        %     - UNC
        %     - NYU, Berkeley, Washington, big data science centers
        %     - PSU/Databrary
        %     - Michigan

\subsection{Conceptual and Theoretical}

\subsubsection{Ethics}
        % * Privacy
        %     - FERPA
        %     - HIPAA
        %     - Differing standards across countries
        %     - Commercial vs. academic research practices, cultures
        %     - "Linking" data improves potential inferences, but increases privacy risk.
        %     - Can minors consent to "indefinite" data storage?
        % * Security
        %     - Consent
        %         + Open Humans Project, Personal Genome Project
        %         + Databrary

\subsubsection{Challenges of conducting big data developmental research}
     %    * Cohort effects, development is developing
     %    * What is normative? What is not?
     %    * Instruments/measures public and open and free vs. private and closed and expensive
     % Most of psychology (especially developmental?) is WEIRD
    	% - How do large scale patterns arise from micro processes
    	% - The end of theory? [^5], [^6], Chris Anderson.
    	% - Questions for Big Data (boyd & Crawford), [^7] 
    	% - Social Sciences are underpowered, most research findings are false. Ioanidis. [^9]. Center at Stanford on research methodology. METRICS. [^10]

\section{Conclusion}

\bibliography{big-data}

\end{document}

### 2015-07-07 wiley-big-data-devel/big-data.tex
more abt devel sci & big data

@@
 
-So, big data looms large in the minds of public agencies who fund most research on human development.
+Clearly, big data looms large in the minds of public officials whose agenciesfund most research on human development.
 In recognizing this, we note that the recent focus on big data has a history longer than the 21st century \cite{Press2013a}.
@@
 In turn, researchers in developmental science have continued to pioneer advanced analytical techniques that unpack and explain developmental phenomena across multiple levels of analysis and across wide-ranging timescales \cite{CDS2014, QuantDev}.
+Developmental scientists have also been at the vanguard of efforts to create ways to share research data that preserve privacy and uphold ethical research principles \cite{Adolph2012, AERA2011, AERAVideo2015, Bakeman2011, Databrary2015}.
 

### 2015-07-08 wiley-big-data-devel/big-data.tex
add what is big data

--- 2015-07-07 wiley-big-data-devel/big-data.tex
@@
 
-A brash new kid has moved in down the street, and she has stolen some limelight from the old guard.
+A brash new kid has moved in down the street, and she has stolen the limelight from the old guard.
 She's a data scientist, and she's teaching the old dogs of statistics some new tricks about how to detect, describe, and visualize patterns in complex sets of data.
 Like successful newcomers in many fields, data scientists build on some well-established ideas and tools, but combine and apply them in new ways to new problems \cite{Press2013b}.
-And, like many upstarts in recent memory, data scientists have ridden the wave created by the massive growth in the availability of information over the internet.
+And, like many upstarts in recent memory, data scientists have ridden the wave created by the massive growth in the availability of information over the internet \cite{Borgmann2015}.
 In this paper, I will focus on how the increased availability of and interest in analyzing large, complex data sets has influenced the study of human development.
@@
 
-When the government notices a cultural phenomenon, it is often well-established. 
+``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.
+While much of the public attention focused on big data concerns the influence of social media \cite{meyer_everything_2014}, impacts on economic development \cite{HBR2015} or on data security breaches \cite{Vinton2015}, the topic has also garnered the attention of government officials \cite{NSFBigData}, and private foundations \cite{ArnoldFdnEvidence, GatesFdn, MooreFdn} with interests in funding scientific research.
 In March 2012, the U.S.
@@
 For example, Indiana has embarked on an effort to reduce infant mortality using big data \cite{Ravindranath2014}.
-Notwithstanding the promises of big data to improving public well-being, governments recognize that privacy concerns loom large \cite{Obama2014}, a topic we will revisit in later sections.
-For researchers, the promise of government interest in big data extends beyond potentially new or expanded funding sources, such as the NIH's Big Data to Knowledge Initiative \cite{BD2K2015} or the now closed National Children's Study \cite{NCS2015}.
-It means potentially new or more widely available or readily accessible data sources accompanied by more stringent requirements for open data sharing \cite{NIMH2015, NSF2011}.
-We will revisit these topics again, as well.
 
-So, big data looms large in the minds of public agencies who fund most research on human development.
-In recognizing this, we note that the recent focus on big data has a history longer than the 21st century \cite{Press2013a}. 
-Indeed, developmental scientists have long recognized the importance of collecting and sharing data sets that are big along at least one dimension -- the number of individuals, number of time points sampled, or the number of measures collected.
-In turn, researchers in developmental science have continued to pioneer advanced analytical techniques that unpack and explain developmental phenomena across multiple levels of analysis and across wide-ranging timescales \cite{CDS2014, QuantDev}.
+Notwithstanding the promises of big data to improving public well-being, governments recognize that privacy and security concerns loom large \cite{Obama2014, Vinton2015} and must be addressed to fully realize the promise of big data science, especially in fields focused on human well-being.
+For researchers, the promise of government interest in big data extends beyond potentially new or expanded funding sources \cite{NSFBigData}, such as the NIH's Big Data to Knowledge Initiative \cite{BD2K2015} or the now closed National Children's Study \cite{NCS2015}.
+It means potentially new or more widely available or readily accessible data sources in open respositories \cite{databrary.org, dataverse.org, ICSPR, macwhinney_childes_2001} accompanied by more stringent requirements for open data sharing from funders \cite{NIMH2015, NSF2011, GatesFdn} and scientific publication outlets \cite{nosek_promoting_2015}.
+At the same time, the collection, storage and open sharing of large data sets about pose significant data management challenges that may require changes in technology use \cite{ProjectJupyter, osf} and scientific practice, including closer collaboration between researchers and librarians and information specialists with expertise in data management, storage and long-term preservation \cite{gordon_researcher-library_2015, lynch_big_2008}.
+We will revisit these topics later.
+
+On the other hand, developmental scientists have long recognized the importance of collecting and sharing data sets that are big along at least one dimension -- the number of individuals, number of time points sampled, or the number of measures collected \cite{macwhinney_childes_2001}.
+In turn, researchers in developmental science continue to pioneer advanced analytical techniques that unpack and explain developmental phenomena across multiple levels of analysis and across wide-ranging timescales \cite{bakeman_sequential_2011, CDS2014, QuantDev}.
+Developmental scientists have also been at the vanguard of efforts to create ways to share research data that preserve privacy and uphold ethical research principles \cite{Adolph2012, AERA2011, AERAVideo2015, Asilomar2014, Databrary2015}.
+Accordingly, the developmental research community may be especially well-positioned to turn big data into deep insight.
 
@@
 
-    % + Three V's [^1],[^2]
-    %    + Volume
-    %    + Velocity
-    %    + Variety
-    % + Add
-    % 	+ variability
-    % 	+ complexity
-    % + Data types
-    % 	+ text, numeric
-    % 	+ image
-    % 	+ audio, video streams
-   	%  + Issues 
-    % 	- Data that cannot be processed by traditional databases or systems
-    % 	- Structured or unstructured data
-   	%  	- creation, capture, storage, sharing, analysis, transfer, visualization, privacy
-   	%  	- 2.5 exabytes (10^8 bytes)/day [^3]
-   	%  	- descriptive statistics; inferential; non-linear models; time series.
-   	%  - In developmental science, this may mean
-    % 	+ Numbers of participants
-    % 	+ Numbers of time points
-    % 	+ Numbers of measures
+Doug Laney \cite{laney01controlling3v} was the first to describe the dimensions of data streams that make them especially ``big'' and which accordingly merit special consideration \cite{ibm_2015}: volume, velocity, and variety.
+So, data streams may be large in quantity, change rapidly, or consist of many distinct components.
+Of course, these dimensions or general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
+What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year \cite{CERN_LHC}.
+Similarly, the measure of volume matters.
+The Inter-university Consortium for Political and Social Research (ICPSR) \cite{ICSPR}, one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections \cite{}
+Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project \cite{METProject}, the large number of files in the repository totalled about 10 TB of digital storage.
+The aspects of volume most pertinent to developmental science include the number of participants or families and the number of measurement time points.
+Similarly, the aspects of data velocity most relevant to developmental researchers are those related to the frequency of measurement.
+In developmental research, velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
+Variety encompasses both the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical.
+To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.
+
+Conventional conceptualizations of data focus on text or numeric information, but increasingly images, and audio \cite{Mehl} and video recordings have become part of the mix of data streams collected by researchers and others to better understand human behavior.
+Video in particular captures the complexity and richness of behavior unlike any other measure, and so video provides the main source of information for many researchers who study behavior in laboratory, classroom, or museum contexts.
+Images and recordings are both larger, in file size, and denser than text or flat-file data. 
+Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders \cite{Databrary2015, openfmri.org} and, of course, sharing these materials for non-research entertainment uses has become astonishingly popular \cite{YouTube2015}.
+New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states \cite{Picard}, activity patterns, and momentary cognitive \cite{Sliwinski}, and emotional states broadly available to researchers.
+Moreover, new web-based tools like Amazon's Mechanical Turk (http://www.mturk.com) or Apple's HealthKit (http://www.applehealthkit.com/) will make it possible for research teams to collect data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
+Fundamentally, big data are those that cannot readily be processed by traditional databases or systems that are widely familiar to developmental researchers.
+
+Data may be big in any number of ways and present themselves in structured or unstructured forms, but they pose significant new challenges to traditional avenues for data creation, capture, storage, sharing, analysis, transfer, visualization, and analysis.
+Big data may also magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
 

### 2015-07-28 wiley-big-data-devel/big-data.tex
nearly complete rough draft

@@
 
-A brash new kid has moved in down the street, and she has stolen the limelight from the old guard.
-She's a data scientist, and she's teaching the old dogs of statistics some new tricks about how to detect, describe, and visualize patterns in complex sets of data.
-Like successful newcomers in many fields, data scientists build on some well-established ideas and tools, but combine and apply them in new ways to new problems \cite{Press2013b}.
-And, like many upstarts in recent memory, data scientists have ridden the wave created by the massive growth in the availability of information over the internet \cite{Borgmann2015}.
-In this paper, I will focus on how the increased availability of and interest in analyzing large, complex data sets has influenced the study of human development.
-I will begin by discussing the recent history of ``big data" and how, in principle, it has implications for developmental science.
-Then, I will survey a set of case studies that illustrate how large, complex data sets have been and are being used currently.
-To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future.
+A search on the term ``big data'' yields more than 49 million hits on Google and 13 million hits on Bing.
+The results return in less than a second.
+The term yields more than 100,000 results on Google Scholar.
+A search of the term using Google's ngram viewer (http://books.google.com/ngram) shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
+Google's ngram viewer indexes terms in digitized books.
+So, whether measured by search engine matches of electronic documents available on the internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
+That use has grown substantially over the past several decades.
+The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases that has occurred largely in the last 20 years (https://en.wikipedia.org/wiki/List_of_websites_founded_before_1995).
+``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015; boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.
 
-\subsection{``Big Data" is big}
+In this paper, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
+I begin by asking what makes data especially big, and what implications the size, density, or complexity of datasets might have for understanding human development.
+Then, I will briefly survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
+To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future of the field.
 
-``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.
-While much of the public attention focused on big data concerns the influence of social media \cite{meyer_everything_2014}, impacts on economic development \cite{HBR2015} or on data security breaches \cite{Vinton2015}, the topic has also garnered the attention of government officials \cite{NSFBigData}, and private foundations \cite{ArnoldFdnEvidence, GatesFdn, MooreFdn} with interests in funding scientific research.
-In March 2012, the U.S.
-Obama Administration announced an initiative to spend \$200 M dollars on a ``big data'' research and development (R\&D) initiative \cite{Gianchandani2012, Obama2012} with the expressed intent of applying data science techniques to large collections of digital data in order ``to solve some of the Nation's most pressing challenges..." \cite{Obama2012}.
-The press release that accompanied the announcement \cite{Obama2012} highlighted some of the big data initiatives already or soon-to-be underway at the federal level, and these include the U.S.
-National Science Foundation (NSF), U.S.
-National Institutes of Health (NIH), the Departments of Energy and Defense and the U.S.
-Geological Survey (USGS).
-The European Union (EU) has also taken public positions \cite{EU2013} about the importance of data to fostering social and economic well-being and has made it a priority to grant more open public access to data sets collected or managed by government entities \cite{EU2015}.
-These potential benefits for open access to public data extend to governments in the developing world \cite{Malik2015}, and to state and local governments in the U.S.
-For example, Indiana has embarked on an effort to reduce infant mortality using big data \cite{Ravindranath2014}.
+I will argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
+Nor, do I find that ``big data'' analyses are all that new in developmental psychology.
+Rather, I find that a while the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily interrelatable datasets.
+Thus, whether significant novel insights into human development emerge from the era of ``big data'' depend not just on the size or complexity of the datasets, but on how widely and openly the data are shared, and how readily able researchers are able to combine or link datasets across levels of analysis.
+In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual esearcher behavior, publisher priorities, and research funding levels.
+Technology may have enabled the ``big data'' era, but the challenges it poses may turn out to be the least important, at least where advancing research in developmental psychology is concerned.
 
-Notwithstanding the promises of big data to improving public well-being, governments recognize that privacy and security concerns loom large \cite{Obama2014, Vinton2015} and must be addressed to fully realize the promise of big data science, especially in fields focused on human well-being.
-For researchers, the promise of government interest in big data extends beyond potentially new or expanded funding sources \cite{NSFBigData}, such as the NIH's Big Data to Knowledge Initiative \cite{BD2K2015} or the now closed National Children's Study \cite{NCS2015}.
-It means potentially new or more widely available or readily accessible data sources in open respositories \cite{databrary.org, dataverse.org, ICSPR, macwhinney_childes_2001} accompanied by more stringent requirements for open data sharing from funders \cite{NIMH2015, NSF2011, GatesFdn} and scientific publication outlets \cite{nosek_promoting_2015}.
-At the same time, the collection, storage and open sharing of large data sets about pose significant data management challenges that may require changes in technology use \cite{ProjectJupyter, osf} and scientific practice, including closer collaboration between researchers and librarians and information specialists with expertise in data management, storage and long-term preservation \cite{gordon_researcher-library_2015, lynch_big_2008}.
-We will revisit these topics later.
+ \subsection{What Does `Big Data` Mean in Developmental Science?}
 
-On the other hand, developmental scientists have long recognized the importance of collecting and sharing data sets that are big along at least one dimension -- the number of individuals, number of time points sampled, or the number of measures collected \cite{macwhinney_childes_2001}.
-In turn, researchers in developmental science continue to pioneer advanced analytical techniques that unpack and explain developmental phenomena across multiple levels of analysis and across wide-ranging timescales \cite{bakeman_sequential_2011, CDS2014, QuantDev}.
-Developmental scientists have also been at the vanguard of efforts to create ways to share research data that preserve privacy and uphold ethical research principles \cite{Adolph2012, AERA2011, AERAVideo2015, Asilomar2014, Databrary2015}.
-Accordingly, the developmental research community may be especially well-positioned to turn big data into deep insight.
+Doug Laney \cite{laney01controlling3v} was the first person to describe the dimensions of data streams that make them especially ``big'': Volume, velocity, and variety.
+Data streams that are large in quantity, change rapidly, or consist of many distinct components merit special consideration \cite{ibm_2015}.
+Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
 
- \subsection{What is big data?}
-
-Doug Laney \cite{laney01controlling3v} was the first to describe the dimensions of data streams that make them especially ``big'' and which accordingly merit special consideration \cite{ibm_2015}: volume, velocity, and variety.
-So, data streams may be large in quantity, change rapidly, or consist of many distinct components.
-Of course, these dimensions or general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
 What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year \cite{CERN_LHC}.
 Similarly, the measure of volume matters.
-The Inter-university Consortium for Political and Social Research (ICPSR) \cite{ICSPR}, one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections \cite{}
+The Inter-university Consortium for Political and Social Research (ICPSR) \cite{ICSPR}, one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections \cite{ICPSR}.
 Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project \cite{METProject}, the large number of files in the repository totalled about 10 TB of digital storage.
-The aspects of volume most pertinent to developmental science include the number of participants or families and the number of measurement time points.
-Similarly, the aspects of data velocity most relevant to developmental researchers are those related to the frequency of measurement.
+Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?
+
+Most developmental researchers study individuals or groups of individuals and families.
+But, others focus on neighborhoods or schools, or even broader social, cultural, political and environmental influences.
+This diversity in the targets of analysis reflects an implicit shared belief that developmental trajectories arise from a variety of influences organized at multiple levels of analysis and operating across multiple time scales.
+Consequently, the aspects of `volume' most pertinent to developmental science include the number of participants or families or the number measurement time points.
+We will see that many developmental studies involve tens of participants, but some notable datasets have hundreds or thousands of participants.
+Similarly, the aspects of data velocity most relevant to developmental researchers are those related to the frequency or spacing of measurements.
 In developmental research, velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
@@
 To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.
+For example, patterns of brain activity change rapidly, while working memory spans typically do not.
+Brain activity and peripheral autonomic nervous system measurements like heart rate, heart period variability, and galvanic skin responses are related to central nervous system activity in causal, but indirect and complex ways.
 
@@
 
-\subsection{Big data about what?}
+\subsection{Big data about what and for what purposes?}
 
-    % + Biology
-    %     * Genes
-    %     * Brain activity, structure
-    %     * Growth and health
-    % + Behavior
-    %     * Perception
-    %     * Action
-    %     * Cognition
-    %     * Language
-    %     * Emotion / Personality / Temperament
-    %     * Friendships, families, & relationships
-    %     * Attitudes 
-    % + Educational and Employment aptitude/outcomes
-    % + Social/economic/cultural context
+Armed with some definitions about what constitutes big data, we can next ask what topics large datasets can describe and help inform.
+The answer spans the range of phenomena covered by developmental science, from genes to geography and from microseconds to millenia.
+In the case studies to follow, we will highlight studies that focus on biological evidence from genes, brain structure and activity patterns, hormonal variation, nutrition, growth, and health.
+Biological evidence may derive from typically developing or atypically developing populations or both.
+% Add citations to examples %
+Often, behavioral evidence accompanies biological evidence, sometimes featuring standardized task assessments.
+But, there are many examples of big datasets consisting solely of behavioral data.
+These provide evidence about patterns of change and stability across the range of phenomena that interest developmental researchers: perception, action, cognition, language, emotion, sleep, personality, temperament, friendship, family, play, school achievement, substance use, sexuality, and sleep.
+Still other large datasets focus on educational and employment aptitudes and outcomes and the social, economic, political, cultural, and historical contexts in which children and youth come of age.
+% Again find examples and cite %
 
-\subsection{Why big data in developmental science?}
-    % + What develops?
-    % + Why does it develop that way?
-    %     * Changes in the environment?
-    %     * Changes in the person?
-    % + More precise estimates of small effects
-    %     * Trends within person
-    %     * across time...
-    %     * Example recent replication of GWAS with 100K, <http://pss.sagepub.com/content/early/2014/10/06/0956797614545132.full
-    % + Multi-level influences on development
-    %     * Influences at different spatial, temporal scales
-    %     * Nesting
-    % + Improve outcomes for children
-    %     * In general
-    %     * Disease/disorder focus
+Large datasets can demonstrate either human diversity or human universals, depending on the topic and area of focus.
+In most cases, however, big datasets in developmental science serve to answer core scientific questions that drive research in this area: What develops? 
+Why does it develop that way? 
+Does the environment change or the person or both, and how so?
+Big datasets enable more precise estimation of small effects, for qualities or conditions rare among individuals.
+High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
+Developmental scientists have long recognized the multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998, gottlieb_normally_1998, oyama_ontogeny_2000, vygotsky_mind_1980}.
+Only recently have the tools and resources needed to collect high volume, velocity, and variety needed to make many of these theoretical notions empirically testable.
+Finally, many, perhaps most big datasets in developmental science have as a primary aim the understanding or treatment of disease or disorder.
+In short, big data in developmental science focuses on answering the biggest, most important questions in the field.
+In the next section, we turn to evaluating specific cases of research projects that have generated big datasets and data repositories that preserve, store, and share them with the scientific community.
 
@@
 
-        % * Genes
-        % 	- EGADs/NEAD (Jenae Neiderhiser)
-        %   - Other twin/genetically informed studies 
-        % * Brain activity/structure
-        %     - Developmental Connectome Project, [website](http://www.developingconnectome.org/)
-        %     - David Edwards, Oxford, Imperial College, Kings College, [Centre for the Developing Brain](http://www.kcl.ac.uk/medicine/research/divisions/imaging/centres/cdb/index.aspx)
-        %     - John Richards' normative structural MRI series.
-        % * Growth and health
-        %     - NICHD Study of Early Child Care and Youth Development (SECCYD)
-        %     - [National Children's Study](http://www.niehs.nih.gov/research/programs/children-study/), [^11]
-        %         - In need of redesign? [^12], according to 2014 report from NAS [^13], 40 academic research centers have been dropped from the program [^14]
-        %         - Similar to Japan Environment and Children's Study [^15]
-        %         - WHO coordination efforts for Birth Cohort Studies [^16]
-        %         - British birth cohort studies [^17], site specificially describing UK longitudinal research datasets [^18]
-        %     - World Health Organization (WHO) Child Growth Standards, <http://www.who.int/childgrowth/en/>
-        %     - WHO Multicentre Growth Reference Study (MGRS), <http://www.who.int/childgrowth/en/>
-        %         - "The MGRS collected primary growth data and related information from approximately 8500 children from widely different ethnic backgrounds and cultural settings (Brazil, Ghana, India, Norway, Oman and the USA)."
-        %     - Adolescent Health, <http://www.cpc.unc.edu/projects/addhealth>
-        %     - Sex, drugs, alcohol and risky behavior
-        %         + Rob Turrisi, Eva Lewkowicz, Chuck Geier
-        %         + PSU Prevention Center
-        %     - Nutrition
+\subsubsection{Epidemiological and Population-based Studies}
+
+Some of the largest and most widely used datasets in developmental science are population-based, often with an epidemiological dimension.
+Several notable examples began data collection in the 1950's or 60's.
+The National Health and Nutrition Examination Survey (NHANES) \cite{NHANES} began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S.
+infants, children, and adults.
+The annual survey combines interviews about demographic, socioeconomic, dietary and health related topics and physical examinations involving medical, dental, and physiological measurements, as well as laboratory tests.
+The study assesses a nationally representative sample of about 5,000 persons located across the country, but 15 counties are surveyed annually.
+Many will be familiar with some of the study's products such as standardized child growth charts (http://www.cdc.gov/growthcharts/cdc_charts.htm), or related initiatives such as the National Youth Fitness Survey (NNYFS) (http://www.cdc.gov/nchs/nnyfs.htm which collected data on physical activity and fitness levels in order to provide an evaluation of the health and fitness of children in the U.S.
+ages 3 to 15.
+The NNYFS collected data on physical activity and fitness levels of our youth through interviews and fitness tests.
+Much of the NHANES and NNYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
+However, access to some data are restricted, available only to individuals via a Research Data Center (RDC) that can provide access.
+<http://wwwn.cdc.gov/nchs/nhanes/search/variablelist.aspx?CycleBeginYear=1999&Component=Non-Public>.
+There are 19 Federal Statistical Research Data Center (RDC) locations across the U.S.
+A Google Scholar search shows more than 110,000 publications mention the NHANES dataset.
+The Panel Study on Income Dynamics (PSID) \cite{PSID} is another large-scale population based effort.
+PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
+The study consists of survey responses to questions about employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, education, and other topics.
+As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}.
+Much of data is available publicly to users who register (http://simba.isr.umich.edu/U/ca.aspx) subject to the provisions of a formal set of use conditions (http://simba.isr.umich.edu/U/CondUse.aspx) that focus largely on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
+In addition, some restricted data elements are only available subject to a formal restricted use contract (http://simba.isr.umich.edu/restricted/RestrictedUse.aspx) between the investigator and the data owner, the University of Michigan.
+
+In contrast the National Longitudinal Survey of Youth (NLSY) \cite{NLSY} has studied a large subset of the same cohort since the study began in 1979.
+The project follows the lives of a sample of almost 10,000 American youth born between 1957-64.
+Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY ata are now available for up to 25 survey rounds on questions relating to employment and related questions.
+The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52 through the National Longitudinal Survey (NLS).
+Most NLSY and NLS data are available free of charge to by means of a web-based data portal \cite{NLSY_data}, access to which users must register for.
+Users must apply for special permission to access geographic-related variables.
+The Center for Human Resource Research (CHRR) at The Ohio State University and the National Opinion Research Center (NORC) at the University of Chicago administer both the data collection and repository functions.
+A Google Scholar search in mid-2015 shows more than 20,000 publications cite the NLSY.
+
+The Early Childhood Longitudinal Study (ECLS) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences.
+The ECLS consists of three cohorts, one focused on a sample of children measured beginning at birth and two cohorts following children from kindergarten.
+The birth cohort a sample of children born in 2001 and followed from birth through kindergarten entry, and unusual among comparable studies includes data from fathers. 
+One of the kindergarten cohorts studied approximately 22,000 children who entered school in 1998-99 and were followe through the eighth grade. 
+A second cohort entered kindergarten in 2010-11 and is being followed through the fifth grade.
+The project has received funding from the National Center for Educational Statistics (https://nces.ed.gov/ecls/), and the data are archived at ICPSR.
+ECLS consists of both public and restricted data, and according to a Google Scholar search has generated more than 8,000 publications, but ICSPR says 167.
+
+The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} was a survey designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
+Undertaken in response to a Congressional mandate, the study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S.
+during the 1994-1995 school year.
+The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
+The Ad Health cohort has been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32. 
+In addition to survey data Add Health includes biomarker data from a blood test taken during Wave IV. 
+Data are available about glucose, hemoglobin (HbA1c) levels, and blood lipids. 
+The Add Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
+As of mid-2015, more than 5,000 publications \cite{add_health_publications} had arisen from the Add Health dataset.
+
+In comparison, the NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
+The goal was to learn more about the kind and quality of child care experiences, and the effects of different child care settings on developmental outcomes. 
+Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, and anthropometric measures, and demographic information.
+The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
+Unlike the Ad Health study, data in SECCYD are not available to the public, but are restricted.
+Access is granted only by application, and approval is granted on a case-by-case basis.
+The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.
+
+Outside the U.S., there are several examples of large-scale longitudinal studies focused on children's growth and health.
+Launched in 2011, the Japan Environment and Children's Study <http://www.env.go.jp/en/chemi/hs/jecs/> is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
+Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
+Health outcomes and exposure measurements will continue until the participating children become 13 years old.
+The study is in its early stages, and has only produced a handful of publications <http://www.env.go.jp/en/chemi/hs/jecs/publications/index.html>.
+Data sharing plans have not been announced.
+World Health Organization (WHO) has offered to help coordinate measurement strategies for large-scale birth cohort studies <http://www.who.int/ceh/cohorts/en/> to ensure that data can be harmonized across them.
+Several birth cohort studies have been undertaken in the U.K., beginning with the National Survey of Health & Development in 1946 (https://en.wikipedia.org/wiki/National_Survey_of_Health_%26_Development).
+The National Child Development Study (NCDS) is a continuing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958. 
+The origins of the NCDS can be found in the Perinatal Mortality Survey (PMS) which was then sponsored by the National Birthday Trust Fund and set up to collect information about the social and obstetric factors associated with stillbirth and death in early infancy.[3] Examples of other topics which have since been included are medical care, health, home environment, educational progress, parental involvement, family relationships, economic activity, income, training and housing.[4]
+Conducted by the Centre for Longitudinal Studies (CLS), the aim of the study is to improve understanding of the factors affecting human development over the whole lifespan. 
+It collects information on physical and educational development, economic circumstances, employment, family life, health behaviour, wellbeing, social participation and attitudes.
+physical, educational and social development.[1] During the period 2002-2004, genetic information on participants was also obtained to examine the genetic effects on common traits and diseases.[2]
+The British Cohort Studies have generated more than 3,500 publications (http://www.cls.ioe.ac.uk/Bibliography.aspx?sitesectionid=647&sitesectiontitle=Bibliography) to-date.
+Data are available from (http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies) by registration.
+Another widely known and used dataset comes from the World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) (http://www.who.int/childgrowth/en/).
+The MGRS collected growth and related data from 8500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
+The data provide body mass index, height for children 5-19, and weight data norms for children 5-10.
+
+Not all large-scale birth cohort studies succeed.
+A notable example in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
+Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children from before birth to age 21. 
+However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel.
+Authorized by the Children’s Health Act of 2000, the NCS was a planned large-scale, long-term study of U.S.
+children and their mothers and fathers designed to study environmental influences on child health and development.
+Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
+Those data are slated for release in a data archive sometime in 2015.
+
+Several themes emerge from this sample of population-based studies.
+Datasets are large in terms of the volume of participants sampled ranging in the thousands to tens of thousands.
+The datasets demonstrate significant variety of measures, including survey-based and bio-marker-based (NHANES, Ad Health, SECCYD) and even video (SECCYD).
+The studies show variety geographic composition, as well.
+Some studies involve tens of sample occasions per individual, the NLSY and Ad Health studies in particular.
+Datasets are housed either in U.S. Government (NHANES, SECCYD) or university-based repositories (PSID, Ad Health, NLSY).
+Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center.
+Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
+% Susan McHale's comment about training. Check the other datasets for this.
+% Limits on geographic linkage, other ways data are anomymized to reduce possibility of identifiability.
+
+\subsubsection{Behavior Genetics}
+
+The next area of developmental science we explore is behavior genetics, where large data samples have long been required in order to detect what can often be small effect sizes.
+Several large-scale genetics studies that are not specifically developmental deserve mention.
+The Genome of the Netherlands Project(http://www.bbmri.nl/en-gb/activities/rainbow-projects/131-genome-of-the-netherlands) is a publicly available dataset from n=250 families, consisting of two parents plus one of their adult children \cite{Genome_Netherlands}.
+The Genes, Environment, and Melanoma Study (GEMS) focuses on risk for melanoma in a sample of 3,700 individuals from the North America, Europe, and Australia (http://gemstudy.org/main/index.html).
+The Psychiatric Genomics Consortium (PGC) (http://www.med.unc.edu/pgc) contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
+Many disorders featured (autism, attention-deficit hyperactivity disorder, and schizophrenia) in the PGC dataset have developmental dimensions.
+Genome-wide association study (GWAS) results from the PGC dataset may be viewed in a specialized web browser tool (http://www.broadinstitute.org/mpg/ricopili/).
+
+The U.K.-based Twins Early Development Study (TEDS) \cite{TEDS} (https://en.wikipedia.org/wiki/Twins_Early_Development_Study) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
+Some 13,000 pairs remain actively involved, providing lab-based behavioral task and survey responses.
+DNA samples from more than 5,000 twins pairs are also available.
+TEDS has generated more than 350 publications to-date.
+Researchers wishing to gain access to the data may do so by submitting a formal data request specifying the aims of the planned research and the variables that needed satisfy the aims. 
+Proposals that do not overlap with analyses already being planned or carried out by the TEDS research team or other collaborators are usually approved.
+
+The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} focuses on genetic and environmental contributions to measures of family relationships and mental health.
+TOSS extends the Twin Moms Project to include a sample of twin fathers (N=320 twin pairs and their families).
+In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (n=3,000 individuals) and their families.
+While still relatively early in data collection, more than 24 publications reference the TOSS dataset.
+Data sharing plans have not yet been announced.
+
+Illustrative U.S. examples include the Colorado Adoption Project (CAP), the Nonshared Environment in Adolescent Development (NEAD) project, and the Early Growth and Development Study (EGADS).
+NEAD studied a sample of 720 families, including twins, siblings, and their parents, in a longitudinal design that involved up to 3 data collection waves.
+Researchers focused on survey, self-report, and experimenter ratings (from video) of family relationships, adolescent adjustment and competence.
+A Google Scholar search shows 73 publications to-date.
+The data will be shared by request to the project team.
+The team plans to share the data openly later in 2015.
+
+Adoption studies provide a unique window on the mix of genetic and non-genetic factors that influence child development.
+Since its inception in 1976, the Colorado Adoption Project (CAP) (http://ibg.colorado.edu/cap/index.html) has enrolled more than 2,400 participants from more than 450 families.
+Data include cognitive abilities, temperament measures, demographic variables, 
+Additionally parents completed a questionnaire assessment which included measures of temperament
+CAP has generated more than 200 published articles (http://ibg.colorado.edu/cap/publishedarticles.html).
+Data collection continues.
+The study aims to be the definitive genetically informed longitudinal study of lifespan development \cite{rhea_colorado_2013}.
+Data from 1976-1989 are archived at Dataverse (https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/00913) and can be accessed by application to Dataverse.
+EGADS is a prospective adoption study of (n=561) triads of birth parents, adoptive parents, and adopted children in the U.S. 
+The study focuses on genetic and non-genetic factors that influence children’s adjustment from infancy to age 9. 
+The data include self- and other-report measures, DNA, and salivary cortisol.
+Data are available by request, with plans for more open data sharing in the near future.
+
+\subsubsection{Brain Imaging}
+
+Several large studies have focused on developmental patterns in brain structure or activity.
+For practical reasons -- motion artifacts make imaging very young children extremely challenging -- most projects have focused on children and adolescents.
+One of the earliest studies of its kind was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
+The study collected structural brain information from 145 4 to 20-year-olds along with behavioral and clinical data.
+More than 100 of the participants were scanned on more than one occasion, with the time between scans approximately two years.
+The original paper describing this study has been cited more than 3,300 times, as of mid-2015.
+The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data appear not to be available for analyses by researchers outside the original investigative team.
+The NIH MRI Study of of Normal Brain Development \cite{NIH_Pediatric_MRI} collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
+A second cohort of children from birth to 4 years was scanned longitudinally with up to 10 scans. 
+Demographic, hormonal, cognitive, affective, and psychiatric data were also collected.
+The data are available to qualified researchers by application, and the dataset has generated some highly cited papers since data collection ended in 2006 \cite{group_total_2012, zielinski_network-level_2010}.
+John Richards and colleagues have combined data from the NIH MRI Study, data collected in their own labs, and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI \cite{Richards_Developmental_MRI_Database, richards_database_2015}.
+These data can be openly shared with appropriate citation.
+More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING) Project \cite{PING} collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from Baltimore, Boston, Honolulu, Los Angeles, New Haven, New York, Sacramento, and San Diego.
+The study was funded by the Eunice Kennedy Shriver National Institute of Child Health and Human Development.
+Data are available to the research community by application.
+As of mid-2015, the project had generated more than 20 publications.
+The Developmental Connectome Project \cite{developmental_connectome}, led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
+This project aims to study the connectome, a map of human brain connectivity, from 20 to 44 weeks post-conception.
+Funded by the European Research Council, the imaging data are accompanied by clinical, behavioral and genetic information.
+Investigators hope that the data set will provide a basis for studying genetic and environmental risks that could lead to neurodevelopmental disorders such as Autistic Spectrum Disorder or Cerebral Palsy.
+The project has already generated 17 publications as of mid-2015.
+
+Looking across the set of large-scale pediatric brain imaging studies described here, several themes emerge.
+One is that these studies require large, complex research teams, usually spanning multiple sites.
+Another is that, with some notable exceptions \cite{richards_database_2015}, when the data are available for use by others, that use is controlled by the original investigative team.
+In some cases, the original investigative team must be included as an author on any publication.
+Another trend among large-scale developmental imaging studies is that the data are usually housed locally, on project-specific sites and not generally on centralized servers that aggregate data across studies and sources \cite{openfmri.org, neurovault.org}.
 
@@
 
-   % + Mass media interest in longitudinal perspective
-   %      * Up Series, <http://en.wikipedia.org/wiki/Up_Series>
-   %      * Boyhood
-   %       * Perception
-   %          - Teller Acuity Card norms
-   %              + <http://www.ncbi.nlm.nih.gov/pubmed/7890497>
-   %              + <http://www.ncbi.nlm.nih.gov/pubmed/7890496>
-   %          - Auditory testing norms
-   %      * Action
-   %          - Gesell?
-   %          - Bayley Scales of Infant and Toddler Development
-   %              + Nancy Bayley bio, <http://www.pearsonclinical.com/education/authors/bayley-nancy.html>
-   %      * Cognition
-   %          - Wechsler
-   %              + <http://www.iupui.edu/~flip/wechsler.html>
-   %              + Wechsler Adult Intelligence Scale (WAIS)
-   %              + Wechsler Intelligence Scale for Children (WISC-III), 6-16 year-olds
-   %              + Wechsler Preschool and Primary Scale of Intelligence (WPPSI-R), 4-6 1/2 years.
-   %      * Language
-   %          - Human Speechome, <http://www.media.mit.edu/cogmac/projects/hsp.html>
-   %          - [CHILDES/TalkBank](http://talkbank.org/)
-   %          - Smith, Aslin, Yu, Frank word learning, grammar
-   %          - MacArthur, CDI and Mike Frank's project (wordbank.stanford.edu)
-   %          - Cathie Tamis-Lemonda
-   %          - Gedeon Deak
-   %          - LENA, <http://www.lenafoundation.org/>
-   %      * Emotion / Personality / Temperament
-   %          - Ecological Momentary Assessment
-   %              + Matthias Mehl
-   %              + Diary studies
-   %          - Infant Behavior Questionnaire, CBC
+In the realm of behavior, there are a number of measures that have become standards whose widespread adoption has resulted in data that are large in volume, velocity or variety.
+Davida Teller pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
+A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.
+The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} arose from the Nancy Bayley's Berkeley Growth Study \cite{jones_berkeley_1941} that studied the behavior, motor coordination, physical activity, cognitive function, and maternal behavior of 61 healthy newborn infants.
+The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
+The Bayley is widely used in clinical and epidemiological settings such as the EPICure <http://www.epicure.ac.uk/> because of the norms.
+However, data about those norms appears to be under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
+Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE), 1972-1992 project on ICPSR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4091?sortBy=19&amp;q=Bayley&amp;searchSource=revise).
+The data are freely available to registered users of ICSPR.
+
+The Wechsler intelligence tests were developed by Dr.
+David Wechsler, beginning in the 1930's.
+The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, while the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
+The current version (WISC V) is provided by Pearson (http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html).
+Data about the norms themselves appears not to be available to researchers.
+However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR (https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/13604?q=wisc&amp;searchSource=find-analyze-home&amp;sortBy=).
+The data are restricted, and thus are only available by specific authorization.
+The Family Life Project also collected and reported WPPSI data.
+
+There are several large-scale datasets associated with language processing.
+The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established.
+It consists of transcripts and audio and video recordings of children's utterances along with recent data from adults with aphasia.
+CHILDES/TalkBank has generated more than ten thousand of citations, according to Google Scholar, and the data are available to the research community, some of them publicly.
+The Human Speechome Project (http://www.media.mit.edu/cogmac/projects/hsp.html) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
+The project has generated more than 78 publications (http://www.media.mit.edu/cogmac/publications.html), but the data have not been made available outside of the original investigative team.
+Data was gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
+% edit previous
+WordBank (http://wordbank.stanford.edu/) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
+Catherine Tamis-Lemonda has released video data (https://nyu.databrary.org/volume/8) from a longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
+The LENA Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed (http://www.lenafoundation.org/customer-resources/videos).
+LENA stands for Language ENvironment Analysis.
+The system provides an automatic language collection and analysis tool for speech language professionals and parents.
+The system was designed and is distributed by the LENA Foundation.
+
+%- Infant Behavior Questionnaire, CBC
+While no archive for IBQ data across studies exists, IBQ data on 1,388 participants (n = 658) who were group matched within site with a group of non-exposed comparison infants (n=730) are archived at ICSPR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34312?sortBy=19&amp;q=infant+behavior+questionnaire&amp;searchSource=revise) as part of the Maternal Lifestyle Study in Four Sites in the United States, 1993-2011.
+MSL was a large NIH longitudinal studies of children with prenatal cocaine exposure (PCE).
+MLS was a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
+IBQ data are also available from ICPSR from the Family Life Project (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=19&amp;q=ibq&amp;searchSource=revise).
 
@@
 
-        %    - Bureau of Labor Statistics (BLS), youth employment, <http://www.bls.gov/news.release/youth.nr0.htm>
-        % * Education
-        %     - MOOCs
-        %         + Stanford, Coursera; Harvard, MIT, EdX.
-        %     - SAT, ACT scores, <http://www.erikthered.com/tutor/sat-act-history.html>
-        %     - Aptitude, <http://timss.bc.edu/>
-        %     - OECD PISA, <http://pisa2000.acer.edu.au/index.php>
-        %     - Measures of Effective Teaching (MET) Study, <http://www.metproject.org/>
-        %         - "The MET project was a research partnership between 3,000 teacher volunteers and dozens of independent research teams.
-The project's goal was to build and test measures of effective teaching to find out how evaluation methods could best be used to tell teachers more about the skills that make them most effective and to help districts identify and develop great teaching.
-Launched in 2009, the study has identified multiple measures and tools that – taken together – can provide an accurate and reliable picture of teaching effectiveness."
-        %         - "The project was funded by the Bill & Melinda Gates Foundation as part of ongoing efforts to give teachers the tools they need to be successful and to improve student achievement in public schools across the United States.
-2.
-WHO WAS INVOLVED IN THE MET PROJECT?Truly understanding the skills and techniques that are hallmarks of great teaching requires collaboration between educators and researchers.
-The MET project was unprecedented: it was a partnership among thousands of teacher volunteers and administrators and union leaders from school districts across the country and dozens of independent researchers and education organizations.
-Teachers – more than 3,000 of them – were at the heart of the study.
-The teacher volunteers were recruited from Charlotte-Mecklenburg Schools, N.C.; Dallas Independent School District, Texas; Denver Public Schools, Colo.; Hillsborough County Public Schools, Fla.; Memphis City Schools, Tenn.; New York City Department of Education, N.Y.; and Pittsburgh Public Schools, Pa.
-Pittsburgh served as the project's pilot district, but no data from this district was analyzed.
-Lead researchers involved in the project were affiliated with Dartmouth College, Harvard University, the University of Michigan, the University of Virginia and Stanford University.
-Participating non-profits and education companies included RAND Corporation, Educational Testing Service, Teachscape, The Danielson Group, The New Teacher Center, National Math & Science Initiative and Westat."
+There are several large datasets with information about employment patterns and educational attainment in children and youth
+The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
+The U.S.
+Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers only under a restricted use agreement.
+The College Board shares SAT and college cost and scholarship data with the research community by application (http://www.collegeboard.com/inquiry/csinquiry_form.html).
+The ACT (http://www.act.org/research/). Privacy policies (http://www.act.org/privacy.html). 
+The Trends in International Math and Science Study (TIMSS) studied mathematics and science achievement in 4th and 8th grade for more than 20 years.
+Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
+Both studies provide data available for public download (http://timss.bc.edu/).
+The 1999 TIMSS video study examined teaching practices in seven countries, and some of the materials are available for public use (http://www.timssvideo.com/).
+The Measures of Effective Teaching (MET, http://www.metproject.org/) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
+The MET video data are stored and shared at ICSPR.
+The OECD-funded PISA dataset is available (http://pisa2000.acer.edu.au/index.php).
+Data from the performance of students in Massive Open Online Courses at Harvard and MIT are available to the research community by application.
+
+The 
+LearnLab!
+
           %      * Attitudes
@@
 
-       % * [Panel study on income dynamics](http://psidonline.isr.umich.edu/)
-       %  * [NIH study of early child care and youth development (SECCYD)](https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx)
        %  * [Family Life Project](http://flp.fpg.unc.edu/)
+
+       The purpose of this project is to study the early development of a group of children who are at risk regarding later successful adjustment and for whom we have little information: children living in rural, largely poor communities.
+A birth cohort of 800 children in three rural counties of North Carolina and 600 children in three rural counties of Pennsylvania were studied over the first 3 years of life.
+       https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=&amp;q=Bayley&amp;searchSource=revise
+
+MacArthur Network on Transitions to Adulthood (http://www.transitions2adulthood.com/). Data not obviously open or available.
+Research Network on Adolescent Development & Juvenile Justice - See more at: https://www.macfound.org/networks/research-network-on-adolescent-development-juvenil/#sthash.PQOwqv14.dpuf
+Research Network on Youth and Participatory Politics - See more at: https://www.macfound.org/networks/research-network-on-youth-and-participatory-politi/#sthash.egU1Wwvy.dpuf
+
        %  + Normative vs. perturbed development
@@
        %      * NPR story, <http://www.npr.org/blogs/health/2014/02/20/280237833/orphans-lonely-beginnings-reveal-how-parents-shape-a-childs-brain>
-       %      * Ros Picard, ASD
 
-\section{The Future of Big Data}
+\section{The Future of Big Data in Development}
+
+Clearly, the collection, analysis, and sharing of large, multilevel datasets has been part of the fabric of developmental science for a long time.
+In this section, we discuss a range of technical, conceptual, and theoretical issues that arise in thinking about the future of big data in developmental science.
 
@@
 
-\subsubsection{Collecting data with New technologies, data streams}
-        % * Social media
-        %     * Twitter, Facebook, Snapchat, etc.
-        % + Other web-based data collection tools
-        %     * Lookit
-        %     * Tablets
-        %     * Portable devices
+Technical issues associated with big data in developmental science center on collection, storage and retrieval, data management, provenance, and analysis.
+
+\subsubsection{File and Data Formats}
+
+\subsubsection{Collecting data with new technologies, data streams}
+
+As described previously, developmental scientists have and continue to collect data at multiple levels of analysis, from genetic, endocrine, and metabolic measures to physical structure, brain activity, behavioral measures, to peer, family, school, neighborhood, and geographic/cultural context information.
+Some researchers believe that advances in understanding about many development processes will require deeper understanding about how these influences combine to influence trajectories of change.
+If true, achieving that understanding will require continued improvements in the mechanisms of measurement, and the tools for converting these measures into electronic data that can be cross-linked and analyzed.
+
+For some some data elements, the devices used to make measurements already provide data and metadata in a structured, organized way.
+For example, genetic analyses from modern gene sequencing tools provide machine-readable outputs that (Rick talk to David Vandenburgh).
+Researchers who analyze hormone levels in saliva or micronutrients in blood, for example, (Rick asking Sam Scott, others).
+Magnetic Resonance Imaging (MRI) systems produce electronic image data and machine-readable subject-level metadata; however, many research teams limit the amount and kind of subject-level metadata they enter into MRI databases because of the possibility of violating research participant confidentiality.
+Physiological measurements like EEG, heart-rate or skin conductance, produce electronic files of the time series of measurements.
+But, unlike MRI, there are no standard file formats, and most data collection systems provide no standard subject-level metadata.
+Thus, the files require significant post-collection data processing prior to analysis. 
+Behavioral measures involving computer-based tasks are commonly used in developmental research.
+But, like physiological measures, many, if not most of these measures involve the use of custom, project-specific software, so the output data files, while often in an electronic form, may or may not be produced in a form that can be easily linked with other data items without significant post-processing.
+
+Recordings such as still images, audio, and video are readily collected or converted to an electronic form.
+There are a number of still image formats, but readily available tools exist to convert between them.
+Audio and video provide a bigger challenge because of the diversity of formats.
+Data archives like Databrary that specialize in video, for example, have chosen to convert all video into a common format in addition to storing files in their original formats for preservation purposes.
+
+At the same time, many, perhaps most developmental researchers continue to collect data elements using paper and pencil measures, making some form of manual data entry essential.
+
+Looking ahead, there is increasing use of web-based tools for data collection (Qualtrix, SurveyMonkey).
+Amazon's Mechanical Turk (http://mturk.com), for example has become a hub for conducting large-scale behavioral science experiments, (see also Todd Gureckis' Psyturk).
+Amazon's terms of use prohibit minors, but some developmental researchers have found ways to carry out studies that require parental consent.
+The LookIt platform developed at MIT allows parents to give video-based informed consent for their child to participate in a behavioral study, such as one involving the measurement of looking times via a laptop or tablet computer's built-in web cam.
+A number of researchers have adapted smartphones to serve as devices to collect survey, behavioral/task-related, and ambient sound information (Sliwinski, Mehl)
+
+Other social media providers like Twitter, Facebook, Snapchat, Instagram, and YouTube collect individual data and metadata.
+Services are free, so business model involves selling data.
+Apple's HealthKit.
+Other wearable devices like FitBit, Mio, collect data and store it in a private, proprietary cloud server.
+Data is electronic, but data may or may not be accessible to user.
+You can see what data Google has about you.
+New initiatives like DataWallet (https://www.datawallet.io/) are on the horizong that may empower individuals to share their data for a fee.
+
+Both old and new data collection systems raise a number of issues.
+How can critical metadata be captured more efficiently without significantly increasing risks to privacy.
+Video cameras that record recording date and time and camera specs, for example.
+Reduce need for post-processing of data, more widespread use of tidy and reproducible data formats.
+Non-proprietary data streams and storage.
+Raise ethical questions about who owns data.
+
+\subsubsection{Storing and Sharing big data}
+
+Big data requires large scale servers.
+If the data are to be served over the web, then the servers require signficant network bandwidth, and all of the other substantial infrastructure required of large-scale data centers.
+
+Researchers who wish to find and reuse or store and share data face a bewildering array of options.
+
+\subsubsubsection{Storing and Sharing Data}
+
+The options include individual or institutional websites, or cloud services like Dropbox, Box, or Amazon Storage services.
+Other researchers may choose institutional archives like Penn State's ScholarSphere (https://scholarsphere.psu.edu/).
+There are domain specific archives like ICSPR, Databrary, TalkBank, or WordBank, or , OpenfMRI and NeuroVault.
+Other researchers may choose to store and share their data with domain-general services like Researchgate.net, FigShare/SlideShare, Dataverse, or the Open Science Framework.
+Other researchers are turning to open source web sites like GitHub.
+Some journals offer or require storage of datasets, but these are typically limited to text-based flat-files used for statistical analyses and not raw images, videos, or physiological time series.
+What to store, when to store (during, after), and how to share, how to get credit/notice/citations for sharing.
+
+\subsubsubsection{Finding Data}
+
+The task confronting researchers who wish to find and reanalyze or reuse data faces challenges, as well.
+How to discover data?
+What are the key terms? Demographic variables like age and sex. Task, context, language?
+Data associated with specific publications, grants, or researchers.
+Obviously repositories.
+But, can we find the data we want?
+Governments also maintain data archives such as the National Database for Autism Research, the National Center for Educational Statistics, the U.S.
+Census Bureau, the Centers for Disease Control, and the Bureau of Labor Statistics, to name a few sources of developmental science-related data discussed in this paper.
+A number of large-scale research project maintain their own data repositories.
+ICSPR, for example, enables variable-level searching, but not all datasets provide it.
+The International Cross Time Cross System Dataset (XTXS) (http://www.intledstatsdatabase.org/) aspires to provide a comprehensive source of data about educational outcomes.
+The APA maintains a web-page with links to datasets (http://www.apa.org/research/responsible/data-links.aspx).
+
+It is much easier to find publications and search the network of references than it is to find data.
+
+Rick's case study in trying to find and download a public dataset from ICPSR.
+
+Bring data to the analyst or analyses to the data?
+Problem will get bigger as datasets grow in size, complexity.
 
 \subsubsection{Coding and analyzing big data}
-        % * Audio
-        % * Video
-        % * Surveys
-        % * Physiological data streams
-        % * Geo-location
 
-\subsection{Storing and sharing big data}
-        % * Where to store
-        % * What/when to share
-        % * How to share
-        % * NDAR
-        % * Dataverse
-        % * ICPSR
-        % * Databases of datasets
-        % * International Cross Time Cross System Dataset (XTXS), <http://www.intledstatsdatabase.org/>
-        % * APA's list, <http://www.apa.org/research/responsible/data-links.aspx>
-        %         * Domain-specific
-        %     - NDAR, etc.
-        %     - MacArthur Network
-        % * Domain-general
-        %     - UNC
-        %     - NYU, Berkeley, Washington, big data science centers
-        %     - PSU/Databrary
-        %     - Michigan
+Most data science work involves `janitor work' (http://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html?_r=0) or data carpentry (http://datacarpentry.org/).
+Turning raw materials into numbers.
+How were the materials converted into numbers?
+Data provenance.
+Term definitions, coding books.
+Example, from video to coded spreadsheets to numbers.
 
-\subsection{Conceptual and Theoretical}
+Emerging sentiment around the use of tools that enable reproducible science.
+RStudio, Jupyter.
+Version control.
+Text-based, automated/scripted workflows. Galaxy?
+Scripting in Datavyu.
 
-\subsubsection{Ethics}
-        % * Privacy
-        %     - FERPA
-        %     - HIPAA
-        %     - Differing standards across countries
-        %     - Commercial vs. academic research practices, cultures
-        %     - "Linking" data improves potential inferences, but increases privacy risk.
-        %     - Can minors consent to "indefinite" data storage?
-        % * Security
-        %     - Consent
-        %         + Open Humans Project, Personal Genome Project
-        %         + Databrary
+Need for tools and workflows that keep track of the history of researcher's actions.
+MS Word's track changes, but for data cleaning, coding, analysis.
+
+\subsection{Ethical, Conceptual and Theoretical Issues}
+
+The collection, analysis, and interpretation of large scale datasets poses a number of ethical, conceptual and theoretical issues.
+
+\subsubsection{Ethics and Privacy}
+
+Developmental researchers often collect data from vulnerable populations of research participants.
+Professional ethics require that special care be taken about what data are collected from research participants in general and who gains access to it.
+The focus in developmental science on studying vulnerable research populations magnifies these concerns.
+
+Differing practices across cultures in terms of privacy pose challenges for collecting and aggregating datasets.
+In the U.S., researchers must navigate a regulatory environment in which different types of data are covered under different sections of Federal law.
+For example, the Federal Educational Rights and Privacy Act (FERPA - http://www2.ed.gov/policy/gen/guid/fpco/ferpa/index.html) governs who has access to a student's educational records.
+The Health Insurance Portability and Accountability Act (HIPAA - http://www.hhs.gov/ocr/privacy/hipaa/understanding/), among other functions, governs the process by which individually identifiable health information may be disclosed and to whom.
+If the data in question are audio or video recordings, different state provisions can come into play.
+For example, in two-party states \cite{http://www.dmlp.org/legal-guide/recording-phone-calls-and-conversations} both the person making the recording and the person or persons being recorded must consent.
+
+Research activities funded by the U.S.
+Federal Government, typically in colleges, universities, or academic or private medical centers, are supervised by an Institutional Review Board or its equivalent.
+These entities are regulated by the U.S. Office of Health and Human Services (http://www.hhs.gov/ohrp/).
+Researchers supervised by IRBs must respect privacy, secure informed consent or assent, and maintain confidentiality. 
+Adhering to these ethical principles has practical consequences for research.
+Researchers have limits on the ways they are able to recruit and identify participants.
+Minors must give informed assent to participate in a study with a parent or guardian giving formal consent.
+Data must be collected in ways that minimimize the likelihood that information given by a participant will be disclosed or a participants' identity revealed outside the research team.
+This usually means that data items that could reveal a participant's identity are removed or altered to reduce the likelihood of disclosure.
+Whether `de-identified' data of this sort can be shared with researchers outside the original team that collected depends on several factors.
+One factor is the sensitivity of the data collected and the likelihood that specific identities or home locations or other items could be revealed.
+Another factor concerns whether participants were informed that de-identified data might be shared outside the research team.
+IRBs may view these matters differently, creating additional complexities.
+Some IRBs may require participants to be informed that de-identified data might be shared outside the IRB-approved research team, and others may deem the analysis of de-identified data as no longer meeting the definition of human subjects research.
+From a big data perspective, if data cannot be shared outside the original IRB-approved research team, then the possible analyses are restricted to the interests, resources, and expertise of that team.
+% What about population data......
+
+Of course, some data types like photographs, audio or video recordings contain identifiable information that cannot be removed or altered without reducing the value to others.
+Thus, data from photographs or recordings requires additional consideration and special care.
+Databrary (http://databrary.org), a digital data library specialized for storing, managing, and sharing video data from developmental research has an access model that empowers researchers who wish to share identifiable research data to do so with explicit permission of the participants.
+Databrary has created template language to help researchers gain participants' permission.
+Further, Databrary restricts access to its identifiable data to researchers who have formally agreed to uphold ethical research principles and whose institutions approve of their access.
+The notion that research participants can consent to share identifiable or potentially identifiable data is relatively new.
+The Personal Genomes Project, Open Humans Project, and Human Connectome Project embody similar principles.
+The experience of Databrary investigators is that a significant proportion -- more than 75 percent -- of research participants and their parents or guardians will consent to sharing identifiable data, mostly video, with other members of the research community.
+
+It is too early to predict whether it will become commonplace for academic developmental researchers to seek explicit permission to share research data, including identifiable elements, with other researchers.
+There are reasons to be optimistic, however.
+In just over a year of operation, Databrary has secured formal agreements with more than 75 institutions in North and South America, Europe, Australia, and Asia allowing more than X researchers to access identifiable data.
+On the other hand, some leading developmental researchers have argued that the families of research participants forge a relationship of trust with a particular research team, formalized through the informed consent document \cite{Eisenberg-APS}.
+The relationship might be harmed or the research project negatively affected if participants were asked to share data with other researchers.
+Sensitive to the latter argument, Databrary has recommended that permission to share be sought separate from consent to participant in research and after a given data collection has ended.
+Moreover, the fact that a significant proportion of families agree to share when asked suggests that the relationship of trust involved in research participation might be extended to a community of researchers, given suitable provisions and constraints.
+Undoubtedly, seeking explicit permission to share on a consistent and widespread basis would resolve any ambiguity about whether a given dataset can be shared with whom and for what sort of purposes.
+
+Greater transparency and more explicit clarification about what data is being collected and for what purposes could be sought from commercial entities, as well.
+Social media companies like Google, Facebook, Twitter, SnapChat, and Instagram have business models that involve the collection, mining, and packaging of data, usually to advertisers, in exchange for services that are free to users.
+While some services attempt to restrict the youngest ages at which users can create accounts, the limits often lack rigor, and there is no parallel to the requirement of adult consent required in formal research contexts.
+The data collection and analyses carried out by private entities are subject to no supervision or formal regulation comparable to academic researchers.
+Instead, data use, analysis, and sharing provisions are governed by terms of use agreements users must indicate agreement with, usually by clicking a button, prior to using a given service.
+Unlike academic settings, where violations of research ethics principles may involve significant consequences for the researchers and institutions, violations of commercial terms of use require aggrieved parties to seek redress through litigation.
+The software industry has recommended its members adopt common data privacy principles \cite{}, but these guidelines are voluntary.
+
+Summary.
+Many big data sets have restrictions on them, aren't permissioned for open, widespread sharing
+Must be used at Restricted Data facilities, can't be used to study individuals.
+Studies of individuals may or may not have permission to share widely.
+New models for gaining permission to share identifiable data on a restricted basis may expand the availability of big data for data reuse.
+Have to ensure that data reuse is not prohibited by formal consent process.
+Have to ensure that data don't have to be destroyed.
+Remaining concerns about possibility that data linkage increases the risks of reidentification.
+There are questions about whether it is essential to reconsent minors when they become adults.
+Reconsenting may be impractical.
+General concern about whether it is possible to consent to sharing of an indefinite nature.
+Security can't be guaranteed, so risks can be minimized, but not reduced.
+Files that are downloaded increase risks, for image streams, it is impractical to prevent image/sound capture.
+Principle of informed consent important to uphold.
+Not clear how fully informed consumers are in using free commercial services.
+
+\subsubsection{Transparency and Reproducibility}
+
+The social and behavioral sciences have seen a string of high profile cases of scientific misconduct in recent years, including cases of fraudulent data \cite{}.
+The credibility problem is magnified by several factors.
+That lack of power and unrestricted exploratory analyses may mean that most research findings are false \cite{Ioanides}.
+The actual effect sizes of published findings is unknown due to a bias toward publishing positive results.
+Most journals reject papers that report failures to replicate published findings, and as a result, few scientists attempt or are recognized and rewarded for conducting genuine replications \cite{Nosek-Spies}.
+The problem is so serious that some have claimed that science as a whole faces a crisis of replicability.
+
+To address the crisis, the Center for Open Science (http://cos.io) has organized several large-scale replication efforts, including some in psychological science under the umbrella of the ``Many Labs'' project \cite{klein, https://osf.io/ct89g/; https://osf.io/8cd4r/}.
+The results of these pre-registered, open, large sample replications have been mixed.
+Some published effects clearly replicated, but others did not.
+
+Whether there are replicability problems developmental science and whether they constitute a crisis is not known.
+Developmental research reflects the same positive effects biases seen in other fields, and the same problem that null results often sit unpublished in file drawers -- the so-called file drawer effect \cite{rosenthal_file_1979}.
+No failures to replicate developmental studies have been reported to Psychfiledrawer.org (http://psychfiledrawer.org/view_article_list.php), a resource designed to bring replications failures to light.
+As some developmental researchers have written (http://deevybee.blogspot.co.uk/2012/01/novelty-interest-and-replicability.html), replicating effects with developmental populations can be especially difficult and so even partial replications are noteworthy.
+No large-scale replication efforts in developmental science have been mounted, but there have been calls for changes in journal practices to give replications have a more privileged place in scientific publications (http://deevybee.blogspot.co.uk/2012/01/novelty-interest-and-replicability.html).
+One barrier to more open data practices appears to be researcher's fears of having one's reputation or abilities publicly undermined \cite{ascoli_ups_2006}.
+So, changing views about replication may require shifts in the scientific culture.
+Researchers should work to reduce the extent of blame levied at researchers whose initial positive findings fail to be replicated by others (http://deevybee.blogspot.com/2015/07/publishing-replication-failures-some.html).
+Technological tools that foster increased openness and transparency and more systematic research data management (OSF and Databrary) will also contribute to changing the scientific practices.
+So will the widespread adoption of more consistent journal practices related to transparent and open scientific practices \cite{nosek_promoting_2015}.
+
+Still, the increasing availability of large-scale datasets about developmental questions promises to magnify problems at the intersection between exploratory and confirmatory research.
+Large volume, higher velocity or variety datasets make it possible to explore and discover novel unpredicted patterns in data.
+But, novel findings might be spurious, and exploratory findings must be properly confirmed.
+While pre-registration and pre-review have been suggested as one way to address the problem of spurious exploratory findings, these tools may not be practical in all cases and could have a chilling effect on discovery.
+In contrast, increased transparency about the process that led to an exploratory finding and the steps taken to confirm it can bolster a finding's credibility.
+Thus, developmental researchers may find it essential to adopt more transparent and reproducible workflows using some of the new tools developed for this purpose (COS; Databrary).
+
+\subsubsection{Community Engagement and the Impetus for Change}
+
+Developmental researchers have clearly shown enthusiasm for sharing with the research community and the public at large the results of their findings via publications, and in some subfields the sharing of data, materials and methods is firmly established.
+Open sharing practices tend to be more common when there is a high cost, centralized source of scientific data that could not conveniently be owned or managed by researchers themselves (e.g., space telescopes or the U.S.
+Census).
+Successful big data projects -- repositories or specific studies or datasets -- appear to have fostered a high level of engagement around a particular resource, often through specific training, and targeted workshops (e.g.
+Ad Health, ICSPR, Databrary, AERA workshop on video data sharing, COS, Data/Software Carpentry).
+In addition to bottom-up/grassroots initiatives, journals and funding agencies continue to play a vital role in creating an impetus for change.
+Funders can require data management plans, mandate data be deposited into particular types of open repositories (Gates Foundation; NIMH call), and provide funding to build and support big data infrastructure.
+Journals can also require that data be deposited in open archives as a condition of publication in addition to adopting other transparent and open science practices for manuscripts they accept.
+However, the problem with data sharing mandates from funders is that there is no specific funding mechanism to provide sufficient ongoing financial support for data archives, and few researchers budget funds to support data management and archiving.
+And, while some journals are willing to take on the burden of storing and sharing data associated with publications, others (Journal of Neuroscience) have gone the other way.
+Thus, in the interests of promoting greater openness and transparency, funders and journals may create unfunded mandates that make it harder for researchers to make discoveries.
+For example, new regulations about when data must be deposited may be unwieldy and impractical for developmental scientists. 
+See Nancy Eisenberg piece on data sharing. (https://www.psychologicalscience.org/index.php/publications/observer/2015/may-june-15/thoughts-on-the-future-of-data-sharing.html).
+APA stance on data sharing. (http://www.apa.org/science/leadership/bsa/data-sharing-report.aspx).
+
+\subsubsection{Ownership and Access}
+
+Who owns research data? (http://ori.hhs.gov/education/products/n_illinois_u/datamanagement/dotopic.html; http://ori.hhs.gov/education/products/n_illinois_u/datamanagement/dotopic.html).
+The scientific community lacks consensus on the answer, and the developmental science community is no exception.
+Federal funding agencies might argue that the public should `own' data paid for by tax dollars, much like other data collected by government agencies like the U.S.
+Census, National Weather Service, and U.S.
+Bureau of Labor Statistics.
+The institutions that employ, receive, and manage federal grants might stake a claim to ownership.
+Most investigators naturally feel a strong sense of ownership over their intellectual products, even though formal copyright is often surrendered in the process of publishing, and that sense extends to data.
+Some have even argued that research participants themselves own their own data, and there are new business models emerging that may soon provide individuals an opportunity to sell data for personal gain (http://www.datawallet.io).
+
+The lack of consensus about who owns data means that access is often limited in ways that impede reuse by others.
+Some investigative teams control who has access to datasets, for what purposes and for how long.
+That control may persist indefinitely.
+Others grant access to data only if co-authorship on any published product is guaranteed.
+While legimate arguments might be made in favor of `embargo' periods that enable teams of researchers to mine and report findings from their research efforts, the ideal of fostering greater data reuse argues for shorter periods.
+Establishing consensus about data ownership and the kind of control investigators can exercise over it will require conversations among researchers, institutions, and funding agencies.
+That consensus may well prove vital to achieving some of the benefits of big data analyses in development.
 
 \subsubsection{Challenges of conducting big data developmental research}
-     %    * Cohort effects, development is developing
-     %    * What is normative? What is not?
-     %    * Instruments/measures public and open and free vs. private and closed and expensive
-     % Most of psychology (especially developmental?) is WEIRD
-    	% - How do large scale patterns arise from micro processes
-    	% - The end of theory? [^5], [^6], Chris Anderson.
-    	% - Questions for Big Data (boyd & Crawford), [^7] 
-    	% - Social Sciences are underpowered, most research findings are false. Ioanidis. [^9]. Center at Stanford on research methodology. METRICS. [^10]
+
+The increasing availability of big datasets for analysis in developmental research poses significant theoretical and conceptual questions alongside the many pragmatic ones already discussed.
+Big(ger) data may help to overcome limitations with our existing knowledge base.
+One challenge that big data may help to address is a particular bias in existing samples.
+Developmental research presumes to study what is normative about changes across time in human behavior.
+But, much of what behavioral and social science has learned about developmental processes comes from samples that represent only a small fraction of the world's population.
+Developmental psychology, like other branches of the field, presents findings from Western, education, industrialized, rich, and democratic (WEIRD) societies (http://www.theatlantic.com/daily-dish/archive/2010/10/western-educated-industrialized-rich-and-democratic/181667/).
+So, to the extent that new tools enable research on development in non-WEIRD cultures and those data can be aggregated and combined, the extent to which developmental science can make claims about universal or near-universal components of developmental processes will be strengthened.
+On the other hand, developmental researchers are well aware of cohort effects -- the notion that developmental processes highly influenced by changing social and cultural norms are themselves subject to change.
+Thus, even the most culturally diverse dataset possible may still yield conclusions that are `locked' in time.
+
+A second challenge larger datasets may help to address is the fact that most social, behavioral \cite{Maxwell-2004} and neuroscience studies \cite{Button-NRN} are underpowered.
+Most worryingly it has been claimed that most published research findings are false in fields with specific characteristics: Sample or sizes are small; more relationships between variables are tested; preselection of tested relationships is less common and exploratory research more common; designs, definitions, outcomes, and analytical modes are less consistent across studies; and when more teams are involved seeking out significant effects \cite{Ioanidis}.
+The collection, analysis, and sharing of larger size datasets may help to ameliorate some, but not all, of these flaws.
+
+Developmental research faces a specific point of tension related to the measurement.
+Many of the measures for which high volume data are available come from proprietary, expensive to license instruments like the Bayley, WIPPSI.
+Where free, academic instruments are in use like the Infant Behavior Questionnaire, there remains no centralized archive for those data.
+Plus, the measures themselves have been revised several times, making it more challenging to compare data collected using different versions, especially across time.
+Similar problems arise when non-proprietary tasks are used.
+Most investigators customize even a well-known task to make it suitable for use with children, and the sharing of research materials is not substantially more widespread than the sharing of data.
+Efforts to encourage researchers to capture and record the conceptual structure of psychological tasks have been undertaking (e.g., The Cognitive Atlas), but they have not been widely adopted even by the target community of researchers in adult cognitive neuroscience.
+
+Finally, some have raised concerns that the rise of especially large datasets means the `end of theory.' \cite{Anderson}
+In a provocative essay in 2008, Anderson argued that large quantities of data mean the traditional model of scientific inquiry involving hypothesis testing will soon give way to model-free descriptions of data.
+Others note that bigger data don't necessarily lead to deeper insights \cite{Graham} (http://www.theguardian.com/news/datablog/2012/mar/09/big-data-theory).
+Boyd and Crawford note \cite{boyd} that beyond its scientific dimensions, `big data' is a sociocultural phenomenon that raises as many questions as it promises answers.
 
 \section{Conclusion}
+
+As boyd and Crawford note, `The era of Big Data has begun.
+Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.' \cite{boyd}
+The clamor extends to the developmental and learning sciences where interests in understanding how to maximize the potential for human achievement and improve health have significant consequences for daily living.
+
+This review has shown that the collection, dissemination and analysis of data sets that are big in volume, velocity, or variety has a long and established history in developmental science.
+Many `big data' studies have had substantial impact on scholarship, as measured by the quantity of published findings, and in some cases, on public understanding and policy.
+
+Research and data management practices have begun to converge on norms that will reduce the costs of preparing data for sharing.
+New ethical procedures
+New data archives, increasing support for sharing, both bottom-up and top down.
+
+Data are still hard to access, even for researchers.
+Most data are not in repositories, file drawer effect.
+Most data not prepared or curated for sharing.
+Data hard to understand without help of original investigators.
+No widespread culture of data sharing except in specific subfields, CHIDLES.
+No unified understanding/cultural consensus about who owns data, when to share, etc.
+
+At the same time, while scientific culture may appear to change slowly, let us remember that the Facebook website was launched in 2004, Twitter in 2006.
+It would be unwise to underestimate the speed with which new technologies and tools can change our lives.
+If developmental researchers can find ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development, we can look forward to a future rich in theory and understanding.
 

### 2015-07-28 wiley-big-data-devel/cuts.tex
nearly complete rough draft

A brash new kid has moved in down the street, and she has stolen the limelight from the old guard.
She's a data scientist, and she's teaching the old dogs of statistics some new tricks about how to detect, describe, and visualize patterns in complex sets of data.
Like successful newcomers in many fields, data scientists build on some well-established ideas and tools, but combine and apply them in new ways to new problems \cite{Press2013b}.
And, like many upstarts in recent memory, data scientists have ridden the wave created by the massive growth in the availability of information over the internet \cite{Borgmann2015}.
In this paper, I will focus on how the increased availability of and interest in analyzing large, complex data sets has influenced the study of human development.
I will begin by discussing the recent history of ``big data" and how, in principle, it has implications for developmental science.
Then, I will survey a set of case studies that illustrate how large, complex data sets have been and are being used currently.
To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future.

\subsection{Recent History}

Much of the public attention focused on big data concerns the influence of social media \cite{meyer_everything_2014}, impacts on economic development \cite{HBR2015}, or on data security breaches \cite{Vinton2015}.
he topic has also garnered the attention of government officials \cite{NSFBigData}, and private foundations \cite{ArnoldFdnEvidence, GatesFdn, MooreFdn} with interests in funding scientific research.
In March 2012, the U.S. Obama Administration announced an initiative to spend \$200 M dollars on a ``big data'' research and development (R\&D) initiative \cite{Gianchandani2012, Obama2012} with the expressed intent of applying data science techniques to large collections of digital data in order ``to solve some of the Nation's most pressing challenges..." \cite{Obama2012}.
The press release that accompanied the announcement \cite{Obama2012} highlighted some of the big data initiatives already or soon-to-be underway at the federal level, and these include the U.S. National Science Foundation (NSF), U.S. National Institutes of Health (NIH), the Departments of Energy and Defense and the U.S. Geological Survey (USGS).
The European Union (EU) has also taken public positions \cite{EU2013} about the importance of data to fostering social and economic well-being and has made it a priority to grant more open public access to data sets collected or managed by government entities \cite{EU2015}.
These potential benefits for open access to public data extend to governments in the developing world \cite{Malik2015}, and to state and local governments in the U.S.
For example, Indiana has embarked on an effort to reduce infant mortality using big data \cite{Ravindranath2014}.

Notwithstanding the promises of big data to improving public well-being, governments recognize that privacy and security concerns loom large \cite{Obama2014, Vinton2015} and must be addressed to fully realize the promise of big data science, especially in fields focused on human well-being.
For researchers, the promise of government interest in big data extends beyond potentially new or expanded funding sources \cite{NSFBigData}, such as the NIH's Big Data to Knowledge Initiative \cite{BD2K2015} or the now closed National Children's Study \cite{NCS2015}.
It means potentially new or more widely available or readily accessible data sources in open respositories \cite{databrary.org, dataverse.org, ICSPR, macwhinney_childes_2001} accompanied by more stringent requirements for open data sharing from funders \cite{NIMH2015, NSF2011, GatesFdn} and scientific publication outlets \cite{nosek_promoting_2015}.
At the same time, the collection, storage and open sharing of large data sets about pose significant data management challenges that may require changes in technology use \cite{ProjectJupyter, osf} and scientific practice, including closer collaboration between researchers and librarians and information specialists with expertise in data management, storage and long-term preservation \cite{gordon_researcher-library_2015, lynch_big_2008}.
We will revisit these topics later.

On the other hand, developmental scientists have long recognized the importance of collecting and sharing data sets that are big along at least one dimension -- the number of individuals, number of time points sampled, or the number of measures collected \cite{macwhinney_childes_2001}.
In turn, researchers in developmental science continue to pioneer advanced analytical techniques that unpack and explain developmental phenomena across multiple levels of analysis and across wide-ranging timescales \cite{bakeman_sequential_2011, CDS2014, QuantDev}.
Developmental scientists have also been at the vanguard of efforts to create ways to share research data that preserve privacy and uphold ethical research principles \cite{Adolph2012, AERA2011, AERAVideo2015, Asilomar2014, Databrary2015}.
Accordingly, the developmental research community may be especially well-positioned to turn big data into deep insight.

### 2015-07-29 wiley-big-data-devel/big-data.tex
thru section 2

@@
 That use has grown substantially over the past several decades.
-The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases that has occurred largely in the last 20 years (https://en.wikipedia.org/wiki/List_of_websites_founded_before_1995).
+The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases.
+Most of the tools that enable it have been created in the last 20 years (https://en.wikipedia.org/wiki/List_of_websites_founded_before_1995).
 ``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015; boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.
@@
 Doug Laney \cite{laney01controlling3v} was the first person to describe the dimensions of data streams that make them especially ``big'': Volume, velocity, and variety.
-Data streams that are large in quantity, change rapidly, or consist of many distinct components merit special consideration \cite{ibm_2015}.
+Data streams that are large in quantity, change rapidly, or consist of many components merit special consideration \cite{ibm_2015}.
 Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
@@
 
-Most developmental researchers study individuals or groups of individuals and families.
-But, others focus on neighborhoods or schools, or even broader social, cultural, political and environmental influences.
-This diversity in the targets of analysis reflects an implicit shared belief that developmental trajectories arise from a variety of influences organized at multiple levels of analysis and operating across multiple time scales.
-Consequently, the aspects of `volume' most pertinent to developmental science include the number of participants or families or the number measurement time points.
-We will see that many developmental studies involve tens of participants, but some notable datasets have hundreds or thousands of participants.
-Similarly, the aspects of data velocity most relevant to developmental researchers are those related to the frequency or spacing of measurements.
-In developmental research, velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
-Variety encompasses both the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical.
+The range of phenomena covered by developmental science spans divergent levels of analysis in space, from genes to geography, and time, from microseconds to millenia, all aimed at answering two questions: What develops?
+Why does it develop this way?
+In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998, gottlieb_normally_1998, oyama_ontogeny_2000, vygotsky_mind_1980}.
+For social and behavioral scientists, the goal is to describing change in the *behavior* or *beliefs* of individuals or groups of individuals and families.
+Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
+Similarly, the assessment of biological data -- genes, hormone levels, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status -- inform thinking about how these within-person factors influence changes within individuals' behavior across time.
+
+Consequently, the aspects of data `volume' pertinent to most developmental scientists include the number of participants or families or the number measurement time points.
+We will see that some notable datasets have hundreds or thousands of participants even though samples of tens of participants are more common.
+High volume datasets enable more precise estimation of small effects, especially for qualities or conditions rare among individuals.
+Similarly, the aspects of data `velocity' most relevant to developmental researchers are those related to the frequency or spacing of measurements.
+Velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
+High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
+`Variety' encompasses the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical/evolutionary and the use of multiple measurements to address the same underlying construct.
 To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.
-For example, patterns of brain activity change rapidly, while working memory spans typically do not.
-Brain activity and peripheral autonomic nervous system measurements like heart rate, heart period variability, and galvanic skin responses are related to central nervous system activity in causal, but indirect and complex ways.
 
-Conventional conceptualizations of data focus on text or numeric information, but increasingly images, and audio \cite{Mehl} and video recordings have become part of the mix of data streams collected by researchers and others to better understand human behavior.
-Video in particular captures the complexity and richness of behavior unlike any other measure, and so video provides the main source of information for many researchers who study behavior in laboratory, classroom, or museum contexts.
-Images and recordings are both larger, in file size, and denser than text or flat-file data. 
-Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders \cite{Databrary2015, openfmri.org} and, of course, sharing these materials for non-research entertainment uses has become astonishingly popular \cite{YouTube2015}.
-New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states \cite{Picard}, activity patterns, and momentary cognitive \cite{Sliwinski}, and emotional states broadly available to researchers.
-Moreover, new web-based tools like Amazon's Mechanical Turk (http://www.mturk.com) or Apple's HealthKit (http://www.applehealthkit.com/) will make it possible for research teams to collect data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
-Fundamentally, big data are those that cannot readily be processed by traditional databases or systems that are widely familiar to developmental researchers.
+The collection, management, and analysis of data high in volume, velocity, variety, variability, and complexity poses significant practical challenges for data collection, capture, storage, sharing, analysis, transfer, visualization, and analysis.
+Big data may also magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
+Big data also pose theoretical challenges -- e.g, how do micro-scale factors influence macro-scale phenomena?
+Nevertheless, big data offer the possibility of tackling some of the most profound and vexing questions in the field if the relationships among data components can be revealed in ways that do not undermine ethical research principles.
+Greater openness and more widespread sharing of research data than currently exists will also be essential to realize this scientific promise.
 
-Data may be big in any number of ways and present themselves in structured or unstructured forms, but they pose significant new challenges to traditional avenues for data creation, capture, storage, sharing, analysis, transfer, visualization, and analysis.
-Big data may also magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
+In the next section, I will review the current state of big data in developmental research.
+In doing so, I summarize information about the size of datasets along the relevant dimensions we have discussed, the availability of datasets for reuse or reanalysis by others, and the impact of particular datasets on scholarship.
 
-\subsection{Big data about what and for what purposes?}
+\section{Big datasets in developmental research}
 
-Armed with some definitions about what constitutes big data, we can next ask what topics large datasets can describe and help inform.
-The answer spans the range of phenomena covered by developmental science, from genes to geography and from microseconds to millenia.
-In the case studies to follow, we will highlight studies that focus on biological evidence from genes, brain structure and activity patterns, hormonal variation, nutrition, growth, and health.
-Biological evidence may derive from typically developing or atypically developing populations or both.
-% Add citations to examples %
-Often, behavioral evidence accompanies biological evidence, sometimes featuring standardized task assessments.
-But, there are many examples of big datasets consisting solely of behavioral data.
-These provide evidence about patterns of change and stability across the range of phenomena that interest developmental researchers: perception, action, cognition, language, emotion, sleep, personality, temperament, friendship, family, play, school achievement, substance use, sexuality, and sleep.
-Still other large datasets focus on educational and employment aptitudes and outcomes and the social, economic, political, cultural, and historical contexts in which children and youth come of age.
-% Again find examples and cite %
+% How big, in what ways are big?
+% How open, in what ways open?
+% Size of impact on scholarship?
+% Domains of interest with big datasets?
+% Government vs. individual investigator?
 
-Large datasets can demonstrate either human diversity or human universals, depending on the topic and area of focus.
-In most cases, however, big datasets in developmental science serve to answer core scientific questions that drive research in this area: What develops? 
-Why does it develop that way? 
-Does the environment change or the person or both, and how so?
-Big datasets enable more precise estimation of small effects, for qualities or conditions rare among individuals.
-High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
-Developmental scientists have long recognized the multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998, gottlieb_normally_1998, oyama_ontogeny_2000, vygotsky_mind_1980}.
-Only recently have the tools and resources needed to collect high volume, velocity, and variety needed to make many of these theoretical notions empirically testable.
-Finally, many, perhaps most big datasets in developmental science have as a primary aim the understanding or treatment of disease or disorder.
-In short, big data in developmental science focuses on answering the biggest, most important questions in the field.
-In the next section, we turn to evaluating specific cases of research projects that have generated big datasets and data repositories that preserve, store, and share them with the scientific community.
+Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
+Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities like the U.S.
+Census Bureau and the U.S.
+Department of Education.
+The collection and dissemination of these datasets is mandated by law, subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
+Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
+Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded, usually by government or private funders, to collect large datasets form the next group.
+These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
+% Mention focus on individuals and families?
+The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
+% Mention Federal regulations that guide research with human subjects.
+Finally, there are large scale datasets collected by private entities for business purposes that include sharing data about individual subscribers or users.
+The data collected by private entities about individuals largely concerns consumer behavior, as the data customers are other commercial entities seeking to sell products or services to individual users.
+But, commercial entities are beginning to collect and store health and fitness related activity (Fitbit, etc.) 
+These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities who provide the services.
 
-\section{Case Studies}
-
-\subsection{Biology and Health}
-
-\subsubsection{Epidemiological and Population-based Studies}
+\subsection{Government-Supervised and Managed Datasets}
 
@@
 
+Not all large-scale birth cohort studies succeed.
+A notable example in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
+Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children from before birth to age 21. 
+However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel.
+Authorized by the Children’s Health Act of 2000, the NCS was a planned large-scale, long-term study of U.S.
+children and their mothers and fathers designed to study environmental influences on child health and development.
+Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
+Those data are slated for release in a data archive sometime in 2015.
+
+\subsubsection{Non-U.S. Examples}
+
+Outside the U.S., there are several examples of large-scale longitudinal studies focused on children's growth and health.
+Launched in 2011, the Japan Environment and Children's Study <http://www.env.go.jp/en/chemi/hs/jecs/> is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
+Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
+Health outcomes and exposure measurements will continue until the participating children become 13 years old.
+The study is in its early stages, and has only produced a handful of publications <http://www.env.go.jp/en/chemi/hs/jecs/publications/index.html>.
+Data sharing plans have not been announced.
+World Health Organization (WHO) has offered to help coordinate measurement strategies for large-scale birth cohort studies <http://www.who.int/ceh/cohorts/en/> to ensure that data can be harmonized across them.
+
+Several birth cohort studies have been undertaken in the U.K., beginning with the National Survey of Health & Development in 1946 (https://en.wikipedia.org/wiki/National_Survey_of_Health_%26_Development).
+The National Child Development Study (NCDS) is a continuing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958. 
+The origins of the NCDS can be found in the Perinatal Mortality Survey (PMS) which was then sponsored by the National Birthday Trust Fund and set up to collect information about the social and obstetric factors associated with stillbirth and death in early infancy.[3] Examples of other topics which have since been included are medical care, health, home environment, educational progress, parental involvement, family relationships, economic activity, income, training and housing.[4]
+Conducted by the Centre for Longitudinal Studies (CLS), the aim of the study is to improve understanding of the factors affecting human development over the whole lifespan. 
+It collects information on physical and educational development, economic circumstances, employment, family life, health behaviour, wellbeing, social participation and attitudes.
+physical, educational and social development.[1] During the period 2002-2004, genetic information on participants was also obtained to examine the genetic effects on common traits and diseases.[2]
+The British Cohort Studies have generated more than 3,500 publications (http://www.cls.ioe.ac.uk/Bibliography.aspx?sitesectionid=647&sitesectiontitle=Bibliography) to-date.
+Data are available from (http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies) by registration.
+Another widely known and used dataset comes from the World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) (http://www.who.int/childgrowth/en/).
+The MGRS collected growth and related data from 8500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
+The data provide body mass index, height for children 5-19, and weight data norms for children 5-10.
+
+The OECD-funded PISA dataset is available (http://pisa2000.acer.edu.au/index.php).
+Data from the performance of students in Massive Open Online Courses at Harvard and MIT are available to the research community by application.
+
+\subsubsection{Summary}
+
+Several themes emerge from this sample of population-based studies.
+Datasets are large in terms of the volume of participants sampled ranging in the thousands to tens of thousands.
+The datasets demonstrate significant variety of measures, including survey-based and bio-marker-based (NHANES, Ad Health, SECCYD) and even video (SECCYD).
+The studies show variety geographic composition, as well.
+Some studies involve tens of sample occasions per individual, the NLSY and Ad Health studies in particular.
+Datasets are housed either in U.S. Government (NHANES, SECCYD) or university-based repositories (PSID, Ad Health, NLSY).
+Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center.
+Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
+% Susan McHale's comment about training. Check the other datasets for this.
+% Limits on geographic linkage, other ways data are anomymized to reduce possibility of identifiability.
+
+There are several large datasets with information about employment patterns and educational attainment in children and youth
+The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
+The U.S.
+Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers only under a restricted use agreement.
+
+\subsection{Academic Investigator Initiated and Managed Datasets}
+
 The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} was a survey designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
@@
 The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.
-
-Outside the U.S., there are several examples of large-scale longitudinal studies focused on children's growth and health.
-Launched in 2011, the Japan Environment and Children's Study <http://www.env.go.jp/en/chemi/hs/jecs/> is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
-Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
-Health outcomes and exposure measurements will continue until the participating children become 13 years old.
-The study is in its early stages, and has only produced a handful of publications <http://www.env.go.jp/en/chemi/hs/jecs/publications/index.html>.
-Data sharing plans have not been announced.
-World Health Organization (WHO) has offered to help coordinate measurement strategies for large-scale birth cohort studies <http://www.who.int/ceh/cohorts/en/> to ensure that data can be harmonized across them.
-Several birth cohort studies have been undertaken in the U.K., beginning with the National Survey of Health & Development in 1946 (https://en.wikipedia.org/wiki/National_Survey_of_Health_%26_Development).
-The National Child Development Study (NCDS) is a continuing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958. 
-The origins of the NCDS can be found in the Perinatal Mortality Survey (PMS) which was then sponsored by the National Birthday Trust Fund and set up to collect information about the social and obstetric factors associated with stillbirth and death in early infancy.[3] Examples of other topics which have since been included are medical care, health, home environment, educational progress, parental involvement, family relationships, economic activity, income, training and housing.[4]
-Conducted by the Centre for Longitudinal Studies (CLS), the aim of the study is to improve understanding of the factors affecting human development over the whole lifespan. 
-It collects information on physical and educational development, economic circumstances, employment, family life, health behaviour, wellbeing, social participation and attitudes.
-physical, educational and social development.[1] During the period 2002-2004, genetic information on participants was also obtained to examine the genetic effects on common traits and diseases.[2]
-The British Cohort Studies have generated more than 3,500 publications (http://www.cls.ioe.ac.uk/Bibliography.aspx?sitesectionid=647&sitesectiontitle=Bibliography) to-date.
-Data are available from (http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies) by registration.
-Another widely known and used dataset comes from the World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) (http://www.who.int/childgrowth/en/).
-The MGRS collected growth and related data from 8500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
-The data provide body mass index, height for children 5-19, and weight data norms for children 5-10.
-
-Not all large-scale birth cohort studies succeed.
-A notable example in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
-Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children from before birth to age 21. 
-However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel.
-Authorized by the Children’s Health Act of 2000, the NCS was a planned large-scale, long-term study of U.S.
-children and their mothers and fathers designed to study environmental influences on child health and development.
-Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
-Those data are slated for release in a data archive sometime in 2015.
-
-Several themes emerge from this sample of population-based studies.
-Datasets are large in terms of the volume of participants sampled ranging in the thousands to tens of thousands.
-The datasets demonstrate significant variety of measures, including survey-based and bio-marker-based (NHANES, Ad Health, SECCYD) and even video (SECCYD).
-The studies show variety geographic composition, as well.
-Some studies involve tens of sample occasions per individual, the NLSY and Ad Health studies in particular.
-Datasets are housed either in U.S. Government (NHANES, SECCYD) or university-based repositories (PSID, Ad Health, NLSY).
-Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center.
-Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
-% Susan McHale's comment about training. Check the other datasets for this.
-% Limits on geographic linkage, other ways data are anomymized to reduce possibility of identifiability.
 
@@
 The data are restricted, and thus are only available by specific authorization.
+
 The Family Life Project also collected and reported WPPSI data.
+
+\subsubsection{Language}
 
@@
 
+\subsubsection{Vulnerable Populations}
+
 %- Infant Behavior Questionnaire, CBC
 While no archive for IBQ data across studies exists, IBQ data on 1,388 participants (n = 658) who were group matched within site with a group of non-exposed comparison infants (n=730) are archived at ICSPR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34312?sortBy=19&amp;q=infant+behavior+questionnaire&amp;searchSource=revise) as part of the Maternal Lifestyle Study in Four Sites in the United States, 1993-2011.
+
 MSL was a large NIH longitudinal studies of children with prenatal cocaine exposure (PCE).
@@
 IBQ data are also available from ICPSR from the Family Life Project (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=19&amp;q=ibq&amp;searchSource=revise).
-
-\subsection{Education and Employment}
-
-There are several large datasets with information about employment patterns and educational attainment in children and youth
-The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
-The U.S.
-Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers only under a restricted use agreement.
-The College Board shares SAT and college cost and scholarship data with the research community by application (http://www.collegeboard.com/inquiry/csinquiry_form.html).
-The ACT (http://www.act.org/research/). Privacy policies (http://www.act.org/privacy.html). 
-The Trends in International Math and Science Study (TIMSS) studied mathematics and science achievement in 4th and 8th grade for more than 20 years.
-Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
-Both studies provide data available for public download (http://timss.bc.edu/).
-The 1999 TIMSS video study examined teaching practices in seven countries, and some of the materials are available for public use (http://www.timssvideo.com/).
-The Measures of Effective Teaching (MET, http://www.metproject.org/) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
-The MET video data are stored and shared at ICSPR.
-The OECD-funded PISA dataset is available (http://pisa2000.acer.edu.au/index.php).
-Data from the performance of students in Massive Open Online Courses at Harvard and MIT are available to the research community by application.
-
-The 
-LearnLab!
-
-          %      * Attitudes
-   %          - Judith Torney-Purta, <http://www.nie.edu.sg/research-publications/cieclopediaorg/cieclopediaorg-a-to-z-listing/Judith-Torney-Purta>
-
-\subsection{Social, Economic, and Cultural Contexts of Development}
 
@@
 
-       %  + Normative vs. perturbed development
-       %      * Bucharest Early Intervention Project (BEIP), <http://www.childrenshospital.org/beip>
-       %      * NPR story, <http://www.npr.org/blogs/health/2014/02/20/280237833/orphans-lonely-beginnings-reveal-how-parents-shape-a-childs-brain>
+* Bucharest Early Intervention Project (BEIP), <http://www.childrenshospital.org/beip>
+* NPR story, <http://www.npr.org/blogs/health/2014/02/20/280237833/orphans-lonely-beginnings-reveal-how-parents-shape-a-childs-brain>
+
+\subsubsection{Education}
+
+%LearnLab, Edith Gumm's new gig
+The Trends in International Math and Science Study (TIMSS) studied mathematics and science achievement in 4th and 8th grade for more than 20 years.
+Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
+Both studies provide data available for public download (http://timss.bc.edu/).
+The 1999 TIMSS video study examined teaching practices in seven countries, and some of the materials are available for public use (http://www.timssvideo.com/).
+The Measures of Effective Teaching (MET, http://www.metproject.org/) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
+The MET video data are stored and shared at ICSPR.
+
+\subsection{Commercially Initiated and Managed Datasets}
+
+The College Board shares SAT and college cost and scholarship data with the research community by application (http://www.collegeboard.com/inquiry/csinquiry_form.html).
+The ACT (http://www.act.org/research/). Privacy policies (http://www.act.org/privacy.html).
+
+Google, Twitter, Amazon, Facebook, SnapChat.
 
@@
 
+Conventional conceptualizations of data focus on text or numeric information, but increasingly images, and audio \cite{Mehl} and video recordings have become part of the mix of data streams collected by researchers and others to better understand human behavior.
+Video in particular captures the complexity and richness of behavior unlike any other measure, and so video provides the main source of information for many researchers who study behavior in laboratory, classroom, or museum contexts.
+Images and recordings are both larger, in file size, and denser than text or flat-file data. 
+Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders \cite{Databrary2015, openfmri.org} and, of course, sharing these materials for non-research entertainment uses has become astonishingly popular \cite{YouTube2015}.
+New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states \cite{Picard}, activity patterns, and momentary cognitive \cite{Sliwinski}, and emotional states broadly available to researchers.
+Moreover, new web-based tools like Amazon's Mechanical Turk (http://www.mturk.com) or Apple's HealthKit (http://www.applehealthkit.com/) will make it possible for research teams to collect data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
+Fundamentally, big data are those that cannot readily be processed by traditional databases or systems that are widely familiar to developmental researchers.
+
 \subsubsection{Collecting data with new technologies, data streams}
@@
 For some some data elements, the devices used to make measurements already provide data and metadata in a structured, organized way.
-For example, genetic analyses from modern gene sequencing tools provide machine-readable outputs that (Rick talk to David Vandenburgh).
+For example, genetic analyses from modern gene sequencing tools provide machine-readable outputs.
 Researchers who analyze hormone levels in saliva or micronutrients in blood, for example, (Rick asking Sam Scott, others).

### 2015-07-29 wiley-big-data-devel/cuts.tex
thru section 2

@@
 Accordingly, the developmental research community may be especially well-positioned to turn big data into deep insight.
+
+\subsection{Big data about what and for what purposes?}
+
+Armed with some definitions about what constitutes big data, we can next ask what topics large datasets can describe and help inform.
+The answer spans the range of phenomena covered by developmental science, from genes to geography and from microseconds to millenia.
+In the case studies to follow, we will highlight studies that focus on biological evidence from genes, brain structure and activity patterns, hormonal variation, nutrition, growth, and health.
+Biological evidence may derive from typically developing or atypically developing populations or both.
+% Add citations to examples %
+Often, behavioral evidence accompanies biological evidence, sometimes featuring standardized task assessments.
+But, there are many examples of big datasets consisting solely of behavioral data.
+These provide evidence about patterns of change and stability across the range of phenomena that interest developmental researchers: perception, action, cognition, language, emotion, sleep, personality, temperament, friendship, family, play, school achievement, substance use, sexuality, and sleep.
+Still other large datasets focus on educational and employment aptitudes and outcomes and the social, economic, political, cultural, and historical contexts in which children and youth come of age.
+% Again find examples and cite %
+
+Large datasets can demonstrate either human diversity or human universals, depending on the topic and area of focus.
+In most cases, however, big datasets in developmental science serve to answer core scientific questions that drive research in this area: What develops? 
+Why does it develop that way? 
+Does the environment change or the person or both, and how so?
+High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
+Finally, many, perhaps most big datasets in developmental science have as a primary aim the understanding or treatment of disease or disorder.
+In short, big data in developmental science focuses on answering the biggest, most important questions in the field.
+In the next section, we turn to evaluating specific cases of research projects that have generated big datasets and data repositories that preserve, store, and share them with the scientific community.
+
+This diversity in the targets of analysis reflects an implicit shared belief that developmental trajectories arise from a variety of influences organized at multiple levels of analysis operating across multiple time scales.
+
+For example, patterns of brain activity change rapidly, while working memory spans and genotypes typically do not.
+Further, brain activity and peripheral autonomic nervous system measurements like heart rate, heart period variability, and galvanic skin responses are related to central nervous system activity in causal, but indirect and complex ways.

### 2015-07-30 wiley-big-data-devel/big-data.tex
thru govt run section

@@
 
-\subsection{Government-Supervised and Managed Datasets}
+\subsection{Government-collected and Managed Datasets}
 
-Some of the largest and most widely used datasets in developmental science are population-based, often with an epidemiological dimension.
-Several notable examples began data collection in the 1950's or 60's.
+Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
 The National Health and Nutrition Examination Survey (NHANES) \cite{NHANES} began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S.
@@
 The annual survey combines interviews about demographic, socioeconomic, dietary and health related topics and physical examinations involving medical, dental, and physiological measurements, as well as laboratory tests.
-The study assesses a nationally representative sample of about 5,000 persons located across the country, but 15 counties are surveyed annually.
-Many will be familiar with some of the study's products such as standardized child growth charts (http://www.cdc.gov/growthcharts/cdc_charts.htm), or related initiatives such as the National Youth Fitness Survey (NNYFS) (http://www.cdc.gov/nchs/nnyfs.htm which collected data on physical activity and fitness levels in order to provide an evaluation of the health and fitness of children in the U.S.
-ages 3 to 15.
-The NNYFS collected data on physical activity and fitness levels of our youth through interviews and fitness tests.
+The study is cross-sectional and assesses a nationally representative sample of about 5,000 persons located across the country, with 15 counties surveyed annually.
+Many will be familiar with some of NHANES's products such as standardized child growth charts (http://www.cdc.gov/growthcharts/cdc_charts.htm).
+The National Youth Fitness Survey \cite{NYFS} is closely associated with NHANES.
+NYFS collected data on physical activity and fitness levels in U.S. children ages 3 to 15 in 2012 in order to provide an evaluation their overall fitness.
 Much of the NHANES and NNYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
-However, access to some data are restricted, available only to individuals via a Research Data Center (RDC) that can provide access.
-<http://wwwn.cdc.gov/nchs/nhanes/search/variablelist.aspx?CycleBeginYear=1999&Component=Non-Public>.
-There are 19 Federal Statistical Research Data Center (RDC) locations across the U.S.
+However, access to some data are restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers (RDC) locations across the U.S.
 A Google Scholar search shows more than 110,000 publications mention the NHANES dataset.
-The Panel Study on Income Dynamics (PSID) \cite{PSID} is another large-scale population based effort.
-PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
-The study consists of survey responses to questions about employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, education, and other topics.
-As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}.
-Much of data is available publicly to users who register (http://simba.isr.umich.edu/U/ca.aspx) subject to the provisions of a formal set of use conditions (http://simba.isr.umich.edu/U/CondUse.aspx) that focus largely on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
-In addition, some restricted data elements are only available subject to a formal restricted use contract (http://simba.isr.umich.edu/restricted/RestrictedUse.aspx) between the investigator and the data owner, the University of Michigan.
 
-In contrast the National Longitudinal Survey of Youth (NLSY) \cite{NLSY} has studied a large subset of the same cohort since the study began in 1979.
-The project follows the lives of a sample of almost 10,000 American youth born between 1957-64.
-Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY ata are now available for up to 25 survey rounds on questions relating to employment and related questions.
+The National Longitudinal Survey of Youth (NLSY) \cite{NLSY} provides an example of a large-scale study that has examined the same cohort -- almost 10,000 American youth born between 1957-64 -- on a annual basis since the study's beginning in 1979.
+Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY data are now available for up to 25 survey rounds on questions relating to employment and related issues.
 The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52 through the National Longitudinal Survey (NLS).
-Most NLSY and NLS data are available free of charge to by means of a web-based data portal \cite{NLSY_data}, access to which users must register for.
-Users must apply for special permission to access geographic-related variables.
+Most NLSY and NLS data are available free of charge to by means of a web-based data portal \cite{NLSY_data} to which users must apply for access.
+Access to geographic-related variables requires special permission.
 The Center for Human Resource Research (CHRR) at The Ohio State University and the National Opinion Research Center (NORC) at the University of Chicago administer both the data collection and repository functions.
-A Google Scholar search in mid-2015 shows more than 20,000 publications cite the NLSY.
+A Google Scholar search in mid-2015 shows that more than 20,000 publications cite the NLSY.
 
@@
 The project has received funding from the National Center for Educational Statistics (https://nces.ed.gov/ecls/), and the data are archived at ICPSR.
-ECLS consists of both public and restricted data, and according to a Google Scholar search has generated more than 8,000 publications, but ICSPR says 167.
+ECLS consists of both publicly available and restricted data.
+A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications (https://www.icpsr.umich.edu/icpsrweb/ICPSR/biblio/series/00191/resources?sortBy=1).
+
+There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
+The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
+The U.S.
+Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
+
+There are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
+The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) \cite{MGRS} collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
+MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
+Launched in 2011, the Japan Environment and Children's Study \cite{JECS} is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
+Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
+Health outcomes and exposure measurements will continue until the participating children become 13 years old.
+The study is in its early stages.
+The study has produced a handful of publications \cite{JECS_pubs}, and no data sharing plans have been announced.
+Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health & Development \cite{NSHD_UK} and the National Child Development Study (NCDS) \cite{NCDS_UK}.
+NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
+Originally focused on improving infant mortality, the study has expanded to focus on factors predicting wellbeing across the lifespan.
+Measures include genetic, health, education, social, and economic factors.
+The study is funded by the British government and data are available by registration from the U.K. Data Service \cite{UK_cohort_data}.
+The British Cohort Studies have generated more than 3,500 publications to-date \cite{UK_cohort_citations}.
+Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset \cite{PISA}, which generates more than 5,000 search hits.
 
 Not all large-scale birth cohort studies succeed.
-A notable example in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
-Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children from before birth to age 21. 
+A notable failure in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
+Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children and their parents from before birth to age 21. 
 However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel. 
-Authorized by the Children’s Health Act of 2000, the NCS was a planned large-scale, long-term study of U.S.
-children and their mothers and fathers designed to study environmental influences on child health and development.
 Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
@@
 
-\subsubsection{Non-U.S. Examples}
-
-Outside the U.S., there are several examples of large-scale longitudinal studies focused on children's growth and health.
-Launched in 2011, the Japan Environment and Children's Study <http://www.env.go.jp/en/chemi/hs/jecs/> is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
-Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
-Health outcomes and exposure measurements will continue until the participating children become 13 years old.
-The study is in its early stages, and has only produced a handful of publications <http://www.env.go.jp/en/chemi/hs/jecs/publications/index.html>.
-Data sharing plans have not been announced.
-World Health Organization (WHO) has offered to help coordinate measurement strategies for large-scale birth cohort studies <http://www.who.int/ceh/cohorts/en/> to ensure that data can be harmonized across them.
-
-Several birth cohort studies have been undertaken in the U.K., beginning with the National Survey of Health & Development in 1946 (https://en.wikipedia.org/wiki/National_Survey_of_Health_%26_Development).
-The National Child Development Study (NCDS) is a continuing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958. 
-The origins of the NCDS can be found in the Perinatal Mortality Survey (PMS) which was then sponsored by the National Birthday Trust Fund and set up to collect information about the social and obstetric factors associated with stillbirth and death in early infancy.[3] Examples of other topics which have since been included are medical care, health, home environment, educational progress, parental involvement, family relationships, economic activity, income, training and housing.[4]
-Conducted by the Centre for Longitudinal Studies (CLS), the aim of the study is to improve understanding of the factors affecting human development over the whole lifespan. 
-It collects information on physical and educational development, economic circumstances, employment, family life, health behaviour, wellbeing, social participation and attitudes.
-physical, educational and social development.[1] During the period 2002-2004, genetic information on participants was also obtained to examine the genetic effects on common traits and diseases.[2]
-The British Cohort Studies have generated more than 3,500 publications (http://www.cls.ioe.ac.uk/Bibliography.aspx?sitesectionid=647&sitesectiontitle=Bibliography) to-date.
-Data are available from (http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies) by registration.
-Another widely known and used dataset comes from the World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) (http://www.who.int/childgrowth/en/).
-The MGRS collected growth and related data from 8500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
-The data provide body mass index, height for children 5-19, and weight data norms for children 5-10.
-
-The OECD-funded PISA dataset is available (http://pisa2000.acer.edu.au/index.php).
-Data from the performance of students in Massive Open Online Courses at Harvard and MIT are available to the research community by application.
-
 \subsubsection{Summary}
 
-Several themes emerge from this sample of population-based studies.
-Datasets are large in terms of the volume of participants sampled ranging in the thousands to tens of thousands.
-The datasets demonstrate significant variety of measures, including survey-based and bio-marker-based (NHANES, Ad Health, SECCYD) and even video (SECCYD).
-The studies show variety geographic composition, as well.
-Some studies involve tens of sample occasions per individual, the NLSY and Ad Health studies in particular.
-Datasets are housed either in U.S. Government (NHANES, SECCYD) or university-based repositories (PSID, Ad Health, NLSY).
+While no means exhaustive in its coverage, several themes emerge from this sample of U.S. and international population-based studies.
+Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
+Some studies involve tens of sample occasions per individual over extended periods of time, and show significant variety in the measures collected.
 Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center.
 Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
-% Susan McHale's comment about training. Check the other datasets for this.
-% Limits on geographic linkage, other ways data are anomymized to reduce possibility of identifiability.
-
-There are several large datasets with information about employment patterns and educational attainment in children and youth
-The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
-The U.S.
-Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers only under a restricted use agreement.
+We turn next to datasets whose collection and management is under the control of individual, usually academic investigators.
 
 \subsection{Academic Investigator Initiated and Managed Datasets}
+
+The number and variety of topics studied
+
+The Panel Study on Income Dynamics (PSID) \cite{PSID} is another large-scale population based effort.
+PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
+The study consists of survey responses to questions about employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, education, and other topics.
+As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}.
+Much of data is available publicly to users who register (http://simba.isr.umich.edu/U/ca.aspx) subject to the provisions of a formal set of use conditions (http://simba.isr.umich.edu/U/CondUse.aspx) that focus largely on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
+In addition, some restricted data elements are only available subject to a formal restricted use contract (http://simba.isr.umich.edu/restricted/RestrictedUse.aspx) between the investigator and the data owner, the University of Michigan.
 
@@
 
+The purpose of this project is to study the early development of a group of children who are at risk regarding later successful adjustment and for whom we have little information: children living in rural, largely poor communities.
+A birth cohort of 800 children in three rural counties of North Carolina and 600 children in three rural counties of Pennsylvania were studied over the first 3 years of life.
+https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=&amp;q=Bayley&amp;searchSource=revise
+
 \subsubsection{Behavior Genetics}
 
-The next area of developmental science we explore is behavior genetics, where large data samples have long been required in order to detect what can often be small effect sizes.
+Large data samples have long been required in behavior genetics order to detect what can often be small effect sizes.
 Several large-scale genetics studies that are not specifically developmental deserve mention.
@@
 
-The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} focuses on genetic and environmental contributions to measures of family relationships and mental health.
+The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} focuses on genetic and environmental ontributions to measures of family relationships and mental health.
 TOSS extends the Twin Moms Project to include a sample of twin fathers (N=320 twin pairs and their families).
@@
 
-Several large studies have focused on developmental patterns in brain structure or activity.
+Several large-scale studies have focused on developmental patterns in brain structure or activity.
 For practical reasons -- motion artifacts make imaging very young children extremely challenging -- most projects have focused on children and adolescents.
@@
 The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data appear not to be available for analyses by researchers outside the original investigative team.
+
 The NIH MRI Study of of Normal Brain Development \cite{NIH_Pediatric_MRI} collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
@@
 These data can be openly shared with appropriate citation.
+
 More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING) Project \cite{PING} collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from Baltimore, Boston, Honolulu, Los Angeles, New Haven, New York, Sacramento, and San Diego.
@@
 As of mid-2015, the project had generated more than 20 publications.
+
 The Developmental Connectome Project \cite{developmental_connectome}, led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
@@
 CHILDES/TalkBank has generated more than ten thousand of citations, according to Google Scholar, and the data are available to the research community, some of them publicly.
+
 The Human Speechome Project (http://www.media.mit.edu/cogmac/projects/hsp.html) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
@@
 Data was gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
+
 % edit previous
@@
 Catherine Tamis-Lemonda has released video data (https://nyu.databrary.org/volume/8) from a longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
+
 The LENA Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed (http://www.lenafoundation.org/customer-resources/videos).

### 2015-08-03 wiley-big-data-devel/big-data.tex
nearly complete draft

@@
 Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
+The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years \cite{TIMSS}.
+TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commerical partners.
+The 2001 TIMSS assessment involved more than 20,000 U.S. students from more than 1,000 schools plus data from more than 60 countries.
+The next TIMSS data collection is in 2015 and will include students in grades 4, 8 and 12.
+Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
+Both studies provide data available for public download.
 
-There are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
+In addition to TIMSS, there are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
 The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) \cite{MGRS} collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
@@
 
-The number and variety of topics studied
+Investigator-managed big data projects cover a wide range of topics and and even wider range of research methodologies.
+In several cases, the studies involve partnerships between government agencies and academic investigators.
+The Panel Study on Income Dynamics (PSID) \cite{PSID} is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
+PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
+As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}. 
+Much of data is available publicly to registered users and agree to the terms of use conditions \cite{PSID_term} that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
+In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement \cite{PSID_restricted_use} between the investigator and the data owner, the University of Michigan.
 
-The Panel Study on Income Dynamics (PSID) \cite{PSID} is another large-scale population based effort.
-PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
-The study consists of survey responses to questions about employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, education, and other topics.
-As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}.
-Much of data is available publicly to users who register (http://simba.isr.umich.edu/U/ca.aspx) subject to the provisions of a formal set of use conditions (http://simba.isr.umich.edu/U/CondUse.aspx) that focus largely on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
-In addition, some restricted data elements are only available subject to a formal restricted use contract (http://simba.isr.umich.edu/restricted/RestrictedUse.aspx) between the investigator and the data owner, the University of Michigan.
-
-The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} was a survey designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
+The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} began as a survey study designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
 Undertaken in response to a Congressional mandate, the study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S.
@@
 The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
-The Ad Health cohort has been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32. 
-In addition to survey data Add Health includes biomarker data from a blood test taken during Wave IV. 
-Data are available about glucose, hemoglobin (HbA1c) levels, and blood lipids. 
-The Add Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
+The Ad Health cohort has subsequently been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32.
+During Wave IV, biomarker data from a blood test has made available data about glucose, hemoglobin (HbA1c) levels, blood lipids, and related measures. 
+The Ad Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
 As of mid-2015, more than 5,000 publications \cite{add_health_publications} had arisen from the Add Health dataset.
 
-In comparison, the NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
-The goal was to learn more about the kind and quality of child care experiences, and the effects of different child care settings on developmental outcomes. 
+The NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
+The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
 Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, and anthropometric measures, and demographic information.
 The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
-Unlike the Ad Health study, data in SECCYD are not available to the public, but are restricted.
-Access is granted only by application, and approval is granted on a case-by-case basis.
+Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, and approval is granted on a case-by-case basis.
+Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
 The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.
 
-The purpose of this project is to study the early development of a group of children who are at risk regarding later successful adjustment and for whom we have little information: children living in rural, largely poor communities.
-A birth cohort of 800 children in three rural counties of North Carolina and 600 children in three rural counties of Pennsylvania were studied over the first 3 years of life.
-https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=&amp;q=Bayley&amp;searchSource=revise
+The Family Life Project (FLP) studied the early development children living in rural, largely poor communities in North Carolina and Pennsylvania. 
+A birth cohort of 800 children were studied over the first 3 years of life in a series of home, child care visits and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
+Data collected included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
+These features make the FLP large in volume, velocity, and variety.
+Data from Phase I of the FLP are archived at ICPSR \cite{Family_Life_data}, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
+The FPL site lists about 50 publications through 2013.
 
@@
 The Genome of the Netherlands Project(http://www.bbmri.nl/en-gb/activities/rainbow-projects/131-genome-of-the-netherlands) is a publicly available dataset from n=250 families, consisting of two parents plus one of their adult children \cite{Genome_Netherlands}.
-The Genes, Environment, and Melanoma Study (GEMS) focuses on risk for melanoma in a sample of 3,700 individuals from the North America, Europe, and Australia (http://gemstudy.org/main/index.html).
-The Psychiatric Genomics Consortium (PGC) (http://www.med.unc.edu/pgc) contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
+The Genes, Environment, and Melanoma Study (GEMS) focuses on risk for melanoma in a sample of 3,700 individuals from the North America, Europe, and Australia \cite{GEMS}.
+The Psychiatric Genomics Consortium (PGC) \cite{PGC} contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
 Many disorders featured (autism, attention-deficit hyperactivity disorder, and schizophrenia) in the PGC dataset have developmental dimensions.
@@
 
-The U.K.-based Twins Early Development Study (TEDS) \cite{TEDS} (https://en.wikipedia.org/wiki/Twins_Early_Development_Study) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
+The U.K.-based Twins Early Development Study (TEDS) \cite{TEDS} contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
 Some 13,000 pairs remain actively involved, providing lab-based behavioral task and survey responses.
@@
 
-The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} focuses on genetic and environmental ontributions to measures of family relationships and mental health.
+The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} is another twin-focused study aimed at uncovering genetic and environmental ontributions to measures of family relationships and mental health.
 TOSS extends the Twin Moms Project to include a sample of twin fathers (N=320 twin pairs and their families).
@@
 
+Adoption studies provide a unique window on the mix of genetic and non-genetic factors that influence child development.
 Illustrative U.S. examples include the Colorado Adoption Project (CAP), the Nonshared Environment in Adolescent Development (NEAD) project, and the Early Growth and Development Study (EGADS).
+Since its inception in 1976, the Colorado Adoption Project (CAP) \cite{CAP} has enrolled more than 2,400 participants from more than 450 families.
+Data include cognitive abilities, temperament measures, and demographic variables.
+CAP has generated more than 200 published articles \cite{CAP_pubs}.
+Data collection continues.
+The study aims to be the definitive genetically informed longitudinal study of lifespan development \cite{rhea_colorado_2013}.
+Data from 1976-1989 are archived at Dataverse (https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/00913) and can be accessed by application to Dataverse.
 NEAD studied a sample of 720 families, including twins, siblings, and their parents, in a longitudinal design that involved up to 3 data collection waves.
@@
 The team plans to share the data openly later in 2015.
-
-Adoption studies provide a unique window on the mix of genetic and non-genetic factors that influence child development.
-Since its inception in 1976, the Colorado Adoption Project (CAP) (http://ibg.colorado.edu/cap/index.html) has enrolled more than 2,400 participants from more than 450 families.
-Data include cognitive abilities, temperament measures, demographic variables, 
-Additionally parents completed a questionnaire assessment which included measures of temperament
-CAP has generated more than 200 published articles (http://ibg.colorado.edu/cap/publishedarticles.html).
-Data collection continues.
-The study aims to be the definitive genetically informed longitudinal study of lifespan development \cite{rhea_colorado_2013}.
-Data from 1976-1989 are archived at Dataverse (https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/00913) and can be accessed by application to Dataverse.
 EGADS is a prospective adoption study of (n=561) triads of birth parents, adoptive parents, and adopted children in the U.S. 
@@
 Several large-scale studies have focused on developmental patterns in brain structure or activity.
-For practical reasons -- motion artifacts make imaging very young children extremely challenging -- most projects have focused on children and adolescents.
 One of the earliest studies of its kind was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
@@
 
-\subsection{Behavior}
+\subsubsection{Language, Cognition, Temperament}
 
-In the realm of behavior, there are a number of measures that have become standards whose widespread adoption has resulted in data that are large in volume, velocity or variety.
-Davida Teller pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
-A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.
-The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} arose from the Nancy Bayley's Berkeley Growth Study \cite{jones_berkeley_1941} that studied the behavior, motor coordination, physical activity, cognitive function, and maternal behavior of 61 healthy newborn infants.
-The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
-The Bayley is widely used in clinical and epidemiological settings such as the EPICure <http://www.epicure.ac.uk/> because of the norms.
-However, data about those norms appears to be under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
-Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE), 1972-1992 project on ICPSR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4091?sortBy=19&amp;q=Bayley&amp;searchSource=revise).
-The data are freely available to registered users of ICSPR.
-
-The Wechsler intelligence tests were developed by Dr.
-David Wechsler, beginning in the 1930's.
-The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, while the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
-The current version (WISC V) is provided by Pearson (http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html).
-Data about the norms themselves appears not to be available to researchers.
-However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR (https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/13604?q=wisc&amp;searchSource=find-analyze-home&amp;sortBy=).
-The data are restricted, and thus are only available by specific authorization.
-
-The Family Life Project also collected and reported WPPSI data.
-
-\subsubsection{Language}
-
-There are several large-scale datasets associated with language processing.
+The child language community has pioneered the aggregation and sharing of datasets.
 The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established.
 It consists of transcripts and audio and video recordings of children's utterances along with recent data from adults with aphasia.
-CHILDES/TalkBank has generated more than ten thousand of citations, according to Google Scholar, and the data are available to the research community, some of them publicly.
+CHILDES/TalkBank has generated more than ten thousand of citations, according to Google Scholar, and the data are available to the research community, many of them publicly.
 
+Other large-scale sources of language-related data are more recent.
+WordBank (http://wordbank.stanford.edu/) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
+Catherine Tamis-Lemonda has released video data (https://nyu.databrary.org/volume/8) from an NSF-funded longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
 The Human Speechome Project (http://www.media.mit.edu/cogmac/projects/hsp.html) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
@@
 Data was gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
-
-% edit previous
-WordBank (http://wordbank.stanford.edu/) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
-Catherine Tamis-Lemonda has released video data (https://nyu.databrary.org/volume/8) from a longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
-
-The LENA Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed (http://www.lenafoundation.org/customer-resources/videos).
+The LENA Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed \cite{LENA}.
 LENA stands for Language ENvironment Analysis.
@@
 The system was designed and is distributed by the LENA Foundation.
+No large-scale archive for LENA data currently exists, but some researchers who use the tool have begun to explore the creation of a data repository specialized for these sorts of data.
 
-\subsubsection{Vulnerable Populations}
+In other areas of cognition, a number of measures that have become standards whose widespread adoption has resulted in data that are large in volume, velocity or variety.
+Davida Teller pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
+A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.
 
-%- Infant Behavior Questionnaire, CBC
-While no archive for IBQ data across studies exists, IBQ data on 1,388 participants (n = 658) who were group matched within site with a group of non-exposed comparison infants (n=730) are archived at ICSPR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34312?sortBy=19&amp;q=infant+behavior+questionnaire&amp;searchSource=revise) as part of the Maternal Lifestyle Study in Four Sites in the United States, 1993-2011.
+The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} arose from the Nancy Bayley's Berkeley Growth Study \cite{jones_berkeley_1941} that studied the behavior, motor coordination, physical activity, cognitive function, and maternal behavior of 61 healthy newborn infants.
+The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
+The Bayley is widely used in clinical and epidemiological settings such as the EPICure \cite{EPICure} because of the norms even though there are questions about its validity \cite{hack_poor_2005}.
+However, data about those norms appears to be under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
+Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE) project on ICPSR \cite{CARE}.
+The data are freely available to registered users of ICSPR.
 
-MSL was a large NIH longitudinal studies of children with prenatal cocaine exposure (PCE).
-MLS was a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
-IBQ data are also available from ICPSR from the Family Life Project (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=19&amp;q=ibq&amp;searchSource=revise).
+The Wechsler intelligence tests were developed by Dr. David Wechsler, beginning in the 1930's. 
+The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, while the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
+The current version (WISC V) is provided by Pearson \cite{WISC_V}.
+Data about the norms themselves appears not to be available to researchers.
+However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR \cite{PHDCN}. 
+The data are restricted, and thus are only available by specific authorization.
 
-%  * [Family Life Project](http://flp.fpg.unc.edu/)
-
-The purpose of this project is to study the early development of a group of children who are at risk regarding later successful adjustment and for whom we have little information: children living in rural, largely poor communities.
-A birth cohort of 800 children in three rural counties of North Carolina and 600 children in three rural counties of Pennsylvania were studied over the first 3 years of life.
-https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=&amp;q=Bayley&amp;searchSource=revise
-
-MacArthur Network on Transitions to Adulthood (http://www.transitions2adulthood.com/). Data not obviously open or available.
-Research Network on Adolescent Development & Juvenile Justice - See more at: https://www.macfound.org/networks/research-network-on-adolescent-development-juvenil/#sthash.PQOwqv14.dpuf
-Research Network on Youth and Participatory Politics - See more at: https://www.macfound.org/networks/research-network-on-youth-and-participatory-politi/#sthash.egU1Wwvy.dpuf
-
-* Bucharest Early Intervention Project (BEIP), <http://www.childrenshospital.org/beip>
-* NPR story, <http://www.npr.org/blogs/health/2014/02/20/280237833/orphans-lonely-beginnings-reveal-how-parents-shape-a-childs-brain>
+In the realm of emotion and temperament, two academic-investigator-initiated measures have been widely adopted.
+The Infant Behavior Questionnaire (IBQ), a parent survey measure, and the Laboratory Temperament Assessment Battery (Lab-TAB) \cite{Lab_TAB} an observational battery, using a set of 3-5 min episodes mimicking situations in everyday life.
+While no archive for IBQ or Lab-TAB data across studies exists, IBQ data on 1,388 participants are archived at ICSPR as part of the Maternal Lifestyle Study (MLS) \cite{MSL}, a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
+IBQ and Lab-TAB data are also available from ICPSR from the Family Life Project.
 
@@
 
-%LearnLab, Edith Gumm's new gig
-The Trends in International Math and Science Study (TIMSS) studied mathematics and science achievement in 4th and 8th grade for more than 20 years.
-Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
-Both studies provide data available for public download (http://timss.bc.edu/).
-The 1999 TIMSS video study examined teaching practices in seven countries, and some of the materials are available for public use (http://www.timssvideo.com/).
-The Measures of Effective Teaching (MET, http://www.metproject.org/) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
+There are several investigator initiatiated or managed education studies that deserve mention because of their volume, velocity, or variety or because of their availability for reuse by others.
+The 1999 TIMSS video study recorded teaching practices in seven countries, and some of the materials are available for public use \cite{TIMSS_video}.
+The TIMSS video study repository is hosted at UCLA with support from the Carnegie Foundation for the Advancement of Teaching, The William and Flora Hewlett Foundation and the Gabriella and Paul Rosenbaum Foundation.
+The Gates Foundation-funded Measures of Effective Teaching (MET) \cite{METProject} Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
 The MET video data are stored and shared at ICSPR.
+The Spatial Intelligence and Learning Center (SILC) \cite{SILC} is an NSF-funded initiative designed to ``develop the science of spatial learning and to use this knowledge to transform educational practice, supporting children and adults in acquiring the scientific, technical, engineering, and mathematical (STEM) skills required for effective participation in an increasingly high-technology society and global economy.''
+The SILC website hosts links to a set of test and instruments useful in research on spatial learning.
+LearnLab \cite{LearnLab}, another NSF-funded center, combines cognitive theory and computational modeling to understand changes in student knowledge in the context of computer-based math, science and language courses.
+LearnLab hosts a repository, DataShop \cite{DataShop}, to store and openly share detailed data from classrooms employing cognitive tutors.
+DataShop enables data storage, management, visualization and analysis in a web-based tool.
 
-\subsection{Commercially Initiated and Managed Datasets}
+\subsubsection{Summary}
 
-The College Board shares SAT and college cost and scholarship data with the research community by application (http://www.collegeboard.com/inquiry/csinquiry_form.html).
-The ACT (http://www.act.org/research/). Privacy policies (http://www.act.org/privacy.html).
+Reviewing the set of investigator initiated or managed datasets, we see examples of datasets large in volume, velocity, and variety.
+Measures employed are also diverse, from surveys, to observational measures, including video recordings, to both free academic and commercial population-normed test instruments, to biological measurements of physiology, brain structure or function.
+The extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.
 
-Google, Twitter, Amazon, Facebook, SnapChat.
+\subsection{Non-Academically Initiated or Managed Sources of Big Data}
+
+Finally, we mention briefly some large-volume sources of developmental data that are collected and managed by private, non-academic or government entities.
+More than 1.6 million high school students \cite{lewin_more_2013} take standardized tests or provide financial aid information via measures developed and managed by The College Board or the ACT, both of which are private entities.
+The College Board shares Scholastic Aptitude Test (SAT) and college cost and scholarship data with the research community by application \cite{College_Board_data}.
+So does the ACT \cite{ACT_research}.
+
+Internet-based service providers operate at an even larger scale.
+Google's Gmail has more than 900 million users worldwide \cite{lardinois_gmail_2015}, while Facebook has more than a billion \cite{facebook_2014}.
+According to the YouGov site \cite{YouGov}, some 17\% of Gmail users in the United Kingdom are 17-24 years of age.
+Facebook's policies require that users be at least 13 years of age \cite{facebook_under_13}, but detailed information about user demographics for Facebook or other social media popular among children and adolescents is not openly available \cite{BI_social_media}.
+Of course, detailed information about users, their characteristics, and preferences is the primary asset social media companies mine and market to the business community.
+Users receive free services in exchange for providing these data.
+Both Google \cite{Google_Research} and Facebook \cite{Facebook_Research} have arms that both conduct research and cooperate with academic researchers, but there has been significant public criticism about the ethics of certain research projects \cite{meyer_everything_2014}.
+The primary criticism concerns whether Facebook users had given informed consent to participate in the manipulation of their newsfeeds in the way that researchers whose work is supervised by an IRB would be required to secure if a similar study were taken in a laboratory context.
+
+The scale of data collected and managed by non-academic entities dwarfs that of government or academically-managed initiatives.
+Because the data are collected for proprietary business purposes, it is difficult to assess their current or potential impact on the study of human development, but in the final section we turn to a number of technical, conceptual, and theoretical issues that bear on the question.
 
@@
 
-\subsubsection{File and Data Formats}
+\subsubsection{Collection from Multiple Sources and in Diverse Formats}
 
-Conventional conceptualizations of data focus on text or numeric information, but increasingly images, and audio \cite{Mehl} and video recordings have become part of the mix of data streams collected by researchers and others to better understand human behavior.
-Video in particular captures the complexity and richness of behavior unlike any other measure, and so video provides the main source of information for many researchers who study behavior in laboratory, classroom, or museum contexts.
-Images and recordings are both larger, in file size, and denser than text or flat-file data. 
-Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders \cite{Databrary2015, openfmri.org} and, of course, sharing these materials for non-research entertainment uses has become astonishingly popular \cite{YouTube2015}.
-New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states \cite{Picard}, activity patterns, and momentary cognitive \cite{Sliwinski}, and emotional states broadly available to researchers.
-Moreover, new web-based tools like Amazon's Mechanical Turk (http://www.mturk.com) or Apple's HealthKit (http://www.applehealthkit.com/) will make it possible for research teams to collect data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
-Fundamentally, big data are those that cannot readily be processed by traditional databases or systems that are widely familiar to developmental researchers.
+Developmental scientists collect data from sources representing multiple levels of analysis.
+Increasingly, the devices used to make measurements provide data and metadata in a structured, organized, machine-readable formats.
 
-\subsubsection{Collecting data with new technologies, data streams}
+While some researchers continue to use paper and pencil measures to collect survey information, many universities now have site-licenses for web-based tools \cite{SurveyMonkey, Qualtrics} that significantly reduce the amount of manual labor involved in preparing a survey and processing completed data for analysis.
+Behavioral measures involving computer-based tasks are commonly used in developmental research, but most involve the use of custom, project-specific software.
+So the output data files, while often in an electronic form, may require significant post-processing to be linked with other data.
+On the other hand, new tools like Amazon's Mechanical Turk \cite{MTurk} are empowering behavioral researchers to conduct large-scale behavioral science experiments using tools specialized for psychological research \cite{PsiTurk}, and with data delivered in well-structured electronic formats.
+Amazon's terms of use prohibit minors, but developmental researchers have found ways to secure video-based informed consent from parents to enable their children to participate in looking time studies \cite{Lookit} over the web.
 
-As described previously, developmental scientists have and continue to collect data at multiple levels of analysis, from genetic, endocrine, and metabolic measures to physical structure, brain activity, behavioral measures, to peer, family, school, neighborhood, and geographic/cultural context information.
-Some researchers believe that advances in understanding about many development processes will require deeper understanding about how these influences combine to influence trajectories of change.
-If true, achieving that understanding will require continued improvements in the mechanisms of measurement, and the tools for converting these measures into electronic data that can be cross-linked and analyzed.
+The use of video and audio \cite{Mehl} recordings have long been a mainstay of developmental research.
+Video in particular captures the complexity and richness of behavior unlike any other measure, and so video provides a uniquely valuable source of information for many researchers who study behavior in laboratory, classroom, or museum contexts.
+Images and recordings are both larger, in file size, and denser than text or flat-file data, and come in a multiplicity of formats. 
+Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders \cite{Databrary2015, openfmri.org}.
 
-For some some data elements, the devices used to make measurements already provide data and metadata in a structured, organized way.
-For example, genetic analyses from modern gene sequencing tools provide machine-readable outputs.
-Researchers who analyze hormone levels in saliva or micronutrients in blood, for example, (Rick asking Sam Scott, others).
+Lab-based tools for conducting physiological measurements like EEG, heart-rate or skin conductance, produce electronic files of the time series of measurements.
+Genetic analyses from modern gene sequencing tools and reports from tissue, blood, or salivary samples typically yield machine-readable outputs.
 Magnetic Resonance Imaging (MRI) systems produce electronic image data and machine-readable subject-level metadata; however, many research teams limit the amount and kind of subject-level metadata they enter into MRI databases because of the possibility of violating research participant confidentiality.
-Physiological measurements like EEG, heart-rate or skin conductance, produce electronic files of the time series of measurements.
 But, unlike MRI, there are no standard file formats, and most data collection systems provide no standard subject-level metadata.
 Thus, the files require significant post-collection data processing prior to analysis.
-Behavioral measures involving computer-based tasks are commonly used in developmental research.
-But, like physiological measures, many, if not most of these measures involve the use of custom, project-specific software, so the output data files, while often in an electronic form, may or may not be produced in a form that can be easily linked with other data items without significant post-processing.
 
-Recordings such as still images, audio, and video are readily collected or converted to an electronic form.
-There are a number of still image formats, but readily available tools exist to convert between them.
-Audio and video provide a bigger challenge because of the diversity of formats.
-Data archives like Databrary that specialize in video, for example, have chosen to convert all video into a common format in addition to storing files in their original formats for preservation purposes.
+New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states \cite{Empatica}, activity patterns, and momentary cognitive \cite{Sliwinski}, and emotional states broadly available to researchers.
+Moreover, web-based tools like Amazon's Mechanical Turk \cite{MTurk} or Apple's HealthKit \cite{HealthKit} are already making it possible for research teams to collect data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
 
-At the same time, many, perhaps most developmental researchers continue to collect data elements using paper and pencil measures, making some form of manual data entry essential.
+\subsubsection{Storage and Retrieval}
 
-Looking ahead, there is increasing use of web-based tools for data collection (Qualtrix, SurveyMonkey).
-Amazon's Mechanical Turk (http://mturk.com), for example has become a hub for conducting large-scale behavioral science experiments, (see also Todd Gureckis' Psyturk).
-Amazon's terms of use prohibit minors, but some developmental researchers have found ways to carry out studies that require parental consent.
-The LookIt platform developed at MIT allows parents to give video-based informed consent for their child to participate in a behavioral study, such as one involving the measurement of looking times via a laptop or tablet computer's built-in web cam.
-A number of researchers have adapted smartphones to serve as devices to collect survey, behavioral/task-related, and ambient sound information (Sliwinski, Mehl)
+Developmental researchers who wish to store and share big data face a bewildering array of options.
+These include individual or institutional websites, institutional repositories \cite{ScholarSphere}, cloud services (Dropbox, Box, or Amazon), domain specific repositories (ICSPR, Databrary, TalkBank, WordBank, OpenfMRI), domain general services (Researchgate.net, FigShare/SlideShare, Dataverse, and the Open Science Framework), and open source software web sites like GitHub.
+Some journals offer or require storage of datasets, but these are typically limited to text-based flat-files used for statistical analyses and not raw images, videos, or physiological time series.
+The diversity of storage options can pose daunting challenges for researchers and institutions.
+Identifiable and sensitive data must be kept secure.
+Storage solution smust meet the needs of researchers during the active data collection phase of a study while not posing insurmountable hurdles to data sharing down the line.
+The effort to reconcile these sometimes competing demands led Databrary, for example, to build tools that allow researchers to upload session-level video and flat-file data to a secure web-based server as it is collected, thus minimizing post-study data curation.
+The Open Science Framework offers similar functionality for non-identifiable data.
 
-Other social media providers like Twitter, Facebook, Snapchat, Instagram, and YouTube collect individual data and metadata.
-Services are free, so business model involves selling data.
-Apple's HealthKit.
-Other wearable devices like FitBit, Mio, collect data and store it in a private, proprietary cloud server.
-Data is electronic, but data may or may not be accessible to user.
-You can see what data Google has about you.
-New initiatives like DataWallet (https://www.datawallet.io/) are on the horizong that may empower individuals to share their data for a fee.
+Where and how data are stored is only one part of the problem.
+To foster increased reuse, data must be discoverable and accessible to researchers.
+At present, it is far easier to search and discover research publications relevant to a particular topic using web-based search tools than it is to find data.
+There are several reasons.
+With some exceptions noted earlier, many research publications do not use data that are readily available to investigators outside of the research team.
+While practices are changing, datasets that are available may lack persistent, citable, identifiers.
+When data from a publication are potentially available to other researchers, access is often restricted and requires a specific, sometimes time consuming application to a data repository or to the original data producer.
+In contrast, Databrary allows researchers access to a library of data under the same access agreement, an innovation aimed at accelerating reuse.
+Another barrier to reuse is the difficulty in finding data that meet specific task or demographic criteria.
+Some repositories like ICPSR and the National Database for Autism Research (NDAR) maintain extensive standardized metadata about tasks and participant demographics.
+This can help investigators search for specific data sources.
+But, even within domain-specific repositories variable-level searching is not available for all datasets.
+The problems of where to store and how to find and retrieve data will get bigger as datasets grow in size and complexity.
 
-Both old and new data collection systems raise a number of issues.
-How can critical metadata be captured more efficiently without significantly increasing risks to privacy.
-Video cameras that record recording date and time and camera specs, for example.
-Reduce need for post-processing of data, more widespread use of tidy and reproducible data formats.
-Non-proprietary data streams and storage.
-Raise ethical questions about who owns data.
+\subsubsection{Coding, Analysis, and Provenance}
 
-\subsubsection{Storing and Sharing big data}
+Even easy-to-find datasets must be processed prior to analysis.
+Indeed, most data science work involves `janitor work' \cite{lohr_for_2014}.
+The process of curation involves carefully documenting how raw information from a data stream was transformed into information used in formal analyses. 
+Can the provenance of the data be recorded in ways that others can understand, reproduce, and rely upon?
+For example, physiological data are often filtered.
+Video data are usually coded by human observers, and the codes transformed into quantitative measurements.
+What were the variables, units of measurements, calibration properties of the instrument, definition of key terms and codes?
+Well-curated datasets usually report these components, but that curation takes time and specialized expertise that many individual investigators lack.
 
-Big data requires large scale servers.
-If the data are to be served over the web, then the servers require signficant network bandwidth, and all of the other substantial infrastructure required of large-scale data centers.
+Several new software tools have recently emerged that make it easier for data analysts to produce and reproduce self-documenting data workflows.
+For example, the free RStudio \cite{RStudio} and Jupyter \cite{Jupyter} environments make it possible for analysts to create the equivalent of electronic notebooks that combine data, annotations and observations, statistical analyses, and visualizations in human-friendly web or document formats.
+The free, open-source Datavyu \cite{Datavyu} video coding tool allows automated data analysis and export schemes to be created via the Ruby scripting language.
+Many developmental researchers may be unfamiliar with these sorts of tools, but volunteer groups like Software Carpentry \cite{Software_Carpentry} have begun to provide researchers on-site training in the use of tools for reproducible research workflows, including the use of version control and workflow scripting.
 
-Researchers who wish to find and reuse or store and share data face a bewildering array of options.
+\subsubsection{Summary}
 
-\subsubsubsection{Storing and Sharing Data}
+% Summarize technical issues
 
-The options include individual or institutional websites, or cloud services like Dropbox, Box, or Amazon Storage services.
-Other researchers may choose institutional archives like Penn State's ScholarSphere (https://scholarsphere.psu.edu/).
-There are domain specific archives like ICSPR, Databrary, TalkBank, or WordBank, or , OpenfMRI and NeuroVault.
-Other researchers may choose to store and share their data with domain-general services like Researchgate.net, FigShare/SlideShare, Dataverse, or the Open Science Framework.
-Other researchers are turning to open source web sites like GitHub.
-Some journals offer or require storage of datasets, but these are typically limited to text-based flat-files used for statistical analyses and not raw images, videos, or physiological time series.
-What to store, when to store (during, after), and how to share, how to get credit/notice/citations for sharing.
+\subsection{Research Ethics and Practice}
 
-\subsubsubsection{Finding Data}
-
-The task confronting researchers who wish to find and reanalyze or reuse data faces challenges, as well.
-How to discover data?
-What are the key terms? Demographic variables like age and sex. Task, context, language?
-Data associated with specific publications, grants, or researchers.
-Obviously repositories.
-But, can we find the data we want?
-Governments also maintain data archives such as the National Database for Autism Research, the National Center for Educational Statistics, the U.S.
-Census Bureau, the Centers for Disease Control, and the Bureau of Labor Statistics, to name a few sources of developmental science-related data discussed in this paper.
-A number of large-scale research project maintain their own data repositories.
-ICSPR, for example, enables variable-level searching, but not all datasets provide it.
-The International Cross Time Cross System Dataset (XTXS) (http://www.intledstatsdatabase.org/) aspires to provide a comprehensive source of data about educational outcomes.
-The APA maintains a web-page with links to datasets (http://www.apa.org/research/responsible/data-links.aspx).
-
-It is much easier to find publications and search the network of references than it is to find data.
-
-Rick's case study in trying to find and download a public dataset from ICPSR.
-
-Bring data to the analyst or analyses to the data?
-Problem will get bigger as datasets grow in size, complexity.
-
-\subsubsection{Coding and analyzing big data}
-
-Most data science work involves `janitor work' (http://www.nytimes.com/2014/08/18/technology/for-big-data-scientists-hurdle-to-insights-is-janitor-work.html?_r=0) or data carpentry (http://datacarpentry.org/).
-Turning raw materials into numbers.
-How were the materials converted into numbers?
-Data provenance.
-Term definitions, coding books.
-Example, from video to coded spreadsheets to numbers.
-
-Emerging sentiment around the use of tools that enable reproducible science.
-RStudio, Jupyter.
-Version control.
-Text-based, automated/scripted workflows. Galaxy?
-Scripting in Datavyu.
-
-Need for tools and workflows that keep track of the history of researcher's actions.
-MS Word's track changes, but for data cleaning, coding, analysis.
-
-\subsection{Ethical, Conceptual and Theoretical Issues}
-
-The collection, analysis, and interpretation of large scale datasets poses a number of ethical, conceptual and theoretical issues.
+The collection, analysis, and interpretation of large scale datasets poses a number questions related to research ethics, participant privacy, and scientific transparency.
 
@@
 If the data in question are audio or video recordings, different state provisions can come into play.
-For example, in two-party states \cite{http://www.dmlp.org/legal-guide/recording-phone-calls-and-conversations} both the person making the recording and the person or persons being recorded must consent.
+For example, in two-party states \cite{DMLP} both the person making the recording and the person or persons being recorded must consent, making some forms of data collection using recordings problematic, at best.
 
@@
 These entities are regulated by the U.S. Office of Health and Human Services (http://www.hhs.gov/ohrp/).
-Researchers supervised by IRBs must respect privacy, secure informed consent or assent, and maintain confidentiality. 
+Researchers supervised by IRBs must respect participants' privacy, secure informed consent or assent, and maintain confidentiality. 
 Adhering to these ethical principles has practical consequences for research.
@@
 From a big data perspective, if data cannot be shared outside the original IRB-approved research team, then the possible analyses are restricted to the interests, resources, and expertise of that team.
-% What about population data......
 
@@
 Thus, data from photographs or recordings requires additional consideration and special care.
-Databrary (http://databrary.org), a digital data library specialized for storing, managing, and sharing video data from developmental research has an access model that empowers researchers who wish to share identifiable research data to do so with explicit permission of the participants.
+Databrary \cite{databrary.org}, a digital data library specialized for storing, managing, and sharing video data from developmental research has an access model that empowers researchers who wish to share identifiable research data to do so with explicit permission of the participants.
 Databrary has created template language to help researchers gain participants' permission.
@@
 There are reasons to be optimistic, however.
-In just over a year of operation, Databrary has secured formal agreements with more than 75 institutions in North and South America, Europe, Australia, and Asia allowing more than X researchers to access identifiable data.
+In just over a year of operation, Databrary has secured formal agreements with more than 75 institutions in North and South America, Europe, Australia, and Asia allowing more than 100 researchers to access identifiable data.
 On the other hand, some leading developmental researchers have argued that the families of research participants forge a relationship of trust with a particular research team, formalized through the informed consent document \cite{Eisenberg-APS}.
@@
 
-\subsubsection{Challenges of conducting big data developmental research}
+New initiatives like DataWallet (https://www.datawallet.io/) are on the horizon that may empower individuals to share their data for a fee.
+
+\subsection{Conceptual and Theoretical Issues}
 
@@
 But, much of what behavioral and social science has learned about developmental processes comes from samples that represent only a small fraction of the world's population.
-Developmental psychology, like other branches of the field, presents findings from Western, education, industrialized, rich, and democratic (WEIRD) societies (http://www.theatlantic.com/daily-dish/archive/2010/10/western-educated-industrialized-rich-and-democratic/181667/).
+Developmental psychology, like other branches of the field, presents findings from Western, education, industrialized, rich, and democratic (WEIRD) societies \cite{henrich_weirdest_2010}.
 So, to the extent that new tools enable research on development in non-WEIRD cultures and those data can be aggregated and combined, the extent to which developmental science can make claims about universal or near-universal components of developmental processes will be strengthened.
@@
 
-As boyd and Crawford note, `The era of Big Data has begun.
-Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.' \cite{boyd}
+As boyd and Crawford note, ``The era of Big Data has begun.
+Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.'' \cite{boyd}
 The clamor extends to the developmental and learning sciences where interests in understanding how to maximize the potential for human achievement and improve health have significant consequences for daily living.

### 2015-08-26 wiley-big-data-devel/big-data.tex
full draft

@@
 \usepackage{graphicx}
+\usepackage[hyphens]{url}
+\usepackage{hyperref}
+
+\hypersetup{breaklinks=true}
 
@@
 
-\oneauthor{Rick O. Gilmore}
-\twoaffiliations{The Pennsylvania State University, The Databrary Project}
+\author{Rick O. Gilmore}
+\affiliation{{The Pennsylvania State University}, {The Databrary Project}}
 
-\abstract{}
+%\abstract{}
 
@@
 
-A search on the term ``big data'' yields more than 49 million hits on Google and 13 million hits on Bing.
+A search on the term ``big data'' yields more than 49 million hits on Google, more than 100,000 results on Google Scholar, and 13 million hits on Bing.
 The results return in less than a second.
-The term yields more than 100,000 results on Google Scholar.
-A search of the term using Google's ngram viewer (http://books.google.com/ngram) shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
-Google's ngram viewer indexes terms in digitized books.
-So, whether measured by search engine matches of electronic documents available on the internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
+A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
+So, whether measured by search engine matches of electronic documents available on the Internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
 That use has grown substantially over the past several decades.
 The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases.
-Most of the tools that enable it have been created in the last 20 years (https://en.wikipedia.org/wiki/List_of_websites_founded_before_1995).
-``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015; boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012, Marcus2013} and specialist press \cite{HBR2015, Press2013a}.
+Most of the tools that enable it have been created in the last 20 years.
+``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.
 
-In this paper, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
+In this review, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
 I begin by asking what makes data especially big, and what implications the size, density, or complexity of datasets might have for understanding human development.
-Then, I will briefly survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
-To close, I will discuss a range of issues that the emergence of ``big data" approaches to developmental science pose for the future of the field.
+Then, I survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
+I close by discussing a range of issues that the emergence of ``big data'' approaches to developmental science pose for the future of the field.
 
-I will argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
-Nor, do I find that ``big data'' analyses are all that new in developmental psychology.
-Rather, I find that a while the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily interrelatable datasets.
+I argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
+Nor are ``big data'' analyses all that new in developmental psychology.
+Rather, while some the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
 Thus, whether significant novel insights into human development emerge from the era of ``big data'' depend not just on the size or complexity of the datasets, but on how widely and openly the data are shared, and how readily able researchers are able to combine or link datasets across levels of analysis.
-In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual esearcher behavior, publisher priorities, and research funding levels.
-Technology may have enabled the ``big data'' era, but the challenges it poses may turn out to be the least important, at least where advancing research in developmental psychology is concerned.
+In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual researcher behavior, publisher priorities, and research funding levels.
+Technology may accelerate the ``big data'' era, but the challenges it poses may turn out to be the least important where advancing research in developmental psychology is concerned.
 
@@
 
-Doug Laney \cite{laney01controlling3v} was the first person to describe the dimensions of data streams that make them especially ``big'': Volume, velocity, and variety.
-Data streams that are large in quantity, change rapidly, or consist of many components merit special consideration \cite{ibm_2015}.
+According to Laney \citeyear{laney01controlling3v} the volume, velocity, and variety of data streams make data big.
 Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
+What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year (\url{http://home.web.cern.ch/about/computing}).
+Similarly, the measure of volume matters.
+The Inter-university Consortium for Political and Social Research (ICPSR) (\url{https://www.icpsr.umich.edu}), one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections.
+Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project (\url{http://www.metproject.org}), the large number of files in the repository totaled about 10 TB of digital storage.
 
-What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year \cite{CERN_LHC}.
-Similarly, the measure of volume matters.
-The Inter-university Consortium for Political and Social Research (ICPSR) \cite{ICSPR}, one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections \cite{ICPSR}.
-Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project \cite{METProject}, the large number of files in the repository totalled about 10 TB of digital storage.
 Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?
-
 The range of phenomena covered by developmental science spans divergent levels of analysis in space, from genes to geography, and time, from microseconds to millenia, all aimed at answering two questions: What develops?
 Why does it develop this way?
-In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998, gottlieb_normally_1998, oyama_ontogeny_2000, vygotsky_mind_1980}.
-For social and behavioral scientists, the goal is to describing change in the *behavior* or *beliefs* of individuals or groups of individuals and families.
+In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
+For social and behavioral scientists, the goal is to describe change in the \emph{behavior} or \emph{beliefs} of individuals or groups of individuals and families.
 Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
@@
 Consequently, the aspects of data `volume' pertinent to most developmental scientists include the number of participants or families or the number measurement time points.
-We will see that some notable datasets have hundreds or thousands of participants even though samples of tens of participants are more common.
+Some notable datasets have hundreds or thousands of participants even though samples of tens of participants are more common.
 High volume datasets enable more precise estimation of small effects, especially for qualities or conditions rare among individuals.
@@
 
-In the next section, I will review the current state of big data in developmental research.
-In doing so, I summarize information about the size of datasets along the relevant dimensions we have discussed, the availability of datasets for reuse or reanalysis by others, and the impact of particular datasets on scholarship.
-
 \section{Big datasets in developmental research}
-
-% How big, in what ways are big?
-% How open, in what ways open?
-% Size of impact on scholarship?
-% Domains of interest with big datasets?
-% Government vs. individual investigator?
 
@@
 These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
-% Mention focus on individuals and families?
 The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
-% Mention Federal regulations that guide research with human subjects.
 Finally, there are large scale datasets collected by private entities for business purposes that include sharing data about individual subscribers or users.
 The data collected by private entities about individuals largely concerns consumer behavior, as the data customers are other commercial entities seeking to sell products or services to individual users.
-But, commercial entities are beginning to collect and store health and fitness related activity (Fitbit, etc.) 
-These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities who provide the services.
+But, commercial entities are beginning to collect and store health and fitness related activity (e.g., Fitbit, etc.) 
+These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities that provide the services.
 
@@
 Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
-The National Health and Nutrition Examination Survey (NHANES) \cite{NHANES} began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S.
+The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S.
 infants, children, and adults.
@@
 The study is cross-sectional and assesses a nationally representative sample of about 5,000 persons located across the country, with 15 counties surveyed annually.
-Many will be familiar with some of NHANES's products such as standardized child growth charts (http://www.cdc.gov/growthcharts/cdc_charts.htm).
-The National Youth Fitness Survey \cite{NYFS} is closely associated with NHANES.
+Many will be familiar with some of NHANES's products such as standardized child growth charts, and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}).
 NYFS collected data on physical activity and fitness levels in U.S. children ages 3 to 15 in 2012 in order to provide an evaluation their overall fitness.
-Much of the NHANES and NNYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
+Much of the NHANES and NYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
 However, access to some data are restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers (RDC) locations across the U.S.
-A Google Scholar search shows more than 110,000 publications mention the NHANES dataset.
+A Google Scholar search in mid-2015 shows more than 110,000 publications mention the NHANES dataset.
 
-The National Longitudinal Survey of Youth (NLSY) \cite{NLSY} provides an example of a large-scale study that has examined the same cohort -- almost 10,000 American youth born between 1957-64 -- on a annual basis since the study's beginning in 1979.
+The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) provides an example of a large-scale study that has examined the same cohort -- almost 10,000 American youth born between 1957-64 -- on a annual basis since the study's beginning in 1979.
 Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY data are now available for up to 25 survey rounds on questions relating to employment and related issues.
-The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52 through the National Longitudinal Survey (NLS).
-Most NLSY and NLS data are available free of charge to by means of a web-based data portal \cite{NLSY_data} to which users must apply for access.
+The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52.
+Most NLSY and NLS data are available free of charge to by means of a web-based data portal to which users must apply for access.
 Access to geographic-related variables requires special permission.
@@
 
-The Early Childhood Longitudinal Study (ECLS) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences.
+The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences.
 The ECLS consists of three cohorts, one focused on a sample of children measured beginning at birth and two cohorts following children from kindergarten.
@@
 ECLS consists of both publicly available and restricted data.
-A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications (https://www.icpsr.umich.edu/icpsrweb/ICPSR/biblio/series/00191/resources?sortBy=1).
+A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications.
 
 There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
-The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
+The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (\url{http://www.bls.gov/cps/demographics.htm}).
 The U.S.
 Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
-The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years \cite{TIMSS}.
-TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commerical partners.
+The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu/}).
+TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commercial partners.
 The 2001 TIMSS assessment involved more than 20,000 U.S. students from more than 1,000 schools plus data from more than 60 countries.
@@
 In addition to TIMSS, there are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
-The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) \cite{MGRS} collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
+The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
 MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
-Launched in 2011, the Japan Environment and Children's Study \cite{JECS} is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
+Launched in 2011, the Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
 Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
 Health outcomes and exposure measurements will continue until the participating children become 13 years old.
-The study is in its early stages.
-The study has produced a handful of publications \cite{JECS_pubs}, and no data sharing plans have been announced.
-Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health & Development \cite{NSHD_UK} and the National Child Development Study (NCDS) \cite{NCDS_UK}.
+The study has produced a handful of publications, and no data sharing plans have been announced.
+Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
 NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
@@
 Measures include genetic, health, education, social, and economic factors.
-The study is funded by the British government and data are available by registration from the U.K. Data Service \cite{UK_cohort_data}.
-The British Cohort Studies have generated more than 3,500 publications to-date \cite{UK_cohort_citations}.
-Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset \cite{PISA}, which generates more than 5,000 search hits.
+The study is funded by the British government and data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
+The British Cohort Studies have generated more than 3,500 publications to-date.
+Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.
 
-Not all large-scale birth cohort studies succeed.
-A notable failure in the U.S. is the National Children's Study](http://www.nichd.nih.gov/research/NCS/Pages/default.aspx).
+Not all large-scale birth cohort studies succeed, of course.
+A notable failure in the U.S. is the National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
 Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children and their parents from before birth to age 21. 
@@
 
-\subsubsection{Summary}
-
 While no means exhaustive in its coverage, several themes emerge from this sample of U.S. and international population-based studies.
@@
 Some studies involve tens of sample occasions per individual over extended periods of time, and show significant variety in the measures collected.
-Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center.
+Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center where special security provisions apply.
 Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
@@
 
-\subsection{Academic Investigator Initiated and Managed Datasets}
+\subsection{Public/Private Partnerships and Investigator-driven Project}
 
-Investigator-managed big data projects cover a wide range of topics and and even wider range of research methodologies.
-In several cases, the studies involve partnerships between government agencies and academic investigators.
-The Panel Study on Income Dynamics (PSID) \cite{PSID} is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
+Several big data studies in developmental science involve partnerships between government agencies and academic investigators.
+Others involve large-scale data collections that are initiated and managed by individual investigators.
+These projects are diverse in focus and methodology, including population-based studies with a health focus, to behavior genetics, brain imaging, cognition and temperament, and education and employment.
+
+\subsubsection{Population-Based Studies}
+
+A population-based example is the Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
 PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
-As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}. 
-Much of data is available publicly to registered users and agree to the terms of use conditions \cite{PSID_term} that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
-In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement \cite{PSID_restricted_use} between the investigator and the data owner, the University of Michigan.
+As of mid-2015, PSID had generated more than 3,900 publications. 
+Much of data is available publicly to registered users and agree to the terms of use conditions that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
+In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement between the investigator and the data owner, the University of Michigan.
 
-The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} began as a survey study designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
-Undertaken in response to a Congressional mandate, the study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S.
-during the 1994-1995 school year.
+The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) is another population-based public/private partnership.
+SECCYD prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
+The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
+Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometric and demographic information.
+The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
+Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, with approval granted on a case-by-case basis.
+Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
+The SECCYD has generated more than 214 publications as of mid-2015.
+
+The National Longitudinal Study of Adolescent Health (Ad Health; \url{http://www.nichd.nih.gov/news/releases/Pages/adolescent.aspx}) represents a third example.
+Ad Health began as a survey study undertaken in response to a Congressional mandate designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
+The study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year.
 The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
@@
 During Wave IV, biomarker data from a blood test has made available data about glucose, hemoglobin (HbA1c) levels, blood lipids, and related measures. 
-The Ad Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
-As of mid-2015, more than 5,000 publications \cite{add_health_publications} had arisen from the Add Health dataset.
+The Ad Health data are archived at ICSPR and are available for public use.
+As of mid-2015, more than 5,000 publications had arisen from the Add Health dataset.
 
-The NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
-The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
-Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, and anthropometric measures, and demographic information.
-The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
-Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, and approval is granted on a case-by-case basis.
-Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
-The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.
-
-The Family Life Project (FLP) studied the early development children living in rural, largely poor communities in North Carolina and Pennsylvania. 
+Investigator-initiated and managed, the Family Life Project (FLP; \url{http://flp.fpg.unc.edu}) took a population-based approach to studying the early development of children living in rural, largely poor communities in North Carolina and Pennsylvania.
 A birth cohort of 800 children were studied over the first 3 years of life in a series of home, child care visits and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
 Data collected included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
-These features make the FLP large in volume, velocity, and variety.
-Data from Phase I of the FLP are archived at ICPSR \cite{Family_Life_data}, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
-The FPL site lists about 50 publications through 2013.
+Data from Phase I of the FLP are archived at ICPSR, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
+The FLP site lists about 50 publications through 2013.
 
@@
 Several large-scale genetics studies that are not specifically developmental deserve mention.
-The Genome of the Netherlands Project(http://www.bbmri.nl/en-gb/activities/rainbow-projects/131-genome-of-the-netherlands) is a publicly available dataset from n=250 families, consisting of two parents plus one of their adult children \cite{Genome_Netherlands}.
-The Genes, Environment, and Melanoma Study (GEMS) focuses on risk for melanoma in a sample of 3,700 individuals from the North America, Europe, and Australia \cite{GEMS}.
-The Psychiatric Genomics Consortium (PGC) \cite{PGC} contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
+The Genome of the Netherlands Project(\url{http://www.nlgenome.nl}) is a publicly available dataset from 250 families, consisting of two parents plus one of their adult children.
+The Psychiatric Genomics Consortium (PGC; \url{http://www.med.unc.edu/pgc}) contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
 Many disorders featured (autism, attention-deficit hyperactivity disorder, and schizophrenia) in the PGC dataset have developmental dimensions.
-Genome-wide association study (GWAS) results from the PGC dataset may be viewed in a specialized web browser tool (http://www.broadinstitute.org/mpg/ricopili/).
+Genome-wide association study (GWAS) results from the PGC dataset may be viewed in a specialized web browser tool (\url{http://www.broadinstitute.org/mpg/ricopili/}).
 
-The U.K.-based Twins Early Development Study (TEDS) \cite{TEDS} contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
+The U.K.-based Twins Early Development Study (TEDS; \url{http://www.teds.ac.uk/}) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
 Some 13,000 pairs remain actively involved, providing lab-based behavioral task and survey responses.
@@
 
-The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} is another twin-focused study aimed at uncovering genetic and environmental ontributions to measures of family relationships and mental health.
-TOSS extends the Twin Moms Project to include a sample of twin fathers (N=320 twin pairs and their families).
-In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (n=3,000 individuals) and their families.
+The Twin and Offspring Study in Sweden (TOSS; \url{http://ki.se/en/meb/twin-offspring-study-in-sweden-toss}) is another twin-focused study aimed at uncovering genetic and environmental ontributions to measures of family relationships and mental health.
+TOSS extends the Twin Moms Project to include a sample of twin fathers (320 twin pairs and their families).
+In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (3,000 individuals in all) and their families.
 While still relatively early in data collection, more than 24 publications reference the TOSS dataset.
@@
 Illustrative U.S. examples include the Colorado Adoption Project (CAP), the Nonshared Environment in Adolescent Development (NEAD) project, and the Early Growth and Development Study (EGADS).
-Since its inception in 1976, the Colorado Adoption Project (CAP) \cite{CAP} has enrolled more than 2,400 participants from more than 450 families.
+Since its inception in 1976, the Colorado Adoption Project (CAP; \url{http://ibg.colorado.edu/cap}) has enrolled more than 2,400 participants from more than 450 families.
 Data include cognitive abilities, temperament measures, and demographic variables.
-CAP has generated more than 200 published articles \cite{CAP_pubs}.
-Data collection continues.
-The study aims to be the definitive genetically informed longitudinal study of lifespan development \cite{rhea_colorado_2013}.
-Data from 1976-1989 are archived at Dataverse (https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/00913) and can be accessed by application to Dataverse.
+CAP has generated more than 200 published articles, and data collection continues.
+Data from 1976-1989 are archived at Dataverse and can be accessed by application.
 NEAD studied a sample of 720 families, including twins, siblings, and their parents, in a longitudinal design that involved up to 3 data collection waves.
@@
 The team plans to share the data openly later in 2015.
-EGADS is a prospective adoption study of (n=561) triads of birth parents, adoptive parents, and adopted children in the U.S. 
+EGADS is a prospective adoption study of 561 triads of birth parents, adoptive parents, and adopted children in the U.S. 
 The study focuses on genetic and non-genetic factors that influence children’s adjustment from infancy to age 9. 
@@
 
-The NIH MRI Study of of Normal Brain Development \cite{NIH_Pediatric_MRI} collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
-A second cohort of children from birth to 4 years was scanned longitudinally with up to 10 scans. 
+The NIH MRI Study of of Normal Brain Development (\url{http://pediatricmri.nih.gov/nihpd/info}) collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
+A second cohort of children from birth to 4 years was scanned longitudinally with up to 10 scans per participant. 
 Demographic, hormonal, cognitive, affective, and psychiatric data were also collected.
-The data are available to qualified researchers by application, and the dataset has generated some highly cited papers since data collection ended in 2006 \cite{group_total_2012, zielinski_network-level_2010}.
-John Richards and colleagues have combined data from the NIH MRI Study, data collected in their own labs, and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI \cite{Richards_Developmental_MRI_Database, richards_database_2015}.
+The data are available to qualified researchers by application.
+John Richards and colleagues have combined data from the NIH MRI Study, data collected in their own labs, and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI (\url{http://jerlab.psych.sc.edu/NeurodevelopmentalMRIDatabase/}).
 These data can be openly shared with appropriate citation.
 
-More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING) Project \cite{PING} collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from Baltimore, Boston, Honolulu, Los Angeles, New Haven, New York, Sacramento, and San Diego.
-The study was funded by the Eunice Kennedy Shriver National Institute of Child Health and Human Development.
+More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING; \url{http://pingstudy.ucsd.edu/}) Project collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from Baltimore, Boston, Honolulu, Los Angeles, New Haven, New York, Sacramento, and San Diego.
 Data are available to the research community by application.
@@
 
-The Developmental Connectome Project \cite{developmental_connectome}, led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
-This project aims to study the connectome, a map of human brain connectivity, from 20 to 44 weeks post-conception.
+Outside of the U.S., The Developing Human Connectome Project (\url{http://www.developingconnectome.org}), led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
+This project aims to study the connectome, a map of human brain connectivity in fetuses from 20 to 44 weeks post-conception.
 Funded by the European Research Council, the imaging data are accompanied by clinical, behavioral and genetic information.
@@
 
-Looking across the set of large-scale pediatric brain imaging studies described here, several themes emerge.
-One is that these studies require large, complex research teams, usually spanning multiple sites.
-Another is that, with some notable exceptions \cite{richards_database_2015}, when the data are available for use by others, that use is controlled by the original investigative team.
-In some cases, the original investigative team must be included as an author on any publication.
-Another trend among large-scale developmental imaging studies is that the data are usually housed locally, on project-specific sites and not generally on centralized servers that aggregate data across studies and sources \cite{openfmri.org, neurovault.org}.
-
 \subsubsection{Language, Cognition, Temperament}
@@
 The child language community has pioneered the aggregation and sharing of datasets.
-The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established.
+The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established in the behavioral sciences.
 It consists of transcripts and audio and video recordings of children's utterances along with recent data from adults with aphasia.
@@
 Other large-scale sources of language-related data are more recent.
-WordBank (http://wordbank.stanford.edu/) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
-Catherine Tamis-Lemonda has released video data (https://nyu.databrary.org/volume/8) from an NSF-funded longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
-The Human Speechome Project (http://www.media.mit.edu/cogmac/projects/hsp.html) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
-The project has generated more than 78 publications (http://www.media.mit.edu/cogmac/publications.html), but the data have not been made available outside of the original investigative team.
+WordBank (\url{http://wordbank.stanford.edu/}) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
+Catherine Tamis-Lemonda has released video data (\url{https://nyu.databrary.org/volume/8}) from an NSF-funded longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
+The Human Speechome Project (\url{http://www.media.mit.edu/cogmac/projects/hsp.html}) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
+The project has generated more than 78 publications, but the data have not been made available outside of the original investigative team.
 Data was gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
-The LENA Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed \cite{LENA}.
-LENA stands for Language ENvironment Analysis.
+The LENA (Language ENvironment Analysis; \url{http://www.lenafoundation.org/}) Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed.
 The system provides an automatic language collection and analysis tool for speech language professionals and parents.
-The system was designed and is distributed by the LENA Foundation.
 No large-scale archive for LENA data currently exists, but some researchers who use the tool have begun to explore the creation of a data repository specialized for these sorts of data.
@@
 The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
-The Bayley is widely used in clinical and epidemiological settings such as the EPICure \cite{EPICure} because of the norms even though there are questions about its validity \cite{hack_poor_2005}.
-However, data about those norms appears to be under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
-Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE) project on ICPSR \cite{CARE}.
+The Bayley is widely used in clinical and epidemiological settings such as the EPICure (\url{http://www.epicure.ac.uk/}) because of the norms even though there are questions about its validity \cite{hack_poor_2005}.
+However, data about those norms is under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
+Some Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE; \url{http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4091}) project on ICPSR.
 The data are freely available to registered users of ICSPR.
 
-The Wechsler intelligence tests were developed by Dr. David Wechsler, beginning in the 1930's. 
+The Wechsler intelligence tests are another widely used standardized instrument.
+The tests were developed by Dr. David Wechsler, beginning in the 1930's. 
 The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, while the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
-The current version (WISC V) is provided by Pearson \cite{WISC_V}.
+The current version (WISC-V) is published by Pearson (\url{http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html}).
 Data about the norms themselves appears not to be available to researchers.
-However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR \cite{PHDCN}. 
+However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR.
 The data are restricted, and thus are only available by specific authorization.
@@
 In the realm of emotion and temperament, two academic-investigator-initiated measures have been widely adopted.
-The Infant Behavior Questionnaire (IBQ), a parent survey measure, and the Laboratory Temperament Assessment Battery (Lab-TAB) \cite{Lab_TAB} an observational battery, using a set of 3-5 min episodes mimicking situations in everyday life.
-While no archive for IBQ or Lab-TAB data across studies exists, IBQ data on 1,388 participants are archived at ICSPR as part of the Maternal Lifestyle Study (MLS) \cite{MSL}, a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
+The Infant Behavior Questionnaire (IBQ) is a parent survey measure, and the Laboratory Temperament Assessment Battery (Lab-TAB; \url{http://www.uta.edu/faculty/jgagne/labtab}) is an observational battery, using a set of 3-5 min episodes mimicking situations in everyday life.
+While no archive for IBQ or Lab-TAB data across studies exists, IBQ data on 1,388 participants are archived at ICSPR as part of the Maternal Lifestyle Study (MLS; \url{https://neonatal.rti.org/about/mls_background.cfm}), a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
 IBQ and Lab-TAB data are also available from ICPSR from the Family Life Project.
 
-\subsubsection{Education}
+\subsubsection{Education and Learning Science}
 
-There are several investigator initiatiated or managed education studies that deserve mention because of their volume, velocity, or variety or because of their availability for reuse by others.
-The 1999 TIMSS video study recorded teaching practices in seven countries, and some of the materials are available for public use \cite{TIMSS_video}.
-The TIMSS video study repository is hosted at UCLA with support from the Carnegie Foundation for the Advancement of Teaching, The William and Flora Hewlett Foundation and the Gabriella and Paul Rosenbaum Foundation.
-The Gates Foundation-funded Measures of Effective Teaching (MET) \cite{METProject} Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
+Several studies in the education and learning sciences studies deserve mention because of their volume, velocity, or variety or because of their availability for reuse by others.
+The 1999 TIMSS video study recorded math and science teaching practices in seven countries, and some of the materials are available for public use through a repository hosted at UCLA (\url{http://www.timssvideo.com/}).
+The Gates Foundation-funded Measures of Effective Teaching (MET; \url{http://www.metproject.org/}) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
 The MET video data are stored and shared at ICSPR.
-The Spatial Intelligence and Learning Center (SILC) \cite{SILC} is an NSF-funded initiative designed to ``develop the science of spatial learning and to use this knowledge to transform educational practice, supporting children and adults in acquiring the scientific, technical, engineering, and mathematical (STEM) skills required for effective participation in an increasingly high-technology society and global economy.''
-The SILC website hosts links to a set of test and instruments useful in research on spatial learning.
-LearnLab \cite{LearnLab}, another NSF-funded center, combines cognitive theory and computational modeling to understand changes in student knowledge in the context of computer-based math, science and language courses.
-LearnLab hosts a repository, DataShop \cite{DataShop}, to store and openly share detailed data from classrooms employing cognitive tutors.
+The Spatial Intelligence and Learning Center (SILC; \url{http://spatiallearning.org/}) is an NSF-funded initiative designed to develop a multi-disciplinary science of spatial learning, including the development of tests and instruments useful in research.
+LearnLab (\url{http://learnlab.org}), another NSF-funded center, combines cognitive theory and computational modeling to understand changes in student knowledge in the context of computer-based math, science and language courses.
+LearnLab hosts a repository, DataShop (\url{http://learnlab.org/technologies/datashop/index.php}), to store and openly share detailed data from classrooms employing cognitive tutors.
 DataShop enables data storage, management, visualization and analysis in a web-based tool.
@@
 
-Reviewing the set of investigator initiated or managed datasets, we see examples of datasets large in volume, velocity, and variety.
-Measures employed are also diverse, from surveys, to observational measures, including video recordings, to both free academic and commercial population-normed test instruments, to biological measurements of physiology, brain structure or function.
-The extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.
+The studies involving partnerships with government agencies or those initiated and managed by individual investigators exemplify the high volume, velocity, variety, and variability of big datasets.
+The studies employ a range observational measures, including video recordings, free academic and commercial population-normed test instruments, biological measurements of physiology, brain structure or function.
+On the other hand, the extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.
+Only some of the datasets are stored in data repositories, for example.
+This means that many large-scale developmental dataset are usually housed locally, on project-specific sites and not generally on centralized servers that aggregate data across studies and sources.
+In many cases, the secondary use of the data by other researchers is controlled by the original investigative team, including the kinds of questions that third parties may ask.
+In some cases, the original investigative team must be included as an author on any publication.
+Perhaps as a consequence of these factors, these studies have by and large generated fewer scientific publications than government-sponsored efforts.
+In a later section, I discuss specific barriers to more widespread data sharing.
 
@@
 
-Finally, we mention briefly some large-volume sources of developmental data that are collected and managed by private, non-academic or government entities.
+Some large-volume sources of developmental data are collected and managed by private, non-academic or government entities.
 More than 1.6 million high school students \cite{lewin_more_2013} take standardized tests or provide financial aid information via measures developed and managed by The College Board or the ACT, both of which are private entities.
-The College Board shares Scholastic Aptitude Test (SAT) and college cost and scholarship data with the research community by application \cite{College_Board_data}.
-So does the ACT \cite{ACT_research}.
+The College Board shares Scholastic Aptitude Test (SAT) and college cost and scholarship data with the research community by application.
+So does the ACT (\url{http://www.act.org/research/}).
 
-Internet-based service providers operate at an even larger scale.
+Internet-based for-profit service providers operate at an even larger scale.
 Google's Gmail has more than 900 million users worldwide \cite{lardinois_gmail_2015}, while Facebook has more than a billion \cite{facebook_2014}.
-According to the YouGov site \cite{YouGov}, some 17\% of Gmail users in the United Kingdom are 17-24 years of age.
-Facebook's policies require that users be at least 13 years of age \cite{facebook_under_13}, but detailed information about user demographics for Facebook or other social media popular among children and adolescents is not openly available \cite{BI_social_media}.
+According to the YouGov site (\url{https://yougov.co.uk}), some 17\% of Gmail users in the United Kingdom are 17-24 years of age.
+Facebook's policies require that users be at least 13 years of age (\url{https://www.facebook.com/help/157793540954833}), but detailed information about user demographics for Facebook or other social media popular among children and adolescents is not openly available.
 Of course, detailed information about users, their characteristics, and preferences is the primary asset social media companies mine and market to the business community.
 Users receive free services in exchange for providing these data.
-Both Google \cite{Google_Research} and Facebook \cite{Facebook_Research} have arms that both conduct research and cooperate with academic researchers, but there has been significant public criticism about the ethics of certain research projects \cite{meyer_everything_2014}.
+Both Google and Facebook have arms that both conduct research and cooperate with academic researchers, but there has been significant public criticism about the ethics of certain research projects \cite{meyer_everything_2014}.
 The primary criticism concerns whether Facebook users had given informed consent to participate in the manipulation of their newsfeeds in the way that researchers whose work is supervised by an IRB would be required to secure if a similar study were taken in a laboratory context.
@@
 The scale of data collected and managed by non-academic entities dwarfs that of government or academically-managed initiatives.
-Because the data are collected for proprietary business purposes, it is difficult to assess their current or potential impact on the study of human development, but in the final section we turn to a number of technical, conceptual, and theoretical issues that bear on the question.
+Because the data are collected for proprietary business purposes, it is difficult to assess their current or potential impact on the scholarship of human development.
 
@@
 
-While some researchers continue to use paper and pencil measures to collect survey information, many universities now have site-licenses for web-based tools \cite{SurveyMonkey, Qualtrics} that significantly reduce the amount of manual labor involved in preparing a survey and processing completed data for analysis.
+While some researchers continue to use paper and pencil measures to collect survey information, many universities now have site-licenses for web-based tools like SurveyMonkey and Qualtrics that significantly reduce the amount of manual labor involved in preparing a survey and processing completed data for analysis.
 Behavioral measures involving computer-based tasks are commonly used in developmental research, but most involve the use of custom, project-specific software.
-So the output data files, while often in an electronic form, may require significant post-processing to be linked with other data.
-On the other hand, new tools like Amazon's Mechanical Turk \cite{MTurk} are empowering behavioral researchers to conduct large-scale behavioral science experiments using tools specialized for psychological research \cite{PsiTurk}, and with data delivered in well-structured electronic formats.
-Amazon's terms of use prohibit minors, but developmental researchers have found ways to secure video-based informed consent from parents to enable their children to participate in looking time studies \cite{Lookit} over the web.
+So, the output data files, while often in an electronic form, may require significant post-processing to be linked with other data.
+On the other hand, new tools like Amazon's Mechanical Turk (\url{http://www.mturk.com}) or Apple's HealthKit (\url{https://developer.apple.com/healthkit/}) are empowering behavioral and health researchers to conduct large-scale behavioral science experiments using tools specialized for psychological research (e.g., \url{https://psiturk.org}), and with data delivered in well-structured electronic formats.
+Amazon's terms of use prohibit minors, but developmental researchers have found ways to secure video-based informed consent from parents to enable their children to participate in looking time studies (\url{https://lookit.mit.edu}) over the web.
 
-The use of video and audio \cite{Mehl} recordings have long been a mainstay of developmental research.
+The use of video and audio recordings have long been a mainstay of developmental research.
 Video in particular captures the complexity and richness of behavior unlike any other measure, and so video provides a uniquely valuable source of information for many researchers who study behavior in laboratory, classroom, or museum contexts.
 Images and recordings are both larger, in file size, and denser than text or flat-file data, and come in a multiplicity of formats. 
-Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders \cite{Databrary2015, openfmri.org}.
+Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders .
 
@@
 
-New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states \cite{Empatica}, activity patterns, and momentary cognitive \cite{Sliwinski}, and emotional states broadly available to researchers.
-Moreover, web-based tools like Amazon's Mechanical Turk \cite{MTurk} or Apple's HealthKit \cite{HealthKit} are already making it possible for research teams to collect data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
+New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states (e.g., \url{https://www.empatica.com/}), activity patterns, and momentary cognitive, and emotional states broadly available to researchers.
+These tools are beginning to enable the collection of data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
 
@@
 Developmental researchers who wish to store and share big data face a bewildering array of options.
-These include individual or institutional websites, institutional repositories \cite{ScholarSphere}, cloud services (Dropbox, Box, or Amazon), domain specific repositories (ICSPR, Databrary, TalkBank, WordBank, OpenfMRI), domain general services (Researchgate.net, FigShare/SlideShare, Dataverse, and the Open Science Framework), and open source software web sites like GitHub.
+These include individual or institutional websites, institutional repositories (e.g., \url{https://scholarsphere.psu.edu/}), cloud services (Dropbox, Box, or Amazon), domain or measure-specific repositories (ICSPR, Databrary.org, TalkBank.org, WordBank.org, OpenfMRI.org), domain general services (Researchgate.net, FigShare/SlideShare, Dataverse, and the Open Science Framework), and open source software web sites like GitHub.
 Some journals offer or require storage of datasets, but these are typically limited to text-based flat-files used for statistical analyses and not raw images, videos, or physiological time series.
@@
 Identifiable and sensitive data must be kept secure.
-Storage solution smust meet the needs of researchers during the active data collection phase of a study while not posing insurmountable hurdles to data sharing down the line.
-The effort to reconcile these sometimes competing demands led Databrary, for example, to build tools that allow researchers to upload session-level video and flat-file data to a secure web-based server as it is collected, thus minimizing post-study data curation.
-The Open Science Framework offers similar functionality for non-identifiable data.
+Storage solutions must meet the needs of researchers during the active data collection phase of a study while not posing insurmountable hurdles to data sharing down the line.
+The effort to reconcile these sometimes competing demands led Databrary (\url{http://databrary.org}), for example, to build tools that allow researchers to upload session-level video and flat-file data to a secure web-based server as it is collected, thus minimizing post-study data curation.
+The Open Science Framework (\url{https://osf.io}) offers similar data management functionality for non-identifiable data.
 
@@
 Can the provenance of the data be recorded in ways that others can understand, reproduce, and rely upon?
-For example, physiological data are often filtered.
+For example, physiological data are often filtered, sometimes by the recording devices.
 Video data are usually coded by human observers, and the codes transformed into quantitative measurements.
@@
 Several new software tools have recently emerged that make it easier for data analysts to produce and reproduce self-documenting data workflows.
-For example, the free RStudio \cite{RStudio} and Jupyter \cite{Jupyter} environments make it possible for analysts to create the equivalent of electronic notebooks that combine data, annotations and observations, statistical analyses, and visualizations in human-friendly web or document formats.
-The free, open-source Datavyu \cite{Datavyu} video coding tool allows automated data analysis and export schemes to be created via the Ruby scripting language.
-Many developmental researchers may be unfamiliar with these sorts of tools, but volunteer groups like Software Carpentry \cite{Software_Carpentry} have begun to provide researchers on-site training in the use of tools for reproducible research workflows, including the use of version control and workflow scripting.
+For example, the free RStudio (\url{https://www.rstudio.com}) and Jupyter (\url{https://jupyter.org}) environments make it possible for analysts to create the equivalent of electronic notebooks that combine data, annotations and observations, statistical analyses, and visualizations in human-friendly web or document formats.
+The free, open-source Datavyu (\url{http://datavyu.org}) video coding tool allows automated data analysis and export schemes to be created via the Ruby scripting language.
+Many developmental researchers may be unfamiliar with these sorts of tools, but volunteer groups like Software Carpentry (\url{https://software-carpentry.org}) have begun to provide researchers on-site training in the use of tools for reproducible research workflows, including the use of version control and workflow scripting.
+Similarly, Databrary and the Center for Open Science (\url{http://centerforopenscience.org}) have initiated a series of conference-based and regional workshops to provide hands-on researcher training.
+But, the use of tools that produce well-curated, reproducible scientific workflows remains rare among mainstream developmental researchers.
 
@@
 
-% Summarize technical issues
+Technical issues will continue to slow progress in many areas of developmental research that are generated and big data.
+Critical challenges include getting data into open, standard, and easily manipulable electronic formats as soon as possible in the research cycle; the development and widespread adoption of data storage platforms or repositories that provide a degree of metadata standardization and enable search and discovery; the creation and adoption of data management practices that make curation part of the typical research workflow; and the creation of a cohort of developmental researchers who have the training and expertise to implement these techniques in their own labs.
+On the other hand, there is demonstrable progress on many of these fronts, and therefore many reasons to be optimistic that the technical challenges can be overcome.
 
@@
 
-The collection, analysis, and interpretation of large scale datasets poses a number questions related to research ethics, participant privacy, and scientific transparency.
-
-\subsubsection{Ethics and Privacy}
-
+Clearly the collection, analysis, and interpretation of large scale datasets pose a number questions related to research ethics, participant privacy, and scientific transparency.
 Developmental researchers often collect data from vulnerable populations of research participants.
@@
 In the U.S., researchers must navigate a regulatory environment in which different types of data are covered under different sections of Federal law.
-For example, the Federal Educational Rights and Privacy Act (FERPA - http://www2.ed.gov/policy/gen/guid/fpco/ferpa/index.html) governs who has access to a student's educational records.
-The Health Insurance Portability and Accountability Act (HIPAA - http://www.hhs.gov/ocr/privacy/hipaa/understanding/), among other functions, governs the process by which individually identifiable health information may be disclosed and to whom.
+For example, the Federal Educational Rights and Privacy Act (FERPA; \url{http://www2.ed.gov/policy/gen/guid/fpco/ferpa/index.html}) governs who has access to a student's educational records.
+The Health Insurance Portability and Accountability Act (HIPAA; \url{http://www.hhs.gov/ocr/privacy/hipaa/understanding/}), among other functions, governs the process by which individually identifiable health information may be disclosed and to whom.
+The Code of Federal Regulations (CFR) Title 45 (\url{http://www.hhs.gov/ohrp/humansubjects/guidance/45cfr46.html}) governs research with human participants.
 If the data in question are audio or video recordings, different state provisions can come into play.
-For example, in two-party states \cite{DMLP} both the person making the recording and the person or persons being recorded must consent, making some forms of data collection using recordings problematic, at best.
+For example, in two-party states (\url{http://www.dmlp.org/legal-guide/recording-phone-calls-and-conversations}) both the person making the recording and the person or persons being recorded must consent, making some forms of data collection using recordings problematic, at best.
 
@@
 Federal Government, typically in colleges, universities, or academic or private medical centers, are supervised by an Institutional Review Board or its equivalent.
-These entities are regulated by the U.S. Office of Health and Human Services (http://www.hhs.gov/ohrp/).
+These entities are regulated by the U.S. Office of Health and Human Services (\url{http://www.hhs.gov/ohrp/}).
 Researchers supervised by IRBs must respect participants' privacy, secure informed consent or assent, and maintain confidentiality. 
@@
 IRBs may view these matters differently, creating additional complexities.
-Some IRBs may require participants to be informed that de-identified data might be shared outside the IRB-approved research team, and others may deem the analysis of de-identified data as no longer meeting the definition of human subjects research.
+Some IRBs may require participants to be informed that de-identified data might be shared outside the IRB-approved research team, and others may deem the analysis of de-identified data no longer meets the definition of human subjects research.
 From a big data perspective, if data cannot be shared outside the original IRB-approved research team, then the possible analyses are restricted to the interests, resources, and expertise of that team.
@@
 Thus, data from photographs or recordings requires additional consideration and special care.
-Databrary \cite{databrary.org}, a digital data library specialized for storing, managing, and sharing video data from developmental research has an access model that empowers researchers who wish to share identifiable research data to do so with explicit permission of the participants.
+Databrary, a digital data library specialized for storing, managing, and sharing video data from developmental research has an access model that empowers researchers who wish to share identifiable research data to do so with explicit permission of the participants.
 Databrary has created template language to help researchers gain participants' permission.
@@
 The notion that research participants can consent to share identifiable or potentially identifiable data is relatively new.
-The Personal Genomes Project, Open Humans Project, and Human Connectome Project embody similar principles.
+The Personal Genomes Project (\url{http://www.personalgenomes.org}), Open Humans Project (\url{https://www.openhumans.org}), and Human Connectome Project (\url{http://www.humanconnectomeproject.org}) embody similar principles.
 The experience of Databrary investigators is that a significant proportion -- more than 75 percent -- of research participants and their parents or guardians will consent to sharing identifiable data, mostly video, with other members of the research community.
 
-It is too early to predict whether it will become commonplace for academic developmental researchers to seek explicit permission to share research data, including identifiable elements, with other researchers.
+It is too early to predict whether it will become commonplace for academic developmental researchers to seek explicit permission to share research data that includes identifiable elements, with other researchers.
 There are reasons to be optimistic, however.
 In just over a year of operation, Databrary has secured formal agreements with more than 75 institutions in North and South America, Europe, Australia, and Asia allowing more than 100 researchers to access identifiable data.
-On the other hand, some leading developmental researchers have argued that the families of research participants forge a relationship of trust with a particular research team, formalized through the informed consent document \cite{Eisenberg-APS}.
+On the other hand, some leading developmental researchers have argued that the families of research participants forge a relationship of trust with a particular research team, formalized through the informed consent document \cite{eisenberg_thoughts_2015}.
 The relationship might be harmed or the research project negatively affected if participants were asked to share data with other researchers.
@@
 Unlike academic settings, where violations of research ethics principles may involve significant consequences for the researchers and institutions, violations of commercial terms of use require aggrieved parties to seek redress through litigation.
-The software industry has recommended its members adopt common data privacy principles \cite{}, but these guidelines are voluntary.
+The White House has recommended data privacy principles \cite{data-privacy-networked-world} that some software companies have adopted voluntarily.
 
-Summary.
-Many big data sets have restrictions on them, aren't permissioned for open, widespread sharing
-Must be used at Restricted Data facilities, can't be used to study individuals.
-Studies of individuals may or may not have permission to share widely.
-New models for gaining permission to share identifiable data on a restricted basis may expand the availability of big data for data reuse.
-Have to ensure that data reuse is not prohibited by formal consent process.
-Have to ensure that data don't have to be destroyed.
-Remaining concerns about possibility that data linkage increases the risks of reidentification.
-There are questions about whether it is essential to reconsent minors when they become adults.
-Reconsenting may be impractical.
-General concern about whether it is possible to consent to sharing of an indefinite nature.
-Security can't be guaranteed, so risks can be minimized, but not reduced.
-Files that are downloaded increase risks, for image streams, it is impractical to prevent image/sound capture.
-Principle of informed consent important to uphold.
-Not clear how fully informed consumers are in using free commercial services.
+Research ethics continue to evolve, but several aspects of the current state-of-the art stand out.
+Many big data sets in developmental science have restrictions on access either because the data were collected under Federal regulations that prohibit releasing data that could identify individuals or because the participants were not asked for permission to share individual-level data with other researchers.
+These restrictions impede the secondary reuse of existing dataset.
+New models for gaining permission from participants or their families to share identifiable data on a restricted basis with the research community have been developed and are being implemented in a growing number of laboratories.
+The widespread adoption of these practices promises to expand the availability of big data for reuse in the future.
+
+Unresolved issues that could impact the availability of big data in the future include whether linkage across streams increases the risks of reidentification, whether it is essential to reconsent minors when they become adults, a notion most researchers find totally impractical and a significant barrier to data sharing, and a general concern about the ethics of granting consent to share data for an indefinite period.
+Since data security can't be guaranteed, risks can only be minimized and managed, but not eliminated.
+Finally, there are unresolved questions about privacy protections in the consumer domain that have the potential to influence academic research.
 
@@
 
-The social and behavioral sciences have seen a string of high profile cases of scientific misconduct in recent years, including cases of fraudulent data \cite{}.
+Another dimension of scientific ethics concerns transparency and reproducibility.
+The social and behavioral sciences have seen a string of high profile cases of scientific misconduct in recent years, including cases of fraudulent data \cite{singal_case_2015, bhattacharjee_diederik_2013}.
 The credibility problem is magnified by several factors.
-That lack of power and unrestricted exploratory analyses may mean that most research findings are false \cite{Ioanides}.
+That lack of power and unrestricted exploratory analyses may mean that most research findings are false \cite{ioannidis_why_2005}.
 The actual effect sizes of published findings is unknown due to a bias toward publishing positive results.
-Most journals reject papers that report failures to replicate published findings, and as a result, few scientists attempt or are recognized and rewarded for conducting genuine replications \cite{Nosek-Spies}.
+Most journals reject papers that report failures to replicate published findings, and as a result, few scientists attempt or are recognized and rewarded for conducting genuine replications \cite{nosek_scientific_2012}.
 The problem is so serious that some have claimed that science as a whole faces a crisis of replicability.
 
-To address the crisis, the Center for Open Science (http://cos.io) has organized several large-scale replication efforts, including some in psychological science under the umbrella of the ``Many Labs'' project \cite{klein, https://osf.io/ct89g/; https://osf.io/8cd4r/}.
+To address the crisis, the Center for Open Science has organized several large-scale replication efforts, including some in psychological science under the umbrella of the ``Many Labs'' project (\url{https://osf.io/ct89g/}; \url{https://osf.io/8cd4r/}).
 The results of these pre-registered, open, large sample replications have been mixed.
@@
 Developmental research reflects the same positive effects biases seen in other fields, and the same problem that null results often sit unpublished in file drawers -- the so-called file drawer effect \cite{rosenthal_file_1979}.
-No failures to replicate developmental studies have been reported to Psychfiledrawer.org (http://psychfiledrawer.org/view_article_list.php), a resource designed to bring replications failures to light.
-As some developmental researchers have written (http://deevybee.blogspot.co.uk/2012/01/novelty-interest-and-replicability.html), replicating effects with developmental populations can be especially difficult and so even partial replications are noteworthy.
-No large-scale replication efforts in developmental science have been mounted, but there have been calls for changes in journal practices to give replications have a more privileged place in scientific publications (http://deevybee.blogspot.co.uk/2012/01/novelty-interest-and-replicability.html).
+No failures to replicate developmental studies have been reported to Psychfiledrawer.org (\url{http://psychfiledrawer.org}), a resource designed to bring replications failures to light.
+As some developmental researchers have written \cite{bishop-blog-2012}, replicating effects with developmental populations can be especially difficult and so even partial replications are noteworthy.
+No large-scale replication efforts in developmental science have been mounted, but there have been calls for changes in journal practices to give replications have a more privileged place in scientific publications \cite{bishop-blog-2012}.
 One barrier to more open data practices appears to be researcher's fears of having one's reputation or abilities publicly undermined \cite{ascoli_ups_2006}.
 So, changing views about replication may require shifts in the scientific culture.
-Researchers should work to reduce the extent of blame levied at researchers whose initial positive findings fail to be replicated by others (http://deevybee.blogspot.com/2015/07/publishing-replication-failures-some.html).
+Researchers should work to reduce the extent of blame levied at researchers whose initial positive findings fail to be replicated by others \cite{bishop-blog-2015}.
 Technological tools that foster increased openness and transparency and more systematic research data management (OSF and Databrary) will also contribute to changing the scientific practices.
@@
 Census).
-Successful big data projects -- repositories or specific studies or datasets -- appear to have fostered a high level of engagement around a particular resource, often through specific training, and targeted workshops (e.g.
-Ad Health, ICSPR, Databrary, AERA workshop on video data sharing, COS, Data/Software Carpentry).
+Successful big data projects -- repositories or specific studies or datasets -- appear to have fostered a high level of engagement around a particular resource, often through specific training, and targeted workshops.
+
 In addition to bottom-up/grassroots initiatives, journals and funding agencies continue to play a vital role in creating an impetus for change.
-Funders can require data management plans, mandate data be deposited into particular types of open repositories (Gates Foundation; NIMH call), and provide funding to build and support big data infrastructure.
+Funders can require data management plans and mandate that data and research products be deposited into particular types of open repositories as the Gates Foundation has done, and provide funding to build and support big data infrastructure.
 Journals can also require that data be deposited in open archives as a condition of publication in addition to adopting other transparent and open science practices for manuscripts they accept.
-However, the problem with data sharing mandates from funders is that there is no specific funding mechanism to provide sufficient ongoing financial support for data archives, and few researchers budget funds to support data management and archiving.
-And, while some journals are willing to take on the burden of storing and sharing data associated with publications, others (Journal of Neuroscience) have gone the other way.
+However, the problem with data sharing mandates from funders is that there is no specific mechanism to provide sufficient ongoing financial support for data archives, and few researchers budget funds to support data management and archiving.
+And, while some journals are willing to take on the burden of storing and sharing data associated with publications, others \cite{maunsell_announcement_2010} have gone the other way and refuse to accept supplemental materials.
 Thus, in the interests of promoting greater openness and transparency, funders and journals may create unfunded mandates that make it harder for researchers to make discoveries.
-For example, new regulations about when data must be deposited may be unwieldy and impractical for developmental scientists. 
-See Nancy Eisenberg piece on data sharing. (https://www.psychologicalscience.org/index.php/publications/observer/2015/may-june-15/thoughts-on-the-future-of-data-sharing.html).
-APA stance on data sharing. (http://www.apa.org/science/leadership/bsa/data-sharing-report.aspx).
+For example, new regulations specifying when data must be deposited may be unwieldy and impractical for some developmental scientists to carry out their work \cite{eisenberg_thoughts_2015, apa_data_sharing_work_group_data_2015}.
 
-\subsubsection{Ownership and Access}
-
-Who owns research data? (http://ori.hhs.gov/education/products/n_illinois_u/datamanagement/dotopic.html; http://ori.hhs.gov/education/products/n_illinois_u/datamanagement/dotopic.html).
-The scientific community lacks consensus on the answer, and the developmental science community is no exception.
+These issues are complicated by lack of consensus about who `owns' research data \cite{who_owns_research_data}.
 Federal funding agencies might argue that the public should `own' data paid for by tax dollars, much like other data collected by government agencies like the U.S.
@@
 Most investigators naturally feel a strong sense of ownership over their intellectual products, even though formal copyright is often surrendered in the process of publishing, and that sense extends to data.
-Some have even argued that research participants themselves own their own data, and there are new business models emerging that may soon provide individuals an opportunity to sell data for personal gain (http://www.datawallet.io).
+Some have even argued that research participants themselves own their own data, and there are new business models emerging that may soon provide individuals an opportunity to sell data for personal gain (\url{http://www.datawallet.io}).
 
@@
 That consensus may well prove vital to achieving some of the benefits of big data analyses in development.
-
-New initiatives like DataWallet (https://www.datawallet.io/) are on the horizon that may empower individuals to share their data for a fee.
 
@@
 
-A second challenge larger datasets may help to address is the fact that most social, behavioral \cite{Maxwell-2004} and neuroscience studies \cite{Button-NRN} are underpowered.
-Most worryingly it has been claimed that most published research findings are false in fields with specific characteristics: Sample or sizes are small; more relationships between variables are tested; preselection of tested relationships is less common and exploratory research more common; designs, definitions, outcomes, and analytical modes are less consistent across studies; and when more teams are involved seeking out significant effects \cite{Ioanidis}.
+A second challenge larger datasets may help to address is the fact that most social, behavioral \cite{maxwell_persistence_2004} and neuroscience studies \cite{button_power_2013} are underpowered.
+Most worryingly it has been claimed that most published research findings are false in fields with specific characteristics: Sample or sizes are small; more relationships between variables are tested; preselection of tested relationships is less common and exploratory research more common; designs, definitions, outcomes, and analytical modes are less consistent across studies; and when more teams are involved seeking out significant effects \cite{ioannidis_why_2005}.
 The collection, analysis, and sharing of larger size datasets may help to ameliorate some, but not all, of these flaws.
@@
 
-Finally, some have raised concerns that the rise of especially large datasets means the `end of theory.' \cite{Anderson}
+Finally, some have raised concerns that the rise of especially large datasets means the `end of theory.' \cite{anderson_end_2008}
 In a provocative essay in 2008, Anderson argued that large quantities of data mean the traditional model of scientific inquiry involving hypothesis testing will soon give way to model-free descriptions of data.
-Others note that bigger data don't necessarily lead to deeper insights \cite{Graham} (http://www.theguardian.com/news/datablog/2012/mar/09/big-data-theory).
-Boyd and Crawford note \cite{boyd} that beyond its scientific dimensions, `big data' is a sociocultural phenomenon that raises as many questions as it promises answers.
+Others note that bigger data don't necessarily lead to deeper insights \cite{graham_big_2012}.
+Boyd and Crawford note \cite{boyd_critical_2012} that beyond its scientific dimensions, `big data' is a sociocultural phenomenon that raises as many questions as it promises answers.
 
@@
 
-As boyd and Crawford note, ``The era of Big Data has begun.
-Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.'' \cite{boyd}
+As boyd and Crawford \citeyear{boyd_critical_2012} note, ``The era of Big Data has begun.
+Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.''
 The clamor extends to the developmental and learning sciences where interests in understanding how to maximize the potential for human achievement and improve health have significant consequences for daily living.
@@
 Many `big data' studies have had substantial impact on scholarship, as measured by the quantity of published findings, and in some cases, on public understanding and policy.
+For the most part, studies that have had the largest impact have been ones funded by and managed by government entities, perhaps in part to the widespread access to data for reuse by others.
+Those investigator-initiated and managed projects that have had the largest intellectual impacts have created significant intellectual communities around the datasets, communities that extend the beyond the boundaries of the original investigative teams.
+While technical issues about data formats, storage, cleaning, provenance, visualization and provenance remain, there is significant progress.
+Developmental researchers have available a growing array of well-established data repositories (ICSPR, Dataverse) and new data storage/management tools (Databrary, OSF).
+Research and data management practices have begun to converge on norms that will reduce the costs of preparing data for sharing in the future.
+New ethical procedures for seeking informed consent to share identifiable data have been developed and are being implemented in diverse research contexts.
 
-Research and data management practices have begun to converge on norms that will reduce the costs of preparing data for sharing.
-New ethical procedures
-New data archives, increasing support for sharing, both bottom-up and top down.
+On the other hand, there are many barriers to progress in the analysis of big datasets in developmental science.
+Most developmental science data are hard to find, and cumbersome to access, even for researchers.
+Most data linked to publications are not stored in open data repositories. 
+This means that virtually all of the data from unpublished studies remains unavailable, making the size of the file drawer effect unknown.
+Most investigators do not currently employ workflows that make it easy to share data or document analysis pathways.
+With rare exceptions clustered around specific datasets there is no widespread culture of data sharing, and indeed some level of bias against the use of secondary data.
+Finally, there is no unified understanding or consensus within developmental science about who owns research data, whether it is essential or merely wise to share data, and when in the research cycle data should be shared.
 
-Data are still hard to access, even for researchers.
-Most data are not in repositories, file drawer effect.
-Most data not prepared or curated for sharing.
-Data hard to understand without help of original investigators.
-No widespread culture of data sharing except in specific subfields, CHIDLES.
-No unified understanding/cultural consensus about who owns data, when to share, etc.
-
-At the same time, while scientific culture may appear to change slowly, let us remember that the Facebook website was launched in 2004, Twitter in 2006.
-It would be unwise to underestimate the speed with which new technologies and tools can change our lives.
+At the same time, while scientific culture may appear to change slowly, remember that Facebook was launched in 2004, Twitter in 2006, and the iPhone in 2007.
+It would be unwise to underestimate the speed with which new technologies, tools, and cultural practices can change.
 If developmental researchers can find ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development, we can look forward to a future rich in theory and understanding.

### 2015-09-02 wiley-big-data-devel/big-data.tex
final draft

@@
-\documentclass[letterpaper,man,apacite]{apa6}
+\documentclass[letterpaper,man,apacite,natbib]{apa6}
 
@@
 \usepackage[hyphens]{url}
-\usepackage{hyperref}
+%\usepackage{hyperref}
 
-\hypersetup{breaklinks=true}
+% \hypersetup{breaklinks=true}
 
@@
 
-%\abstract{}
+\abstract{The use of the term ``big data'' has grown substantially over the past several decades, and is now widespread. 
+In this review, I ask what makes data ``big'' and the implications of size, density, or complexity of datasets for the science of human development.
+A survey of existing data sets illustrates how large, complex, multi-level, multi-measure data have been used to reveal the complexities of human development.
+``Big data'' poses significant technical, policy, ethics, transparency, cultural, and conceptual issues that must be addressed.
+Most big developmental science data are currently hard to find and cumbersome to access; the field lacks a culture of data sharing; and there is no consensus about who owns or should control research data.
+But, these barriers are rapidly dissolving.
+Developmental researchers are finding new ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development.
+This promises a future in which ``big data'' can lead to deep insight about some of the most vexing questions in behavioral science.
+}
 
 \authornote{Rick O. Gilmore is in the Department of Psychology, The Pennsylvania State University, University Park, PA 16802, rogilmore@psu.edu.
-Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.}
+This work was supported by the National Science Foundation (BCS-1238599), the Eunice Kennedy Shriver National Institute of Child Health and Human Development (U01-HD-076595), and the Society for Research in Child Development.
+Any opinions, findings, and conclusions or recommendations expressed in the material contributed here are those of the author and do not necessarily reflect the views of the National Science Foundation, the Eunice Kennedy Shriver National Institute of Child Health and Human Development, or the Society for Research in Child Development.
+I thank Karen E. Adolph for her comments on the manuscript.}
 
@@
 The results return in less than a second.
-A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term right before 1900 with a steady rise in frequency from the 1950s to just after 2000.
-So, whether measured by search engine matches of electronic documents available on the Internet, scholarly documents, or digitized books, the term ``big data'' gets widespread use now.
-That use has grown substantially over the past several decades.
-The fact that these basic facts about a particular phrase can be determined in an instant speaks to the rapid pace of change in networked computing, search engines, and databases.
+A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term around 1900 with a steady rise in frequency from the 1950s to around 2000.
+Clearly, as measured by search engine matches of electronic documents on the Internet, scholarly documents, or digitized books, the use of the term ``big data'' has grown substantially over the past several decades, and is now widespread.
+Moreover, the fact that these basic statistics about a particular phrase can be determined in an instant speaks to the rapid progress in networked computing, search engines, and databases.
 Most of the tools that enable it have been created in the last 20 years.
@@
 
-In this review, I focus on how the increased availability of and interest in analyzing large, complex, multi-level, multi-measure data sets has influenced, continues to influence, and promises to alter the study of human development.
-I begin by asking what makes data especially big, and what implications the size, density, or complexity of datasets might have for understanding human development.
-Then, I survey a set of case studies that illustrate how large, complex data sets have been and are being used currently to reveal the complexities of developmental processes.
-I close by discussing a range of issues that the emergence of ``big data'' approaches to developmental science pose for the future of the field.
+In this review, I show how increased availability of and interest in big data sets promises to alter the study of human development.
+I begin by asking what makes data ``big'' and the implications of size, density, or complexity of datasets for understanding human development.
+Then, I survey case studies to illustrate how large, complex, multi-level, multi-measure data sets have been used to reveal the complexities of human development.
+I conclude by discussing key questions that ``big data'' approaches pose for the future of developmental science pose.
 
-I argue against the claim that the emergence of an era of ``big data'' signals the end of theory \cite{anderson_end_2008} while acknowledging the lure of that notion to some.
-Nor are ``big data'' analyses all that new in developmental psychology.
-Rather, while some the promises of ``big data'' to revolutionize scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
-Thus, whether significant novel insights into human development emerge from the era of ``big data'' depend not just on the size or complexity of the datasets, but on how widely and openly the data are shared, and how readily able researchers are able to combine or link datasets across levels of analysis.
+I disagree with the claim that an era of ``big data'' signals the end of theory \cite{anderson_end_2008}.
+Nor are ``big data'' analyses especially new in developmental science.
+Although some claims about ``big data'' revolutionizing scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
+Significant novel insights into human development emerging from the era of ``big data'' depend on the size, density, and complexity of the datasets, on how widely and openly the data are shared, and on how readily researchers are able to combine or link datasets across levels of analysis.
 In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual researcher behavior, publisher priorities, and research funding levels.
-Technology may accelerate the ``big data'' era, but the challenges it poses may turn out to be the least important where advancing research in developmental psychology is concerned.
+Technology may accelerate the ``big data'' era, but the challenges it poses turn out to be relatively unimportant for advancing research in developmental psychology.
 
-\subsection{What Does `Big Data` Mean in Developmental Science?}
+\subsection{What Does ``Big Data'' Mean in Developmental Science?}
 
@@
 Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
-What is high volume data to a a developmental psychologist, an archive of 10 terabytes (TB) of video and flat-file data, for example, is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year (\url{http://home.web.cern.ch/about/computing}).
-Similarly, the measure of volume matters.
+High volume data for a developmental psychologist--an archive of 10 terabytes (TB) of video and flat-file data, for example--is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year (\url{http://home.web.cern.ch/about/computing}).
+Similarly, what is ``big'' depends on how volume is measured.
 The Inter-university Consortium for Political and Social Research (ICPSR) (\url{https://www.icpsr.umich.edu}), one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections.
-Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project (\url{http://www.metproject.org}), the large number of files in the repository totaled about 10 TB of digital storage.
+Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project (\url{http://www.metproject.org}), the entire repository totaled about 10 TB of digital storage.
 
 Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?
-The range of phenomena covered by developmental science spans divergent levels of analysis in space, from genes to geography, and time, from microseconds to millenia, all aimed at answering two questions: What develops?
-Why does it develop this way?
+The range of phenomena covered by developmental science spans multiple levels of analysis in space (from genes to geography) and time (from microseconds to millennia) all aimed at answering two questions: ``What develops?'' and  ``Why does it develop this way?''
 In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
-For social and behavioral scientists, the goal is to describe change in the \emph{behavior} or \emph{beliefs} of individuals or groups of individuals and families.
+For social and behavioral scientists, the goal is to describe change in the psychological processes of individuals, groups of individuals, or families.
 Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
-Similarly, the assessment of biological data -- genes, hormone levels, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status -- inform thinking about how these within-person factors influence changes within individuals' behavior across time.
+Similarly, the assessment of biological data--genes, hormones, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status--inform thinking about how within-person factors influence changes behavior.
 
-Consequently, the aspects of data `volume' pertinent to most developmental scientists include the number of participants or families or the number measurement time points.
-Some notable datasets have hundreds or thousands of participants even though samples of tens of participants are more common.
-High volume datasets enable more precise estimation of small effects, especially for qualities or conditions rare among individuals.
-Similarly, the aspects of data `velocity' most relevant to developmental researchers are those related to the frequency or spacing of measurements.
+Consequently, the aspects of data \emph{volume} pertinent to most developmental scientists include the number of participants or families and/or the number of measurement time points.
+Some datasets have hundreds or thousands of participants.
+Such high volume datasets enable more precise estimation of small effects, especially for rare qualities or conditions.
+Similarly, the aspects of data \emph{velocity} most relevant to developmental researchers relate to the frequency or spacing of measurements.
 Velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
-High volume or velocity data may inform the estimation of trends within or between people across time \cite{rietveld_replicability_2014}.
-`Variety' encompasses the range of measurement types employed across developmental research: biological, behavioral, contextual, and cultural/historical/evolutionary and the use of multiple measurements to address the same underlying construct.
-To the ``three Vs'', variability, or the rate of change over time, and complexity, the mutual interdependence of individual measures, might also be added.
+High volume or velocity data may inform the estimation of trends within and between people across time \cite{rietveld_replicability_2014}.
+\emph{Variety} encompasses the range of measurement types employed across developmental research--Biological, behavioral, contextual, and cultural/historical/evolutionary--and the use of different types of measurements to address the same underlying construct.
+In addition to the ``three Vs'', intra-individual \emph{variability} and complexity, or the mutual interdependence of individual measures, might also be included.
 
@@
 Big data may also magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
-Big data also pose theoretical challenges -- e.g, how do micro-scale factors influence macro-scale phenomena?
-Nevertheless, big data offer the possibility of tackling some of the most profound and vexing questions in the field if the relationships among data components can be revealed in ways that do not undermine ethical research principles.
-Greater openness and more widespread sharing of research data than currently exists will also be essential to realize this scientific promise.
+Big data also pose theoretical challenges: For example, how do micro-scale factors influence macro-scale phenomena?
+Nevertheless, big data offer the possibility of tackling some of the most profound and vexing questions in the field if the relations among data components can be revealed in ways that do not undermine ethical research principles.
+Greater openness and more widespread sharing of research data than is current practice will be essential to realize this scientific promise.
+Developmental researchers can look to the recent past for examples of big datasets that have already been collected and in some cases, widely shared for inspiration.
 
-\section{Big datasets in developmental research}
+\section{Big Datasets in Developmental Research}
 
 Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
-Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities like the U.S.
+Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities such as the U.S.
 Census Bureau and the U.S.
 Department of Education.
-The collection and dissemination of these datasets is mandated by law, subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
+The collection and dissemination of these datasets is mandated by law and subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
 Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
@@
 The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
-Finally, there are large scale datasets collected by private entities for business purposes that include sharing data about individual subscribers or users.
-The data collected by private entities about individuals largely concerns consumer behavior, as the data customers are other commercial entities seeking to sell products or services to individual users.
+Finally, some large-scale datasets are collected by private entities for business purposes and may include sharing data about individual subscribers or users.
+The data collected by private entities about individuals largely concern consumer behavior because the data customers are other commercial entities seeking to sell products or services to individual users.
 But, commercial entities are beginning to collect and store health and fitness related activity (e.g., Fitbit, etc.) 
-These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities that provide the services.
+These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the commercial entities that provide the services.
 
@@
 Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
-The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 with a focus on characterizing the health and nutritional status of U.S.
+The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 on the health and nutritional status of U.S.
 infants, children, and adults.
-The annual survey combines interviews about demographic, socioeconomic, dietary and health related topics and physical examinations involving medical, dental, and physiological measurements, as well as laboratory tests.
-The study is cross-sectional and assesses a nationally representative sample of about 5,000 persons located across the country, with 15 counties surveyed annually.
-Many will be familiar with some of NHANES's products such as standardized child growth charts, and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}).
-NYFS collected data on physical activity and fitness levels in U.S. children ages 3 to 15 in 2012 in order to provide an evaluation their overall fitness.
-Much of the NHANES and NYFS data is available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
-However, access to some data are restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers (RDC) locations across the U.S.
-A Google Scholar search in mid-2015 shows more than 110,000 publications mention the NHANES dataset.
+The annual survey combines interviews with a geographically diverse representative sample of about 5,000 people.
+NHANES focuses on demographic, socioeconomic, dietary and health-related topics and physical examinations involving medical, dental, and physiological measurements, and laboratory tests of behavior.
+The products include standardized child growth charts and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}), which reports physical activity and fitness levels in 3 to 15-year-olds
+Much of the NHANES and NYFS data are available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
+However, access to some data is restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers locations across the U.S.
+By mid-2015 more than 110,000 publications had cited the NHANES dataset.
 
-The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) provides an example of a large-scale study that has examined the same cohort -- almost 10,000 American youth born between 1957-64 -- on a annual basis since the study's beginning in 1979.
-Hosted by the U.S. National Bureau of Labor Statistics (BLS), NLSY data are now available for up to 25 survey rounds on questions relating to employment and related issues.
-The NLSY has also been extended to encompass women born in 1922-37 and 1943-55 and men born 1906-21 and 1941-52.
-Most NLSY and NLS data are available free of charge to by means of a web-based data portal to which users must apply for access.
-Access to geographic-related variables requires special permission.
-The Center for Human Resource Research (CHRR) at The Ohio State University and the National Opinion Research Center (NORC) at the University of Chicago administer both the data collection and repository functions.
-A Google Scholar search in mid-2015 shows that more than 20,000 publications cite the NLSY.
+In contrast to the cross-sectional NHANES, The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) has examined the same cohort--almost 10,000 American youth born between 1957 and 1964--on a annual basis since the study began in 1979.
+NLSY data are now available for 25 survey rounds on questions relating to employment and other issues; extensions encompass wider age ranges.
+Most NLSY and NLS data are available free of charge by means of a web-based data portal to which users must apply for access, but access to geographic-related variables requires special permission.
+By mid-2015 more than 20,000 publications had cited the NLSY.
 
-The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) is another large-scale population-based study that has followed the same groups of young people over time focusing on child development, school readiness, and early school experiences.
-The ECLS consists of three cohorts, one focused on a sample of children measured beginning at birth and two cohorts following children from kindergarten.
-The birth cohort a sample of children born in 2001 and followed from birth through kindergarten entry, and unusual among comparable studies includes data from fathers. 
-One of the kindergarten cohorts studied approximately 22,000 children who entered school in 1998-99 and were followe through the eighth grade. 
-A second cohort entered kindergarten in 2010-11 and is being followed through the fifth grade.
-The project has received funding from the National Center for Educational Statistics (https://nces.ed.gov/ecls/), and the data are archived at ICPSR.
-ECLS consists of both publicly available and restricted data.
+The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) is another longitudinal population-based study 
+It focuses on child development, school readiness, and early school experiences in separate cohorts of children followed either from birth or from kindergarten that in some cases exceed 20,000 individuals.
+ECLS consists of both publicly available and restricted data archived at ICPSR.
 A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications.
 
-There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
+Several large government-collected and shared datasets contain information about employment patterns and educational attainment in children and youth.
 The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (\url{http://www.bls.gov/cps/demographics.htm}).
 The U.S.
-Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
-The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu/}).
-TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commercial partners.
-The 2001 TIMSS assessment involved more than 20,000 U.S. students from more than 1,000 schools plus data from more than 60 countries.
-The next TIMSS data collection is in 2015 and will include students in grades 4, 8 and 12.
-Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
-Both studies provide data available for public download.
+Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on U.S.
+primary, secondary, and post-secondary education, and many of the datasets are either public; data with personally identifiable data are available to researchers under a restricted use agreement.
+The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu/}), with more than 20,000 U.S.
+students included in the 2001 sample.
+De-identified data from TIMSS and its sister study, the Progress in International Reading Literacy Study (PIRLS; \url{http://timssandpirls.bc.edu}) are available for public download.
 
-In addition to TIMSS, there are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
-The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
-MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
-Launched in 2011, the Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development.
-Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
-Health outcomes and exposure measurements will continue until the participating children become 13 years old.
-The study has produced a handful of publications, and no data sharing plans have been announced.
-Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
-NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
-Originally focused on improving infant mortality, the study has expanded to focus on factors predicting wellbeing across the lifespan.
-Measures include genetic, health, education, social, and economic factors.
-The study is funded by the British government and data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
+Some large-scale longitudinal studies focused on children's growth, health, and education include participants from outside the U.S.
+The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected physical growth and related data from 8,500 children in Brazil, Ghana, India, Norway, Oman, and the U.S.
+The Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) involves 100,000 parent-child pairs with the goal of evaluating the impact of environmental factors on children's health and development.
+Large-scale birth cohort studies in the U.K. include the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
+NCDS is an ongoing longitudinal study that follows everyone in Great Britain who was born in one particular week in 1958, focusing on factors predicting wellbeing across the lifespan.
+NCDS data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
 The British Cohort Studies have generated more than 3,500 publications to-date.
-Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.
+The Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.
 
-Not all large-scale birth cohort studies succeed, of course.
-A notable failure in the U.S. is the National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
-Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children and their parents from before birth to age 21. 
-However, the NIH Director decided to close the NCS on December 12, 2014 following the recommendations of an advisory panel. 
-Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to the NCS study's closure.
+Not all large-scale birth cohort studies succeed.
+A notable failure is the National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
+Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children prenatally until age 21.
+However, the NIH Director decided to close the NCS in 2014 following the recommendations of an advisory panel. 
+Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to study closure.
 Those data are slated for release in a data archive sometime in 2015.
 
-While no means exhaustive in its coverage, several themes emerge from this sample of U.S. and international population-based studies.
+Several themes emerge from this sample of U.S. and international population-based studies.
 Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
-Some studies involve tens of sample occasions per individual over extended periods of time, and show significant variety in the measures collected.
+Some studies involve dozens of samples per individual over extended periods of time, and show significant variety in the measures collected.
 Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center where special security provisions apply.
-Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
-We turn next to datasets whose collection and management is under the control of individual, usually academic investigators.
+Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact scientific discovery.
+I turn next to datasets whose collection and management is under the control of individual investigators.
 
@@
 Several big data studies in developmental science involve partnerships between government agencies and academic investigators.
-Others involve large-scale data collections that are initiated and managed by individual investigators.
-These projects are diverse in focus and methodology, including population-based studies with a health focus, to behavior genetics, brain imaging, cognition and temperament, and education and employment.
+Others involve large-scale data collections initiated and managed by individual investigators and funded by government or foundation sources.
+These projects are diverse in focus and methods, including population-based studies with a health focus, behavior genetics, brain imaging, cognition and temperament, and education and employment.
 
@@
 
-A population-based example is the Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
+The Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is a population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
 PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
 As of mid-2015, PSID had generated more than 3,900 publications. 
-Much of data is available publicly to registered users and agree to the terms of use conditions that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data.
-In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement between the investigator and the data owner, the University of Michigan.
+The data is available publicly to registered users who agree to the terms of use conditions that ensure participants are not reidentified and that users properly cite and acknowledge use of the data. 
+Some restricted data elements (e.g., geospatial data, school-level identifiers) are only available subject to a formal restricted use agreement between the investigator and the data owner, the University of Michigan.
 
-The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) is another population-based public/private partnership.
-SECCYD prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
-The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
-Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometric and demographic information.
-The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
-Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, with approval granted on a case-by-case basis.
-Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
-The SECCYD has generated more than 214 publications as of mid-2015.
-
-The National Longitudinal Study of Adolescent Health (Ad Health; \url{http://www.nichd.nih.gov/news/releases/Pages/adolescent.aspx}) represents a third example.
-Ad Health began as a survey study undertaken in response to a Congressional mandate designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
-The study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year.
-The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
-The Ad Health cohort has subsequently been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32.
-During Wave IV, biomarker data from a blood test has made available data about glucose, hemoglobin (HbA1c) levels, blood lipids, and related measures. 
+The National Longitudinal Study of Adolescent Health (Ad Health; \url{http://www.nichd.nih.gov/news/releases/Pages/adolescent.aspx}) began as a survey study undertaken in response to a Congressional mandate designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
+Parts of the Ad Health cohort have been followed into young adulthood with four in-home interviews, and in a recent wave, biomarker data from a blood test has been made available. 
 The Ad Health data are archived at ICSPR and are available for public use.
@@
 
-Investigator-initiated and managed, the Family Life Project (FLP; \url{http://flp.fpg.unc.edu}) took a population-based approach to studying the early development of children living in rural, largely poor communities in North Carolina and Pennsylvania.
-A birth cohort of 800 children were studied over the first 3 years of life in a series of home, child care visits and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
-Data collected included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
-Data from Phase I of the FLP are archived at ICPSR, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
+The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
+The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
+Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometrics, and demographic information.
+Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, with approval granted on a case-by-case basis.
+Staff at the National Institute of Child Health and Human Development manage the data access and approval process.
+The SECCYD has generated more than 214 publications as of mid-2015.
+
+Investigator-initiated and managed, the Family Life Project (FLP; \url{http://flp.fpg.unc.edu}) used a population-based approach to study the early development of children living in rural, largely poor communities in North Carolina and Pennsylvania.
+A birth cohort of 800 children were studied in a series of home, child care visits, and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
+Data included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
+Data from Phase I of the FLP are archived at ICPSR, but are subject to restrictions on access due to the sensitive and identifiable nature of the data.
 The FLP site lists about 50 publications through 2013.
@@
 
-Large data samples have long been required in behavior genetics order to detect what can often be small effect sizes.
+Large data samples have long been required in behavior genetics to detect small effect sizes.
 Several large-scale genetics studies that are not specifically developmental deserve mention.
 The Genome of the Netherlands Project(\url{http://www.nlgenome.nl}) is a publicly available dataset from 250 families, consisting of two parents plus one of their adult children.
-The Psychiatric Genomics Consortium (PGC; \url{http://www.med.unc.edu/pgc}) contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
-Many disorders featured (autism, attention-deficit hyperactivity disorder, and schizophrenia) in the PGC dataset have developmental dimensions.
-Genome-wide association study (GWAS) results from the PGC dataset may be viewed in a specialized web browser tool (\url{http://www.broadinstitute.org/mpg/ricopili/}).
+The Psychiatric Genomics Consortium (PGC; \url{http://www.med.unc.edu/pgc}) contains data from more than 170,000 individuals with psychiatric diagnoses or who are at-risk.
+Many disorders in the PGC dataset have developmental dimensions(autism, attention-deficit hyperactivity disorder, and schizophrenia).
+Results may be viewed in a specialized web browser tool (\url{http://www.broadinstitute.org/mpg/ricopili/}).
 
-The U.K.-based Twins Early Development Study (TEDS; \url{http://www.teds.ac.uk/}) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
-Some 13,000 pairs remain actively involved, providing lab-based behavioral task and survey responses.
-DNA samples from more than 5,000 twins pairs are also available.
+The U.K.-based Twins Early Development Study (TEDS; \url{http://www.teds.ac.uk/}) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996, with lab-based behavioral task, survey, and DNA samples from a subset of participants.
 TEDS has generated more than 350 publications to-date.
-Researchers wishing to gain access to the data may do so by submitting a formal data request specifying the aims of the planned research and the variables that needed satisfy the aims. 
+Researchers wishing to gain access to the data may do so by submitting a formal data request specifying the aims of the planned research and the variables needed to satisfy the aims. 
 Proposals that do not overlap with analyses already being planned or carried out by the TEDS research team or other collaborators are usually approved.
 
-The Twin and Offspring Study in Sweden (TOSS; \url{http://ki.se/en/meb/twin-offspring-study-in-sweden-toss}) is another twin-focused study aimed at uncovering genetic and environmental ontributions to measures of family relationships and mental health.
+The Twin and Offspring Study in Sweden (TOSS; \url{http://ki.se/en/meb/twin-offspring-study-in-sweden-toss}) is another twin-focused study aimed at uncovering genetic and environmental contributions to measures of family relationships and mental health.
 TOSS extends the Twin Moms Project to include a sample of twin fathers (320 twin pairs and their families).
@@
 
-Adoption studies provide a unique window on the mix of genetic and non-genetic factors that influence child development.
-Illustrative U.S. examples include the Colorado Adoption Project (CAP), the Nonshared Environment in Adolescent Development (NEAD) project, and the Early Growth and Development Study (EGADS).
-Since its inception in 1976, the Colorado Adoption Project (CAP; \url{http://ibg.colorado.edu/cap}) has enrolled more than 2,400 participants from more than 450 families.
-Data include cognitive abilities, temperament measures, and demographic variables.
+Among behavior genetics studies, those focusing on adoption provide a unique window on the mix of genetic and non-genetic factors that influence child development.
+The Colorado Adoption Project (CAP; \url{http://ibg.colorado.edu/cap}) provides an illustrative example.
+Since its inception in 1976, CAP has enrolled more than 2,400 participants from more than 450 families, collecting data on cognitive abilities, temperament, and demographics.
 CAP has generated more than 200 published articles, and data collection continues.
 Data from 1976-1989 are archived at Dataverse and can be accessed by application.
-NEAD studied a sample of 720 families, including twins, siblings, and their parents, in a longitudinal design that involved up to 3 data collection waves.
-Researchers focused on survey, self-report, and experimenter ratings (from video) of family relationships, adolescent adjustment and competence.
-A Google Scholar search shows 73 publications to-date.
-The data will be shared by request to the project team.
-The team plans to share the data openly later in 2015.
-EGADS is a prospective adoption study of 561 triads of birth parents, adoptive parents, and adopted children in the U.S. 
-The study focuses on genetic and non-genetic factors that influence children’s adjustment from infancy to age 9. 
-The data include self- and other-report measures, DNA, and salivary cortisol.
-Data are available by request, with plans for more open data sharing in the near future.
 
@@
 Several large-scale studies have focused on developmental patterns in brain structure or activity.
-One of the earliest studies of its kind was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
+One of the earliest studies was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
 The study collected structural brain information from 145 4 to 20-year-olds along with behavioral and clinical data.
-More than 100 of the participants were scanned on more than one occasion, with the time between scans approximately two years.
+More than 100 of the participants were scanned on more than once, about two years apart.
 The original paper describing this study has been cited more than 3,300 times, as of mid-2015.
-The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data appear not to be available for analyses by researchers outside the original investigative team.
+The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data are not available for analyses by researchers outside the original investigative team.
 
-The NIH MRI Study of of Normal Brain Development (\url{http://pediatricmri.nih.gov/nihpd/info}) collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
-A second cohort of children from birth to 4 years was scanned longitudinally with up to 10 scans per participant. 
+The NIH MRI Study of Normal Brain Development (\url{http://pediatricmri.nih.gov/nihpd/info}) collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
+A second cohort of children from birth to 4 years was scanned longitudinally with up to 10 scans per child. 
 Demographic, hormonal, cognitive, affective, and psychiatric data were also collected.
@@
 John Richards and colleagues have combined data from the NIH MRI Study, data collected in their own labs, and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI (\url{http://jerlab.psych.sc.edu/NeurodevelopmentalMRIDatabase/}).
-These data can be openly shared with appropriate citation.
+These data can be openly accessed with appropriate citation.
 
-More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING; \url{http://pingstudy.ucsd.edu/}) Project collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from Baltimore, Boston, Honolulu, Los Angeles, New Haven, New York, Sacramento, and San Diego.
+More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING; \url{http://pingstudy.ucsd.edu/}) Project collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from 8 U.S.
+cities.
 Data are available to the research community by application.
@@
 
-Outside of the U.S., The Developing Human Connectome Project (\url{http://www.developingconnectome.org}), led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
+Outside the U.S., The Developing Human Connectome Project (\url{http://www.developingconnectome.org}), led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
 This project aims to study the connectome, a map of human brain connectivity in fetuses from 20 to 44 weeks post-conception.
@@
 Investigators hope that the data set will provide a basis for studying genetic and environmental risks that could lead to neurodevelopmental disorders such as Autistic Spectrum Disorder or Cerebral Palsy.
-The project has already generated 17 publications as of mid-2015.
+The project has generated 17 publications as of mid-2015.
 
-\subsubsection{Language, Cognition, Temperament}
+\subsubsection{Language, Cognition, and Temperament}
 
-The child language community has pioneered the aggregation and sharing of datasets.
-The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established in the behavioral sciences.
+The child language community pioneered the aggregation and sharing of datasets.
+The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established archives in the behavioral sciences.
 It consists of transcripts and audio and video recordings of children's utterances along with recent data from adults with aphasia.
-CHILDES/TalkBank has generated more than ten thousand of citations, according to Google Scholar, and the data are available to the research community, many of them publicly.
+CHILDES/TalkBank has generated more than 10,000 citations, and the datasets are available to the research community, many of them publicly.
 
-Other large-scale sources of language-related data are more recent.
-WordBank (\url{http://wordbank.stanford.edu/}) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
+WordBank (\url{http://wordbank.stanford.edu/}) is an open access archive that consists of data from more than 40,000 samples of the MacArthur-Bates Communicative Development Inventory (CDI).
 Catherine Tamis-Lemonda has released video data (\url{https://nyu.databrary.org/volume/8}) from an NSF-funded longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
-The Human Speechome Project (\url{http://www.media.mit.edu/cogmac/projects/hsp.html}) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
-The project has generated more than 78 publications, but the data have not been made available outside of the original investigative team.
-Data was gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
-The LENA (Language ENvironment Analysis; \url{http://www.lenafoundation.org/}) Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed.
+The Human Speechome Project (\url{http://www.media.mit.edu/cogmac/projects/hsp.html}) at MIT's Media Lab recorded 10 hours of video from one child's home on a daily basis from birth to age three.
+The project has generated more than 78 publications, but the data are not available outside of the original investigative team.
+Data were gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
+The LENA (Language ENvironment Analysis; \url{http://www.lenafoundation.org/}) Foundation created a technology framework that allows children's speech in natural settings to be recorded and analyzed.
 The system provides an automatic language collection and analysis tool for speech language professionals and parents.
@@
 
-In other areas of cognition, a number of measures that have become standards whose widespread adoption has resulted in data that are large in volume, velocity or variety.
-Davida Teller pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
+In other areas of cognition, a number of measures that have become standards and their widespread adoption has resulted in data that are large in volume, velocity or variety.
+Davida Teller \cite{teller1986assessment} pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
 A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.
 
-The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} arose from the Nancy Bayley's Berkeley Growth Study \cite{jones_berkeley_1941} that studied the behavior, motor coordination, physical activity, cognitive function, and maternal behavior of 61 healthy newborn infants.
-The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
-The Bayley is widely used in clinical and epidemiological settings such as the EPICure (\url{http://www.epicure.ac.uk/}) because of the norms even though there are questions about its validity \cite{hack_poor_2005}.
+The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} is widely used in clinical and epidemiological settings such as the EPICure (\url{http://www.epicure.ac.uk/}) because of the published norms even though there are questions about its validity \cite{hack_poor_2005}.
 However, data about those norms is under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
@@
 
-The Wechsler intelligence tests are another widely used standardized instrument.
-The tests were developed by Dr. David Wechsler, beginning in the 1930's. 
-The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, while the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
+The standardized Wechsler intelligence tests were developed in the 1930's. 
+The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, and the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
 The current version (WISC-V) is published by Pearson (\url{http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html}).
-Data about the norms themselves appears not to be available to researchers.
-However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR.
-The data are restricted, and thus are only available by specific authorization.
+Data about the norms are not available to researchers.
+However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is housed at ICSPR and are accessible by specific authorization.
 
@@
 The Infant Behavior Questionnaire (IBQ) is a parent survey measure, and the Laboratory Temperament Assessment Battery (Lab-TAB; \url{http://www.uta.edu/faculty/jgagne/labtab}) is an observational battery, using a set of 3-5 min episodes mimicking situations in everyday life.
-While no archive for IBQ or Lab-TAB data across studies exists, IBQ data on 1,388 participants are archived at ICSPR as part of the Maternal Lifestyle Study (MLS; \url{https://neonatal.rti.org/about/mls_background.cfm}), a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
+IBQ data on 1,388 participants are archived at ICSPR as part of the Maternal Lifestyle Study (MLS; \url{https://neonatal.rti.org/about/mls_background.cfm}), a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
 IBQ and Lab-TAB data are also available from ICPSR from the Family Life Project.
@@
 The 1999 TIMSS video study recorded math and science teaching practices in seven countries, and some of the materials are available for public use through a repository hosted at UCLA (\url{http://www.timssvideo.com/}).
-The Gates Foundation-funded Measures of Effective Teaching (MET; \url{http://www.metproject.org/}) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
+The Gates Foundation Measures of Effective Teaching (MET; \url{http://www.metproject.org/}) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing currently.
 The MET video data are stored and shared at ICSPR.
@@
 The studies involving partnerships with government agencies or those initiated and managed by individual investigators exemplify the high volume, velocity, variety, and variability of big datasets.
-The studies employ a range observational measures, including video recordings, free academic and commercial population-normed test instruments, biological measurements of physiology, brain structure or function.
-On the other hand, the extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.
+The studies employ a range observational measures, including video recordings, population-normed test instruments, biological measurements of physiology, and brain structure or function.
+However, the extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.
 Only some of the datasets are stored in data repositories, for example.
-This means that many large-scale developmental dataset are usually housed locally, on project-specific sites and not generally on centralized servers that aggregate data across studies and sources.
-In many cases, the secondary use of the data by other researchers is controlled by the original investigative team, including the kinds of questions that third parties may ask.
-In some cases, the original investigative team must be included as an author on any publication.
-Perhaps as a consequence of these factors, these studies have by and large generated fewer scientific publications than government-sponsored efforts.
-In a later section, I discuss specific barriers to more widespread data sharing.
+Many large-scale developmental dataset are usually housed locally, on project-specific sites, not on centralized servers that aggregate data across studies and sources.
+In many cases, the original investigative team retains control over the use of data by other researchers, including the kinds of questions that third parties may ask.
+In some cases, the original investigative team must be included as an author on publications.
+Perhaps as a consequence of these factors, these datasets have generated fewer scientific publications than government-sponsored efforts.
 
@@
 Some large-volume sources of developmental data are collected and managed by private, non-academic or government entities.
-More than 1.6 million high school students \cite{lewin_more_2013} take standardized tests or provide financial aid information via measures developed and managed by The College Board or the ACT, both of which are private entities.
+More than 1.6 million high school students \cite{lewin_more_2013} take standardized tests or provide financial aid information via measures developed and managed by The College Board and ACT, Inc.
 The College Board shares Scholastic Aptitude Test (SAT) and college cost and scholarship data with the research community by application.
@@
 Internet-based for-profit service providers operate at an even larger scale.
-Google's Gmail has more than 900 million users worldwide \cite{lardinois_gmail_2015}, while Facebook has more than a billion \cite{facebook_2014}.
-According to the YouGov site (\url{https://yougov.co.uk}), some 17\% of Gmail users in the United Kingdom are 17-24 years of age.
+Google's Gmail has more than 900 million users worldwide \cite{lardinois_gmail_2015}, and Facebook has more than a billion \cite{facebook_2014}.
+According to the YouGov site (\url{https://yougov.co.uk}), 17\% of Gmail users in the United Kingdom are 17-24 years of age.
 Facebook's policies require that users be at least 13 years of age (\url{https://www.facebook.com/help/157793540954833}), but detailed information about user demographics for Facebook or other social media popular among children and adolescents is not openly available.
@@
 Users receive free services in exchange for providing these data.
-Both Google and Facebook have arms that both conduct research and cooperate with academic researchers, but there has been significant public criticism about the ethics of certain research projects \cite{meyer_everything_2014}.
-The primary criticism concerns whether Facebook users had given informed consent to participate in the manipulation of their newsfeeds in the way that researchers whose work is supervised by an IRB would be required to secure if a similar study were taken in a laboratory context.
+Both Google and Facebook have arms that conduct research and cooperate with academic researchers albeit with significant public criticism about the ethics of certain research projects \cite{meyer_everything_2014}.
+The primary criticism concerns whether Facebook users had given informed consent to participate in the manipulation of their newsfeeds as would be required by research ethics boards if a similar study were undertaken in a laboratory context.
 
@@
 Clearly, the collection, analysis, and sharing of large, multilevel datasets has been part of the fabric of developmental science for a long time.
-In this section, we discuss a range of technical, conceptual, and theoretical issues that arise in thinking about the future of big data in developmental science.
+In this section, I discuss a range of technical, conceptual, and theoretical issues that arise in thinking about the future of big data in developmental science.
 
@@
 Developmental scientists collect data from sources representing multiple levels of analysis.
-Increasingly, the devices used to make measurements provide data and metadata in a structured, organized, machine-readable formats.
+Increasingly, measurement devices provide data and metadata in structured, organized, and machine-readable formats.
 
-While some researchers continue to use paper and pencil measures to collect survey information, many universities now have site-licenses for web-based tools like SurveyMonkey and Qualtrics that significantly reduce the amount of manual labor involved in preparing a survey and processing completed data for analysis.
-Behavioral measures involving computer-based tasks are commonly used in developmental research, but most involve the use of custom, project-specific software.
+Although some researchers continue to use paper and pencil measures to collect survey information, many universities now have site-licenses for web-based tools such as SurveyMonkey and Qualtrics that  reduce the manual labor involved in preparing a survey and processing completed data for analysis.
+Behavioral measures involving computer-based tasks are commonly used in developmental research, but most rely on custom, project-specific software.
 So, the output data files, while often in an electronic form, may require significant post-processing to be linked with other data.
-On the other hand, new tools like Amazon's Mechanical Turk (\url{http://www.mturk.com}) or Apple's HealthKit (\url{https://developer.apple.com/healthkit/}) are empowering behavioral and health researchers to conduct large-scale behavioral science experiments using tools specialized for psychological research (e.g., \url{https://psiturk.org}), and with data delivered in well-structured electronic formats.
+New tools like Amazon's Mechanical Turk (\url{http://www.mturk.com}) or Apple's HealthKit (\url{https://developer.apple.com/healthkit/}) are empowering behavioral and health researchers to conduct large-scale behavioral science experiments using tools specialized for psychological research (e.g., \url{https://psiturk.org}), and with data delivered in well-structured electronic formats.
 Amazon's terms of use prohibit minors, but developmental researchers have found ways to secure video-based informed consent from parents to enable their children to participate in looking time studies (\url{https://lookit.mit.edu}) over the web.
 
-The use of video and audio recordings have long been a mainstay of developmental research.
-Video in particular captures the complexity and richness of behavior unlike any other measure, and so video provides a uniquely valuable source of information for many researchers who study behavior in laboratory, classroom, or museum contexts.
-Images and recordings are both larger, in file size, and denser than text or flat-file data, and come in a multiplicity of formats. 
-Thus, developing tools that enable the storage and sharing of images, including brain images, audio, and video data has become a high priority for researchers and research funders .
+Video and audio recordings are a long-standing mainstay of developmental research.
+Video captures the complexity and richness of behavior unlike any other measure, and so video provides a uniquely valuable source of information for researchers who study behavior in laboratory, home, classroom, or museum contexts.
+Images and recordings are larger in file size, denser than text or flat-file data, and come in a diverse formats. 
+Thus, a high priority is the development of tools to enable storage and sharing of images (including brain images), and audio and video data.
 
-Lab-based tools for conducting physiological measurements like EEG, heart-rate or skin conductance, produce electronic files of the time series of measurements.
+Lab-based tools for conducting physiological measurements such as EEG, heart-rate or skin conductance, produce electronic files of these time series.
 Genetic analyses from modern gene sequencing tools and reports from tissue, blood, or salivary samples typically yield machine-readable outputs.
@@
 
-New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promises to make new big data streams about individual participants' locations, physiological states (e.g., \url{https://www.empatica.com/}), activity patterns, and momentary cognitive, and emotional states broadly available to researchers.
-These tools are beginning to enable the collection of data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety available for analysis.
+New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promise to make big data streams about individual participants' locations, physiological states (e.g., \url{https://www.empatica.com/}), activity patterns, and momentary cognitive, and emotional states broadly available to researchers.
+These tools will enable the collection of data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety of data available for analysis.
 
@@
 Developmental researchers who wish to store and share big data face a bewildering array of options.
-These include individual or institutional websites, institutional repositories (e.g., \url{https://scholarsphere.psu.edu/}), cloud services (Dropbox, Box, or Amazon), domain or measure-specific repositories (ICSPR, Databrary.org, TalkBank.org, WordBank.org, OpenfMRI.org), domain general services (Researchgate.net, FigShare/SlideShare, Dataverse, and the Open Science Framework), and open source software web sites like GitHub.
-Some journals offer or require storage of datasets, but these are typically limited to text-based flat-files used for statistical analyses and not raw images, videos, or physiological time series.
+These include individual or institutional websites, institutional repositories (e.g., \url{https://scholarsphere.psu.edu/}), cloud services (Dropbox, Box, or Amazon), domain or measure-specific repositories (ICSPR, Databrary.org, TalkBank.org, WordBank.org, OpenfMRI.org), domain general services (Researchgate.net, FigShare/SlideShare, Dataverse, and the Open Science Framework), and open source software web sites (GitHub).
+Some journals offer or require data storage, but these are typically limited to text-based flat-files used for statistical analyses and do not include raw images, videos, or physiological time series.
 The diversity of storage options can pose daunting challenges for researchers and institutions.
@@
 Storage solutions must meet the needs of researchers during the active data collection phase of a study while not posing insurmountable hurdles to data sharing down the line.
-The effort to reconcile these sometimes competing demands led Databrary (\url{http://databrary.org}), for example, to build tools that allow researchers to upload session-level video and flat-file data to a secure web-based server as it is collected, thus minimizing post-study data curation.
+The effort to reconcile these competing demands led Databrary (\url{http://databrary.org}), for example, to build tools that allow researchers to upload session-level video and flat-file data to a secure web-based server as it is collected, thus minimizing post-study data curation.
 The Open Science Framework (\url{https://osf.io}) offers similar data management functionality for non-identifiable data.
 
-Where and how data are stored is only one part of the problem.
+Where and how data are stored is only part of the problem.
 To foster increased reuse, data must be discoverable and accessible to researchers.
@@
 There are several reasons.
-With some exceptions noted earlier, many research publications do not use data that are readily available to investigators outside of the research team.
-While practices are changing, datasets that are available may lack persistent, citable, identifiers.
-When data from a publication are potentially available to other researchers, access is often restricted and requires a specific, sometimes time consuming application to a data repository or to the original data producer.
-In contrast, Databrary allows researchers access to a library of data under the same access agreement, an innovation aimed at accelerating reuse.
-Another barrier to reuse is the difficulty in finding data that meet specific task or demographic criteria.
-Some repositories like ICPSR and the National Database for Autism Research (NDAR) maintain extensive standardized metadata about tasks and participant demographics.
-This can help investigators search for specific data sources.
+Most research publications do not use data that are readily available to investigators outside of the research team.
+Datasets that are available may lack persistent, citable, identifiers.
+When data from a publication are available to other researchers, access is often restricted and requires a specific, time consuming application to a data repository or to the original data producer.
+In contrast, Databrary allows researchers access to a library of data under a single access agreement, an innovation aimed at accelerating reuse.
+Another barrier to reuse is the difficulty of finding data that meet specific task or demographic criteria.
+Some repositories such as ICPSR and the National Database for Autism Research (NDAR) maintain extensive standardized metadata about tasks and participant demographics.
+This can help investigators to search for specific data sources.
 But, even within domain-specific repositories variable-level searching is not available for all datasets.
-The problems of where to store and how to find and retrieve data will get bigger as datasets grow in size and complexity.
+The problems of where to store and how to find and retrieve data will increase as datasets grow in size and complexity.
 
@@
 Even easy-to-find datasets must be processed prior to analysis.
-Indeed, most data science work involves `janitor work' \cite{lohr_for_2014}.
+Indeed, most data science involves ``janitor work'' \cite{lohr_for_2014}.
 The process of curation involves carefully documenting how raw information from a data stream was transformed into information used in formal analyses. 
 Can the provenance of the data be recorded in ways that others can understand, reproduce, and rely upon?
-For example, physiological data are often filtered, sometimes by the recording devices.
-Video data are usually coded by human observers, and the codes transformed into quantitative measurements.
-What were the variables, units of measurements, calibration properties of the instrument, definition of key terms and codes?
+For example, physiological data are often filtered and smoothed, sometimes by the recording devices.
+Video data are usually coded by human observers and the codes transformed into quantitative measurements.
+What were the variables, units of measurements, calibration properties of the instrument, and definitions of key terms and codes?
 Well-curated datasets usually report these components, but that curation takes time and specialized expertise that many individual investigators lack.
 
-Several new software tools have recently emerged that make it easier for data analysts to produce and reproduce self-documenting data workflows.
-For example, the free RStudio (\url{https://www.rstudio.com}) and Jupyter (\url{https://jupyter.org}) environments make it possible for analysts to create the equivalent of electronic notebooks that combine data, annotations and observations, statistical analyses, and visualizations in human-friendly web or document formats.
-The free, open-source Datavyu (\url{http://datavyu.org}) video coding tool allows automated data analysis and export schemes to be created via the Ruby scripting language.
-Many developmental researchers may be unfamiliar with these sorts of tools, but volunteer groups like Software Carpentry (\url{https://software-carpentry.org}) have begun to provide researchers on-site training in the use of tools for reproducible research workflows, including the use of version control and workflow scripting.
-Similarly, Databrary and the Center for Open Science (\url{http://centerforopenscience.org}) have initiated a series of conference-based and regional workshops to provide hands-on researcher training.
+Several software tools have recently emerged that make it easier for researchers to produce and reproduce self-documenting data workflows.
+For example, the free RStudio (\url{https://www.rstudio.com}) and Jupyter (\url{https://jupyter.org}) environments allow researchers to create electronic notebooks that combine data, annotations and observations, statistical analyses, and visualizations in human-friendly web or document formats.
+The free, open-source Datavyu (\url{http://datavyu.org}) video coding tool allows automated data analysis and export schemes to be created with the Ruby scripting language.
+Many developmental researchers may be unfamiliar with these sorts of tools, but volunteer groups such as Software Carpentry (\url{https://software-carpentry.org}) provide researchers with on-site training in the use of tools for reproducible research workflows, including the use of version control and workflow scripting.
+Similarly, Databrary and the Center for Open Science (\url{http://centerforopenscience.org}) have initiated open office hours and conference-based and regional workshops to provide hands-on researcher training.
 But, the use of tools that produce well-curated, reproducible scientific workflows remains rare among mainstream developmental researchers.
@@
 
-Technical issues will continue to slow progress in many areas of developmental research that are generated and big data.
-Critical challenges include getting data into open, standard, and easily manipulable electronic formats as soon as possible in the research cycle; the development and widespread adoption of data storage platforms or repositories that provide a degree of metadata standardization and enable search and discovery; the creation and adoption of data management practices that make curation part of the typical research workflow; and the creation of a cohort of developmental researchers who have the training and expertise to implement these techniques in their own labs.
-On the other hand, there is demonstrable progress on many of these fronts, and therefore many reasons to be optimistic that the technical challenges can be overcome.
+Technical issues will continue to slow progress in many areas of developmental research that depend on big data.
+Critical challenges include getting data into open, standard, and easily manipulated electronic formats as soon as possible in the research cycle; the development and widespread adoption of data storage platforms or repositories that provide a degree of metadata standardization and enable search and discovery; the creation and adoption of data management practices that make curation part of the research workflow; and the creation of a cohort of developmental researchers who have the training and expertise to implement these techniques in their own labs.
+There is demonstrable progress on many of these fronts, and therefore cause to be optimistic that the technical challenges can be overcome.
 
@@
 
-Clearly the collection, analysis, and interpretation of large scale datasets pose a number questions related to research ethics, participant privacy, and scientific transparency.
-Developmental researchers often collect data from vulnerable populations of research participants.
-Professional ethics require that special care be taken about what data are collected from research participants in general and who gains access to it.
+Clearly the collection, analysis, and interpretation of large scale datasets present issues related to research ethics, participant privacy, and scientific transparency.
+Professional ethics require that special care be taken about what data are collected from research participants and who gains access to it.
 The focus in developmental science on studying vulnerable research populations magnifies these concerns.
@@
 In the U.S., researchers must navigate a regulatory environment in which different types of data are covered under different sections of Federal law.
-For example, the Federal Educational Rights and Privacy Act (FERPA; \url{http://www2.ed.gov/policy/gen/guid/fpco/ferpa/index.html}) governs who has access to a student's educational records.
-The Health Insurance Portability and Accountability Act (HIPAA; \url{http://www.hhs.gov/ocr/privacy/hipaa/understanding/}), among other functions, governs the process by which individually identifiable health information may be disclosed and to whom.
+For example, the Federal Educational Rights and Privacy Act (FERPA; \url{http://www2.ed.gov/policy/gen/guid/fpco/ferpa/index.html}) governs access to student educational records.
+The Health Insurance Portability and Accountability Act (HIPAA; \url{http://www.hhs.gov/ocr/privacy/hipaa/understanding/}) governs the disclosure of individually identifiable health information may be disclosed and to whom.
 The Code of Federal Regulations (CFR) Title 45 (\url{http://www.hhs.gov/ohrp/humansubjects/guidance/45cfr46.html}) governs research with human participants.
 If the data in question are audio or video recordings, different state provisions can come into play.
-For example, in two-party states (\url{http://www.dmlp.org/legal-guide/recording-phone-calls-and-conversations}) both the person making the recording and the person or persons being recorded must consent, making some forms of data collection using recordings problematic, at best.
+For example, in two-party states (\url{http://www.dmlp.org/legal-guide/recording-phone-calls-and-conversations}) both the person making the recording and the person(s) being recorded must consent, making some forms of data collection using recordings problematic.
 
-Research activities funded by the U.S.
-Federal Government, typically in colleges, universities, or academic or private medical centers, are supervised by an Institutional Review Board or its equivalent.
+Research activities funded by the U.S. Federal Government are supervised by an Institutional Review Board or its equivalent.
 These entities are regulated by the U.S. Office of Health and Human Services (\url{http://www.hhs.gov/ohrp/}).
-Researchers supervised by IRBs must respect participants' privacy, secure informed consent or assent, and maintain confidentiality. 
-Adhering to these ethical principles has practical consequences for research.
-Researchers have limits on the ways they are able to recruit and identify participants.
+Researchers supervised by IRBs must respect participants' privacy, secure informed consent, and maintain confidentiality. 
+These ethical principles have practical consequences for research.
+They limit the ways researchers may recruit participants.
 Minors must give informed assent to participate in a study with a parent or guardian giving formal consent.
-Data must be collected in ways that minimimize the likelihood that information given by a participant will be disclosed or a participants' identity revealed outside the research team.
+Data must be collected in ways that minimize the likelihood that information given by a participant will be disclosed or a participants' identity revealed outside the research team.
 This usually means that data items that could reveal a participant's identity are removed or altered to reduce the likelihood of disclosure.
-Whether `de-identified' data of this sort can be shared with researchers outside the original team that collected depends on several factors.
+Whether deidentified data of this sort can be shared with researchers outside the original team that collected depends on several factors.
 One factor is the sensitivity of the data collected and the likelihood that specific identities or home locations or other items could be revealed.
-Another factor concerns whether participants were informed that de-identified data might be shared outside the research team.
-IRBs may view these matters differently, creating additional complexities.
-Some IRBs may require participants to be informed that de-identified data might be shared outside the IRB-approved research team, and others may deem the analysis of de-identified data no longer meets the definition of human subjects research.
+Another factor concerns whether participants were informed that deidentified data might be shared outside the research team.
+IRBs may view these matters differently, creating additional complexities for large-scale projects that span geographic areas.
+Some IRBs may require participants to be informed that deidentified data might be shared outside the IRB-approved research team, and others may deem that the analysis of deidentified data no longer meets the definition of human subjects research.
+Many big datasets in developmental science have restrictions on access either because the data were collected under Federal regulations that prohibit releasing identifiable data or because the participants were not asked for permission to share data with other researchers.
 From a big data perspective, if data cannot be shared outside the original IRB-approved research team, then the possible analyses are restricted to the interests, resources, and expertise of that team.
 
-Of course, some data types like photographs, audio or video recordings contain identifiable information that cannot be removed or altered without reducing the value to others.
+Of course, some data types like photographs and audio or video recordings contain identifiable information that cannot be removed or altered without reducing the value to others.
 Thus, data from photographs or recordings requires additional consideration and special care.
@@
 Databrary has created template language to help researchers gain participants' permission.
-Further, Databrary restricts access to its identifiable data to researchers who have formally agreed to uphold ethical research principles and whose institutions approve of their access.
+Furthermore, Databrary restricts access to identifiable data to researchers who have formally agreed to uphold ethical research principles and whose institutions approve of their access.
 The notion that research participants can consent to share identifiable or potentially identifiable data is relatively new.
 The Personal Genomes Project (\url{http://www.personalgenomes.org}), Open Humans Project (\url{https://www.openhumans.org}), and Human Connectome Project (\url{http://www.humanconnectomeproject.org}) embody similar principles.
-The experience of Databrary investigators is that a significant proportion -- more than 75 percent -- of research participants and their parents or guardians will consent to sharing identifiable data, mostly video, with other members of the research community.
+The experience of Databrary investigators is that a significant proportion of research participants and their parents or guardians will consent to sharing identifiable data, mostly video, with other members of the research community.
 
-It is too early to predict whether it will become commonplace for academic developmental researchers to seek explicit permission to share research data that includes identifiable elements, with other researchers.
-There are reasons to be optimistic, however.
-In just over a year of operation, Databrary has secured formal agreements with more than 75 institutions in North and South America, Europe, Australia, and Asia allowing more than 100 researchers to access identifiable data.
-On the other hand, some leading developmental researchers have argued that the families of research participants forge a relationship of trust with a particular research team, formalized through the informed consent document \cite{eisenberg_thoughts_2015}.
+It is too early to predict whether it will become commonplace for academic developmental researchers to seek explicit permission to share identifiable research data with other researchers.
+But, there are reasons to be optimistic.
+In just over a year of operation, Databrary has secured formal agreements with more than 80 institutions in North and South America, Europe, Australia, and Asia allowing more than 140 researchers to access identifiable data.
+However, some leading developmental researchers have argued that the families of research participants forge a relationship of trust with a particular research team, formalized through the informed consent document \cite{eisenberg_thoughts_2015}.
 The relationship might be harmed or the research project negatively affected if participants were asked to share data with other researchers.
-Sensitive to the latter argument, Databrary has recommended that permission to share be sought separate from consent to participant in research and after a given data collection has ended.
-Moreover, the fact that a significant proportion of families agree to share when asked suggests that the relationship of trust involved in research participation might be extended to a community of researchers, given suitable provisions and constraints.
+Sensitive to the latter argument, Databrary has recommended that permission to share be sought separately from consent to participant in research and after a given data collection has ended.
+Moreover, the fact that most families agree to share when asked suggests that the relationship of trust involved in research participation might be extended to a community of researchers, given suitable provisions and constraints.
 Undoubtedly, seeking explicit permission to share on a consistent and widespread basis would resolve any ambiguity about whether a given dataset can be shared with whom and for what sort of purposes.
 
-Greater transparency and more explicit clarification about what data is being collected and for what purposes could be sought from commercial entities, as well.
+Greater transparency and more explicit clarification about what data is being collected and for what purposes could be sought from commercial entities as well.
 Social media companies like Google, Facebook, Twitter, SnapChat, and Instagram have business models that involve the collection, mining, and packaging of data, usually to advertisers, in exchange for services that are free to users.
-While some services attempt to restrict the youngest ages at which users can create accounts, the limits often lack rigor, and there is no parallel to the requirement of adult consent required in formal research contexts.
-The data collection and analyses carried out by private entities are subject to no supervision or formal regulation comparable to academic researchers.
-Instead, data use, analysis, and sharing provisions are governed by terms of use agreements users must indicate agreement with, usually by clicking a button, prior to using a given service.
+Although some services attempt to restrict the ages at which users can create accounts, the limits often lack rigor, and there is no parallel to the requirement of adult consent required in formal research contexts.
+The data collection and analyses carried out by private entities are subject to no supervision or formal regulation comparable to academic research.
+Instead, data use, analysis, and sharing provisions are governed by terms of use agreements that users acknowledge by clicking a button prior to using a given service.
 Unlike academic settings, where violations of research ethics principles may involve significant consequences for the researchers and institutions, violations of commercial terms of use require aggrieved parties to seek redress through litigation.
@@
 
-Research ethics continue to evolve, but several aspects of the current state-of-the art stand out.
-Many big data sets in developmental science have restrictions on access either because the data were collected under Federal regulations that prohibit releasing data that could identify individuals or because the participants were not asked for permission to share individual-level data with other researchers.
-These restrictions impede the secondary reuse of existing dataset.
-New models for gaining permission from participants or their families to share identifiable data on a restricted basis with the research community have been developed and are being implemented in a growing number of laboratories.
-The widespread adoption of these practices promises to expand the availability of big data for reuse in the future.
-
-Unresolved issues that could impact the availability of big data in the future include whether linkage across streams increases the risks of reidentification, whether it is essential to reconsent minors when they become adults, a notion most researchers find totally impractical and a significant barrier to data sharing, and a general concern about the ethics of granting consent to share data for an indefinite period.
-Since data security can't be guaranteed, risks can only be minimized and managed, but not eliminated.
-Finally, there are unresolved questions about privacy protections in the consumer domain that have the potential to influence academic research.
+Unresolved issues that could impact the availability of big data in the future include whether linkage across streams increases risk of reidentification, whether it is essential to reconsent minors when they become adults, a notion most researchers find totally impractical and a significant barrier to data sharing, and a general concern about the ethics of granting consent to share data for an indefinite period.
+Because data security cannot ever be guaranteed, risks can be minimized and managed, but not entirely eliminated. 
+Finally, there are unresolved questions about privacy protections in the consumer domain that have the potential to influence academic research
 
@@
 
-Another dimension of scientific ethics concerns transparency and reproducibility.
-The social and behavioral sciences have seen a string of high profile cases of scientific misconduct in recent years, including cases of fraudulent data \cite{singal_case_2015, bhattacharjee_diederik_2013}.
+An important dimension of scientific ethics concerns transparency and reproducibility.
+The social and behavioral sciences have incurred an unfortunate string of high profile cases of scientific misconduct in recent years, including cases of fraudulent data \cite{singal_case_2015, bhattacharjee_diederik_2013}.
 The credibility problem is magnified by several factors.
-That lack of power and unrestricted exploratory analyses may mean that most research findings are false \cite{ioannidis_why_2005}.
-The actual effect sizes of published findings is unknown due to a bias toward publishing positive results.
-Most journals reject papers that report failures to replicate published findings, and as a result, few scientists attempt or are recognized and rewarded for conducting genuine replications \cite{nosek_scientific_2012}.
-The problem is so serious that some have claimed that science as a whole faces a crisis of replicability.
+Lack of power and unrestricted exploratory analyses may mean that most research findings are false \cite{ioannidis_why_2005}.
+The actual effect sizes of published findings are unknown due to a bias toward publishing positive results.
+Most journals reject papers that report failures to replicate published findings, and as a result, few scientists attempt replications or are recognized and rewarded for doing so \cite{nosek_scientific_2012}.
+The problem is so serious that some have claimed that science as a whole faces a crisis of reproducibility.
 
-To address the crisis, the Center for Open Science has organized several large-scale replication efforts, including some in psychological science under the umbrella of the ``Many Labs'' project (\url{https://osf.io/ct89g/}; \url{https://osf.io/8cd4r/}).
-The results of these pre-registered, open, large sample replications have been mixed.
-Some published effects clearly replicated, but others did not.
+To address this problem, the Center for Open Science has organized several large-scale replication efforts, including some in psychological science under the ``Many Labs'' project (\url{https://osf.io/ct89g/}; \url{https://osf.io/8cd4r/}).
+The results of these pre-registered, open, large sample replications have been mixed \cite{collaboration_estimating_2015}.
+Some published effects were replicated, but others were not.
 
-Whether there are replicability problems developmental science and whether they constitute a crisis is not known.
-Developmental research reflects the same positive effects biases seen in other fields, and the same problem that null results often sit unpublished in file drawers -- the so-called file drawer effect \cite{rosenthal_file_1979}.
-No failures to replicate developmental studies have been reported to Psychfiledrawer.org (\url{http://psychfiledrawer.org}), a resource designed to bring replications failures to light.
+Whether there are replicability problems exist in developmental science and whether they constitute a crisis is unknown.
+Developmental research reflects the same positive effects biases seen in other fields, and the same problem that null results often sit unpublished in file drawers--the so-called file drawer effect \cite{rosenthal_file_1979}.
+No failures to replicate developmental studies have been reported to Psychfiledrawer.org (\url{http://psychfiledrawer.org}), a resource designed to bring replication failures to light.
 As some developmental researchers have written \cite{bishop-blog-2012}, replicating effects with developmental populations can be especially difficult and so even partial replications are noteworthy.
-No large-scale replication efforts in developmental science have been mounted, but there have been calls for changes in journal practices to give replications have a more privileged place in scientific publications \cite{bishop-blog-2012}.
-One barrier to more open data practices appears to be researcher's fears of having one's reputation or abilities publicly undermined \cite{ascoli_ups_2006}.
+No large-scale replication efforts in developmental science have been mounted, but there have been calls for changes in journal practices to give replications a more privileged place in scientific publications \cite{bishop-blog-2012}.
+One barrier to more open data practices appears to be researcher's fears of having their reputation or abilities publicly undermined \cite{ascoli_ups_2006}.
 So, changing views about replication may require shifts in the scientific culture.
@@
 Still, the increasing availability of large-scale datasets about developmental questions promises to magnify problems at the intersection between exploratory and confirmatory research.
-Large volume, higher velocity or variety datasets make it possible to explore and discover novel unpredicted patterns in data.
+Large volume, high velocity, and high variety datasets make it possible to explore and discover novel unpredicted patterns in data.
 But, novel findings might be spurious, and exploratory findings must be properly confirmed.
-While pre-registration and pre-review have been suggested as one way to address the problem of spurious exploratory findings, these tools may not be practical in all cases and could have a chilling effect on discovery.
+Whereas pre-registration and pre-review have been suggested as one way to address the problem of spurious exploratory findings, these tools are not practical in all cases and could have a chilling effect on discovery.
 In contrast, increased transparency about the process that led to an exploratory finding and the steps taken to confirm it can bolster a finding's credibility.
@@
 
-Developmental researchers have clearly shown enthusiasm for sharing with the research community and the public at large the results of their findings via publications, and in some subfields the sharing of data, materials and methods is firmly established.
-Open sharing practices tend to be more common when there is a high cost, centralized source of scientific data that could not conveniently be owned or managed by researchers themselves (e.g., space telescopes or the U.S.
+Developmental researchers have clearly shown enthusiasm for sharing the results of their findings via publications, and in some subfields, the sharing of data, materials and methods is firmly established.
+Open sharing practices tend to be more common when there is a high cost, centralized source of scientific data that could not conveniently be owned or managed by individual researchers (e.g., space telescopes or the U.S.
 Census).
-Successful big data projects -- repositories or specific studies or datasets -- appear to have fostered a high level of engagement around a particular resource, often through specific training, and targeted workshops.
 
 In addition to bottom-up/grassroots initiatives, journals and funding agencies continue to play a vital role in creating an impetus for change.
-Funders can require data management plans and mandate that data and research products be deposited into particular types of open repositories as the Gates Foundation has done, and provide funding to build and support big data infrastructure.
-Journals can also require that data be deposited in open archives as a condition of publication in addition to adopting other transparent and open science practices for manuscripts they accept.
-However, the problem with data sharing mandates from funders is that there is no specific mechanism to provide sufficient ongoing financial support for data archives, and few researchers budget funds to support data management and archiving.
-And, while some journals are willing to take on the burden of storing and sharing data associated with publications, others \cite{maunsell_announcement_2010} have gone the other way and refuse to accept supplemental materials.
-Thus, in the interests of promoting greater openness and transparency, funders and journals may create unfunded mandates that make it harder for researchers to make discoveries.
-For example, new regulations specifying when data must be deposited may be unwieldy and impractical for some developmental scientists to carry out their work \cite{eisenberg_thoughts_2015, apa_data_sharing_work_group_data_2015}.
+Funders can require data management plans and mandate that data and research products be deposited into particular types of open repositories and provide funding to build and support big data infrastructure.
+Journals can also require that data be deposited in open archives as a condition of publication in addition to adopting other transparent and open science practices for manuscripts they accept (e.g.
+PLoS).
+However, the problem with data sharing mandates from funders is that there is no specific mechanism to provide sufficient ongoing financial support for data archives.
+Few researchers budget funds to support data management and archiving.
+Some journals are willing to shoulder the burden of storing and sharing data associated with publications, but others refuse to accept supplemental materials of any kind \cite{maunsell_announcement_2010}.
+Thus, in the interest of promoting greater openness and transparency, funders and journals may create unfunded mandates that make it harder for researchers to make discoveries.
+For example, new regulations specifying when data must be deposited may be unwieldy and impractical for developmental scientists to carry out their work \cite{eisenberg_thoughts_2015, apa_data_sharing_work_group_data_2015}.
 
-These issues are complicated by lack of consensus about who `owns' research data \cite{who_owns_research_data}.
-Federal funding agencies might argue that the public should `own' data paid for by tax dollars, much like other data collected by government agencies like the U.S.
+These issues are complicated by lack of consensus about who \emph{owns} research data \cite{who_owns_research_data}.
+Federal funding agencies might argue that the public should own research data paid for by tax dollars, much like other data collected by government agencies such as the U.S.
 Census, National Weather Service, and U.S.
@@
 The institutions that employ, receive, and manage federal grants might stake a claim to ownership.
-Most investigators naturally feel a strong sense of ownership over their intellectual products, even though formal copyright is often surrendered in the process of publishing, and that sense extends to data.
+Most investigators naturally feel a strong sense of ownership over their intellectual products, although formal copyright is often surrendered in the process of publishing, and that sense extends to data.
 Some have even argued that research participants themselves own their own data, and there are new business models emerging that may soon provide individuals an opportunity to sell data for personal gain (\url{http://www.datawallet.io}).
@@
 Others grant access to data only if co-authorship on any published product is guaranteed.
-While legimate arguments might be made in favor of `embargo' periods that enable teams of researchers to mine and report findings from their research efforts, the ideal of fostering greater data reuse argues for shorter periods.
+Although legitimate arguments might be made in favor of embargo periods that enable teams of researchers to mine and report findings from their research efforts, the ideal of fostering greater data reuse argues for the shortest possible periods.
 Establishing consensus about data ownership and the kind of control investigators can exercise over it will require conversations among researchers, institutions, and funding agencies.
@@
 One challenge that big data may help to address is a particular bias in existing samples.
-Developmental research presumes to study what is normative about changes across time in human behavior.
-But, much of what behavioral and social science has learned about developmental processes comes from samples that represent only a small fraction of the world's population.
+Developmental research typically purports to study what is normative about changes across time in human behavior.
+But, much of what we have learned about developmental processes comes from samples that represent only a small fraction of the world's population \cite{karasik2010weird, fernald2010getting}
 Developmental psychology, like other branches of the field, presents findings from Western, education, industrialized, rich, and democratic (WEIRD) societies \cite{henrich_weirdest_2010}.
-So, to the extent that new tools enable research on development in non-WEIRD cultures and those data can be aggregated and combined, the extent to which developmental science can make claims about universal or near-universal components of developmental processes will be strengthened.
-On the other hand, developmental researchers are well aware of cohort effects -- the notion that developmental processes highly influenced by changing social and cultural norms are themselves subject to change.
-Thus, even the most culturally diverse dataset possible may still yield conclusions that are `locked' in time.
+So, to the extent that new tools enable research on development in non-WEIRD cultures and those data can be aggregated and combined will strengthen the ability to make claims about universal or near-universal components of developmental processes.
+However, developmental researchers are well aware of cohort effects--the notion that developmental processes can be influenced by changing social and cultural norms.
+Thus, even the most culturally diverse dataset possible may still yield conclusions that are locked in time.
 
 A second challenge larger datasets may help to address is the fact that most social, behavioral \cite{maxwell_persistence_2004} and neuroscience studies \cite{button_power_2013} are underpowered.
-Most worryingly it has been claimed that most published research findings are false in fields with specific characteristics: Sample or sizes are small; more relationships between variables are tested; preselection of tested relationships is less common and exploratory research more common; designs, definitions, outcomes, and analytical modes are less consistent across studies; and when more teams are involved seeking out significant effects \cite{ioannidis_why_2005}.
-The collection, analysis, and sharing of larger size datasets may help to ameliorate some, but not all, of these flaws.
+Most worryingly many published research findings are false in fields that rely on small sample sizes, test multiple relationships between variables, engage in exploratory research, use diverse research designs, definitions, outcomes, and analytical modes across studies, and when more labs seek out significant effects \cite{ioannidis_why_2005}.
+The collection, analysis, and sharing of larger datasets will help to ameliorate these factors.
 
 Developmental research faces a specific point of tension related to the measurement.
-Many of the measures for which high volume data are available come from proprietary, expensive to license instruments like the Bayley, WIPPSI.
-Where free, academic instruments are in use like the Infant Behavior Questionnaire, there remains no centralized archive for those data.
+Many of the measures for which high volume data are available come from proprietary, expensive instruments such as the Bayley and the WIPPSI.
+Free, academic instruments such as the Infant Behavior Questionnaire have no centralized data archive.
 Plus, the measures themselves have been revised several times, making it more challenging to compare data collected using different versions, especially across time.
 Similar problems arise when non-proprietary tasks are used.
-Most investigators customize even a well-known task to make it suitable for use with children, and the sharing of research materials is not substantially more widespread than the sharing of data.
-Efforts to encourage researchers to capture and record the conceptual structure of psychological tasks have been undertaking (e.g., The Cognitive Atlas), but they have not been widely adopted even by the target community of researchers in adult cognitive neuroscience.
+Most investigators customize even a well-known task to make it suitable for use with children, and the sharing of research materials is just as limited as the sharing of data.
+Efforts to encourage researchers to capture and record the conceptual structure of psychological tasks have been undertaken (e.g., The Cognitive Atlas; \url{http://www.cognitiveatlas.org}), but have not yet been widely adopted.
 
-Finally, some have raised concerns that the rise of especially large datasets means the `end of theory.' \cite{anderson_end_2008}
-In a provocative essay in 2008, Anderson argued that large quantities of data mean the traditional model of scientific inquiry involving hypothesis testing will soon give way to model-free descriptions of data.
+Finally, some critics have raised concerns that the rise big data means the ``end of theory'' \cite{anderson_end_2008}.
+In a provocative essay Anderson \citeyear{anderson_end_2008} argued that large quantities of data mean the traditional model of scientific inquiry involving hypothesis testing will soon give way to model-free descriptions of data.
 Others note that bigger data don't necessarily lead to deeper insights \cite{graham_big_2012}.
-Boyd and Crawford note \cite{boyd_critical_2012} that beyond its scientific dimensions, `big data' is a sociocultural phenomenon that raises as many questions as it promises answers.
+Boyd and Crawford \cite{boyd_critical_2012} note that beyond its scientific dimensions, big data is a sociocultural phenomenon that raises as many questions as it promises answers.
+Developmental science has a rich and rigorous intellectual history in which theory and experiment play central roles.
+That tradition is likely to continue.
 
@@
 
-As boyd and Crawford \citeyear{boyd_critical_2012} note, ``The era of Big Data has begun.
-Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.''
-The clamor extends to the developmental and learning sciences where interests in understanding how to maximize the potential for human achievement and improve health have significant consequences for daily living.
+As boyd and Crawford \citeyear{boyd_critical_2012}  ``The era of Big Data has begun.
+Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.'' (p.
+662).
+The clamor extends to the developmental and learning sciences where improving health and maximizing the potential for human achievement have significant consequences for daily living.
 
 This review has shown that the collection, dissemination and analysis of data sets that are big in volume, velocity, or variety has a long and established history in developmental science.
-Many `big data' studies have had substantial impact on scholarship, as measured by the quantity of published findings, and in some cases, on public understanding and policy.
-For the most part, studies that have had the largest impact have been ones funded by and managed by government entities, perhaps in part to the widespread access to data for reuse by others.
+Many `big data' studies have had substantial impact on scholarship, and in some cases, on public understanding and policy.
+For the most part, studies with the largest impact (as measured by the quantity of published papers) have been ones funded by and managed by government entities, perhaps due to widespread access to data for reuse.
 Those investigator-initiated and managed projects that have had the largest intellectual impacts have created significant intellectual communities around the datasets, communities that extend the beyond the boundaries of the original investigative teams.
-While technical issues about data formats, storage, cleaning, provenance, visualization and provenance remain, there is significant progress.
-Developmental researchers have available a growing array of well-established data repositories (ICSPR, Dataverse) and new data storage/management tools (Databrary, OSF).
+Although technical issues about data formats, storage, cleaning, visualization and provenance remain, there is significant progress.
+Developmental researchers have available a growing array of well-established data repositories (CHILDES, Databrary, Dataverse, ICPSR) and new data storage/management tools (Databrary, OSF).
 Research and data management practices have begun to converge on norms that will reduce the costs of preparing data for sharing in the future.
-New ethical procedures for seeking informed consent to share identifiable data have been developed and are being implemented in diverse research contexts.
+New ethical procedures for seeking informed consent to share identifiable data have been developed and are being implemented in diverse research contexts that promise to accelerate future reuse.
 
-On the other hand, there are many barriers to progress in the analysis of big datasets in developmental science.
+On the other hand, many barriers remain.
 Most developmental science data are hard to find, and cumbersome to access, even for researchers.
@@
 This means that virtually all of the data from unpublished studies remains unavailable, making the size of the file drawer effect unknown.
-Most investigators do not currently employ workflows that make it easy to share data or document analysis pathways.
+Most investigators do not currently employ workflows that make it easy to share data or to document analysis pathways.
 With rare exceptions clustered around specific datasets there is no widespread culture of data sharing, and indeed some level of bias against the use of secondary data.
 Finally, there is no unified understanding or consensus within developmental science about who owns research data, whether it is essential or merely wise to share data, and when in the research cycle data should be shared.
+These factors limit the potential for discovery that the era of big data so seductively promises, especially in the face of what can seem a glacial pace of change in scientific culture.
 
-At the same time, while scientific culture may appear to change slowly, remember that Facebook was launched in 2004, Twitter in 2006, and the iPhone in 2007.
+Nevertheless, we should remember that Facebook was launched in 2004, Twitter in 2006, and the iPhone in 2007.
 It would be unwise to underestimate the speed with which new technologies, tools, and cultural practices can change.
-If developmental researchers can find ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development, we can look forward to a future rich in theory and understanding.
+If developmental researchers can find ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development, as many are beginning to do, we can look forward to a future rich in theory and understanding.
 

### 2015-09-03 wiley-big-data-devel/big-data.tex
submitted manuscript

@@
 %\usepackage{hyperref}
-
 % \hypersetup{breaklinks=true}
@@
 
-\abstract{The use of the term ``big data'' has grown substantially over the past several decades, and is now widespread. 
-In this review, I ask what makes data ``big'' and the implications of size, density, or complexity of datasets for the science of human development.
-A survey of existing data sets illustrates how large, complex, multi-level, multi-measure data have been used to reveal the complexities of human development.
-``Big data'' poses significant technical, policy, ethics, transparency, cultural, and conceptual issues that must be addressed.
-Most big developmental science data are currently hard to find and cumbersome to access; the field lacks a culture of data sharing; and there is no consensus about who owns or should control research data.
-But, these barriers are rapidly dissolving.
-Developmental researchers are finding new ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development.
-This promises a future in which ``big data'' can lead to deep insight about some of the most vexing questions in behavioral science.
+\abstract{The use of the term ``big data'' has grown substantially over the past several decades and is now widespread. 
+In this review, I ask what makes data ``big'' and what implications of size, density, or complexity of datasets have for the science of human development.
+A survey of existing data sets illustrates how existing large, complex, multi-level, and multi-measure data can reveal the complexities of developmental processes.
+At the same time, significant technical, policy, ethics, transparency, cultural, and conceptual issues associated with the use of big data must be addressed.
+Most big developmental science data are currently hard to find and cumbersome to access, the field lacks a culture of data sharing, and there is no consensus about who owns or should control research data.
+But, these barriers are dissolving.
+Developmental researchers are finding new ways to collect, manage, store, share, and enable others to reuse data.
+This promises a future in which big data can lead to deeper insights about some of the most profound questions in behavioral science.
 }
@@
 \section{Introduction}
-
-A search on the term ``big data'' yields more than 49 million hits on Google, more than 100,000 results on Google Scholar, and 13 million hits on Bing.
+A search on the term ``big data'' yields more than 49 million hits on Google (\url{http://www.google.com}), more than 147,000 results on Google Scholar (\url{http://scholar.google.com}), and 14 million hits on Bing (\url{http://bing.com}).
 The results return in less than a second.
 A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term around 1900 with a steady rise in frequency from the 1950s to around 2000.
-Clearly, as measured by search engine matches of electronic documents on the Internet, scholarly documents, or digitized books, the use of the term ``big data'' has grown substantially over the past several decades, and is now widespread.
+Clearly, as measured by search engine matches of electronic documents on the Internet, scholarly documents, or digitized books, the use of the term has grown substantially over the past several decades, and is now widespread.
 Moreover, the fact that these basic statistics about a particular phrase can be determined in an instant speaks to the rapid progress in networked computing, search engines, and databases.
 Most of the tools that enable it have been created in the last 20 years.
-``Big data'' has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.
+Big data has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.
 
-In this review, I show how increased availability of and interest in big data sets promises to alter the study of human development.
+In this review, I show how the increased availability of and interest in big data sets promises to alter the study of human development.
 I begin by asking what makes data ``big'' and the implications of size, density, or complexity of datasets for understanding human development.
-Then, I survey case studies to illustrate how large, complex, multi-level, multi-measure data sets have been used to reveal the complexities of human development.
-I conclude by discussing key questions that ``big data'' approaches pose for the future of developmental science pose.
+Then, I survey case studies to illustrate how large, complex, multi-level, multi-measure data sets have been used to reveal some of the complexities of developmental processes.
+I conclude by discussing key questions that big data approaches pose for the future of developmental science.
 
-I disagree with the claim that an era of ``big data'' signals the end of theory \cite{anderson_end_2008}.
-Nor are ``big data'' analyses especially new in developmental science.
-Although some claims about ``big data'' revolutionizing scientific understanding are exaggerated \cite{boyd_critical_2012}, developmental science tackles questions that can benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
-Significant novel insights into human development emerging from the era of ``big data'' depend on the size, density, and complexity of the datasets, on how widely and openly the data are shared, and on how readily researchers are able to combine or link datasets across levels of analysis.
-In turn, these specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual researcher behavior, publisher priorities, and research funding levels.
-Technology may accelerate the ``big data'' era, but the challenges it poses turn out to be relatively unimportant for advancing research in developmental psychology.
+In fact, big data analyses are not especially new in developmental science.
+The field tackles questions that benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
+Big data does not signal the end of theory \cite{anderson_end_2008}, nor will it necessarily revolutionize scientific understanding \cite{boyd_critical_2012}. 
+Rather, significant novel insights emerging from the era of big data will depend not just on the size, density, and complexity of the datasets, but on how widely and openly data are shared, and on how readily researchers are able to combine or link datasets across levels of analysis.
+These specific innovations depend largely on small, probably manageable, but nonetheless thorny problems related to policy, scientific culture, individual researcher behavior, publisher priorities, and research funding levels.
+Thus, technology may accelerate the big data era, but the challenges it poses may turn out to be less important for advancing research in developmental psychology than changes in scientific culture.
 
 \subsection{What Does ``Big Data'' Mean in Developmental Science?}
-
 According to Laney \citeyear{laney01controlling3v} the volume, velocity, and variety of data streams make data big.
-Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of a specific research context.
-High volume data for a developmental psychologist--an archive of 10 terabytes (TB) of video and flat-file data, for example--is tiny to a physicist working on the Large Hadron Collider (LHC) that generates 30 petabytes in a year (\url{http://home.web.cern.ch/about/computing}).
-Similarly, what is ``big'' depends on how volume is measured.
-The Inter-university Consortium for Political and Social Research (ICPSR) (\url{https://www.icpsr.umich.edu}), one of the largest data repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections.
+Of course, general statements about the total quantity of data generated per day \cite{ibm_2015} make little sense outside of specific research contexts.
+High volume data for a developmental psychologist---an archive of 10 terabytes (TB) of video and flat-file data, for example---represents a tiny fraction of the 30 petabytes per year (\url{http://home.web.cern.ch/about/computing}) available to a physicist working on the Large Hadron Collider (LHC).
+Similarly, what constitutes big depends on how one measures volume.
+The Inter-university Consortium for Political and Social Research (ICPSR) (\url{https://www.icpsr.umich.edu}), one of the largest and oldest repositories for data from the social sciences, consists of more than 500,000 files in 16 specialized data collections.
 Yet, until the recent acquisition of video data from the Gates Foundation-funded Methods of Effective Teaching (MET) Project (\url{http://www.metproject.org}), the entire repository totaled about 10 TB of digital storage.
 
-Even more important than the quantity of information stored is the kind of information: What are the targets of inquiry?
-The range of phenomena covered by developmental science spans multiple levels of analysis in space (from genes to geography) and time (from microseconds to millennia) all aimed at answering two questions: ``What develops?'' and  ``Why does it develop this way?''
-In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising at different temporal and spatial scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
-For social and behavioral scientists, the goal is to describe change in the psychological processes of individuals, groups of individuals, or families.
-Data about neighborhoods, schools, and broader social, cultural, political and environmental influences primarily inform thinking about the development of individual or family behavior.
-Similarly, the assessment of biological data--genes, hormones, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status--inform thinking about how within-person factors influence changes behavior.
+Even more important than the quantity of information stored is the kind of information.
+Developmental science spans phenomena across multiple levels of analysis in space (from genes to geography) and time (from microseconds to millennia) all aimed at answering two questions: ``What develops?'' and  ``Why does it develop this way?''
+In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising across scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
+Most seek to describe change in the psychological processes of individuals, groups of individuals, or families.
+Data about neighborhoods, schools, and the broader social, cultural, political and environment primarily inform thinking about the development of individual or family behavior.
+Similarly, biological data---genes, hormones, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status---reveals the influence of within-person factors on changes in behavior.
 
-Consequently, the aspects of data \emph{volume} pertinent to most developmental scientists include the number of participants or families and/or the number of measurement time points.
+Consequently, the aspects of data \emph{volume} salient to most developmental scientists include the number of participants or families and/or the number of measurement time points.
 Some datasets have hundreds or thousands of participants.
-Such high volume datasets enable more precise estimation of small effects, especially for rare qualities or conditions.
+Such high volume datasets enable the precise estimation of small effects, especially for rare qualities or conditions.
 Similarly, the aspects of data \emph{velocity} most relevant to developmental researchers relate to the frequency or spacing of measurements.
 Velocity can span many orders of magnitude, from physiological measurements collected at millisecond time scales to longitudinal research spanning years or decades.
-High volume or velocity data may inform the estimation of trends within and between people across time \cite{rietveld_replicability_2014}.
-\emph{Variety} encompasses the range of measurement types employed across developmental research--Biological, behavioral, contextual, and cultural/historical/evolutionary--and the use of different types of measurements to address the same underlying construct.
-In addition to the ``three Vs'', intra-individual \emph{variability} and complexity, or the mutual interdependence of individual measures, might also be included.
+High volume or velocity data informs the estimation of trends within and between people across time \cite{rietveld_replicability_2014}.
+\emph{Variety} encompasses the range of measurement types employed across developmental research---biological, behavioral, contextual, and cultural/historical/evolutionary---and the use of different types of measurements to address the same underlying construct.
+In addition to the ``three Vs'', intra-individual \emph{variability} and \emph{complexity}, or the mutual interdependence of individual measures, might also be included.
 
-The collection, management, and analysis of data high in volume, velocity, variety, variability, and complexity poses significant practical challenges for data collection, capture, storage, sharing, analysis, transfer, visualization, and analysis.
-Big data may also magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
-Big data also pose theoretical challenges: For example, how do micro-scale factors influence macro-scale phenomena?
-Nevertheless, big data offer the possibility of tackling some of the most profound and vexing questions in the field if the relations among data components can be revealed in ways that do not undermine ethical research principles.
-Greater openness and more widespread sharing of research data than is current practice will be essential to realize this scientific promise.
-Developmental researchers can look to the recent past for examples of big datasets that have already been collected and in some cases, widely shared for inspiration.
-
+The collection, management, and analysis of data high in volume, velocity, variety, variability, and complexity poses significant practical challenges for data collection, capture, storage, transfer, sharing, visualization, and analysis.
+Big data magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
+Big data pose theoretical challenges: For example, how do micro-scale factors influence macro-scale phenomena?
+Nevertheless, big data offer developmental researchers the opportunity to tackle some of the most profound and vexing questions in the behavioral science---if the relations among data components can be revealed in ways that do not undermine ethical research principles.
+Realizing this promise will require greater openness and more widespread sharing of research data than current practice.
+But, researchers may draw inspiration from examples of big datasets addressing developmental questions that have already been collected, and in some cases, widely shared.
 \section{Big Datasets in Developmental Research}
-
 Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
-Many of the largest existing datasets available to developmental researchers come from data collected and disseminated by government entities such as the U.S.
-Census Bureau and the U.S.
-Department of Education.
+Many of the largest existing sources come from data collected and disseminated by government entities such as the U.S. Census Bureau and the U.S. Department of Education.
 The collection and dissemination of these datasets is mandated by law and subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
 Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
-Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded, usually by government or private funders, to collect large datasets form the next group.
+Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded by government or private foundations form a second group.
 These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
-The extent to which datasets collected by government-grant-funded research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
-Finally, some large-scale datasets are collected by private entities for business purposes and may include sharing data about individual subscribers or users.
-The data collected by private entities about individuals largely concern consumer behavior because the data customers are other commercial entities seeking to sell products or services to individual users.
-But, commercial entities are beginning to collect and store health and fitness related activity (e.g., Fitbit, etc.) 
-These datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the commercial entities that provide the services.
-
+The extent to which datasets collected by non-government research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
+Large-scale datasets collected by private entities for business purposes form a third group.
+The data collected by private entities about individuals largely concern consumer behavior although health and fitness related data are an emerging area (e.g., Fitbit; \url{http://www.fitbit.com}).  
+Commercial datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the commercial entities that provide the services.
 \subsection{Government-collected and Managed Datasets}
-
-Some of the largest and most widely used datasets in developmental science are population-based, often with a focus on health.
+Population-based studies constitute some of the largest and most widely used datasets in developmental science.
 The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 on the health and nutritional status of U.S.
 infants, children, and adults.
-The annual survey combines interviews with a geographically diverse representative sample of about 5,000 people.
-NHANES focuses on demographic, socioeconomic, dietary and health-related topics and physical examinations involving medical, dental, and physiological measurements, and laboratory tests of behavior.
-The products include standardized child growth charts and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}), which reports physical activity and fitness levels in 3 to 15-year-olds
+The annual survey combines interviews from a geographically diverse representative sample of about 5,000 people.
+NHANES focuses on demographic, socioeconomic, dietary and health-related topics, physical examinations involving medical, dental, and physiological measurements, and laboratory tests of behavior.
+The products include standardized child growth charts and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}), which reports physical activity and fitness levels in 3 to 15-year-olds.
 Much of the NHANES and NYFS data are available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
@@
 
-In contrast to the cross-sectional NHANES, The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) has examined the same cohort--almost 10,000 American youth born between 1957 and 1964--on a annual basis since the study began in 1979.
-NLSY data are now available for 25 survey rounds on questions relating to employment and other issues; extensions encompass wider age ranges.
-Most NLSY and NLS data are available free of charge by means of a web-based data portal to which users must apply for access, but access to geographic-related variables requires special permission.
+In contrast to the cross-sectional NHANES, The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) has surveyed the same cohort---almost 10,000 American youth born between 1957 and 1964--- about employment-related topics on a annual basis since the study began in 1979.
+Most NLSY and NLS data are available free of charge by means of a web-based data portal to which users must apply for access; access to geographic-related variables requires special permission.
 By mid-2015 more than 20,000 publications had cited the NLSY.
 
-The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) is another longitudinal population-based study 
-It focuses on child development, school readiness, and early school experiences in separate cohorts of children followed either from birth or from kindergarten that in some cases exceed 20,000 individuals.
+The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) focuses on child development, school readiness, and early school experiences in separate cohorts of children followed either from birth or from kindergarten.
+Some cohorts exceed 20,000 individuals.
 ECLS consists of both publicly available and restricted data archived at ICPSR.
@@
 The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (\url{http://www.bls.gov/cps/demographics.htm}).
-The U.S.
-Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on U.S.
-primary, secondary, and post-secondary education, and many of the datasets are either public; data with personally identifiable data are available to researchers under a restricted use agreement.
-The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu/}), with more than 20,000 U.S.
-students included in the 2001 sample.
-De-identified data from TIMSS and its sister study, the Progress in International Reading Literacy Study (PIRLS; \url{http://timssandpirls.bc.edu}) are available for public download.
+The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes variables related to U.S. primary, secondary, and post-secondary education.
+Many datasets are public while data with personally identifiable data are available to researchers under a restricted use agreement.
 
-Some large-scale longitudinal studies focused on children's growth, health, and education include participants from outside the U.S.
+Several large-scale longitudinal studies focused on children's growth, health, and education include participants from outside the U.S.
 The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected physical growth and related data from 8,500 children in Brazil, Ghana, India, Norway, Oman, and the U.S.
-The Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) involves 100,000 parent-child pairs with the goal of evaluating the impact of environmental factors on children's health and development.
+The ongoing Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) involves 100,000 parent-child pairs with the goal of evaluating the impact of environmental factors on children's health and development.
 Large-scale birth cohort studies in the U.K. include the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
-NCDS is an ongoing longitudinal study that follows everyone in Great Britain who was born in one particular week in 1958, focusing on factors predicting wellbeing across the lifespan.
+NCDS is an ongoing longitudinal study that follows everyone in Great Britain who was born in one particular week in 1958, focusing on factors that predict wellbeing across the lifespan.
 NCDS data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
-The British Cohort Studies have generated more than 3,500 publications to-date.
+The British Cohort Studies such as NCDS have generated more than 3,500 publications to-date.
 The Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.
@@
 Not all large-scale birth cohort studies succeed.
-A notable failure is the National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
+A notable failure is the U.S. National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
 Authorized by the Children's Health Act of 2000, the NCS would have followed 100,000 children prenatally until age 21.
 However, the NIH Director decided to close the NCS in 2014 following the recommendations of an advisory panel. 
-Questionnaire, physical measures, biospecimens and environmental sample data from up to 5,726 participants in the NCS Vanguard study were collected in 2009-2014 prior to study closure.
+Questionnaire, physical measures, biospecimens and environmental data from up to 5,726 participants were collected in 2009-2014 prior to study closure.
 Those data are slated for release in a data archive sometime in 2015.
 
-Several themes emerge from this sample of U.S. and international population-based studies.
+Several themes emerge from this sample of government-led studies.
 Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
@@
 Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center where special security provisions apply.
-Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact scientific discovery.
+Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
 I turn next to datasets whose collection and management is under the control of individual investigators.
-
-\subsection{Public/Private Partnerships and Investigator-driven Project}
-
+\subsection{Public/Private Partnerships and Investigator-driven Projects}
 Several big data studies in developmental science involve partnerships between government agencies and academic investigators.
@@
 These projects are diverse in focus and methods, including population-based studies with a health focus, behavior genetics, brain imaging, cognition and temperament, and education and employment.
-
 \subsubsection{Population-Based Studies}
-
 The Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is a population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
-PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
+PSID began in 1968 with a nationally representative sample of 5,000 families and more than 18,000 individuals.
 As of mid-2015, PSID had generated more than 3,900 publications. 
@@
 
-The National Longitudinal Study of Adolescent Health (Ad Health; \url{http://www.nichd.nih.gov/news/releases/Pages/adolescent.aspx}) began as a survey study undertaken in response to a Congressional mandate designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors.
-Parts of the Ad Health cohort have been followed into young adulthood with four in-home interviews, and in a recent wave, biomarker data from a blood test has been made available. 
-The Ad Health data are archived at ICSPR and are available for public use.
+The National Longitudinal Study of Adolescent Health (Add Health; \url{http://www.cpc.unc.edu/projects/addhealth}) began as a survey study undertaken in response to a Congressional mandate.
+Designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors, parts of the Add Health cohort have been followed into young adulthood with four in-home interviews.
+In a recent wave, biomarker data have been made available. 
+The Add Health data are archived at ICSPR and are available for public use.
 As of mid-2015, more than 5,000 publications had arisen from the Add Health dataset.
@@
 The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
-The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
+The study sought to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
 Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometrics, and demographic information.
-Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, with approval granted on a case-by-case basis.
+Unlike the Add Health study, data in SECCYD are restricted, available only by application, with approval granted on a case-by-case basis.
 Staff at the National Institute of Child Health and Human Development manage the data access and approval process.
@@
 The FLP site lists about 50 publications through 2013.
-
 \subsubsection{Behavior Genetics}
-
 Large data samples have long been required in behavior genetics to detect small effect sizes.
 Several large-scale genetics studies that are not specifically developmental deserve mention.
-The Genome of the Netherlands Project(\url{http://www.nlgenome.nl}) is a publicly available dataset from 250 families, consisting of two parents plus one of their adult children.
+The Genome of the Netherlands Project (\url{http://www.nlgenome.nl}) is a publicly available dataset from 250 families representing two parents plus one of their adult children.
 The Psychiatric Genomics Consortium (PGC; \url{http://www.med.unc.edu/pgc}) contains data from more than 170,000 individuals with psychiatric diagnoses or who are at-risk.
-Many disorders in the PGC dataset have developmental dimensions(autism, attention-deficit hyperactivity disorder, and schizophrenia).
+Many disorders in the PGC dataset have developmental dimensions (autism, attention-deficit hyperactivity disorder, and schizophrenia).
 Results may be viewed in a specialized web browser tool (\url{http://www.broadinstitute.org/mpg/ricopili/}).
 
-The U.K.-based Twins Early Development Study (TEDS; \url{http://www.teds.ac.uk/}) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996, with lab-based behavioral task, survey, and DNA samples from a subset of participants.
+The U.K.-based Twins Early Development Study (TEDS; \url{http://www.teds.ac.uk/}) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996, with lab-based behavioral task and DNA samples from a subset of participants.
 TEDS has generated more than 350 publications to-date.
@@
 
-The Twin and Offspring Study in Sweden (TOSS; \url{http://ki.se/en/meb/twin-offspring-study-in-sweden-toss}) is another twin-focused study aimed at uncovering genetic and environmental contributions to measures of family relationships and mental health.
+The Twin and Offspring Study in Sweden (TOSS; \url{http://ki.se/en/meb/twin-offspring-study-in-sweden-toss}) aimed seeks to uncover genetic and environmental contributions to measures of family relationships and mental health.
 TOSS extends the Twin Moms Project to include a sample of twin fathers (320 twin pairs and their families).
 In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (3,000 individuals in all) and their families.
-While still relatively early in data collection, more than 24 publications reference the TOSS dataset.
+While relatively early in data collection, more than 24 publications reference the TOSS dataset.
 Data sharing plans have not yet been announced.
@@
 Data from 1976-1989 are archived at Dataverse and can be accessed by application.
-
 \subsubsection{Brain Imaging}
-
 Several large-scale studies have focused on developmental patterns in brain structure or activity.
@@
 The study collected structural brain information from 145 4 to 20-year-olds along with behavioral and clinical data.
-More than 100 of the participants were scanned on more than once, about two years apart.
+More than 100 of the participants were scanned more than once, about two years apart.
 The original paper describing this study has been cited more than 3,300 times, as of mid-2015.
@@
 The data are available to qualified researchers by application.
-John Richards and colleagues have combined data from the NIH MRI Study, data collected in their own labs, and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI (\url{http://jerlab.psych.sc.edu/NeurodevelopmentalMRIDatabase/}).
+John Richards and colleagues have combined data from the NIH MRI Study with data collected in their own labs and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI (\url{http://jerlab.psych.sc.edu/NeurodevelopmentalMRIDatabase/}).
 These data can be openly accessed with appropriate citation.
 
-More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING; \url{http://pingstudy.ucsd.edu/}) Project collected multimodal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function in more than 1,000 participants 3-20 years of age recruited from 8 U.S.
+More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING; \url{http://pingstudy.ucsd.edu/}) Project collected multi-modal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function from more than 1,000 participants 3-20 years of age recruited from 8 U.S.
 cities.
@@
 Outside the U.S., The Developing Human Connectome Project (\url{http://www.developingconnectome.org}), led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
-This project aims to study the connectome, a map of human brain connectivity in fetuses from 20 to 44 weeks post-conception.
+This project aims to study the connectome, a map of human brain connectivity, in fetuses from 20 to 44 weeks post-conception.
 Funded by the European Research Council, the imaging data are accompanied by clinical, behavioral and genetic information.
@@
 The project has generated 17 publications as of mid-2015.
+\subsubsection{Language, Cognition, and Temperament}
+The child language community pioneered the aggregation and sharing of developmental datasets about individual participants.
+CHILDES/TalkBank \cite{macwhinney_childes_2001} supports one of the largest and most well-established data archives in the psychological sciences.
+It consists of transcripts and audio and video recordings of children's utterances and from adults with aphasia.
+CHILDES/TalkBank has generated more than 10,000 citations, and the data are available to the research community, much of it publicly.
 
-\subsubsection{Language, Cognition, and Temperament}
-
-The child language community pioneered the aggregation and sharing of datasets.
-The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established archives in the behavioral sciences.
-It consists of transcripts and audio and video recordings of children's utterances along with recent data from adults with aphasia.
-CHILDES/TalkBank has generated more than 10,000 citations, and the datasets are available to the research community, many of them publicly.
-
-WordBank (\url{http://wordbank.stanford.edu/}) is an open access archive that consists of data from more than 40,000 samples of the MacArthur-Bates Communicative Development Inventory (CDI).
+WordBank (\url{http://wordbank.stanford.edu/}) consists of data from more than 40,000 samples of the MacArthur-Bates Communicative Development Inventory (CDI), a widely used test of early language function.
 Catherine Tamis-Lemonda has released video data (\url{https://nyu.databrary.org/volume/8}) from an NSF-funded longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
@@
 The project has generated more than 78 publications, but the data are not available outside of the original investigative team.
-Data were gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
-The LENA (Language ENvironment Analysis; \url{http://www.lenafoundation.org/}) Foundation created a technology framework that allows children's speech in natural settings to be recorded and analyzed.
+The LENA (Language ENvironment Analysis; \url{http://www.lenafoundation.org}) Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed.
 The system provides an automatic language collection and analysis tool for speech language professionals and parents.
-No large-scale archive for LENA data currently exists, but some researchers who use the tool have begun to explore the creation of a data repository specialized for these sorts of data.
+No large-scale archive for LENA data currently exists, but some researchers who use the tool have begun to explore the creation of a data repository specialized for it.
 
-In other areas of cognition, a number of measures that have become standards and their widespread adoption has resulted in data that are large in volume, velocity or variety.
+In other areas of cognition, a number of measures that have become standards.
+Their widespread adoption has resulted in data that are large in volume, velocity or variety.
 Davida Teller \cite{teller1986assessment} pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
@@
 
-The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} is widely used in clinical and epidemiological settings such as the EPICure (\url{http://www.epicure.ac.uk/}) because of the published norms even though there are questions about its validity \cite{hack_poor_2005}.
+The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} is widely used in clinical and epidemiological settings such as the EPICure (\url{http://www.epicure.ac.uk}) because of its published norms even though there are questions about the test's validity \cite{hack_poor_2005}.
 However, data about those norms is under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
@@
 The standardized Wechsler intelligence tests were developed in the 1930's. 
-The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, and the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
+The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6-16, and the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4-6 1/2 years.
 The current version (WISC-V) is published by Pearson (\url{http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html}).
-Data about the norms are not available to researchers.
+Like the Bayley, data about Wechsler norms are not available to researchers.
 However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is housed at ICSPR and are accessible by specific authorization.
@@
 IBQ and Lab-TAB data are also available from ICPSR from the Family Life Project.
-
 \subsubsection{Education and Learning Science}
-
 Several studies in the education and learning sciences studies deserve mention because of their volume, velocity, or variety or because of their availability for reuse by others.
-The 1999 TIMSS video study recorded math and science teaching practices in seven countries, and some of the materials are available for public use through a repository hosted at UCLA (\url{http://www.timssvideo.com/}).
-The Gates Foundation Measures of Effective Teaching (MET; \url{http://www.metproject.org/}) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing currently.
+The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu}), with more than 20,000 U.S.
+students included in the 2001 sample.
+De-identified data from TIMSS and its sister study, the Progress in International Reading Literacy Study (PIRLS; \url{http://timssandpirls.bc.edu}) are available for public download.
+The 1999 TIMSS video study recorded math and science teaching practices in seven countries, and some of the materials are available for public use through a repository hosted at UCLA (\url{http://www.timssvideo.com}).
+The Gates Foundation Measures of Effective Teaching (MET; \url{http://www.metproject.org}) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing currently.
 The MET video data are stored and shared at ICSPR.
-The Spatial Intelligence and Learning Center (SILC; \url{http://spatiallearning.org/}) is an NSF-funded initiative designed to develop a multi-disciplinary science of spatial learning, including the development of tests and instruments useful in research.
+The Spatial Intelligence and Learning Center (SILC; \url{http://spatiallearning.org}) is an NSF-funded initiative designed to develop a multi-disciplinary science of spatial learning, including the development of tests and instruments useful in research.
 LearnLab (\url{http://learnlab.org}), another NSF-funded center, combines cognitive theory and computational modeling to understand changes in student knowledge in the context of computer-based math, science and language courses.
-LearnLab hosts a repository, DataShop (\url{http://learnlab.org/technologies/datashop/index.php}), to store and openly share detailed data from classrooms employing cognitive tutors.
+LearnLab hosts a repository, DataShop (\url{http://learnlab.org/technologies/datashop/index.php}), to store and openly share detailed data from classrooms employing computer-based cognitive tutors.
 DataShop enables data storage, management, visualization and analysis in a web-based tool.
-
 \subsubsection{Summary}
-
-The studies involving partnerships with government agencies or those initiated and managed by individual investigators exemplify the high volume, velocity, variety, and variability of big datasets.
-The studies employ a range observational measures, including video recordings, population-normed test instruments, biological measurements of physiology, and brain structure or function.
+Studies involving partnerships with government agencies or those initiated and managed by individual investigators exemplify the high volume, velocity, variety, and variability of big datasets.
+The studies employ a range of observational measures, including video recordings, population-normed test instruments, biological measurements of physiology, and brain structure or function.
 However, the extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.
 Only some of the datasets are stored in data repositories, for example.
-Many large-scale developmental dataset are usually housed locally, on project-specific sites, not on centralized servers that aggregate data across studies and sources.
+Many large-scale developmental datasets are housed locally, on project-specific web sites, not on centralized servers that aggregate data across studies and sources.
 In many cases, the original investigative team retains control over the use of data by other researchers, including the kinds of questions that third parties may ask.
 In some cases, the original investigative team must be included as an author on publications.
-Perhaps as a consequence of these factors, these datasets have generated fewer scientific publications than government-sponsored efforts.
+Perhaps as a consequence of these factors, these datasets have generated fewer scientific publications than government-led efforts.
 
 \subsection{Non-Academically Initiated or Managed Sources of Big Data}
-
 Some large-volume sources of developmental data are collected and managed by private, non-academic or government entities.
-More than 1.6 million high school students \cite{lewin_more_2013} take standardized tests or provide financial aid information via measures developed and managed by The College Board and ACT, Inc.
+For example, more than 1.6 million high school students \cite{lewin_more_2013} take standardized tests or provide financial aid information via measures developed and managed by The College Board or ACT, Inc.
 The College Board shares Scholastic Aptitude Test (SAT) and college cost and scholarship data with the research community by application.
-So does the ACT (\url{http://www.act.org/research/}).
+So does the ACT (\url{http://www.act.org/research}).
 
 Internet-based for-profit service providers operate at an even larger scale.
-Google's Gmail has more than 900 million users worldwide \cite{lardinois_gmail_2015}, and Facebook has more than a billion \cite{facebook_2014}.
+Google's Gmail has more than 900 million users worldwide \cite{lardinois_gmail_2015}; Facebook has more than a billion \cite{facebook_2014}.
 According to the YouGov site (\url{https://yougov.co.uk}), 17\% of Gmail users in the United Kingdom are 17-24 years of age.
 Facebook's policies require that users be at least 13 years of age (\url{https://www.facebook.com/help/157793540954833}), but detailed information about user demographics for Facebook or other social media popular among children and adolescents is not openly available.
-Of course, detailed information about users, their characteristics, and preferences is the primary asset social media companies mine and market to the business community.
+Of course, detailed information about users, their characteristics and preferences is the primary asset social media companies mine and market.
 Users receive free services in exchange for providing these data.
@@
 Because the data are collected for proprietary business purposes, it is difficult to assess their current or potential impact on the scholarship of human development.
-
 \section{The Future of Big Data in Development}
-
 Clearly, the collection, analysis, and sharing of large, multilevel datasets has been part of the fabric of developmental science for a long time.
 In this section, I discuss a range of technical, conceptual, and theoretical issues that arise in thinking about the future of big data in developmental science. 
-
 \subsection{Technical}
-
-Technical issues associated with big data in developmental science center on collection, storage and retrieval, data management, provenance, and analysis.
-
+Technical issues associated with big data in developmental science center on collection, storage and retrieval, data management, provenance, and analysis \cite{goodman_ten_2014}. 
 \subsubsection{Collection from Multiple Sources and in Diverse Formats}
-
 Developmental scientists collect data from sources representing multiple levels of analysis.
@@
 
-Although some researchers continue to use paper and pencil measures to collect survey information, many universities now have site-licenses for web-based tools such as SurveyMonkey and Qualtrics that  reduce the manual labor involved in preparing a survey and processing completed data for analysis.
-Behavioral measures involving computer-based tasks are commonly used in developmental research, but most rely on custom, project-specific software.
+Although some researchers continue to use paper and pencil measures to collect survey information, many universities now have site-licenses for web-based tools such as SurveyMonkey and Qualtrics.
+These reduce the manual labor involved in preparing a survey and processing completed data for analysis. 
+Developmental research commonly use behavioral measures involving computer-based tasks, but most rely on custom, project-specific software.
 So, the output data files, while often in an electronic form, may require significant post-processing to be linked with other data.
-New tools like Amazon's Mechanical Turk (\url{http://www.mturk.com}) or Apple's HealthKit (\url{https://developer.apple.com/healthkit/}) are empowering behavioral and health researchers to conduct large-scale behavioral science experiments using tools specialized for psychological research (e.g., \url{https://psiturk.org}), and with data delivered in well-structured electronic formats.
+Some researchers have begun to use tools such as Amazon's Mechanical Turk (\url{http://www.mturk.com}) or Apple's HealthKit (\url{https://developer.apple.com/healthkit/}) to conduct large-scale behavioral science experiments.
+These sites deliver data in well-structured electronic formats, sometimes using tools specialized for psychological research (e.g., \url{https://psiturk.org}).
 Amazon's terms of use prohibit minors, but developmental researchers have found ways to secure video-based informed consent from parents to enable their children to participate in looking time studies (\url{https://lookit.mit.edu}) over the web.
 
-Video and audio recordings are a long-standing mainstay of developmental research.
+Large numbers of developmental researchers collect video and audio recordings.
 Video captures the complexity and richness of behavior unlike any other measure, and so video provides a uniquely valuable source of information for researchers who study behavior in laboratory, home, classroom, or museum contexts.
-Images and recordings are larger in file size, denser than text or flat-file data, and come in a diverse formats. 
-Thus, a high priority is the development of tools to enable storage and sharing of images (including brain images), and audio and video data.
+Images and recordings generate large, dense files and come in a diverse formats.
+With few notable exceptions (e.g., Databrary, \url{http://databrary.org}, and the MET Project) most existing data archives support the storage and sharing of text files, but not images (including brain images), audio, and video data.
 
-Lab-based tools for conducting physiological measurements such as EEG, heart-rate or skin conductance, produce electronic files of these time series.
 Genetic analyses from modern gene sequencing tools and reports from tissue, blood, or salivary samples typically yield machine-readable outputs.
@@
 But, unlike MRI, there are no standard file formats, and most data collection systems provide no standard subject-level metadata.
+Lab-based tools for conducting physiological measurements such as EEG, heart-rate or skin conductance, produce electronic files.
 Thus, the files require significant post-collection data processing prior to analysis.
 
-New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promise to make big data streams about individual participants' locations, physiological states (e.g., \url{https://www.empatica.com/}), activity patterns, and momentary cognitive, and emotional states broadly available to researchers.
+New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promise to make big data streams about individual participants' locations, physiological states (e.g., \url{https://www.empatica.com}), activity patterns, and momentary cognitive, and emotional states broadly available to researchers.
 These tools will enable the collection of data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety of data available for analysis.   
-
 \subsubsection{Storage and Retrieval}
-
 Developmental researchers who wish to store and share big data face a bewildering array of options.
-These include individual or institutional websites, institutional repositories (e.g., \url{https://scholarsphere.psu.edu/}), cloud services (Dropbox, Box, or Amazon), domain or measure-specific repositories (ICSPR, Databrary.org, TalkBank.org, WordBank.org, OpenfMRI.org), domain general services (Researchgate.net, FigShare/SlideShare, Dataverse, and the Open Science Framework), and open source software web sites (GitHub).
+These include individual or institutional websites, institutional repositories (e.g., \url{https://scholarsphere.psu.edu}), cloud services (Dropbox, Box, or Amazon), domain or measure-specific repositories (ICSPR, Databrary.org, TalkBank.org, WordBank.org, OpenfMRI.org), domain general services (Researchgate.net, FigShare/SlideShare, Dataverse, and the Open Science Framework), and open source software web sites (GitHub).
 Some journals offer or require data storage, but these are typically limited to text-based flat-files used for statistical analyses and do not include raw images, videos, or physiological time series.
@@
 Storage solutions must meet the needs of researchers during the active data collection phase of a study while not posing insurmountable hurdles to data sharing down the line.
-The effort to reconcile these competing demands led Databrary (\url{http://databrary.org}), for example, to build tools that allow researchers to upload session-level video and flat-file data to a secure web-based server as it is collected, thus minimizing post-study data curation.
+The effort to reconcile these competing demands led Databrary (\url{http://databrary.org}) to build tools that allow researchers to upload session-level video and flat-file data to a secure web-based server as the data are collected, thereby minimizing post-study data curation.
 The Open Science Framework (\url{https://osf.io}) offers similar data management functionality for non-identifiable data.
@@
 Where and how data are stored is only part of the problem.
-To foster increased reuse, data must be discoverable and accessible to researchers.
+To foster increased reuse, data must be made discoverable and accessible to other researchers.
 At present, it is far easier to search and discover research publications relevant to a particular topic using web-based search tools than it is to find data.
@@
 Most research publications do not use data that are readily available to investigators outside of the research team.
-Datasets that are available may lack persistent, citable, identifiers.
+Available datasets may lack persistent, citable, searchable, identifiers (e.g., digital object identifiers or DOIs).
 When data from a publication are available to other researchers, access is often restricted and requires a specific, time consuming application to a data repository or to the original data producer.
@@
 Another barrier to reuse is the difficulty of finding data that meet specific task or demographic criteria.
-Some repositories such as ICPSR and the National Database for Autism Research (NDAR) maintain extensive standardized metadata about tasks and participant demographics.
+Some repositories such as ICPSR and the National Database for Autism Research (NDAR; \url{https://ndar.nih.gov}) maintain extensive standardized metadata about tasks and participant demographics.
 This can help investigators to search for specific data sources.
-But, even within domain-specific repositories variable-level searching is not available for all datasets.
+But, not all datasets support variable-level search, and supplementing datasets with extensive metadata requires expertise and financial resources many research teams lack.
 The problems of where to store and how to find and retrieve data will increase as datasets grow in size and complexity.
-
 \subsubsection{Coding, Analysis, and Provenance}
-
 Even easy-to-find datasets must be processed prior to analysis.
@@
 For example, physiological data are often filtered and smoothed, sometimes by the recording devices.
-Video data are usually coded by human observers and the codes transformed into quantitative measurements.
+Video data are usually edited and coded by human observers and the codes transformed into quantitative measurements.
 What were the variables, units of measurements, calibration properties of the instrument, and definitions of key terms and codes?
-Well-curated datasets usually report these components, but that curation takes time and specialized expertise that many individual investigators lack.
+Well-curated datasets usually report these components, but curation takes time and specialized expertise that many individual investigators lack.
 
-Several software tools have recently emerged that make it easier for researchers to produce and reproduce self-documenting data workflows.
-For example, the free RStudio (\url{https://www.rstudio.com}) and Jupyter (\url{https://jupyter.org}) environments allow researchers to create electronic notebooks that combine data, annotations and observations, statistical analyses, and visualizations in human-friendly web or document formats.
+Several software tools have recently emerged that make it easier for researchers to produce and reproduce self-documenting data workflows, thus reducing the curational burden.
+For example, the free RStudio (\url{https://www.rstudio.com}) and Jupyter (\url{https://jupyter.org}) environments allow researchers to create electronic notebooks that combine data, annotations, observations, statistical analyses, and visualizations in human-friendly formats.
 The free, open-source Datavyu (\url{http://datavyu.org}) video coding tool allows automated data analysis and export schemes to be created with the Ruby scripting language.
@@
 Similarly, Databrary and the Center for Open Science (\url{http://centerforopenscience.org}) have initiated open office hours and conference-based and regional workshops to provide hands-on researcher training.
-But, the use of tools that produce well-curated, reproducible scientific workflows remains rare among mainstream developmental researchers.
-
+Still, the use of tools that produce well-curated, reproducible scientific workflows remains rare among mainstream developmental researchers.
 \subsubsection{Summary}
-
 Technical issues will continue to slow progress in many areas of developmental research that depend on big data.
-Critical challenges include getting data into open, standard, and easily manipulated electronic formats as soon as possible in the research cycle; the development and widespread adoption of data storage platforms or repositories that provide a degree of metadata standardization and enable search and discovery; the creation and adoption of data management practices that make curation part of the research workflow; and the creation of a cohort of developmental researchers who have the training and expertise to implement these techniques in their own labs.
+Critical challenges include getting data into open, standard, and easily manipulated electronic formats as soon as possible in the research cycle; the development and widespread adoption of data storage platforms or repositories that provide metadata standardization and enable search and discovery; the creation and adoption of data management practices that make curation part of the research workflow; and the creation of a cohort of developmental researchers who have the training and expertise to implement these techniques in their own labs.
 There is demonstrable progress on many of these fronts, and therefore cause to be optimistic that the technical challenges can be overcome.
-
 \subsection{Research Ethics and Practice}
-
 Clearly the collection, analysis, and interpretation of large scale datasets present issues related to research ethics, participant privacy, and scientific transparency.
@@
 
-Research activities funded by the U.S. Federal Government are supervised by an Institutional Review Board or its equivalent.
-These entities are regulated by the U.S. Office of Health and Human Services (\url{http://www.hhs.gov/ohrp/}).
+Research activities funded by the U.S.
+Federal Government must be supervised by an Institutional Review Board or its equivalent, and many institutions conducting research with human subjects that is not federally-funded follow the same procedures.
+IRBs are regulated by the U.S. Office of Health and Human Services (\url{http://www.hhs.gov/ohrp/}).
 Researchers supervised by IRBs must respect participants' privacy, secure informed consent, and maintain confidentiality. 
 These ethical principles have practical consequences for research.
-They limit the ways researchers may recruit participants.
+They limit the ways that researchers may recruit participants.
 Minors must give informed assent to participate in a study with a parent or guardian giving formal consent.
 Data must be collected in ways that minimize the likelihood that information given by a participant will be disclosed or a participants' identity revealed outside the research team.
-This usually means that data items that could reveal a participant's identity are removed or altered to reduce the likelihood of disclosure.
-Whether deidentified data of this sort can be shared with researchers outside the original team that collected depends on several factors.
-One factor is the sensitivity of the data collected and the likelihood that specific identities or home locations or other items could be revealed.
+This usually means that researchers remove or alter data items that could reveal a participant's identity to reduce the likelihood of disclosure.
+Whether deidentified data of this sort can be shared with researchers outside the original IRB-approved team that collected it depends on several factors.
+One factor is the sensitivity of the data collected and the likelihood that specific identities or home locations could be revealed.
 Another factor concerns whether participants were informed that deidentified data might be shared outside the research team.
 IRBs may view these matters differently, creating additional complexities for large-scale projects that span geographic areas.
-Some IRBs may require participants to be informed that deidentified data might be shared outside the IRB-approved research team, and others may deem that the analysis of deidentified data no longer meets the definition of human subjects research.
-Many big datasets in developmental science have restrictions on access either because the data were collected under Federal regulations that prohibit releasing identifiable data or because the participants were not asked for permission to share data with other researchers.
+Some IRBs may require participants to be informed that deidentified data might be shared outside the IRB-approved research team, and others may deem that the analysis of deidentified data no longer meets the definition of human subjects research and thus requires no additional approval.
+Many big datasets in developmental science have restrictions on access either because the data were collected under Federal regulations that prohibit releasing individually identifiable data or because the participants were not asked for permission to share data with other researchers.
 From a big data perspective, if data cannot be shared outside the original IRB-approved research team, then the possible analyses are restricted to the interests, resources, and expertise of that team.
@@
 Thus, data from photographs or recordings requires additional consideration and special care.
-Databrary, a digital data library specialized for storing, managing, and sharing video data from developmental research has an access model that empowers researchers who wish to share identifiable research data to do so with explicit permission of the participants.
-Databrary has created template language to help researchers gain participants' permission.
+Databrary, a digital data library specialized for storing, managing, and sharing video data from developmental research, has an access model that empowers researchers who wish to share identifiable research data to do so with explicit permission of the participants.
+Databrary has created template language to help researchers secure and document participants' permission.
 Furthermore, Databrary restricts access to identifiable data to researchers who have formally agreed to uphold ethical research principles and whose institutions approve of their access.
@@
 The experience of Databrary investigators is that a significant proportion of research participants and their parents or guardians will consent to sharing identifiable data, mostly video, with other members of the research community.
-
 It is too early to predict whether it will become commonplace for academic developmental researchers to seek explicit permission to share identifiable research data with other researchers.
@@
 In just over a year of operation, Databrary has secured formal agreements with more than 80 institutions in North and South America, Europe, Australia, and Asia allowing more than 140 researchers to access identifiable data.
+
 However, some leading developmental researchers have argued that the families of research participants forge a relationship of trust with a particular research team, formalized through the informed consent document \cite{eisenberg_thoughts_2015}.
 The relationship might be harmed or the research project negatively affected if participants were asked to share data with other researchers.
-Sensitive to the latter argument, Databrary has recommended that permission to share be sought separately from consent to participant in research and after a given data collection has ended.
-Moreover, the fact that most families agree to share when asked suggests that the relationship of trust involved in research participation might be extended to a community of researchers, given suitable provisions and constraints.
-Undoubtedly, seeking explicit permission to share on a consistent and widespread basis would resolve any ambiguity about whether a given dataset can be shared with whom and for what sort of purposes.
+Sensitive to the latter argument, Databrary recommends that permission to share be sought separately from consent to participant in research and after a given data collection episode has ended.
+The fact that most families agree to share when asked suggests that the relationship of trust involved in research participation might be extended to a community of researchers, given suitable provisions and constraints.
+Undoubtedly, seeking explicit permission to share on a consistent and widespread basis would resolve any remaining ambiguity about whether a given dataset can be shared with whom and for what sort of purposes.
 
@@
 Although some services attempt to restrict the ages at which users can create accounts, the limits often lack rigor, and there is no parallel to the requirement of adult consent required in formal research contexts.
-The data collection and analyses carried out by private entities are subject to no supervision or formal regulation comparable to academic research.
+The data collection and analyses carried out by private entities are not subject to supervision or formal regulation comparable to academic research.
 Instead, data use, analysis, and sharing provisions are governed by terms of use agreements that users acknowledge by clicking a button prior to using a given service.
@@
 
-Unresolved issues that could impact the availability of big data in the future include whether linkage across streams increases risk of reidentification, whether it is essential to reconsent minors when they become adults, a notion most researchers find totally impractical and a significant barrier to data sharing, and a general concern about the ethics of granting consent to share data for an indefinite period.
-Because data security cannot ever be guaranteed, risks can be minimized and managed, but not entirely eliminated. 
-Finally, there are unresolved questions about privacy protections in the consumer domain that have the potential to influence academic research
-
+Unresolved issues that could impact the availability of big data in the future include whether linkage across streams increases the risk of reidentification, whether it is essential to reconsent minors when they become adults, a notion most researchers find totally impractical and a significant barrier to data sharing, and a general concern about the ethics of granting consent to share data for an indefinite period.
+Because data security cannot ever be guaranteed, risks can only be minimized and managed, but not entirely eliminated. 
+Finally, there are unresolved questions about privacy protections in the consumer domain that have the potential to influence public attitudes toward academic research \cite{meyer_everything_2014}.
 \subsubsection{Transparency and Reproducibility}
-
-An important dimension of scientific ethics concerns transparency and reproducibility.
+Another important dimension of scientific ethics concerns transparency and reproducibility.
 The social and behavioral sciences have incurred an unfortunate string of high profile cases of scientific misconduct in recent years, including cases of fraudulent data \cite{singal_case_2015, bhattacharjee_diederik_2013}.
 The credibility problem is magnified by several factors.
-Lack of power and unrestricted exploratory analyses may mean that most research findings are false \cite{ioannidis_why_2005}.
-The actual effect sizes of published findings are unknown due to a bias toward publishing positive results.
+Lack of power and unrestricted exploratory analyses may mean that most published research findings are false \cite{ioannidis_why_2005}, and true effect sizes are unknown due to a bias toward publishing positive results.
 Most journals reject papers that report failures to replicate published findings, and as a result, few scientists attempt replications or are recognized and rewarded for doing so \cite{nosek_scientific_2012}.
@@
 
-Whether there are replicability problems exist in developmental science and whether they constitute a crisis is unknown.
-Developmental research reflects the same positive effects biases seen in other fields, and the same problem that null results often sit unpublished in file drawers--the so-called file drawer effect \cite{rosenthal_file_1979}.
-No failures to replicate developmental studies have been reported to Psychfiledrawer.org (\url{http://psychfiledrawer.org}), a resource designed to bring replication failures to light.
+Whether there replicability problems exist in developmental science and whether they constitute a crisis is unknown.
+Undoubtedly, developmental research reflects the same positive effects biases seen in other fields, and the same problem that null results often sit unpublished in file drawers---the so-called file drawer effect \cite{rosenthal_file_1979}.
+Still, no failures to replicate developmental studies have been reported to Psychfiledrawer.org (\url{http://psychfiledrawer.org}), a resource designed to bring replication failures to light.
 As some developmental researchers have written \cite{bishop-blog-2012}, replicating effects with developmental populations can be especially difficult and so even partial replications are noteworthy.
@@
 Researchers should work to reduce the extent of blame levied at researchers whose initial positive findings fail to be replicated by others \cite{bishop-blog-2015}.
-Technological tools that foster increased openness and transparency and more systematic research data management (OSF and Databrary) will also contribute to changing the scientific practices.
+Technological tools that foster increased openness and transparency and more systematic research data management (OSF and Databrary) will also contribute to changing scientific practices.
 So will the widespread adoption of more consistent journal practices related to transparent and open scientific practices \cite{nosek_promoting_2015}.
@@
 In contrast, increased transparency about the process that led to an exploratory finding and the steps taken to confirm it can bolster a finding's credibility.
-Thus, developmental researchers may find it essential to adopt more transparent and reproducible workflows using some of the new tools developed for this purpose (COS; Databrary).
-
+Thus, developmental researchers may find it essential to adopt more transparent and reproducible workflows using some of the new tools developed specifically for this purpose (e.g., OSF, Databrary, RStudio, Jupyter).
 \subsubsection{Community Engagement and the Impetus for Change}
-
 Developmental researchers have clearly shown enthusiasm for sharing the results of their findings via publications, and in some subfields, the sharing of data, materials and methods is firmly established.
@@
 
-In addition to bottom-up/grassroots initiatives, journals and funding agencies continue to play a vital role in creating an impetus for change.
-Funders can require data management plans and mandate that data and research products be deposited into particular types of open repositories and provide funding to build and support big data infrastructure.
-Journals can also require that data be deposited in open archives as a condition of publication in addition to adopting other transparent and open science practices for manuscripts they accept (e.g.
+In addition to bottom-up/grassroots initiatives, journals and funding agencies continue to play a vital role in creating an impetus for change in data practices.
+Many funders require data management plans, mandate that data and research products be deposited into particular types of open repositories, and provide funding to build and support big data infrastructure.
+Journals are beginning to require that data be deposited in open archives as a condition of publication in addition to adopting other transparent and open science practices for manuscripts they accept (e.g.
 PLoS).
-However, the problem with data sharing mandates from funders is that there is no specific mechanism to provide sufficient ongoing financial support for data archives.
-Few researchers budget funds to support data management and archiving.
+One problem with data sharing mandates from funders is that there is no specific mechanism to provide ongoing financial support to data archives.
+Another is that few researchers budget funds to support data management and archiving and with increasing competition for grants, may be reluctant to do so.
 Some journals are willing to shoulder the burden of storing and sharing data associated with publications, but others refuse to accept supplemental materials of any kind \cite{maunsell_announcement_2010}.
@@
 That consensus may well prove vital to achieving some of the benefits of big data analyses in development. 
-
 \subsection{Conceptual and Theoretical Issues}
-
-The increasing availability of big datasets for analysis in developmental research poses significant theoretical and conceptual questions alongside the many pragmatic ones already discussed.
+The increasing availability of big datasets for analysis in developmental research poses significant theoretical and conceptual questions alongside the pragmatic ones already discussed.
 Big(ger) data may help to overcome limitations with our existing knowledge base.
-One challenge that big data may help to address is a particular bias in existing samples.
+Specifically, big data may help mitigate a particular bias in existing samples.
 Developmental research typically purports to study what is normative about changes across time in human behavior.
-But, much of what we have learned about developmental processes comes from samples that represent only a small fraction of the world's population \cite{karasik2010weird, fernald2010getting}
-Developmental psychology, like other branches of the field, presents findings from Western, education, industrialized, rich, and democratic (WEIRD) societies \cite{henrich_weirdest_2010}.
+But, much of what we have learned about developmental processes comes from samples that represent only a small fraction of the world's population \cite{karasik2010weird, fernald2010getting}.
+Developmental psychology, like other branches of the psychological science, presents findings from Western, education, industrialized, rich, and democratic (WEIRD) societies \cite{henrich_weirdest_2010}.
 So, to the extent that new tools enable research on development in non-WEIRD cultures and those data can be aggregated and combined will strengthen the ability to make claims about universal or near-universal components of developmental processes.
-However, developmental researchers are well aware of cohort effects--the notion that developmental processes can be influenced by changing social and cultural norms.
-Thus, even the most culturally diverse dataset possible may still yield conclusions that are locked in time.
+However, developmental researchers are well aware of cohort effects---the notion that developmental processes can be influenced by changing social and cultural norms.
+Thus, even the most culturally diverse dataset may still yield conclusions that are locked in time.
 
@@
 Most worryingly many published research findings are false in fields that rely on small sample sizes, test multiple relationships between variables, engage in exploratory research, use diverse research designs, definitions, outcomes, and analytical modes across studies, and when more labs seek out significant effects \cite{ioannidis_why_2005}.
-The collection, analysis, and sharing of larger datasets will help to ameliorate these factors.
+Developmental research reflects many of these characteristics, but the collection, analysis, and sharing of larger datasets should work to reduce their impact.
 
-Developmental research faces a specific point of tension related to the measurement.
-Many of the measures for which high volume data are available come from proprietary, expensive instruments such as the Bayley and the WIPPSI.
+Developmental research faces a specific point of tension related to measurement.
+Many of the measures for which high volume data are available come from proprietary, expensive instruments such as the Bayley and the WIPPSI for which baseline data about population norms are unavailable.
 Free, academic instruments such as the Infant Behavior Questionnaire have no centralized data archive.
@@
 Most investigators customize even a well-known task to make it suitable for use with children, and the sharing of research materials is just as limited as the sharing of data.
-Efforts to encourage researchers to capture and record the conceptual structure of psychological tasks have been undertaken (e.g., The Cognitive Atlas; \url{http://www.cognitiveatlas.org}), but have not yet been widely adopted.
+Efforts to encourage researchers to capture and record the conceptual structure of psychological tasks have been undertaken (e.g., The Cognitive Atlas; \url{http://www.cognitiveatlas.org}) but are commonly used.
 
-Finally, some critics have raised concerns that the rise big data means the ``end of theory'' \cite{anderson_end_2008}.
+Finally, some critics have raised concerns that the rise of big data means the ``end of theory'' \cite{anderson_end_2008}.
 In a provocative essay Anderson \citeyear{anderson_end_2008} argued that large quantities of data mean the traditional model of scientific inquiry involving hypothesis testing will soon give way to model-free descriptions of data.
-Others note that bigger data don't necessarily lead to deeper insights \cite{graham_big_2012}.
-Boyd and Crawford \cite{boyd_critical_2012} note that beyond its scientific dimensions, big data is a sociocultural phenomenon that raises as many questions as it promises answers.
-Developmental science has a rich and rigorous intellectual history in which theory and experiment play central roles.
+Others note that bigger data do not necessarily lead to deeper insights \cite{graham_big_2012}.
+While some data intensive fields, largely in computer science, have adopted theory-free approaches to discovery, developmental science has a rich and rigorous intellectual history in which theory and experiment play central, essential role in scholarly discourse.
 That tradition is likely to continue.
-
 \section{Conclusion}
-
-As boyd and Crawford \citeyear{boyd_critical_2012}  ``The era of Big Data has begun.
+As boyd and Crawford \citeyear{boyd_critical_2012} observe ``The era of Big Data has begun.
 Computer scientists, physicists, economists, mathematicians, political scientists, bio-informaticists, sociologists, and other scholars are clamoring for access to the massive quantities of information produced by and about people, things, and their interactions.'' (p.
 662).
-The clamor extends to the developmental and learning sciences where improving health and maximizing the potential for human achievement have significant consequences for daily living.
+The clamor extends to the developmental and learning sciences where discoveries have the potential to improve health and maximizing the potential for human achievement.
 
-This review has shown that the collection, dissemination and analysis of data sets that are big in volume, velocity, or variety has a long and established history in developmental science.
-Many `big data' studies have had substantial impact on scholarship, and in some cases, on public understanding and policy.
-For the most part, studies with the largest impact (as measured by the quantity of published papers) have been ones funded by and managed by government entities, perhaps due to widespread access to data for reuse.
-Those investigator-initiated and managed projects that have had the largest intellectual impacts have created significant intellectual communities around the datasets, communities that extend the beyond the boundaries of the original investigative teams.
-Although technical issues about data formats, storage, cleaning, visualization and provenance remain, there is significant progress.
-Developmental researchers have available a growing array of well-established data repositories (CHILDES, Databrary, Dataverse, ICPSR) and new data storage/management tools (Databrary, OSF).
-Research and data management practices have begun to converge on norms that will reduce the costs of preparing data for sharing in the future.
-New ethical procedures for seeking informed consent to share identifiable data have been developed and are being implemented in diverse research contexts that promise to accelerate future reuse.
+However, that potential is limited because most developmental science data are hard to find and cumbersome to access, even for researchers.
+Data that are available have restrictions that largely prohibit analyses at the level of individual participants.
+Most data linked to publications are not stored in open data repositories.
+Virtually all of the data from unpublished studies remains unavailable, making the size of the file drawer effect unknown.
+Most investigators do not currently employ workflows that make it easy to share data or to document analysis pathways.
+With rare exceptions clustered around specific datasets, there is no widespread culture of data sharing, and in some subfields a degree of bias against the use of secondary data.
+Finally, there is no unified understanding or consensus within developmental science about who owns research data, whether it is essential or merely wise to share data, and when in the research cycle data should be shared.
+These factors limit the potential for discovery that the era of big data so seductively promises.
 
-On the other hand, many barriers remain.
-Most developmental science data are hard to find, and cumbersome to access, even for researchers.
-Most data linked to publications are not stored in open data repositories. 
-This means that virtually all of the data from unpublished studies remains unavailable, making the size of the file drawer effect unknown.
-Most investigators do not currently employ workflows that make it easy to share data or to document analysis pathways.
-With rare exceptions clustered around specific datasets there is no widespread culture of data sharing, and indeed some level of bias against the use of secondary data.
-Finally, there is no unified understanding or consensus within developmental science about who owns research data, whether it is essential or merely wise to share data, and when in the research cycle data should be shared.
-These factors limit the potential for discovery that the era of big data so seductively promises, especially in the face of what can seem a glacial pace of change in scientific culture.
+Still, this review has shown that the collection, dissemination and analysis of data sets that are big in volume, velocity, or variety has a long and established history in developmental science.
+Many big data studies have had substantial impact on scholarship, and in some cases, on public policy.
+For the most part, studies with the largest impact (as measured by the quantity of published papers) have been ones funded by and managed by government entities.
+Investigator-initiated projects with the largest impacts have attracted significant intellectual communities around the datasets that extend the beyond the boundaries of the original investigative teams.
+Thus, the impact of existing big datasets appears tightly linked to the degree to which information from them is widely shared.
+This suggests that the future of big data approaches in developmental science depends on the extent to which barriers to data sharing can be overcome.
 
-Nevertheless, we should remember that Facebook was launched in 2004, Twitter in 2006, and the iPhone in 2007.
-It would be unwise to underestimate the speed with which new technologies, tools, and cultural practices can change.
-If developmental researchers can find ways to collect, manage, store, share, and enable others to reuse data about the multiple facets of human development, as many are beginning to do, we can look forward to a future rich in theory and understanding.
+Technical issues about data formats, storage, cleaning, visualization, and provenance remain, but significant progress has been made in addressing them.
+Developmental researchers have available a growing array of data repositories (CHILDES, Databrary, Dataverse, ICPSR) and new data management tools (Databrary, OSF).
+Research and data management practices have begun to converge on norms that will reduce the costs of preparing data for sharing in the future \cite{goodman_ten_2014}.
+New ethical procedures for seeking informed consent to share identifiable data have been developed and are being implemented in diverse research contexts.
+These promise to accelerate the reuse of data which has previously been difficult or impossible to share widely.
+
+We should remember that Facebook was launched in 2004, Twitter in 2006, and the iPhone in 2007.
+It would be short-sighted to underestimate the speed with which new technologies, tools, and cultural practices can change.
+If developmental researchers can find ways to collect, manage, store, share, and enable others to build upon data about the multiple facets of human development, as many are beginning to do, we should look forward to a future rich in theory and understanding.
 

### 2015-11-28 wiley-big-data-devel/big-data.tex
ms revision

@@
 \usepackage{graphicx}
+\usepackage{array} % for table
 \usepackage[hyphens]{url}
+\usepackage{pdflscape}
+
 %\usepackage{hyperref}
@@
 \abstract{The use of the term ``big data'' has grown substantially over the past several decades and is now widespread. 
-In this review, I ask what makes data ``big'' and what implications of size, density, or complexity of datasets have for the science of human development.
+In this review, I ask what makes data ``big'' and what implications size, density, or complexity of datasets have for the science of human development.
 A survey of existing data sets illustrates how existing large, complex, multi-level, and multi-measure data can reveal the complexities of developmental processes.
@@
 A search of the term using Google's Ngram viewer (\url{https://books.google.com/ngrams}) that indexes terms in digitized books shows the first appearance of the term around 1900 with a steady rise in frequency from the 1950s to around 2000.
-Clearly, as measured by search engine matches of electronic documents on the Internet, scholarly documents, or digitized books, the use of the term has grown substantially over the past several decades, and is now widespread.
-Moreover, the fact that these basic statistics about a particular phrase can be determined in an instant speaks to the rapid progress in networked computing, search engines, and databases.
+Clearly, as measured by search engine matches of electronic documents on the Internet, scholarly documents, or digitized books, the use of the term ``big data'' has grown substantially over the past several decades, and is now widespread.
+Moreover, the fact that these basic statistics about the use a particular phrase can be determined in an instant speaks to the rapid progress in networked computing, search engines, and databases.
 Most of the tools that enable it have been created in the last 20 years.
-Big data has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.
+In turn, big data has become a significant cultural phenomenon \cite{borgman_big_2015,boyd_critical_2012}, with frequent feature articles in the popular \cite{lohr_big_2012,Marcus2013} and specialist press \cite{HBR2015,Press2013a}.
 
 In this review, I show how the increased availability of and interest in big data sets promises to alter the study of human development.
-I begin by asking what makes data ``big'' and the implications of size, density, or complexity of datasets for understanding human development.
-Then, I survey case studies to illustrate how large, complex, multi-level, multi-measure data sets have been used to reveal some of the complexities of developmental processes.
-I conclude by discussing key questions that big data approaches pose for the future of developmental science.
+I begin by asking what makes data ``big'' and what implications the size, density, or complexity of datasets have for understanding human development.
+Then, I review and evaluate some of the existing big datasets in developmental science.
+I conclude by discussing key questions that big data approaches pose for the future of the field.
 
-In fact, big data analyses are not especially new in developmental science.
-The field tackles questions that benefit substantially from larger, richer, more widely shared, and more readily inter-operable datasets.
-Big data does not signal the end of theory \cite{anderson_end_2008}, nor will it necessarily revolutionize scientific understanding \cite{boyd_critical_2012}. 
+We will see that big data analyses in developmental science are not especially new.
+The field tackles questions that have benefited and will continue to benefit from large, rich, widely shared, and readily inter-operable datasets.
+So, big data approaches to development do not signal the end of theory \cite{anderson_end_2008}, nor will they necessarily revolutionize scientific understanding \cite{boyd_critical_2012}. 
 Rather, significant novel insights emerging from the era of big data will depend not just on the size, density, and complexity of the datasets, but on how widely and openly data are shared, and on how readily researchers are able to combine or link datasets across levels of analysis.
@@
 
-Even more important than the quantity of information stored is the kind of information.
-Developmental science spans phenomena across multiple levels of analysis in space (from genes to geography) and time (from microseconds to millennia) all aimed at answering two questions: ``What develops?'' and  ``Why does it develop this way?''
-In seeking answers, developmental scientists have long recognized the importance of multiple, nested influences on developmental processes arising across scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
+Even more important than the quantity of information stored is the kind.
+Developmental science spans phenomena across multiple levels of analysis in space (from genes to geography) and time (from microseconds to millennia) all aimed at answering two questions: ``What develops?'' and  ``How does development occur''?
+In seeking answers, scientists have long recognized the importance of multiple, nested influences on developmental processes arising across scales \cite{elman_rethinking_1998,gottlieb_normally_1998,oyama_ontogeny_2000,vygotsky_mind_1980}.
 Most seek to describe change in the psychological processes of individuals, groups of individuals, or families.
 Data about neighborhoods, schools, and the broader social, cultural, political and environment primarily inform thinking about the development of individual or family behavior.
-Similarly, biological data---genes, hormones, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status---reveals the influence of within-person factors on changes in behavior.
+Similarly, biological data---genes, hormones, physiological responses, brain activity, body dimensions, brain structure, disease or disorder status--- are brought to bear to reveal the influence of within-person factors on changes in behavior.
 
@@
 Big data magnify the challenges facing researchers in maintaining participant privacy, in part because the more data that are collected, the more likely it is that individual identities can be discovered \cite{sweeney_identifiability}.
-Big data pose theoretical challenges: For example, how do micro-scale factors influence macro-scale phenomena?
+Big data pose theoretical challenges, too: For example, how do micro-scale factors influence macro-scale phenomena?
 Nevertheless, big data offer developmental researchers the opportunity to tackle some of the most profound and vexing questions in the behavioral science---if the relations among data components can be revealed in ways that do not undermine ethical research principles.
@@
 But, researchers may draw inspiration from examples of big datasets addressing developmental questions that have already been collected, and in some cases, widely shared.
+The next section highlights several.
+
 \section{Big Datasets in Developmental Research}
-Existing big datasets in developmental science fall into one of three broad categories depending on who collected the data and what mandates or restrictions apply to who may access it.
-Many of the largest existing sources come from data collected and disseminated by government entities such as the U.S. Census Bureau and the U.S. Department of Education.
+Existing big datasets in developmental science fall into one of several broad categories depending on who collected the data, what mandates or restrictions apply to who may access it, and the diversity of measures represented.
+
+\subsection{Government-Collected and Managed Datasets}
+Many of the largest existing sources come from data collected and disseminated by government entities such as the U.S. Census Bureau.
+Table 1 summarizes information about some representative large, developmentally-focused datasets whose collection and hosting is managed by governmental entities.
+Several themes emerge from this sample.
+Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
+Some studies involve dozens of samples per individual over extended periods of time, and show significant variety in the measures collected.
 The collection and dissemination of these datasets is mandated by law and subject to provisions of a complex set of statutory and regulatory requirements designed to protect individual respondents' identities.
 Government-collected datasets tend to be the most open and widely available to researchers, but by their very design and legal mandate the datasets are not intended for answering questions about individuals, families, or small groups.
-Datasets collected by individual researchers or teams of researchers employed by institutions, usually colleges or universities, who are funded by government or private foundations form a second group.
-These datasets tend to be somewhat smaller in size, but significantly more varied in terms of the types of information collected, the duration or intensity of data collection, the means of collection, the extent to which data are made available to individuals outside the research team, and the impact on scholarship.
-The extent to which datasets collected by non-government research teams are shared with other researchers depends on a set of factors including institutional research ethics and data privacy policies, funder and journal requirements, and individual researcher's support for data sharing.
-Large-scale datasets collected by private entities for business purposes form a third group.
-The data collected by private entities about individuals largely concern consumer behavior although health and fitness related data are an emerging area (e.g., Fitbit; \url{http://www.fitbit.com}).  
-Commercial datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the commercial entities that provide the services.
-\subsection{Government-collected and Managed Datasets}
-Population-based studies constitute some of the largest and most widely used datasets in developmental science.
-The National Health and Nutrition Examination Survey (NHANES; \url{http://www.cdc.gov/nchs/nhanes.htm}) began collecting data in 1959 on the health and nutritional status of U.S.
-infants, children, and adults.
-The annual survey combines interviews from a geographically diverse representative sample of about 5,000 people.
-NHANES focuses on demographic, socioeconomic, dietary and health-related topics, physical examinations involving medical, dental, and physiological measurements, and laboratory tests of behavior.
-The products include standardized child growth charts and the closely associated National Youth Fitness Survey (NYFS; \url{http://www.cdc.gov/nchs/nnyfs.htm}), which reports physical activity and fitness levels in 3 to 15-year-olds.
-Much of the NHANES and NYFS data are available to the public through the National Center for Health Statistics at the U.S. Centers for Disease Control.
-However, access to some data is restricted, available only to individuals via one of the 19 Federal Statistical Research Data Centers locations across the U.S.
-By mid-2015 more than 110,000 publications had cited the NHANES dataset.
+Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval, and in the U.S., possibly travel to a Federal Research Data Center where special security provisions apply.
 
-In contrast to the cross-sectional NHANES, The National Longitudinal Survey of Youth (NLSY; \url{http://www.bls.gov/nls/}) has surveyed the same cohort---almost 10,000 American youth born between 1957 and 1964--- about employment-related topics on a annual basis since the study began in 1979.
-Most NLSY and NLS data are available free of charge by means of a web-based data portal to which users must apply for access; access to geographic-related variables requires special permission.
-By mid-2015 more than 20,000 publications had cited the NLSY.
-
-The Early Childhood Longitudinal Study (ECLS; \url{https://nces.ed.gov/ecls/}) focuses on child development, school readiness, and early school experiences in separate cohorts of children followed either from birth or from kindergarten.
-Some cohorts exceed 20,000 individuals.
-ECLS consists of both publicly available and restricted data archived at ICPSR.
-A Google Scholar search generates more than 8,000 hits, of which the ICPSR archive can verify 167 project-related publications.
-
-Several large government-collected and shared datasets contain information about employment patterns and educational attainment in children and youth.
-The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (\url{http://www.bls.gov/cps/demographics.htm}).
-The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes variables related to U.S. primary, secondary, and post-secondary education.
-Many datasets are public while data with personally identifiable data are available to researchers under a restricted use agreement.
-
-Several large-scale longitudinal studies focused on children's growth, health, and education include participants from outside the U.S.
-The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS; \url{http://www.who.int/childgrowth/en/}) collected physical growth and related data from 8,500 children in Brazil, Ghana, India, Norway, Oman, and the U.S.
-The ongoing Japan Environment and Children's Study (JECS; \url{http://www.env.go.jp/en/chemi/hs/jecs/}) involves 100,000 parent-child pairs with the goal of evaluating the impact of environmental factors on children's health and development.
-Large-scale birth cohort studies in the U.K. include the National Survey of Health \& Development and the National Child Development Study (NCDS; \url{http://www.cls.ioe.ac.uk}).
-NCDS is an ongoing longitudinal study that follows everyone in Great Britain who was born in one particular week in 1958, focusing on factors that predict wellbeing across the lifespan.
-NCDS data are available by registration from the U.K. Data Service (\url{http://ukdataservice.ac.uk/get-data/key-data/cohort-and-longitudinal-studies}).
-The British Cohort Studies such as NCDS have generated more than 3,500 publications to-date.
-The Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset (\url{http://pisa2000.acer.edu.au/index.php}), which generates more than 5,000 search hits.
-
-Not all large-scale birth cohort studies succeed.
+Not all large-scale government-initiated studies succeed.
 A notable failure is the U.S. National Children's Study (\url{http://www.nichd.nih.gov/research/NCS/Pages/default.aspx}).
@@
 Those data are slated for release in a data archive sometime in 2015.
+A comparable study in the U.K.
+targeting 80,000 was cancelled in October 2015, just 8 months after launch, for failures to recruit sufficient numbers of participant families \cite{pearson_massive_2015}.
+These failures highlight the significant challenges associated with designing and successfully implementing large-scale birth-cohort studies.
 
-Several themes emerge from this sample of government-led studies.
-Datasets are large in terms of the volume of participants sampled, ranging in the thousands to tens of thousands.
-Some studies involve dozens of samples per individual over extended periods of time, and show significant variety in the measures collected.
-Most data are available to the public via web-based download or browsing/analysis portal, although access to data deemed sensitive may require a specific application and approval and possibly travel to a Research Data Center where special security provisions apply.
-Finally, the datasets have generated research publications that range from the hundreds to the tens of thousands, suggesting that the collection, curation, and preservation of these resources has had a significant impact on scientific discovery.
-I turn next to datasets whose collection and management is under the control of individual investigators.
-\subsection{Public/Private Partnerships and Investigator-driven Projects}
-Several big data studies in developmental science involve partnerships between government agencies and academic investigators.
-Others involve large-scale data collections initiated and managed by individual investigators and funded by government or foundation sources.
-These projects are diverse in focus and methods, including population-based studies with a health focus, behavior genetics, brain imaging, cognition and temperament, and education and employment.
-\subsubsection{Population-Based Studies}
-The Panel Study on Income Dynamics (PSID; \url{https://psidonline.isr.umich.edu}) is a population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
-PSID began in 1968 with a nationally representative sample of 5,000 families and more than 18,000 individuals.
-As of mid-2015, PSID had generated more than 3,900 publications. 
-The data is available publicly to registered users who agree to the terms of use conditions that ensure participants are not reidentified and that users properly cite and acknowledge use of the data. 
-Some restricted data elements (e.g., geospatial data, school-level identifiers) are only available subject to a formal restricted use agreement between the investigator and the data owner, the University of Michigan.
+\subsection{Researcher-Initiated and Managed Datasets}
+Datasets initiated and collected by academic or medical researchers form a second group.
+Table 2 summarizes information about some representative large, developmentally-focused datasets whose collection was initiated by individual researchers, and the data themselves are managed by non-governmental entities.
+These tend to be smaller than those collected by government agencies, but the data collected are more varied in type, means of collection, and duration or intensity.
+For example, investigator-initiated studies commonly collect observational measures, including video recordings, population-normed test instruments, biological measurements of physiology, genetics, and brain structure or function.
+Unfortunately, the extent to which these data are available for secondary reuse, and the process for acquiring access is also more variable than for datasets initiated and managed by government entities.
+Institutional research ethics and data privacy policies, funder and journal requirements, and individual researchers' support for data sharing influence whether, when, and with whom data are shared.
+In many cases, the original investigative team retains control over the use of data by other researchers, including the kinds of questions that third parties may ask.
+Some require the original investigative team to be included as an author on publications.
+Most large-scale investigator-initiated developmental datasets are housed locally, on project-specific web sites, not on centralized servers that aggregate data across studies and sources.
+Only some are stored in open public data repositories, for example.
+Catherine Tamis-Lemonda's MetroBaby dataset \cite{1372d9e6-d07e-4068-8264-7a5c629799c2} hosted on Databrary is a notable exception.
 
-The National Longitudinal Study of Adolescent Health (Add Health; \url{http://www.cpc.unc.edu/projects/addhealth}) began as a survey study undertaken in response to a Congressional mandate.
-Designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors, parts of the Add Health cohort have been followed into young adulthood with four in-home interviews.
-In a recent wave, biomarker data have been made available. 
-The Add Health data are archived at ICSPR and are available for public use.
-As of mid-2015, more than 5,000 publications had arisen from the Add Health dataset.
+\subsection{Measure-Specific Data}
+Datasets representing a single test or form of measurement constitute another group.
+Table 3 summarizes information about some measures commonly used in developmental science research and datasets created around them.
+Many measures in this category derive from the use of standardized instruments with group norms.
+It is considered best-practice in many research communities to employ widely adopted, standardized behavioral tests with well-characterized psychometric properties and developmental, usually age-based, norms.
+This allows researchers to compare patterns of performance between groups.
+Perhaps surprisingly, most of the raw data underlying the norms remain private.
+So, with few exceptions, researchers seeking access to measure-specific data collected by others will find it almost impossible.
+A number of standardized measures are published by commercial entities, and so economic interests may conflict with the ideal of greater data availability.
+However, widespread data sharing remains relatively rare even where measures developed by academic researchers and made freely available are concerned.
+Data sharing initiatives among child language researchers (CHILDES; WordBank; HomeBank) are notable exceptions.
 
-The NICHD Study of Early Child Care and Youth Development (SECCYD; \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx}) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
-The study sought to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
-Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, anthropometrics, and demographic information.
-Unlike the Add Health study, data in SECCYD are restricted, available only by application, with approval granted on a case-by-case basis.
-Staff at the National Institute of Child Health and Human Development manage the data access and approval process.
-The SECCYD has generated more than 214 publications as of mid-2015.
+\subsection{Commercial Datasets}
+Large-scale datasets collected by private entities for business purposes form a final group.
+The data collected by private entities about individuals largely concern consumer behavior although health and fitness related data constitute a third category (e.g., Fitbit; \url{http://www.fitbit.com}).
+A growing number of smartphone apps enable parents to collect data on their own children for personal purposes (e.g. \url{https://www.baby-connect.com}).
+Commercial datasets are sometimes made available to academic researchers, but the policies that govern data access are under the control of the entities that provide the services.
 
-Investigator-initiated and managed, the Family Life Project (FLP; \url{http://flp.fpg.unc.edu}) used a population-based approach to study the early development of children living in rural, largely poor communities in North Carolina and Pennsylvania.
-A birth cohort of 800 children were studied in a series of home, child care visits, and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
-Data included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
-Data from Phase I of the FLP are archived at ICPSR, but are subject to restrictions on access due to the sensitive and identifiable nature of the data.
-The FLP site lists about 50 publications through 2013.
-\subsubsection{Behavior Genetics}
-Large data samples have long been required in behavior genetics to detect small effect sizes.
-Several large-scale genetics studies that are not specifically developmental deserve mention.
-The Genome of the Netherlands Project (\url{http://www.nlgenome.nl}) is a publicly available dataset from 250 families representing two parents plus one of their adult children.
-The Psychiatric Genomics Consortium (PGC; \url{http://www.med.unc.edu/pgc}) contains data from more than 170,000 individuals with psychiatric diagnoses or who are at-risk.
-Many disorders in the PGC dataset have developmental dimensions (autism, attention-deficit hyperactivity disorder, and schizophrenia).
-Results may be viewed in a specialized web browser tool (\url{http://www.broadinstitute.org/mpg/ricopili/}).
-
-The U.K.-based Twins Early Development Study (TEDS; \url{http://www.teds.ac.uk/}) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996, with lab-based behavioral task and DNA samples from a subset of participants.
-TEDS has generated more than 350 publications to-date.
-Researchers wishing to gain access to the data may do so by submitting a formal data request specifying the aims of the planned research and the variables needed to satisfy the aims. 
-Proposals that do not overlap with analyses already being planned or carried out by the TEDS research team or other collaborators are usually approved.
-
-The Twin and Offspring Study in Sweden (TOSS; \url{http://ki.se/en/meb/twin-offspring-study-in-sweden-toss}) aimed seeks to uncover genetic and environmental contributions to measures of family relationships and mental health.
-TOSS extends the Twin Moms Project to include a sample of twin fathers (320 twin pairs and their families).
-In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (3,000 individuals in all) and their families.
-While relatively early in data collection, more than 24 publications reference the TOSS dataset.
-Data sharing plans have not yet been announced.
-
-Among behavior genetics studies, those focusing on adoption provide a unique window on the mix of genetic and non-genetic factors that influence child development.
-The Colorado Adoption Project (CAP; \url{http://ibg.colorado.edu/cap}) provides an illustrative example.
-Since its inception in 1976, CAP has enrolled more than 2,400 participants from more than 450 families, collecting data on cognitive abilities, temperament, and demographics.
-CAP has generated more than 200 published articles, and data collection continues.
-Data from 1976-1989 are archived at Dataverse and can be accessed by application.
-\subsubsection{Brain Imaging}
-Several large-scale studies have focused on developmental patterns in brain structure or activity.
-One of the earliest studies was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
-The study collected structural brain information from 145 4 to 20-year-olds along with behavioral and clinical data.
-More than 100 of the participants were scanned more than once, about two years apart.
-The original paper describing this study has been cited more than 3,300 times, as of mid-2015.
-The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data are not available for analyses by researchers outside the original investigative team.
-
-The NIH MRI Study of Normal Brain Development (\url{http://pediatricmri.nih.gov/nihpd/info}) collected multi-modal structural MRI from 554 children from 4 to 18 years of age.
-A second cohort of children from birth to 4 years was scanned longitudinally with up to 10 scans per child.
-Demographic, hormonal, cognitive, affective, and psychiatric data were also collected.
-The data are available to qualified researchers by application.
-John Richards and colleagues have combined data from the NIH MRI Study with data collected in their own labs and other public sources to create average structural brain templates that can provide more accurate bases for developmental functional neuroimaging studies using EEG and fMRI (\url{http://jerlab.psych.sc.edu/NeurodevelopmentalMRIDatabase/}).
-These data can be openly accessed with appropriate citation.
-
-More recently, the Pediatric Imaging, Neurocognition, and Genetics (PING; \url{http://pingstudy.ucsd.edu/}) Project collected multi-modal neuroimaging data, genotypes, neurodevelopmental histories, and information about cognitive and social and emotional function from more than 1,000 participants 3-20 years of age recruited from 8 U.S.
-cities.
-Data are available to the research community by application.
-As of mid-2015, the project had generated more than 20 publications.
-
-Outside the U.S., The Developing Human Connectome Project (\url{http://www.developingconnectome.org}), led by investigators at King’s College London, Imperial College London, and Oxford University goes against the trend of imaging studies that focus on older children.
-This project aims to study the connectome, a map of human brain connectivity, in fetuses from 20 to 44 weeks post-conception.
-Funded by the European Research Council, the imaging data are accompanied by clinical, behavioral and genetic information.
-Investigators hope that the data set will provide a basis for studying genetic and environmental risks that could lead to neurodevelopmental disorders such as Autistic Spectrum Disorder or Cerebral Palsy.
-The project has generated 17 publications as of mid-2015.
-\subsubsection{Language, Cognition, and Temperament}
-The child language community pioneered the aggregation and sharing of developmental datasets about individual participants.
-CHILDES/TalkBank \cite{macwhinney_childes_2001} supports one of the largest and most well-established data archives in the psychological sciences.
-It consists of transcripts and audio and video recordings of children's utterances and from adults with aphasia.
-CHILDES/TalkBank has generated more than 10,000 citations, and the data are available to the research community, much of it publicly.
-
-WordBank (\url{http://wordbank.stanford.edu/}) consists of data from more than 40,000 samples of the MacArthur-Bates Communicative Development Inventory (CDI), a widely used test of early language function.
-Catherine Tamis-Lemonda has released video data (\url{https://nyu.databrary.org/volume/8}) from an NSF-funded longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
-The Human Speechome Project (\url{http://www.media.mit.edu/cogmac/projects/hsp.html}) at MIT's Media Lab recorded 10 hours of video from one child's home on a daily basis from birth to age three.
-The project has generated more than 78 publications, but the data are not available outside of the original investigative team.
-The LENA (Language ENvironment Analysis; \url{http://www.lenafoundation.org}) Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed.
-The system provides an automatic language collection and analysis tool for speech language professionals and parents.
-No large-scale archive for LENA data currently exists, but some researchers who use the tool have begun to explore the creation of a data repository specialized for it.
-
-In other areas of cognition, a number of measures that have become standards.
-Their widespread adoption has resulted in data that are large in volume, velocity or variety.
-Davida Teller \cite{teller1986assessment} pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
-A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.
-
-The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} is widely used in clinical and epidemiological settings such as the EPICure (\url{http://www.epicure.ac.uk}) because of its published norms even though there are questions about the test's validity \cite{hack_poor_2005}.
-However, data about those norms is under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
-Some Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE; \url{http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4091}) project on ICPSR.
-The data are freely available to registered users of ICSPR.
-
-The standardized Wechsler intelligence tests were developed in the 1930's. 
-The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6-16, and the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4-6 1/2 years.
-The current version (WISC-V) is published by Pearson (\url{http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html}).
-Like the Bayley, data about Wechsler norms are not available to researchers.
-However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is housed at ICSPR and are accessible by specific authorization.
-
-In the realm of emotion and temperament, two academic-investigator-initiated measures have been widely adopted.
-The Infant Behavior Questionnaire (IBQ) is a parent survey measure, and the Laboratory Temperament Assessment Battery (Lab-TAB; \url{http://www.uta.edu/faculty/jgagne/labtab}) is an observational battery, using a set of 3-5 min episodes mimicking situations in everyday life.
-IBQ data on 1,388 participants are archived at ICSPR as part of the Maternal Lifestyle Study (MLS; \url{https://neonatal.rti.org/about/mls_background.cfm}), a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
-IBQ and Lab-TAB data are also available from ICPSR from the Family Life Project.
-\subsubsection{Education and Learning Science}
-Several studies in the education and learning sciences studies deserve mention because of their volume, velocity, or variety or because of their availability for reuse by others.
-The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years (\url{http://timss.bc.edu}), with more than 20,000 U.S.
-students included in the 2001 sample.
-De-identified data from TIMSS and its sister study, the Progress in International Reading Literacy Study (PIRLS; \url{http://timssandpirls.bc.edu}) are available for public download.
-The 1999 TIMSS video study recorded math and science teaching practices in seven countries, and some of the materials are available for public use through a repository hosted at UCLA (\url{http://www.timssvideo.com}).
-The Gates Foundation Measures of Effective Teaching (MET; \url{http://www.metproject.org}) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing currently.
-The MET video data are stored and shared at ICSPR.
-The Spatial Intelligence and Learning Center (SILC; \url{http://spatiallearning.org}) is an NSF-funded initiative designed to develop a multi-disciplinary science of spatial learning, including the development of tests and instruments useful in research.
-LearnLab (\url{http://learnlab.org}), another NSF-funded center, combines cognitive theory and computational modeling to understand changes in student knowledge in the context of computer-based math, science and language courses.
-LearnLab hosts a repository, DataShop (\url{http://learnlab.org/technologies/datashop/index.php}), to store and openly share detailed data from classrooms employing computer-based cognitive tutors.
-DataShop enables data storage, management, visualization and analysis in a web-based tool.
-\subsubsection{Summary}
-Studies involving partnerships with government agencies or those initiated and managed by individual investigators exemplify the high volume, velocity, variety, and variability of big datasets.
-The studies employ a range of observational measures, including video recordings, population-normed test instruments, biological measurements of physiology, and brain structure or function.
-However, the extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.
-Only some of the datasets are stored in data repositories, for example.
-Many large-scale developmental datasets are housed locally, on project-specific web sites, not on centralized servers that aggregate data across studies and sources.
-In many cases, the original investigative team retains control over the use of data by other researchers, including the kinds of questions that third parties may ask.
-In some cases, the original investigative team must be included as an author on publications.
-Perhaps as a consequence of these factors, these datasets have generated fewer scientific publications than government-led efforts.
-
-\subsection{Non-Academically Initiated or Managed Sources of Big Data}
 Some large-volume sources of developmental data are collected and managed by private, non-academic or government entities.
@@
 The primary criticism concerns whether Facebook users had given informed consent to participate in the manipulation of their newsfeeds as would be required by research ethics boards if a similar study were undertaken in a laboratory context.
+Clearly, the scale of data collected and managed by non-academic entities dwarfs that of other providers.
+Because the data are collected for proprietary business purposes, it is difficult to assess their current or potential impact on the scholarship of human development.
 
-The scale of data collected and managed by non-academic entities dwarfs that of government or academically-managed initiatives.
-Because the data are collected for proprietary business purposes, it is difficult to assess their current or potential impact on the scholarship of human development.
 \section{The Future of Big Data in Development}
-Clearly, the collection, analysis, and sharing of large, multilevel datasets has been part of the fabric of developmental science for a long time.
+Clearly, the collection, analysis, and sharing of large datasets has been part of the fabric of developmental science for a long time.
 In this section, I discuss a range of technical, conceptual, and theoretical issues that arise in thinking about the future of big data in developmental science. 
@@
 So, the output data files, while often in an electronic form, may require significant post-processing to be linked with other data.
-Some researchers have begun to use tools such as Amazon's Mechanical Turk (\url{http://www.mturk.com}) or Apple's HealthKit (\url{https://developer.apple.com/healthkit/}) to conduct large-scale behavioral science experiments.
-These sites deliver data in well-structured electronic formats, sometimes using tools specialized for psychological research (e.g., \url{https://psiturk.org}).
+Some researchers have begun to use tools such as Amazon's Mechanical Turk (\url{http://www.mturk.com}) or Apple's HealthKit (\url{https://developer.apple.com/healthkit/}) to conduct large-scale behavioral science experiments (e.g., \url{https://autismandbeyond.researchkit.duke.edu}).
+These sites deliver data in well-structured electronic formats, sometimes using tools specialized for psychological research (e.g., PsiTurk, \url{https://psiturk.org}).
 Amazon's terms of use prohibit minors, but developmental researchers have found ways to secure video-based informed consent from parents to enable their children to participate in looking time studies (\url{https://lookit.mit.edu}) over the web.
@@
 
-New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promise to make big data streams about individual participants' locations, physiological states (e.g., \url{https://www.empatica.com}), activity patterns, and momentary cognitive, and emotional states broadly available to researchers.
-These tools will enable the collection of data from large numbers of participants in short periods of time, significantly enlarging the volume, velocity, and variety of data available for analysis.   
+New technologies, specifically the widespread use of smart mobile devices with embedded sensors, promise to make big data streams about individual participants' locations, physiological states (e.g., \url{https://www.empatica.com}, \url{https://autismandbeyond.researchkit.duke.edu}), activity patterns, facial expressions, and momentary cognitive, and emotional states broadly available to researchers.
+For example, a new class of wearable devices for infants and children has arisen (e.g., \url{https://www.owletcare.com}, \url{http://mimobaby.com}, \url{http://www.sproutling.com}), coupled to parent-controlled child tracking apps mentioned previously.
+These tools enable the collection of data from large numbers of people in short periods of time, significantly enlarging the volume, velocity, and variety of data available for analysis.
+Whether and how the data can be made available for academic and medical research in developmental science remains an open question.
+
 \subsubsection{Storage and Retrieval}
@@
 Because data security cannot ever be guaranteed, risks can only be minimized and managed, but not entirely eliminated.
-Finally, there are unresolved questions about privacy protections in the consumer domain that have the potential to influence public attitudes toward academic research \cite{meyer_everything_2014}.
+The DataTags Project (\url{http://datatags.org} at Harvard provides example of a practical solution that may help researchers navigate the complexities of sharing data in the future. 
+DataTags seeks to make the process of determining what risks particular datasets pose and provide a practical way of ``tagging'' datasets based on that level of risk.
+
+Of course, there are unresolved questions about privacy protections in the consumer domain that have the potential to influence public attitudes toward academic research \cite{meyer_everything_2014}.
 \subsubsection{Transparency and Reproducibility}
@@
 Researchers should work to reduce the extent of blame levied at researchers whose initial positive findings fail to be replicated by others \cite{bishop-blog-2015}.
-Technological tools that foster increased openness and transparency and more systematic research data management (OSF and Databrary) will also contribute to changing scientific practices.
-So will the widespread adoption of more consistent journal practices related to transparent and open scientific practices \cite{nosek_promoting_2015}.
+Technological tools that foster increased openness and transparency and more systematic research data management (OSF and Databrary) will also contribute to changing scientific practices, as will the widespread adoption of more consistent journal practices related to these issues \cite{nosek_promoting_2015}.
 
@@
 Establishing consensus about data ownership and the kind of control investigators can exercise over it will require conversations among researchers, institutions, and funding agencies.
-That consensus may well prove vital to achieving some of the benefits of big data analyses in development. 
+That consensus may well prove vital to achieving some of the benefits of big data analyses in developmental science. 
 \subsection{Conceptual and Theoretical Issues}
@@
 
-A second challenge larger datasets may help to address is the fact that most social, behavioral \cite{maxwell_persistence_2004} and neuroscience studies \cite{button_power_2013} are underpowered.
-Most worryingly many published research findings are false in fields that rely on small sample sizes, test multiple relationships between variables, engage in exploratory research, use diverse research designs, definitions, outcomes, and analytical modes across studies, and when more labs seek out significant effects \cite{ioannidis_why_2005}.
+Another challenge larger datasets may help to address is the fact that most social, behavioral \cite{maxwell_persistence_2004} and neuroscience studies \cite{button_power_2013} are underpowered.
+Most worryingly, many published research findings are false in fields that rely on small sample sizes, test multiple relationships between variables, engage in exploratory research, use diverse research designs, definitions, outcomes, and analytical modes across studies, and when more labs seek out significant effects \cite{ioannidis_why_2005}.
 Developmental research reflects many of these characteristics, but the collection, analysis, and sharing of larger datasets should work to reduce their impact.
 
-Developmental research faces a specific point of tension related to measurement.
+Developmental research based on big data faces a specific point of tension related to measurement.
 Many of the measures for which high volume data are available come from proprietary, expensive instruments such as the Bayley and the WIPPSI for which baseline data about population norms are unavailable.
@@
 
-Finally, some critics have raised concerns that the rise of big data means the ``end of theory'' \cite{anderson_end_2008}.
+Although new technologies make it possible to carry out large-scale experimental studies with developmental populations (e.g., LookIt, PsiTurk), big data techniques often invoke some form of correlational analysis.
+This makes causal inference problematic at best.
+Indeed, some critics have raised concerns that the rise of big data means the ``end of theory'' \cite{anderson_end_2008}.
 In a provocative essay Anderson \citeyear{anderson_end_2008} argued that large quantities of data mean the traditional model of scientific inquiry involving hypothesis testing will soon give way to model-free descriptions of data.
 Others note that bigger data do not necessarily lead to deeper insights \cite{graham_big_2012}.
-While some data intensive fields, largely in computer science, have adopted theory-free approaches to discovery, developmental science has a rich and rigorous intellectual history in which theory and experiment play central, essential role in scholarly discourse.
-That tradition is likely to continue.
+Some data intensive fields, largely in computer science, have adopted theory-free approaches to discovery.
+But, developmental science has a rich and rigorous intellectual history in which theory, correlational analyses, and experiments play central, essential roles in scholarly discourse.
+It`s vital that tradition continue.
 \section{Conclusion}
@@
 \bibliography{big-data}
-
+\input{table-1.tex}
+\input{table-2.tex}
+\input{table-3.tex}
 \end{document}

### 2015-11-28 wiley-big-data-devel/table-1.tex
ms revision

% Government-hosted databases Datasets table
\begin{table}[h!]
\caption{Illustrative big datasets hosted by governmental entities}
\tiny
\centering
\begin{tabular}{| p{2.1in} | p{2.1in} | p{2.1in} | } 
\hline
Dataset & Sample & URL \\
\hline\hline
Current Population Survey (CPS) & & \url{http://www.bls.gov/cps/demographics.htm} \\
\hline
Danish National Birth Cohort & n=100,000 women recruited 1999-2002, assessed during pregnancy and when children were 6 mos, 18 mos, 7 yrs, and 11 yrs & \url{http://www.ssi.dk/English/RandD/Research%20areas/Epidemiology/DNBC/} \\
\hline
Early Childhood Longitudinal Study (ECLS) & Birth cohort (ECLS-B): n=14,000 children born in 2001 from birth through kindergarten; ECLS-K cohort: kindergarten through 8th grade; ECLS-K:2011: kindergarten through 5th grade. & \url{http://nces.ed.gov/ecls/} \\
\hline
Japan Environment and Children`s Study (JECS) & n=100,000 parent-child pairs & \url{http://www.env.go.jp/en/chemi/hs/jecs/} \\
\hline
National Health and Nutrition Examination Survey (NHANES) & nationally-representative sample of n=5,000  & \url{http://www.cdc.gov/nchs/nhanes.htm} \\ 
\hline
National Longitudinal Survey of Youth (NLSY) & n=10,000 youth born 1957-1964 surveyed beginning in 1979. & \url{http://www.bls.gov/nls/} \\
\hline
National Youth Fitness Survey (NYFS) & n=1,650 3-15 year-olds surveyed in 2012 & \url{http://www.cdc.gov/nchs/nnyfs.htm} \\
\hline
NICHD Study of Early Child Care and Youth Development (SECCYD) & n=1,364 infants born in 1991 & \url{https://www.nichd.nih.gov/research/supported/Pages/seccyd.aspx} \\
\hline
NIH MRI Study of Normal Brain Development & n=554 4-18 year-olds & \url{http://pediatricmri.nih.gov/nihpd/info} \\
\hline
NIMH National Database for Autism Research (NDAR) & n>85,000 individuals & \url{http://ndar.nih.gov} \\
\hline
Norwegian Mother and Child Cohort Study (MoBa) & n>90,000 pregnant women and n>70,000 men, recruited 1999-2008 & \url{http://www.fhi.no/eway/} \\
\hline
WHO Multicentre Growth Reference Study (MGRS) & n=8,500 0-24 mo-olds and 18-71 mo-olds & \url{http://www.who.int/childgrowth/en/} \\
\hline
\end{tabular}
\label{table:1}
\end{table}

### 2015-11-28 wiley-big-data-devel/table-2.tex
ms revision

% Datasets hosted by non-governmental entities
\begin{table}[h!]
\caption{Illustrative big developmental datasets hosted by non-governmental entities.}
\tiny
\centering
\begin{tabular}{| p{2.1in} | p{2.1in} | p{2.1in} | } 
\hline
Dataset & Sample & URL \\
\hline\hline
National Longitudinal Study of Adolescent Health (AddHealth) & Wave I: n=90,118 adolescents in-school questionnaires, n=20,745 in-home interviews; Wave II: n=14,738 adolescents in-home interviews; Wave III
n=15,197 young adults in-home interviews and biomarker collection; Wave IV:
15,701 adult in-Home interviews and biomarker collection. & \url{http://www.cpc.unc.edu/projects/addhealth}\\
\hline
Child Language Data Exchange System (CHILDES) & n>3,000 6 mos - 8 yrs & \url{http://childes.psy.cmu.edu/} \\
\hline
Colorado Adoption Project (CAP) & n=450 families (>2,400 individuals) & http://ibg.colorado.edu/cap/ \\
\hline
Databrary & >3,400 hours of video from n>2,900 participants, 2 mos - 50+ yrs; experimental displays & \url{http://databrary.org} \\
\hline
Developing Human Connectome Project & 20 to 44 weeks post-conception & \url{http://www.developingconnectome.org} \\
\hline
Family Life Project (FLP) & n=800 sampled at 2, 6, 15, 24, and 36 mos & \url{http://flp.fpg.unc.edu} \\
\hline
Genome of the Netherlands Project (GoNL) & n=250 familes & \url{http://www.nlgenome.nl} \\
\hline
Human Speechome Project & 10 hrs video/day from 1 child from 0-3 yrs & \url{http://www.media.mit.edu/cogmac/projects/hsp.html} \\
\hline
Maternal Lifestyle Study (MLS) & n=1,388 infants (and mothers/caregivers) assessed at 4, 8, 10, 12, 18, 24, 30, and 36 mos & \url{https://neonatal.rti.org/about/mls_background.cfm} \\
\hline
Measures of Effective Teaching Project (MET) & n>3,000 teachers recorded beginning in 2009 & \url{http://www.metproject.org} \\
\hline
MetroBaby & n>1,000 racially and ethnically diverse children video-recorded at home or in the lab at 14, 24, 36, 52, 64, and 76 months & (\url{http://doi.org/10.17910/B7CC74} \\
\hline
Panel Study of Income Dynamics (PSID) & n=5,000 families (>18,000 individuals) & \url{https://psidonline.isr.umich.edu} \\
\hline
Pediatric Imaging, Neurocognition, and Genetics (PING) & n=1,000 3-20 year-olds & \url{http://pingstudy.ucsd.edu} \\
\hline
Progress in International Reading Literacy (PIRLS) & & \url{http://timssandpirls.bc.edu} \\
\hline
Psychiatric Genomics Consortium (PGC) & n>900,000 & \url{http://www.med.unc.edu/pgc} \\
\hline
Trends in International Math and Science Study (TIMSS) & n>20,000 per sample & \url{http://timss.bc.edu} \\
\hline
Trends in International Math and Science Study Video (TIMSSVideo) & teachers recorded in n=7 countries & \url{http://www.timssvideo.com} \\
\hline
Twins Early Development Study (TEDS) & survey data from n=15,000 twins born in 1994-1996; lab-based behavioral task and DNA samples from a smaller subset & \url{http://www.teds.ac.uk} \\
\hline
Twin and Offspring Study in Sweden (TOSS) & target of n=900 parents and families (~3,000 individuals) & \url{http://ki.se/en/meb/twin-offspring-study-in-sweden-toss} \\
\hline
\end{tabular}
\label{table:2}
\end{table}

### 2015-11-28 wiley-big-data-devel/table-3.tex
ms revision

\begin{table}[h!]
\caption{Illustrative measure-based datasets.}
\tiny
\centering
\begin{tabular}{| p{3.25in} | p{3.25in} | }
\hline
Measure & Comments \\
\hline\hline
Bayley Scales of Infant and Toddler Development & \cite{bayley2006bayley}. Data underlying published norms not available, but some study-specific data is available: \url{http://www.epicure.ac.uk} and \url{http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4091}. \\
\hline
\hline
Computer-based cognitive tutor data & LearnLab (\url{http://learnlab.org/technologies/datashop/index.php}) hosts DataShop repository/data analysis tool. \\
\hline
Infant Behavior Questionnaire (IBQ) & \cite{rothbart1981measurement}. Information at \url{http://www.bowdoin.edu/~sputnam/rothbart-temperament-questionnaires/instrument-descriptions/infant-behavior-questionnaire.html}. IBQ data on n=1,388 participants archived at ICSPR as part of the Maternal Lifestyle Study (MLS; \url{https://neonatal.rti.org/about/mls_background.cfm}) \\
\hline
Laboratory Temperament Assessment Battery (Lab-TAB) & \cite{goldsmith1993laboratory}. Information available at \url{http://www.uta.edu/faculty/jgagne/labtab} \\
\hline
MacArthur Communicative Inventory (CDI) & n>40,000 samples archived at WordBank (\url{http://wordbank.stanford.edu}) \\
\hline
Spatial Learning & NSF-funded Spatial Intelligence and Learning Center (SILC; \url{http://spatiallearning.org}) stores information about research tests and intruments \\ 
\hline
Teller Acuity Cards & Publications citing norms for 0-4 year-olds have been published \cite{teller1986assessment, mayer_monocular_1995, salomao_large_1995} \\
\hline 
Wechsler Intelligence Scale for Children-III (WISC-III) & Designed for children ages 6-16. Data underlying norms not available. Published by Pearson (\url{http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html})\\
\hline
Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) & Designed for children aged 4-6 1/2 years. Data underling norms not available. \\
\hline
\end{tabular}
\label{table:3}
\end{table}

### 2016-03-27 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
initial commit

---
title: "video-and-reproducible-science"
author: "Karen E. Adolph & Rick O. Gilmore"
date: "`r Sys.time()`"
output: 
  ioslides_presentation:
    widescreen: TRUE
    css: css/databrary.css
    logo: img/databrary-leaf-large.png
bibliography: bib/databrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Acknowledgements

- [NSF BCS-1238599](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1238599&HistoricalAwards=false)
- [NICHD U01-HD-076595](https://projectreporter.nih.gov/project_info_description.cfm?aid=8531595&icde=15908155&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC)
- [Society for Research in Child Development (SRCD)](http;//srcd.org)

## Video and Reproducible Workflows

## Reproducibility in Psychological Science

## Data Sharing Sharpens Focus on Substance

## Video Uniquely Captures Behavior

## Video Collection & Sharing Aids Reproducibility

- Methodological details not captured in write-ups
- Unequivocally & unambiguously demonstrates phenomenon

## Children's Scale Errors

<div class="centered">
<a href="http://doi.org/10.1126/science.1093567">
<img src="img/deloache.science.jpg" height=500px>

[[@deloache_scale_2004]](http://doi.org/10.1126/science.1093567)
</a>
</div>

## Children's Scale Errors

<div class="centered">
<video width="640" height="480" controls>
  <source src="https://nyu.databrary.org/slot/9850/-/asset/11552/download?inline=true" type="video/mp4">
Your browser does not support the video tag.
</video>
[[@b03c5c6c-31b8-445e-9e4f-8917b735bb53]](http://doi.org/10.17910/B7H019)
</div>

## [Databrary.org](http://databrary.org)

- Digital data library specialized for video
- Video/audio + participant/context metadata
- Restricted access for research/educational use

## How Databrary Aids Reproducibility

## Analysis Pipeline Using Databrary

## Scripting Video Data Analyses

- [Datavyu.org](http://datavyu.org)
- Video coding tool
- Java with Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts)
- Free, open source, <https://github.com/databrary/datavyu>

## Datavyu Script Example 1

- Checking/correcting errors

## Datavyu Script Example 2

- Bulk export for analysis

## Take Homes

- Video captures behavior
- Databrary securely stores, shares video
- Video + Databrary -> reproducible and transparent workflows
- Video + Datavyu -> reproducible and transparent workflows

## References {.smaller}

### 2016-03-27 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
Edit reproducible science slide

@@
 
-## Reproducibility in Psychological Science
+## Is Psychological Science Reproducible?
 
-## Data Sharing Sharpens Focus on Substance
+<div class="centered">
+<a href="http://cos.io">
+<img src="https://cos.io/static/img/icons/cos_wide.png">
+</div>
+</a>
+
+- Nosek et al., [[@collaboration_estimating_2015]](http://doi.org/10.1126/science.aac4716)
+- vs. Gilbert et al. [[@gilbert_comment_2016]](http://doi.org/10.1126/science.aad7243)
+- Sharing [data](https://osf.io/ezcuj/wiki/home/?_ga=1.260625677.1251838540.1458403228) focuses attention on substantive issues
 
@@
 <a href="http://doi.org/10.1126/science.1093567">
-<img src="img/deloache.science.jpg" height=500px>
+<img src="img/deloache.science.jpg" height=400px>
+</a>
 
 [[@deloache_scale_2004]](http://doi.org/10.1126/science.1093567)
-</a>
 </div>
 
-## Children's Scale Errors
+## Children's Scale Errors {.vcenter}
 

### 2016-04-08 psu-orp/video-sharing-psu.Rmd
psu orp

---
title: "video-data-sharing-psu"
author: Rick Gilmore
date: "`r Sys.time()`"
output:
  ioslides_presentation:
    css: gilmore.css
    incremental: false
    widescreen: true
    logo: img/databrary-leaf-large.png
---

## Sharing video data @ PSU

- Why data sharing?
- What is Databrary?
- Databrary access model
- PSU-specific issues

## Why data sharing?

- Funder mandates
    + [NIMH](http://grants.nih.gov/grants/guide/notice-files/NOT-MH-15-012.html)
    + *"Investigators are expected to share with other researchers, at no more than incremental cost and within a reasonable time, the primary data, samples, physical collections and other supporting materials created or gathered in the course of work under NSF grants."* [NSF Data Sharing Policy](https://www.nsf.gov/bfa/dias/policy/dmp.jsp)

## Why data sharing?

- Accelerates discovery
- Increases transparency, reproducibility
- Increases benefits of participation in research

## What is [Databrary](http://databrary.org)?

- Digital data library specialized for storing, managing, & sharing video
- Supported by [NICHD U01-HD-076595-01](https://projectreporter.nih.gov/project_info_description.cfm?aid=8531595&icde=15908155&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC) and [NSF BCS-1238599](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1238599&HistoricalAwards=false)
- Policies for sharing identifiable video a specific aim

## Databrary access model

- Goals
    + Encourage re-use, accelerate discovery
    + Protect participants
    + Protect data contributors

## Hurdles

- Video contains faces, voices, names spoken out loud, images of homes
- Blurring expensive, undermines value of reuse
- Images personally identifying information (PII) under [HIPAA](http://www.hhs.gov/hipaa/for-professionals/privacy/special-topics/de-identification/index.html)

## Approach

- Restrict access to authorized researchers
- Seek permission to share
- Release levels reflect disclosure risk

## Limit access to researchers with institutional authorization

- [Databrary Access Agreement](https://databrary.org/access/policies/agreement.html) (NYU <-> PSU)
- Legal agreement; @ PSU OSP signs
- Authorized investigators must be PIs; can approve affiliates
- [User guide](https://databrary.org/access/guide.html)
- [Institution guide](https://databrary.org/access/guide/institutions.html)

## Access agreement permits

- Non-/pre-research uses
- Research use with IRB approval
- Contributions with IRB approval

## Seek permission to share

- Build on informed consent, video/photo release practices
- Less ambiguous than checklists
- [Databrary template](https://databrary.org/access/policies/release-template.html)
- Sample [script](https://databrary.org/access/guide/investigators/release/asking/script.html) and [video](https://databrary.org/access/guide/investigators/release/asking/examples.html)

## Seeking permission to share

- Separate release for sharing from consent to participate
    + Recommend asking after
- Ask sharing permission within consent document
- Separate protocol for sharing

## Release levels reflect disclosure risk

- No sharing, data remains within IRB-approved team
    + Unreleased (didn't ask; info lost)
    + Private (refused)
- Sharing outside IRB-approved protocol
    + Authorized users
    + Excerpts
    + Public

## Databrary to-date

<div class="centered">
<img src="img/databrary.jpg" width=800px>
</div>

## PSU Issues

- How to ask, document permission to share
- Compromising ability to share by making future data uses too specific
- Data destruction requirements incompatible with data archiving
- Where can video data be stored
- Can ["old" data be shared](https://databrary.org/access/guide/investigators/release/grandfathering-data.html) if consent == Databrary release

## PSU Issues

- How identifiable are video data without [other identifiers](http://www.hhs.gov/ohrp/policy/cdebiol.html)?
- Who owns research data?

### 2016-04-29 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
Update slides

@@
 
+## “Video Doesn’t Lie"
+
+<div class=centered>
+**Reproducible Workflows With [Databrary](http://databrary.org)**
+
+Karen E. Adolph (NYU)</br>
+David Millman (NYU)</br>
+Rick O. Gilmore (Penn State)</br>
+
+<http://github.com/databrary/presentations/nyu-data-science-reproducibility-16/video-reproducibility>
+</div>
+
 ## Acknowledgements
@@
 - [Society for Research in Child Development (SRCD)](http;//srcd.org)
-
-## Video and Reproducible Workflows
 
@@
 ## Video Uniquely Captures Behavior
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/9840/-/asset/11193/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+[[@c7fbeef0-d8a9-4b99-9e17-f598507486b7]](http://doi.org/10.17910/B7H019)
+</div>
+
+https://nyu.databrary.org/slot/9840/-/asset/11193/download?inline=true
+
+## [[@c7fbeef0-d8a9-4b99-9e17-f598507486b7]](http://doi.org/10.17910/B7H019)
+
+<div class=centered>
+<a href="https://nyu.databrary.org/volume/33">
+<img src="img/frank-databrary.jpg" height=550px>
+</a>
+</div>
 
@@
 
-## How Databrary Aids Reproducibility
+## Databrary Improves Reproducibility
 
-## Analysis Pipeline Using Databrary
+- "Active" curation of data *as it's collected*
+- Organize, share, standardize participant metadata
+- Share materials, displays
+
+## Databrary Improves Reproducibility
+
+- External links to code repos
+- DOIs for shared datasets
+- Long-term preservation via partnership with NYU Libraries
 
@@
 
+- Most analyses of video use specialized tools
 - [Datavyu.org](http://datavyu.org)
-- Video coding tool
+- Free, open source, <https://github.com/databrary/datavyu>
 - Java with Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts)
-- Free, open source, <https://github.com/databrary/datavyu>
 
@@
 
-## Datavyu Script Example 2
-
-- Bulk export for analysis
-
 ## Take Homes
@@
 - Databrary securely stores, shares video
-- Video + Databrary -> reproducible and transparent workflows
-- Video + Datavyu -> reproducible and transparent workflows
+- Video + Databrary -> reproducible & transparent workflows
+- Video + Datavyu -> reproducible & transparent workflows
+- Databrary helps make psychological science more transparent & reproducible
 

### 2016-04-30 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
Add better URL to talk repo
Embed more videos, iframes

@@
 ---
-title: "video-and-reproducible-science"
-author: "Karen E. Adolph & Rick O. Gilmore"
+title: "Video Doesn't Lie"
+author: "Karen E. Adolph, David Millman, and Rick O. Gilmore"
 date: "`r Sys.time()`"
@@
 
-## “Video Doesn’t Lie"
+----
 
 <div class=centered>
+**Video Doesn't Lie:**</br>
 **Reproducible Workflows With [Databrary](http://databrary.org)**
@@
 
-<http://github.com/databrary/presentations/nyu-data-science-reproducibility-16/video-reproducibility>
+<http://github.com/databrary/presentations/nyu-data-science-reproducibility-16/>
 </div>
@@
 - Methodological details not captured in write-ups
-- Unequivocally & unambiguously demonstrates phenomenon
+- Unequivocally & unambiguously demonstrates phenomena
 
-## Children's Scale Errors
+----
 
@@
 
-## Children's Scale Errors {.vcenter}
+----
 
@@
 
+----
+
+<iframe src="http://databrary.org" width=700px height=550px>
+</iframe>
+
 ## [Databrary.org](http://databrary.org)
@@
 
-- "Active" curation of data *as it's collected*
-- Organize, share, standardize participant metadata
-- Share materials, displays
+- "Active" curation of data **as it's collected**
+- Organize, share, standardized participant metadata
+
+## [[@15586370-3088-4527-bd0e-88ec0d8d673c]](http://dx.doi.org/10.17910/B7WC7S)
+
+<iframe src="https://nyu.databrary.org/volume/2" width=750px height=500px>
+</iframe>
 
@@
 
+- Share materials, displays
 - External links to code repos
@@
 
-## Scripting Video Data Analyses
+## Getting Permission to Share
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://databrary.org/video/example-video-1.mp4" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+<https://databrary.org/access/guide.html>
+</div>
+
+## Reproducible video coding workflows
 
@@
 
-## Datavyu Script Example 1
+----
 
-- Checking/correcting errors
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="http://datavyu.org/user-guide/_static/video/run-a-script.mp4" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+</div>
 
@@
 - Video + Databrary -> reproducible & transparent workflows
+
+## Take homes
+
 - Video + Datavyu -> reproducible & transparent workflows

### 2016-04-30 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
One more iframe

@@
 
-<iframe src="http://databrary.org" width=700px height=550px>
+<iframe src="http://databrary.org" width=640px height=480px>
 </iframe>
@@
 
-<div class="centered">
-<video width="640" height="480" controls>
-  <source src="http://datavyu.org/user-guide/_static/video/run-a-script.mp4" type="video/mp4">
-Your browser does not support the video tag.
-</video>
-</div>
+<iframe src="http://datavyu.org/" width=640px height=480px>
+</iframe>
 

### 2016-04-30 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
KEA comments

@@
 author: "Karen E. Adolph, David Millman, and Rick O. Gilmore"
-date: "`r Sys.time()`"
 output: 
@@
 
-----
+# Reproducible workflows with [Databrary](http://databrary.org)
 
-<div class=centered>
-**Video Doesn't Lie:**</br>
-**Reproducible Workflows With [Databrary](http://databrary.org)**
-
-Karen E. Adolph (NYU)</br>
-David Millman (NYU)</br>
-Rick O. Gilmore (Penn State)</br>
-
-<http://github.com/databrary/presentations/nyu-data-science-reproducibility-16/>
-</div>
-
-## Acknowledgements
+## Acknowledgments
 
@@
 - [NICHD U01-HD-076595](https://projectreporter.nih.gov/project_info_description.cfm?aid=8531595&icde=15908155&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC)
-- [Society for Research in Child Development (SRCD)](http;//srcd.org)
+- [Society for Research in Child Development (SRCD)](http://srcd.org)
+- NYU Libraries, ITS, IRB, and OSP
 
-## Is Psychological Science Reproducible?
+## Is psychological science reproducible?
 
@@
 
-- Nosek et al., [[@collaboration_estimating_2015]](http://doi.org/10.1126/science.aac4716)
+- Nosek et al. [[@collaboration_estimating_2015]](http://doi.org/10.1126/science.aac4716)
 - vs. Gilbert et al. [[@gilbert_comment_2016]](http://doi.org/10.1126/science.aad7243)
@@
 
-## Video Uniquely Captures Behavior
+----
+
+<iframe src="http://databrary.org" width=640px height=400px>
+</iframe>
+
+## [Databrary.org](http://databrary.org)
+
+- Digital data library specialized for research video
+- Video/audio + participant/context metadata
+- Restricted access for research/educational use
+- Policy framework for sharing identifiable data
+
+## Video uniquely captures behavior
 
 <div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/9840/-/asset/11192/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+[[@c7fbeef0-d8a9-4b99-9e17-f598507486b7]](http://doi.org/10.17910/B7H019)
+</div>
+
+## Video uniquely captures behavior
+
+<div class="centered">
+
 <video width="640" height="480" controls>
@@
 <a href="https://nyu.databrary.org/volume/33">
-<img src="img/frank-databrary.jpg" height=550px>
+<img src="img/frank-databrary.jpg" height=450px>
 </a>
@@
 
-## Video Collection & Sharing Aids Reproducibility
-
-- Methodological details not captured in write-ups
-- Unequivocally & unambiguously demonstrates phenomena
-
-----
+## Unequivocally & unambiguously demonstrates phenomena
 
@@
 <a href="http://doi.org/10.1126/science.1093567">
-<img src="img/deloache.science.jpg" height=400px>
+<img src="img/deloache.science.jpg" height=350px>
 </a>
@@
 <video width="640" height="480" controls>
-  <source src="https://nyu.databrary.org/slot/9850/-/asset/11552/download?inline=true" type="video/mp4">
+  <source src="https://nyu.databrary.org/slot/9850/-/asset/11550/download?inline=true" type="video/mp4">
 Your browser does not support the video tag.
@@
 
-<iframe src="http://databrary.org" width=640px height=480px>
+<iframe src="http://doi.org/10.17910/B7H019" width=640px height=400px>
 </iframe>
 
-## [Databrary.org](http://databrary.org)
+## Video depicts unreported methodological details
 
-- Digital data library specialized for video
-- Video/audio + participant/context metadata
-- Restricted access for research/educational use
+Fall from video, bridge 20 going into bridge 14, then fall. Subject 1. Cari brown sleeveless shirt. John assist.
+<https://nyu.databrary.org/volume/5>
 
-## Databrary Improves Reproducibility
+## [Datavyu](http://datavyu.org) makes video coding reproducible
 
-- "Active" curation of data **as it's collected**
-- Organize, share, standardized participant metadata
+- Raw research video must be coded
+- [Datavyu.org](http://datavyu.org) is a free, [open source](https://github.com/databrary/datavyu) coding tool
+- Add codes, annotations time-locked to video segments
+- Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts)
 
-## [[@15586370-3088-4527-bd0e-88ec0d8d673c]](http://dx.doi.org/10.17910/B7WC7S)
+## Reproducible video coding
 
-<iframe src="https://nyu.databrary.org/volume/2" width=750px height=500px>
-</iframe>
+Static image of fall from video.
 
-## Databrary Improves Reproducibility
-
-- Share materials, displays
-- External links to code repos
-- DOIs for shared datasets
-- Long-term preservation via partnership with NYU Libraries
-
-## Getting Permission to Share
+## Securing permission to share while limiting access
 
@@
 </video>
-
-<https://databrary.org/access/guide.html>
 </div>
 
-## Reproducible video coding workflows
+## Databrary enhances reproducibility
 
-- Most analyses of video use specialized tools
-- [Datavyu.org](http://datavyu.org)
-- Free, open source, <https://github.com/databrary/datavyu>
-- Java with Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts)
+- "Active" curation of data **as it is collected**
+- Organize, share, standardized participant metadata
+- Sharing based on access level & participant permission
 
-----
+## [[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
 
-<iframe src="http://datavyu.org/" width=640px height=480px>
+<iframe src="https://nyu.databrary.org/volume/5" width=750px height=500px>
 </iframe>
 
-## Take Homes
+## Databrary enhances reproducibility
 
-- Video captures behavior
-- Databrary securely stores, shares video
-- Video + Databrary -> reproducible & transparent workflows
+> - Share raw research videos
+> - Share video excerpts for teaching and learning
+> - Share materials, coding spreadsheets, displays
+> - Share links to papers, code repositories
+> - Shared datasets get DOIs
+> - Long-term preservation via partnership with NYU Libraries
 
@@
 
-- Video + Datavyu -> reproducible & transparent workflows
-- Databrary helps make psychological science more transparent & reproducible
+- Video uniquely captures complexity & richness of behavior
+- Databrary securely stores, shares video
+- Databrary and Datavyu enhance reproducibility of video-based research
+- Databrary makes psychological science more transparent & reproducible
 

### 2016-05-01 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
fix Datavyu section

@@
 
-# Reproducible workflows with [Databrary](http://databrary.org)
-
 ## Acknowledgments
-
 - [NSF BCS-1238599](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1238599&HistoricalAwards=false)
@@
 ## Is psychological science reproducible?
-
 <div class="centered">
@@
 ----
-
 <iframe src="http://databrary.org" width=640px height=400px>
@@
 ## [Databrary.org](http://databrary.org)
-
 - Digital data library specialized for research video
@@
 ## Video uniquely captures behavior
-
 <div class="centered">
@@
 ## Video uniquely captures behavior
-
 <div class="centered">
-
 <video width="640" height="480" controls>
@@
 ## [[@c7fbeef0-d8a9-4b99-9e17-f598507486b7]](http://doi.org/10.17910/B7H019)
-
 <div class=centered>
@@
 ## Unequivocally & unambiguously demonstrates phenomena
-
 <div class="centered">
@@
 ----
-
 <div class="centered">
@@
 ----
-
 <iframe src="http://doi.org/10.17910/B7H019" width=640px height=400px>
@@
 
-## Video depicts unreported methodological details
-
-Fall from video, bridge 20 going into bridge 14, then fall. Subject 1. Cari brown sleeveless shirt. John assist.
-<https://nyu.databrary.org/volume/5>
+## Video depicts unreported methodological details [[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/9855/0,85147/asset/38977/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+</div>
 
 ## [Datavyu](http://datavyu.org) makes video coding reproducible
+- Raw research video must be coded
+- [Datavyu.org](http://datavyu.org) = free, [open source](https://github.com/databrary/datavyu) coding tool
+- Add codes, annotations time-locked to video segments
+- Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts) reproducible workflows
 
-- Raw research video must be coded
-- [Datavyu.org](http://datavyu.org) is a free, [open source](https://github.com/databrary/datavyu) coding tool
-- Add codes, annotations time-locked to video segments
-- Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts)
+----
+<div class="centered">
+<a href="http://datavyu.org/user-guide/index.html">
+<img src="http://datavyu.org/user-guide/_images/annotated-datavyu-desktop.png" width=800px>
+</a>
+</div>
 
 ## Reproducible video coding
+<div class="centered">
+<img src="img/crawl-1-hand-fall.jpg">
 
-Static image of fall from video.
+[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
+</div>
 
-## Securing permission to share while limiting access
+----
+<div class="centered">
+<img src="img/timeline-crawl-1.jpg">
 
+[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
+</div>
+----
+<div class="centered">
+<img src="img/datavyu-ss-bridge-crawl-1-trial10.jpg" width=1000px>
+
+[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
+</div>
+----
+<div class="centered">
+<img src="img/databrary-vol-5-materials.jpg" width=1000px>
+
+[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
+</div>
+
+## Securing permission to share while limiting access {.flexbox .vcenter}
 <div class="centered">
@@
 ## Databrary enhances reproducibility
-
 - "Active" curation of data **as it is collected**
 - Organize, share, standardized participant metadata
-- Sharing based on access level & participant permission
+- Sharing based on 
+    + user access level
+    + participant permission
 
 ## [[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
-
 <iframe src="https://nyu.databrary.org/volume/5" width=750px height=500px>
@@
 - Video uniquely captures complexity & richness of behavior
-- Databrary securely stores, shares video
+- Databrary securely stores, shares video + metadata
 - Databrary and Datavyu enhance reproducibility of video-based research

### 2016-05-02 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
add figure builds

@@
 - [Society for Research in Child Development (SRCD)](http://srcd.org)
-- NYU Libraries, ITS, IRB, and OSP
+- NYU Libraries, NYU IT, IRB, and OSP
 
@@
 ## [Datavyu](http://datavyu.org) makes video coding reproducible
+
 - Raw research video must be coded
@@
 ----
+
 <div class="centered">
 <a href="http://datavyu.org/user-guide/index.html">
-<img src="http://datavyu.org/user-guide/_images/annotated-datavyu-desktop.png" width=800px>
+<img src="http://datavyu.org/user-guide/_images/annotated-datavyu-desktop.png" width=1000px>
 </a>
@@
 
-## Reproducible video coding
+## Datavyu facilitates reproducible video coding
+
 <div class="centered">
-<img src="img/crawl-1-hand-fall.jpg">
+<img src="img/crawl-1-hand-fall.jpg" height=550px>
 
@@
 ----
+
 <div class="centered">
-<img src="img/timeline-crawl-1.jpg">
+<img src="img/timeline-crawl-0.jpg" height=550px>
+</div>
 
-[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
+----
+
+<div class="centered">
+<img src="img/timeline-crawl-1.jpg" height=550px>
 </div>
+
 ----
+
 <div class="centered">
-<img src="img/datavyu-ss-bridge-crawl-1-trial10.jpg" width=1000px>
+<img src="img/datavyu-ss-bridge-crawl-1-trial10.jpg" height=550px>
+</div>
 
-[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
-</div>
 ----
+
 <div class="centered">
@@
 
-## Securing permission to share while limiting access {.flexbox .vcenter}
-<div class="centered">
-<video width="640" height="480" controls>
-  <source src="https://databrary.org/video/example-video-1.mp4" type="video/mp4">
-Your browser does not support the video tag.
-</video>
-</div>
+## Databrary enhances reproducibility
 
-## Databrary enhances reproducibility
 - "Active" curation of data **as it is collected**
@@
 </iframe>
+
+----
+
+<div class="centered">
+<img src="img/vol-5-spreadsheet-0.jpg">
+</div>
+
+----
+
+<div class="centered">
+<img src="img/vol-5-spreadsheet-1.jpg">
+</div>
+
+----
+
+<div class="centered">
+<img src="img/vol-5-spreadsheet-2.jpg">
+</div>
+
+----
+
+<div class="centered">
+<img src="img/vol-5-spreadsheet-3.jpg">
+</div>
+
+----
+
+<div class="centered">
+<img src="img/vol-5-spreadsheet-4.jpg">
+</div>
+
+## Standardized -- reproducible -- release levels
+
+<a href="https://databrary.org/access/guide/investigators/release/release-levels.html">
+<div class="centered">
+<img src="img/release-levels.jpg">
+</a>
+</div>
 
@@
 - Databrary makes psychological science more transparent & reproducible
+- <http://github.com/databrary/presentations/nyu-data-science-reproducibility-16>
+
+## To boldly go where no science has gone before...
+
+- Reproducibility of numerical computation ≠ 
+- Reproducibility of behavioral measures, manipulations
+- **All** behavioral researchers should 
+    + share reproducible workflows of data, analyses *AND*
+    + videos of behavioral measures, displays
 

### 2016-05-02 nyu-data-science-reproducibility-16/be-bold.Rmd
add be bold

---
title: "To boldly go where no science..."
author: "Karen E. Adolph, David Millman, and Rick O. Gilmore"
output: 
  ioslides_presentation:
    widescreen: TRUE
    css: css/databrary.css
    logo: img/databrary-leaf-large.png
bibliography: bib/databrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Acknowledgments

- [NSF BCS-1238599](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1238599&HistoricalAwards=false)
- [NICHD U01-HD-076595](https://projectreporter.nih.gov/project_info_description.cfm?aid=8531595&icde=15908155&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC)
- [Society for Research in Child Development (SRCD)](http://srcd.org)
- NYU Libraries, NYU IT, IRB, and OSP

## TL;DR

- Video is essential for reproducible behavioral science
    + Even where video is *not* the primary raw data
- [Databrary](http://databrary.org) specializes in storing and sharing research video
- Video is identifiable, but it *can* be shared
- *All* behavioral scientists should collect and share videos of their research

## Is behavioral science reproducible?

- Nosek et al. [[@collaboration_estimating_2015]](http://doi.org/10.1126/science.aac4716)
- vs. Gilbert et al. [[@gilbert_comment_2016]](http://doi.org/10.1126/science.aad7243)

## Why we're unsure

- Behavior rich, complex
- Numeric, text-based measures reduce that complexity
- **Video captures and preserves it**

## Why we're unsure

- Replications can fail due to methdological differences
- Methods sections don't always report essential details
- **Video captures and preserves it**

## **Essentials** for a reproducible behavioral science

- Video record all tasks, measures, and behaviors
- Share the recordings
- Share all questionnaires, tasks, displays
- Share statistical, computational, data workflows
- Prepare to share from the beginning
- Seek permission to share data

## [Databrary.org](http://databrary.org)

- Digital data library specialized for research video
- Video/audio + participant/context metadata
- Restricted access for research/educational use
- Policy framework for sharing identifiable data
- Developmental science focus, but not exclusive

----

<iframe src="http://databrary.org" width=640px height=400px>
</iframe>

## How Databrary overcomes barriers to sharing video

- Policies for sharing identifiable video data
- Tools for coding video
- Tools for "active curation"" during data collection

## Policies

- Restrict access to authorized researchers (& affiliates)
    + Institutional [access agreement](https://databrary.org/access/policies/agreement.html)
- Seek permission to share data from participants
    + Release [template](https://databrary.org/access/policies/release-template.html); [script](https://databrary.org/access/guide/investigators/release/asking/script.html); sample [video](https://databrary.org/video/example-video-1.mp4)

## Standardized release levels

<div class="centered">
<a href="https://databrary.org/access/guide/investigators/release/release-levels.html">
<img src="img/release-levels.jpg">
</a>
</div>

## Tools for coding video

- Raw research video must be coded
- [Datavyu.org](http://datavyu.org) = free, [open source](https://github.com/databrary/datavyu) coding tool
- Add codes, annotations time-locked to video segments
- Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts) reproducible workflows

----

<div class="centered">
<a href="http://datavyu.org/user-guide/index.html">
<img src="http://datavyu.org/user-guide/_images/annotated-datavyu-desktop.png" width=1000px>
</a>
</div>

## Curating data **as it is collected**

- Organize, share, standardized participant metadata
- Sharing based on 
    + user access level
    + participant permission

## [[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)

<iframe src="https://nyu.databrary.org/volume/5" width=750px height=500px>
</iframe>

----

<div class="centered">
<img src="img/vol-5-spreadsheet-0.jpg">
</div>

----

<div class="centered">
<img src="img/vol-5-spreadsheet-1.jpg">
</div>

----

<div class="centered">
<img src="img/vol-5-spreadsheet-2.jpg">
</div>

----

<div class="centered">
<img src="img/vol-5-spreadsheet-3.jpg">
</div>

----

<div class="centered">
<img src="img/vol-5-spreadsheet-4.jpg">
</div>

----

<div class="centered">
<img src="img/databrary-vol-5-materials.jpg">

[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
</div>

## Take homes

- Video uniquely captures complexity & richness of behavior
- Video captures vital methodological details in behavioral research
- Identifiable video (and other data) can be shared
- Video data sharing essential for a reproducible behavioral science

## **Essentials** for a reproducible behavioral science

- Video record all tasks, measures, and behaviors
- Share the recordings
- Share all questionnaires, tasks, displays
- Share statistical, computational, data workflows
- Prepare to share from the beginning
- Seek permission to share data

## Join us!

<iframe src="https://databrary.org/about/jobs.html" width=640px height=400px>
</iframe>

## References {.smaller}

### 2016-05-02 nyu-data-science-reproducibility-16/video-reproducibility.Rmd
add be bold

@@
 ## Acknowledgments
+
 - [NSF BCS-1238599](http://www.nsf.gov/awardsearch/showAward?AWD_ID=1238599&HistoricalAwards=false)
@@
 ----
+
 <iframe src="http://doi.org/10.17910/B7H019" width=640px height=400px>
@@
 
+## Join us!
+
+<iframe src="https://databrary.org/about/jobs.html" width=640px height=400px>
+</iframe>
+
 ## References {.smaller}

### 2016-05-02 nyu-data-science-reproducibility-16/be-bold.Rmd
revise be-bold

@@
 
-## TL;DR
+## TL; DR
 
@@
 
-- Replications can fail due to methdological differences
+- Replications can fail due to methodological differences
 - Methods sections don't always report essential details
@@
 
-## [Databrary.org](http://databrary.org)
+## [Databrary.org](http://databrary.org) makes this possible
 
@@
 - Video/audio + participant/context metadata
+- Share displays, materials, text-based data files
 - Restricted access for research/educational use
 - Policy framework for sharing identifiable data
-- Developmental science focus, but not exclusive
-
-----
-
-<iframe src="http://databrary.org" width=640px height=400px>
-</iframe>
+- Developmental focus, but not exclusive
 
@@
 - Policies for sharing identifiable video data
-- Tools for coding video
-- Tools for "active curation"" during data collection
+- Tools for reproducibly coding video
+- Tools for "active curation" == during data collection
 
@@
 - Raw research video must be coded
-- [Datavyu.org](http://datavyu.org) = free, [open source](https://github.com/databrary/datavyu) coding tool
+- [Datavyu](http://datavyu.org) a free, [open source](https://github.com/databrary/datavyu) coding tool
 - Add codes, annotations time-locked to video segments
+- Turn behavior into quantifiable data
 - Ruby API for [scripting](https://github.com/databrary/Datavyu-Example-Scripts) reproducible workflows
@@
 
-- Organize, share, standardized participant metadata
+- After-the-fact curation burdensome
+- Databrary organizes, shares, standardized participant metadata
 - Sharing based on
@@
 - Video uniquely captures complexity & richness of behavior
-- Video captures vital methodological details in behavioral research
+- Video captures vital methodological details
 - Identifiable video (and other data) can be shared
+- Preparing to share reduces the burden
 - Video data sharing essential for a reproducible behavioral science
 
-## **Essentials** for a reproducible behavioral science
-
-- Video record all tasks, measures, and behaviors
-- Share the recordings
-- Share all questionnaires, tasks, displays
-- Share statistical, computational, data workflows
-- Prepare to share from the beginning
-- Seek permission to share data
-
-## Join us!
-
-<iframe src="https://databrary.org/about/jobs.html" width=640px height=400px>
-</iframe>
-
 ## References {.smaller}

### 2016-05-03 nyu-data-science-reproducibility-16/be-bold.Rmd
update README, add ? to release levels fig

@@
 - Replications can fail due to methodological differences
-- Methods sections don't always report essential details
+- Methods sections can't possibly report essential details
 - **Video captures and preserves it**
 
-## **Essentials** for a reproducible behavioral science
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/9855/0,85147/asset/38977/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+[[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)
+</div>
+
+## A reproducible behavioral science *must*...
 
@@
 
+----
+
+<iframe src="http://databrary.org" width=640px height=400px>
+</iframe>
+
 ## References {.smaller}

### 2016-05-03 nyu-data-science-reproducibility-16/be-bold.Rmd
add Databrary leaf

@@
 - Numeric, text-based measures reduce that complexity
-- **Video captures and preserves it**
+- **Video captures and preserves it
 
@@
 
+----
+
+<iframe src="https://databrary.org/about/jobs.html" width=640px height=400px>
+</iframe>
+
+----
+
+<div class="centered">
+<img src="img/databrary-leaf-large.png" height=450>
+
+<http://github.com/databrary/presentations/nyu-data-science-reproducibility-16/>
+</div>
+
 ## References {.smaller}

### 2016-08-11 nysds-2016/gilmore-nysds-2016.Rmd
initial NYSDS 16 commit

--- 2016-05-03 nyu-data-science-reproducibility-16/be-bold.Rmd
@@
 ---
-title: "To boldly go where no science..."
-author: "Karen E. Adolph, David Millman, and Rick O. Gilmore"
+title: "gilmore-nysds-2016"
+author: "Rick O. Gilmore, Karen E. Adolph, & David S. Millman"
+date: "`r Sys.Date()`"
 output:
@@
 ```
+
+----
+
+**Curating identifiable data**
 
@@
 - [Databrary](http://databrary.org) specializes in storing and sharing research video
-- Video is identifiable, but it *can* be shared
-- *All* behavioral scientists should collect and share videos of their research
+- Video is identifiable, but it *can* be securely stored, managed, and shared
+- **All** behavioral scientists should collect and share videos of their research
 
@@
 - **Video captures and preserves it**
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/9840/-/asset/11192/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+[[@c7fbeef0-d8a9-4b99-9e17-f598507486b7]](http://doi.org/10.17910/B7H019)
+</div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/9840/-/asset/11193/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+[[@c7fbeef0-d8a9-4b99-9e17-f598507486b7]](http://doi.org/10.17910/B7H019)
+</div>
 
@@
 
-- Video record all tasks, measures, and behaviors
-- Share the recordings
-- Share all questionnaires, tasks, displays
-- Share statistical, computational, data workflows
-- Prepare to share from the beginning
-- Seek permission to share data
+>- Video record all tasks, measures, and behaviors
+>- Share the recordings openly with other researchers
+>- Share all questionnaires, tasks, displays
+>- Share statistical, computational, data workflows
+>- Prepare to share from the beginning
+>- Seek permission to share data from participants
 
@@
 - Share displays, materials, text-based data files
+
+----
+
 - Restricted access for research/educational use
@@
 - Developmental focus, but not exclusive
+
+----
+
+- High capacity, centralized storage
+- Transcoding to common, interoperable formats
+- Long-term preservation
+
+## Databrary's progress
+
+- Support from NSF and NICHD in late 2012, early 2013.
+- Website "live" early 2014.
+- Rapid growth in investigators, institutions, datasets.
+
+----
+
+<iframe src="https://databrary.org" width=750px height=500px>
+</iframe>
 
@@
 - Tools for "active curation" == during data collection
+- Tools for searching, filtering
 
@@
 </div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/223/-/asset/47075/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+</div>
+</iframe>
+
+## Video is just a spatially and temporally rich time series
+
+<div class="centered">
+<img src="http://www.psyrecon.de/uploads/pics/psychophysiologischeVariablen_01.jpg">
+</div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37958/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+</div>
+</iframe>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37962/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+</div>
+</iframe>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37968/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+</div>
+</iframe>
+
+## Datavyu can import multiple time series
+
+- Heart rate, skin conductance, brain activation
+- Eye position, diameter
+- Body position, movement
 
@@
 
+## Common metadata representation makes search, filtering easy
+
+<div class="centered">
+<img src="img/search=1-2-y-olds.jpg">
+**"Show me all the datasets that have videos of 1-2 year-old children"**
+</div>
+
 ## Take homes
@@
 - Video captures vital methodological details
-- Identifiable video (and other data) can be shared
+- Identifiable video (and other data) can be shared securely and ethically
 - Preparing to share reduces the burden
@@
 
+----
+
+<iframe src="https://databrary.org/about/jobs.html" width=640px height=400px>
+</iframe>
+
+----
+
+<div class="centered">
+<img src="img/databrary-leaf-large.png" height=450>
+
+<http://github.com/databrary/presentations/nysds-2016/>
+</div>
+
 ## References {.smaller}

### 2016-08-11 nysds-2016/gilmore-nysds-2016.Rmd
1st full draft of NYSDS 16

@@
 
-----
+## Curating identifiable data with Databrary
 
-**Curating identifiable data**
+<div class="centered">
+<img src="img/databrary-leaf-large.png" height=450>
+</div>
 
@@
 
-----
+## [Databrary.org](http://databrary.org) makes this possible
 
@@
 <a href="https://databrary.org/access/guide/investigators/release/release-levels.html">
-<img src="img/release-levels.jpg">
+<img src="img/release-levels-revised.jpg" width=1000px>
 </a>
@@
 <div class="centered">
-<video width="640" height="480" controls>
-  <source src="https://nyu.databrary.org/slot/223/-/asset/47075/download?inline=true" type="video/mp4">
+<video width="700" height="550" controls>
+  <source src="https://nyu.databrary.org/slot/11652/307774,376273/asset/47075/download?inline=true" type="video/mp4">
 Your browser does not support the video tag.
@@
 <div class="centered">
-<img src="http://www.psyrecon.de/uploads/pics/psychophysiologischeVariablen_01.jpg">
+<a href="http://www.psyrecon.de/uploads/pics/psychophysiologischeVariablen_01.jpg">
+<img src="http://www.psyrecon.de/uploads/pics/psychophysiologischeVariablen_01.jpg" height=500px>
+</a>
 </div>
@@
 - Body position, movement
+- Coming soon: Visualizing Datavyu tags within Databrary
 
@@
 
-## Common metadata representation makes search, filtering easy
+## Searching, filtering by participant characteristics
 
 <div class="centered">
-<img src="img/search=1-2-y-olds.jpg">
-**"Show me all the datasets that have videos of 1-2 year-old children"**
+<img src="img/search-1-2-y-olds.jpg" height=450px>
 </div>
@@
 - Preparing to share reduces the burden
-- Video data sharing essential for a reproducible behavioral science
+- Video data sharing essential for a reproducible science
 

### 2016-08-12 nysds-2016/gilmore-nysds-2016.Rmd
More tweaks

@@
 >- Share the recordings openly with other researchers
+
+## *Journal of Visualized Experiments*
+
+<iframe src="http://www.jove.com/" width=750px height=500px>
+</iframe>
+
+## But, a reproducible behavioral science must *also*...
+
 >- Share all questionnaires, tasks, displays
@@
 
-## [Databrary.org](http://databrary.org) makes this possible
+## [Databrary.org](http://databrary.org)
 
@@
 
+## Databrary's structure
+
+- Datasets have own page; shared datasets have DOIs.
+- Data about testing *sessions* (locations, dates/times, people) stored in *spreadsheet*
+- Session data organized in *timeline*
+- Store data AND materials (displays, protocols, etc.)
+
 ## [[@ddff6ac6-cd1b-4365-89aa-70ea654b4ed9]](http://dx.doi.org/10.17910/B7MW2K)

### 2016-08-15 nysds-2016/gilmore-nysds-2016.Rmd
full draft

@@
 
+<div class="notes">
+- I want to acknowledge support from NSF, NICHD, and the Society for Research in Child Development. 
+- I'd also like to thank the NYU Libraries, NYU IT, IRB, and OSP for their advice and support.
+- Finally, thank you to the organizers and the sponsors for making this meeting possible.
+</div>
+
 ## TL; DR
 
-- Video is essential for reproducible behavioral science
+- Psychology is harder than physics
+- Video uniquely informative about behavior
+- Video is essential for reproducible science of behavior
     + Even where video is *not* the primary raw data
-- [Databrary](http://databrary.org) specializes in storing and sharing research video
+- [Databrary](http://databrary.org) specializes in storing and sharing research video + metadata
+
+<div class="notes">
+- It's late on the 2nd day, and you may be tired of hearing people talk to you. So, here's the "too long; didn't read" summary.
+</div>
+
+## TL; DR
+
 - Video is identifiable, but it *can* be securely stored, managed, and shared
-- **All** behavioral scientists should collect and share videos of their research
+- **All** ~~behavioral~~ scientists should collect and share videos of their research methods.
+- Video (+ other streams) can serve as the core of a multivariate/multi-level big data science of behavior
+
+## Why psychology is harder than physics {.flexbox .vcenter}
+
+<div class="centered">
+<img src="img/psych-harder-1.jpg" width=800px//>
+</div>
+
+## Why psychology is harder than physics {.flexbox .vcenter}
+
+<div class="centered">
+<img src="img/psych-harder-2.jpg" width=800px//>
+</div>
+
+## Video is a uniquely informative source of evidence about behavior
+
+----
+
+<div class="centered">
+<a href="http://doi.org/10.1126/science.1093567">
+<img src="img/deloache.science.jpg" height=350px>
+</a>
+
+[[@deloache_scale_2004]](http://doi.org/10.1126/science.1093567)
+</div>
+
+<div class="notes">
+- If a picture is worth 1,000 words, a video is worth 1,000 pictures.
+</div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/9850/-/asset/11550/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+[[@b03c5c6c-31b8-445e-9e4f-8917b735bb53]](http://doi.org/10.17910/B7H019)
+</div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/11647/0,84928/asset/40653/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+[[@b155b094-099c-4015-8d05-143257d99f28]](http://doi.org/10.17910/B7.236)
+</div>
+
+<div class="notes">
+- Here's another example from Chen Yu and Linda Smith.
+- It shows the results from two, temporally synched head-mounted eye trackers worn by a mother and her infant while the two play with a set of toys.
+- Do infants look at the toys when their mothers direct them to? Do infants look at toys when moms speak?
+- Might this form of "joint attention" be a mechanism for language learning?
+</div>
+
+----
+
+<div class="centered">
+<iframe src="http://pediatrics.aappublications.org/content/early/2016/08/11/peds.2016-1533" width=750px height=500px>
+</iframe>
+
+[[@batra_nocturnal_2016]](http://doi.org/10.1542/peds.2016-1533)
+</div>
+
+<div class="notes">
+- Video data can answer questions of considerable importance to public health.
+- Do parents put infants to sleep in postures that are safe or risky?
+- These authors got permission to record babies in their cribs to find out, and the answer is surprising -- not as often as they probably should.
+- These particular videos aren't shared -- they were collected by my colleagues at Penn State long before we were even thinking about Databrary -- but without video, the question could not really be asked.
+</div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37958/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+[[@6243604f-bfd6-488f-b744-50305e9eefaa]](http://doi.org/10.17910/B7G59R)
+</div>
+
+<div class="notes">
+- Here are videos of a gaze time series for participants from one of three groups of children: typically developing, those with Down Syndrome, and those with autism spectrum disorder.
+- See if you can tell which is which.
+- This is group 1.
+</div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37962/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+[[@6243604f-bfd6-488f-b744-50305e9eefaa]](http://doi.org/10.17910/B7G59R)
+</div>
+
+<div class="notes">
+- This is group 2.
+</div>
+
+----
+
+<div class="centered">
+<video width="640" height="480" controls>
+  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37968/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+[[@6243604f-bfd6-488f-b744-50305e9eefaa]](http://doi.org/10.17910/B7G59R)
+</div>
+
+<div class="notes">
+- This is group 3.
+- Can you tell which is the typically developing group? It was group 1. Group 2 was the Down syndrome group and 3 was the group with ASD.
+- Sometimes video data can show *similarities* even when we might otherwise expect differences.
+</div>
 
@@
 
-## Why we're unsure
+<div class="notes">
+- Now, I'm certain that many of you are familiar with the recent back and forth about the degree to which psychological science is reproducible.
+</div>
+
+----
+
+<div class="centered">
+<a href="https://www.theguardian.com/science/2015/aug/27/study-delivers-bleak-verdict-on-validity-of-psychology-experiment-results">
+<img src="img/bleak-verdict.jpg" height=450px>
+</a>
+</div>
+
+----
+
+<div class="centered">
+<a href="http://www.nature.com/news/psychology-s-reproducibility-problem-is-exaggerated-say-psychologists-1.19498">
+<img src="img/reproducibility-exaggerated.jpg" height=450px>
+</a>
+</div>
+
+## Why the dispute; what to do?
 
@@
 - Numeric, text-based measures reduce that complexity
+
 - **Video captures and preserves it**
+
+<div class="notes">
+- Why can't we agree?
+- One reason is that behavior is rich, multitextured, and complex.
+- Our measures often reduce that complexity.
+- But **video** can preserve it
+- Oftentimes, video tells the story better than words
+</div>
 
@@
 [[@c7fbeef0-d8a9-4b99-9e17-f598507486b7]](http://doi.org/10.17910/B7H019)
+</div>
+
+<div class="notes">
+- Here's an example from Michael Frank.
+- What is this boy doing?
 </div>
@@
 
-## Why we're unsure
+<div class="notes">
+- What is he doing now?
+</div>
+
+----
+
+<iframe src="https://nyu.databrary.org/volume/33" width=640px height=400px>
+</iframe>
+
+<div class="notes">
+- Now you completely understand the title of this paper -- "Representing exact number visually using mental abacus" -- don't you?
+</div>
+
+## Why the dispute; what to do?
 
 - Replications can fail due to methodological differences
-- Methods sections can't possibly report essential details
+- Methods sections can't possibly report all essential details
 - **Video captures and preserves it**
 
-----
+<div class="notes>
+- Even when researcher try to replicate results with help from the original investigators, the results can fall short?
+- Why? Methods sections can't possibly report all essential details due to time and space constraints.
+</div>
+
+## Adolph study
 
@@
 
+<div class="notes">
+- Here is a study from Karen Adolph's lab group examining how well infants can perceive when it is safe or risky to cross a bridge.
+- This is the raw video. Notice that it consists of multiple camera angles.
+- Watch how the experimenters coordinate their actions, keep the baby happy and entertained, adjust the apparatus between trials, and so forth.
+</div>
+
 ## A reproducible behavioral science *must*...
 
->- Video record all tasks, measures, and behaviors
->- Share the recordings openly with other researchers
+- Video record all tasks, measures, and behaviors
+- Share the recordings openly with other researchers
 
@@
 
-## But, a reproducible behavioral science must *also*...
+<div class="notes">
+- One solution is to create a journal that publishes video records of scientific protocols. The Journal of Visualized Experiments does just this, and it does it well.
+- But, the high production values of JOVE come at equally high cost to researchers and libraries.
+</div>
 
->- Share all questionnaires, tasks, displays
->- Share statistical, computational, data workflows
->- Prepare to share from the beginning
->- Seek permission to share data from participants
+## A reproducible behavioral science must *also*...
+
+- Share all questionnaires, tasks, displays
+- Share statistical, computational, data workflows
+- Make it easy to share from the beginning
+- Seek permission to share data from participants
+- Store data securely
+
+----
+
+<iframe src="http://databrary.org" width=640px height=400px>
+</iframe>
 
@@
 - Share displays, materials, text-based data files
-
-## [Databrary.org](http://databrary.org) makes this possible
-
-- Restricted access for research/educational use
 - Policy framework for sharing identifiable data
@@
 
-## [Databrary.org](http://databrary.org) makes this possible
+## [Databrary.org](http://databrary.org) facilitates data sharing, re-use, preservation
 
@@
 
-## Databrary's progress
-
-- Support from NSF and NICHD in late 2012, early 2013.
-- Website "live" early 2014.
-- Rapid growth in investigators, institutions, datasets.
-
-----
-
-<iframe src="https://databrary.org" width=750px height=500px>
-</iframe>
-
-## How Databrary overcomes barriers to sharing video
+## Databrary helps to overcomes barriers to sharing video
 
@@
 - Restrict access to authorized researchers (& affiliates)
+    + PIs and the trainees they supervise
     + Institutional [access agreement](https://databrary.org/access/policies/agreement.html)
@@
 
-## Standardized release levels
+## Standardized (reproducible) release levels
 
@@
 
-- Raw research video must be coded
+- Raw research video must be coded by human observers
 - [Datavyu](http://datavyu.org) a free, [open source](https://github.com/databrary/datavyu) coding tool
@@
 
-## Video is just a spatially and temporally rich time series
-
-<div class="centered">
-<a href="http://www.psyrecon.de/uploads/pics/psychophysiologischeVariablen_01.jpg">
-<img src="http://www.psyrecon.de/uploads/pics/psychophysiologischeVariablen_01.jpg" height=500px>
-</a>
+<div class="notes">
+- Here's an example of a short segment of video that has been coded for a some behaviors of interest -- speech, actions involving objects, and locomotion.
+- Notice how the annotations, coupled to the video segments, reveals the complex interrelationships among behaviors.
+- And, we've only focused on a subset of them.
 </div>
 
-----
-
-<div class="centered">
-<video width="640" height="480" controls>
-  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37958/download?inline=true" type="video/mp4">
-Your browser does not support the video tag.
-</video>
-</div>
-</iframe>
-
-----
-
-<div class="centered">
-<video width="640" height="480" controls>
-  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37962/download?inline=true" type="video/mp4">
-Your browser does not support the video tag.
-</video>
-</div>
-</iframe>
-
-----
-
-<div class="centered">
-<video width="640" height="480" controls>
-  <source src="https://nyu.databrary.org/slot/11141/0,5000/asset/37968/download?inline=true" type="video/mp4">
-Your browser does not support the video tag.
-</video>
-</div>
-</iframe>
-
-## Datavyu can import multiple time series
-
-- Heart rate, skin conductance, brain activation
-- Eye position, diameter
-- Body position, movement
-- Coming soon: Visualizing Datavyu tags within Databrary
-
-## Curating data **as it is collected**
+## Tools for curating data **as it is collected**
 
@@
     + participant permission
+
+<div class="notes">
+- Behavioral researchers spend hours coding videos, and so spending even more time after the fact to curate it for sharing with others is hugely burdensome.
+- So, Databrary tries to make it easy and convenient for researchers to curate their data while it is collected.
+</div>
 
@@
 
+<div class="notes"
+- Here is an illustrative dataset.
+- Datasets have own page; shared datasets have DOIs.
+- Datsets can have *highlights* audio, video, or photos that represent the main points of a study.
+- Data about testing *sessions* (locations, dates/times, people) stored in *spreadsheet*
+- Session data organized in *timeline*
+- Store data AND materials (displays, protocols, etc.)
+</div>
 ----
@@
 
-## Searching, filtering by participant characteristics
+----
+
+<div class="centered">
+<iframe src="https://nyu.databrary.org/volume/81/slot/7580/-?asset=15968" width=750px height=500px>
+</iframe>
+
+[[@264e9563-8bc9-4863-b70f-e92203c3f283]](http://doi.org/10.17910/B7988V)
+</div>
+
+<div class="notes">
+- Here's an example of a session from one of my own shared datasets.
+- It represents a set of videos collected from head-mounted cameras worn by infants in India.
+- Notice that there are many video segments, and Databrary represents the temporal relationships among them.
+</div>
+
+## Tools for searching, filtering by participant characteristics
 
@@
 </div>
+
+<div class="notes">
+- It's all well and good to put data into a repository, but how do we get it out?
+- It's been clear for some time that searching and filtering have revolutionized the way we access knowledge in almost every domain.
+- I suggest that the same will be true in data science.
+- At the moment, Databrary offers the ability to search for datasets based on participant or file characteristics.
+- But, this search capability will surely grow, and so in the very near future, a researcher will be able to search for individual sessions or even segments of time when a particular event is occurring.
+</div>
+
+----
+
+<iframe width="560" height="315" src="https://www.youtube.com/embed/ZAmZucyzyZM" frameborder="0" allowfullscreen></iframe>
+
+## "Big" questions about behavior require big(ger) data
+
+<div class="centered">
+<img src="http://lifesciences.ieee.org/images/newsletter/apr2012/thakor_02_lg.jpg" height=450px>
+</div>
+
+## "Big" questions about behavior require big(ger) data
+
+- Multiple levels of spatial, temporal resolution
+- Linked data sets and streams
+- Key linkage variables: *when* (time), *where* (location), *who* (age, gender)
+- Video is a spatially and temporally rich time series
+
+<div class="notes">
+- Video is just one of a number of spatially and temporally rich time series that scientists collect about human behavior.
+- Going forward, we need to build infrastructure that allows data repositories to capture and researchers to visualize and analyze multivariate time series.
+</div>
+
+## How data scientists can help...
+
+- Semi-automated video annotation
+- Speech stream extraction from natural video, transcription, ms-level precision
+- Multivariate time series visualization, data reduction, analysis
+- Linking data, while preserving privacy
 
@@
 - Preparing to share reduces the burden
-- Video data sharing essential for a reproducible science
+
+## Take homes
+
+- Video data sharing essential for a reproducible, robust behavioral science
+- Video can serve as the core of a big data revolution in behavioral science
+
+----
+
+<iframe src="http://wires.wiley.com/WileyCDA/WiresArticle/articles.html?doi=10.1002%2Fwcs.1379" width=640px height=400px>
+</iframe>
 
@@
 
-<iframe src="https://databrary.org/about/jobs.html" width=640px height=400px>
-</iframe>
+<div class="centered">
+<img src="img/databrary-leaf-large.png" height=450>
+</div>
 
@@
 
-<div class="centered">
-<img src="img/databrary-leaf-large.png" height=450>
-
-<http://github.com/databrary/presentations/nysds-2016/>
-</div>
+<iframe src="https://github.com/databrary/presentations/tree/master/nysds-2016" width=640px height=400px>
+</iframe>
 

### 2016-08-16 nysds-2016/gilmore-nysds-2016.Rmd
final draft
final pre-talk

@@
 
-<div class="notes>
-- Even when researcher try to replicate results with help from the original investigators, the results can fall short?
-- Why? Methods sections can't possibly report all essential details due to time and space constraints.
-</div>
-
-## Adolph study
+----
 
@@
 
+## What is the input to the developing brain?
+
+<div class="centered">
+<video width="700" height="550" controls>
+  <source src="https://nyu.databrary.org/slot/11680/0,24500/asset/41871/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+[[@264e9563-8bc9-4863-b70f-e92203c3f283]](http://doi.org/10.17910/B7988V)
+</div>
+
+----
+
+<div class="centered">
+<video width="700" height="550" controls>
+  <source src="https://nyu.databrary.org/slot/11680/25500,50000/asset/41873/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+[[@264e9563-8bc9-4863-b70f-e92203c3f283]](http://doi.org/10.17910/B7988V)
+</div>
+
+----
+
+<div class="centered">
+<video width="700" height="550" controls>
+  <source src="https://nyu.databrary.org/slot/11680/51000,75500/asset/41875/download?inline=true" type="video/mp4">
+Your browser does not support the video tag.
+</video>
+
+[[@264e9563-8bc9-4863-b70f-e92203c3f283]](http://doi.org/10.17910/B7988V)
+</div>
+
+----
+
+<iframe src="http://ieeexplore.ieee.org/xpl/articleDetails.jsp?arnumber=7345450" width=640px height=400px>
+</iframe>
+
 ## Tools for searching, filtering by participant characteristics
@@
 - But, this search capability will surely grow, and so in the very near future, a researcher will be able to search for individual sessions or even segments of time when a particular event is occurring.
+- The results of this search and filtering will be a custom dataset that can be reanalyzed and reused.
 </div>
-
-----
-
-<iframe width="560" height="315" src="https://www.youtube.com/embed/ZAmZucyzyZM" frameborder="0" allowfullscreen></iframe>
 
@@
 - Linked data sets and streams
-- Key linkage variables: *when* (time), *where* (location), *who* (age, gender)
+- Key linkage variables: *when* (time), *where* (location), *who* (age, gender, race/ethnicity, ...)
 - Video is a spatially and temporally rich time series
@@
 
-<iframe src="https://github.com/databrary/presentations/tree/master/nysds-2016" width=640px height=400px>
-</iframe>
+<div class="centered">
+<img src="img/github-screenshot.jpg" width=900px>
+
+</div>
 
